---
title: "Analysis 5: Subgroup Fairness Analysis"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.width=8, fig.height=5)
```

## Overview

Subgroup performance of the Analysis-1 nomogram's out-of-sample risk is assessed here, stratified by sex and age band, along the three axes that together define a fair clinical risk model:

1. **Discrimination parity** --- AUROC per subgroup (does the score rank cases equally well in every group?).
2. **Calibration parity** --- calibration-in-the-large, the ratio of mean predicted risk to observed event rate (does the score systematically over- or under-predict in any group?).
3. **Error-rate parity** --- at a single shared decision threshold, the sensitivity (true-positive rate) and false-positive rate per subgroup, an equal-opportunity / equalized-odds view (does a deployed alarm miss cases or raise false alarms unevenly across groups?).

Discrimination alone is insufficient: two subgroups can share an AUROC yet, at the same alarm threshold, differ sharply in how often the model misses a case or fires falsely. The shared threshold is fixed once on the full cohort to yield 50 per cent overall sensitivity, matching the practical operating point of Analysis 4. External validation on eICU is deferred (no local database available).

## Load Libraries

```{r load-libraries, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({ library(pROC) })
```

## Locate and Read Data

```{r load-data}
find_data <- function(f) {
  cands <- c(file.path("../../../processed_data", f),
             file.path("processed_data", f), f)
  hit <- cands[file.exists(cands)][1]; if (is.na(hit)) stop("cannot find ", f); hit
}
OUT <- dirname(find_data("10_analysis1_oos_pred.parquet"))

pred <- as.data.frame(arrow::read_parquet(find_data("10_analysis1_oos_pred.parquet")))
lm6  <- as.data.frame(arrow::read_parquet(find_data("08_landmark_h6.parquet")))[, c("stay_id","age","gender")]
d <- merge(pred, lm6, by="stay_id")
d$age_band <- cut(d$age, c(-Inf,50,65,80,Inf),
                  labels=c("<50","50-64","65-79","80+"))
```

## Shared Decision Threshold

Error-rate parity is only meaningful at a common operating point, so a single threshold is fixed on the whole cohort before any subgroup split. It is set to the value that flags 50 per cent of the true onset cases overall (the median predicted risk among positives), matching the practical alarm point of Analysis 4. The same threshold is then applied unchanged to every subgroup.

```{r shared-threshold}
THR <- as.numeric(median(d$risk_en[d$onset_within_h == 1]))
cat(sprintf("Shared decision threshold (50%% overall sensitivity): %.4f\n", THR))
```

## Define Subgroup Metrics Function

For each subgroup the function reports discrimination (AUROC), calibration-in-the-large (mean predicted risk over observed rate; 1.0 = the average prediction matches the true rate, above 1 = over-prediction, below 1 = under-prediction), and, at the shared threshold, error-rate parity via the sensitivity (true-positive rate) and false-positive rate.

```{r define-metrics}
grp_metrics <- function(sub, label, thr) {
  y <- sub$onset_within_h; p <- sub$risk_en
  au <- tryCatch(as.numeric(auc(roc(y, p, quiet=TRUE, direction="<"))), error=function(e) NA)
  pos <- y == 1; neg <- y == 0; flag <- p >= thr
  tpr <- if (any(pos)) mean(flag[pos]) else NA   # sensitivity / equal opportunity
  fpr <- if (any(neg)) mean(flag[neg]) else NA   # false-positive rate
  data.frame(subgroup=label, n=nrow(sub), event_rate=round(mean(y),4),
             auroc=round(au,4), mean_pred=round(mean(p),4),
             obs_rate=round(mean(y),4),
             calib_ratio=round(mean(p)/mean(y),3),
             sensitivity=round(tpr,4), fpr=round(fpr,4))
}
```

## Compute Subgroup Metrics and Write Results

Fairness metrics are computed for the full cohort and for each sex and age-band subgroup, then saved to `analysis5_subgroups.csv`.

```{r compute-metrics}
rows <- list(grp_metrics(d, "ALL", THR))
for (g in sort(unique(d$gender)))
  rows[[length(rows)+1]] <- grp_metrics(d[d$gender==g,], paste0("sex=",g), THR)
for (b in levels(d$age_band))
  rows[[length(rows)+1]] <- grp_metrics(d[d$age_band==b,], paste0("age=",b), THR)
tab <- do.call(rbind, rows)
print(tab, row.names=FALSE)
write.csv(tab, file.path(OUT, "16_analysis5_subgroups.csv"), row.names=FALSE)

cat(sprintf("\nEqual-opportunity spread: sensitivity range %.3f-%.3f, FPR range %.3f-%.3f\n",
            min(tab$sensitivity[-1], na.rm=TRUE), max(tab$sensitivity[-1], na.rm=TRUE),
            min(tab$fpr[-1], na.rm=TRUE), max(tab$fpr[-1], na.rm=TRUE)))
cat("\nAnalysis 5 (fairness) complete. External eICU validation deferred (no data).\n")
```
