---
title: "01 Setup - Environment Setup and Data Verification"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)   # student ID for reproducibility

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook checks that all required packages, data directories, and file schemas are in place before the pipeline runs.

```{r load-config}
source("config.R")
source("utils.R")

step("=== Phase 1: Environment Setup ===")
```

## Required Packages

All required packages are checked and missing ones are installed automatically.

```{r check-packages}
REQUIRED_PKGS <- c(
  "data.table", "arrow", "dplyr", "tidyr", "ggplot2", "lubridate",
  "survival", "nnet", "VGAM", "discSurv", "rms", "cmprsk",
  "xgboost", "pmsampsize", "dcurves", "timeROC", "CalibrationCurves",
  "purrr", "tibble", "readr"
)

step("Checking required packages...")
pkg_status <- vapply(REQUIRED_PKGS, requireNamespace, logical(1), quietly = TRUE)
missing_pkgs <- names(pkg_status)[!pkg_status]

if (length(missing_pkgs) > 0) {
  v2_log("Installing missing packages: ", paste(missing_pkgs, collapse = ", "))
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org", quiet = TRUE)
}

pkg_report <- data.frame(
  package = REQUIRED_PKGS,
  status  = ifelse(pkg_status, "OK", "MISSING"),
  stringsAsFactors = FALSE
)
knitr::kable(pkg_report, caption = "Package availability")
```

## Data Directories

MIMIC-IV (Johnson et al., 2023) and eICU-CRD (Pollard et al., 2018) must be accessible at the paths defined in `config.R`.

```{r check-dirs}
step("Verifying data directories...")

check_dir <- function(label, path) {
  exists <- dir.exists(path)
  v2_log(sprintf("  %-30s %s  (%s)", label,
                 if (exists) "[OK]" else "[MISSING]", path))
  exists
}

mimic_ok <- check_dir("MIMIC-IV root",  MIMIC_DIR)
hosp_ok  <- check_dir("MIMIC-IV/hosp", HOSP_DIR)
icu_ok   <- check_dir("MIMIC-IV/icu",  ICU_DIR)
eicu_ok  <- check_dir("eICU root",      EICU_DIR)

dir_report <- data.frame(
  directory = c("MIMIC-IV root", "MIMIC-IV/hosp", "MIMIC-IV/icu", "eICU root"),
  path      = c(MIMIC_DIR, HOSP_DIR, ICU_DIR, EICU_DIR),
  status    = c(mimic_ok, hosp_ok, icu_ok, eicu_ok)
)
knitr::kable(dir_report, caption = "Data directory status")
```

## Key File Spot-Check

Checks that key source tables exist and reports their file sizes.

```{r check-mimic-files}
step("Spot-checking key MIMIC-IV files...")

KEY_FILES <- list(
  admissions         = hosp_file("admissions.csv"),
  patients           = hosp_file("patients.csv"),
  icustays           = icu_file("icustays.csv"),
  chartevents        = icu_file("chartevents.csv"),
  labevents          = hosp_file("labevents.csv"),
  prescriptions      = hosp_file("prescriptions.csv"),
  microbiologyevents = hosp_file("microbiologyevents.csv"),
  inputevents        = icu_file("inputevents.csv")
)

file_report <- lapply(names(KEY_FILES), function(nm) {
  path   <- KEY_FILES[[nm]]
  exists <- file.exists(path)
  sz     <- if (exists) file.size(path) else 0
  v2_log(sprintf("  %-25s %s  (%.1f MB)", nm,
                 if (exists) "[OK]" else "[MISSING]", sz / 1e6))
  data.frame(table = nm, exists = exists, size_mb = round(sz / 1e6, 1))
})
file_df <- do.call(rbind, file_report)
knitr::kable(file_df, caption = "MIMIC-IV key files")
```

```{r check-eicu-files}
step("Spot-checking eICU files...")

EICU_KEY_FILES <- list(
  patient       = file.path(EICU_DIR, "patient.csv.gz"),
  lab           = file.path(EICU_DIR, "lab.csv.gz"),
  medication    = file.path(EICU_DIR, "medication.csv.gz"),
  vitalPeriodic = file.path(EICU_DIR, "vitalPeriodic.csv.gz"),
  microLab      = file.path(EICU_DIR, "microLab.csv.gz"),
  diagnosis     = file.path(EICU_DIR, "diagnosis.csv.gz")
)

eicu_report <- lapply(names(EICU_KEY_FILES), function(nm) {
  path   <- EICU_KEY_FILES[[nm]]
  exists <- file.exists(path)
  sz     <- if (exists) file.size(path) else 0
  v2_log(sprintf("  %-25s %s  (%.1f MB)", nm,
                 if (exists) "[OK]" else "[MISSING]", sz / 1e6))
  data.frame(table = nm, exists = exists, size_mb = round(sz / 1e6, 1))
})
eicu_df <- do.call(rbind, eicu_report)
knitr::kable(eicu_df, caption = "eICU key files")
```

## Schema Verification

Confirms that key columns exist and that `anchor_year_group` (needed for temporal splitting) is present.

```{r check-schema}
step("Peeking at icustays to verify schema...")

icustays_path <- icu_file("icustays.csv")
if (file.exists(icustays_path)) {
  stays_sample <- data.table::fread(icustays_path, nrows = 5, showProgress = FALSE)
  cat("icustays columns:\n")
  print(names(stays_sample))

  patients_path <- hosp_file("patients.csv")
  if (file.exists(patients_path)) {
    pts_sample      <- data.table::fread(patients_path, nrows = 5, showProgress = FALSE)
    has_anchor      <- "anchor_year_group" %in% names(pts_sample)
    v2_log("anchor_year_group present: ", has_anchor)
    cat(sprintf("anchor_year_group present: %s\n", has_anchor))
  }
} else {
  cat("icustays.csv not found - check MIMIC_DIR path.\n")
}
```

## Output Directory Writability

Verifies the output directory is writable.

```{r check-output-dir}
step("Checking output directory...")

test_file <- file.path(OUTPUT_DIR, ".write_test")
tryCatch({
  writeLines("test", test_file)
  file.remove(test_file)
  cat(sprintf("Output dir writable: OK\n%s\n", OUTPUT_DIR))
}, error = function(e) {
  cat(sprintf("Output dir NOT writable: %s\n", e$message))
})
```

## Summary

```{r summary}
step("=== Setup Summary ===")

summary_df <- data.frame(
  item    = c("Project root", "MIMIC-IV", "eICU", "Output dir", "R version", "Label variants"),
  value   = c(
    PROJECT_ROOT,
    if (hosp_ok && icu_ok) "AVAILABLE" else "MISSING",
    if (eicu_ok) "AVAILABLE" else "MISSING",
    OUTPUT_DIR,
    as.character(getRversion()),
    "A (Narrow), B (Seymour-Standard), C (Liberal)"
  )
)
knitr::kable(summary_df, caption = "Setup summary")

save_parquet(file_df, "01_setup_file_report", OUTPUT_DIR)
cat("\nSaved: 01_setup_file_report.parquet\n")
```
