---
title: "02 Cohort Labels - Cohort Extraction Under Sepsis-3 and Alternative Label Variants"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook applies three Sepsis-3 label variants to the same adult ICU cohort from MIMIC-IV. Following Cohen et al. (2024), the variants use different antibiotic-culture windows and SOFA evaluation periods, producing cohorts that differ in size, event rate, and onset timing.

All three cohorts share the same inclusion criteria (adults $\geq$ 18, first ICU stay, LOS $\geq$ 4 h). The label-variance summary table produced here (M1 deliverable) quantifies these differences before any modelling.

Because all three are Sepsis-3 readings anchored to the same culture-plus-antibiotic trigger, two further variants are built alongside them under Protocol Amendment 2: **E (Anchor-Only)**, the suspicion-of-infection time with the organ-dysfunction requirement removed, and **D (Deterioration)**, acute organ dysfunction with no infection criterion and no treatment timestamp. They are reported separately and never enter the pre-registered H1 comparison.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(dplyr)
})

source("config.R")
source("utils.R")

step(sprintf("=== Phase 2a: Cohort extraction under %d label variants ===",
             length(active_variants())))
```

## Base Tables

The three source tables (patients, admissions, and ICU stays) provide demographics, hospital-level outcomes, and the time spine for each ICU episode.

```{r load-base-tables}
step("Loading base tables...")

patients <- read_mimic(hosp_file("patients.csv"),
                       cols = c("subject_id", "gender", "anchor_age",
                                "anchor_year", "anchor_year_group"))

admissions <- read_mimic(hosp_file("admissions.csv"),
                         cols = c("subject_id", "hadm_id", "admittime",
                                  "dischtime", "race", "language",
                                  "insurance", "hospital_expire_flag"))
admissions[, admittime := parse_dt(admittime)]
admissions[, dischtime := parse_dt(dischtime)]

icustays <- read_mimic(icu_file("icustays.csv"),
                       cols = c("subject_id", "hadm_id", "stay_id",
                                "intime", "outtime", "los"))
icustays[, intime    := parse_dt(intime)]
icustays[, outtime   := parse_dt(outtime)]
icustays[, los_hours := hours_diff(intime, outtime)]
```

## Base Cohort Criteria

Standard inclusion filters: adults ($\geq$ 18), first ICU stay per patient, and minimum ICU LOS $\geq$ 4 hours.

```{r build-base-cohort}
step("Building base cohort (adult, first ICU stay, min LOS)...")

stays <- merge(icustays, patients, by = "subject_id", all.x = TRUE)
stays <- merge(stays, admissions[, .(hadm_id, race, language, insurance,
                                      hospital_expire_flag)],
               by = "hadm_id", all.x = TRUE)

stays <- stays[anchor_age >= MIN_AGE]

setorderv(stays, c("subject_id", "intime"))
stays[, icu_rank := seq_len(.N), by = subject_id]
if (FIRST_ICU_STAY_ONLY) stays <- stays[icu_rank == 1]

stays <- stays[los_hours >= MIN_ICU_LOS_HOURS]

cat(sprintf("Base cohort: %s stays, %s unique patients\n",
            format(nrow(stays), big.mark = ","),
            format(stays[, uniqueN(subject_id)], big.mark = ",")))
```

## Suspected Infection

Sepsis-3 (Singer et al., 2016) requires temporal co-occurrence of an antibiotic order and a microbiology culture. All antibiotic-culture pairs within each admission are generated here; variant-specific time-window filters are applied in the cohort-building step. Surveillance cultures are excluded.

```{r suspected-infection}
step("Identifying suspected infection episodes...")

prescriptions <- read_mimic(hosp_file("prescriptions.csv"),
                             cols = c("hadm_id", "starttime", "stoptime",
                                      "drug", "route"))
prescriptions[, starttime := parse_dt(starttime)]
prescriptions[, is_abx := match_antibiotics(drug, ANTIBIOTICS)]
abx <- prescriptions[is_abx == TRUE & !is.na(starttime),
                     .(hadm_id, abx_time = starttime)]
abx <- unique(abx)
cat(sprintf("Antibiotic orders: %s\n", format(nrow(abx), big.mark = ",")))

microbio <- read_mimic(hosp_file("microbiologyevents.csv"),
                       cols = c("hadm_id", "charttime", "chartdate",
                                "spec_type_desc"))
microbio[, charttime := parse_dt(charttime)]
if (nrow(microbio) > 0)
  microbio <- microbio[!(spec_type_desc %in% SURVEILLANCE_SPEC_TYPES)]
microbio[is.na(charttime) & !is.na(chartdate),
         charttime := as.POSIXct(chartdate, format = "%Y-%m-%d", tz = "UTC")]
cultures <- microbio[!is.na(charttime), .(hadm_id, culture_time = charttime)]
cultures <- unique(cultures)
cat(sprintf("Culture samples: %s\n", format(nrow(cultures), big.mark = ",")))

abx[, abx_id := .I]
cultures[, cul_id := .I]

suspicion_raw <- merge(abx, cultures, by = "hadm_id", allow.cartesian = TRUE)
suspicion_raw[, dt_abx_before_cul := hours_diff(abx_time, culture_time)]
suspicion_raw[, dt_cul_before_abx := hours_diff(culture_time, abx_time)]
cat(sprintf("ABX-culture pairs (raw): %s\n",
            format(nrow(suspicion_raw), big.mark = ",")))
```

## SOFA Components

The SOFA score (Vincent et al., 1996) quantifies organ dysfunction across six systems. Lab values (platelets, bilirubin, creatinine, PaO$_2$/FiO$_2$), vitals (MAP, GCS), and vasopressor records are loaded and filtered to the cohort.

```{r sofa-components}
step("Computing SOFA components for all stays...")

cohort_hadm_ids <- stays[, unique(hadm_id)]
cat(sprintf("Pre-filtering large tables to %d cohort hadm_ids...\n",
            length(cohort_hadm_ids)))

lab_sofa_items <- c(LAB_ITEMIDS$platelets, LAB_ITEMIDS$bilirubin_total,
                    LAB_ITEMIDS$creatinine, LAB_ITEMIDS$pao2, LAB_ITEMIDS$fio2_lab)

v2_log("Loading labevents for SOFA components via DuckDB...")
labevents <- read_large_csv_filtered(
  path     = hosp_file("labevents.csv"),
  cols     = c("hadm_id", "itemid", "charttime", "valuenum"),
  itemids  = lab_sofa_items,
  hadm_ids = cohort_hadm_ids
)
labevents[, charttime := parse_dt(charttime)]
setkeyv(labevents, c("hadm_id", "charttime"))

chart_sofa_items <- c(CHART_ITEMIDS$map_invasive, CHART_ITEMIDS$map_noninvasive,
                      CHART_ITEMIDS$gcs_eye, CHART_ITEMIDS$gcs_verbal,
                      CHART_ITEMIDS$gcs_motor)

v2_log("Loading chartevents for SOFA components (MAP, GCS) via DuckDB...")
chartevents_sofa <- read_large_csv_filtered(
  path     = icu_file("chartevents.csv"),
  cols     = c("stay_id", "hadm_id", "itemid", "charttime", "valuenum"),
  itemids  = chart_sofa_items,
  hadm_ids = cohort_hadm_ids
)
chartevents_sofa[, charttime := parse_dt(charttime)]
setkeyv(chartevents_sofa, c("hadm_id", "charttime"))

v2_log("Loading inputevents for vasopressors...")
inputevents <- read_mimic(icu_file("inputevents.csv"),
                          cols = c("stay_id", "hadm_id", "itemid",
                                   "starttime", "endtime", "amount",
                                   "amountuom", "rate", "rateuom"))
inputevents <- inputevents[itemid %in% unlist(VASOPRESSOR_ITEMIDS) &
                             hadm_id %in% cohort_hadm_ids]
inputevents[, starttime := parse_dt(starttime)]
inputevents[, endtime   := parse_dt(endtime)]
setkeyv(inputevents, c("hadm_id", "starttime"))
```

## Vectorised SOFA Delta

Sepsis-3 requires a SOFA increase $\geq$ 2 points relative to baseline. `data.table::foverlaps` performs interval joins to aggregate measurements within evaluation and baseline windows without per-row loops.

```{r sofa-delta, code_folding="hide"}
source("clinical_scores.R")

lab_pt <- labevents[, .(hadm_id, t_start = charttime, t_end = charttime,
                         itemid, valuenum)]
setkeyv(lab_pt, c("hadm_id", "t_start", "t_end"))

chart_pt <- chartevents_sofa[, .(hadm_id, t_start = charttime, t_end = charttime,
                                  itemid, valuenum)]
setkeyv(chart_pt, c("hadm_id", "t_start", "t_end"))

vaso_iv <- inputevents[, .(hadm_id, t_start = starttime,
                             t_end = fifelse(is.na(endtime), starttime + 3600, endtime))]
setkeyv(vaso_iv, c("hadm_id", "t_start", "t_end"))

sofa_from_agg <- function(agg_dt) {
  w   <- dcast(agg_dt, stay_id ~ itemid, value.var = "valuenum",
               fun.aggregate = mean, na.rm = TRUE)
  ids <- names(w)[names(w) != "stay_id"]

  get_v <- function(iids) {
    cs <- intersect(as.character(as.integer(iids)), ids)
    if (length(cs) == 0) return(rep(NA_real_, nrow(w)))
    if (length(cs) == 1) return(as.numeric(w[[cs]]))
    rowMeans(w[, ..cs], na.rm = TRUE)
  }

  pao2_v <- get_v(LAB_ITEMIDS$pao2)
  fio2_v <- get_v(LAB_ITEMIDS$fio2_lab)
  fio2_f <- ifelse(!is.na(fio2_v) & fio2_v > 1, fio2_v / 100, fio2_v)
  pf     <- ifelse(!is.na(pao2_v) & !is.na(fio2_f) & fio2_f > 0,
                   pao2_v / fio2_f, NA_real_)

  panel <- data.frame(
    pf_ratio   = pf,
    platelets  = get_v(LAB_ITEMIDS$platelets),
    bilirubin  = get_v(LAB_ITEMIDS$bilirubin_total),
    creatinine = get_v(LAB_ITEMIDS$creatinine),
    map = {
      mi <- get_v(CHART_ITEMIDS$map_invasive)
      mn <- get_v(CHART_ITEMIDS$map_noninvasive)
      pmin(mi, mn, na.rm = TRUE)
    },
    gcs = {
      e  <- get_v(CHART_ITEMIDS$gcs_eye)
      v2 <- get_v(CHART_ITEMIDS$gcs_verbal)
      m  <- get_v(CHART_ITEMIDS$gcs_motor)
      rowSums(cbind(e, v2, m), na.rm = TRUE)
    },
    vaso_any       = FALSE,
    norepi_epi_any = FALSE
  )
  s <- score_sofa(panel)
  data.table(stay_id = w$stay_id, sofa_total = s$sofa_total)
}

compute_sofa_delta_vectorized <- function(rows, v) {
  if (nrow(rows) == 0) return(list(sofa_window = numeric(0), sofa_base = numeric(0)))

  eval_wins <- rows[, .(hadm_id, stay_id,
                         t_start = sofa_window_start, t_end = sofa_window_end,
                         wtype = "eval")]
  if (v$sofa_baseline == "admission") {
    base_wins <- rows[, .(hadm_id, stay_id,
                           t_start = intime, t_end = intime + 3600 * 4,
                           wtype = "base")]
  } else {
    base_wins <- rows[, .(hadm_id, stay_id,
                           t_start = sofa_window_start - 3600 * 24,
                           t_end   = sofa_window_start,
                           wtype = "base")]
  }
  all_wins <- rbindlist(list(eval_wins, base_wins))
  all_wins <- all_wins[!is.na(t_start) & !is.na(t_end) & t_end >= t_start]
  setkeyv(all_wins, c("hadm_id", "t_start", "t_end"))

  lab_m   <- foverlaps(lab_pt,   all_wins, type = "within", nomatch = 0L)
  chart_m <- foverlaps(chart_pt, all_wins, type = "within", nomatch = 0L)
  vaso_m  <- foverlaps(vaso_iv,  all_wins, type = "any",    nomatch = 0L)

  all_m <- rbindlist(list(
    lab_m[,   .(stay_id, wtype, itemid, valuenum)],
    chart_m[, .(stay_id, wtype, itemid, valuenum)]
  ))
  if (nrow(all_m) == 0) {
    zeros <- data.table(stay_id = rows$stay_id, sofa_total = 0)
    return(list(sofa_window = zeros$sofa_total, sofa_base = zeros$sofa_total))
  }

  agg <- all_m[, .(valuenum = mean(valuenum, na.rm = TRUE)),
               by = .(stay_id, wtype, itemid)]
  agg[, itemid := as.character(itemid)]

  if (nrow(vaso_m) > 0)
    vaso_flag <- unique(vaso_m[, .(stay_id, wtype)])[, vaso_any := TRUE]

  compute_for_type <- function(wt) {
    sub <- agg[wtype == wt]
    if (nrow(sub) == 0) return(data.table(stay_id = rows$stay_id, sofa_total = 0))
    res   <- sofa_from_agg(sub)
    all_s <- data.table(stay_id = rows$stay_id)
    res   <- merge(all_s, res, by = "stay_id", all.x = TRUE)
    res[is.na(sofa_total), sofa_total := 0]
    res
  }

  eval_sofa <- compute_for_type("eval")
  base_sofa <- compute_for_type("base")
  list(sofa_window = eval_sofa$sofa_total, sofa_base = base_sofa$sofa_total)
}
```

## Build Cohort Per Variant

Each variant applies its own antibiotic-culture time window and SOFA evaluation period, producing a distinct set of labelled onsets. Onsets outside the ICU stay or within the first hour are excluded.

```{r build-cohorts}
build_cohort_for_variant <- function(variant_id) {
  v    <- variant_def(variant_id)
  ty   <- variant_type(variant_id)
  step(sprintf("Building cohort for Variant %s (%s)...", variant_id, v$name))

  # Variant D carries no suspicion anchor at all: onset is derived per hour from
  # physiology in 03_person_hours.Rmd, where the hourly panel exists. The cohort
  # is the same base cohort with an empty onset column.
  if (ty == "deterioration") {
    cohort <- copy(stays)
    cohort[, `:=`(t_susp = as.POSIXct(NA), onset_time = as.POSIXct(NA),
                  onset_hour = NA_integer_,
                  died_in_hosp = hospital_expire_flag == 1L,
                  label_variant = variant_id, label_name = v$name)]
    cat(sprintf("Variant %s: %d total stays, onsets derived in stage 03\n",
                variant_id, nrow(cohort)))
    return(cohort)
  }

  susp <- suspicion_raw[
    (dt_abx_before_cul >= 0 & dt_abx_before_cul <= v$abx_before_culture_h) |
    (dt_cul_before_abx >= 0 & dt_cul_before_abx <= v$culture_before_abx_h)
  ]
  susp[dt_abx_before_cul >= 0,                         t_susp := pmax(abx_time, culture_time)]
  susp[dt_cul_before_abx >= 0 & is.na(t_susp),         t_susp := pmax(abx_time, culture_time)]
  susp <- susp[!is.na(t_susp)]
  setorderv(susp, c("hadm_id", "t_susp"))
  susp_first <- susp[, .SD[1], by = hadm_id]

  cohort <- merge(stays, susp_first[, .(hadm_id, t_susp)],
                  by = "hadm_id", all.x = TRUE)

  v2_log(sprintf("  Stays with suspected infection: %s / %s",
                 format(cohort[!is.na(t_susp), .N], big.mark = ","),
                 format(nrow(cohort), big.mark = ",")))

  # Variant E is the anchor limb on its own: the suspicion time *is* the onset,
  # with no organ-dysfunction requirement.
  if (ty == "anchor_only") {
    cohort[, onset_time := t_susp]
  } else {

  cohort[, sofa_window_start := t_susp - lubridate::hours(v$sofa_window_before_h)]
  cohort[, sofa_window_end   := t_susp + lubridate::hours(v$sofa_window_after_h)]

  susp_idx <- which(!is.na(cohort$t_susp))
  if (length(susp_idx) > 0) {
    v2_log(sprintf("  Computing SOFA delta (vectorized) for %d stays...",
                   length(susp_idx)))
    sofa_res <- compute_sofa_delta_vectorized(cohort[susp_idx], v)
    cohort[susp_idx, sofa_in_window := sofa_res$sofa_window]
    cohort[susp_idx, sofa_baseline  := sofa_res$sofa_base]
    cohort[susp_idx, sofa_delta     := sofa_in_window - sofa_baseline]
    cohort[susp_idx,
           onset_time := fifelse(sofa_delta >= SOFA_INCREASE_THRESHOLD,
                                 t_susp, as.POSIXct(NA))]
  }

  }

  cohort[!is.na(onset_time) & (onset_time < intime | onset_time > outtime),
         onset_time := NA_real_]
  cohort[!is.na(onset_time),
         onset_hour := as.integer(hours_diff(intime, onset_time))]
  cohort[!is.na(onset_hour) & onset_hour <= 1, onset_time := NA_real_]
  cohort[!is.na(onset_hour) & onset_hour <= 1, onset_hour := NA_integer_]

  cohort[, died_in_hosp   := hospital_expire_flag == 1L]
  cohort[, label_variant  := variant_id]
  cohort[, label_name     := v$name]

  n_events <- cohort[!is.na(onset_time), .N]
  cat(sprintf("Variant %s: %d total stays, %d sepsis onsets (%.1f%%)\n",
              variant_id, nrow(cohort), n_events,
              100 * n_events / nrow(cohort)))
  cohort
}

run_variants   <- active_variants()
cohorts        <- lapply(run_variants, build_cohort_for_variant)
names(cohorts) <- run_variants

for (vid in names(cohorts))
  save_parquet(cohorts[[vid]], paste0("02_cohort_", vid), DATA_DIR)
```

## Label-Variance Table (M1 Deliverable)

The label-variance table summarises cohort size, event count, event rate, and onset-time distribution for each variant. This is the empirical foundation for the label-dominance hypothesis (H1).

```{r label-variance-table}
step("Building M1 label-variance table...")

summarise_variant <- function(vid) {
  c_dt <- cohorts[[vid]]
  data.frame(
    variant         = vid,
    name            = variant_name(vid),
    n_stays         = nrow(c_dt),
    n_sepsis_onsets = c_dt[!is.na(onset_time), .N],
    pct_sepsis      = round(100 * c_dt[!is.na(onset_time), .N] / nrow(c_dt), 1),
    n_non_sepsis    = c_dt[is.na(onset_time), .N],
    n_deaths_no_sep = c_dt[is.na(onset_time) & died_in_hosp == TRUE, .N],
    median_onset_h  = round(median(c_dt[!is.na(onset_hour), onset_hour], na.rm = TRUE), 1),
    iqr_onset_h_lo  = round(quantile(c_dt[!is.na(onset_hour), onset_hour], 0.25, na.rm = TRUE), 1),
    iqr_onset_h_hi  = round(quantile(c_dt[!is.na(onset_hour), onset_hour], 0.75, na.rm = TRUE), 1)
  )
}

prereg_run <- intersect(names(cohorts), names(LABEL_VARIANTS))
if (length(prereg_run) > 0) {
  label_variance <- do.call(rbind, lapply(prereg_run, summarise_variant))
  knitr::kable(label_variance, caption = "M1 Label-Variance Table (cohort size × event count × onset timing)")
  save_parquet(label_variance, "02_label_variance_table", DATA_DIR)
  cat("\nSaved: 02_label_variance_table.parquet\n")
}
```

## Alternative-Label Variance Table (Protocol Amendment 2)

Variants A--C are three readings of one definition and share a single
culture-plus-antibiotic anchor, so part of their disagreement is definitional
rather than empirical. Variant E keeps that anchor and drops the
organ-dysfunction requirement; Variant D keeps organ dysfunction and drops the
anchor entirely. Variant D's onsets are assigned per hour in
`03_person_hours.Rmd` and so are not counted here.

```{r alt-label-variance-table}
alt_run <- intersect(names(cohorts), names(ALT_LABEL_VARIANTS))
if (length(alt_run) > 0) {
  alt_variance <- do.call(rbind, lapply(alt_run, summarise_variant))
  alt_variance$label_family <- vapply(alt_run, variant_type, character(1))
  knitr::kable(alt_variance,
               caption = "Alternative (non-Sepsis-3) label variants; Variant D onsets are assigned in stage 03")
  save_parquet(alt_variance, "02_alt_label_variance_table", DATA_DIR)
  cat("\nSaved: 02_alt_label_variance_table.parquet\n")
} else {
  cat("No alternative-label variants selected in this run.\n")
}
```
