---
title: "03 Person-Hours - Person-Hour Table Construction"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)   # student ID for reproducibility

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook expands each ICU stay into an hourly grid, attaches time-varying vitals and labs via LOCF (last-observation-carried-forward), and codes each row with its competing-event status (D'Agostino et al., 1990; Heyard et al., 2019).

Four mutually exclusive states at each hour:

| Code | Event | Interpretation |
|------|-------|----------------|
| 0 | In-progress | Patient remains at risk (censored at this hour) |
| 1 | Sepsis onset | Sepsis-3 criteria met in this hour |
| 2 | Discharge alive | ICU discharge without developing sepsis |
| 3 | Death without sepsis | In-ICU mortality without prior sepsis onset |

Treatment anchor timestamps (first antibiotic, first culture) are recorded for the treatment-anchored evaluation window (Kamran et al., 2024).

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(dplyr)
})

source("config.R")
source("utils.R")
source("clinical_scores.R")

step("=== Phase 2b: Person-hour table construction ===")
```

## Load Cohort IDs

Cohort stay and admission IDs are collected so the large source tables can be filtered in a single DuckDB pass.

```{r load-cohort-ids}
step("Loading cohort IDs for pre-filtering...")

all_stay_ids <- unique(unlist(lapply(names(LABEL_VARIANTS), function(vid) {
  p <- file.path(OUTPUT_DIR, paste0("02_cohort_", vid, ".parquet"))
  if (!file.exists(p)) return(integer(0))
  as.integer(arrow::read_parquet(p, col_select = "stay_id")$stay_id)
})))

all_hadm_ids <- unique(unlist(lapply(names(LABEL_VARIANTS), function(vid) {
  p <- file.path(OUTPUT_DIR, paste0("02_cohort_", vid, ".parquet"))
  if (!file.exists(p)) return(integer(0))
  as.integer(arrow::read_parquet(p, col_select = "hadm_id")$hadm_id)
})))

cat(sprintf("Cohort: %d stays, %d hadm_ids\n",
            length(all_stay_ids), length(all_hadm_ids)))
```

## Load and Pre-Filter Large Tables

Chartevents and labevents are loaded via DuckDB with pushdown filters on item IDs and cohort membership. Temperature F is converted to C, and FiO$_2$ percentages are rescaled to fractions.

```{r load-chartevents}
step("Loading chartevents (vitals) via DuckDB...")

all_chart_ids <- unlist(CHART_ITEMIDS)
chartevents   <- read_large_csv_filtered(
  path     = icu_file("chartevents.csv"),
  cols     = c("stay_id", "hadm_id", "itemid", "charttime", "valuenum"),
  itemids  = all_chart_ids,
  stay_ids = all_stay_ids
)
chartevents[, charttime := parse_dt(charttime)]
chartevents[, valuenum  := as.numeric(valuenum)]

# Unit conversions
temp_f_ids <- CHART_ITEMIDS$temp_f
chartevents[itemid %in% temp_f_ids, valuenum := (valuenum - 32) * 5 / 9]
chartevents[itemid %in% temp_f_ids, itemid   := CHART_ITEMIDS$temp_c]

fio2_ids <- CHART_ITEMIDS$fio2
chartevents[itemid %in% fio2_ids & valuenum > 1, valuenum := valuenum / 100]

cat(sprintf("chartevents filtered: %s rows\n", format(nrow(chartevents), big.mark = ",")))
```

```{r load-labevents}
step("Loading labevents via DuckDB...")

all_lab_ids <- unlist(LAB_ITEMIDS)
labevents   <- read_large_csv_filtered(
  path     = hosp_file("labevents.csv"),
  cols     = c("hadm_id", "itemid", "charttime", "valuenum"),
  itemids  = all_lab_ids,
  hadm_ids = all_hadm_ids
)
labevents[, charttime := parse_dt(charttime)]
labevents[, valuenum  := as.numeric(valuenum)]
cat(sprintf("labevents filtered: %s rows\n", format(nrow(labevents), big.mark = ",")))
```

```{r load-vasopressors}
step("Loading inputevents (vasopressors)...")

all_vaso_ids <- unlist(VASOPRESSOR_ITEMIDS)
inputevents  <- read_mimic(icu_file("inputevents.csv"),
                           cols = c("stay_id", "hadm_id", "itemid",
                                    "starttime", "endtime", "rate"))
inputevents  <- inputevents[itemid %in% all_vaso_ids]
inputevents[, starttime := parse_dt(starttime)]
inputevents[, endtime   := parse_dt(endtime)]
```

```{r load-treatment-anchors}
step("Loading treatment anchor timestamps...")

prescriptions <- read_mimic(hosp_file("prescriptions.csv"),
                             cols = c("hadm_id", "starttime", "drug"))
prescriptions[, starttime := parse_dt(starttime)]
prescriptions[, is_abx := match_antibiotics(drug, ANTIBIOTICS)]
first_abx <- prescriptions[is_abx == TRUE & !is.na(starttime),
                            .(first_abx_time = min(starttime, na.rm = TRUE)),
                            by = hadm_id]

microbio <- read_mimic(hosp_file("microbiologyevents.csv"),
                       cols = c("hadm_id", "charttime", "spec_type_desc"))
microbio[, charttime := parse_dt(charttime)]
microbio <- microbio[!(spec_type_desc %in% SURVEILLANCE_SPEC_TYPES) & !is.na(charttime)]
first_culture <- microbio[, .(first_culture_time = min(charttime, na.rm = TRUE)), by = hadm_id]
```

## Per-Hour Covariate Extraction

A dense hourly grid is built for each stay. Vitals and labs are aggregated to hourly means, then forward-filled (LOCF). Missingness indicators flag unmeasured hours. Six-hour rolling slopes capture vital-sign trends.

```{r hourly-covariates-fn, code_folding="hide"}
build_hourly_covariates <- function(stays_dt) {
  step(sprintf("  Building hourly grid for %d stays...", nrow(stays_dt)))
  grid <- build_person_hour_grid(stays_dt, max_hours = MAX_PANEL_HOURS)
  grid <- merge(grid, stays_dt[, .(stay_id, hadm_id, intime)],
                by = "stay_id", all.x = TRUE)
  grid[, hour_time := intime + lubridate::hours(hour)]

  # Chart vitals → hourly mean
  chart_var_map <- data.table(
    itemid   = as.integer(unlist(CHART_ITEMIDS)),
    var_name = rep(names(CHART_ITEMIDS), lengths(CHART_ITEMIDS))
  )
  ce_mapped <- merge(chartevents, chart_var_map, by = "itemid")
  ce_merged <- merge(ce_mapped, stays_dt[, .(stay_id, intime)], by = "stay_id")
  ce_merged[, obs_hour := floor(hours_diff(intime, charttime))]
  ce_merged <- ce_merged[obs_hour >= 0 & obs_hour <= MAX_PANEL_HOURS]

  gcs_events <- ce_merged[var_name %in% c("gcs_eye", "gcs_verbal", "gcs_motor")]
  gcs_hr     <- gcs_events[, .(gcs = sum(valuenum, na.rm = TRUE)), by = .(stay_id, obs_hour)]
  gcs_hr[gcs < 3 | gcs > 15, gcs := NA_real_]

  other_chart <- ce_merged[!var_name %in% c("gcs_eye", "gcs_verbal", "gcs_motor", "temp_f")]
  chart_hr <- dcast(
    other_chart[, .(value = mean(valuenum, na.rm = TRUE)),
                by = .(stay_id, obs_hour, var_name)],
    stay_id + obs_hour ~ var_name, value.var = "value", fun.aggregate = mean
  )

  # Consolidate dual MAP / SBP itemids
  if ("map_invasive" %in% names(chart_hr) && "map_noninvasive" %in% names(chart_hr)) {
    chart_hr[, map := pmin(map_invasive, map_noninvasive, na.rm = TRUE)]
    chart_hr[, c("map_invasive", "map_noninvasive") := NULL]
  } else if ("map_invasive"    %in% names(chart_hr)) { setnames(chart_hr, "map_invasive",    "map") }
  else if   ("map_noninvasive" %in% names(chart_hr)) { setnames(chart_hr, "map_noninvasive", "map") }

  if ("sbp_invasive" %in% names(chart_hr) && "sbp_noninvasive" %in% names(chart_hr)) {
    chart_hr[, sbp := pmin(sbp_invasive, sbp_noninvasive, na.rm = TRUE)]
    chart_hr[, c("sbp_invasive", "sbp_noninvasive") := NULL]
  } else if ("sbp_invasive"    %in% names(chart_hr)) { setnames(chart_hr, "sbp_invasive",    "sbp") }
  else if   ("sbp_noninvasive" %in% names(chart_hr)) { setnames(chart_hr, "sbp_noninvasive", "sbp") }

  chart_hr <- merge(chart_hr, gcs_hr, by = c("stay_id", "obs_hour"), all = TRUE)

  # Lab values → hourly mean
  le_merged <- merge(labevents, stays_dt[, .(stay_id, hadm_id, intime)], by = "hadm_id")
  le_merged[, obs_hour := floor(hours_diff(intime, charttime))]
  le_merged <- le_merged[obs_hour >= 0 & obs_hour <= MAX_PANEL_HOURS]
  lab_var_map <- data.table(
    itemid   = as.integer(unlist(LAB_ITEMIDS)),
    var_name = rep(names(LAB_ITEMIDS), lengths(LAB_ITEMIDS))
  )
  le_mapped <- merge(le_merged, lab_var_map, by = "itemid")
  lab_hr <- dcast(
    le_mapped[, .(value = mean(valuenum, na.rm = TRUE)),
              by = .(stay_id, obs_hour, var_name)],
    stay_id + obs_hour ~ var_name, value.var = "value", fun.aggregate = mean
  )
  if ("pao2" %in% names(lab_hr) && "fio2_lab" %in% names(lab_hr)) {
    lab_hr[, fio2_frac := ifelse(fio2_lab > 1, fio2_lab / 100, fio2_lab)]
    lab_hr[, pf_ratio  := ifelse(!is.na(pao2) & !is.na(fio2_frac) & fio2_frac > 0,
                                  pao2 / fio2_frac, NA_real_)]
  } else { lab_hr[, pf_ratio := NA_real_] }

  # Vasopressors → any active in this hour
  inp_merged <- merge(inputevents, stays_dt[, .(stay_id, intime)], by = "stay_id")
  inp_merged[, start_h := floor(hours_diff(intime, starttime))]
  inp_merged[, end_h   := floor(hours_diff(intime, endtime))]
  inp_merged[is.na(end_h), end_h := start_h]
  vaso_hrs <- inp_merged[, .(hour = seq(max(0, start_h), min(MAX_PANEL_HOURS, end_h))),
                          by = stay_id]
  vaso_hrs[, vaso_any := TRUE]
  vaso_hrs <- vaso_hrs[hour >= 0]
  setnames(vaso_hrs, "hour", "obs_hour")

  norepi_ids <- unlist(list(norepinephrine = VASOPRESSOR_ITEMIDS$norepinephrine,
                             epinephrine    = VASOPRESSOR_ITEMIDS$epinephrine))
  vaso_norepi <- inp_merged[itemid %in% norepi_ids,
                              .(hour = seq(max(0, start_h), min(MAX_PANEL_HOURS, end_h))),
                              by = stay_id]
  vaso_norepi[, norepi_epi_any := TRUE]
  if (nrow(vaso_norepi) > 0) setnames(vaso_norepi, "hour", "obs_hour")

  # Merge onto grid
  setnames(grid, "hour", "obs_hour")
  panel <- merge(grid, chart_hr, by = c("stay_id", "obs_hour"), all.x = TRUE)
  lab_keep <- setdiff(names(lab_hr), c("fio2_lab", "fio2_frac", "pao2"))
  panel <- merge(panel, lab_hr[, ..lab_keep], by = c("stay_id", "obs_hour"), all.x = TRUE)

  if (nrow(vaso_hrs) > 0) {
    panel <- merge(panel, unique(vaso_hrs), by = c("stay_id", "obs_hour"), all.x = TRUE)
    panel[is.na(vaso_any), vaso_any := FALSE]
  } else { panel[, vaso_any := FALSE] }

  if (nrow(vaso_norepi) > 0) {
    panel <- merge(panel, unique(vaso_norepi), by = c("stay_id", "obs_hour"), all.x = TRUE)
    panel[is.na(norepi_epi_any), norepi_epi_any := FALSE]
  } else { panel[, norepi_epi_any := FALSE] }

  # Forward-fill within stay
  fill_cols <- intersect(c("hr", "resp_rate", "spo2", "map", "sbp", "temp_c", "fio2",
                            "gcs", "pf_ratio", "lactate", "creatinine", "bilirubin_total",
                            "platelets", "wbc", "pao2"), names(panel))
  if (length(fill_cols) > 0) panel <- ffill_by_stay(panel, fill_cols)

  # Missingness indicators
  miss_labs <- c("lactate", "creatinine", "bilirubin_total", "platelets", "wbc", "pf_ratio")
  for (ml in intersect(miss_labs, names(panel)))
    panel[[paste0(ml, "_measured")]] <- as.integer(!is.na(panel[[ml]]))

  # 6-hour rolling slopes
  slope_cols <- intersect(c("hr", "resp_rate", "map", "temp_c", "gcs"), names(panel))
  for (col in slope_cols)
    panel <- rolling_mean_by_stay(panel, col, n_h = SLOPE_WINDOW_H,
                                  col_out = paste0(col, "_slope6h"))

  setnames(panel, "obs_hour", "hour")
  panel
}
```

## Build Person-Hour Tables Per Variant

For each label variant, the hourly panel is assembled, treatment anchors merged, competing events coded, and clinical scores (SOFA, NEWS2, qSOFA, SIRS) computed. Post-outcome rows are dropped. A temporal train/test split based on `anchor_year_group` follows Guo et al. (2022).

```{r build-person-hours}
for (vid in names(LABEL_VARIANTS)) {
  step(sprintf("=== Person-hours for Variant %s (%s) ===",
               vid, LABEL_VARIANTS[[vid]]$name))

  cohort <- load_parquet(paste0("02_cohort_", vid), OUTPUT_DIR)
  panel  <- build_hourly_covariates(cohort)

  panel <- merge(panel,
                 cohort[, .(stay_id, anchor_year_group,
                             onset_hour, died_in_hosp, los_hours,
                             gender, anchor_age, race, language, insurance,
                             label_variant, label_name)],
                 by = "stay_id", all.x = TRUE)

  # Treatment anchor timestamps
  cohort_hadm_only <- unique(cohort[, .(hadm_id)])
  abx_anchor  <- merge(cohort_hadm_only, first_abx,     by = "hadm_id", all.x = TRUE)
  cult_anchor <- merge(cohort_hadm_only, first_culture,  by = "hadm_id", all.x = TRUE)
  panel <- merge(panel, abx_anchor,  by = "hadm_id", all.x = TRUE)
  panel <- merge(panel, cult_anchor, by = "hadm_id", all.x = TRUE)
  panel[, first_abx_hour     := floor(hours_diff(intime, first_abx_time))]
  panel[, first_culture_hour := floor(hours_diff(intime, first_culture_time))]

  # Competing event coding
  panel[, event := 0L]
  panel[!is.na(onset_hour) & hour == onset_hour, event := 1L]
  panel_last_hour <- panel[, .(max_hour = max(hour)), by = stay_id]
  panel <- merge(panel, panel_last_hour, by = "stay_id", all.x = TRUE)
  panel[hour == max_hour & is.na(onset_hour) & died_in_hosp == TRUE,  event := 3L]
  panel[hour == max_hour & is.na(onset_hour) & died_in_hosp == FALSE, event := 2L]

  # Drop post-outcome rows
  panel[!is.na(onset_hour), at_risk := hour <= onset_hour]
  panel[is.na(onset_hour),  at_risk := TRUE]
  panel <- panel[at_risk == TRUE]
  panel[, at_risk := NULL]

  # SOFA
  if ("bilirubin_total" %in% names(panel) && !"bilirubin" %in% names(panel))
    panel[, bilirubin := bilirubin_total]
  sofa_scores <- score_sofa(as.data.frame(panel))
  panel[, names(sofa_scores) := sofa_scores]

  # NEWS2
  nr <- nrow(panel)
  get_panel_col <- function(col) { v <- panel[[col]]; if (is.null(v)) rep(NA_real_, nr) else v }
  suppl_o2   <- { fio2_v <- get_panel_col("fio2"); !is.na(fio2_v) & fio2_v > 0.21 }
  avpu_alert <- { gcs_v  <- get_panel_col("gcs");  !is.na(gcs_v)  & gcs_v >= 14 }
  news2_scores <- news2(get_panel_col("resp_rate"), get_panel_col("spo2"),
                        suppl_o2, get_panel_col("sbp"), get_panel_col("hr"),
                        get_panel_col("temp_c"), avpu_alert)
  panel[, names(news2_scores) := news2_scores]

  panel[, qsofa_score := qsofa(get_panel_col("resp_rate"),
                                get_panel_col("gcs"), get_panel_col("map"))]
  panel[, sirs_score  := sirs(get_panel_col("hr"), get_panel_col("resp_rate"),
                               get_panel_col("temp_c"), get_panel_col("wbc"))]

  # Temporal split
  panel[, temporal_split := ifelse(anchor_year_group %in% TEMPORAL_TEST_GROUPS,
                                    "test", "train")]

  # Random split (patient-level, for H4 split-optimism quantification; Guo et al., 2022)
  unique_stays <- unique(panel$stay_id)
  set.seed(RANDOM_SPLIT_SEED)
  test_stays <- sample(unique_stays, size = round(length(unique_stays) * RANDOM_SPLIT_TEST_FRAC))
  panel[, random_split := ifelse(stay_id %in% test_stays, "test", "train")]
  cat(sprintf("Random split: %d train stays, %d test stays (seed=%d, frac=%.2f)\n",
              length(unique_stays) - length(test_stays), length(test_stays),
              RANDOM_SPLIT_SEED, RANDOM_SPLIT_TEST_FRAC))

  drop_cols <- intersect(c("max_hour", "intime.x", "intime.y"), names(panel))
  if (length(drop_cols) > 0) panel[, (drop_cols) := NULL]

  cat(sprintf("Person-hour table V%s: %s rows, %d cols\n",
              vid, format(nrow(panel), big.mark = ","), ncol(panel)))
  cat(sprintf("Events: sepsis=%d, discharge=%d, death=%d, censored=%d\n",
              panel[event == 1, .N], panel[event == 2, .N],
              panel[event == 3, .N], panel[event == 0, .N]))

  save_parquet(panel, paste0("03_person_hours_", vid), OUTPUT_DIR)
  rm(panel, cohort); gc()
}
```
