---
title: "05 Primary Model - Dynamic Discrete-Time Competing-Risks Hazard Model"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

The primary model is a **multinomial pooled logistic regression** (D'Agostino et al., 1990; Heyard et al., 2019). If this interpretable regression matches gradient-boosted trees (H6), black-box complexity is unnecessary.

For patient $i$ at hour $t$, with competing events $r \in \{1\text{=sepsis}, 2\text{=discharge}, 3\text{=death}\}$:

$$\lambda_r(t \mid X_{it}) = \frac{\exp\bigl(\alpha_r(t) + \boldsymbol{\beta}_r^\top \mathbf{X}_{it}\bigr)}{1 + \sum_{s} \exp\bigl(\alpha_s(t) + \boldsymbol{\beta}_s^\top \mathbf{X}_{it}\bigr)}$$

$\alpha_r(t)$ is the baseline log-hazard (natural cubic spline in hours since admission) and $\boldsymbol{\beta}_r$ are cause-specific covariate coefficients. Standard errors use a cluster-robust sandwich estimator (clustered on `stay_id`).

Models are fitted per label variant: Variant B is primary; A and C support the cross-label coefficient comparison (H2). The alternative-label variants of Protocol Amendment 2 (D, E) are fitted with exactly the same specification so that the label is the only thing that differs; their coefficients are kept in a separate file and take no part in H2.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(nnet)
  library(rms)
  library(survival)
  library(sandwich)
  library(lmtest)
  library(dplyr)
  library(splines)
})

source("config.R")
source("utils.R")

step("=== Phase 3a: Primary discrete-time competing-risks model ===")
```

## Feature Set

Predictors include vitals, lab values, vasopressor exposure, and 6-hour rolling slopes. Missingness indicators let the model distinguish normal values from unmeasured ones.

```{r feature-set}
BASE_FEATURES <- c(
  "hr", "resp_rate", "spo2", "map", "temp_c", "gcs",
  "lactate", "creatinine", "bilirubin_total", "platelets", "wbc",
  "vaso_any", "norepi_epi_any",
  "hr_slope6h", "resp_rate_slope6h", "map_slope6h"
)
MISSINGNESS_FEATURES <- c(
  "lactate_measured", "creatinine_measured", "bilirubin_total_measured",
  "platelets_measured", "wbc_measured", "pf_ratio_measured"
)

cat("Base features:", paste(BASE_FEATURES, collapse = ", "), "\n")
cat("Missingness indicators:", paste(MISSINGNESS_FEATURES, collapse = ", "), "\n")
```

## Fit Primary Model Per Variant

Per variant: load training person-hours, impute remaining NAs to zero, fit the multinomial model via `nnet::multinom`, compute cluster-robust SEs, predict on the test set, and save outputs. A binary logistic fallback handles non-convergence.

```{r fit-primary-model, code_folding="hide"}
fit_primary_model <- function(vid, arm = "full") {
  step(sprintf("--- Fitting primary model: Variant %s (%s), arm = %s ---",
               vid, variant_name(vid), arm))

  ph <- as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
  ))

  train <- ph[temporal_split == "train"]
  cat(sprintf("  Training rows: %s, events (sepsis): %d\n",
              format(nrow(train), big.mark = ","), sum(train$event == 1)))

  train[, outcome := factor(event, levels = c(0, 1, 2, 3),
                             labels = c("inprogress", "sepsis", "discharge", "death"))]

  avail_base  <- intersect(BASE_FEATURES,        names(train))
  avail_miss  <- intersect(MISSINGNESS_FEATURES, names(train))
  require_features(BASE_FEATURES, MISSINGNESS_FEATURES, available = names(train))

  # Protocol Amendment 5: the ablated arm drops the action-derived features
  # (vasopressor exposure and the order indicators) and nothing else. The
  # required-feature check above still runs against the full declared set, so a
  # column vanishing from the panel is caught in both arms.
  all_features <- arm_features(c(avail_base, avail_miss), arm)
  if (identical(arm, "ablated"))
    cat(sprintf("  Ablated arm: %d action-derived features dropped, %d remain\n",
                length(intersect(ACTION_DERIVED_FEATURES, c(avail_base, avail_miss))),
                length(all_features)))

  for (f in all_features) {
    if (is.numeric(train[[f]])) train[is.na(get(f)), (f) := 0]
    if (is.logical(train[[f]])) train[is.na(get(f)), (f) := FALSE]
  }

  n_knots  <- 5
  hour_max <- max(train$hour, na.rm = TRUE)
  knots    <- quantile(train$hour, seq(0.1, 0.9, length.out = n_knots), na.rm = TRUE)
  cat(sprintf("  Spline knots at hours: %s\n", paste(round(knots, 1), collapse = ", ")))

  feat_formula <- paste(all_features, collapse = " + ")
  formula_str  <- sprintf(
    "outcome ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
    paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
  )
  mod_formula <- as.formula(formula_str)

  v2_log(sprintf("  Fitting multinomial logistic regression (ridge decay = %g)...",
                 PRIMARY_MODEL_DECAY))
  fit <- tryCatch(
    nnet::multinom(mod_formula, data = as.data.frame(train),
                   trace = FALSE, MaxNWts = 50000, maxit = PRIMARY_MODEL_MAXIT,
                   decay = PRIMARY_MODEL_DECAY),
    error = function(e) {
      v2_log("  nnet::multinom failed: ", e$message, level = "ERROR")
      NULL
    }
  )

  check_multinom_fit(fit, vid)

  if (is.null(fit)) {
    v2_log("  Fallback: binary logistic for sepsis onset only")
    train[, outcome_bin := as.integer(event == 1)]
    formula_bin <- as.formula(sprintf(
      "outcome_bin ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
      paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
    ))
    fit <- glm(formula_bin, data = as.data.frame(train),
               family = binomial(link = "logit"))
    fit$is_binary_fallback <- TRUE
  }

  # Cluster-robust SEs
  v2_log("  Computing cluster-robust SEs (clustered on stay_id)...")
  vcov_robust <- tryCatch({
    if (inherits(fit, "multinom"))
      sandwich::vcovHC(fit, type = "HC3")
    else
      sandwich::vcovCL(fit, cluster = ~stay_id, data = as.data.frame(train))
  }, error = function(e) {
    v2_log("  vcov_robust failed: ", e$message, level = "WARN"); NULL
  })

  # Coefficient table
  coefs_raw <- if (inherits(fit, "multinom")) coef(fit) else coef(summary(fit))
  if (is.matrix(coefs_raw)) {
    coef_df    <- as.data.table(as.data.frame(coefs_raw), keep.rownames = "outcome")
    coef_table <- melt(coef_df, id.vars = "outcome",
                       variable.name = "term", value.name = "coef")
    setDF(coef_table)
  } else {
    coef_table         <- as.data.frame(coefs_raw)
    coef_table$term    <- rownames(coef_table)
    coef_table$outcome <- NA_character_
  }
  coef_table$variant <- vid
  coef_table$model   <- "discrete_time_competing_risks"
  coef_table$arm     <- arm

  # Test-set predictions
  test <- ph[temporal_split == "test"]
  for (f in all_features) {
    if (is.numeric(test[[f]])) test[is.na(get(f)), (f) := 0]
    if (is.logical(test[[f]])) test[is.na(get(f)), (f) := FALSE]
  }
  v2_log(sprintf("  Predicting on test set (%s rows)...", format(nrow(test), big.mark = ",")))
  if (inherits(fit, "multinom")) {
    pred_probs <- predict(fit, newdata = as.data.frame(test), type = "probs")
    test[, pred_sepsis := if (is.matrix(pred_probs)) pred_probs[, "sepsis"]
                           else as.numeric(pred_probs)]
  } else {
    test[, pred_sepsis := predict(fit, newdata = as.data.frame(test), type = "response")]
  }

  save_cols <- c("stay_id", "hour", "event", "pred_sepsis", "temporal_split", "random_split",
                 "anchor_year_group", "onset_hour",
                 intersect(c("sofa_total", "news2_total", "qsofa_score"), names(test)))
  sfx <- arm_suffix(arm)
  save_parquet(test[, ..save_cols],
               paste0("05_predictions_primary", sfx, "_", vid), DATA_DIR)
  save_parquet(coef_table, paste0("05_coef_table_primary", sfx, "_", vid), DATA_DIR)
  saveRDS(fit, file.path(DATA_DIR, paste0("05_model_primary", sfx, "_", vid, ".rds")))

  cat(sprintf("  Variant %s model saved (arm = %s).\n", vid, arm))

  # Deliberately do NOT return the fit. A multinom object on 2.3 M person-hours
  # carries its fitted values and residuals (~1.5 GB live, 450 MB on disk), and
  # the caller keeps one per variant for the whole chunk. With five variants
  # that retained ~7 GB and made the random-split re-fit below hit R's 16 GB
  # vector limit. Only `coef_table` is used downstream (the H2 chunk); the fit
  # itself is already persisted to `05_model_primary_<vid>.rds`.
  out <- list(coef_table = coef_table, vcov_robust = vcov_robust)
  rm(fit, train, test, ph); invisible(gc(FALSE))
  out
}
```

```{r run-all-variants}
run_variants <- active_variants()
run_arms     <- active_arms()
v2_log(sprintf("Arms in this run: %s", paste(run_arms, collapse = ", ")))

models <- NULL
if ("full" %in% run_arms) {
  models        <- lapply(run_variants, fit_primary_model, arm = "full")
  names(models) <- run_variants
}

# The ablated arm is Protocol Amendment 5 and post-hoc. It writes to its own
# files and never overwrites the pre-registered fit, so `V2_ARMS=ablated`
# refreshes it while leaving every headline number on disk untouched.
models_ablated <- NULL
if ("ablated" %in% run_arms) {
  ablate_variants        <- intersect(run_variants, names(LABEL_VARIANTS))
  models_ablated         <- lapply(ablate_variants, fit_primary_model, arm = "ablated")
  names(models_ablated)  <- ablate_variants
}
```

## Random-Split Re-Fit (H4: Split Optimism)

The same primary model is re-fitted on a random patient-level split (Guo et al., 2022). Only Variant B is needed — H4 compares temporal vs. random AUROC under a fixed label and model class.

```{r fit-random-split}
step("--- Re-fitting primary model on random split (H4) ---")

fit_primary_random <- function(vid = "B") {
  ph <- as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
  ))

  if (!"random_split" %in% names(ph)) {
    v2_log("  random_split column not found — skipping H4 re-fit.", level = "WARN")
    return(NULL)
  }

  train <- ph[random_split == "train"]
  cat(sprintf("  Random-split training rows: %s, events (sepsis): %d\n",
              format(nrow(train), big.mark = ","), sum(train$event == 1)))

  train[, outcome := factor(event, levels = c(0, 1, 2, 3),
                             labels = c("inprogress", "sepsis", "discharge", "death"))]

  avail_base  <- intersect(BASE_FEATURES,        names(train))
  avail_miss  <- intersect(MISSINGNESS_FEATURES, names(train))
  all_features <- c(avail_base, avail_miss)
  require_features(BASE_FEATURES, MISSINGNESS_FEATURES, available = names(train))

  for (f in all_features) {
    if (is.numeric(train[[f]])) train[is.na(get(f)), (f) := 0]
    if (is.logical(train[[f]])) train[is.na(get(f)), (f) := FALSE]
  }

  n_knots  <- 5
  hour_max <- max(train$hour, na.rm = TRUE)
  knots    <- quantile(train$hour, seq(0.1, 0.9, length.out = n_knots), na.rm = TRUE)

  feat_formula <- paste(all_features, collapse = " + ")
  formula_str  <- sprintf(
    "outcome ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
    paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
  )
  mod_formula <- as.formula(formula_str)

  fit <- tryCatch(
    nnet::multinom(mod_formula, data = as.data.frame(train),
                   trace = FALSE, MaxNWts = 50000, maxit = PRIMARY_MODEL_MAXIT,
                   decay = PRIMARY_MODEL_DECAY),
    error = function(e) {
      v2_log("  Random-split multinom failed: ", e$message, level = "ERROR")
      NULL
    }
  )

  check_multinom_fit(fit, paste0(vid, " (random split)"))

  if (is.null(fit)) {
    v2_log("  Fallback: binary logistic for random split")
    train[, outcome_bin := as.integer(event == 1)]
    formula_bin <- as.formula(sprintf(
      "outcome_bin ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
      paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
    ))
    fit <- glm(formula_bin, data = as.data.frame(train),
               family = binomial(link = "logit"))
    fit$is_binary_fallback <- TRUE
  }

  test <- ph[random_split == "test"]
  for (f in all_features) {
    if (is.numeric(test[[f]])) test[is.na(get(f)), (f) := 0]
    if (is.logical(test[[f]])) test[is.na(get(f)), (f) := FALSE]
  }
  v2_log(sprintf("  Predicting on random-split test set (%s rows)...",
                 format(nrow(test), big.mark = ",")))
  if (inherits(fit, "multinom")) {
    pred_probs <- predict(fit, newdata = as.data.frame(test), type = "probs")
    test[, pred_sepsis := if (is.matrix(pred_probs)) pred_probs[, "sepsis"]
                           else as.numeric(pred_probs)]
  } else {
    test[, pred_sepsis := predict(fit, newdata = as.data.frame(test), type = "response")]
  }

  save_cols <- c("stay_id", "hour", "event", "pred_sepsis", "random_split",
                 "anchor_year_group", "onset_hour",
                 intersect(c("sofa_total", "news2_total", "qsofa_score"), names(test)))
  save_parquet(test[, ..save_cols],
               paste0("05_predictions_primary_random_", vid), DATA_DIR)
  saveRDS(fit, file.path(DATA_DIR, paste0("05_model_primary_random_", vid, ".rds")))
  cat(sprintf("  Random-split model for Variant %s saved.\n", vid))
  rm(fit, train, test, ph); invisible(gc(FALSE))
  invisible(TRUE)
}

# Release whatever the per-variant loop still holds before the random-split fit
# allocates its own design matrix.
invisible(gc(FALSE))

# The random-split arm exists only to quantify optimism under the pre-registered
# specification (H4), so it is tied to the full arm. Under V2_ARMS=ablated it is
# skipped rather than refitted, which is what keeps an ablation-only pass from
# touching any H4 input.
random_split_done <- if (!"full" %in% run_arms) {
  cat("Full arm not selected; keeping the existing random-split fit.\n")
  NULL
} else if ("B" %in% run_variants) fit_primary_random("B") else {
  cat("Variant B not selected in this run; keeping the existing random-split fit.\n")
  NULL
}
```

## Cross-Label Coefficient Stability

The combined coefficient table across variants feeds the H2 (coefficient instability) test: if a covariate changes sign across label variants, the label choice qualitatively alters the model's clinical logic.

```{r cross-label-coefs}
step("Computing cross-label coefficient stability (→ H2)...")

bind_coefs <- function(vids, src = models) data.table::rbindlist(
  lapply(vids, function(vid) as.data.table(src[[vid]]$coef_table)), fill = TRUE
)

# H2 is defined on the pre-registered family, so the alternative-label
# coefficients are written to their own file and never mixed in.
prereg_run <- if (is.null(models)) {
  character(0)
} else {
  intersect(run_variants, names(LABEL_VARIANTS))
}
if (length(prereg_run) > 0) {
  all_coefs <- bind_coefs(prereg_run)
  save_parquet(all_coefs, "05_coef_all_variants", DATA_DIR)
  cat("Cross-label coefficient tables saved for H2 testing.\n")

  # Show sepsis-outcome coefficients for Variant B as a preview
  coef_b_sepsis <- all_coefs[outcome == "sepsis" & variant == "B" &
                                !grepl("^ns\\(|^\\(Intercept\\)", as.character(term))]
  if (nrow(coef_b_sepsis) > 0) {
    knitr::kable(head(coef_b_sepsis[order(-abs(coef)), .(term, coef, variant)], 15),
                 caption = "Top 15 sepsis-outcome coefficients (Variant B, by |coef|)")
  }
} else {
  cat("No pre-registered variants in this run; 05_coef_all_variants left unchanged.\n")
}

alt_run <- if (is.null(models)) {
  character(0)
} else {
  intersect(run_variants, names(ALT_LABEL_VARIANTS))
}
if (length(alt_run) > 0) {
  save_parquet(bind_coefs(alt_run), "05_coef_alt_variants", DATA_DIR)
  cat("Alternative-label coefficient tables saved (Protocol Amendment 2).\n")
}
```

### Ablated-Arm Stability (Protocol Amendment 5)

The same instability rule applied to the ablated fit. The question is whether
the cross-label instability observed in the full specification is a property of
the model as a whole or specifically of the features that record clinician
behaviour: if the latter, removing those features should leave a markedly more
stable set of coefficients behind.

```{r ablated-coef-stability}
if (!is.null(models_ablated) && length(models_ablated) >= 2) {
  abl_coefs <- bind_coefs(names(models_ablated), src = models_ablated)
  save_parquet(abl_coefs, "05_coef_all_variants_ablated", DATA_DIR)

  # coef_stability_flags() is the same function 12_inference.Rmd applies to the
  # full arm, so the two counts differ only by the feature set.
  abl_flags <- coef_stability_flags(abl_coefs, variants = names(models_ablated))
  save_parquet(abl_flags, "05_h2_stability_ablated", DATA_DIR)
  save_csv(abl_flags,     "05_h2_stability_ablated", DATA_DIR)

  cat(sprintf("Ablated arm: of %d shared covariates, %d reverse sign, %d change magnitude >=2x, %d do either\n",
              nrow(abl_flags), sum(abl_flags$sign_flip), sum(abl_flags$mag_flip),
              sum(abl_flags$sign_flip | abl_flags$mag_flip)))
  knitr::kable(abl_flags[sign_flip | mag_flip],
               caption = "Unstable covariates remaining after ablation")
} else {
  cat("Ablated arm not run (or fewer than two variants); ablated stability unchanged.\n")
}
```
