---
title: "06 Comparators - Rule-Based, Cox, and GBT Comparator Models"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)   # student ID for reproducibility

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

All comparators share the same cohort, labels, features, temporal split, and metrics as the primary model (Reyna et al., 2020).

| Comparator | Rationale |
|---|---|
| **NEWS2** | Rule-based sepsis screening score (Evans et al., 2021). Pre-computed. |
| **qSOFA, SIRS** | Descriptive context only. |
| **Static Cox PH** | Weak baseline with time-fixed covariates. PH assumption expected to be violated. |
| **XGBoost** | ML comparator per PhysioNet conventions (Reyna et al., 2020). Tests H6. |

Variant B is the primary comparison; A and C feed the variance decomposition.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(survival)
  library(xgboost)
  library(rms)
  library(dplyr)
})

source("config.R")
source("utils.R")
source("clinical_scores.R")

step("=== Phase 3b: Comparator models ===")
```

## Feature Set

Same feature set as the primary model.

```{r feature-set}
BASE_FEATURES <- c(
  "hr", "resp_rate", "spo2", "map", "temp_c", "gcs",
  "lactate", "creatinine", "bilirubin_total", "platelets", "wbc",
  "vaso_any", "norepi_epi_any",
  "hr_slope6h", "resp_rate_slope6h", "map_slope6h",
  "lactate_measured", "creatinine_measured", "bilirubin_total_measured",
  "platelets_measured", "wbc_measured", "pf_ratio_measured"
)
```

## Fit Comparators Per Variant

Per variant: NEWS2/qSOFA/SIRS scores are read from the person-hour table (no fitting); Cox is fitted with Schoenfeld testing; XGBoost uses early stopping, class-imbalance weighting, and AUCPR.

```{r fit-comparators-fn, code_folding="hide"}
fit_comparators <- function(vid) {
  step(sprintf("--- Comparators for Variant %s ---", vid))

  ph <- as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
  ))

  train <- ph[temporal_split == "train"]
  test  <- ph[temporal_split == "test"]

  avail_feats <- intersect(BASE_FEATURES, names(ph))
  for (f in avail_feats) {
    if (is.numeric(train[[f]])) train[is.na(get(f)), (f) := 0]
    if (is.logical(train[[f]])) train[is.na(get(f)), (f) := FALSE]
    if (is.numeric(test[[f]]))  test[is.na(get(f)),  (f) := 0]
    if (is.logical(test[[f]]))  test[is.na(get(f)),  (f) := FALSE]
  }

  train[, event_bin := as.integer(event == 1)]
  test[,  event_bin := as.integer(event == 1)]

  results <- list()

  # --- NEWS2 (pre-computed in 03_person_hours.Rmd) ---
  if ("news2_total" %in% names(test)) {
    step("  Comparator: NEWS2")
    test[, pred_news2  := as.numeric(news2_total) / 20]
    test[, alert_news2 := as.integer(news2_total >= 5)]
    results[["news2"]] <- test[, .(stay_id, hour, event, event_bin,
                                    pred_news2, alert_news2)]
    cat(sprintf("    NEWS2 alert rate: %.1f%%\n",
                100 * mean(test$alert_news2, na.rm = TRUE)))
  }

  if ("qsofa_score" %in% names(test)) {
    test[, pred_qsofa := qsofa_score / 3]
    results[["qsofa"]] <- test[, .(stay_id, hour, event, event_bin, pred_qsofa)]
  }

  if ("sirs_score" %in% names(test)) {
    test[, pred_sirs := sirs_score / 4]
    results[["sirs"]] <- test[, .(stay_id, hour, event, event_bin, pred_sirs)]
  }

  # --- Static Cox PH ---
  step("  Comparator: Static Cox PH")
  cox_feats   <- intersect(c("hr", "resp_rate", "spo2", "map", "temp_c", "gcs",
                               "lactate", "creatinine", "bilirubin_total",
                               "platelets", "wbc", "vaso_any"), names(train))
  cox_fit <- tryCatch({
    train_cox <- train[, .SD, .SDcols = c("stay_id", "hour", "event_bin", cox_feats)]
    train_cox[, stop := hour + 1L]
    cox_formula_tv <- as.formula(paste(
      "Surv(hour, stop, event_bin) ~", paste(cox_feats, collapse = " + ")
    ))
    survival::coxph(cox_formula_tv, data = as.data.frame(train_cox),
                    cluster = stay_id, ties = "efron")
  }, error = function(e) {
    v2_log("  Cox fit error: ", e$message, level = "WARN"); NULL
  })

  if (!is.null(cox_fit)) {
    schoen_test <- tryCatch(survival::cox.zph(cox_fit, transform = "log"),
                            error = function(e) NULL)
    if (!is.null(schoen_test)) {
      schoen_df            <- as.data.frame(schoen_test$table)
      schoen_df$covariate  <- rownames(schoen_df)
      schoen_df$variant    <- vid
      save_parquet(schoen_df, paste0("06_cox_schoenfeld_", vid), DATA_DIR)
      cat(sprintf("  Cox PH test: global p = %.3f (p < 0.05 = PH violated, expected)\n",
                  schoen_test$table["GLOBAL", "p"]))
    }
    test_cox <- test[, .SD, .SDcols = c("stay_id", "hour", "event_bin", cox_feats)]
    test_cox[, stop := hour + 1L]
    lp <- tryCatch(
      predict(cox_fit, newdata = as.data.frame(test_cox), type = "lp"),
      error = function(e) rep(NA_real_, nrow(test_cox))
    )
    test[, pred_cox := plogis(lp)]
    results[["cox"]] <- test[, .(stay_id, hour, event, event_bin, pred_cox)]
    saveRDS(cox_fit, file.path(DATA_DIR, paste0("06_model_cox_", vid, ".rds")))
    cox_coef           <- as.data.frame(coef(summary(cox_fit)))
    cox_coef$covariate <- rownames(cox_coef)
    cox_coef$variant   <- vid
    save_parquet(cox_coef, paste0("06_coef_cox_", vid), DATA_DIR)
  }

  # --- Gradient-Boosted Trees ---
  step("  Comparator: Gradient-Boosted Trees (XGBoost)")
  gbt_feats <- avail_feats
  X_train   <- as.matrix(train[, ..gbt_feats])
  y_train   <- train$event_bin
  X_test    <- as.matrix(test[, ..gbt_feats])
  y_test    <- test$event_bin

  X_train[is.na(X_train)] <- 0
  X_test[is.na(X_test)]   <- 0
  X_train <- apply(X_train, 2, as.numeric)
  X_test  <- apply(X_test,  2, as.numeric)

  dtrain <- xgboost::xgb.DMatrix(data = X_train, label = y_train)
  dtest  <- xgboost::xgb.DMatrix(data = X_test,  label = y_test)

  n_pos      <- sum(y_train, na.rm = TRUE)
  n_neg      <- sum(!y_train, na.rm = TRUE)
  scale_pos  <- n_neg / max(n_pos, 1)

  gbt_params <- list(
    objective        = "binary:logistic",
    eval_metric      = "aucpr",
    max_depth        = 6,
    eta              = 0.05,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    min_child_weight = 5,
    scale_pos_weight = scale_pos,
    nthread          = 4
  )

  gbt_fit <- tryCatch({
    watchlist <- list(train = dtrain, test = dtest)
    xgboost::xgb.train(
      params                = gbt_params,
      data                  = dtrain,
      nrounds               = 300,
      evals                 = watchlist,
      early_stopping_rounds = 30,
      verbose               = 0
    )
  }, error = function(e) {
    v2_log("  XGBoost error: ", e$message, level = "WARN"); NULL
  })

  if (!is.null(gbt_fit)) {
    pred_gbt <- predict(gbt_fit, dtest)
    test[, pred_gbt := pred_gbt]
    results[["gbt"]] <- test[, .(stay_id, hour, event, event_bin, pred_gbt)]
    saveRDS(gbt_fit, file.path(DATA_DIR, paste0("06_model_gbt_", vid, ".rds")))
    feat_imp <- xgboost::xgb.importance(feature_names = gbt_feats, model = gbt_fit)
    if (!is.null(feat_imp)) {
      feat_imp$variant <- vid
      save_parquet(as.data.frame(feat_imp),
               paste0("06_gbt_feature_importance_", vid), DATA_DIR)
    }
    cat(sprintf("  GBT fitted: best iteration %d\n", gbt_fit$best_iteration))
  }

  # Combine comparator predictions
  all_preds <- test[, .(stay_id, hour, event, event_bin, onset_hour,
                          anchor_year_group, temporal_split,
                          first_abx_hour = if ("first_abx_hour" %in% names(test))
                                             first_abx_hour else NA,
                          first_culture_hour = if ("first_culture_hour" %in% names(test))
                                                first_culture_hour else NA)]
  for (nm in names(results)) {
    pred_col <- grep("^pred_", names(results[[nm]]), value = TRUE)
    if (length(pred_col) > 0)
      all_preds <- merge(all_preds, results[[nm]][, c("stay_id", "hour", pred_col),
                                                   with = FALSE],
                         by = c("stay_id", "hour"), all.x = TRUE)
  }
  all_preds[, variant := vid]
  save_parquet(all_preds, paste0("06_predictions_comparators_", vid), DATA_DIR)
  cat(sprintf("  Comparator predictions saved for Variant %s\n", vid))
  all_preds
}
```

```{r run-comparators}
all_comparator_preds        <- lapply(names(LABEL_VARIANTS), fit_comparators)
names(all_comparator_preds) <- names(LABEL_VARIANTS)
```

## Random-Split Comparators (H4)

GBT and rule-based comparators are re-fitted on the random patient-level split for Variant B, so H4 can compare temporal vs. random AUROC across all models.

```{r fit-random-comparators}
step("--- Random-split comparators for Variant B (H4) ---")

vid <- "B"
ph <- as.data.table(arrow::read_parquet(
  file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
))

if ("random_split" %in% names(ph)) {
  train_r <- ph[random_split == "train"]
  test_r  <- ph[random_split == "test"]

  avail_feats <- intersect(BASE_FEATURES, names(ph))
  for (f in avail_feats) {
    if (is.numeric(train_r[[f]])) train_r[is.na(get(f)), (f) := 0]
    if (is.logical(train_r[[f]])) train_r[is.na(get(f)), (f) := FALSE]
    if (is.numeric(test_r[[f]]))  test_r[is.na(get(f)),  (f) := 0]
    if (is.logical(test_r[[f]]))  test_r[is.na(get(f)),  (f) := FALSE]
  }
  train_r[, event_bin := as.integer(event == 1)]
  test_r[,  event_bin := as.integer(event == 1)]

  results_r <- list()

  # NEWS2/qSOFA/SIRS (no re-fit needed)
  if ("news2_total" %in% names(test_r)) {
    test_r[, pred_news2 := as.numeric(news2_total) / 20]
    results_r[["news2"]] <- test_r[, .(stay_id, hour, event, event_bin, pred_news2)]
  }
  if ("qsofa_score" %in% names(test_r))
    test_r[, pred_qsofa := qsofa_score / 3]
  if ("sirs_score" %in% names(test_r))
    test_r[, pred_sirs := sirs_score / 4]

  # Cox PH on random split
  cox_feats_r <- intersect(c("hr", "resp_rate", "spo2", "map", "temp_c", "gcs",
                              "lactate", "creatinine", "bilirubin_total",
                              "platelets", "wbc", "vaso_any"), names(train_r))
  cox_fit_r <- tryCatch({
    train_cox_r <- train_r[, .SD, .SDcols = c("stay_id", "hour", "event_bin", cox_feats_r)]
    train_cox_r[, stop := hour + 1L]
    cox_formula_r <- as.formula(paste(
      "Surv(hour, stop, event_bin) ~", paste(cox_feats_r, collapse = " + ")
    ))
    survival::coxph(cox_formula_r, data = as.data.frame(train_cox_r),
                    cluster = stay_id, ties = "efron")
  }, error = function(e) { v2_log("  Random Cox error: ", e$message, level = "WARN"); NULL })

  if (!is.null(cox_fit_r)) {
    test_cox_r <- test_r[, .SD, .SDcols = c("stay_id", "hour", "event_bin", cox_feats_r)]
    test_cox_r[, stop := hour + 1L]
    lp_r <- tryCatch(
      predict(cox_fit_r, newdata = as.data.frame(test_cox_r), type = "lp"),
      error = function(e) rep(NA_real_, nrow(test_cox_r))
    )
    test_r[, pred_cox := plogis(lp_r)]
    results_r[["cox"]] <- test_r[, .(stay_id, hour, event, event_bin, pred_cox)]
  }

  # GBT on random split
  gbt_feats_r <- avail_feats
  X_train_r   <- as.matrix(train_r[, ..gbt_feats_r])
  y_train_r   <- train_r$event_bin
  X_test_r    <- as.matrix(test_r[, ..gbt_feats_r])
  y_test_r    <- test_r$event_bin
  X_train_r[is.na(X_train_r)] <- 0
  X_test_r[is.na(X_test_r)]   <- 0
  X_train_r <- apply(X_train_r, 2, as.numeric)
  X_test_r  <- apply(X_test_r,  2, as.numeric)

  dtrain_r <- xgboost::xgb.DMatrix(data = X_train_r, label = y_train_r)
  dtest_r  <- xgboost::xgb.DMatrix(data = X_test_r,  label = y_test_r)

  n_pos_r     <- sum(y_train_r, na.rm = TRUE)
  n_neg_r     <- sum(!y_train_r, na.rm = TRUE)
  scale_pos_r <- n_neg_r / max(n_pos_r, 1)

  gbt_params_r <- list(
    objective        = "binary:logistic",
    eval_metric      = "aucpr",
    max_depth        = 6,
    eta              = 0.05,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    min_child_weight = 5,
    scale_pos_weight = scale_pos_r,
    nthread          = 4
  )

  gbt_fit_r <- tryCatch({
    watchlist_r <- list(train = dtrain_r, test = dtest_r)
    xgboost::xgb.train(
      params                = gbt_params_r,
      data                  = dtrain_r,
      nrounds               = 300,
      evals                 = watchlist_r,
      early_stopping_rounds = 30,
      verbose               = 0
    )
  }, error = function(e) { v2_log("  Random GBT error: ", e$message, level = "WARN"); NULL })

  if (!is.null(gbt_fit_r)) {
    test_r[, pred_gbt := predict(gbt_fit_r, dtest_r)]
    results_r[["gbt"]] <- test_r[, .(stay_id, hour, event, event_bin, pred_gbt)]
    cat(sprintf("  Random-split GBT: best iteration %d\n", gbt_fit_r$best_iteration))
  }

  # Combine random-split comparator predictions
  all_preds_r <- test_r[, .(stay_id, hour, event, event_bin, onset_hour,
                             anchor_year_group, random_split,
                             first_abx_hour = if ("first_abx_hour" %in% names(test_r))
                                                first_abx_hour else NA,
                             first_culture_hour = if ("first_culture_hour" %in% names(test_r))
                                                   first_culture_hour else NA)]
  for (nm in names(results_r)) {
    pred_col <- grep("^pred_", names(results_r[[nm]]), value = TRUE)
    if (length(pred_col) > 0)
      all_preds_r <- merge(all_preds_r, results_r[[nm]][, c("stay_id", "hour", pred_col),
                                                         with = FALSE],
                           by = c("stay_id", "hour"), all.x = TRUE)
  }
  all_preds_r[, variant := vid]
  save_parquet(all_preds_r, paste0("06_predictions_comparators_random_", vid), DATA_DIR)
  cat(sprintf("  Random-split comparator predictions saved for Variant %s\n", vid))
} else {
  cat("random_split column not found in person-hours; skipping H4 comparators.\n")
}
```

## Feature Importance (GBT)

GBT feature importance provides a cross-check: if GBT relies on the same features as the multinomial regression, both model classes are learning similar structure.

```{r gbt-importance}
imp_files <- list.files(DATA_DIR, pattern = "^06_gbt_feature_importance_.*\\.parquet$", full.names = TRUE)
if (length(imp_files) > 0) {
  imp_all <- rbindlist(lapply(imp_files, function(f) as.data.table(arrow::read_parquet(f))))
  if (nrow(imp_all) > 0) {
    top_imp <- imp_all[variant == "B"][order(-Gain)][1:min(20, .N)]
    knitr::kable(top_imp[, .(Feature, Gain = round(Gain, 4), Cover = round(Cover, 4),
                              Frequency = round(Frequency, 4))],
                 caption = "Top 20 GBT features by Gain (Variant B)")
  }
}
```
