---
title: "07 Sample Size - Riley et al. (2019) Power Calculation"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)   # student ID for reproducibility

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

Riley et al. (2019) provide minimum sample size criteria for prediction models: (i) shrinkage $\geq 0.9$, (ii) Nagelkerke $R^2$ optimism $\leq 0.05$, and (iii) adequate event-rate precision.

The calculation is repeated per label variant. The effective sample size depends on independent **onset events**, not total person-hour rows, because rows within a stay are correlated. Anticipated C-statistic = 0.846 (Moor et al., 2023).

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(pmsampsize)
})

source("config.R")
source("utils.R")

step("=== Sample Size Calculation (Riley et al., 2019) ===")
```

## Label-Variance Recap

Cohort and event counts from the label-variance table (produced in 02) are the inputs.

```{r load-label-variance}
lv_path <- file.path(OUTPUT_DIR, "02_label_variance_table.parquet")
if (!file.exists(lv_path))
  stop("02_label_variance_table.parquet not found. Run 02_cohort_labels.Rmd first.")

lv <- as.data.table(arrow::read_parquet(lv_path))
knitr::kable(lv, caption = "Label-variance table (from 02\\_cohort\\_labels.Rmd)")
```

## Sample Size per Variant

For each variant, the training-set person-hour table is loaded and `pmsampsize` called with the anticipated C-statistic of 0.846. Events-per-variable (EPV) is reported as a secondary reference.

```{r sample-size-calc}
results <- list()

for (i in seq_len(nrow(lv))) {
  row <- lv[i]
  vid <- row$variant

  ph_path <- file.path(OUTPUT_DIR, paste0("03_person_hours_", vid, ".parquet"))
  if (!file.exists(ph_path)) {
    v2_log(sprintf("  Variant %s: person_hours not found, skipping", vid), level = "WARN")
    next
  }

  ph       <- arrow::read_parquet(ph_path, col_select = c("stay_id", "event", "temporal_split"))
  ph_train <- ph[ph$temporal_split == "train", ]

  n_person_hours <- nrow(ph_train)
  n_events       <- sum(ph_train$event == 1, na.rm = TRUE)
  n_stays        <- length(unique(ph_train$stay_id))
  event_rate_ph  <- n_events / n_person_hours

  cat(sprintf("\n--- Variant %s (%s) ---\n", vid, row$name))
  cat(sprintf("  Training person-hours: %s\n", format(n_person_hours, big.mark = ",")))
  cat(sprintf("  Training stays:        %s\n", format(n_stays,         big.mark = ",")))
  cat(sprintf("  Onset events (train):  %s\n", format(n_events,        big.mark = ",")))
  cat(sprintf("  Per-hour event rate:   %.4f%%\n", 100 * event_rate_ph))

  ss_res <- tryCatch(
    pmsampsize::pmsampsize(
      type       = "b",
      rsquared   = NULL,
      parameters = P_CANDIDATE_PREDICTORS,
      shrinkage  = 0.9,
      prevalence = event_rate_ph,
      cstatistic = ANTICIPATED_AUC,
      seed       = 42
    ),
    error = function(e) {
      v2_log(sprintf("  pmsampsize error: %s", e$message), level = "WARN")
      NULL
    }
  )

  if (!is.null(ss_res)) {
    min_n_ph <- ss_res$sample_size
    cat(sprintf("  Minimum person-hours (Riley): %s\n", format(min_n_ph, big.mark = ",")))
    cat(sprintf("  Available person-hours:       %s\n", format(n_person_hours, big.mark = ",")))
    cat(sprintf("  Sufficient (person-hours):    %s\n",
                if (n_person_hours >= min_n_ph) "YES" else "NO - variant may be underpowered"))
    cat(sprintf("  NOTE: effective information = onset EVENTS (%d), not person-hours.\n",
                n_events))
  }

  epv <- if (P_CANDIDATE_PREDICTORS > 0) n_events / P_CANDIDATE_PREDICTORS else NA
  cat(sprintf("  Events-per-variable (EPV): %.1f (reference only)\n", epv))

  results[[vid]] <- data.frame(
    variant             = vid,
    name                = row$name,
    n_train_person_h    = n_person_hours,
    n_train_stays       = n_stays,
    n_onset_events      = n_events,
    per_hour_event_rate = round(event_rate_ph, 5),
    min_n_riley         = if (!is.null(ss_res)) ss_res$sample_size else NA,
    sufficient          = if (!is.null(ss_res)) n_person_hours >= ss_res$sample_size else NA,
    epv                 = round(epv, 1),
    anticipated_auc     = ANTICIPATED_AUC,
    n_candidate_preds   = P_CANDIDATE_PREDICTORS
  )
}
```

## Summary and Interpretation

The table below reports minimum required person-hours (Riley criterion), available training person-hours, and whether the sample is sufficient.

```{r summary-table}
if (length(results) > 0) {
  result_df <- do.call(rbind, results)
  knitr::kable(result_df,
               caption = "Sample size adequacy (Riley et al., 2019) per label variant")
  save_parquet(result_df, "07_sample_size", OUTPUT_DIR)
  cat("\nSaved: 07_sample_size.parquet\n")
}
```
