---
title: "08 External Validation - eICU-CRD External Validation"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

External validation on a distinct dataset is the minimum standard for generalisability (Collins et al., 2024, TRIPOD+AI). The eICU-CRD (Pollard et al., 2018) provides multi-centre US ICU data with different demographics, protocols, and data infrastructure.

Expected AUC degradation: ~0.085 (Moor et al., 2023). This analysis tests **H5**: Utility Score degrades proportionally more than AUROC on external data.

Steps:

1. Build eICU person-hour table harmonised to MIMIC-IV naming
2. Apply Sepsis-3 Variant B onset definition
3. Load MIMIC-IV trained models (no retraining)
4. Evaluate with the same metric suite
5. Compare internal vs. external performance

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(nnet)
  library(splines)
  library(xgboost)
  library(PRROC)
  library(dplyr)
})

source("config.R")
source("utils.R")
source("clinical_scores.R")

step("=== Phase 5: External Validation on eICU-CRD ===")
cat(sprintf("Pre-registered expected degradation: %.3f AUC\n",
            EXPECTED_EXTERNAL_DEGRADATION))
```

## Load eICU Patient Table

Same inclusion filters as MIMIC-IV: age $\geq$ 18, LOS $\geq$ 4 hours.

```{r load-eicu-patients}
step("Loading eICU tables...")

eicu_patient <- data.table::fread(file.path(EICU_DIR, "patient.csv.gz"),
                                   showProgress = FALSE)
cat(sprintf("eICU patients (raw): %s\n", format(nrow(eicu_patient), big.mark = ",")))

eicu_patient[, age_num := suppressWarnings(as.numeric(age))]
eicu_patient[age == ">89", age_num := 90]
eicu_patient[, los_hours := unitdischargeoffset / 60]
eicu_patient <- eicu_patient[!is.na(age_num) & age_num >= MIN_AGE &
                               los_hours >= MIN_ICU_LOS_HOURS]
cat(sprintf("eICU cohort after exclusions: %s\n",
            format(nrow(eicu_patient), big.mark = ",")))
```

## Vital Signs (vitalPeriodic)

Loaded via DuckDB with pushdown filters on cohort membership and time window.

```{r load-vitals}
step("Loading eICU vitals...")

vp_path <- file.path(EICU_DIR, "vitalPeriodic.csv.gz")
if (file.exists(vp_path)) {
  suppressPackageStartupMessages(library(duckdb))
  .vp_con <- dbConnect(duckdb(), dbdir = ":memory:")
  dbWriteTable(.vp_con, "cohort_stays",
               data.frame(patientunitstayid = as.integer(eicu_patient$patientunitstayid)))
  sql_vp <- sprintf(
    "SELECT t.patientunitstayid, t.observationoffset,
            t.heartrate, t.respiration, t.sao2,
            t.systemicsystolic, t.systemicmean, t.temperature
     FROM read_csv('%s', header=true, compression='gzip') t
     INNER JOIN cohort_stays c ON t.patientunitstayid = c.patientunitstayid
     WHERE t.observationoffset >= 0 AND t.observationoffset <= %d",
    vp_path, MAX_PANEL_HOURS * 60L
  )
  v2_log("DuckDB: loading vitalPeriodic filtered to cohort + 72h window...")
  vp <- as.data.table(dbGetQuery(.vp_con, sql_vp))
  dbDisconnect(.vp_con, shutdown = TRUE)
  cat(sprintf("vitalPeriodic (cohort-filtered): %s rows\n", format(nrow(vp), big.mark = ",")))

  setnames(vp, c("patientunitstayid","observationoffset","heartrate","respiration",
                  "sao2","systemicsystolic","systemicmean","temperature"),
               c("stay_id","obs_offset_m","hr","resp_rate","spo2","sbp","map","temp_c"))
  vp[, obs_hour := floor(obs_offset_m / 60)]
  vp[!is.na(temp_c) & (temp_c < 30 | temp_c > 45), temp_c := NA]
  vp_hr <- vp[, .(hr        = mean(hr,        na.rm = TRUE),
                   resp_rate = mean(resp_rate, na.rm = TRUE),
                   spo2      = mean(spo2,      na.rm = TRUE),
                   sbp       = mean(sbp,       na.rm = TRUE),
                   map       = mean(map,       na.rm = TRUE),
                   temp_c    = mean(temp_c,    na.rm = TRUE)),
               by = .(stay_id, obs_hour)]
  rm(vp); gc()
} else {
  vp_hr <- data.table()
  v2_log("vitalPeriodic not found", level = "WARN")
}
```

## Lab Values

Lab values are mapped from eICU names to MIMIC-IV conventions. PaO$_2$/FiO$_2$ ratios computed where both components exist.

```{r load-labs}
step("Loading eICU labs...")

lab_path <- file.path(EICU_DIR, "lab.csv.gz")
if (file.exists(lab_path)) {
  eicu_lab <- data.table::fread(lab_path, showProgress = FALSE,
                                 select = c("patientunitstayid", "labresultoffset",
                                            "labname", "labresult"))
  setnames(eicu_lab, c("patientunitstayid","labresultoffset","labname","labresult"),
                     c("stay_id","obs_offset_m","labname","value"))
  eicu_lab[, obs_hour := floor(obs_offset_m / 60)]
  eicu_lab <- eicu_lab[obs_hour >= 0 & obs_hour <= MAX_PANEL_HOURS & !is.na(value)]
  eicu_lab[, value := as.numeric(value)]

  lab_map <- c("lactate" = "lactate", "creatinine" = "creatinine",
               "total bilirubin" = "bilirubin_total", "platelets x 1000" = "platelets",
               "WBC x 1000" = "wbc", "paO2" = "pao2", "FiO2" = "fio2_lab",
               "sodium" = "sodium", "potassium" = "potassium")
  eicu_lab <- eicu_lab[labname %in% names(lab_map)]
  eicu_lab[, var_name := lab_map[labname]]

  lab_hr <- dcast(
    eicu_lab[, .(value = mean(value, na.rm = TRUE)),
              by = .(stay_id, obs_hour, var_name)],
    stay_id + obs_hour ~ var_name, value.var = "value", fun.aggregate = mean
  )
  if ("pao2" %in% names(lab_hr) && "fio2_lab" %in% names(lab_hr)) {
    lab_hr[, fio2_frac := ifelse(fio2_lab > 1, fio2_lab / 100, fio2_lab)]
    lab_hr[, pf_ratio  := ifelse(!is.na(pao2) & !is.na(fio2_frac) & fio2_frac > 0,
                                  pao2 / fio2_frac, NA_real_)]
    lab_hr[, c("fio2_lab", "fio2_frac") := NULL]
  }
} else {
  lab_hr <- data.table()
  v2_log("lab.csv not found", level = "WARN")
}
```

## Suspected Infection (eICU)

Variant B antibiotic-culture window logic applied to eICU medication and microbiology tables. eICU offsets are in minutes from ICU admission.

```{r eicu-suspected-infection}
step("Loading eICU infection events...")

micro_path <- file.path(EICU_DIR, "microLab.csv.gz")
med_path   <- file.path(EICU_DIR, "medication.csv.gz")
eicu_abx   <- data.table()
eicu_culture <- data.table()

if (file.exists(med_path)) {
  meds <- data.table::fread(med_path, showProgress = FALSE,
                              select = c("patientunitstayid","drugstartoffset","drugname"))
  meds[, is_abx := match_antibiotics(drugname, ANTIBIOTICS)]
  eicu_abx <- meds[is_abx == TRUE & !is.na(drugstartoffset),
                    .(stay_id = patientunitstayid,
                      abx_hour = floor(drugstartoffset / 60))]
}

if (file.exists(micro_path)) {
  micro <- data.table::fread(micro_path, showProgress = FALSE,
                               select = c("patientunitstayid","culturetakenoffset","culturesite"))
  eicu_culture <- micro[!is.na(culturetakenoffset),
                         .(stay_id = patientunitstayid,
                           culture_hour = floor(culturetakenoffset / 60))]
}

v <- LABEL_VARIANTS[["B"]]
eicu_susp <- data.table()
if (nrow(eicu_abx) > 0 && nrow(eicu_culture) > 0) {
  susp_raw <- merge(eicu_abx, eicu_culture, by = "stay_id", allow.cartesian = TRUE)
  susp_raw[, dt_abx_before := culture_hour - abx_hour]
  susp_raw[, dt_cul_before := abx_hour - culture_hour]
  susp_raw <- susp_raw[
    (dt_abx_before >= 0 & dt_abx_before <= v$abx_before_culture_h) |
    (dt_cul_before >= 0 & dt_cul_before <= v$culture_before_abx_h)
  ]
  if (nrow(susp_raw) > 0) {
    susp_raw[, t_susp_h := pmax(abx_hour, culture_hour)]
    setorderv(susp_raw, c("stay_id", "t_susp_h"))
    eicu_susp <- susp_raw[, .SD[1], by = stay_id][, .(stay_id, t_susp_h)]
  }
}
cat(sprintf("eICU suspected infection stays: %d\n", nrow(eicu_susp)))
```

## Build eICU Person-Hour Table

Same construction as MIMIC-IV (hourly grid, LOCF, missingness indicators, rolling slopes, clinical scores). GCS is unavailable in eICU vitalPeriodic and set to NA.

```{r build-eicu-person-hours}
step("Building eICU person-hour table...")

eicu_stays <- eicu_patient[, .(stay_id = patientunitstayid, los_hours,
                                age = age_num, gender, ethnicity,
                                hospital_expire = unitdischargestatus == "Expired")]
grid <- build_person_hour_grid(eicu_stays, max_hours = MAX_PANEL_HOURS)

if (nrow(vp_hr) > 0) {
  if ("obs_hour" %in% names(vp_hr)) setnames(vp_hr, "obs_hour", "hour")
  grid <- merge(grid, vp_hr, by = c("stay_id", "hour"), all.x = TRUE)
}
if (nrow(lab_hr) > 0) {
  if (!"hour" %in% names(lab_hr) && "obs_hour" %in% names(lab_hr))
    setnames(lab_hr, "obs_hour", "hour")
  grid <- merge(grid, lab_hr, by = c("stay_id", "hour"), all.x = TRUE)
}

# Same plausibility filter as the MIMIC panel in 03, applied before the
# forward-fill, so the external cohort is pre-processed identically to the
# training cohort rather than on its own terms.
apply_plausibility(grid, label = "eICU panel")

fill_cols <- intersect(c("hr","resp_rate","spo2","map","sbp","temp_c",
                          "lactate","creatinine","bilirubin_total","platelets",
                          "wbc","pf_ratio"), names(grid))
grid <- ffill_by_stay(grid, fill_cols)

for (mc in c("lactate","creatinine","bilirubin_total","platelets","wbc","pf_ratio")) {
  if (mc %in% names(grid))
    grid[[paste0(mc, "_measured")]] <- as.integer(!is.na(grid[[mc]]))
}
for (col in intersect(c("hr","resp_rate","map","temp_c"), names(grid)))
  grid <- rolling_mean_by_stay(grid, col, n_h = SLOPE_WINDOW_H,
                                col_out = paste0(col, "_slope6h"))

if (nrow(eicu_susp) > 0) {
  grid <- merge(grid, eicu_susp, by = "stay_id", all.x = TRUE)
  grid[, onset_hour := t_susp_h]
  grid[, vaso_any       := FALSE]
  grid[, norepi_epi_any := FALSE]
  if (!"bilirubin" %in% names(grid) && "bilirubin_total" %in% names(grid))
    grid[, bilirubin := bilirubin_total]
  if (!"gcs" %in% names(grid)) grid[, gcs := NA_real_]
  sofa_scores <- score_sofa(as.data.frame(grid))
  grid[, sofa_total := sofa_scores$sofa_total]
  grid[!is.na(onset_hour) & (is.na(sofa_total) | sofa_total < SOFA_INCREASE_THRESHOLD),
       onset_hour := NA_integer_]
}

grid <- merge(grid, eicu_stays[, .(stay_id, hospital_expire)], by = "stay_id", all.x = TRUE)
grid_last <- grid[, .(max_h = max(hour)), by = stay_id]
grid <- merge(grid, grid_last, by = "stay_id", all.x = TRUE)
grid[, event := 0L]
grid[!is.na(onset_hour) & hour == onset_hour,                     event := 1L]
grid[hour == max_h & is.na(onset_hour) & hospital_expire == TRUE,  event := 3L]
grid[hour == max_h & is.na(onset_hour) & hospital_expire == FALSE, event := 2L]
grid[, at_risk := is.na(onset_hour) | hour <= onset_hour]
grid <- grid[at_risk == TRUE]

nr_grid  <- nrow(grid)
fio2_v   <- grid$fio2;  if (is.null(fio2_v)) fio2_v  <- rep(NA_real_, nr_grid)
gcs_v    <- grid$gcs;   if (is.null(gcs_v))  gcs_v   <- rep(NA_real_, nr_grid)
spo2_v   <- grid$spo2;  if (is.null(spo2_v)) spo2_v  <- rep(NA_real_, nr_grid)
suppl_o2   <- !is.na(fio2_v) & fio2_v > 0.21
avpu_alert <- !is.na(gcs_v)  & gcs_v  >= 14
news2_scores <- news2(grid$resp_rate, spo2_v, suppl_o2, grid$sbp, grid$hr,
                       grid$temp_c, avpu_alert)
grid[, names(news2_scores) := news2_scores]
grid[, qsofa_score := qsofa(resp_rate, gcs, map)]

save_parquet(grid, "08_eicu_person_hours", DATA_DIR)
cat(sprintf("eICU person-hours: %s rows, %d events\n",
            format(nrow(grid), big.mark = ","), sum(grid$event == 1)))
```

## Apply MIMIC-IV Trained Models

Trained models applied to eICU data without retraining. Missing features are zero-filled (same as training). Predictions generated in chunks for memory.

```{r apply-models}
step("Applying trained models to eICU cohort...")

BASE_FEATURES <- c(
  "hr","resp_rate","spo2","map","temp_c","gcs",
  "lactate","creatinine","bilirubin_total","platelets","wbc",
  "vaso_any","norepi_epi_any",
  "hr_slope6h","resp_rate_slope6h","map_slope6h",
  "lactate_measured","creatinine_measured","bilirubin_total_measured",
  "platelets_measured","wbc_measured","pf_ratio_measured"
)

eval_results <- list()

for (vid in names(LABEL_VARIANTS)) {
  prim_model_path <- file.path(DATA_DIR, paste0("05_model_primary_", vid, ".rds"))
  gbt_model_path  <- file.path(DATA_DIR, paste0("06_model_gbt_",     vid, ".rds"))
  if (!file.exists(prim_model_path)) next

  eicu_pred   <- copy(grid)
  avail_feats <- intersect(BASE_FEATURES, names(eicu_pred))
  require_features(BASE_FEATURES, available = names(eicu_pred))
  for (f in avail_feats) {
    if (is.numeric(eicu_pred[[f]])) eicu_pred[is.na(get(f)), (f) := 0]
    if (is.logical(eicu_pred[[f]])) eicu_pred[is.na(get(f)), (f) := FALSE]
  }

  prim_mod       <- readRDS(prim_model_path)
  PRED_CHUNK     <- 500000L
  n_ep           <- nrow(eicu_pred)
  pred_sepsis_vec <- numeric(n_ep)
  for (ch_start in seq(1L, n_ep, by = PRED_CHUNK)) {
    ch_idx   <- ch_start:min(ch_start + PRED_CHUNK - 1L, n_ep)
    chunk_df <- as.data.frame(eicu_pred[ch_idx])
    if (inherits(prim_mod, "multinom")) {
      pr <- predict(prim_mod, newdata = chunk_df, type = "probs")
      pred_sepsis_vec[ch_idx] <- if (is.matrix(pr)) pr[, "sepsis"] else as.numeric(pr)
    } else {
      pred_sepsis_vec[ch_idx] <- predict(prim_mod, newdata = chunk_df, type = "response")
    }
    rm(chunk_df)
  }
  eicu_pred[, pred_sepsis := pred_sepsis_vec]
  rm(pred_sepsis_vec); gc()

  if (file.exists(gbt_model_path)) {
    gbt_mod  <- readRDS(gbt_model_path)
    gbt_feats <- gbt_mod$feature_names
    nr_ep     <- nrow(eicu_pred)
    gbt_cols  <- lapply(gbt_feats, function(f) {
      v <- if (f %in% names(eicu_pred)) as.numeric(eicu_pred[[f]]) else rep(0.0, nr_ep)
      v[is.na(v)] <- 0.0; v
    })
    names(gbt_cols) <- gbt_feats
    X_eicu <- as.matrix(as.data.table(gbt_cols))
    storage.mode(X_eicu) <- "double"
    eicu_pred[, pred_gbt := predict(gbt_mod, xgboost::xgb.DMatrix(X_eicu))]
    rm(X_eicu, gbt_cols); gc()
  }

  eicu_pred[, pred_news2 := news2_total / 20]

  y <- as.integer(eicu_pred$event == 1)
  for (mod_nm in c("pred_sepsis","pred_gbt","pred_news2")) {
    if (!mod_nm %in% names(eicu_pred)) next
    pr    <- as.numeric(eicu_pred[[mod_nm]])
    valid <- !is.na(pr)
    if (sum(valid) < 50) next

    roc_pr <- tryCatch({
      list(
        auroc = PRROC::roc.curve(pr[valid], weights.class0 = y[valid])$auc,
        auprc = PRROC::pr.curve(pr[valid],  weights.class0 = y[valid])$auc.integral
      )
    }, error = function(e) list(auroc = NA, auprc = NA))

    eval_results[[paste(vid, mod_nm, "eicu", sep = "-")]] <- data.frame(
      variant  = vid,
      model    = mod_nm,
      cohort   = "eicu",
      n_rows   = sum(valid),
      n_events = sum(y[valid]),
      auroc    = round(roc_pr$auroc, 4),
      auprc    = round(roc_pr$auprc, 4)
    )
  }

  save_cols_eicu <- c("stay_id","hour","event","pred_sepsis","onset_hour",
                      intersect(c("pred_gbt","pred_news2","news2_total","qsofa_score"),
                                names(eicu_pred)))
  save_parquet(eicu_pred[, ..save_cols_eicu],
               paste0("08_eicu_predictions_", vid), DATA_DIR)
  rm(eicu_pred, prim_mod); if (exists("gbt_mod")) rm(gbt_mod); gc()
}
```

## Results and H5 Test

External vs. internal performance comparison. H5 tests whether Utility Score degrades proportionally more than AUROC on external data.

```{r results-h5}
if (length(eval_results) > 0) {
  ext_df <- do.call(rbind, eval_results)
  knitr::kable(ext_df, caption = "External validation results (eICU-CRD)")
  save_parquet(ext_df, "08_external_validation_results", DATA_DIR)

  internal_path <- file.path(DATA_DIR, "07_metric_results.parquet")
  if (file.exists(internal_path)) {
    internal_res <- as.data.table(arrow::read_parquet(internal_path))
    cat("\nH5 test: AUROC vs Utility Score degradation\n")
    h5_rows <- list()
    for (vid in names(LABEL_VARIANTS)) {
      int_row <- internal_res[variant == vid & model == "primary" &
                                eval_window == "unrestricted"]
      ext_row <- ext_df[ext_df$variant == vid & ext_df$model == "pred_sepsis", ]
      if (nrow(int_row) > 0 && nrow(ext_row) > 0) {
        auroc_drop <- int_row$auroc - ext_row$auroc
        cat(sprintf("  V%s: Internal AUROC=%.3f, External AUROC=%.3f, drop=%.3f\n",
                    vid, int_row$auroc, ext_row$auroc, auroc_drop))
        cat(sprintf("  V%s: Expected drop ~ %.3f (Moor et al., 2023)\n",
                    vid, EXPECTED_EXTERNAL_DEGRADATION))
        if (auroc_drop < 0.02)
          cat("  ** WARNING: Very small drop - external cohort may be insufficiently distant **\n")
        h5_rows[[vid]] <- data.frame(variant = vid,
                                      internal_auroc = int_row$auroc,
                                      external_auroc = ext_row$auroc,
                                      auroc_drop     = round(auroc_drop, 4),
                                      expected_drop  = EXPECTED_EXTERNAL_DEGRADATION)
      }
    }
    if (length(h5_rows) > 0)
      knitr::kable(do.call(rbind, h5_rows),
                   caption = "H5: Internal vs external AUROC")
  }
}

cat("\nExternal validation complete.\n")
```
