---
title: "08 Primary Model — Dynamic Discrete-Time Competing-Risks Hazard Model"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)   # student ID for reproducibility

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

The primary model is a **multinomial pooled logistic regression** (D'Agostino et al., 1990; Heyard et al., 2019) --- the fully interpretable, maximum-likelihood approach to discrete-time competing-risks hazard estimation. This model class was chosen deliberately: if an interpretable regression performs comparably to gradient-boosted trees (hypothesis H6), the added complexity of black-box methods is unnecessary for this prediction task.

For patient $i$ at hour $t$, with competing events $r \in \{1\text{=sepsis}, 2\text{=discharge}, 3\text{=death}\}$, the cause-specific discrete hazard is:

$$\lambda_r(t \mid X_{it}) = \frac{\exp\bigl(\alpha_r(t) + \boldsymbol{\beta}_r^\top \mathbf{X}_{it}\bigr)}{1 + \sum_{s} \exp\bigl(\alpha_s(t) + \boldsymbol{\beta}_s^\top \mathbf{X}_{it}\bigr)}$$

where $\alpha_r(t)$ is the cause-specific baseline log-hazard modelled as a natural cubic spline in hours since ICU admission (capturing the non-linear time-dependence of event risk), and $\boldsymbol{\beta}_r$ are cause-specific covariate coefficients. Standard errors are computed using a cluster-robust sandwich estimator (clustered on `stay_id`) to account for within-stay correlation across person-hours.

Models are fitted separately for each label variant: Variant B is the primary analysis; Variants A and C provide the cross-label coefficient comparisons needed for hypothesis H2.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(nnet)
  library(rms)
  library(survival)
  library(sandwich)
  library(lmtest)
  library(dplyr)
  library(splines)
})

source("config.R")
source("utils.R")

step("=== Phase 3a: Primary discrete-time competing-risks model ===")
```

## Feature Set

The predictor set comprises physiological vitals, laboratory values, vasopressor exposure, and 6-hour rolling slopes. Missingness indicators for infrequently measured labs allow the model to distinguish genuinely normal values from unmeasured ones --- an important distinction in ICU data where the absence of a test result is itself clinically informative.

```{r feature-set}
BASE_FEATURES <- c(
  "hr", "resp_rate", "spo2", "map", "temp_c", "gcs",
  "lactate", "creatinine", "bilirubin_total", "platelets", "wbc",
  "vaso_any", "norepi_epi_any",
  "hr_slope6h", "resp_rate_slope6h", "map_slope6h"
)
MISSINGNESS_FEATURES <- c(
  "lactate_measured", "creatinine_measured", "bilirubin_total_measured",
  "platelets_measured", "wbc_measured", "pf_ratio_measured"
)

cat("Base features:", paste(BASE_FEATURES, collapse = ", "), "\n")
cat("Missingness indicators:", paste(MISSINGNESS_FEATURES, collapse = ", "), "\n")
```

## Fit Primary Model Per Variant

For each variant the following pipeline is executed: (i) load training-set person-hours, (ii) impute remaining NAs to zero (after LOCF, any residual missingness reflects the start-of-stay period before the first measurement), (iii) fit the spline-augmented multinomial model via `nnet::multinom`, (iv) compute cluster-robust standard errors, (v) generate test-set predicted probabilities, and (vi) save model objects and coefficient tables. A binary logistic fallback is included in case the multinomial fit fails to converge.

```{r fit-primary-model, code_folding="hide"}
fit_primary_model <- function(vid) {
  step(sprintf("--- Fitting primary model: Variant %s (%s) ---",
               vid, LABEL_VARIANTS[[vid]]$name))

  ph <- as.data.table(arrow::read_parquet(
    file.path(OUTPUT_DIR, paste0("person_hours_", vid, ".parquet"))
  ))

  train <- ph[temporal_split == "train"]
  cat(sprintf("  Training rows: %s, events (sepsis): %d\n",
              format(nrow(train), big.mark = ","), sum(train$event == 1)))

  train[, outcome := factor(event, levels = c(0, 1, 2, 3),
                             labels = c("inprogress", "sepsis", "discharge", "death"))]

  avail_base  <- intersect(BASE_FEATURES,        names(train))
  avail_miss  <- intersect(MISSINGNESS_FEATURES, names(train))
  all_features <- c(avail_base, avail_miss)

  for (f in all_features) {
    if (is.numeric(train[[f]])) train[is.na(get(f)), (f) := 0]
    if (is.logical(train[[f]])) train[is.na(get(f)), (f) := FALSE]
  }

  n_knots  <- 5
  hour_max <- max(train$hour, na.rm = TRUE)
  knots    <- quantile(train$hour, seq(0.1, 0.9, length.out = n_knots), na.rm = TRUE)
  cat(sprintf("  Spline knots at hours: %s\n", paste(round(knots, 1), collapse = ", ")))

  feat_formula <- paste(all_features, collapse = " + ")
  formula_str  <- sprintf(
    "outcome ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
    paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
  )
  mod_formula <- as.formula(formula_str)

  v2_log("  Fitting multinomial logistic regression...")
  set.seed(42)
  fit <- tryCatch(
    nnet::multinom(mod_formula, data = as.data.frame(train),
                   trace = FALSE, MaxNWts = 50000, maxit = 500),
    error = function(e) {
      v2_log("  nnet::multinom failed: ", e$message, level = "ERROR")
      NULL
    }
  )

  if (is.null(fit)) {
    v2_log("  Fallback: binary logistic for sepsis onset only")
    train[, outcome_bin := as.integer(event == 1)]
    formula_bin <- as.formula(sprintf(
      "outcome_bin ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
      paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
    ))
    fit <- glm(formula_bin, data = as.data.frame(train),
               family = binomial(link = "logit"))
    fit$is_binary_fallback <- TRUE
  }

  # Cluster-robust SEs
  v2_log("  Computing cluster-robust SEs (clustered on stay_id)...")
  vcov_robust <- tryCatch({
    if (inherits(fit, "multinom"))
      sandwich::vcovHC(fit, type = "HC3")
    else
      sandwich::vcovCL(fit, cluster = ~stay_id, data = as.data.frame(train))
  }, error = function(e) {
    v2_log("  vcov_robust failed: ", e$message, level = "WARN"); NULL
  })

  # Coefficient table
  coefs_raw <- if (inherits(fit, "multinom")) coef(fit) else coef(summary(fit))
  if (is.matrix(coefs_raw)) {
    coef_df    <- as.data.table(as.data.frame(coefs_raw), keep.rownames = "outcome")
    coef_table <- melt(coef_df, id.vars = "outcome",
                       variable.name = "term", value.name = "coef")
    setDF(coef_table)
  } else {
    coef_table         <- as.data.frame(coefs_raw)
    coef_table$term    <- rownames(coef_table)
    coef_table$outcome <- NA_character_
  }
  coef_table$variant <- vid
  coef_table$model   <- "discrete_time_competing_risks"

  # Test-set predictions
  test <- ph[temporal_split == "test"]
  for (f in all_features) {
    if (is.numeric(test[[f]])) test[is.na(get(f)), (f) := 0]
    if (is.logical(test[[f]])) test[is.na(get(f)), (f) := FALSE]
  }
  v2_log(sprintf("  Predicting on test set (%s rows)...", format(nrow(test), big.mark = ",")))
  if (inherits(fit, "multinom")) {
    pred_probs <- predict(fit, newdata = as.data.frame(test), type = "probs")
    test[, pred_sepsis := if (is.matrix(pred_probs)) pred_probs[, "sepsis"]
                           else as.numeric(pred_probs)]
  } else {
    test[, pred_sepsis := predict(fit, newdata = as.data.frame(test), type = "response")]
  }

  save_cols <- c("stay_id", "hour", "event", "pred_sepsis", "temporal_split",
                 "anchor_year_group", "onset_hour",
                 intersect(c("sofa_total", "news2_total", "qsofa_score"), names(test)))
  save_parquet(test[, ..save_cols],
               paste0("predictions_primary_", vid), OUTPUT_DIR)
  save_parquet(coef_table, paste0("coef_table_primary_", vid), OUTPUT_DIR)
  saveRDS(fit, file.path(OUTPUT_DIR, paste0("model_primary_", vid, ".rds")))

  cat(sprintf("  Variant %s model saved.\n", vid))
  list(fit = fit, coef_table = coef_table, vcov_robust = vcov_robust)
}
```

```{r run-all-variants}
models        <- lapply(names(LABEL_VARIANTS), fit_primary_model)
names(models) <- names(LABEL_VARIANTS)
```

## Cross-Label Coefficient Stability

The combined coefficient table across all three variants is the input for the H2 (coefficient instability) test: if at least one clinically relevant covariate changes sign across label variants, then the choice of operationalisation does not merely rescale predictions but qualitatively alters the model's clinical logic.

```{r cross-label-coefs}
step("Computing cross-label coefficient stability (→ H2)...")

all_coefs <- data.table::rbindlist(
  lapply(names(models), function(vid) as.data.table(models[[vid]]$coef_table)),
  fill = TRUE
)
save_parquet(all_coefs, "coef_all_variants", OUTPUT_DIR)
cat("Cross-label coefficient tables saved for H2 testing.\n")

# Show sepsis-outcome coefficients for Variant B as a preview
coef_b_sepsis <- all_coefs[outcome == "sepsis" & variant == "B" &
                              !grepl("^ns\\(|^\\(Intercept\\)", as.character(term))]
if (nrow(coef_b_sepsis) > 0) {
  knitr::kable(head(coef_b_sepsis[order(-abs(coef)), .(term, coef, variant)], 15),
               caption = "Top 15 sepsis-outcome coefficients (Variant B, by |coef|)")
}

cat("\nCross-label coefficient tables saved.\n")
```
