---
title: "09 Variance Decomposition - Attribution of Performance Variation"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

Central output of V2: a variance decomposition attributing performance variation to each analytical decision. Published sepsis studies vary across multiple factors simultaneously (Cohen et al., 2024); this analysis varies one factor at a time.

Five factors:

| Factor | Levels |
|--------|--------|
| Model class | Discrete-time hazard, Cox PH, GBT, NEWS2 |
| Label operationalisation | Variant A (Narrow), B (Seymour-Standard), C (Liberal) |
| Validation split | Temporal, random |
| Prediction-time anchoring | Treatment-anchored, unrestricted |
| Evaluation metric | Utility Score, calibration, AUROC, AUPRC |

Six pre-registered hypotheses structure the interpretation:

| \# | Claim | Threshold |
|----|-------|-----------|
| H1 | Label-variant spread $\geq$ model-class spread in AUROC | label spread $\geq$ model spread |
| H2 | Coefficient instability across labels | $\geq$ 1 covariate changes sign |
| H3 | Treatment-anchored AUROC $<$ unrestricted | $\Delta \geq 0.03$ |
| H4 | Temporal-split AUROC $<$ random-split | $\Delta \geq 0.02$ |
| H5 | Utility degradation $>$ AUROC degradation (internal $\to$ external) | ratio comparison |
| H6 | $|\text{primary} - \text{GBT}| \leq 0.02$ AUROC | $|\Delta| \leq 0.02$ |

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
})

source("config.R")
source("utils.R")

step("=== Phase 6a: Variance Decomposition ===")
```

## Load Metric Results

Metric results from 10, external validation from 11, and coefficient table from 08 are loaded.

```{r load-metrics}
metric_path <- file.path(DATA_DIR, "07_metric_results.parquet")
ext_path    <- file.path(DATA_DIR, "08_external_validation_results.parquet")
coef_path   <- file.path(DATA_DIR, "05_coef_all_variants.parquet")

if (!file.exists(metric_path))
  stop("07_metric_results.parquet not found. Run 07_metrics_suite.Rmd first.")

metrics    <- as.data.table(arrow::read_parquet(metric_path))
ext_res    <- if (file.exists(ext_path))  as.data.table(arrow::read_parquet(ext_path))  else NULL
coef_table <- if (file.exists(coef_path)) as.data.table(arrow::read_parquet(coef_path)) else NULL

cat(sprintf("Metric results loaded: %d rows\n", nrow(metrics)))
knitr::kable(metrics[, .(variant, model, eval_window, auroc, auprc,
                           utility_normalised, alert_burden, lead_time_h)],
             caption = "All evaluation results (from 07\\_metrics\\_suite.Rmd)")
```

## Variance by Factor

Each factor's spread is the range of AUROC (or Utility Score) across that factor's levels, with others held at reference values (Variant B, primary model, temporal split, unrestricted).

```{r variance-by-factor}
# Factor 1: MODEL CLASS - label=B, temporal, unrestricted
model_class_rows   <- metrics[variant == "B" & eval_window == "unrestricted"]
model_spread_auroc   <- if (nrow(model_class_rows) > 0) diff(range(model_class_rows$auroc, na.rm = TRUE)) else NA
model_spread_utility <- if (nrow(model_class_rows) > 0) diff(range(model_class_rows$utility_normalised, na.rm = TRUE)) else NA

# Factor 2: LABEL VARIANT - model=primary, temporal, unrestricted
label_rows         <- metrics[model == "primary" & eval_window == "unrestricted"]
label_spread_auroc   <- if (nrow(label_rows) > 0) diff(range(label_rows$auroc, na.rm = TRUE)) else NA
label_spread_utility <- if (nrow(label_rows) > 0) diff(range(label_rows$utility_normalised, na.rm = TRUE)) else NA

# Factor 3: PREDICTION-TIME ANCHORING - model=primary, variant=B, temporal
anchor_rows_B  <- metrics[model == "primary" & variant == "B"]
anchor_unrestr <- anchor_rows_B[eval_window == "unrestricted",         auroc]
anchor_treatm  <- anchor_rows_B[eval_window == "treatment_anchored",   auroc]
h3_delta_auroc <- if (length(anchor_unrestr) > 0 && length(anchor_treatm) > 0)
                    anchor_unrestr - anchor_treatm else NA

# Factor 4: VALIDATION SPLIT (temporal vs random)
random_path <- file.path(DATA_DIR, "07_metric_results_random.parquet")
if (file.exists(random_path)) {
  metrics_random <- as.data.table(arrow::read_parquet(random_path))
  temporal_auroc <- metrics[model == "primary" & variant == "B" &
                               eval_window == "unrestricted", auroc]
  random_auroc   <- metrics_random[model == "primary" & variant == "B" &
                                    eval_window == "unrestricted", auroc]
  h4_delta_auroc <- if (length(random_auroc) > 0 && length(temporal_auroc) > 0)
                      random_auroc - temporal_auroc else NA
} else {
  h4_delta_auroc <- NA
  cat("Random split results not found; H4 cannot be tested.\n")
}

# Factor 5: METRIC
internal_primary_B <- metrics[model == "primary" & variant == "B" & eval_window == "unrestricted"]
h5_internal_auroc  <- if (nrow(internal_primary_B) > 0) internal_primary_B$auroc else NA
h5_internal_util   <- if (nrow(internal_primary_B) > 0) internal_primary_B$utility_normalised else NA
if (!is.null(ext_res)) {
  ext_primary_B <- ext_res[variant == "B" & model == "pred_sepsis"]
  h5_ext_auroc  <- if (nrow(ext_primary_B) > 0) ext_primary_B$auroc else NA
} else { h5_ext_auroc <- NA }

# Factor 6: MODEL CLASS NULL (H6)
if (nrow(model_class_rows) > 0) {
  prim_auroc <- model_class_rows[model == "primary", auroc]
  gbt_auroc  <- model_class_rows[model == "gbt",     auroc]
  h6_delta   <- if (length(prim_auroc) > 0 && length(gbt_auroc) > 0)
                  abs(prim_auroc - gbt_auroc) else NA
} else { h6_delta <- NA }
```

## The Variance Decomposition Table

Primary deliverable. Each row quantifies how much a single analytical decision shifts reported performance.

```{r decomp-table}
decomp_table <- data.frame(
  factor = c(
    "Model class (label=B, temporal, unrestricted)",
    "Label operationalisation (model=primary, temporal, unrestricted)",
    "Prediction-time anchoring (model=primary, variant=B, temporal)",
    "Validation split (temporal vs random, model=primary, variant=B)",
    "Metric (AUROC vs Utility, internal→external, model=primary, variant=B)",
    "Model-class null |primary − GBT| (variant=B, temporal, unrestricted)"
  ),
  spread_auroc = round(c(model_spread_auroc, label_spread_auroc,
                          h3_delta_auroc, h4_delta_auroc,
                          if (!is.na(h5_internal_auroc) && !is.na(h5_ext_auroc))
                            h5_internal_auroc - h5_ext_auroc else NA,
                          h6_delta), 4),
  spread_utility = round(c(model_spread_utility, label_spread_utility,
                             NA, NA, NA, NA), 4),
  hypothesis  = c("-", "H1", "H3", "H4", "H5", "H6"),
  h_threshold = c("-", "label ≥ model", "Δ ≥ 0.03", "Δ ≥ 0.02",
                   "utility drop > AUROC drop", "|Δ| ≤ 0.02")
)

knitr::kable(decomp_table, caption = "THE VARIANCE DECOMPOSITION TABLE")
save_parquet(decomp_table, "09_variance_decomposition_table", DATA_DIR)
```

## Hypothesis Test Verdicts

Each hypothesis is evaluated against its threshold. "CONFIRMED" = effect meets/exceeds threshold; "NOT confirmed" = falls short. These are descriptive verdicts, not formal statistical tests.

```{r hypothesis-verdicts}
step("Hypothesis test verdicts...")
verdicts <- list()

# H1
if (!is.na(label_spread_auroc) && !is.na(model_spread_auroc)) {
  h1_result <- label_spread_auroc >= model_spread_auroc
  cat(sprintf("H1 (Label dominance): label_spread=%.3f, model_spread=%.3f → %s\n",
              label_spread_auroc, model_spread_auroc,
              if (h1_result) "CONFIRMED" else "NOT confirmed"))
  verdicts[["H1"]] <- data.frame(hypothesis = "H1", finding = "Label dominance",
                                  threshold = "label >= model",
                                  observed  = sprintf("label=%.3f, model=%.3f",
                                                      label_spread_auroc, model_spread_auroc),
                                  confirmed = h1_result)
}

# H2
if (!is.null(coef_table) && "term" %in% names(coef_table)) {
  coef_sep <- coef_table[outcome == "sepsis" &
                           !grepl("^ns\\(|^\\(Intercept\\)", as.character(term))]
  if (nrow(coef_sep) > 0) {
    common_terms <- Reduce(intersect, lapply(c("A","B","C"), function(v)
      unique(as.character(coef_sep[variant == v, term]))))
    sign_changes <- character(0)
    for (tm in common_terms) {
      vals  <- sapply(c("A","B","C"), function(v) {
        r <- coef_sep[variant == v & as.character(term) == tm, coef]
        if (length(r) == 0) NA_real_ else as.numeric(r[1])
      })
      signs <- sign(vals[!is.na(vals)])
      if (length(signs) >= 2 && diff(range(signs)) != 0)
        sign_changes <- c(sign_changes, tm)
    }
    h2_result <- length(sign_changes) >= 1
    cat(sprintf("H2 (Coefficient instability): %d sign-changes: %s → %s\n",
                length(sign_changes), paste(head(sign_changes, 3), collapse = ", "),
                if (h2_result) "CONFIRMED" else "NOT confirmed"))
    verdicts[["H2"]] <- data.frame(hypothesis = "H2", finding = "Coefficient instability",
                                    threshold = ">=1 covariate changes sign",
                                    observed  = paste(sign_changes, collapse = "; "),
                                    confirmed = h2_result)
  }
}

# H3
if (!is.na(h3_delta_auroc)) {
  h3_result <- h3_delta_auroc >= 0.03
  cat(sprintf("H3 (Leakage): unrestricted − anchored AUROC Δ=%.3f → %s\n",
              h3_delta_auroc, if (h3_result) "CONFIRMED" else "NOT confirmed (< 0.03)"))
  verdicts[["H3"]] <- data.frame(hypothesis = "H3", finding = "Leakage via treatment anchoring",
                                  threshold = "Δ ≥ 0.03",
                                  observed  = sprintf("Δ=%.3f", h3_delta_auroc),
                                  confirmed = h3_result)
}

# H4
if (!is.na(h4_delta_auroc)) {
  h4_result <- h4_delta_auroc >= 0.02
  cat(sprintf("H4 (Split optimism): random − temporal AUROC Δ=%.3f → %s\n",
              h4_delta_auroc, if (h4_result) "CONFIRMED" else "NOT confirmed (< 0.02)"))
  verdicts[["H4"]] <- data.frame(hypothesis = "H4", finding = "Split optimism",
                                  threshold = "Δ ≥ 0.02",
                                  observed  = sprintf("Δ=%.3f", h4_delta_auroc),
                                  confirmed = h4_result)
}

# H6
if (!is.na(h6_delta)) {
  h6_result <- h6_delta <= 0.02
  cat(sprintf("H6 (Model-class null): |primary − GBT| AUROC=%.3f → %s\n",
              h6_delta, if (h6_result) "NOT rejected (interpretable ≈ GBT)"
                        else "REJECTED (GBT wins by > 0.02)"))
  verdicts[["H6"]] <- data.frame(hypothesis = "H6", finding = "Model-class null",
                                  threshold = "|Δ| ≤ 0.02",
                                  observed  = sprintf("Δ=%.3f", h6_delta),
                                  confirmed = h6_result)
}

if (length(verdicts) > 0) {
  verdicts_df <- do.call(rbind, verdicts)
  knitr::kable(verdicts_df, caption = "Hypothesis test verdicts")
  save_parquet(verdicts_df, "09_hypothesis_verdicts", DATA_DIR)
}
```

## Variance Decomposition Plots

Label-variant effect vs. model-class effect on AUROC. If H1 is confirmed, the label-variant spread exceeds the model-class spread.

```{r variance-plots, fig.cap="Variance decomposition: label-variant and model-class effects on AUROC.", fig.width=6, fig.height=4}
if (nrow(metrics[!is.na(auroc)]) > 0) {
  # Label-variant effect
  tryCatch({
    p_label <- ggplot(metrics[model == "primary" & eval_window == "unrestricted"],
                       aes(x = variant, y = auroc, fill = variant)) +
      geom_col() +
      geom_hline(yintercept = if (!is.na(h5_internal_auroc)) h5_internal_auroc else 0.75,
                 linetype = "dashed") +
      labs(title    = "Label-variant effect on AUROC (primary model, temporal split)",
           subtitle = "H1: label spread should be ≥ model-class spread",
           x = "Label Variant", y = "AUROC") +
      theme_bw() + theme(legend.position = "none")
    print(p_label)
    ggsave(file.path(FIGURES_DIR, "09_label_variance_auroc.png"), p_label,
           width = 6, height = 4, dpi = 150)
  }, error = function(e) cat("Label plot failed:", e$message, "\n"))

  # Model-class effect
  tryCatch({
    p_model <- ggplot(metrics[variant == "B" & eval_window == "unrestricted" & !is.na(auroc)],
                       aes(x = model, y = auroc, fill = model)) +
      geom_col() +
      labs(title    = "Model-class effect on AUROC (Variant B, temporal split)",
           subtitle = "H6: |primary − GBT| ≤ 0.02",
           x = "Model", y = "AUROC") +
      theme_bw() + theme(axis.text.x = element_text(angle = 30, hjust = 1),
                         legend.position = "none")
    print(p_model)
    ggsave(file.path(FIGURES_DIR, "09_model_class_auroc.png"), p_model,
           width = 6, height = 4, dpi = 150)
  }, error = function(e) cat("Model plot failed:", e$message, "\n"))
}

cat("\nVariance decomposition complete.\n")
```

## Label-Anchor Attribution (Protocol Amendment 2)

Everything above varies the label *within* Sepsis-3. Variants A, B and C are
three readings of one definition and all three fire on the same
culture-plus-antibiotic trigger, so the label spread entering H1 is partly a
**definitional** quantity: it measures how much the reading of a definition
matters, not how much the choice of construct matters. Whether the physiology
itself carries the signal cannot be settled inside that family.

Sepsis-3 is a conjunction --- *suspected infection* **and** *acute organ
dysfunction*. This section splits it and fits the same model to each limb alone:

| Variant | Suspicion anchor | Organ dysfunction | Reads a treatment decision? |
|---|---|---|---|
| B (Seymour-Standard) | required | required | yes |
| E (Anchor-Only) | required | not required | yes |
| D (Deterioration) | not required | required | **no** |

Variant D is the treatment-independent labelling the decomposition was missing:
its onsets are derived from recorded physiology alone, so no antibiotic order,
culture draw or drug administration enters the label.

Two quantities are read off the comparison. The **anchor-attributable
fraction** is the share of Variant B's above-chance discrimination that is
reproduced by predicting the treatment anchor on its own,

$$\phi_{\text{anchor}} = \frac{\mathrm{AUROC}_E - 0.5}{\mathrm{AUROC}_B - 0.5},$$

and the **label agreement** statistics report how far the alternative labels
move the onsets themselves.

```{r anchor-attribution}
step("Label-anchor attribution (Protocol Amendment 2)...")

alt_path <- file.path(DATA_DIR, "07_metric_results_altlabel.parquet")
alt_res  <- if (file.exists(alt_path)) as.data.table(arrow::read_parquet(alt_path)) else NULL

attrib <- NULL
if (is.null(alt_res)) {
  cat("07_metric_results_altlabel.parquet not found; run 07 with variants D and E.\n")
} else {
  ref <- metrics[eval_window == "unrestricted" & variant == "B",
                 .(model, auroc_B = auroc, auprc_B = auprc,
                   utility_B = utility_normalised, n_events_B = n_sepsis_events)]

  alt <- alt_res[eval_window == "unrestricted",
                 .(variant, variant_name, model, auroc, auprc,
                   utility_normalised, n_sepsis_events)]

  attrib <- merge(alt, ref, by = "model")
  attrib[, `:=`(
    delta_auroc_vs_B  = round(auroc - auroc_B, 4),
    anchor_fraction   = ifelse(variant == "E" & auroc_B > 0.5,
                               round((auroc - 0.5) / (auroc_B - 0.5), 3), NA_real_),
    event_rate_ratio  = round(n_sepsis_events / n_events_B, 2)
  )]
  setorder(attrib, variant, -auroc)

  knitr::kable(
    attrib[, .(variant, label = variant_name, model,
               auroc = round(auroc, 4), auroc_B = round(auroc_B, 4),
               delta_auroc_vs_B, n_events = n_sepsis_events,
               event_rate_ratio, anchor_fraction)],
    caption = "Discrimination under each limb of the Sepsis-3 conjunction, against Variant B (temporal test, unrestricted window)"
  )
  save_parquet(attrib, "09_label_anchor_attribution", DATA_DIR)
  save_csv(attrib,     "09_label_anchor_attribution", DATA_DIR)

  phi <- attrib[variant == "E" & model == "primary", anchor_fraction]
  if (length(phi) == 1 && !is.na(phi))
    cat(sprintf("\nAnchor-attributable fraction (primary model): %.1f%% of Variant B's above-chance AUROC is reproduced by the treatment anchor alone.\n",
                100 * phi))
}
```

## Label Agreement Between Definitions

How far the alternative labels move the onsets, measured on the same stays:
the share of stays each pair labels identically (Cohen's $\kappa$ for chance
correction) and, among stays positive under both, the median shift in onset
hour. A label that disagrees with Sepsis-3 about *which* patients deteriorate
is a different construct, not a re-reading of the same one.

```{r label-agreement}
# Onsets come from the person-hour tables where those exist, and otherwise from
# the cohort tables, which carry the same stay-level onset_hour for every
# variant except D (whose onsets are only defined on the hourly panel).
load_onsets <- function(vid) {
  p <- file.path(DATA_DIR, sprintf("03_onsets_%s.parquet", vid))
  if (!file.exists(p)) p <- file.path(DATA_DIR, sprintf("02_cohort_%s.parquet", vid))
  if (!file.exists(p)) return(NULL)
  d <- unique(as.data.table(
    arrow::read_parquet(p, col_select = c("stay_id", "onset_hour"))))
  # Onsets beyond the modelled horizon never become events, so they count as
  # negative here; this keeps D, whose onsets exist only inside the panel,
  # comparable with the variants labelled on the full stay.
  d[!is.na(onset_hour) & onset_hour > MAX_PANEL_HOURS, onset_hour := NA_integer_]
  setnames(d, "onset_hour", paste0("onset_", vid))
  d
}

cohen_kappa <- function(a, b) {
  po <- mean(a == b)
  pe <- mean(a) * mean(b) + mean(!a) * mean(!b)
  if (pe >= 1) return(NA_real_)
  (po - pe) / (1 - pe)
}

agreement <- NULL
ons <- Filter(Negate(is.null), lapply(ALL_VARIANTS, load_onsets))
if (length(ons) < 2 || is.null(load_onsets("B"))) {
  cat("Per-variant onset tables not available; re-run 03_person_hours.Rmd.\n")
} else {
  wide <- Reduce(function(x, y) merge(x, y, by = "stay_id"), ons)
  others <- setdiff(ALL_VARIANTS, "B")
  others <- others[paste0("onset_", others) %in% names(wide)]

  agreement <- rbindlist(lapply(others, function(vid) {
    oc <- paste0("onset_", vid)
    pos_B <- !is.na(wide$onset_B); pos_v <- !is.na(wide[[oc]])
    both  <- pos_B & pos_v
    data.table(
      variant          = vid,
      label            = variant_name(vid),
      label_family     = variant_type(vid),
      n_stays          = nrow(wide),
      n_positive       = sum(pos_v),
      n_positive_B     = sum(pos_B),
      n_positive_both  = sum(both),
      pct_agree        = round(100 * mean(pos_B == pos_v), 1),
      kappa_vs_B       = round(cohen_kappa(pos_B, pos_v), 3),
      # of the stays Sepsis-3 (B) calls septic, how many does this label find
      sensitivity_to_B = round(sum(both) / max(sum(pos_B), 1), 3),
      median_shift_h   = if (sum(both) > 0)
                           round(median(wide[[oc]][both] - wide$onset_B[both]), 1)
                         else NA_real_
    )
  }))

  knitr::kable(agreement, caption = "Onset agreement with Variant B across all label variants")
  save_parquet(agreement, "09_label_agreement", DATA_DIR)
  save_csv(agreement,     "09_label_agreement", DATA_DIR)
}
```

```{r anchor-attribution-plot, fig.cap="Discrimination under each limb of the Sepsis-3 conjunction. Variant B requires suspected infection and organ dysfunction; E keeps only the treatment anchor; D keeps only the physiological deterioration and contains no treatment timestamp.", fig.width=7, fig.height=4}
if (!is.null(attrib) && nrow(attrib) > 0) {
  tryCatch({
    plot_dt <- rbind(
      metrics[eval_window == "unrestricted" & variant == "B",
              .(variant, model, auroc)],
      attrib[, .(variant, model, auroc)]
    )
    plot_dt[, label := factor(
      variant,
      levels = c("E", "B", "D"),
      labels = c("E: anchor only\n(treatment)", "B: Sepsis-3\n(anchor AND dysfunction)",
                 "D: deterioration only\n(treatment-independent)"))]
    plot_dt <- plot_dt[!is.na(label) & !is.na(auroc)]

    p_anchor <- ggplot(plot_dt, aes(x = label, y = auroc, fill = model)) +
      geom_col(position = position_dodge(width = 0.8), width = 0.75) +
      geom_hline(yintercept = 0.5, linetype = "dashed") +
      labs(title    = "Splitting the Sepsis-3 conjunction into its two limbs",
           subtitle = "Temporal test set, unrestricted window; dashed line = chance",
           x = NULL, y = "AUROC", fill = "Model") +
      theme_bw() + theme(axis.text.x = element_text(size = 8))
    print(p_anchor)
    ggsave(file.path(FIGURES_DIR, "09_label_anchor_attribution.png"), p_anchor,
           width = 7, height = 4, dpi = 150)
  }, error = function(e) cat("Anchor attribution plot failed:", e$message, "\n"))
}
```

## Reading the Attribution

Three things follow, and only the third is empirical rather than definitional.

1. **Variant E is not a sepsis model.** Its label is the moment a clinician
   sent a culture and started antibiotics. Whatever discrimination it achieves
   is discrimination about treatment behaviour, and $\phi_{\text{anchor}}$
   states how much of Variant B's apparent skill that behaviour already
   accounts for.
2. **Variant D is not directly comparable to B on AUROC.** Its label is a
   threshold function of recorded physiology, and five of the six SOFA
   components (all but PaO$_2$/FiO$_2$) are model inputs, so a high AUROC would
   be partly self-fulfilling. The comparison that carries information is not
   the level but whether the label changes *which* patients are flagged, which
   the agreement table answers.
3. **The label-sensitivity finding survives only to the extent that D and B
   disagree.** If a treatment-independent label picks out substantially
   different patients at substantially different hours, then label choice is a
   real constraint on what a sepsis model can be trained to do, not an artefact
   of how one definition is read.

```{r attribution-summary}
if (!is.null(attrib)) {
  e_prim <- attrib[variant == "E" & model == "primary"]
  d_prim <- attrib[variant == "D" & model == "primary"]
  b_prim <- metrics[variant == "B" & model == "primary" & eval_window == "unrestricted"]
  if (nrow(e_prim) == 1 && nrow(b_prim) == 1)
    cat(sprintf("Primary model AUROC — B (Sepsis-3): %.3f | E (anchor only): %.3f | D (deterioration): %s\n",
                b_prim$auroc, e_prim$auroc,
                if (nrow(d_prim) == 1) sprintf("%.3f", d_prim$auroc) else "n/a"))
}
if (!is.null(agreement)) {
  d_row <- agreement[variant == "D"]
  if (nrow(d_row) == 1)
    cat(sprintf("Variant D vs B: kappa = %.3f, %.1f%% of B's septic stays also deteriorate, median onset shift %s h\n",
                d_row$kappa_vs_B, 100 * d_row$sensitivity_to_B,
                format(d_row$median_shift_h)))
}
```
