---
title: "10 Metrics Suite - Full Evaluation Metric Suite"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)   # student ID for reproducibility

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook applies a comprehensive metric suite following Wang et al. (2025) and Van Calster et al. (2019).

**Primary:** PhysioNet Utility Score (Reyna et al., 2020), calibration (intercept + slope).

**Secondary:** AUPRC, AUROC, decision-curve net benefit, alert burden ($\leq 1.4$ benchmark; Moor et al., 2023), median lead time.

All metrics are computed in two windows: **unrestricted** (all test hours) and **treatment-anchored** (pre-treatment hours only; Kamran et al., 2024).

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(PRROC)
  library(dcurves)
  library(CalibrationCurves)
  library(dplyr)
  library(ggplot2)
})

source("config.R")
source("utils.R")

step("=== Full Evaluation Metric Suite ===")
```

## Metric Functions

Standalone metric functions, reused across internal, external, and subgroup evaluations.

```{r metric-fns}
# PhysioNet Utility Score
compute_utility_score <- function(ph_pred, pred_col, threshold = 0.5) {
  dt <- copy(as.data.table(ph_pred))
  dt[, .alert := as.integer(get(pred_col) >= threshold)]

  stay_results <- dt[, {
    onset_h <- onset_hour[!is.na(onset_hour)]
    onset_h <- if (length(onset_h) > 0) onset_h[1] else NA_real_
    alert_h <- hour[.alert == 1L]

    if (!is.na(onset_h)) {
      tp_early <- alert_h[alert_h >= (onset_h + UTILITY_TP_MIN) & alert_h <= onset_h]
      tp_late  <- alert_h[alert_h > onset_h & alert_h <= (onset_h + UTILITY_TP_MAX)]
      fp_pre   <- alert_h[alert_h < (onset_h + UTILITY_TP_MIN)]
      if (length(tp_early) > 0) {
        lead <- onset_h - min(tp_early)
        u    <- max(0, (PREDICTION_HORIZON_H - lead) / PREDICTION_HORIZON_H)
      } else if (length(tp_late) > 0) {
        u <- -0.05 * length(tp_late)
      } else {
        u <- -1.0
      }
      u     <- u - 0.05 * length(fp_pre)
      u_max <- 1.0
    } else {
      u     <- -0.05 * length(alert_h)
      u_max <- 0.0
    }
    .(u_stay = u, u_max_stay = u_max)
  }, by = stay_id]

  U_total <- sum(stay_results$u_stay)
  U_max   <- sum(stay_results$u_max_stay)
  list(utility_total      = U_total,
       utility_max        = U_max,
       utility_normalised = if (U_max > 0) U_total / U_max else NA_real_)
}

# Calibration
compute_calibration <- function(y_true, pred_prob) {
  df <- data.frame(y = y_true, p = pred_prob)
  df <- df[!is.na(df$p) & !is.na(df$y), ]
  if (nrow(df) < 10) return(list(intercept = NA, slope = NA))

  calib_model <- tryCatch(
    glm(y ~ offset(qlogis(pmax(pmin(p, 0.999), 0.001))),
        data = df, family = binomial),
    error = function(e) NULL
  )
  intercept <- if (!is.null(calib_model)) coef(calib_model)[1] else NA_real_
  calib_slope <- tryCatch({
    mod <- glm(y ~ qlogis(pmax(pmin(p, 0.999), 0.001)), data = df, family = binomial)
    coef(mod)[2]
  }, error = function(e) NA_real_)

  list(intercept = intercept, slope = calib_slope)
}

# AUROC + AUPRC
compute_roc_pr <- function(y_true, pred_prob) {
  if (sum(y_true, na.rm = TRUE) == 0 || sum(!y_true, na.rm = TRUE) == 0)
    return(list(auroc = NA, auprc = NA))
  valid <- !is.na(pred_prob) & !is.na(y_true)
  if (sum(valid) < 10) return(list(auroc = NA, auprc = NA))
  roc_obj <- tryCatch(
    PRROC::roc.curve(scores.class0 = pred_prob[valid], weights.class0 = y_true[valid]),
    error = function(e) NULL
  )
  pr_obj <- tryCatch(
    PRROC::pr.curve(scores.class0 = pred_prob[valid], weights.class0 = y_true[valid]),
    error = function(e) NULL
  )
  list(auroc = if (!is.null(roc_obj)) roc_obj$auc else NA_real_,
       auprc = if (!is.null(pr_obj))  pr_obj$auc.integral else NA_real_)
}

# Alert burden
compute_alert_burden <- function(y_true, alerts) {
  tp <- sum(alerts == 1 & y_true == 1, na.rm = TRUE)
  fp <- sum(alerts == 1 & y_true == 0, na.rm = TRUE)
  if (tp == 0) return(Inf)
  fp / tp
}

# Lead time to clinical recognition
compute_lead_time <- function(ph_pred, pred_col, threshold = 0.5) {
  sepsis_stays <- ph_pred[!is.na(onset_hour)]
  if (nrow(sepsis_stays) == 0) return(NA_real_)
  lead_times <- sepsis_stays[, {
    rows_ord    <- .SD[order(hour)]
    first_alert <- rows_ord[get(pred_col) >= threshold, min(hour)]
    if (is.infinite(first_alert) || is.na(first_alert)) NA_real_
    else as.numeric(onset_hour[1]) - as.numeric(first_alert)
  }, by = stay_id]
  median(lead_times$V1, na.rm = TRUE)
}
```

## Evaluation Loop

The full models x variants x windows grid is evaluated. Decision-curve analysis runs where both primary and GBT predictions are available.

```{r evaluate-all}
MODEL_NAMES <- c(primary = "pred_sepsis", news2 = "pred_news2",
                  cox = "pred_cox", gbt = "pred_gbt",
                  qsofa = "pred_qsofa", sirs = "pred_sirs")

all_results <- list()

for (vid in names(LABEL_VARIANTS)) {
  step(sprintf("--- Evaluating Variant %s (%s) ---", vid, LABEL_VARIANTS[[vid]]$name))

  prim_pred_path <- file.path(OUTPUT_DIR, paste0("08_predictions_primary_",     vid, ".parquet"))
  comp_pred_path <- file.path(OUTPUT_DIR, paste0("09_predictions_comparators_", vid, ".parquet"))

  if (!file.exists(prim_pred_path)) {
    v2_log(sprintf("  Skipping Variant %s: primary predictions not found", vid))
    next
  }

  ph_prim <- as.data.table(arrow::read_parquet(prim_pred_path))
  ph_comp <- if (file.exists(comp_pred_path))
               as.data.table(arrow::read_parquet(comp_pred_path)) else NULL

  ph_all <- ph_prim
  if (!is.null(ph_comp)) {
    merge_cols <- c("stay_id", "hour", setdiff(names(ph_comp), names(ph_prim)))
    ph_all <- merge(ph_all, ph_comp[, ..merge_cols], by = c("stay_id", "hour"), all.x = TRUE)
  }

  eval_windows <- list(
    unrestricted = ph_all,
    treatment_anchored = {
      if ("first_abx_hour" %in% names(ph_all)) {
        anchor_h <- if ("first_culture_hour" %in% names(ph_all))
                      pmin(ph_all$first_abx_hour, ph_all$first_culture_hour, na.rm = TRUE)
                    else ph_all$first_abx_hour
        ph_all[is.na(anchor_h) | hour < anchor_h]
      } else ph_all
    }
  )

  for (win_name in names(eval_windows)) {
    data_win <- eval_windows[[win_name]]
    if (nrow(data_win) == 0) next

    for (mod_name in names(MODEL_NAMES)) {
      pred_col <- MODEL_NAMES[[mod_name]]
      if (!pred_col %in% names(data_win)) next
      valid <- !is.na(data_win[[pred_col]])
      if (sum(valid) < 100) next

      y    <- as.integer(data_win$event == 1)
      pred <- as.numeric(data_win[[pred_col]])

      roc_pr  <- compute_roc_pr(y[valid], pred[valid])
      calib   <- compute_calibration(y[valid], pred[valid])
      alerts  <- as.integer(pred >= 0.5)
      burden  <- compute_alert_burden(y[valid], alerts[valid])
      utility <- tryCatch(
        compute_utility_score(data_win[valid], pred_col, threshold = 0.5),
        error = function(e) list(utility_normalised = NA, utility_total = NA)
      )
      lead_t  <- tryCatch(
        compute_lead_time(data_win[valid], pred_col),
        error = function(e) NA_real_
      )

      result_row <- data.frame(
        variant            = vid,
        variant_name       = LABEL_VARIANTS[[vid]]$name,
        model              = mod_name,
        eval_window        = win_name,
        split_type         = "temporal",
        n_rows             = sum(valid),
        n_sepsis_events    = sum(y[valid]),
        auroc              = round(roc_pr$auroc, 4),
        auprc              = round(roc_pr$auprc, 4),
        calib_intercept    = round(calib$intercept, 4),
        calib_slope        = round(calib$slope, 4),
        utility_normalised = round(utility$utility_normalised, 4),
        alert_burden       = round(burden, 2),
        lead_time_h        = round(lead_t, 2)
      )
      all_results[[paste(vid, mod_name, win_name, sep = "-")]] <- result_row
    }
  }

  # Decision-curve analysis
  ph_test_for_dca <- ph_all[!is.na(pred_sepsis)]
  if (nrow(ph_test_for_dca) > 0 && "pred_gbt" %in% names(ph_test_for_dca)) {
    ph_test_for_dca[, outcome := as.integer(event == 1)]
    dca_result <- tryCatch(
      dcurves::dca(outcome ~ pred_sepsis + pred_gbt,
                   data       = as.data.frame(ph_test_for_dca),
                   thresholds = seq(0.01, 0.20, 0.01)),
      error = function(e) NULL
    )
    if (!is.null(dca_result)) {
      dca_df <- tryCatch(dca_result$dca, error = function(e) as.data.frame(dca_result))
      if (!is.null(dca_df) && nrow(dca_df) > 0) {
        dca_df$variant <- vid
        save_parquet(as.data.frame(dca_df), paste0("10_dca_", vid), OUTPUT_DIR)
      }
    }
  }
}
```

## Results and Interpretation

Central performance comparison. Key checks: (i) primary vs. GBT Utility Score (H6), (ii) treatment-anchored vs. unrestricted (H3, leakage), (iii) calibration slope near 1.0.

```{r show-results}
if (length(all_results) > 0) {
  results_df <- do.call(rbind, all_results)
  save_parquet(results_df, "10_metric_results", OUTPUT_DIR)

  display_cols <- c("variant", "model", "eval_window", "auroc", "auprc",
                    "utility_normalised", "alert_burden", "lead_time_h")
  knitr::kable(
    results_df[order(results_df$variant, results_df$eval_window, results_df$model),
               display_cols],
    caption = "Evaluation results: all models x variants x windows (temporal split)"
  )
  cat("\nSaved: 10_metric_results.parquet\n")
} else {
  cat("No results to display. Check that 08 and 09 have been run.\n")
}
```

## Random-Split Evaluation (H4: Split Optimism)

The same metric suite is applied to random-split predictions (Variant B only). The output `10_metric_results_random.parquet` feeds the variance decomposition for H4 (Guo et al., 2022).

```{r evaluate-random-split}
step("--- Random-split evaluation for H4 ---")

random_results <- list()
vid <- "B"

prim_random_path <- file.path(OUTPUT_DIR, paste0("08_predictions_primary_random_", vid, ".parquet"))
comp_random_path <- file.path(OUTPUT_DIR, paste0("09_predictions_comparators_random_", vid, ".parquet"))

if (file.exists(prim_random_path)) {
  ph_prim_r <- as.data.table(arrow::read_parquet(prim_random_path))
  ph_comp_r <- if (file.exists(comp_random_path))
                 as.data.table(arrow::read_parquet(comp_random_path)) else NULL

  ph_all_r <- ph_prim_r
  if (!is.null(ph_comp_r)) {
    merge_cols_r <- c("stay_id", "hour", setdiff(names(ph_comp_r), names(ph_prim_r)))
    ph_all_r <- merge(ph_all_r, ph_comp_r[, ..merge_cols_r],
                      by = c("stay_id", "hour"), all.x = TRUE)
  }

  eval_windows_r <- list(
    unrestricted = ph_all_r,
    treatment_anchored = {
      if ("first_abx_hour" %in% names(ph_all_r)) {
        anchor_h_r <- if ("first_culture_hour" %in% names(ph_all_r))
                        pmin(ph_all_r$first_abx_hour, ph_all_r$first_culture_hour, na.rm = TRUE)
                      else ph_all_r$first_abx_hour
        ph_all_r[is.na(anchor_h_r) | hour < anchor_h_r]
      } else ph_all_r
    }
  )

  for (win_name in names(eval_windows_r)) {
    data_win_r <- eval_windows_r[[win_name]]
    if (nrow(data_win_r) == 0) next

    for (mod_name in names(MODEL_NAMES)) {
      pred_col <- MODEL_NAMES[[mod_name]]
      if (!pred_col %in% names(data_win_r)) next
      valid <- !is.na(data_win_r[[pred_col]])
      if (sum(valid) < 100) next

      y    <- as.integer(data_win_r$event == 1)
      pred <- as.numeric(data_win_r[[pred_col]])

      roc_pr  <- compute_roc_pr(y[valid], pred[valid])
      calib   <- compute_calibration(y[valid], pred[valid])
      alerts  <- as.integer(pred >= 0.5)
      burden  <- compute_alert_burden(y[valid], alerts[valid])
      utility <- tryCatch(
        compute_utility_score(data_win_r[valid], pred_col, threshold = 0.5),
        error = function(e) list(utility_normalised = NA, utility_total = NA)
      )
      lead_t  <- tryCatch(
        compute_lead_time(data_win_r[valid], pred_col),
        error = function(e) NA_real_
      )

      result_row <- data.frame(
        variant            = vid,
        variant_name       = LABEL_VARIANTS[[vid]]$name,
        model              = mod_name,
        eval_window        = win_name,
        split_type         = "random",
        n_rows             = sum(valid),
        n_sepsis_events    = sum(y[valid]),
        auroc              = round(roc_pr$auroc, 4),
        auprc              = round(roc_pr$auprc, 4),
        calib_intercept    = round(calib$intercept, 4),
        calib_slope        = round(calib$slope, 4),
        utility_normalised = round(utility$utility_normalised, 4),
        alert_burden       = round(burden, 2),
        lead_time_h        = round(lead_t, 2)
      )
      random_results[[paste(vid, mod_name, win_name, "random", sep = "-")]] <- result_row
    }
  }

  if (length(random_results) > 0) {
    random_results_df <- do.call(rbind, random_results)
    save_parquet(random_results_df, "10_metric_results_random", OUTPUT_DIR)

    knitr::kable(
      random_results_df[order(random_results_df$eval_window, random_results_df$model),
                        c("variant", "model", "eval_window", "auroc", "auprc",
                          "utility_normalised", "alert_burden", "lead_time_h")],
      caption = "Evaluation results: random split (Variant B only, H4)"
    )
    cat("\nSaved: 10_metric_results_random.parquet\n")
  }
} else {
  cat("Random-split primary predictions not found; skipping H4 evaluation.\n")
  cat("Run 08_primary_model.Rmd first to generate random-split predictions.\n")
}
```
