---
title: "11 Clinical Utility Metrics - Calibration Curves, Net Benefit, Operating Points, Event-Level Detection"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "85%", fig.width = 7, fig.height = 4.2
)
```

## Introduction

Notebook 07 reports discrimination (AUROC, AUPRC), the PhysioNet Utility
Score, calibration intercept and slope, and a ratio-based alert burden. That
suite leaves six clinically relevant questions unanswered, and this notebook
answers them:

1. **Calibration curves.** Intercept and slope are two numbers summarising a
   whole curve. The flexible calibration curve is plotted directly, and
   summarised with the Integrated Calibration Index (ICI), E50, E90 and the
   scaled Brier score (Austin & Steyerberg, 2019; Van Calster et al., 2019).
2. **Decision-curve / net-benefit analysis.** Notebook 07 evaluated net
   benefit on a threshold grid of 0.01--0.20. Event prevalence in the
   person-hour table is roughly 0.2 per cent, so every threshold on that grid
   is an order of magnitude above the prevalence and the whole curve is
   trivially below "treat none". Net benefit is recomputed here on a
   prevalence-appropriate grid (Vickers & Elkin, 2006; Vickers et al., 2019).
3. **PPV at a fixed sensitivity.** Alerting systems are commissioned at an
   operating point, not at a default 0.5 cut-off. Thresholds are chosen to
   hit sensitivity 0.50 / 0.80 / 0.90 and PPV, specificity and number needed
   to evaluate are reported there.
4. **False alarms per patient-day.** The ratio-based burden (false alerts per
   true alert) is not time-normalised, so it cannot be compared against
   ward-level alarm-fatigue thresholds. Both alert-hours and distinct
   alarm *episodes* per patient-day are computed here.
5. **Per-patient (event-level) detection.** All of notebook 07's
   discrimination is per person-hour. Episode-level sensitivity ("what
   fraction of septic patients were alerted before onset?"), per-patient PPV
   and the fraction of non-septic stays that received at least one alert are
   reported alongside the per-hour figures.
6. **Lead time.** Reported as a full distribution (median, IQR, proportion
   detected at least 3 h ahead), not a single median buried in a wide table.

Everything is computed on the temporal test set (2017--2019) using the
predictions already written by notebooks 05, 06 and 08. No model is refitted.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(arrow)
  library(splines)
  library(ggplot2)
  library(dplyr)
})

source("config.R")
source("utils.R")

step("=== Clinical Utility Metric Suite (11) ===")

# Sensitivity targets at which operating points are defined
SENS_TARGETS <- c(0.50, 0.80, 0.90)

# Headline operating point reported in the thesis
HEADLINE_SENS <- 0.80

# Models carried through this notebook
MODEL_NAMES <- c(primary = "pred_sepsis", gbt = "pred_gbt",
                  news2 = "pred_news2", cox = "pred_cox",
                  qsofa = "pred_qsofa", sirs = "pred_sirs")

# Models shown in the plots (the primary model and its two headline comparators)
PLOT_MODELS <- c("primary", "gbt", "news2")

MODEL_LABELS <- c(primary = "Primary (pooled logistic)", gbt = "GBT",
                   news2 = "NEWS2", cox = "Cox", qsofa = "qSOFA", sirs = "SIRS")
```

## Metric Functions

```{r metric-fns}
# --------------------------------------------------------------------------- #
# Small helpers
# --------------------------------------------------------------------------- #
clamp_p <- function(p, eps = 1e-6) pmin(pmax(p, eps), 1 - eps)

#' Count, for each threshold, how many predictions are >= that threshold,
#' and how many of those are events. O(n log n) once, then O(k log n).
#' Returns a data.table with threshold, n_alert, tp, fp.
counts_at_thresholds <- function(y, p, thresholds) {
  ord    <- order(p, decreasing = TRUE)
  p_desc <- p[ord]
  y_desc <- y[ord]
  cum_tp <- cumsum(y_desc)
  n      <- length(p_desc)
  p_asc  <- rev(p_desc)

  # number of predictions strictly below t  ->  n_alert = n - that
  n_below <- findInterval(thresholds, p_asc, left.open = TRUE)
  k       <- n - n_below
  tp      <- ifelse(k > 0, cum_tp[pmax(k, 1)], 0)

  data.table(threshold = thresholds,
             n_alert   = k,
             tp        = tp,
             fp        = k - tp)
}

# --------------------------------------------------------------------------- #
# 1. Calibration: flexible curve + ICI / E50 / E90 / Brier
#    Austin & Steyerberg (2019). The calibration curve is a restricted cubic
#    spline of the outcome on the logit of the predicted risk.
# --------------------------------------------------------------------------- #
calibration_curve <- function(y, p, n_bins = 20L, df_spline = 3L) {
  keep <- !is.na(y) & !is.na(p)
  y <- as.integer(y[keep]); p <- as.numeric(p[keep])
  if (length(y) < 100 || sum(y) < 10 || length(unique(p)) < 20)
    return(list(summary = NULL, curve = NULL, bins = NULL))

  pc <- clamp_p(p)
  lp <- qlogis(pc)

  fit <- tryCatch(
    glm(y ~ ns(lp, df = df_spline), family = binomial),
    error = function(e) NULL
  )
  if (is.null(fit)) return(list(summary = NULL, curve = NULL, bins = NULL))

  obs <- as.numeric(fitted(fit))
  d   <- abs(obs - pc)

  # Intercept / slope (recomputed here so the curve and the summary agree)
  int_fit   <- tryCatch(coef(glm(y ~ offset(lp), family = binomial))[1],
                        error = function(e) NA_real_)
  slope_fit <- tryCatch(coef(glm(y ~ lp, family = binomial))[2],
                        error = function(e) NA_real_)

  brier      <- mean((pc - y)^2)
  prev       <- mean(y)
  brier_null <- prev * (1 - prev)

  summ <- data.table(
    n            = length(y),
    n_events     = sum(y),
    prevalence   = prev,
    calib_int    = as.numeric(int_fit),
    calib_slope  = as.numeric(slope_fit),
    ici          = mean(d),
    e50          = as.numeric(stats::quantile(d, 0.50)),
    e90          = as.numeric(stats::quantile(d, 0.90)),
    emax         = max(d),
    brier        = brier,
    brier_scaled = 1 - brier / brier_null
  )

  # Smooth curve on a grid over the observed prediction range
  grid_lp <- seq(quantile(lp, 0.001), quantile(lp, 0.999), length.out = 200)
  grid_p  <- plogis(grid_lp)
  pr      <- predict(fit, newdata = data.frame(lp = grid_lp),
                     type = "link", se.fit = TRUE)
  curve <- data.table(
    pred     = grid_p,
    obs      = plogis(pr$fit),
    obs_low  = plogis(pr$fit - 1.96 * pr$se.fit),
    obs_high = plogis(pr$fit + 1.96 * pr$se.fit)
  )

  # Decile (default: vigintile) bins for the overlaid points
  brk <- unique(stats::quantile(pc, probs = seq(0, 1, length.out = n_bins + 1)))
  if (length(brk) > 2) {
    bin <- cut(pc, breaks = brk, include.lowest = TRUE, labels = FALSE)
    bins <- data.table(bin = bin, y = y, p = pc)[
      !is.na(bin), .(pred = mean(p), obs = mean(y), n = .N), by = bin]
    bins[, `:=`(obs_low  = pmax(0, obs - 1.96 * sqrt(obs * (1 - obs) / n)),
                obs_high = pmin(1, obs + 1.96 * sqrt(obs * (1 - obs) / n)))]
  } else bins <- NULL

  list(summary = summ, curve = curve, bins = bins)
}

# --------------------------------------------------------------------------- #
# 2. Net benefit (Vickers & Elkin, 2006)
#    NB(t) = TP/N - (FP/N) * t/(1-t);  NB_all(t) = prev - (1-prev)*t/(1-t)
# --------------------------------------------------------------------------- #
net_benefit_curve <- function(y, p, thresholds) {
  keep <- !is.na(y) & !is.na(p)
  y <- as.integer(y[keep]); p <- as.numeric(p[keep])
  n <- length(y)
  if (n < 100 || sum(y) == 0) return(NULL)

  cnt  <- counts_at_thresholds(y, p, thresholds)
  w    <- thresholds / (1 - thresholds)
  prev <- mean(y)

  data.table(
    threshold      = thresholds,
    net_benefit    = cnt$tp / n - (cnt$fp / n) * w,
    nb_treat_all   = prev - (1 - prev) * w,
    nb_treat_none  = 0,
    alert_rate     = cnt$n_alert / n,
    prevalence     = prev
  )
}

# --------------------------------------------------------------------------- #
# 3. Operating point at a fixed sensitivity
#    Threshold = the largest cut-off achieving at least the target sensitivity.
# --------------------------------------------------------------------------- #
operating_point <- function(y, p, target_sens, hours_per_row = 1) {
  keep <- !is.na(y) & !is.na(p)
  y <- as.integer(y[keep]); p <- as.numeric(p[keep])
  n  <- length(y); n_pos <- sum(y); n_neg <- n - n_pos
  if (n_pos < 5 || n_neg < 5) return(NULL)

  # P(p >= thr | y = 1) >= target  ->  thr = quantile of positives
  thr <- as.numeric(stats::quantile(p[y == 1], probs = 1 - target_sens,
                                    type = 1, names = FALSE))

  alert <- p >= thr
  tp <- sum(alert & y == 1); fp <- sum(alert & y == 0)
  fn <- n_pos - tp;          tn <- n_neg - fp

  sens <- tp / n_pos
  spec <- tn / n_neg
  ppv  <- if ((tp + fp) > 0) tp / (tp + fp) else NA_real_
  npv  <- if ((tn + fn) > 0) tn / (tn + fn) else NA_real_

  patient_days <- n * hours_per_row / 24

  data.table(
    target_sens        = target_sens,
    threshold          = thr,
    n_hours            = n,
    n_event_hours      = n_pos,
    patient_days       = patient_days,
    sensitivity        = sens,
    specificity        = spec,
    ppv                = ppv,
    npv                = npv,
    nne                = if (!is.na(ppv) && ppv > 0) 1 / ppv else NA_real_,
    alert_hours        = tp + fp,
    alert_rate         = (tp + fp) / n,
    false_alert_hours  = fp,
    fa_per_patient_day = fp / patient_days,
    alerts_per_pt_day  = (tp + fp) / patient_days,
    alert_burden_ratio = if (tp > 0) fp / tp else Inf
  )
}

# --------------------------------------------------------------------------- #
# 4. Per-patient (event-level) detection at a given threshold
#
#    A stay is a sepsis stay if it contributes at least one event hour.
#    Detection windows follow the Utility Score convention (Reyna et al., 2020):
#      - actionable window : [onset - 6h, onset]
#      - any pre-onset     : first alert at or before onset
#    A contiguous run of alert hours counts as ONE alarm episode; an episode
#    is a true alarm if it overlaps [onset - 6h, onset + 3h].
# --------------------------------------------------------------------------- #
patient_level_metrics <- function(dt, pred_col, thr) {
  d <- dt[!is.na(get(pred_col)), .(stay_id, hour,
                                    ev    = as.integer(event == 1),
                                    onset = as.numeric(onset_hour),
                                    alert = as.integer(get(pred_col) >= thr))]
  if (nrow(d) == 0) return(NULL)
  setorder(d, stay_id, hour)

  # alarm episodes: contiguous runs of alert hours within a stay
  d[, run := rleid(stay_id, alert)]
  ep <- d[alert == 1L, .(stay_id = stay_id[1],
                          h_start = min(hour), h_end = max(hour),
                          onset   = onset[1]), by = run]
  ep[, true_alarm := !is.na(onset) &
       h_end >= (onset + UTILITY_TP_MIN) & h_start <= (onset + UTILITY_TP_MAX)]

  stay <- d[, .(
    is_sepsis = as.integer(any(ev == 1L)),
    onset     = as.numeric(onset[1]),
    n_hours   = .N,
    any_alert = as.integer(any(alert == 1L))
  ), by = stay_id]

  # first alert inside the actionable window [onset - 6 h, onset], and
  # first alert at any point at or before onset
  win <- d[alert == 1L & !is.na(onset) & hour >= (onset + UTILITY_TP_MIN) &
             hour <= onset,
           .(first_win_alert = as.numeric(min(hour))), by = stay_id]
  pre <- d[alert == 1L & !is.na(onset) & hour <= onset,
           .(first_pre_alert = as.numeric(min(hour))), by = stay_id]

  stay <- merge(stay, win, by = "stay_id", all.x = TRUE)
  stay <- merge(stay, pre, by = "stay_id", all.x = TRUE)
  stay[, detected_window := as.integer(is_sepsis == 1L & !is.na(first_win_alert))]
  stay[, detected_pre    := as.integer(is_sepsis == 1L & !is.na(first_pre_alert))]
  stay[, lead_time_h     := ifelse(detected_pre == 1L, onset - first_pre_alert,
                                   NA_real_)]

  n_sepsis    <- stay[, sum(is_sepsis == 1L)]
  n_nonsepsis <- stay[, sum(is_sepsis == 0L)]
  n_alerted   <- stay[, sum(any_alert == 1L)]
  patient_days <- stay[, sum(n_hours)] / 24

  fa_episodes <- ep[true_alarm == FALSE, .N]
  tp_episodes <- ep[true_alarm == TRUE,  .N]

  lt <- stay[detected_pre == 1L & is.finite(lead_time_h), lead_time_h]

  list(
    summary = data.table(
      threshold             = thr,
      n_stays               = nrow(stay),
      n_sepsis_stays        = n_sepsis,
      n_nonsepsis_stays     = n_nonsepsis,
      patient_days          = patient_days,
      # event-level sensitivity
      episode_sens_window   = if (n_sepsis > 0)
        stay[, sum(detected_window)] / n_sepsis else NA_real_,
      episode_sens_pre      = if (n_sepsis > 0)
        stay[, sum(detected_pre)] / n_sepsis else NA_real_,
      # event-level PPV and false-alarm exposure
      stays_alerted         = n_alerted,
      frac_stays_alerted    = n_alerted / nrow(stay),
      frac_nonsepsis_alerted = if (n_nonsepsis > 0)
        stay[is_sepsis == 0L, sum(any_alert)] / n_nonsepsis else NA_real_,
      patient_ppv           = if (n_alerted > 0)
        stay[, sum(detected_pre)] / n_alerted else NA_real_,
      # alarm episodes
      n_alarm_episodes      = nrow(ep),
      n_false_alarm_episodes = fa_episodes,
      n_true_alarm_episodes  = tp_episodes,
      fa_episodes_per_pt_day = fa_episodes / patient_days,
      # lead time
      lead_median_h         = if (length(lt) > 0) median(lt) else NA_real_,
      lead_q25_h            = if (length(lt) > 0) as.numeric(quantile(lt, .25)) else NA_real_,
      lead_q75_h            = if (length(lt) > 0) as.numeric(quantile(lt, .75)) else NA_real_,
      frac_lead_ge_3h       = if (n_sepsis > 0) sum(lt >= 3) / n_sepsis else NA_real_,
      frac_lead_ge_6h       = if (n_sepsis > 0) sum(lt >= 6) / n_sepsis else NA_real_
    ),
    lead_times = lt
  )
}
```

## Data Assembly

The same merge as notebook 07: primary-model predictions joined to the
comparator predictions on `(stay_id, hour)`. Only the unrestricted evaluation
window is used — the treatment-anchored window contains zero event hours
(notebook 07, H3), so no operating point can be defined in it.

```{r assemble}
load_variant <- function(vid) {
  prim_path <- file.path(DATA_DIR, paste0("05_predictions_primary_",     vid, ".parquet"))
  comp_path <- file.path(DATA_DIR, paste0("06_predictions_comparators_", vid, ".parquet"))
  if (!file.exists(prim_path)) return(NULL)

  ph <- as.data.table(arrow::read_parquet(prim_path))
  if (file.exists(comp_path)) {
    cp <- as.data.table(arrow::read_parquet(comp_path))
    add <- c("stay_id", "hour", setdiff(names(cp), names(ph)))
    ph <- merge(ph, cp[, ..add], by = c("stay_id", "hour"), all.x = TRUE)
  }
  ph
}

variants <- names(LABEL_VARIANTS)
ph_list  <- setNames(lapply(variants, load_variant), variants)
ph_list  <- ph_list[!vapply(ph_list, is.null, logical(1))]

for (vid in names(ph_list)) {
  v2_log(sprintf("Variant %s: %s test hours, %s event hours, %s stays",
                 vid,
                 format(nrow(ph_list[[vid]]), big.mark = ","),
                 format(sum(ph_list[[vid]]$event == 1), big.mark = ","),
                 format(uniqueN(ph_list[[vid]]$stay_id), big.mark = ",")))
}
```

## Calibration Curves

Predicted risk against observed event rate. The diagonal is perfect
calibration. The spline curve is the flexible calibration curve; the points
are twenty equal-sized bins of predicted risk with binomial confidence
intervals. Axes are restricted to the observed prediction range, because at a
0.2 per cent event rate no model ever predicts a risk near 1.

```{r calibration, fig.height=4.6}
calib_summ  <- list()
calib_curve <- list()
calib_bins  <- list()

for (vid in names(ph_list)) {
  ph <- ph_list[[vid]]
  y  <- as.integer(ph$event == 1)
  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ph)) next
    res <- calibration_curve(y, as.numeric(ph[[pcol]]))
    if (is.null(res$summary)) next
    key <- paste(vid, mod, sep = "-")
    calib_summ[[key]]  <- cbind(variant = vid, model = mod, res$summary)
    if (mod %in% PLOT_MODELS) {
      calib_curve[[key]] <- cbind(variant = vid, model = mod, res$curve)
      if (!is.null(res$bins))
        calib_bins[[key]] <- cbind(variant = vid, model = mod, res$bins)
    }
  }
}

calib_summ_dt <- rbindlist(calib_summ, fill = TRUE)
save_parquet(calib_summ_dt, "11_calibration_summary", DATA_DIR)
save_csv(calib_summ_dt,     "11_calibration_summary", DATA_DIR)

curve_dt <- rbindlist(calib_curve, fill = TRUE)
bins_dt  <- rbindlist(calib_bins,  fill = TRUE)

if (nrow(curve_dt) > 0) {
  curve_dt[, model_lab := factor(MODEL_LABELS[model],
                                  levels = MODEL_LABELS[PLOT_MODELS])]
  bins_dt[,  model_lab := factor(MODEL_LABELS[model],
                                  levels = MODEL_LABELS[PLOT_MODELS])]
  curve_dt[, variant_lab := paste0("Variant ", variant)]
  bins_dt[,  variant_lab := paste0("Variant ", variant)]

  # Predicted-risk scales differ by orders of magnitude between the pooled
  # logistic model (calibrated to a 0.2% hourly event rate) and GBT, so every
  # panel needs its own scale. facet_grid(scales = "free") only frees axes per
  # row/column, so facet_wrap over the model-variant interaction is used
  # instead. The dashed diagonal is perfect calibration in every panel.
  mk_panel <- function(d) {
    d[, panel_lab := factor(
      paste0(model_lab, ", Variant ", variant),
      levels = unlist(lapply(MODEL_LABELS[PLOT_MODELS], function(m)
        paste0(m, ", Variant ", sort(unique(curve_dt$variant))))))]
    d[!is.na(panel_lab)]
  }
  curve_dt <- mk_panel(curve_dt)
  bins_dt  <- mk_panel(bins_dt)

  p_cal <- ggplot(curve_dt, aes(pred, obs, colour = model_lab, fill = model_lab)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey40", linewidth = 0.4) +
    geom_ribbon(aes(ymin = obs_low, ymax = obs_high), alpha = 0.18,
                colour = NA) +
    geom_line(linewidth = 0.7) +
    geom_point(data = bins_dt, aes(pred, obs), size = 1.0, alpha = 0.7,
               inherit.aes = TRUE) +
    facet_wrap(~ panel_lab, scales = "free", ncol = 3) +
    expand_limits(x = 0, y = 0) +
    labs(x = "Predicted risk of sepsis onset within 6 h",
         y = "Observed event rate",
         colour = NULL, fill = NULL) +
    theme_bw() +
    theme(legend.position = "none",
          panel.grid.minor = element_blank(),
          strip.text = element_text(size = 7.5),
          axis.text = element_text(size = 6.5))

  n_panels <- length(unique(curve_dt$panel_lab))
  ggsave(file.path(FIGURES_DIR, "11_calibration_curves.png"), p_cal,
         width = 9, height = 2.4 * ceiling(n_panels / 3) + 0.4, dpi = 150)
  print(p_cal)
}
```

```{r calibration-table}
if (nrow(calib_summ_dt) > 0) {
  tab <- calib_summ_dt[model %in% PLOT_MODELS,
    .(variant, model = MODEL_LABELS[model],
      Intercept = round(calib_int, 3), Slope = round(calib_slope, 3),
      ICI = signif(ici, 3), E50 = signif(e50, 3), E90 = signif(e90, 3),
      Brier = signif(brier, 4), `Brier scaled` = round(brier_scaled, 4))]
  knitr::kable(tab[order(variant, model)],
    caption = paste("Calibration summary. ICI = Integrated Calibration Index",
                    "(mean absolute difference between predicted risk and the",
                    "flexible calibration curve). E50/E90 = median and 90th",
                    "percentile of that difference. Scaled Brier = 1 - Brier /",
                    "Brier(null); values at or below 0 mean the model is no",
                    "better than predicting the prevalence for everyone."))
}
```

## Decision-Curve Analysis

Net benefit is evaluated on a grid running from one tenth of the event
prevalence to roughly ten times it. This is the range in which a
sepsis-alerting threshold could plausibly be set: a threshold probability of
$t$ encodes a willingness to accept $\left(1-t\right)/t$ false alarms per
true detection. At a 0.2 per cent event rate, a threshold of 0.01 already
encodes a willingness to tolerate only 99 false alarms per detection, which
is why the 0.01--0.20 grid used in notebook 07 produced nothing but
negative curves.

```{r dca, fig.height=4.6}
nb_list <- list()

for (vid in names(ph_list)) {
  ph   <- ph_list[[vid]]
  y    <- as.integer(ph$event == 1)
  prev <- mean(y)
  thr_grid <- unique(sort(c(
    seq(prev / 10, prev * 10, length.out = 80),
    prev
  )))
  thr_grid <- thr_grid[thr_grid > 0 & thr_grid < 0.5]

  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ph)) next
    nb <- net_benefit_curve(y, as.numeric(ph[[pcol]]), thr_grid)
    if (is.null(nb)) next
    nb_list[[paste(vid, mod, sep = "-")]] <- cbind(variant = vid, model = mod, nb)
  }
}

nb_dt <- rbindlist(nb_list, fill = TRUE)
save_parquet(nb_dt, "11_net_benefit", DATA_DIR)

if (nrow(nb_dt) > 0) {
  # Treat-all and treat-none are plotted as named series so that the reader can
  # see directly which model curves lie on top of a default strategy.
  ref <- unique(nb_dt[, .(variant, threshold, nb_treat_all)])
  series <- rbind(
    nb_dt[model %in% PLOT_MODELS,
          .(variant, threshold, nb = net_benefit, series = MODEL_LABELS[model])],
    ref[,  .(variant, threshold, nb = nb_treat_all, series = "Treat all")],
    ref[,  .(variant, threshold, nb = 0,           series = "Treat none")]
  )
  ser_levels <- c(unname(MODEL_LABELS[PLOT_MODELS]), "Treat all", "Treat none")
  series[, series := factor(series, levels = ser_levels)]
  series[, variant_lab := paste0("Variant ", variant)]

  # The maximum attainable net benefit is the prevalence, so the informative
  # band is a small multiple of it. Curves are clipped below -1.75 x prevalence
  # so that the near-zero region stays legible; a line that leaves the bottom
  # of a panel is worse than treating nobody from that threshold onwards.
  series <- merge(series, nb_dt[, .(prev = prevalence[1]), by = variant],
                  by = "variant", all.x = TRUE)
  series[, nb_clip := fifelse(nb < -1.75 * prev, NA_real_, nb)]

  p_dca <- ggplot(series, aes(threshold, nb_clip, colour = series,
                              linetype = series, linewidth = series)) +
    geom_line(na.rm = TRUE) +
    facet_wrap(~ variant_lab, nrow = 1, scales = "free") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_colour_manual(values = setNames(
      c(scales::hue_pal()(length(PLOT_MODELS)), "grey25", "grey55"), ser_levels)) +
    scale_linetype_manual(values = setNames(
      c(rep("solid", length(PLOT_MODELS)), "dotdash", "dashed"), ser_levels)) +
    scale_linewidth_manual(values = setNames(
      c(rep(0.75, length(PLOT_MODELS)), 0.5, 0.5), ser_levels)) +
    labs(x = "Threshold probability", y = "Net benefit",
         colour = NULL, linetype = NULL, linewidth = NULL) +
    theme_bw() +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  ggsave(file.path(FIGURES_DIR, "11_decision_curves.png"), p_dca,
         width = 9, height = 4, dpi = 150)
  print(p_dca)
}
```

```{r dca-table}
# Net benefit at the threshold equal to the event prevalence. At t = prevalence
# both defaults have net benefit exactly 0, so any positive value is a real
# gain over both, and the maximum attainable net benefit is the prevalence
# itself. Standardised net benefit (Vickers et al., 2019) = NB / prevalence,
# i.e. the fraction of the attainable benefit that the model captures.
if (nrow(nb_dt) > 0) {
  nb_prev <- nb_dt[, {
    i <- which.min(abs(threshold - prevalence[1]))
    .(threshold   = threshold[i],
      prevalence  = prevalence[1],
      net_benefit = net_benefit[i],
      nb_treat_all = nb_treat_all[i],
      snb         = net_benefit[i] / prevalence[1],
      alert_rate  = alert_rate[i],
      # does the alert rate respond to the threshold at all?
      alert_rate_range = max(alert_rate) - min(alert_rate))
  }, by = .(variant, model)]

  save_parquet(nb_prev, "11_net_benefit_at_prevalence", DATA_DIR)
  save_csv(nb_prev,     "11_net_benefit_at_prevalence", DATA_DIR)

  knitr::kable(
    nb_prev[order(variant, model),
      .(variant, model = MODEL_LABELS[model],
        Threshold = signif(threshold, 3),
        `Net benefit` = signif(net_benefit, 3),
        `Treat all` = signif(nb_treat_all, 2),
        `Std NB` = round(snb, 3),
        `Alert rate` = round(alert_rate, 3),
        `Alert rate span` = round(alert_rate_range, 3))],
    caption = paste("Net benefit at a threshold probability equal to the event",
                    "prevalence. Both default strategies have net benefit 0 at",
                    "this threshold, so Std NB (net benefit divided by the",
                    "prevalence) is the fraction of the attainable benefit",
                    "captured. 'Alert rate span' is the change in alert rate",
                    "across the whole threshold grid: a span of 0 means the",
                    "score cannot be tuned anywhere in the plausible range and",
                    "the decision curve collapses onto a default strategy."))
}
```

## Operating Points at Fixed Sensitivity

For each model the threshold is set to the largest cut-off achieving the
target per-hour sensitivity, and PPV, specificity, number needed to evaluate
(NNE $=1/\text{PPV}$), alert rate and time-normalised false-alarm rate are
read off at that point.

```{r operating-points}
op_list <- list()

for (vid in names(ph_list)) {
  ph <- ph_list[[vid]]
  y  <- as.integer(ph$event == 1)
  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ph)) next
    p <- as.numeric(ph[[pcol]])
    for (ts in SENS_TARGETS) {
      op <- operating_point(y, p, ts)
      if (is.null(op)) next
      op_list[[paste(vid, mod, ts, sep = "-")]] <-
        cbind(variant = vid, model = mod, op)
    }
  }
}

op_dt <- rbindlist(op_list, fill = TRUE)
save_parquet(op_dt, "11_operating_points", DATA_DIR)
save_csv(op_dt,     "11_operating_points", DATA_DIR)

if (nrow(op_dt) > 0) {
  show <- op_dt[target_sens == HEADLINE_SENS & model %in% PLOT_MODELS,
    .(variant, model = MODEL_LABELS[model],
      Threshold   = signif(threshold, 3),
      Sens        = round(sensitivity, 3),
      Spec        = round(specificity, 3),
      PPV         = signif(ppv, 3),
      NNE         = round(nne, 0),
      `Alerts/pt-day` = round(alerts_per_pt_day, 2),
      `FA/pt-day` = round(fa_per_patient_day, 2),
      `FP:TP`     = round(alert_burden_ratio, 0))]
  knitr::kable(show[order(variant, model)],
    caption = sprintf(paste("Operating characteristics at a fixed per-hour",
                            "sensitivity of %.0f%%. NNE = number of alerted",
                            "person-hours per true event hour. FA/pt-day =",
                            "false-alert hours per patient-day."),
                      HEADLINE_SENS * 100))
}
```

```{r sens-tradeoff, fig.height=3.6}
if (nrow(op_dt) > 0) {
  tr <- op_dt[model %in% PLOT_MODELS]
  tr[, model_lab := factor(MODEL_LABELS[model], levels = MODEL_LABELS[PLOT_MODELS])]
  tr[, variant_lab := paste0("Variant ", variant)]

  p_tr <- ggplot(tr, aes(sensitivity, fa_per_patient_day,
                         colour = model_lab, group = model_lab)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.6) +
    facet_wrap(~ variant_lab, nrow = 1) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(x = "Per-hour sensitivity", y = "False alerts per patient-day",
         colour = NULL) +
    theme_bw() +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  ggsave(file.path(FIGURES_DIR, "11_false_alarms_vs_sensitivity.png"), p_tr,
         width = 9, height = 3.4, dpi = 150)
  print(p_tr)
}
```

## Per-Patient (Event-Level) Detection

Every figure above is per person-hour. Under the discrete-time
competing-risks framing, `event == 1` marks the sepsis onset hour itself
(states 2 and 3 are discharge alive and death without sepsis), so each septic
stay contributes exactly one event hour. Per-hour sensitivity is therefore a
strict quantity: the alert must fire *in the onset hour itself*. Per-patient
sensitivity is the clinically meaningful counterpart: it asks only whether
the patient was alerted at all before onset. The gap between the two, and the
price paid for it in false alarms, is the point of this section.

```{r patient-level}
pat_list  <- list()
lead_list <- list()

for (vid in names(ph_list)) {
  ph <- ph_list[[vid]]
  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ph)) next
    for (ts in SENS_TARGETS) {
      thr <- op_dt[variant == vid & model == mod & target_sens == ts, threshold]
      if (length(thr) == 0) next
      res <- tryCatch(patient_level_metrics(ph, pcol, thr[1]),
                      error = function(e) { v2_log(sprintf(
                        "patient_level_metrics failed (%s/%s/%.2f): %s",
                        vid, mod, ts, e$message), level = "WARN"); NULL })
      if (is.null(res)) next
      key <- paste(vid, mod, ts, sep = "-")
      pat_list[[key]] <- cbind(variant = vid, model = mod,
                                target_sens = ts, res$summary)
      if (ts == HEADLINE_SENS && mod %in% PLOT_MODELS && length(res$lead_times) > 0)
        lead_list[[key]] <- data.table(variant = vid, model = mod,
                                        lead_time_h = res$lead_times)
    }
  }
}

pat_dt <- rbindlist(pat_list, fill = TRUE)
save_parquet(pat_dt, "11_patient_level_metrics", DATA_DIR)
save_csv(pat_dt,     "11_patient_level_metrics", DATA_DIR)

if (nrow(pat_dt) > 0) {
  show_p <- pat_dt[target_sens == HEADLINE_SENS & model %in% PLOT_MODELS,
    .(variant, model = MODEL_LABELS[model],
      `Sepsis stays`      = n_sepsis_stays,
      `Episode sens (6h)` = round(episode_sens_window, 3),
      `Episode sens (any pre)` = round(episode_sens_pre, 3),
      `Patient PPV`       = signif(patient_ppv, 3),
      `% stays alerted`   = round(100 * frac_stays_alerted, 1),
      `% non-sepsis alerted` = round(100 * frac_nonsepsis_alerted, 1),
      `FA episodes/pt-day` = round(fa_episodes_per_pt_day, 2))]
  knitr::kable(show_p[order(variant, model)],
    caption = paste("Per-patient detection at the same thresholds as the",
                    "operating-point table. Episode sens (6h) = fraction of",
                    "septic stays alerted inside the actionable window",
                    "[onset - 6 h, onset]. Episode sens (any pre) = alerted at",
                    "any point before onset. FA episodes = contiguous runs of",
                    "alert hours that do not overlap [onset - 6 h, onset + 3 h]."))
}
```

### Per-Hour versus Per-Patient, Side by Side

```{r hour-vs-patient}
if (nrow(op_dt) > 0 && nrow(pat_dt) > 0) {
  side <- merge(
    op_dt[target_sens == HEADLINE_SENS,
          .(variant, model, target_sens,
            hour_sens = sensitivity, hour_ppv = ppv,
            hour_burden = alert_burden_ratio, fa_per_patient_day)],
    pat_dt[target_sens == HEADLINE_SENS,
           .(variant, model, target_sens,
             pt_sens = episode_sens_pre, pt_ppv = patient_ppv,
             fa_episodes_per_pt_day)],
    by = c("variant", "model", "target_sens"))

  save_parquet(side, "11_hour_vs_patient", DATA_DIR)
  save_csv(side,     "11_hour_vs_patient", DATA_DIR)

  knitr::kable(
    side[model %in% PLOT_MODELS][order(variant, model),
      .(variant, model = MODEL_LABELS[model],
        `Sens (hour)` = round(hour_sens, 3),
        `Sens (patient)` = round(pt_sens, 3),
        `PPV (hour)` = signif(hour_ppv, 3),
        `PPV (patient)` = signif(pt_ppv, 3),
        `FA hours/pt-day` = round(fa_per_patient_day, 2),
        `FA episodes/pt-day` = round(fa_episodes_per_pt_day, 2))],
    caption = paste("Per-hour and per-patient views of the same operating",
                    "point. Patient-level sensitivity is much higher than",
                    "per-hour sensitivity because one alert anywhere before",
                    "onset counts the whole episode as detected; patient-level",
                    "PPV is correspondingly higher than per-hour PPV because a",
                    "multi-hour alert run counts once."))
}
```

## Lead Time to Onset

Lead time is the interval between the first alert issued at or before onset
and the onset hour itself, computed at the fixed-sensitivity threshold and
over the septic stays that were alerted at all.

```{r lead-time, fig.height=3.6}
lead_dt <- rbindlist(lead_list, fill = TRUE)

lead_summary <- pat_dt[target_sens == HEADLINE_SENS,
  .(variant, model, n_sepsis_stays,
    detected_pre = round(episode_sens_pre, 3),
    lead_median_h, lead_q25_h, lead_q75_h,
    frac_lead_ge_3h = round(frac_lead_ge_3h, 3),
    frac_lead_ge_6h = round(frac_lead_ge_6h, 3))]
save_parquet(lead_summary, "11_lead_time_summary", DATA_DIR)
save_csv(lead_summary,     "11_lead_time_summary", DATA_DIR)

if (nrow(lead_dt) > 0) {
  lead_dt[, model_lab := factor(MODEL_LABELS[model], levels = MODEL_LABELS[PLOT_MODELS])]
  lead_dt[, variant_lab := paste0("Variant ", variant)]

  p_lead <- ggplot(lead_dt[lead_time_h >= 0 & lead_time_h <= 48],
                   aes(lead_time_h, fill = model_lab)) +
    geom_histogram(binwidth = 2, position = "identity", alpha = 0.55,
                   colour = NA) +
    facet_wrap(~ variant_lab, nrow = 1, scales = "free_y") +
    labs(x = "Lead time before sepsis onset (h)",
         y = "Septic stays", fill = NULL) +
    theme_bw() +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  ggsave(file.path(FIGURES_DIR, "11_lead_time_distribution.png"), p_lead,
         width = 9, height = 3.4, dpi = 150)
  print(p_lead)
}

knitr::kable(
  lead_summary[model %in% PLOT_MODELS][order(variant, model),
    .(variant, model = MODEL_LABELS[model], n_sepsis_stays, detected_pre,
      `Median lead (h)` = lead_median_h,
      `IQR` = paste0(lead_q25_h, "-", lead_q75_h),
      `>= 3 h` = frac_lead_ge_3h, `>= 6 h` = frac_lead_ge_6h)],
  caption = sprintf(paste("Lead time at the %.0f%% sensitivity operating",
                          "point. '>= 3 h' and '>= 6 h' are fractions of ALL",
                          "septic stays (not just detected ones) alerted at",
                          "least that far ahead of onset."),
                    HEADLINE_SENS * 100))
```

## External Validation: Clinical Utility on eICU-CRD

The same operating-point and per-patient machinery applied to the eICU
predictions from notebook 08, so that the clinical-utility picture is
available externally as well as internally.

```{r external}
ext_op  <- list()
ext_pat <- list()

for (vid in variants) {
  ext_path <- file.path(DATA_DIR, paste0("08_eicu_predictions_", vid, ".parquet"))
  if (!file.exists(ext_path)) next
  ext <- as.data.table(arrow::read_parquet(ext_path))
  y   <- as.integer(ext$event == 1)
  if (sum(y) < 10) next

  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ext)) next
    op <- operating_point(y, as.numeric(ext[[pcol]]), HEADLINE_SENS)
    if (is.null(op)) next
    ext_op[[paste(vid, mod, sep = "-")]] <- cbind(variant = vid, model = mod,
                                                   cohort = "eICU", op)
    res <- tryCatch(patient_level_metrics(ext, pcol, op$threshold),
                    error = function(e) NULL)
    if (!is.null(res))
      ext_pat[[paste(vid, mod, sep = "-")]] <-
        cbind(variant = vid, model = mod, cohort = "eICU",
              target_sens = HEADLINE_SENS, res$summary)
  }
  rm(ext); gc(verbose = FALSE)
}

ext_op_dt  <- rbindlist(ext_op,  fill = TRUE)
ext_pat_dt <- rbindlist(ext_pat, fill = TRUE)

if (nrow(ext_op_dt) > 0) {
  save_parquet(ext_op_dt,  "11_external_operating_points", DATA_DIR)
  save_csv(ext_op_dt,      "11_external_operating_points", DATA_DIR)
  knitr::kable(
    ext_op_dt[model %in% PLOT_MODELS][order(variant, model),
      .(variant, model = MODEL_LABELS[model],
        Threshold = signif(threshold, 3), Sens = round(sensitivity, 3),
        PPV = signif(ppv, 4), NNE = round(nne, 0),
        `FA/pt-day` = round(fa_per_patient_day, 2))],
    caption = sprintf(paste("eICU-CRD operating characteristics at the same",
                            "%.0f%% target sensitivity (thresholds re-derived",
                            "on the external cohort)."), HEADLINE_SENS * 100))
}

if (nrow(ext_pat_dt) > 0) {
  save_parquet(ext_pat_dt, "11_external_patient_level", DATA_DIR)
  save_csv(ext_pat_dt,     "11_external_patient_level", DATA_DIR)
  knitr::kable(
    ext_pat_dt[model %in% PLOT_MODELS][order(variant, model),
      .(variant, model = MODEL_LABELS[model], n_sepsis_stays,
        `Episode sens (6h)` = round(episode_sens_window, 3),
        `Episode sens (any pre)` = round(episode_sens_pre, 3),
        `Patient PPV` = signif(patient_ppv, 3),
        `FA episodes/pt-day` = round(fa_episodes_per_pt_day, 2))],
    caption = "eICU-CRD per-patient detection at the external operating point.")
}
```

## Cross-Check Against Notebook 07

The net-benefit computation here is implemented directly from the
Vickers--Elkin formula rather than through `dcurves`. The two are compared at
the thresholds they have in common, as a correctness check.

```{r crosscheck}
chk_rows <- list()
for (vid in variants) {
  f <- file.path(DATA_DIR, paste0("07_dca_", vid, ".parquet"))
  if (!file.exists(f)) next
  old <- as.data.table(arrow::read_parquet(f))
  old <- old[variable %in% c("pred_sepsis", "pred_gbt")]
  if (nrow(old) == 0) next
  old[, model := ifelse(variable == "pred_sepsis", "primary", "gbt")]

  new <- nb_dt[variant == vid & model %in% c("primary", "gbt")]
  if (nrow(new) == 0) next

  for (m in unique(old$model)) {
    # NB: the 07 DCA table carries a column literally called `n`, so local
    # objects here are named to avoid colliding with it inside `[.data.table`.
    o_dt   <- old[model == m]
    new_dt <- new[model == m]
    thr_new <- new_dt$threshold
    thr_lo  <- min(thr_new); thr_hi <- max(thr_new)
    # nearest new threshold to each old threshold within the shared range
    o_dt <- o_dt[threshold <= thr_hi & threshold >= thr_lo]
    if (nrow(o_dt) == 0) next
    idx <- vapply(o_dt$threshold, function(t) which.min(abs(thr_new - t)),
                  integer(1))
    chk_rows[[paste(vid, m)]] <- data.table(
      variant = vid, model = m, n_compared = nrow(o_dt),
      max_abs_diff = signif(max(abs(o_dt$net_benefit -
                                      new_dt$net_benefit[idx])), 3))
  }
}

if (length(chk_rows) > 0) {
  knitr::kable(rbindlist(chk_rows),
    caption = paste("Agreement between the dcurves output stored by notebook",
                    "07 and the net benefit recomputed here, over the shared",
                    "threshold range. Differences reflect only the threshold",
                    "grid offset, not a difference in definition."))
} else {
  cat("No overlapping threshold range between the 07 grid (0.01-0.20) and the",
      "prevalence-scaled grid used here; nothing to cross-check.\n")
}
```

## Summary

```{r summary-out}
cat("Artefacts written to", DATA_DIR, ":\n")
for (f in c("11_calibration_summary", "11_net_benefit", "11_operating_points",
            "11_patient_level_metrics", "11_hour_vs_patient",
            "11_lead_time_summary", "11_external_operating_points",
            "11_external_patient_level")) {
  p <- file.path(DATA_DIR, paste0(f, ".parquet"))
  if (file.exists(p)) cat("  -", basename(p), "\n")
}
cat("\nFigures written to", FIGURES_DIR, ":\n")
for (f in c("11_calibration_curves.png", "11_decision_curves.png",
            "11_false_alarms_vs_sensitivity.png",
            "11_lead_time_distribution.png")) {
  p <- file.path(FIGURES_DIR, f)
  if (file.exists(p)) cat("  -", f, "\n")
}
```
