---
title: "12 Variance Decomposition - Attribution of Performance Variation"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)   # student ID for reproducibility

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

Central output of V2: a variance decomposition attributing performance variation to each analytical decision. Published sepsis studies vary across multiple factors simultaneously (Cohen et al., 2024); this analysis varies one factor at a time.

Five factors:

| Factor | Levels |
|--------|--------|
| Model class | Discrete-time hazard, Cox PH, GBT, NEWS2 |
| Label operationalisation | Variant A (Narrow), B (Seymour-Standard), C (Liberal) |
| Validation split | Temporal, random |
| Prediction-time anchoring | Treatment-anchored, unrestricted |
| Evaluation metric | Utility Score, calibration, AUROC, AUPRC |

Six pre-registered hypotheses structure the interpretation:

| \# | Claim | Threshold |
|----|-------|-----------|
| H1 | Label-variant spread $\geq$ model-class spread in AUROC | label spread $\geq$ model spread |
| H2 | Coefficient instability across labels | $\geq$ 1 covariate changes sign |
| H3 | Treatment-anchored AUROC $<$ unrestricted | $\Delta \geq 0.03$ |
| H4 | Temporal-split AUROC $<$ random-split | $\Delta \geq 0.02$ |
| H5 | Utility degradation $>$ AUROC degradation (internal $\to$ external) | ratio comparison |
| H6 | $|\text{primary} - \text{GBT}| \leq 0.02$ AUROC | $|\Delta| \leq 0.02$ |

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
})

source("config.R")
source("utils.R")

step("=== Phase 6a: Variance Decomposition ===")
```

## Load Metric Results

Metric results from 10, external validation from 11, and coefficient table from 08 are loaded.

```{r load-metrics}
metric_path <- file.path(OUTPUT_DIR, "10_metric_results.parquet")
ext_path    <- file.path(OUTPUT_DIR, "11_external_validation_results.parquet")
coef_path   <- file.path(OUTPUT_DIR, "08_coef_all_variants.parquet")

if (!file.exists(metric_path))
  stop("10_metric_results.parquet not found. Run 10_metrics_suite.Rmd first.")

metrics    <- as.data.table(arrow::read_parquet(metric_path))
ext_res    <- if (file.exists(ext_path))  as.data.table(arrow::read_parquet(ext_path))  else NULL
coef_table <- if (file.exists(coef_path)) as.data.table(arrow::read_parquet(coef_path)) else NULL

cat(sprintf("Metric results loaded: %d rows\n", nrow(metrics)))
knitr::kable(metrics[, .(variant, model, eval_window, auroc, auprc,
                           utility_normalised, alert_burden, lead_time_h)],
             caption = "All evaluation results (from 10_metrics_suite.Rmd)")
```

## Variance by Factor

Each factor's spread is the range of AUROC (or Utility Score) across that factor's levels, with others held at reference values (Variant B, primary model, temporal split, unrestricted).

```{r variance-by-factor}
# Factor 1: MODEL CLASS - label=B, temporal, unrestricted
model_class_rows   <- metrics[variant == "B" & eval_window == "unrestricted"]
model_spread_auroc   <- if (nrow(model_class_rows) > 0) diff(range(model_class_rows$auroc, na.rm = TRUE)) else NA
model_spread_utility <- if (nrow(model_class_rows) > 0) diff(range(model_class_rows$utility_normalised, na.rm = TRUE)) else NA

# Factor 2: LABEL VARIANT - model=primary, temporal, unrestricted
label_rows         <- metrics[model == "primary" & eval_window == "unrestricted"]
label_spread_auroc   <- if (nrow(label_rows) > 0) diff(range(label_rows$auroc, na.rm = TRUE)) else NA
label_spread_utility <- if (nrow(label_rows) > 0) diff(range(label_rows$utility_normalised, na.rm = TRUE)) else NA

# Factor 3: PREDICTION-TIME ANCHORING - model=primary, variant=B, temporal
anchor_rows_B  <- metrics[model == "primary" & variant == "B"]
anchor_unrestr <- anchor_rows_B[eval_window == "unrestricted",         auroc]
anchor_treatm  <- anchor_rows_B[eval_window == "treatment_anchored",   auroc]
h3_delta_auroc <- if (length(anchor_unrestr) > 0 && length(anchor_treatm) > 0)
                    anchor_unrestr - anchor_treatm else NA

# Factor 4: VALIDATION SPLIT (temporal vs random)
random_path <- file.path(OUTPUT_DIR, "10_metric_results_random.parquet")
if (file.exists(random_path)) {
  metrics_random <- as.data.table(arrow::read_parquet(random_path))
  temporal_auroc <- metrics[model == "primary" & variant == "B" &
                               eval_window == "unrestricted", auroc]
  random_auroc   <- metrics_random[model == "primary" & variant == "B" &
                                    eval_window == "unrestricted", auroc]
  h4_delta_auroc <- if (length(random_auroc) > 0 && length(temporal_auroc) > 0)
                      random_auroc - temporal_auroc else NA
} else {
  h4_delta_auroc <- NA
  cat("Random split results not found; H4 cannot be tested.\n")
}

# Factor 5: METRIC
internal_primary_B <- metrics[model == "primary" & variant == "B" & eval_window == "unrestricted"]
h5_internal_auroc  <- if (nrow(internal_primary_B) > 0) internal_primary_B$auroc else NA
h5_internal_util   <- if (nrow(internal_primary_B) > 0) internal_primary_B$utility_normalised else NA
if (!is.null(ext_res)) {
  ext_primary_B <- ext_res[variant == "B" & model == "pred_sepsis"]
  h5_ext_auroc  <- if (nrow(ext_primary_B) > 0) ext_primary_B$auroc else NA
} else { h5_ext_auroc <- NA }

# Factor 6: MODEL CLASS NULL (H6)
if (nrow(model_class_rows) > 0) {
  prim_auroc <- model_class_rows[model == "primary", auroc]
  gbt_auroc  <- model_class_rows[model == "gbt",     auroc]
  h6_delta   <- if (length(prim_auroc) > 0 && length(gbt_auroc) > 0)
                  abs(prim_auroc - gbt_auroc) else NA
} else { h6_delta <- NA }
```

## The Variance Decomposition Table

Primary deliverable. Each row quantifies how much a single analytical decision shifts reported performance.

```{r decomp-table}
decomp_table <- data.frame(
  factor = c(
    "Model class (label=B, temporal, unrestricted)",
    "Label operationalisation (model=primary, temporal, unrestricted)",
    "Prediction-time anchoring (model=primary, variant=B, temporal)",
    "Validation split (temporal vs random, model=primary, variant=B)",
    "Metric (AUROC vs Utility, internal→external, model=primary, variant=B)",
    "Model-class null |primary − GBT| (variant=B, temporal, unrestricted)"
  ),
  spread_auroc = round(c(model_spread_auroc, label_spread_auroc,
                          h3_delta_auroc, h4_delta_auroc,
                          if (!is.na(h5_internal_auroc) && !is.na(h5_ext_auroc))
                            h5_internal_auroc - h5_ext_auroc else NA,
                          h6_delta), 4),
  spread_utility = round(c(model_spread_utility, label_spread_utility,
                             NA, NA, NA, NA), 4),
  hypothesis  = c("-", "H1", "H3", "H4", "H5", "H6"),
  h_threshold = c("-", "label ≥ model", "Δ ≥ 0.03", "Δ ≥ 0.02",
                   "utility drop > AUROC drop", "|Δ| ≤ 0.02")
)

knitr::kable(decomp_table, caption = "THE VARIANCE DECOMPOSITION TABLE")
save_parquet(decomp_table, "12_variance_decomposition_table", OUTPUT_DIR)
```

## Hypothesis Test Verdicts

Each hypothesis is evaluated against its threshold. "CONFIRMED" = effect meets/exceeds threshold; "NOT confirmed" = falls short. These are descriptive verdicts, not formal statistical tests.

```{r hypothesis-verdicts}
step("Hypothesis test verdicts...")
verdicts <- list()

# H1
if (!is.na(label_spread_auroc) && !is.na(model_spread_auroc)) {
  h1_result <- label_spread_auroc >= model_spread_auroc
  cat(sprintf("H1 (Label dominance): label_spread=%.3f, model_spread=%.3f → %s\n",
              label_spread_auroc, model_spread_auroc,
              if (h1_result) "CONFIRMED" else "NOT confirmed"))
  verdicts[["H1"]] <- data.frame(hypothesis = "H1", finding = "Label dominance",
                                  threshold = "label >= model",
                                  observed  = sprintf("label=%.3f, model=%.3f",
                                                      label_spread_auroc, model_spread_auroc),
                                  confirmed = h1_result)
}

# H2
if (!is.null(coef_table) && "term" %in% names(coef_table)) {
  coef_sep <- coef_table[outcome == "sepsis" &
                           !grepl("^ns\\(|^\\(Intercept\\)", as.character(term))]
  if (nrow(coef_sep) > 0) {
    common_terms <- Reduce(intersect, lapply(c("A","B","C"), function(v)
      unique(as.character(coef_sep[variant == v, term]))))
    sign_changes <- character(0)
    for (tm in common_terms) {
      vals  <- sapply(c("A","B","C"), function(v) {
        r <- coef_sep[variant == v & as.character(term) == tm, coef]
        if (length(r) == 0) NA_real_ else as.numeric(r[1])
      })
      signs <- sign(vals[!is.na(vals)])
      if (length(signs) >= 2 && diff(range(signs)) != 0)
        sign_changes <- c(sign_changes, tm)
    }
    h2_result <- length(sign_changes) >= 1
    cat(sprintf("H2 (Coefficient instability): %d sign-changes: %s → %s\n",
                length(sign_changes), paste(head(sign_changes, 3), collapse = ", "),
                if (h2_result) "CONFIRMED" else "NOT confirmed"))
    verdicts[["H2"]] <- data.frame(hypothesis = "H2", finding = "Coefficient instability",
                                    threshold = ">=1 covariate changes sign",
                                    observed  = paste(sign_changes, collapse = "; "),
                                    confirmed = h2_result)
  }
}

# H3
if (!is.na(h3_delta_auroc)) {
  h3_result <- h3_delta_auroc >= 0.03
  cat(sprintf("H3 (Leakage): unrestricted − anchored AUROC Δ=%.3f → %s\n",
              h3_delta_auroc, if (h3_result) "CONFIRMED" else "NOT confirmed (< 0.03)"))
  verdicts[["H3"]] <- data.frame(hypothesis = "H3", finding = "Leakage via treatment anchoring",
                                  threshold = "Δ ≥ 0.03",
                                  observed  = sprintf("Δ=%.3f", h3_delta_auroc),
                                  confirmed = h3_result)
}

# H4
if (!is.na(h4_delta_auroc)) {
  h4_result <- h4_delta_auroc >= 0.02
  cat(sprintf("H4 (Split optimism): random − temporal AUROC Δ=%.3f → %s\n",
              h4_delta_auroc, if (h4_result) "CONFIRMED" else "NOT confirmed (< 0.02)"))
  verdicts[["H4"]] <- data.frame(hypothesis = "H4", finding = "Split optimism",
                                  threshold = "Δ ≥ 0.02",
                                  observed  = sprintf("Δ=%.3f", h4_delta_auroc),
                                  confirmed = h4_result)
}

# H6
if (!is.na(h6_delta)) {
  h6_result <- h6_delta <= 0.02
  cat(sprintf("H6 (Model-class null): |primary − GBT| AUROC=%.3f → %s\n",
              h6_delta, if (h6_result) "NOT rejected (interpretable ≈ GBT)"
                        else "REJECTED (GBT wins by > 0.02)"))
  verdicts[["H6"]] <- data.frame(hypothesis = "H6", finding = "Model-class null",
                                  threshold = "|Δ| ≤ 0.02",
                                  observed  = sprintf("Δ=%.3f", h6_delta),
                                  confirmed = h6_result)
}

if (length(verdicts) > 0) {
  verdicts_df <- do.call(rbind, verdicts)
  knitr::kable(verdicts_df, caption = "Hypothesis test verdicts")
  save_parquet(verdicts_df, "12_hypothesis_verdicts", OUTPUT_DIR)
}
```

## Variance Decomposition Plots

Label-variant effect vs. model-class effect on AUROC. If H1 is confirmed, the label-variant spread exceeds the model-class spread.

```{r variance-plots, fig.cap="Variance decomposition: label-variant and model-class effects on AUROC.", fig.width=6, fig.height=4}
if (nrow(metrics[!is.na(auroc)]) > 0) {
  # Label-variant effect
  tryCatch({
    p_label <- ggplot(metrics[model == "primary" & eval_window == "unrestricted"],
                       aes(x = variant, y = auroc, fill = variant)) +
      geom_col() +
      geom_hline(yintercept = if (!is.na(h5_internal_auroc)) h5_internal_auroc else 0.75,
                 linetype = "dashed") +
      labs(title    = "Label-variant effect on AUROC (primary model, temporal split)",
           subtitle = "H1: label spread should be ≥ model-class spread",
           x = "Label Variant", y = "AUROC") +
      theme_bw() + theme(legend.position = "none")
    print(p_label)
    ggsave(file.path(OUTPUT_DIR, "12_label_variance_auroc.png"), p_label,
           width = 6, height = 4, dpi = 150)
  }, error = function(e) cat("Label plot failed:", e$message, "\n"))

  # Model-class effect
  tryCatch({
    p_model <- ggplot(metrics[variant == "B" & eval_window == "unrestricted" & !is.na(auroc)],
                       aes(x = model, y = auroc, fill = model)) +
      geom_col() +
      labs(title    = "Model-class effect on AUROC (Variant B, temporal split)",
           subtitle = "H6: |primary − GBT| ≤ 0.02",
           x = "Model", y = "AUROC") +
      theme_bw() + theme(axis.text.x = element_text(angle = 30, hjust = 1),
                         legend.position = "none")
    print(p_model)
    ggsave(file.path(OUTPUT_DIR, "12_model_class_auroc.png"), p_model,
           width = 6, height = 4, dpi = 150)
  }, error = function(e) cat("Model plot failed:", e$message, "\n"))
}

cat("\nVariance decomposition complete.\n")
```
