---
title: "13 Equity Analysis - Subgroup Performance and Label Sensitivity"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)   # student ID for reproducibility

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

Wang et al. (2022) showed that sepsis prediction models perform worse for Asian, Hispanic, and Spanish-speaking patients. This notebook extends their work with:

1. **Subgroup performance**: AUROC, AUPRC, and alert burden evaluated per demographic subgroup (race/ethnicity, language, insurance).

2. **Subgroup label-sensitivity**: How the fraction labelled septic varies across the three label variants *within each subgroup*. Non-uniform sensitivity means label choice has equity implications beyond aggregate event-rate differences.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(PRROC)
  library(ggplot2)
  library(tidyr)
  library(dplyr)
})

source("config.R")
source("utils.R")

step("=== Phase 6b: Equity Analysis ===")
```

## Subgroup Performance

### Metric Helper

Computes AUROC, AUPRC, and alert burden for any subgroup slice with minimum-sample guards.

```{r subgroup-helper}
compute_subgroup_metrics <- function(dt, pred_col, subgroup_col, subgroup_val) {
  sub <- dt[get(subgroup_col) == subgroup_val & !is.na(get(pred_col))]
  if (nrow(sub) < 50) return(NULL)

  y     <- as.integer(sub$event == 1)
  pred  <- as.numeric(sub[[pred_col]])
  alert <- as.integer(pred >= 0.5)
  n_pos <- sum(y)
  if (n_pos == 0) return(NULL)

  roc_pr <- tryCatch({
    list(
      auroc = PRROC::roc.curve(pred, weights.class0 = y)$auc,
      auprc = PRROC::pr.curve(pred,  weights.class0 = y)$auc.integral
    )
  }, error = function(e) list(auroc = NA, auprc = NA))

  tp     <- sum(alert == 1 & y == 1, na.rm = TRUE)
  fp     <- sum(alert == 1 & y == 0, na.rm = TRUE)
  burden <- if (tp == 0) Inf else fp / tp

  data.frame(
    subgroup_col = subgroup_col,
    subgroup_val = subgroup_val,
    n_rows       = nrow(sub),
    n_stays      = length(unique(sub$stay_id)),
    n_sepsis     = n_pos,
    auroc        = round(roc_pr$auroc, 4),
    auprc        = round(roc_pr$auprc, 4),
    alert_burden = round(burden, 2)
  )
}
```

### Performance by Subgroup (Variant B, Primary Model)

Top subgroup levels (by sample size) are evaluated across primary, GBT, and NEWS2.

```{r subgroup-performance}
step("Subgroup performance analysis...")

prim_path <- file.path(OUTPUT_DIR, "08_predictions_primary_B.parquet")
comp_path <- file.path(OUTPUT_DIR, "09_predictions_comparators_B.parquet")

perf_df <- NULL

if (!file.exists(prim_path)) {
  cat("Primary model predictions not found; skipping subgroup performance.\n")
} else {
  ph <- as.data.table(arrow::read_parquet(prim_path))
  if (file.exists(comp_path)) {
    comp       <- as.data.table(arrow::read_parquet(comp_path))
    merge_cols <- c("stay_id", "hour", setdiff(names(comp), names(ph)))
    ph <- merge(ph, comp[, ..merge_cols], by = c("stay_id", "hour"), all.x = TRUE)
  }

  ph_demo_path <- file.path(OUTPUT_DIR, "03_person_hours_B.parquet")
  if (file.exists(ph_demo_path)) {
    ph_demo <- as.data.table(arrow::read_parquet(ph_demo_path,
                                                   col_select = c("stay_id","race",
                                                                   "language","insurance")))
    ph_demo <- unique(ph_demo)
    ph <- merge(ph, ph_demo, by = "stay_id", all.x = TRUE)
  }

  perf_results <- list()
  models_eval  <- c("pred_sepsis", "pred_gbt", "pred_news2")

  for (subg_col in EQUITY_VARS) {
    if (!subg_col %in% names(ph)) next
    val_counts <- ph[, .N, by = subg_col][order(-N)]
    top_vals   <- head(val_counts[[subg_col]], 8)
    for (val in top_vals) {
      for (mod_col in models_eval) {
        if (!mod_col %in% names(ph)) next
        res <- compute_subgroup_metrics(ph, mod_col, subg_col, val)
        if (!is.null(res)) {
          res$model   <- mod_col
          res$variant <- "B"
          perf_results[[paste(subg_col, val, mod_col, sep = "-")]] <- res
        }
      }
    }
  }

  if (length(perf_results) > 0) {
    perf_df <- do.call(rbind, perf_results)
    save_parquet(perf_df, "13_subgroup_performance", OUTPUT_DIR)

    knitr::kable(
      perf_df[order(perf_df$subgroup_col, perf_df$subgroup_val), ],
      caption = "Subgroup performance (primary model, Variant B)"
    )
  }
}
```

### AUROC by Race

```{r auroc-by-race, fig.cap="AUROC stratified by race/ethnicity for the primary model (Variant B). Dashed line indicates the overall mean.", fig.width=8, fig.height=5}
if (!is.null(perf_df) && "race" %in% perf_df$subgroup_col) {
  race_df <- perf_df[perf_df$subgroup_col == "race" & perf_df$model == "pred_sepsis", ]
  if (nrow(race_df) > 0) {
    tryCatch({
      p <- ggplot(race_df, aes(x = reorder(subgroup_val, -auroc), y = auroc,
                                fill = subgroup_val)) +
        geom_col() +
        geom_hline(yintercept = mean(race_df$auroc, na.rm = TRUE),
                   linetype = "dashed", color = "gray40") +
        labs(title    = "AUROC by race/ethnicity (primary model, Variant B)",
             subtitle = "Wang et al. (2022): performance degraded for Asian, Hispanic, Spanish-speaking",
             x = "Race/Ethnicity", y = "AUROC") +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 40, hjust = 1), legend.position = "none")
      print(p)
      ggsave(file.path(OUTPUT_DIR, "13_auroc_by_race.png"), p, width = 8, height = 5, dpi = 150)
    }, error = function(e) cat("Race plot failed:", e$message, "\n"))
  }
}
```

## Subgroup Label-Sensitivity

For each subgroup, the fraction labelled septic is computed under all three label variants. A large spread (pp range across A, B, C) means the label choice disproportionately affects that group.

```{r label-sensitivity}
step("Subgroup label-sensitivity analysis...")

label_sens_results <- list()

for (subg_col in EQUITY_VARS) {
  cohort_data <- lapply(names(LABEL_VARIANTS), function(vid) {
    coh_path <- file.path(OUTPUT_DIR, paste0("02_cohort_", vid, ".parquet"))
    if (!file.exists(coh_path)) return(NULL)
    coh_full <- tryCatch(as.data.table(arrow::read_parquet(coh_path)),
                         error = function(e) NULL)
    if (is.null(coh_full)) return(NULL)
    want_cols <- intersect(c("stay_id","subject_id", subg_col, "onset_hour",
                              "label_variant"), names(coh_full))
    coh <- coh_full[, ..want_cols]
    coh[, variant   := vid]
    coh[, has_sepsis := !is.na(onset_hour)]
    coh
  })
  cohort_data <- Filter(Negate(is.null), cohort_data)
  if (length(cohort_data) == 0) next

  all_cohort <- rbindlist(cohort_data, fill = TRUE)
  if (!subg_col %in% names(all_cohort)) next

  val_counts <- all_cohort[variant == "B" & !is.na(get(subg_col)), .N, by = subg_col]
  top_vals   <- head(val_counts[order(-N)][[subg_col]], 8)

  for (val in top_vals) {
    subg_dt <- all_cohort[get(subg_col) == val]
    if (nrow(subg_dt) == 0) next

    by_variant <- subg_dt[, .(
      n_stays    = .N,
      n_sepsis   = sum(has_sepsis),
      pct_sepsis = round(100 * sum(has_sepsis) / .N, 1)
    ), by = variant]

    if (nrow(by_variant) >= 2) {
      pct_range <- diff(range(by_variant$pct_sepsis))
      get_pct   <- function(v) {
        r <- by_variant[variant == v, pct_sepsis]
        if (length(r) > 0) r[1] else NA_real_
      }
      label_sens_results[[paste(subg_col, val, sep = "-")]] <- data.frame(
        subgroup_col    = subg_col,
        subgroup_val    = as.character(val),
        pct_A           = get_pct("A"),
        pct_B           = get_pct("B"),
        pct_C           = get_pct("C"),
        label_range_pct = round(pct_range, 1)
      )
    }
  }
}

if (length(label_sens_results) > 0) {
  ls_df <- do.call(rbind, label_sens_results)
  save_parquet(ls_df, "13_subgroup_label_sensitivity", OUTPUT_DIR)

  knitr::kable(
    ls_df[order(-abs(ls_df$label_range_pct)), ],
    caption = "Subgroup label-sensitivity (% sepsis by variant A/B/C per demographic group)"
  )

  high_sens <- ls_df[!is.na(ls_df$label_range_pct) & ls_df$label_range_pct > 5, ]
  if (nrow(high_sens) > 0) {
    cat("\nSubgroups with highest label sensitivity (>5 pp range):\n")
    knitr::kable(high_sens)
  }
}
```

### Label Sensitivity vs. AUROC Deficit

Tests whether subgroups with higher label sensitivity also have lower AUROC. A negative trend would suggest label ambiguity compounds prediction difficulty.

```{r label-sens-vs-auroc, fig.cap="Subgroup label sensitivity (x) vs. AUROC (y). Each point is a demographic subgroup; a downward trend suggests label ambiguity compounds prediction difficulty.", fig.width=8, fig.height=6}
if (length(label_sens_results) > 0 && !is.null(perf_df)) {
  merged_eq <- merge(ls_df,
                     perf_df[perf_df$model == "pred_sepsis",
                              c("subgroup_col","subgroup_val","auroc")],
                     by = c("subgroup_col","subgroup_val"), all.x = TRUE)

  if (nrow(merged_eq[!is.na(merged_eq$auroc), ]) > 0) {
    tryCatch({
      p_eq <- ggplot(merged_eq[!is.na(merged_eq$auroc), ],
                     aes(x = label_range_pct, y = auroc,
                         label = paste0(subgroup_col, ":", subgroup_val))) +
        geom_point(aes(color = subgroup_col), size = 3) +
        geom_smooth(method = "lm", se = TRUE, color = "gray40") +
        geom_text(hjust = 1.1, size = 2.5) +
        labs(title    = "Label sensitivity vs. AUROC by subgroup",
             subtitle = "Novel conjunction: Wang et al. (2022) established each separately",
             x = "Label sensitivity (pp range across variants A–C)",
             y = "AUROC (primary model, Variant B)") +
        theme_bw()
      print(p_eq)
      ggsave(file.path(OUTPUT_DIR, "13_label_sens_vs_auroc.png"), p_eq,
             width = 8, height = 6, dpi = 150)
    }, error = function(e) cat("Equity plot failed:", e$message, "\n"))
  }
}

cat("\nEquity analysis and all V2 analyses complete.\n")
```
