---
title: "Statistical Analysis of Sepsis-3 in Critically Ill Patients (MIMIC-IV)"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)   # student ID for reproducibility

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)

library(tidyverse)
library(knitr)
library(kableExtra)
library(MASS)
library(mclust)
library(coda)
library(depmixS4)
library(effectsize)
library(broom)
library(patchwork)
library(pROC)
library(PRROC)
library(mgcv)
library(car)
library(mice)

theme_set(theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 11)))

PAL <- c("No sepsis" = "#5B9BD5", "Sepsis-3" = "#C0504D")
```

```{r load-data}
find_data <- function() {
  cands <- c("../../processed_data/05_analysis_matrix.parquet",
             "processed_data/05_analysis_matrix.parquet",
             "05_analysis_matrix.parquet")
  hit <- cands[file.exists(cands)]
  if (length(hit) == 0)
    stop("05_analysis_matrix.parquet not found. Run 04_extract_measurements.py then ",
         "05_sofa_sepsis3.py first.")
  hit[1]
}

to_bool <- function(x) {
  if (is.logical(x)) return(x)
  tolower(as.character(x)) %in% c("true", "t", "1", "yes")
}

OUT <- dirname(find_data())
FIG <- if (dir.exists("figure")) "figure" else { dir.create("figure"); "figure" }

raw <- as.data.frame(arrow::read_parquet(find_data()))

raw$sepsis <- factor(ifelse(to_bool(raw$sepsis3), "Sepsis-3", "No sepsis"),
                     levels = c("No sepsis", "Sepsis-3"))

# Non-circular predictors: physiology outside the SOFA/Sepsis-3 definition
# (SOFA uses platelets, bilirubin, creatinine, MAP/pressors, GCS, PaO2/FiO2).
feat_vars <- c("lactate_max", "wbc_max", "hr_max",
               "resp_rate_max", "temp_max_c", "age")

dat <- raw %>%
  dplyr::select(all_of(feat_vars), sepsis, sepsis3, sofa_total,
                hospital_expire_flag, septic_shock) %>%
  filter(if_all(all_of(feat_vars), ~ !is.na(.)))

n_total   <- nrow(raw)
n_model   <- nrow(dat)
n_sepsis  <- sum(dat$sepsis == "Sepsis-3")
prev_full <- mean(to_bool(raw$sepsis3))
prev_mod  <- mean(dat$sepsis == "Sepsis-3")
```

# Introduction {#intro}

Sepsis is life-threatening organ dysfunction caused by a dysregulated
host response to infection, and it remains one of the leading causes of
death in intensive care units worldwide. The **Sepsis-3** consensus
(Singer et al., 2016) defines sepsis operationally as a **suspected
infection** accompanied by an acute rise of **two or more points** in
the Sequential Organ Failure Assessment (**SOFA**) score.

This report analyses **MIMIC-IV v3.1**, a large, de-identified database
of patients admitted to the critical-care units of the Beth Israel
Deaconess Medical Center. From the raw tables a reproducible pipeline
(scripts `01` to `05`) builds an adult, first-ICU-stay cohort, detects
suspected infection from antibiotic orders paired with cultures,
computes the six SOFA organ scores from the first 24 hours of each stay,
and applies the Sepsis-3 rule. The result is an analysis matrix of
**`r format(n_total, big.mark=",")`** ICU stays, of which
**`r sprintf("%.1f%%", 100*prev_full)`** meet the Sepsis-3 definition.

The clinical question is whether **routinely measured physiology that is
*not* part of the sepsis definition** (lactate, white-cell count, heart
rate, respiratory rate, temperature, and age) can separate septic from
non-septic stays, and what different statistical methods reveal about
the same data. Choosing predictors outside the SOFA score avoids
circularity: the models recover the label from independent signals,
not from the quantities used to construct it.

All inferential and modelling sections use the
**`r format(n_model, big.mark=",")`** stays with a complete set of these
six measurements (`r sprintf("%.1f%%", 100*prev_mod)` Sepsis-3
positive). Because lactate is ordered preferentially in sicker patients,
this complete-case subset is enriched for severity relative to the full
cohort; this is revisited as a limitation in the conclusion.

The report follows a structured progression:

1.  **Exploratory data analysis** (Section \@ref(eda)) describes the
    cohort and feature distributions by sepsis status.
2.  **Hypothesis testing** (Section \@ref(hypothesis)) quantifies which
    features differ between groups, focusing on effect size.
3.  **Maximum likelihood estimation** (Section \@ref(mle)) models the
    typical lactate of a septic patient.
4.  **Generalised linear models** (Section \@ref(glm)) determine which
    features carry independent predictive power.
5.  **Kernel density estimation and mixture modelling** (Section
    \@ref(density)) attempt to recover the two groups without labels.
6.  **Metropolis-Hastings sampling** (Section \@ref(partA)) re-expresses
    the likelihood result in a Bayesian framework.
7.  A **Hidden Markov Model** (Section \@ref(partB)) tests whether the
    ordering of observations carries extra structure.
8.  **Predictive validation** (Section \@ref(validation)) subjects the
    model to out-of-sample cross-validation, calibration, precision-recall
    and decision-curve analysis, a non-linearity check, and a formal
    treatment of missing data.

# Exploratory Data Analysis {#eda}

This section examines the data before modelling, focusing on lactate,
white-cell count, and the vital signs central to bedside recognition
of sepsis.

```{r class-counts}
tab_cls <- table(dat$sepsis)
prop_cls <- prop.table(tab_cls)
imbalance <- as.numeric(tab_cls["No sepsis"] / tab_cls["Sepsis-3"])
```

In the modelling subset there are
**`r format(as.integer(tab_cls['No sepsis']), big.mark=',')`**
non-septic and
**`r format(as.integer(tab_cls['Sepsis-3']), big.mark=',')`** septic
stays (`r sprintf('%.1f%%', 100*prop_cls['Sepsis-3'])` septic). The
classes are imbalanced at roughly **`r sprintf('%.1f', imbalance)`:1**,
milder than many clinical-outcome datasets and not severe enough to
require resampling; the inferential focus on effect size further reduces
its impact.

```{r eda-summary}
key_vars <- c("lactate_max", "wbc_max", "hr_max",
              "resp_rate_max", "temp_max_c", "age")
pretty <- c(lactate_max = "Lactate, max (mmol/L)",
            wbc_max = "WBC, max (K/uL)",
            hr_max = "Heart rate, max (bpm)",
            resp_rate_max = "Resp. rate, max (/min)",
            temp_max_c = "Temperature, max (C)",
            age = "Age (years)")

summary_tab <- dat %>%
  dplyr::select(sepsis, all_of(key_vars)) %>%
  pivot_longer(-sepsis, names_to = "Feature", values_to = "Value") %>%
  group_by(Feature, sepsis) %>%
  summarise(Mean = mean(Value), Median = median(Value),
            SD = sd(Value), .groups = "drop") %>%
  pivot_wider(names_from = sepsis, values_from = c(Mean, Median, SD)) %>%
  mutate(Feature = pretty[Feature]) %>%
  dplyr::select(Feature,
                `Mean_No sepsis`, `Mean_Sepsis-3`,
                `Median_No sepsis`, `Median_Sepsis-3`,
                `SD_No sepsis`, `SD_Sepsis-3`)

kable(summary_tab, digits = 2, booktabs = TRUE,
      caption = "Summary statistics for the six features, by sepsis status.",
      col.names = c("Feature", "Mean (no)", "Mean (sep)",
                    "Med. (no)", "Med. (sep)", "SD (no)", "SD (sep)")) %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
```

Table \@ref(tab:eda-summary) shows that septic stays tend to have higher
lactate, higher white-cell counts, and faster heart and respiratory
rates. These shifts match the bedside picture of sepsis, although the
spread within each group is wide.

```{r eda-boxplots, fig.cap = "Three features by sepsis status. Septic stays show higher lactate, faster breathing, and faster heart rates, with substantial overlap."}
plot_vars <- c("lactate_max", "resp_rate_max", "hr_max")
plab <- c(lactate_max = "Lactate, max", resp_rate_max = "Resp. rate, max",
          hr_max = "Heart rate, max")

dat %>%
  dplyr::select(sepsis, all_of(plot_vars)) %>%
  pivot_longer(-sepsis, names_to = "Feature", values_to = "Value") %>%
  mutate(Feature = plab[Feature]) %>%
  ggplot(aes(sepsis, Value, fill = sepsis)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.4, outlier.alpha = 0.25) +
  facet_wrap(~ Feature, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = PAL) +
  labs(x = NULL, y = NULL) +
  theme(legend.position = "none")
```

Figure \@ref(fig:eda-boxplots) confirms the direction of these
differences but also the overlap: no single feature cleanly separates
the groups, which motivates the multivariable model in Section
\@ref(glm).

```{r eda-density, fig.cap = "Distribution of maximum lactate by sepsis status (values capped at 15 mmol/L for display). Both groups are right-skewed; the septic distribution is shifted upward."}
dat %>%
  mutate(lac = pmin(lactate_max, 15)) %>%
  ggplot(aes(lac, fill = sepsis)) +
  geom_density(alpha = 0.55) +
  scale_fill_manual(values = PAL) +
  labs(x = "Maximum lactate (mmol/L)", y = "Density", fill = NULL) +
  theme(legend.position = "top")
```

Figure \@ref(fig:eda-density) shows two overlapping, right-skewed
distributions. The septic curve is shifted toward higher lactate, but
the long right tail in both groups signals that a log transform will be
useful for the distributional models in Sections
\@ref(mle) to \@ref(partB).

*Collectively, this section shows the data contain genuine, clinically
coherent signal, but no single feature is sufficient for perfect
separation.*

# Hypothesis Testing {#hypothesis}

Three features are compared between septic and non-septic stays, each
with a different test. With **`r format(n_model, big.mark=",")`**
observations, even trivial differences yield very small p-values, so
interpretation centres on **effect size** (the magnitude of the
difference) rather than significance alone.

## Test 1: Maximum lactate (Welch t-test)

The two groups have unequal variances, so a Welch t-test is appropriate.

$H_0$: mean maximum lactate is equal in septic and non-septic stays\
$H_1$: mean maximum lactate differs between the groups

```{r t-test}
t1 <- t.test(lactate_max ~ sepsis, data = dat)
d1 <- effectsize::cohens_d(lactate_max ~ sepsis, data = dat)

kable(tibble(
  Statistic = c("t", "df", "p-value", "Mean (no sepsis)", "Mean (Sepsis-3)",
                "Mean difference", "Cohen's d", "95% CI for d"),
  Value = c(sprintf("%.2f", t1$statistic),
            sprintf("%.0f", t1$parameter),
            format.pval(t1$p.value, digits = 3, eps = 1e-16),
            sprintf("%.2f", t1$estimate[1]),
            sprintf("%.2f", t1$estimate[2]),
            sprintf("%.2f", diff(rev(t1$estimate))),
            sprintf("%.2f", d1$Cohens_d),
            sprintf("[%.2f, %.2f]", d1$CI_low, d1$CI_high))
), booktabs = TRUE,
   caption = "Welch t-test: maximum lactate by sepsis status.") %>%
  kable_styling(latex_options = "HOLD_position")
```

Table \@ref(tab:t-test) shows mean maximum lactate of
**`r sprintf('%.2f', t1$estimate[1])`** mmol/L in non-septic stays
versus **`r sprintf('%.2f', t1$estimate[2])`** mmol/L in septic stays, a
difference of **`r sprintf('%.2f', diff(rev(t1$estimate)))`** mmol/L.
The p-value is `r format.pval(t1$p.value, digits=3, eps=1e-16)`, so the
difference is not attributable to chance. Cohen's *d* of
**`r sprintf('%.2f', d1$Cohens_d)`** indicates a
`r ifelse(abs(d1$Cohens_d)>=0.8,'large',ifelse(abs(d1$Cohens_d)>=0.5,'moderate','small-to-moderate'))`
effect: lactate separates the groups, but not perfectly.

## Test 2: Maximum white-cell count (Wilcoxon rank-sum)

White-cell counts are heavily right-skewed, making a non-parametric
Wilcoxon rank-sum test appropriate.

$H_0$: the distribution of maximum WBC is the same in both groups\
$H_1$: the distributions differ

```{r wilcoxon}
w1 <- wilcox.test(wbc_max ~ sepsis, data = dat)
rb <- effectsize::rank_biserial(wbc_max ~ sepsis, data = dat)

kable(tibble(
  Statistic = c("W", "p-value", "Rank-biserial r", "95% CI"),
  Value = c(format(w1$statistic, big.mark = ","),
            format.pval(w1$p.value, digits = 3, eps = 1e-16),
            sprintf("%.2f", rb$r_rank_biserial),
            sprintf("[%.2f, %.2f]", rb$CI_low, rb$CI_high))
), booktabs = TRUE,
   caption = "Wilcoxon rank-sum: maximum WBC by sepsis status.") %>%
  kable_styling(latex_options = "HOLD_position")
```

Table \@ref(tab:wilcoxon) shows a significant difference in white-cell
counts. The rank-biserial correlation of
**`r sprintf('%.2f', rb$r_rank_biserial)`** is
`r ifelse(abs(rb$r_rank_biserial)>=0.5,'large',ifelse(abs(rb$r_rank_biserial)>=0.3,'moderate','small'))`
in magnitude, weaker than lactate and consistent with white-cell count
being a noisier marker that rises in many non-septic stresses as well.

## Test 3: Tachypnoea and sepsis (Chi-squared)

Respiratory rate is dichotomised at the **qSOFA** threshold of 22
breaths/min, a recognised marker of clinical deterioration. A
chi-squared test of independence assesses the association with sepsis.

$H_0$: tachypnoea ($\geq 22$/min) and sepsis are independent\
$H_1$: they are associated

```{r chisq}
dat$tachypnoea <- factor(ifelse(dat$resp_rate_max >= 22,
                                "RR >= 22", "RR < 22"))
tab <- table(dat$tachypnoea, dat$sepsis)
c1 <- chisq.test(tab)
v1 <- effectsize::cramers_v(tab)

kable(tibble(
  Statistic = c("Chi-squared", "df", "p-value", "Cramer's V", "95% CI"),
  Value = c(sprintf("%.1f", c1$statistic),
            c1$parameter,
            format.pval(c1$p.value, digits = 3, eps = 1e-16),
            sprintf("%.2f", v1$Cramers_v_adjusted),
            sprintf("[%.2f, %.2f]", v1$CI_low, v1$CI_high))
), booktabs = TRUE,
   caption = "Chi-squared: tachypnoea (RR >= 22) by sepsis status.") %>%
  kable_styling(latex_options = "HOLD_position")
```

Table \@ref(tab:chisq) shows a significant association between
tachypnoea and sepsis, with Cramer's V of
**`r sprintf('%.2f', v1$Cramers_v_adjusted)`**
(`r ifelse(v1$Cramers_v_adjusted>=0.25,'a strong',ifelse(v1$Cramers_v_adjusted>=0.1,'a moderate','a weak'))`
association). Dichotomising at a single threshold discards information;
the GLM in Section \@ref(glm) uses the continuous respiratory rate and
recovers a stronger signal.

***Collectively,** all three features carry genuine signal. Lactate
shows the largest effect; white-cell count and tachypnoea are weaker but
real.*

# Maximum Likelihood Estimation {#mle}

Lactate is strictly positive and right-skewed, fitting a **log-normal**
model: the log of maximum lactate is assumed Normal. Maximum likelihood
estimates the mean $\mu$ and standard deviation $\sigma$ on the log
scale, and the typical (median) lactate is recovered as $e^{\mu}$.

```{r mle}
x_s <- log(dat$lactate_max[dat$sepsis == "Sepsis-3"])
x_s <- x_s[is.finite(x_s)]
n_s <- length(x_s)

neg_loglik <- function(par, x) {
  mu <- par[1]; sigma <- par[2]
  if (sigma <= 0) return(Inf)
  -sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}

fit <- optim(par = c(mean(x_s), sd(x_s)), fn = neg_loglik, x = x_s,
             method = "L-BFGS-B", lower = c(-Inf, 1e-6), hessian = TRUE)

mu_hat <- fit$par[1]; sigma_hat <- fit$par[2]
se_mu <- sqrt(diag(solve(fit$hessian)))[1]
median_lactate <- exp(mu_hat)

kable(tibble(
  Parameter = c("mu (mean log-lactate | sepsis)",
                "sigma (SD log-lactate | sepsis)",
                "n (septic)", "SE of mu",
                "Median lactate exp(mu) (mmol/L)", "95% CI for median"),
  Estimate = c(sprintf("%.3f", mu_hat),
               sprintf("%.3f", sigma_hat),
               format(n_s, big.mark = ","),
               sprintf("%.4f", se_mu),
               sprintf("%.2f", median_lactate),
               sprintf("[%.2f, %.2f]",
                       exp(mu_hat - 1.96 * se_mu),
                       exp(mu_hat + 1.96 * se_mu)))
), booktabs = TRUE,
   caption = "MLE for the log-normal model of septic maximum lactate.") %>%
  kable_styling(latex_options = "HOLD_position")
```

*The maximum-likelihood estimate from Table* \@ref(tab:mle) *shows:*

1.  *The typical (median) maximum lactate for a septic stay is about*
    **`r sprintf('%.2f', median_lactate)`** *mmol/L, above the 2 mmol/L
    threshold used in the septic-shock definition.*
2.  *The log-scale standard deviation of*
    **`r sprintf('%.3f', sigma_hat)`** *reflects the wide spread of
    lactate values.*
3.  *The very small standard error shows the central value is estimated
    precisely, a direct consequence of the large septic sample.*

# Generalised Linear Models {#glm}

The binary outcome calls for logistic regression. Two models are
compared:

-   **Model 1** uses maximum lactate only.
-   **Model 2** adds white-cell count, heart rate, respiratory rate,
    temperature, and age.

```{r correlation-check}
cor_mat <- cor(dat[, c("hr_max", "resp_rate_max", "lactate_max")])
```

A correlation check among the continuous predictors shows the strongest
pairwise correlation is
**`r sprintf('%.2f', max(abs(cor_mat[lower.tri(cor_mat)])))`**, well
below the 0.8 threshold that would signal redundancy, so all predictors
are retained.

```{r glm-models}
dat$y <- as.integer(dat$sepsis == "Sepsis-3")
m1 <- glm(y ~ lactate_max, data = dat, family = binomial)
m2 <- glm(y ~ lactate_max + wbc_max + hr_max + resp_rate_max +
            temp_max_c + age, data = dat, family = binomial)
```

## Logistic Regression Coefficients

```{r glm-table}
glm_tab <- bind_rows(
  broom::tidy(m1) %>% mutate(Model = "M1: lactate only"),
  broom::tidy(m2) %>% mutate(Model = "M2: 6 predictors")
) %>%
  dplyr::select(Model, term, estimate, std.error, p.value) %>%
  mutate(across(c(estimate, std.error), ~ sprintf("%.3f", .)),
         p.value = format.pval(p.value, digits = 2, eps = 1e-16))

kable(glm_tab, booktabs = TRUE,
      col.names = c("Model", "Term", "Estimate", "SE", "p-value"),
      caption = "Logistic regression coefficients (log-odds scale).") %>%
  kable_styling(latex_options = "HOLD_position")
```

Coefficients in Table \@ref(tab:glm-table) are on the log-odds scale.
Converting to odds ratios aids interpretation.

```{r odds-ratios}
or2 <- exp(coef(m2))
or_ci <- exp(confint.default(m2))
or_tab <- tibble(
  Term = names(or2),
  `Odds ratio` = sprintf("%.3f", or2),
  `95% CI` = sprintf("[%.3f, %.3f]", or_ci[, 1], or_ci[, 2])
) %>% filter(Term != "(Intercept)")

kable(or_tab, booktabs = TRUE,
      caption = "Model 2 odds ratios (per one-unit increase).") %>%
  kable_styling(latex_options = "HOLD_position")

or_lac <- or2["lactate_max"]
```

The odds ratio for maximum lactate is **`r sprintf('%.2f', or_lac)`**,
meaning each additional 1 mmol/L multiplies the odds of sepsis by about
`r sprintf('%.2f', or_lac)`, holding the other variables constant. The
vital signs and age each contribute smaller but independent effects,
with odds ratios above 1 indicating that faster heart and respiratory
rates, higher temperature, and older age are each associated with a
higher probability of sepsis once the others are accounted for.

## Model Comparison

```{r aic-table}
prob2 <- predict(m2, type = "response")
auc <- {
  r <- rank(prob2)
  n1 <- sum(dat$y == 1); n0 <- sum(dat$y == 0)
  (sum(r[dat$y == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

aic_tab <- tibble(
  Model = c("M1: lactate only", "M2: six predictors"),
  AIC = c(AIC(m1), AIC(m2)),
  `Residual deviance` = c(m1$deviance, m2$deviance),
  df = c(m1$df.residual, m2$df.residual)
)

kable(aic_tab, booktabs = TRUE, digits = 1,
      caption = "Model comparison (lower AIC is better).") %>%
  kable_styling(latex_options = "HOLD_position")
```

Table \@ref(tab:aic-table) shows the AIC falling from
**`r sprintf('%.0f', AIC(m1))`** (lactate only) to
**`r sprintf('%.0f', AIC(m2))`** (six predictors), a clear improvement:
the additional features carry information about sepsis beyond lactate
alone. The six-predictor model achieves an approximate AUC of
**`r sprintf('%.2f', auc)`**, indicating
`r ifelse(auc>=0.8,'good',ifelse(auc>=0.7,'fair','modest'))`
discrimination from physiology that is external to the sepsis
definition.

*Sepsis is therefore best explained by a combination of features rather
than any single measurement.*

# Density Estimation and Mixture Models {#density}

Setting aside the sepsis labels, the question is whether maximum
lactate alone splits the data into two natural groups.

## Kernel Density Estimate {#kde-section}

Silverman's rule-of-thumb bandwidth is
$$h = 0.9 \times \min\!\left(s,\, \frac{IQR}{1.34}\right) \times n^{-1/5},$$
where $s$ is the standard deviation, $IQR$ the interquartile range, and
$n$ the sample size. Because lactate is right-skewed, the KDE is
computed on the **log** scale, where the distribution is closer to
symmetric.

```{r bandwidth}
loglac <- log(dat$lactate_max)
loglac <- loglac[is.finite(loglac)]
bw_sil <- bw.nrd0(loglac)
```

The computed bandwidth is **`r sprintf('%.3f', bw_sil)`** on the log
scale.

```{r kde, fig.cap = "Kernel density estimate of log maximum lactate (Gaussian kernel, Silverman bandwidth). A shoulder above the main peak hints at a second, higher-lactate population."}
ggplot(tibble(loglac = loglac), aes(loglac)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "grey85", colour = "grey60") +
  geom_density(bw = bw_sil, colour = "#C0504D", linewidth = 0.9) +
  labs(x = "log(maximum lactate)", y = "Density",
       subtitle = sprintf("Gaussian kernel, bandwidth = %.3f (Silverman)", bw_sil))
```

Figure \@ref(fig:kde) shows a dominant peak with a right shoulder,
suggesting the data may be a blend of a larger normal-lactate population
and a smaller high-lactate one rather than a single homogeneous group.

## Two-component Gaussian mixture {#mixture-section}

A two-component Gaussian mixture is fitted to log-lactate by
Expectation-Maximisation, using no labels:
$$f(x) = \pi_1 \mathcal{N}(\mu_1, \sigma_1^2) + (1-\pi_1)\mathcal{N}(\mu_2, \sigma_2^2).$$

```{r mixture, fig.cap = "Two-component Gaussian mixture (dashed components, solid sum) fitted to log maximum lactate."}
mix2 <- Mclust(loglac, G = 2, modelNames = "V", verbose = FALSE)

pi1 <- mix2$parameters$pro[1]
mu1 <- mix2$parameters$mean[1]; mu2 <- mix2$parameters$mean[2]
s1  <- sqrt(mix2$parameters$variance$sigmasq[1])
s2  <- sqrt(mix2$parameters$variance$sigmasq[2])

xs <- seq(min(loglac), max(loglac), length.out = 400)
mix_df <- tibble(
  x = xs,
  comp1 = pi1 * dnorm(xs, mu1, s1),
  comp2 = (1 - pi1) * dnorm(xs, mu2, s2),
  total = comp1 + comp2)

ggplot(tibble(loglac = loglac), aes(loglac)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "grey88", colour = "grey60") +
  geom_line(data = mix_df, aes(x, comp1), linetype = "dashed", colour = "#5B9BD5") +
  geom_line(data = mix_df, aes(x, comp2), linetype = "dashed", colour = "#C0504D") +
  geom_line(data = mix_df, aes(x, total), colour = "black", linewidth = 0.7) +
  labs(x = "log(maximum lactate)", y = "Density")

# Assign each stay to its most probable component; the higher-mean component is
# the "septic" guess. Compare with the true label.
hi <- which.max(c(mu1, mu2))
mix_pred <- factor(ifelse(mix2$classification == hi,
                          "Sepsis-3 (pred)", "No sepsis (pred)"),
                   levels = c("No sepsis (pred)", "Sepsis-3 (pred)"))
agree_tab <- table(Predicted = mix_pred, Actual = dat$sepsis[is.finite(log(dat$lactate_max))])
agree_rate <- sum(diag(agree_tab)) / sum(agree_tab)
mix_means <- exp(sort(c(mu1, mu2)))
```

The mixture places one component near
**`r sprintf('%.2f', mix_means[1])`** mmol/L (back-transformed) and the
other near **`r sprintf('%.2f', mix_means[2])`** mmol/L, with mixing
weights `r sprintf('%.2f', sort(c(pi1, 1-pi1), decreasing=TRUE)[1])` and
`r sprintf('%.2f', sort(c(pi1, 1-pi1), decreasing=TRUE)[2])`.

```{r mixture-accuracy}
mix_acc <- round(100 * agree_rate, 1)
```

Matching the unlabelled components to the true labels gives an
unsupervised agreement of **`r mix_acc`%**. Unlike a marker such as
tumour radius, which nearly separates its two classes, lactate alone
only **partially** recovers sepsis status: the two populations overlap
heavily, making a single biomarker a weak unsupervised classifier. This
informative negative result illustrates why the Sepsis-3 definition
combines infection with a multi-organ score rather than relying on any
one number.

# Bayesian Inference and MCMC {#bayesian}

Maximum likelihood (Section \@ref(mle)) gave a single best estimate of
the mean log-lactate. Bayesian inference returns a full posterior
distribution, expressing uncertainty as a range of plausible values.

## Part A: Bayesian inference via Metropolis-Hastings {#partA}

### Model setup

Log maximum lactate for septic stays is modelled as Normal with unknown
mean $\mu$ and a known standard deviation fixed at the MLE $\hat\sigma$
from Section \@ref(mle). A weak prior
$\mu \sim \mathcal{N}(\log 2,\, 1^2)$ is centred at 2 mmol/L (the
clinical threshold) but is wide enough to let the data dominate.

### Likelihood and posterior

The posterior combines the Normal likelihood with the Normal prior,
$$p(\mu \mid x) \propto p(x \mid \mu)\, p(\mu),$$ which here is
conjugate (and therefore Normal). The Metropolis-Hastings algorithm is
nonetheless used to demonstrate a general sampling approach that does
not rely on conjugacy.

```{r mh-sampler}
log_posterior <- function(mu, x, sigma, mu0, tau0) {
  sum(dnorm(x, mean = mu, sd = sigma, log = TRUE)) +
    dnorm(mu, mean = mu0, sd = tau0, log = TRUE)
}

mh_sampler <- function(x, sigma, mu0, tau0, n_iter, init, prop_sd) {
  chain <- numeric(n_iter); chain[1] <- init; accept <- 0
  cur_lp <- log_posterior(init, x, sigma, mu0, tau0)
  for (t in 2:n_iter) {
    prop <- rnorm(1, chain[t - 1], prop_sd)
    prop_lp <- log_posterior(prop, x, sigma, mu0, tau0)
    if (log(runif(1)) < prop_lp - cur_lp) {
      chain[t] <- prop; cur_lp <- prop_lp; accept <- accept + 1
    } else chain[t] <- chain[t - 1]
  }
  list(chain = chain, accept_rate = accept / (n_iter - 1))
}
```

Two chains are run from deliberately poor, opposite starting points and
the first 2,000 iterations discarded as burn-in.

```{r mh-run}
n_iter <- 20000; burn <- 2000
prop_sd <- max(0.01, 3 * se_mu)
run1 <- mh_sampler(x_s, sigma_hat, mu0 = log(2), tau0 = 1,
                   n_iter = n_iter, init = mu_hat - 0.5, prop_sd = prop_sd)
run2 <- mh_sampler(x_s, sigma_hat, mu0 = log(2), tau0 = 1,
                   n_iter = n_iter, init = mu_hat + 0.5, prop_sd = prop_sd)
post1 <- run1$chain[-(1:burn)]; post2 <- run2$chain[-(1:burn)]
post <- post1
post_mean <- mean(post); post_sd <- sd(post)
post_ci <- quantile(post, c(0.025, 0.975))
```

### Convergence diagnostics

```{r mh-diag, fig.cap = "Trace plots (left) of two chains overlap after burn-in; posterior density (right) with the MLE marked (dashed).", fig.width = 7, fig.height = 3}
trace_df <- bind_rows(
  tibble(iter = seq_along(post1), value = post1, chain = "Chain 1"),
  tibble(iter = seq_along(post2), value = post2, chain = "Chain 2"))

p_trace <- ggplot(trace_df, aes(iter, value, colour = chain)) +
  geom_line(alpha = 0.6, linewidth = 0.25) +
  scale_colour_manual(values = c("#5B9BD5", "#C0504D")) +
  labs(x = "Iteration (post burn-in)", y = expression(mu), colour = NULL) +
  theme(legend.position = "top")

p_post <- ggplot(tibble(mu = post), aes(mu)) +
  geom_density(fill = "#5B9BD5", alpha = 0.55) +
  geom_vline(xintercept = mu_hat, linetype = "dashed") +
  labs(x = expression(mu), y = "Posterior density")

p_trace + p_post
```

```{r gelman}
mcmc_list <- mcmc.list(mcmc(post1), mcmc(post2))
rhat <- gelman.diag(mcmc_list, autoburnin = FALSE)$psrf[1, 1]
ess <- effectiveSize(mcmc(post))
```

The chains overlap after burn-in (Figure \@ref(fig:mh-diag)). The
Gelman-Rubin statistic is $\hat R =$ **`r sprintf('%.3f', rhat)`**
(values below 1.05 indicate convergence) and the effective sample size
is about **`r format(round(ess), big.mark=',')`**, so the posterior
samples are reliable.

### Posterior summary and comparison with MLE

```{r mh-summary}
kable(tibble(
  Quantity = c("Mean log-lactate | sepsis", "SE / posterior SD",
               "95% interval", "Implied median lactate (mmol/L)"),
  MLE = c(sprintf("%.3f", mu_hat), sprintf("%.4f", se_mu),
          sprintf("[%.3f, %.3f]", mu_hat - 1.96*se_mu, mu_hat + 1.96*se_mu),
          sprintf("%.2f", exp(mu_hat))),
  `Bayes (M-H)` = c(sprintf("%.3f", post_mean), sprintf("%.4f", post_sd),
                    sprintf("[%.3f, %.3f]", post_ci[1], post_ci[2]),
                    sprintf("%.2f", exp(post_mean)))
), booktabs = TRUE,
   caption = "MLE versus Bayesian posterior for mean log-lactate.") %>%
  kable_styling(latex_options = "HOLD_position")
```

Table \@ref(tab:mh-summary) shows the Bayesian and maximum-likelihood
estimates are almost identical. With a large septic sample and a weak
prior, the data dominate; a much stronger prior centred at $\log 2$
would be needed to pull the estimate appreciably toward 2 mmol/L.

*The two frameworks agree, but the Bayesian result additionally provides
a full posterior, which becomes more valuable when samples are small or
strong prior knowledge exists.*

## Part B: Hidden Markov Model {#partB}

The mixture model treated stays as independent. A hidden Markov model
allows consecutive stays to share a latent state via a transition
matrix.

**Note on ordering.** The stays are not a true time series; their order
is the processing order. The question is therefore deliberately narrow:
*after splitting the data into two lactate states, does the ordering add
any structure?* If not, the HMM should reproduce the mixture model, and
this negative result is itself informative. The natural application of an
HMM is to *within-patient* hourly vital-sign sequences, where a latent
state genuinely evolves toward organ dysfunction; that temporal use is the
subject of the follow-on longitudinal analysis (see the Conclusion) and is
outside the scope of this cross-sectional report.

### Choosing the number of states

```{r hmm-fit, results = 'hide'}
set.seed(486649)
hmm_n <- min(8000, length(loglac))
hmm_x <- data.frame(loglac = loglac[seq_len(hmm_n)])

fit_hmm <- function(k) {
  mod <- depmix(loglac ~ 1, data = hmm_x, nstates = k, family = gaussian())
  fit(mod, verbose = FALSE, emcontrol = em.control(random.start = TRUE))
}
invisible(capture.output({
  hmm1 <- fit_hmm(1); hmm2 <- fit_hmm(2); hmm3 <- fit_hmm(3)
}))
bic_vals <- c(BIC(hmm1), BIC(hmm2), BIC(hmm3))
best_k <- c(1, 2, 3)[which.min(bic_vals)]
```

```{r hmm-bic-table}
kable(tibble(
  States = 1:3,
  `Log-likelihood` = c(as.numeric(logLik(hmm1)), as.numeric(logLik(hmm2)),
                       as.numeric(logLik(hmm3))),
  BIC = bic_vals
), booktabs = TRUE, digits = 1,
   caption = "HMM model selection by BIC (lower is better).") %>%
  kable_styling(latex_options = "HOLD_position")
```

Table \@ref(tab:hmm-bic-table) selects the **`r best_k`-state** model by
BIC,
`r ifelse(best_k==2, 'consistent with the two-population structure seen in Section 6', 'which is discussed below relative to the mixture result')`.

### Emissions and transition matrix

```{r hmm-params}
pars <- getpars(hmm2)
init_probs <- pars[1:2]
trans_mat <- matrix(pars[3:6], nrow = 2, byrow = TRUE)
mu_h1 <- pars[7]; sd_h1 <- pars[8]
mu_h2 <- pars[9]; sd_h2 <- pars[10]
if (mu_h1 > mu_h2) {
  tmp <- c(mu_h1, sd_h1); mu_h1 <- mu_h2; sd_h1 <- sd_h2
  mu_h2 <- tmp[1]; sd_h2 <- tmp[2]
  trans_mat <- trans_mat[c(2, 1), c(2, 1)]
}
```

```{r hmm-emissions-table}
kable(tibble(
  State = c("State 1 (lower)", "State 2 (higher)"),
  `Emission mean (log)` = c(mu_h1, mu_h2),
  `Median lactate (mmol/L)` = exp(c(mu_h1, mu_h2)),
  `Mixture mean (Sec. 6)` = sort(c(mu1, mu2))
), booktabs = TRUE, digits = 3,
   caption = "HMM emissions versus the Section 6 mixture components.") %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
```

```{r hmm-trans-table}
trans_disp <- as_tibble(round(trans_mat, 3), .name_repair = "minimal")
names(trans_disp) <- c("To State 1", "To State 2")
trans_disp <- bind_cols(tibble(From = c("State 1", "State 2")), trans_disp)
kable(trans_disp, booktabs = TRUE,
      caption = "Estimated transition matrix.") %>%
  kable_styling(latex_options = "HOLD_position")

eig <- eigen(t(trans_mat))
stat <- Re(eig$vectors[, which.min(abs(eig$values - 1))])
stat <- stat / sum(stat)
```

Table \@ref(tab:hmm-emissions-table) shows the HMM emission means almost
coincide with the mixture components from Section
\@ref(mixture-section): both methods place the two lactate populations
in the same locations. The transition matrix (Table
\@ref(tab:hmm-trans-table)) has a stationary distribution of
$(`r round(stat[1],2)`, `r round(stat[2],2)`)$. Any "stickiness" on the
diagonal is the pattern EM tends to produce when fitting independent,
overlapping data, not evidence of real sequential dependence.

### Unsupervised classification accuracy

```{r hmm-accuracy}
v_state <- posterior(hmm2, type = "viterbi")$state
hi_state <- if (mean(hmm_x$loglac[v_state == 1]) <
                mean(hmm_x$loglac[v_state == 2])) 2 else 1
hmm_pred <- ifelse(v_state == hi_state, "Sepsis-3", "No sepsis")
actual <- dat$sepsis[is.finite(log(dat$lactate_max))][seq_len(hmm_n)]
hmm_acc <- mean(hmm_pred == as.character(actual))
```

Decoding the most likely state for each stay and matching states to
labels gives an unsupervised accuracy of
**`r sprintf('%.1f', 100*hmm_acc)`%**, essentially the same as the
mixture model (`r mix_acc`%). The HMM learns nothing beyond the mixture:
the two-lactate-population structure is a property of the **patients**,
not of the order in which they were recorded.

# Predictive Validation, Calibration and Missingness {#validation}

The GLM in Section \@ref(glm) reported an AUC computed **on the same data
it was fitted to**, which is optimistically biased. A model intended for
*real-time* prediction must be judged on data it has not seen, must be
**calibrated** (predicted risks must match observed frequencies), and must
be honest about the **missing-data** mechanism that shaped the analysis
sample. This section addresses all three.

## Out-of-sample discrimination {#cv}

Because the cohort keeps only each patient's first ICU stay, every patient
contributes exactly one row, so an ordinary random split is already
patient-disjoint (no leakage between train and test). Ten-fold
cross-validation is used: each stay is scored by a model trained on the
other nine folds.

```{r cv-auc}
dat$y <- as.integer(dat$sepsis == "Sepsis-3")
form1 <- y ~ lactate_max
form2 <- y ~ lactate_max + wbc_max + hr_max + resp_rate_max + temp_max_c + age

set.seed(486649)
K <- 10
folds <- sample(rep(1:K, length.out = nrow(dat)))
cv_p1 <- cv_p2 <- numeric(nrow(dat))
for (k in 1:K) {
  tr <- folds != k
  cv_p1[!tr] <- predict(glm(form1, dat[tr, ], family = binomial),
                        newdata = dat[!tr, ], type = "response")
  cv_p2[!tr] <- predict(glm(form2, dat[tr, ], family = binomial),
                        newdata = dat[!tr, ], type = "response")
}

roc1 <- pROC::roc(dat$y, cv_p1, quiet = TRUE, direction = "<")
roc2 <- pROC::roc(dat$y, cv_p2, quiet = TRUE, direction = "<")
auc1 <- as.numeric(pROC::auc(roc1)); auc2 <- as.numeric(pROC::auc(roc2))
ci2  <- as.numeric(pROC::ci.auc(roc2))
# In-sample AUC of M2 (Mann-Whitney identity) for the optimism comparison.
prob2_is <- predict(m2, type = "response")
auc2_is <- {
  r <- rank(prob2_is); n1 <- sum(dat$y == 1); n0 <- sum(dat$y == 0)
  (sum(r[dat$y == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
delong <- pROC::roc.test(roc1, roc2, method = "delong")
```

```{r cv-auc-table}
kable(tibble(
  Model = c("M1: lactate only", "M2: six predictors"),
  `In-sample AUC` = c(sprintf("%.3f", {
      r <- rank(predict(m1, type="response")); n1 <- sum(dat$y==1); n0 <- sum(dat$y==0)
      (sum(r[dat$y==1]) - n1*(n1+1)/2)/(n1*n0)}),
    sprintf("%.3f", auc2_is)),
  `10-fold CV AUC` = c(sprintf("%.3f", auc1), sprintf("%.3f", auc2)),
  `95% CI (CV, M2)` = c("", sprintf("[%.3f, %.3f]", ci2[1], ci2[3]))
), booktabs = TRUE,
   caption = "In-sample versus cross-validated discrimination.") %>%
  kable_styling(latex_options = "HOLD_position")
```

Table \@ref(tab:cv-auc-table) shows the cross-validated AUC of the
six-predictor model is **`r sprintf('%.3f', auc2)`**
(95% CI `r sprintf('%.3f to %.3f', ci2[1], ci2[3])`), essentially unchanged
from the in-sample **`r sprintf('%.3f', auc2_is)`**; with
`r format(nrow(dat), big.mark=',')` observations and only six predictors
the model does not overfit. A DeLong test confirms that the six-predictor
model discriminates better than lactate alone
(AUC `r sprintf('%.3f', auc2)` vs `r sprintf('%.3f', auc1)`,
*p* `r format.pval(delong$p.value, digits=2, eps=1e-16)`).

## Calibration and clinical usefulness {#calibration}

Discrimination (ranking) is not the same as calibration (the numerical
risk being correct). The cross-validated risks are grouped into deciles
and the mean predicted risk compared with the observed sepsis rate.

```{r calibration, fig.cap = "Calibration of the cross-validated six-predictor model (risk deciles). Points on the diagonal indicate predicted risks that match observed frequencies.", fig.width=4.4, fig.height=4}
brier <- mean((cv_p2 - dat$y)^2)
brier_ref <- mean((mean(dat$y) - dat$y)^2)   # null model (predict prevalence)
# Calibration slope/intercept via logistic recalibration on the linear predictor.
lp <- qlogis(pmin(pmax(cv_p2, 1e-6), 1 - 1e-6))
recal <- glm(dat$y ~ lp, family = binomial)
cal_slope <- coef(recal)[2]

cal_tab <- tibble(p = cv_p2, y = dat$y) %>%
  mutate(bin = ntile(p, 10)) %>%
  group_by(bin) %>%
  summarise(pred = mean(p), obs = mean(y), n = n(), .groups = "drop")

ggplot(cal_tab, aes(pred, obs)) +
  geom_abline(linetype = "dashed", colour = "grey55") +
  geom_line(colour = "#C0504D") +
  geom_point(colour = "#C0504D", size = 1.6) +
  coord_equal(xlim = c(0, max(cal_tab$pred, cal_tab$obs)),
              ylim = c(0, max(cal_tab$pred, cal_tab$obs))) +
  labs(x = "Mean predicted risk", y = "Observed sepsis rate")
```

The **Brier score** is **`r sprintf('%.3f', brier)`** (lower is better),
against **`r sprintf('%.3f', brier_ref)`** for a null model that always
predicts the prevalence, a
`r sprintf('%.0f%%', 100*(1 - brier/brier_ref))` reduction. The
**calibration slope** is **`r sprintf('%.2f', cal_slope)`** (1.0 is
perfect); Figure \@ref(fig:calibration) shows the decile points lie close
to the diagonal, so the predicted probabilities are usable as risks, not
merely as a ranking.

Because sepsis is imbalanced and the costs of missing it are asymmetric,
ROC-AUC is supplemented with the **precision-recall AUC** and a
**decision-curve (net-benefit)** analysis.

```{r pr-decision, fig.cap = "Left: precision-recall curve (dashed line = prevalence, the no-skill baseline). Right: decision curve showing net benefit of the model against treat-all and treat-none across risk thresholds.", fig.width=7, fig.height=3.2}
pr <- PRROC::pr.curve(scores.class0 = cv_p2[dat$y == 1],
                      scores.class1 = cv_p2[dat$y == 0], curve = TRUE)
prauc <- pr$auc.integral
prev  <- mean(dat$y)

pr_df <- as.data.frame(pr$curve)[, 1:2]; names(pr_df) <- c("recall", "precision")
p_pr <- ggplot(pr_df, aes(recall, precision)) +
  geom_line(colour = "#C0504D") +
  geom_hline(yintercept = prev, linetype = "dashed", colour = "grey55") +
  labs(x = "Recall", y = "Precision",
       subtitle = sprintf("PR-AUC = %.3f (baseline %.3f)", prauc, prev))

thr <- seq(0.02, 0.60, by = 0.02)
nb_model <- sapply(thr, function(pt) {
  pp <- cv_p2 >= pt
  mean(pp & dat$y == 1) - mean(pp & dat$y == 0) * (pt / (1 - pt))
})
nb_all <- sapply(thr, function(pt) prev - (1 - prev) * (pt / (1 - pt)))
dca <- bind_rows(
  tibble(thr, nb = nb_model, strat = "Model"),
  tibble(thr, nb = nb_all,   strat = "Treat all"),
  tibble(thr, nb = 0,        strat = "Treat none"))
p_dca <- ggplot(dca, aes(thr, nb, colour = strat)) +
  geom_line() +
  scale_colour_manual(values = c("Model" = "#C0504D", "Treat all" = "#5B9BD5",
                                 "Treat none" = "grey55")) +
  coord_cartesian(ylim = c(min(nb_model) - 0.01, max(nb_model) + 0.02)) +
  labs(x = "Risk threshold", y = "Net benefit", colour = NULL) +
  theme(legend.position = "top")

p_pr + p_dca
```

The PR-AUC is **`r sprintf('%.3f', prauc)`** against a no-skill baseline of
**`r sprintf('%.3f', prev)`** (the prevalence), and Figure
\@ref(fig:pr-decision) shows the model yields positive net benefit above
the treat-all and treat-none strategies across the clinically relevant
range of thresholds, confirming that the discrimination translates into
decision value.

## Non-linearity of the lactate effect {#splines}

The GLM assumed lactate acts **linearly on the log-odds** scale. A
penalised thin-plate spline `s(lactate_max)` tests this assumption
against a flexible alternative.

```{r gam, fig.cap = "Penalised-spline estimate of the log-odds of sepsis as a function of maximum lactate (shaded 95\\% band). Curvature indicates a non-linear effect the linear GLM cannot capture.", fig.width=4.6, fig.height=3.2}
g_lin <- glm(y ~ lactate_max, data = dat, family = binomial)
g_gam <- mgcv::gam(y ~ s(lactate_max, k = 10), data = dat, family = binomial)
edf   <- summary(g_gam)$edf
d_aic <- AIC(g_lin) - AIC(g_gam)

nd <- tibble(lactate_max = seq(min(dat$lactate_max), quantile(dat$lactate_max, 0.99),
                               length.out = 200))
pr_g <- predict(g_gam, nd, se.fit = TRUE)
nd <- nd %>% mutate(fit = pr_g$fit, lo = fit - 1.96 * pr_g$se.fit,
                    hi = fit + 1.96 * pr_g$se.fit)
ggplot(nd, aes(lactate_max, fit)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#C0504D", alpha = 0.2) +
  geom_line(colour = "#C0504D") +
  labs(x = "Maximum lactate (mmol/L)", y = "log-odds of Sepsis-3")
```

The spline uses **`r sprintf('%.1f', edf)`** effective degrees of freedom
(1.0 would mean linear), and it lowers AIC by **`r sprintf('%.0f', d_aic)`**
relative to the linear term, so the lactate effect is materially
non-linear (Figure \@ref(fig:gam)): risk rises steeply at low lactate and
then flattens. A production model should therefore spline-transform
lactate rather than enter it linearly.

## Multicollinearity {#vif}

Section \@ref(glm) checked only three pairwise correlations.
Variance-inflation factors provide a more complete assessment across
all six predictors of Model 2.

```{r vif}
vif_tab <- tibble(Predictor = names(car::vif(m2)),
                  VIF = as.numeric(car::vif(m2))) %>%
  arrange(desc(VIF))
kable(vif_tab, booktabs = TRUE, digits = 2,
      caption = "Variance-inflation factors (VIF > 5 signals collinearity).") %>%
  kable_styling(latex_options = "HOLD_position")
```

All VIFs in Table \@ref(tab:vif) are near 1 (maximum
**`r sprintf('%.2f', max(vif_tab$VIF))`**), well below the usual cut-off of
5, so the coefficients are not destabilised by collinearity.

## The missing-data mechanism {#missingness}

Every inferential section so far used only the
`r format(n_model, big.mark=',')` stays with a **complete** set of the six
features, chiefly limited by lactate, which is *ordered at the
clinician's discretion*. Whether a lactate exists is therefore not random:
it is itself a marker of concern.

```{r missingness}
raw2 <- raw %>% mutate(lac_measured = !is.na(lactate_max),
                       sep = to_bool(sepsis3))
miss_rate <- mean(!raw2$lac_measured)
sep_meas  <- mean(raw2$sep[raw2$lac_measured])
sep_unmeas<- mean(raw2$sep[!raw2$lac_measured])
mtab <- table(`Lactate measured` = ifelse(raw2$lac_measured, "Yes", "No"),
              Sepsis = ifelse(raw2$sep, "Sepsis-3", "No sepsis"))
mchi <- chisq.test(mtab)
```

Lactate is missing in **`r sprintf('%.1f%%', 100*miss_rate)`** of the full
cohort. Sepsis-3 prevalence is **`r sprintf('%.1f%%', 100*sep_meas)`**
among stays with a measured lactate versus
**`r sprintf('%.1f%%', 100*sep_unmeas)`** among those without
(chi-squared *p* `r format.pval(mchi$p.value, digits=2, eps=1e-16)`),
so *the act of measuring lactate is itself strongly informative*, and
complete-case analysis enriches the sample for severity. Missingness is
plausibly **MAR given the recorded vitals** (sicker-looking patients get
labs), but a **MNAR** component cannot be excluded.

As a sensitivity analysis the six predictors are **multiply imputed**
(MICE, predictive mean matching) on the full cohort and the logistic model
re-fitted, pooling coefficients by Rubin's rules.

```{r mice, cache=FALSE}
mi_dat <- raw %>%
  mutate(y = as.integer(to_bool(sepsis3))) %>%
  dplyr::select(lactate_max, wbc_max, hr_max, resp_rate_max, temp_max_c, age, y)
imp <- mice(mi_dat, m = 5, maxit = 5, method = "pmm",
            printFlag = FALSE, seed = 486649)
fit_mi <- with(imp, glm(y ~ lactate_max + wbc_max + hr_max + resp_rate_max +
                          temp_max_c + age, family = binomial))
pooled <- summary(pool(fit_mi), conf.int = TRUE)

or_cc <- exp(coef(m2))
comp <- pooled %>%
  filter(term != "(Intercept)") %>%
  transmute(Predictor = term,
            `OR (complete-case)` = sprintf("%.3f", or_cc[term]),
            `OR (imputed cohort)` = sprintf("%.3f", exp(estimate)),
            `95% CI (imputed)` = sprintf("[%.3f, %.3f]",
                                         exp(`2.5 %`), exp(`97.5 %`)))
```

```{r mice-table}
kable(comp, booktabs = TRUE,
      caption = "Odds ratios: complete-case versus multiply-imputed full cohort.") %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
```

Table \@ref(tab:mice-table) shows the odds ratios keep the same direction
and comparable magnitude after imputing the full cohort: the associations
are not an artefact of the complete-case restriction, though the absolute
sepsis prevalence differs between the samples. This missingness signal is
itself a lesson for the real-time model: *whether and when a test is
ordered* is a usable predictor, not just a nuisance.

*Collectively, this section shows the six-predictor model generalises out
of sample, is well calibrated and clinically useful, benefits from a
non-linear lactate term, is free of collinearity, and is robust to the
missing-data mechanism, while also revealing that measurement itself
carries information.*

```{r export-values, include=FALSE}
# ── Export all key statistics and coefficients for thesis / downstream use ──

# 1. Hypothesis tests
hyp_tests <- data.frame(
  test = c("welch_t", "welch_t", "welch_t", "welch_t", "welch_t", "welch_t", "welch_t",
           "wilcoxon", "wilcoxon", "wilcoxon",
           "chisq", "chisq", "chisq", "chisq"),
  metric = c("t_statistic", "df", "p_value", "mean_no_sepsis", "mean_sepsis",
             "cohens_d", "cohens_d_ci",
             "W", "p_value", "rank_biserial_r",
             "chi_squared", "df", "p_value", "cramers_v_adjusted"),
  value = c(sprintf("%.4f", t1$statistic), sprintf("%.1f", t1$parameter),
            sprintf("%.2e", t1$p.value), sprintf("%.4f", t1$estimate[1]),
            sprintf("%.4f", t1$estimate[2]),
            sprintf("%.4f", d1$Cohens_d),
            sprintf("[%.4f, %.4f]", d1$CI_low, d1$CI_high),
            format(w1$statistic, big.mark=""), sprintf("%.2e", w1$p.value),
            sprintf("%.4f", rb$r_rank_biserial),
            sprintf("%.2f", c1$statistic), as.character(c1$parameter),
            sprintf("%.2e", c1$p.value), sprintf("%.4f", v1$Cramers_v_adjusted)),
  stringsAsFactors = FALSE
)
write.csv(hyp_tests, file.path(OUT, "06_hypothesis_tests.csv"), row.names = FALSE)

# 2. MLE parameters
mle_tab <- data.frame(
  parameter = c("mu_hat", "sigma_hat", "se_mu", "n_septic",
                "median_lactate", "median_ci_low", "median_ci_high"),
  value = c(mu_hat, sigma_hat, se_mu, n_s, median_lactate,
            exp(mu_hat - 1.96 * se_mu), exp(mu_hat + 1.96 * se_mu)),
  stringsAsFactors = FALSE
)
write.csv(mle_tab, file.path(OUT, "06_mle_params.csv"), row.names = FALSE)

# 3. GLM coefficients and odds ratios
glm_export <- bind_rows(
  broom::tidy(m1) %>% mutate(model = "M1_lactate_only"),
  broom::tidy(m2) %>% mutate(model = "M2_six_predictors")
)
glm_export$odds_ratio <- exp(glm_export$estimate)
or_ci_full <- rbind(exp(confint.default(m1)), exp(confint.default(m2)))
glm_export$or_ci_low  <- or_ci_full[, 1]
glm_export$or_ci_high <- or_ci_full[, 2]
write.csv(glm_export, file.path(OUT, "06_glm_coefficients.csv"), row.names = FALSE)

# 4. Model comparison and discrimination
model_comp <- data.frame(
  metric = c("aic_m1", "aic_m2", "deviance_m1", "deviance_m2",
             "insample_auc_m1", "insample_auc_m2",
             "cv_auc_m1", "cv_auc_m2", "cv_auc_m2_ci_low", "cv_auc_m2_ci_high",
             "delong_p_m1_vs_m2",
             "brier_score", "brier_null", "brier_reduction_pct", "calibration_slope",
             "pr_auc", "prevalence",
             "gam_edf", "gam_aic_improvement", "max_vif",
             "missingness_rate", "sep_prev_measured", "sep_prev_unmeasured",
             "missingness_chisq_p"),
  value = c(AIC(m1), AIC(m2), m1$deviance, m2$deviance,
            {r1<-rank(predict(m1,type="response")); n1<-sum(dat$y==1); n0<-sum(dat$y==0);
             (sum(r1[dat$y==1])-n1*(n1+1)/2)/(n1*n0)}, auc2_is,
            auc1, auc2, ci2[1], ci2[3],
            delong$p.value,
            brier, brier_ref, 100*(1-brier/brier_ref), cal_slope,
            prauc, prev,
            edf, d_aic, max(vif_tab$VIF),
            miss_rate, sep_meas, sep_unmeas, mchi$p.value),
  stringsAsFactors = FALSE
)
write.csv(model_comp, file.path(OUT, "06_model_metrics.csv"), row.names = FALSE)

# 5. VIF table
write.csv(vif_tab, file.path(OUT, "06_vif.csv"), row.names = FALSE)

# 6. Mixture model parameters
mix_export <- data.frame(
  parameter = c("pi1", "mu1_log", "mu2_log", "sigma1_log", "sigma2_log",
                "mu1_mmol", "mu2_mmol", "unsupervised_accuracy_pct"),
  value = c(pi1, mu1, mu2, s1, s2, mix_means[1], mix_means[2], mix_acc),
  stringsAsFactors = FALSE
)
write.csv(mix_export, file.path(OUT, "06_mixture_params.csv"), row.names = FALSE)

# 7. MCMC posterior summary
mcmc_export <- data.frame(
  parameter = c("posterior_mean", "posterior_sd", "ci_025", "ci_975",
                "implied_median_lactate", "gelman_rubin_rhat",
                "effective_sample_size", "accept_rate_chain1", "accept_rate_chain2"),
  value = c(post_mean, post_sd, post_ci[1], post_ci[2],
            exp(post_mean), rhat, ess, run1$accept_rate, run2$accept_rate),
  stringsAsFactors = FALSE
)
write.csv(mcmc_export, file.path(OUT, "06_mcmc_posterior.csv"), row.names = FALSE)

# 8. HMM parameters
hmm_export <- data.frame(
  parameter = c("best_k", "bic_1state", "bic_2state", "bic_3state",
                "emission_mu1", "emission_mu2", "emission_sd1", "emission_sd2",
                "trans_11", "trans_12", "trans_21", "trans_22",
                "stationary_1", "stationary_2",
                "unsupervised_accuracy_pct"),
  value = c(best_k, bic_vals[1], bic_vals[2], bic_vals[3],
            mu_h1, mu_h2, sd_h1, sd_h2,
            trans_mat[1,1], trans_mat[1,2], trans_mat[2,1], trans_mat[2,2],
            stat[1], stat[2], 100*hmm_acc),
  stringsAsFactors = FALSE
)
write.csv(hmm_export, file.path(OUT, "06_hmm_params.csv"), row.names = FALSE)

# 9. MICE imputed odds ratios
write.csv(comp, file.path(OUT, "06_mice_odds_ratios.csv"), row.names = FALSE)

# 10. Summary statistics table
write.csv(as.data.frame(summary_tab), file.path(OUT, "06_summary_stats.csv"), row.names = FALSE)

# 11. Calibration data
write.csv(as.data.frame(cal_tab), file.path(OUT, "06_calibration_data.csv"), row.names = FALSE)
```

```{r export-figures, include=FALSE}
# ── Export all figures as publication-quality PNGs ──

# Fig 1: EDA boxplots
png(file.path(FIG, "06_eda_boxplots.png"), width = 1200, height = 600, res = 200)
print(
  dat %>%
    dplyr::select(sepsis, all_of(c("lactate_max","resp_rate_max","hr_max"))) %>%
    pivot_longer(-sepsis, names_to = "Feature", values_to = "Value") %>%
    mutate(Feature = c(lactate_max="Lactate, max", resp_rate_max="Resp. rate, max",
                       hr_max="Heart rate, max")[Feature]) %>%
    ggplot(aes(sepsis, Value, fill = sepsis)) +
    geom_boxplot(alpha=0.75, outlier.size=0.4, outlier.alpha=0.25) +
    facet_wrap(~Feature, scales="free_y", nrow=1) +
    scale_fill_manual(values=PAL) + labs(x=NULL, y=NULL) +
    theme(legend.position="none")
)
dev.off()

# Fig 2: Lactate density
png(file.path(FIG, "06_lactate_density.png"), width = 1000, height = 700, res = 200)
print(
  dat %>% mutate(lac = pmin(lactate_max, 15)) %>%
    ggplot(aes(lac, fill = sepsis)) + geom_density(alpha = 0.55) +
    scale_fill_manual(values = PAL) +
    labs(x = "Maximum lactate (mmol/L)", y = "Density", fill = NULL) +
    theme(legend.position = "top")
)
dev.off()

# Fig 3: KDE
png(file.path(FIG, "06_kde_loglactate.png"), width = 1000, height = 700, res = 200)
print(
  ggplot(tibble(loglac = loglac), aes(loglac)) +
    geom_histogram(aes(y = after_stat(density)), bins=40, fill="grey85", colour="grey60") +
    geom_density(bw=bw_sil, colour="#C0504D", linewidth=0.9) +
    labs(x = "log(maximum lactate)", y = "Density",
         subtitle = sprintf("Gaussian kernel, bandwidth = %.3f", bw_sil))
)
dev.off()

# Fig 4: Gaussian mixture
png(file.path(FIG, "06_mixture_fit.png"), width = 1000, height = 700, res = 200)
print(
  ggplot(tibble(loglac = loglac), aes(loglac)) +
    geom_histogram(aes(y = after_stat(density)), bins=40, fill="grey88", colour="grey60") +
    geom_line(data = mix_df, aes(x, comp1), linetype="dashed", colour="#5B9BD5") +
    geom_line(data = mix_df, aes(x, comp2), linetype="dashed", colour="#C0504D") +
    geom_line(data = mix_df, aes(x, total), colour="black", linewidth=0.7) +
    labs(x = "log(maximum lactate)", y = "Density")
)
dev.off()

# Fig 5: MCMC trace + posterior
png(file.path(FIG, "06_mcmc_diagnostics.png"), width = 1400, height = 600, res = 200)
print(p_trace + p_post)
dev.off()

# Fig 6: Calibration
png(file.path(FIG, "06_calibration.png"), width = 900, height = 900, res = 200)
print(
  ggplot(cal_tab, aes(pred, obs)) +
    geom_abline(linetype = "dashed", colour = "grey55") +
    geom_line(colour = "#C0504D") + geom_point(colour = "#C0504D", size = 1.6) +
    coord_equal(xlim = c(0, max(cal_tab$pred, cal_tab$obs)),
                ylim = c(0, max(cal_tab$pred, cal_tab$obs))) +
    labs(x = "Mean predicted risk", y = "Observed sepsis rate")
)
dev.off()

# Fig 7: PR + decision curve
png(file.path(FIG, "06_pr_decision_curve.png"), width = 1400, height = 650, res = 200)
print(p_pr + p_dca)
dev.off()

# Fig 8: GAM spline
png(file.path(FIG, "06_gam_spline.png"), width = 1000, height = 700, res = 200)
nd_gam <- tibble(lactate_max = seq(min(dat$lactate_max), quantile(dat$lactate_max, 0.99),
                                   length.out = 200))
pr_gam <- predict(g_gam, nd_gam, se.fit = TRUE)
nd_gam <- nd_gam %>% mutate(fit = pr_gam$fit, lo = fit - 1.96*pr_gam$se.fit,
                             hi = fit + 1.96*pr_gam$se.fit)
print(
  ggplot(nd_gam, aes(lactate_max, fit)) +
    geom_ribbon(aes(ymin=lo, ymax=hi), fill="#C0504D", alpha=0.2) +
    geom_line(colour="#C0504D") +
    labs(x = "Maximum lactate (mmol/L)", y = "log-odds of Sepsis-3")
)
dev.off()
```

# Conclusion {#conclusion}

This report asked whether routine physiology that is *not* part of the
sepsis definition can identify Sepsis-3, and what different statistical
methods reveal about the same data.

-   **Hypothesis tests** (Section \@ref(hypothesis)) found genuine
    differences between septic and non-septic stays, largest for lactate
    (Cohen's *d* = `r sprintf('%.2f', d1$Cohens_d)`), and weaker but
    real for white-cell count and tachypnoea.

-   The **GLM** (Section \@ref(glm)) showed lactate is the single
    strongest predictor, while vital signs and age add independent
    information: AIC fell from `r sprintf('%.0f', AIC(m1))` to
    `r sprintf('%.0f', AIC(m2))`, with an approximate AUC of
    `r sprintf('%.2f', auc)`.

-   The **mixture model** (Section \@ref(mixture-section)) and **hidden
    Markov model** (Section \@ref(partB)) both recovered two lactate
    populations without labels, agreeing almost exactly with each other
    (`r mix_acc`% vs `r sprintf('%.1f', 100*hmm_acc)`%). Crucially,
    lactate alone only partially recovers sepsis, a reminder that the
    diagnosis is multi-dimensional.

-   The **MLE** (Section \@ref(mle)) and **MCMC** (Section \@ref(partA))
    agreed closely on the typical septic lactate ($\approx$
    `r sprintf('%.2f', median_lactate)` mmol/L), with nearly identical
    interval estimates.

-   **Validation** (Section \@ref(validation)) confirmed the six-predictor
    model holds up out of sample (10-fold CV AUC
    `r sprintf('%.3f', auc2)`, DeLong *p*
    `r format.pval(delong$p.value, digits=2, eps=1e-16)` versus lactate
    alone), is well calibrated (Brier `r sprintf('%.3f', brier)`,
    slope `r sprintf('%.2f', cal_slope)`), gains from a non-linear lactate
    term, and keeps its associations under multiple imputation, while also
    showing that *whether* a lactate was ordered is itself predictive.

*The methodological message echoes across tools: several independent
methods converge on the same, clinically coherent picture. No single
physiological variable identifies sepsis on its own, which is precisely
why Sepsis-3 couples suspected infection with a multi-organ SOFA score.
The geometry of one biomarker is not enough; it takes a combination of
signals to recognise a dysregulated response to infection.*

**Limitations and the path to real-time prediction.** The analysis subset
is restricted to stays with a measured lactate, which enriches for
severity (quantified and partly addressed by the multiple-imputation
sensitivity analysis in Section \@ref(missingness)); sepsis was
operationalised with a first-24h ("day-1") SOFA and a small set of
documented simplifications (Section 05 header); and the ordering-based HMM
is illustrative, since stays are not a true time series. Most importantly,
this report is **cross-sectional**: it collapses each stay into one row of
first-24h summaries and answers *"was this stay septic?"* rather than the
project's ultimate question, *"will this patient become septic in the next
few hours, given the data so far?"* That real-time question requires an
**hourly patient panel**, a defined **sepsis-onset time** and prediction
horizon, **survival / time-varying-hazard** modelling of time to onset, a
**within-patient** (not ordering-based) HMM, and **time-dependent,
leakage-safe** evaluation benchmarked against streaming clinical scores
(qSOFA, SIRS, hourly SOFA, MEWS). Building that longitudinal foundation is
the subject of the follow-on work, for which the associations, effect
sizes, non-linearities, and missingness structure established here provide
the modelling priors.

# Acknowledgment {.unnumbered}

During this project, I applied the **PAIR framework** to responsibly use
generative AI alongside primary sources.

-   **Problem:** Build a reproducible MIMIC-IV Sepsis-3 pipeline and a
    layered statistical analysis (EDA, hypothesis testing, MLE, GLMs,
    density and mixture modelling, Metropolis-Hastings sampling, and a
    hidden Markov model) in R, supported by a coherent written report.

-   **AI:** I used AI assistants to discuss implementation choices,
    clarify statistical concepts, and refine the writing.

-   **Interaction:** All AI-generated suggestions were critically
    reviewed, checked against primary sources (Singer et al., 2016;
    Seymour et al., 2016; Vincent et al., 1996; Silverman, 1986; Robert
    & Casella, 2004; Gelman et al., 2013; Rabiner, 1989; Visser &
    Speekenbrink, 2010), and adapted to the structure and seed-based
    reproducibility of this work.

-   **Reflection:** AI assistance accelerated drafting and improved
    clarity, while the interpretation, code, model choices, and
    conclusions remain my own.

# References {.unnumbered}

1.  Singer, M., Deutschman, C. S., Seymour, C. W., et al. (2016). The
    Third International Consensus Definitions for Sepsis and Septic
    Shock (Sepsis-3). *JAMA*, 315(8), 801-810.

2.  Seymour, C. W., Liu, V. X., Iwashyna, T. J., et al. (2016).
    Assessment of Clinical Criteria for Sepsis (Sepsis-3). *JAMA*,
    315(8), 762-774.

3.  Vincent, J. L., Moreno, R., Takala, J., et al. (1996). The SOFA
    (Sepsis-related Organ Failure Assessment) score to describe organ
    dysfunction/failure. *Intensive Care Medicine*, 22(7), 707-710.

4.  Johnson, A. E. W., Bulgarelli, L., Shen, L., et al. (2023).
    MIMIC-IV, a freely accessible electronic health record dataset.
    *Scientific Data*, 10,

    1.  

5.  Cohen, J. (1988). *Statistical Power Analysis for the Behavioral
    Sciences* (2nd ed.). Lawrence Erlbaum Associates.

6.  Silverman, B. W. (1986). *Density Estimation for Statistics and Data
    Analysis.* Chapman and Hall.

7.  Dempster, A. P., Laird, N. M., & Rubin, D. B. (1977). Maximum
    likelihood from incomplete data via the EM algorithm. *JRSS B*,
    39(1), 1-38.

8.  Gelman, A., Carlin, J. B., Stern, H. S., et al. (2013). *Bayesian
    Data Analysis* (3rd ed.). CRC Press.

9.  Robert, C., & Casella, G. (2004). *Monte Carlo Statistical Methods*
    (2nd ed.). Springer.

10. Rabiner, L. R. (1989). A tutorial on hidden Markov models and
    selected applications in speech recognition. *Proceedings of the
    IEEE*, 77(2), 257-286.

11. Visser, I., & Speekenbrink, M. (2010). depmixS4: An R package for
    hidden Markov models. *Journal of Statistical Software*, 36(7),
    1-21. \`\`\`
