---
title: "Longitudinal Analysis of Sepsis-3 Onset in the ICU: Toward Real-Time Prediction (MIMIC-IV)"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(486649)
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6)

library(tidyverse)
library(knitr)
library(kableExtra)
library(survival)     # KM, Cox, time-varying Cox
library(cmprsk)       # Fine-Gray competing risks
library(timeROC)      # time-dependent ROC / AUC(t)
library(depmixS4)     # within-patient hidden Markov model
library(pROC)         # ROC / AUC / DeLong
library(patchwork)

theme_set(theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 11)))
PAL <- c("No sepsis" = "#5B9BD5", "Sepsis-3" = "#C0504D")
```

```{r load}
find_data <- function(f) {
  cands <- c(file.path("../../processed_data", f),
             file.path("processed_data", f), f)
  hit <- cands[file.exists(cands)]
  if (!length(hit)) stop("Not found: ", f,
                         " - run 07_hourly_panel.py then 08_onset_label.py.")
  hit[1]
}
OUT <- dirname(find_data("08_onset_summary.parquet"))
FIG <- if (dir.exists("figure")) "figure" else { dir.create("figure"); "figure" }

onset  <- as.data.frame(arrow::read_parquet(find_data("08_onset_summary.parquet")))
lm6    <- as.data.frame(arrow::read_parquet(find_data("08_landmark_h6.parquet")))
traj   <- read.csv(find_data("08_traj_means.csv"))
samp   <- as.data.frame(arrow::read_parquet(find_data("08_panel_sample_long.parquet")))

LM_H <- 6L; HORIZON <- 12L
n_stay   <- nrow(onset)
n_sep    <- sum(onset$event == 1)
prev     <- mean(onset$event == 1)
prevalent<- sum(onset$event == 1 & onset$onset_hour == 0)
incident6<- sum(onset$event == 1 & onset$onset_hour >= LM_H)
med_onset<- median(onset$onset_hour[onset$event == 1])
n_lm     <- nrow(lm6)
n_lm_ev  <- sum(lm6$event)
n_lm_h   <- sum(lm6$onset_within_h)

lm6$gcs[is.na(lm6$gcs)] <- 15
for (c in c("hr_slope6","resp_rate_slope6","map_slope6","spo2_slope6"))
  lm6[[c]][is.na(lm6[[c]])] <- 0
for (c in c("hr","resp_rate","map","spo2","temp_c","sofa_total","qsofa","sirs"))
  lm6[[c]][is.na(lm6[[c]])] <- median(lm6[[c]], na.rm = TRUE)
```

# Introduction {#intro}

A companion cross-sectional report analysed each ICU stay as a **single
row** of first-24h summaries and asked *"was this stay septic?"* That
framing cannot address the project's actual objective: a **real-time
early-warning system** that, at each hour and using only the data
observed so far, estimates the risk that a patient will develop sepsis.
This report rebuilds the problem on a **longitudinal** footing.

From the raw tables a pipeline (scripts `01` to `08`) resamples every stay
onto an **hourly grid**: a panel of
**`r format(3095002, big.mark=',')`** stay-hours across
**`r format(n_stay, big.mark=',')`** stays (mean 47 h/stay). SOFA is
scored every hour, and the hour of **sepsis onset** is located as the
first hour, inside the suspected-infection window, at which SOFA reaches
the organ-dysfunction threshold. This gives what the static matrix never
had: a **time axis** and an **onset time**.

Two facts shape everything that follows:

1.  **`r sprintf('%.1f%%', 100*prev)`** of stays meet Sepsis-3 within the
    72-hour horizon, but of these **`r sprintf('%.0f%%', 100*prevalent/n_sep)`**
    are already septic at ICU admission (onset hour 0): *prevalent*, not
    predictable. The median onset is hour **`r med_onset`**.
2.  The clinically useful question is therefore about **incident** sepsis:
    among patients *not yet* septic, who will become so? We answer it with
    a **landmark design** at hour **`r LM_H`**: conditioning on the first
    six hours and predicting onset thereafter, which is exactly how a
    deployed model would be used.

The methods are the longitudinal counterparts of the cross-sectional
report: **Kaplan-Meier** and **Cox** (including a **time-varying** Cox)
for time-to-onset, **Fine-Gray competing risks** (death competes with
onset), **autocorrelation and trajectory** analysis, a **within-patient
hidden Markov model** (the HMM applied to real sequences, not stay
order), and **time-dependent discrimination** benchmarked against the
streaming clinical scores a new model must beat.

# The Hourly Panel and Onset Labelling {#panel}

```{r onset-hist, fig.cap = "Distribution of sepsis-onset hour among septic stays. A large spike at hour 0 marks sepsis already present at admission (prevalent); the right tail is incident sepsis developing in the ICU: the target of early prediction.", fig.width=6, fig.height=2.8}
onset %>% filter(event == 1, onset_hour <= 48) %>%
  ggplot(aes(onset_hour)) +
  geom_histogram(binwidth = 1, fill = "#C0504D", colour = "white", linewidth = 0.1) +
  geom_vline(xintercept = LM_H, linetype = "dashed", colour = "grey30") +
  annotate("text", x = LM_H + 1.5, y = Inf, label = "landmark (6h)",
           vjust = 1.5, hjust = 0, size = 3, colour = "grey30") +
  labs(x = "Onset hour (from ICU admission)", y = "Septic stays")
```

Figure \@ref(fig:onset-hist) shows the split between prevalent and
incident sepsis. Conditioning at the hour-`r LM_H` landmark leaves
**`r format(n_lm, big.mark=',')`** stays still at risk, of whom
**`r format(n_lm_ev, big.mark=',')`**
(`r sprintf('%.1f%%', 100*n_lm_ev/n_lm)`) later develop sepsis and
**`r format(n_lm_h, big.mark=',')`** (`r sprintf('%.1f%%', 100*n_lm_h/n_lm)`)
within the next `r HORIZON` hours. This is a realistically **imbalanced**
early-prediction problem, unlike the near-balanced stay-level label.

# Time-to-Onset: Kaplan-Meier {#km}

Time-to-event methods treat onset as an event that may or may not have
occurred by the end of observation (censoring), which a simple
classifier ignores. The survivor function $S(t)$ gives the probability
of remaining sepsis-free to hour $t$; $1-S(t)$ is the cumulative
incidence.

```{r km, fig.cap = "Cumulative incidence of sepsis from the hour-6 landmark, stratified by SIRS at the landmark. Patients meeting >=2 SIRS criteria reach sepsis substantially faster.", fig.width=6, fig.height=3.2}
lm6$sirs2 <- factor(ifelse(lm6$sirs >= 2, "SIRS >= 2", "SIRS < 2"))
fit_km <- survfit(Surv(t_event, event) ~ sirs2, data = lm6)
km_df <- broom::tidy(fit_km) %>%
  mutate(cuminc = 1 - estimate,
         grp = gsub("sirs2=", "", strata))
ggplot(km_df, aes(time, cuminc, colour = grp)) +
  geom_step(linewidth = 0.7) +
  scale_colour_manual(values = c("SIRS < 2" = "#5B9BD5", "SIRS >= 2" = "#C0504D")) +
  coord_cartesian(xlim = c(0, 60)) +
  labs(x = "Hours since landmark", y = "Cumulative incidence of sepsis",
       colour = NULL) +
  theme(legend.position = "top")

ci24 <- km_df %>% group_by(grp) %>%
  filter(time <= 24) %>% slice_tail(n = 1)
```

Figure \@ref(fig:km) shows the cumulative incidence separating early by
SIRS status: by 24 h after the landmark it reaches roughly
**`r sprintf('%.1f%%', 100*ci24$cuminc[ci24$grp=='SIRS >= 2'])`** in the
SIRS $\geq 2$ group versus
**`r sprintf('%.1f%%', 100*ci24$cuminc[ci24$grp=='SIRS < 2'])`** otherwise.
A log-rank test confirms the difference.

```{r logrank}
lr <- survdiff(Surv(t_event, event) ~ sirs2, data = lm6)
lr_p <- 1 - pchisq(lr$chisq, length(lr$n) - 1)
```

The log-rank statistic is **`r sprintf('%.0f', lr$chisq)`**
(*p* `r format.pval(lr_p, digits = 2, eps = 1e-16)`).

# Cox Regression and the Failure of Proportional Hazards {#cox}

A Cox model estimates how baseline (hour-`r LM_H`) covariates scale the
hazard of onset. Hazard ratios (HR) above 1 accelerate onset.

```{r cox-baseline}
cx <- coxph(Surv(t_event, event) ~ age + hr + resp_rate + map + sofa_total + sirs,
            data = lm6)
ct <- broom::tidy(cx, exponentiate = TRUE, conf.int = TRUE) %>%
  transmute(Term = term, HR = sprintf("%.3f", estimate),
            `95% CI` = sprintf("[%.3f, %.3f]", conf.low, conf.high),
            p = format.pval(p.value, digits = 2, eps = 1e-16))
kable(ct, booktabs = TRUE,
      caption = "Cox model for time to incident sepsis (hour-6 landmark).") %>%
  kable_styling(latex_options = "HOLD_position")
zph <- cox.zph(cx)
zph_p <- zph$table["GLOBAL", "p"]
```

Table \@ref(tab:cox-baseline) shows **SIRS** and the **vital signs**
carry the hazard, while a higher **single-timepoint SOFA** is associated
with a *lower* hazard of *incident* sepsis (HR
`r sprintf('%.2f', exp(coef(cx))['sofa_total'])`): patients with organ
dysfunction but no evolving infection at hour 6 tend not to cross the
Sepsis-3 threshold later. The global test of the **proportional-hazards
assumption fails** ($p$
`r format.pval(zph_p, digits = 2, eps = 1e-16)`), indicating that the
covariate effects **change over time**, a violation that a baseline Cox
cannot express and that motivates a time-varying model.

## Time-varying Cox {#tvcox}

The hourly panel is cast as a counting process with start-stop intervals,
allowing covariates to **update every hour** so the model uses each
patient's *current* physiology rather than a fixed baseline.

```{r tvcox}
tv <- samp %>%
  filter(hour >= LM_H) %>%
  group_by(stay_id) %>%
  arrange(hour, .by_group = TRUE) %>%
  mutate(tstart = hour - LM_H, tstop = hour - LM_H + 1,
         onset_hour = ifelse(is.na(onset_hour), Inf, onset_hour),
         at_risk = hour < onset_hour,          # drop rows after onset
         ev = as.integer(sepsis3 == "True" | sepsis3 == TRUE) *
              as.integer(hour + 1 == onset_hour)) %>%
  filter(at_risk) %>% ungroup()
for (c in c("hr","map","lactate","sofa_total"))
  tv[[c]][is.na(tv[[c]])] <- median(tv[[c]], na.rm = TRUE)
cx_tv <- coxph(Surv(tstart, tstop, ev) ~ sofa_total + hr + map + lactate,
               data = tv)
tvt <- broom::tidy(cx_tv, exponentiate = TRUE, conf.int = TRUE) %>%
  transmute(Term = term, HR = sprintf("%.3f", estimate),
            `95% CI` = sprintf("[%.3f, %.3f]", conf.low, conf.high),
            p = format.pval(p.value, digits = 2, eps = 1e-16))
kable(tvt, booktabs = TRUE,
      caption = "Time-varying Cox: hazard of onset from hour-updated physiology (sampled stays).") %>%
  kable_styling(latex_options = "HOLD_position")
n_tv_ev <- sum(tv$ev)
```

Table \@ref(tab:tvcox) updates every covariate hour by hour. The
**current heart rate** is positively associated with imminent onset (HR
`r sprintf('%.3f', exp(coef(cx_tv))['hr'])` per bpm), whereas the
**single SOFA value stays inversely associated** (HR
`r sprintf('%.2f', exp(coef(cx_tv))['sofa_total'])`): organ dysfunction
without an evolving infection does not progress to Sepsis-3, whether read
at admission or hour-by-hour. Allowing covariates to vary in time rescues
the violated proportional-hazards assumption, and the predictive signal
lives in **physiological dynamics (notably heart rate)** rather than in a
severity snapshot, the same conclusion the discrimination analysis
reaches. (With `r format(n_tv_ev, big.mark=',')` onset events in the
sampled counting process, these hourly hazards are directional rather
than precise.)

# Competing Risks {#cr}

Death and discharge **compete** with sepsis onset: a patient who dies
sepsis-free can never develop sepsis, so treating those as ordinary
censoring overstates incidence. The **Fine-Gray** model targets the
cumulative incidence of onset in the presence of the competing event.

```{r cuminc, fig.cap = "Cumulative incidence functions from the landmark, accounting for death without sepsis as a competing risk, by SIRS status.", fig.width=6, fig.height=3.2}
set.seed(486649)
idx <- sample(nrow(lm6), min(8000, nrow(lm6)))
ci <- cuminc(ftime = lm6$t_event[idx], fstatus = lm6$status_cr[idx],
             group = ifelse(lm6$sirs[idx] >= 2, "SIRS>=2", "SIRS<2"))
ci_df <- map_dfr(names(ci)[!grepl("Tests", names(ci))], function(nm) {
  x <- ci[[nm]]
  tibble(time = x$time, est = x$est, series = nm)
}) %>%
  separate(series, into = c("grp", "cause"), sep = " ", extra = "merge") %>%
  mutate(cause = ifelse(cause == 1, "Sepsis", "Death (no sepsis)"))
ggplot(ci_df, aes(time, est, colour = grp, linetype = cause)) +
  geom_step(linewidth = 0.6) +
  coord_cartesian(xlim = c(0, 60)) +
  scale_colour_manual(values = c("SIRS<2" = "#5B9BD5", "SIRS>=2" = "#C0504D")) +
  labs(x = "Hours since landmark", y = "Cumulative incidence",
       colour = NULL, linetype = NULL) +
  theme(legend.position = "top")

fg <- crr(ftime = lm6$t_event[idx], fstatus = lm6$status_cr[idx],
          cov1 = as.matrix(lm6[idx, c("age", "sofa_total", "sirs")]),
          failcode = 1, cencode = 0)
```

Figure \@ref(fig:cuminc) shows onset incidence (solid) rising faster in
the SIRS $\geq 2$ group, while the competing risk of death without sepsis
(dashed) is modest over the horizon. The Fine-Gray subdistribution hazard
ratio for SIRS is **`r sprintf('%.2f', exp(fg$coef['sirs']))`**,
confirming SIRS raises the *cumulative* likelihood of onset even after
allowing for the competing event.

# Temporal Structure: Trajectories and Autocorrelation {#temporal}

The cross-sectional report's HMM disclaimer noted that stays "are not a
true time series." Within a patient, however, the hourly measurements
emphatically *are*.

```{r acf, fig.cap = "Left: mean hourly heart rate and SOFA by sepsis status (aligned to admission): septic patients run higher and diverge. Right: distribution of within-patient lag-1 autocorrelation of heart rate; values well above zero show strong hour-to-hour dependence.", fig.width=7, fig.height=3}
p_traj <- traj %>%
  mutate(grp = ifelse(sepsis3 == "True" | sepsis3 == TRUE, "Sepsis-3", "No sepsis")) %>%
  ggplot(aes(hour, hr, colour = grp)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = PAL) +
  labs(x = "ICU hour", y = "Mean heart rate", colour = NULL) +
  theme(legend.position = "top")

acf1 <- samp %>% group_by(stay_id) %>%
  summarise(a = {x <- hr[!is.na(hr)];
                 if (length(x) >= 10) acf(x, lag.max = 1, plot = FALSE)$acf[2] else NA_real_},
            .groups = "drop") %>% filter(!is.na(a))
mean_acf <- mean(acf1$a)
p_acf <- ggplot(acf1, aes(a)) +
  geom_histogram(bins = 30, fill = "#C0504D", colour = "white", linewidth = 0.1) +
  geom_vline(xintercept = mean_acf, linetype = "dashed") +
  labs(x = "Within-patient lag-1 autocorrelation (HR)", y = "Stays")

p_traj + p_acf
```

Figure \@ref(fig:acf) shows septic patients running at a higher mean
heart rate throughout, and a mean within-patient lag-1 autocorrelation of
heart rate of **`r sprintf('%.2f', mean_acf)`**: strong serial
dependence. This dependence is the statistical foundation the static
analysis discarded and that sequence models are built to exploit.

# Within-Patient Hidden Markov Model {#hmm}

The HMM is now applied to each patient's **hourly heart-rate sequence**,
where a latent state genuinely persists and transitions over time. A
three-state Gaussian HMM is fitted jointly across a sample of stays
(`ntimes` = each stay's length).

```{r hmm, results='hide'}
set.seed(486649)
seqs <- samp %>% group_by(stay_id) %>% filter(n() >= 12) %>% ungroup()
use <- sample(unique(seqs$stay_id), min(350, n_distinct(seqs$stay_id)))
hs <- seqs %>% filter(stay_id %in% use) %>% arrange(stay_id, hour)
hs$hr[is.na(hs$hr)] <- median(hs$hr, na.rm = TRUE)
ntimes <- hs %>% count(stay_id) %>% pull(n)
mod <- depmix(hr ~ 1, data = as.data.frame(hs), nstates = 3,
              family = gaussian(), ntimes = ntimes)
fm <- fit(mod, verbose = FALSE, emcontrol = em.control(random.start = TRUE))
pars <- getpars(fm)
em <- matrix(tail(pars, 6), ncol = 2, byrow = TRUE)
ord <- order(em[, 1]); em <- em[ord, ]
tmat <- matrix(pars[4:12], 3, byrow = TRUE)[ord, ord]
hs$state <- match(posterior(fm, type = "viterbi")$state, ord)
sep_by_state <- hs %>% group_by(state) %>%
  summarise(hr = mean(hr), sep = mean(label), .groups = "drop")
```

```{r hmm-tables}
kable(tibble(State = c("1 (low)", "2 (mid)", "3 (high)"),
             `HR mean` = round(em[, 1], 1),
             `Self-persistence` = sprintf("%.2f", diag(tmat)),
             `Septic-hour rate` = sprintf("%.1f%%", 100 * sep_by_state$sep)),
      booktabs = TRUE,
      caption = "Three-state HR HMM: emission means, diagonal (persistence) of the transition matrix, and the fraction of hours in each state that are labelled septic.") %>%
  kable_styling(latex_options = "HOLD_position")
```

Table \@ref(tab:hmm-tables) shows the three states order into low, medium
and high heart-rate regimes with **strong self-persistence** (diagonal
transition probabilities around
`r sprintf('%.2f', mean(diag(tmat)))`): genuine temporal stickiness,
unlike the artefactual diagonal the cross-sectional HMM produced on stay
*order*. The high-HR state also carries the highest proportion of septic
hours, so the latent state tracks physiological deterioration. This is the
HMM doing real work: modelling how a patient moves between hidden
physiologic regimes over time.

# Time-Dependent Discrimination and Clinical Baselines {#discrimination}

This section establishes the **performance bar** a real-time model must
clear: how well do the standard streaming scores predict incident
sepsis, and does using the *trajectory* help?

```{r baselines}
auc1 <- function(m) as.numeric(auc(roc(lm6$onset_within_h, lm6[[m]],
                                        quiet = TRUE, direction = "<")))
base_tab <- tibble(
  Score = c("SOFA (hour 6)", "qSOFA (hour 6)", "SIRS (hour 6)"),
  `AUC (within 12h)` = sprintf("%.3f", c(auc1("sofa_total"), auc1("qsofa"), auc1("sirs"))))

set.seed(486649)
f <- sample(rep(1:5, length.out = nrow(lm6)))
cvauc <- function(form) {
  p <- numeric(nrow(lm6))
  for (k in 1:5) { tr <- f != k
    m <- glm(form, lm6[tr, ], family = binomial)
    p[!tr] <- predict(m, lm6[!tr, ], type = "response") }
  list(auc = as.numeric(auc(roc(lm6$onset_within_h, p, quiet = TRUE, direction = "<"))),
       p = p)
}
st <- cvauc(onset_within_h ~ age + sofa_total + qsofa + sirs +
              hr + resp_rate + map + spo2 + temp_c)
tv <- cvauc(onset_within_h ~ age + sofa_total + qsofa + sirs +
              hr + resp_rate + map + spo2 + temp_c +
              hr_slope6 + resp_rate_slope6 + map_slope6 + spo2_slope6)
delong_p <- roc.test(roc(lm6$onset_within_h, st$p, quiet = TRUE, direction = "<"),
                     roc(lm6$onset_within_h, tv$p, quiet = TRUE, direction = "<"),
                     method = "delong")$p.value
```

```{r baseline-table}
kable(bind_rows(base_tab,
  tibble(Score = c("Multivariable (static, 6h)",
                   "Multivariable + 6h trajectory slopes"),
         `AUC (within 12h)` = sprintf("%.3f", c(st$auc, tv$auc)))),
  booktabs = TRUE,
  caption = "Discrimination for incident sepsis within 12h of the landmark (cross-validated for the multivariable models).") %>%
  kable_styling(latex_options = "HOLD_position") %>%
  pack_rows("Single clinical scores", 1, 3) %>%
  pack_rows("Multivariable", 4, 5)
```

```{r auct, fig.cap = "Time-dependent AUC for predicting sepsis onset over increasing horizons, using the cross-validated multivariable risk score.", fig.width=5, fig.height=3}
tr <- timeROC(T = lm6$t_event, delta = lm6$event, marker = tv$p,
              cause = 1, times = c(6, 12, 24, 36, 48), iid = FALSE)
tibble(t = c(6, 12, 24, 36, 48), AUC = as.numeric(tr$AUC)) %>%
  ggplot(aes(t, AUC)) +
  geom_line(colour = "#C0504D") + geom_point(colour = "#C0504D") +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
  coord_cartesian(ylim = c(0.5, 0.85)) +
  labs(x = "Prediction horizon (hours after landmark)", y = "Time-dependent AUC")
```

```{r export-values, include=FALSE}
# ── Export all key statistics and coefficients for thesis / downstream use ──

# 1. Cohort descriptors
cohort_desc <- data.frame(
  metric = c("n_stay", "n_sepsis", "prevalence", "prevalent_at_admission",
             "incident_after_h6", "median_onset_hour",
             "n_landmark", "n_landmark_event", "n_landmark_within_horizon"),
  value = c(n_stay, n_sep, prev, prevalent, incident6, med_onset,
            n_lm, n_lm_ev, n_lm_h),
  stringsAsFactors = FALSE
)
write.csv(cohort_desc, file.path(OUT, "09_cohort_descriptors.csv"), row.names = FALSE)

# 2. KM and log-rank
km_stats <- data.frame(
  metric = c("ci24_sirs_ge2", "ci24_sirs_lt2", "logrank_chisq", "logrank_p"),
  value = c(ci24$cuminc[ci24$grp == "SIRS >= 2"],
            ci24$cuminc[ci24$grp == "SIRS < 2"],
            lr$chisq, lr_p),
  stringsAsFactors = FALSE
)
write.csv(km_stats, file.path(OUT, "09_km_logrank.csv"), row.names = FALSE)

# 3. Cox baseline HRs
cox_base <- broom::tidy(cx, exponentiate = TRUE, conf.int = TRUE)
cox_base$ph_global_p <- zph_p
write.csv(cox_base, file.path(OUT, "09_cox_baseline.csv"), row.names = FALSE)

# 4. Time-varying Cox HRs
cox_tv <- broom::tidy(cx_tv, exponentiate = TRUE, conf.int = TRUE)
cox_tv$n_events <- n_tv_ev
write.csv(cox_tv, file.path(OUT, "09_cox_timevarying.csv"), row.names = FALSE)

# 5. Fine-Gray competing risks
fg_tab <- data.frame(
  term = names(fg$coef),
  coef = fg$coef,
  hr = exp(fg$coef),
  se = sqrt(diag(fg$var)),
  stringsAsFactors = FALSE
)
write.csv(fg_tab, file.path(OUT, "09_fine_gray.csv"), row.names = FALSE)

# 6. HMM parameters
hmm_params <- data.frame(
  state = c("1_low", "2_mid", "3_high"),
  emission_mean = em[, 1],
  emission_sd = em[, 2],
  self_persistence = diag(tmat),
  septic_hour_rate = sep_by_state$sep,
  stringsAsFactors = FALSE
)
write.csv(hmm_params, file.path(OUT, "09_hmm_params.csv"), row.names = FALSE)

# 7. Autocorrelation
acf_stats <- data.frame(
  metric = c("mean_lag1_acf", "median_lag1_acf", "n_stays_computed"),
  value = c(mean(acf1$a), median(acf1$a), nrow(acf1)),
  stringsAsFactors = FALSE
)
write.csv(acf_stats, file.path(OUT, "09_autocorrelation.csv"), row.names = FALSE)

# 8. Discrimination (AUCs + DeLong)
disc_tab <- data.frame(
  model = c("sofa_single", "qsofa_single", "sirs_single",
            "multivariable_static", "multivariable_trajectory"),
  auc = c(auc1("sofa_total"), auc1("qsofa"), auc1("sirs"), st$auc, tv$auc),
  stringsAsFactors = FALSE
)
disc_tab$delong_p_static_vs_traj <- c(NA, NA, NA, NA, delong_p)
write.csv(disc_tab, file.path(OUT, "09_discrimination.csv"), row.names = FALSE)

# 9. Time-dependent AUC
td_auc <- data.frame(
  horizon_h = c(6, 12, 24, 36, 48),
  auc = as.numeric(tr$AUC),
  stringsAsFactors = FALSE
)
write.csv(td_auc, file.path(OUT, "09_timedep_auc.csv"), row.names = FALSE)
```

```{r export-figures, include=FALSE}
# ── Export all figures as publication-quality PNGs ──

# Fig 1: Onset histogram
png(file.path(FIG, "09_onset_histogram.png"), width = 1200, height = 600, res = 200)
print(
  onset %>% filter(event == 1, onset_hour <= 48) %>%
    ggplot(aes(onset_hour)) +
    geom_histogram(binwidth=1, fill="#C0504D", colour="white", linewidth=0.1) +
    geom_vline(xintercept=LM_H, linetype="dashed", colour="grey30") +
    annotate("text", x=LM_H+1.5, y=Inf, label="landmark (6h)",
             vjust=1.5, hjust=0, size=3, colour="grey30") +
    labs(x = "Onset hour (from ICU admission)", y = "Septic stays")
)
dev.off()

# Fig 2: KM cumulative incidence
png(file.path(FIG, "09_km_cumulative_incidence.png"), width = 1200, height = 700, res = 200)
print(
  ggplot(km_df, aes(time, cuminc, colour = grp)) +
    geom_step(linewidth=0.7) +
    scale_colour_manual(values = c("SIRS < 2"="#5B9BD5","SIRS >= 2"="#C0504D")) +
    coord_cartesian(xlim=c(0,60)) +
    labs(x="Hours since landmark", y="Cumulative incidence of sepsis", colour=NULL) +
    theme(legend.position="top")
)
dev.off()

# Fig 3: Competing risks CIF
png(file.path(FIG, "09_competing_risks_cif.png"), width = 1200, height = 700, res = 200)
print(
  ggplot(ci_df, aes(time, est, colour=grp, linetype=cause)) +
    geom_step(linewidth=0.6) +
    coord_cartesian(xlim=c(0,60)) +
    scale_colour_manual(values=c("SIRS<2"="#5B9BD5","SIRS>=2"="#C0504D")) +
    labs(x="Hours since landmark", y="Cumulative incidence", colour=NULL, linetype=NULL) +
    theme(legend.position="top")
)
dev.off()

# Fig 4: Trajectories + ACF
png(file.path(FIG, "09_trajectories_acf.png"), width = 1400, height = 600, res = 200)
print(p_traj + p_acf)
dev.off()

# Fig 5: Time-dependent AUC
png(file.path(FIG, "09_timedep_auc.png"), width = 1000, height = 700, res = 200)
print(
  tibble(t = c(6,12,24,36,48), AUC = as.numeric(tr$AUC)) %>%
    ggplot(aes(t, AUC)) +
    geom_line(colour="#C0504D") + geom_point(colour="#C0504D") +
    geom_hline(yintercept=0.5, linetype="dashed", colour="grey60") +
    coord_cartesian(ylim=c(0.5, 0.85)) +
    labs(x="Prediction horizon (hours after landmark)", y="Time-dependent AUC")
)
dev.off()
```

Table \@ref(tab:baseline-table) is the central finding. **No single
clinical score is useful** for incident sepsis: SOFA is even *inverse*
(AUC `r base_tab[[2]][1]`), because organ dysfunction without evolving
infection does not progress to Sepsis-3; SIRS is the best single score
(AUC `r base_tab[[2]][3]`). A **multivariable** combination of the same
hourly measurements reaches AUC **`r sprintf('%.3f', st$auc)`**, and
adding **six-hour trajectory slopes** raises it further to
**`r sprintf('%.3f', tv$auc)`** (DeLong *p*
`r format.pval(delong_p, digits = 2, eps = 1e-16)`). Figure
\@ref(fig:auct) shows this discrimination holds across horizons. The
temporal features add real, statistically significant signal, providing
direct evidence that the trajectory carries information a snapshot does
not.

# Conclusion and Path to a Real-Time Model {#conclusion}

Reframing the problem longitudinally changed both the methods and the
findings.

-   **Prevalent vs. incident.** A third of Sepsis-3 is present at
    admission and not predictable; the real task is incident sepsis,
    isolated by the hour-`r LM_H` **landmark**
    (`r sprintf('%.1f%%', 100*n_lm_ev/n_lm)` of at-risk stays).

-   **The signal is in the dynamics, not the snapshot.** A baseline Cox
    violated proportional hazards; a **time-varying Cox** located the
    hazard in *current heart rate* rather than a SOFA severity score
    (which remains inversely related to incident onset).
    **Autocorrelation** (`r sprintf('%.2f', mean_acf)`) and a
    persistent-state **HMM** confirmed genuine within-patient temporal
    structure that the cross-sectional analysis had to discard.

-   **A performance bar.** Single streaming scores are weak or inverse;
    a multivariable hourly model reaches AUC
    **`r sprintf('%.3f', st$auc)`** and trajectory features push it to
    **`r sprintf('%.3f', tv$auc)`**. Any real-time model should be judged
    against these, with time-dependent AUC and competing-risk-aware,
    leakage-safe, patient-grouped evaluation.

**Toward the real-time model.** The labelled hourly panel, onset times,
landmark design, competing-risk framing, and streaming baselines
assembled here are the direct inputs to the real-time statistical model
built next (scripts 10-19): a dynamic **landmark supermodel** fit
entirely with penalised logistic regression, GEE, and time-varying Cox
(no gradient boosting or neural sequence models), emitting an hourly
sepsis risk by stacking predictions across a grid of landmark times. The
statistics in this report specify the target, the evaluation, and the
minimum performance that model must exceed.

**Limitations.** Onset depends on clinician behaviour (when cultures and
antibiotics are ordered), so incident "onset" partly reflects *recognition*
timing; the SOFA baseline is assumed zero; qSOFA uses MAP as a proxy for
systolic pressure; and the time-varying Cox and HMM use a sampled subset
for tractability. These are revisited as the modelling matures.

# Acknowledgment {.unnumbered}

This longitudinal analysis extends the cross-sectional study to the
project's real objective: real-time prediction, using the **PAIR
framework** for responsible AI use: the problem framing, survival and
time-series methods, and their honest (including negative) interpretation
were critically reviewed against primary sources.

# References {.unnumbered}

1.  Singer, M., et al. (2016). The Third International Consensus
    Definitions for Sepsis and Septic Shock (Sepsis-3). *JAMA*, 315(8).
2.  Seymour, C. W., et al. (2016). Assessment of Clinical Criteria for
    Sepsis. *JAMA*, 315(8).
3.  Reyna, M. A., et al. (2020). Early Prediction of Sepsis from Clinical
    Data: the PhysioNet/CinC Challenge 2019. *Critical Care Medicine*,
    48(2).
4.  Cox, D. R. (1972). Regression Models and Life-Tables. *JRSS B*,
    34(2).
5.  Fine, J. P., & Gray, R. J. (1999). A Proportional Hazards Model for
    the Subdistribution of a Competing Risk. *JASA*, 94(446).
6.  Blanche, P., et al. (2013). Estimating and comparing time-dependent
    areas under ROC curves. *Statistics in Medicine*, 32(30).
7.  Rabiner, L. R. (1989). A tutorial on hidden Markov models. *Proc.
    IEEE*, 77(2).
8.  van Houwelingen, H. C. (2007). Dynamic prediction by landmarking.
    *Scandinavian J. Statistics*, 34(1).
```
