---
title: "Tier 1: Full-Feature Elastic-Net Nomogram at the Hour-6 Landmark"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.width=8, fig.height=5)
```

## Overview

This script builds the full-feature penalized-logistic nomogram at the hour-6 landmark. It establishes the purely statistical upper bound for predicting incident Sepsis-3 within 12 hours of the hour-6 landmark by incorporating heavily-missing labs via an informative-missingness design (median-impute plus "was it measured" indicators). Evaluation uses honest, leakage-safe nested cross-validation with DeLong testing against the earlier baseline model.

## Load Libraries

We load the required packages for elastic-net regression, ROC analysis, precision-recall curves, and regression modelling/nomograms.

```{r load-libraries, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(glmnet); library(pROC); library(PRROC); library(rms)
})
set.seed(486649)
```

## Locate Data and Set Output Directories

This helper function searches candidate paths for data files, and then sets the output and figure directories.

```{r locate-data}
find_data <- function(f) {
  cands <- c(file.path("../../processed_data", f),
             file.path("processed_data", f), f)
  hit <- cands[file.exists(cands)][1]
  if (is.na(hit)) stop("cannot find ", f)
  hit
}
OUT <- dirname(find_data("landmark_h6.csv"))
FIG <- if (dir.exists("figure")) "figure" else { dir.create("figure"); "figure" }
```

## Load Landmark Data

We read in the hour-6 landmark dataset and print a summary of its size and onset rate.

```{r load-data}
lm6 <- read.csv(find_data("landmark_h6.csv"))
cat(sprintf("Loaded landmark_h6: %d stays, onset-within-12h rate %.3f\n",
            nrow(lm6), mean(lm6$onset_within_h)))
```

## Feature Design

The feature matrix combines vitals (near-complete after forward-fill), severity scores (SOFA, qSOFA, SIRS), labs with informative-missingness indicators, and 6-hour trajectory slopes. Labs and slopes use median/zero imputation with a binary "measured" flag to preserve the informative nature of missingness.

```{r feature-design}
# Vitals (near-complete after forward-fill at h6): use directly.
vitals <- c("hr","resp_rate","map","spo2","temp_c","gcs")
# Scores allowed now (nomogram is a full-feature model, not the non-circular one).
scores <- c("sofa_total","qsofa","sirs")
# Labs: 70-88% missing at h6 and the missingness is INFORMATIVE (a drawn lactate
# marks a sicker/suspected patient). Median-impute + a "_msg" measured indicator.
labs   <- c("lactate","creatinine","wbc","pf_ratio")
# 6h trajectory slopes: ~30% missing (fewer than 2 measurements). Impute 0
# (no trend) + a "_msg" indicator.
slopes <- c("hr_slope6","resp_rate_slope6","map_slope6","spo2_slope6")

df <- data.frame(row.names = seq_len(nrow(lm6)))
for (v in c(vitals, scores)) {                       # simple median-impute
  x <- lm6[[v]]; x[is.na(x)] <- median(x, na.rm = TRUE); df[[v]] <- x
}
for (v in labs) {                                    # impute + informative flag
  x <- lm6[[v]]; df[[paste0(v, "_msg")]] <- as.integer(is.na(x))
  x[is.na(x)] <- median(x, na.rm = TRUE); df[[v]] <- x
}
for (v in slopes) {                                  # impute 0 + flag
  x <- lm6[[v]]; df[[paste0(v, "_msg")]] <- as.integer(is.na(x))
  x[is.na(x)] <- 0; df[[v]] <- x
}
df$age    <- lm6$age
df$female <- as.integer(lm6$gender == "F")
y <- lm6$onset_within_h
X <- as.matrix(df)
cat(sprintf("Design matrix: %d features\n", ncol(X)))
```

## Outer 5-Fold CV with Inner Lambda Selection

We run a nested cross-validation loop. The outer loop splits stays into 5 folds, and the inner loop (inside `cv.glmnet`) selects the optimal regularization parameter lambda on each training fold. Out-of-sample predictions are collected for honest evaluation.

```{r outer-cv}
folds <- sample(rep(1:5, length.out = nrow(X)))
oos_en <- numeric(nrow(X))                           # elastic-net OOS risk
ALPHA  <- 0.5                                         # elastic net (LASSO-ish)
for (k in 1:5) {
  tr <- folds != k
  cvfit <- cv.glmnet(X[tr, ], y[tr], family = "binomial",
                     alpha = ALPHA, nfolds = 5, standardize = TRUE)
  oos_en[!tr] <- as.numeric(predict(cvfit, X[!tr, ],
                                    s = "lambda.min", type = "response"))
}
```

## Reconstruct Baseline Models

We re-fit the baseline GLMs from the earlier script (09) on the same outer folds so that a fair DeLong test comparison can be made. Both the static-only and trajectory-augmented baselines are reconstructed.

```{r baseline-models}
# Reconstruct the earlier (script 09) baseline models on the SAME outer
# folds (for DeLong). Fit on the imputed matrix `df` so all three models use
# identical rows (the raw lm6 vitals carry a little missingness that would
# leave NA predictions).
gdat <- cbind(df, y = y)
cvauc_glm <- function(form, data) {
  p <- numeric(nrow(data))
  for (k in 1:5) { tr <- folds != k
    m <- glm(form, data[tr, ], family = binomial)
    p[!tr] <- predict(m, data[!tr, ], type = "response") }
  p
}
p_static <- cvauc_glm(y ~ age + sofa_total + qsofa + sirs +
                        hr + resp_rate + map + spo2 + temp_c, gdat)
p_traj   <- cvauc_glm(y ~ age + sofa_total + qsofa + sirs +
                        hr + resp_rate + map + spo2 + temp_c +
                        hr_slope6 + resp_rate_slope6 + map_slope6 + spo2_slope6,
                      gdat)
```

## Compute Metrics

We calculate AUROC, PR-AUC, and Brier score for each model, then run a DeLong test comparing the elastic-net model against the trajectory baseline. Results are saved to `tier1_metrics.csv`.

```{r metrics}
auc_of <- function(p) as.numeric(auc(roc(y, p, quiet = TRUE, direction = "<")))
prauc_of <- function(p) pr.curve(scores.class0 = p[y==1],
                                 scores.class1 = p[y==0])$auc.integral
brier <- function(p) mean((p - y)^2)

roc_en  <- roc(y, oos_en,   quiet = TRUE, direction = "<")
roc_trj <- roc(y, p_traj,   quiet = TRUE, direction = "<")
dl_vs_traj   <- roc.test(roc_en, roc_trj, method = "delong")$p.value

metrics <- data.frame(
  model = c("Baseline static (glm)", "Baseline + trajectory slopes (glm, 0.755 AUROC)",
            "Tier1 full-feature elastic net"),
  auroc = c(auc_of(p_static), auc_of(p_traj), auc_of(oos_en)),
  prauc = c(prauc_of(p_static), prauc_of(p_traj), prauc_of(oos_en)),
  brier = c(brier(p_static), brier(p_traj), brier(oos_en)))
metrics$delong_p_vs_baseline_traj <- c(NA, NA, dl_vs_traj)
metrics[ ,-1] <- round(metrics[ ,-1], 4)
print(metrics, row.names = FALSE)
write.csv(metrics, file.path(OUT, "tier1_metrics.csv"), row.names = FALSE)
```

## Final Model Coefficients

We fit the elastic-net model on all available data and extract the penalized coefficients and odds ratios. Only features with non-zero coefficients are retained and saved to `tier1_coefs.csv`.

```{r final-model-coefs}
cvfull <- cv.glmnet(X, y, family = "binomial", alpha = ALPHA,
                    nfolds = 5, standardize = TRUE)
co <- as.matrix(coef(cvfull, s = "lambda.min"))
coefs <- data.frame(term = rownames(co), coef = as.numeric(co))
coefs$odds_ratio <- exp(coefs$coef)
coefs <- coefs[coefs$coef != 0, ]
coefs[ ,c("coef","odds_ratio")] <- round(coefs[ ,c("coef","odds_ratio")], 4)
write.csv(coefs, file.path(OUT, "tier1_coefs.csv"), row.names = FALSE)
cat(sprintf("\nSelected features (non-zero at lambda.min): %d / %d\n",
            nrow(coefs) - 1, ncol(X)))
```

## Export Out-of-Sample Predictions

The per-stay out-of-sample risk estimates are exported to `tier1_oos_pred.csv`. These feed into the Tier 4 alarm/utility analysis downstream.

```{r export-oos-predictions}
write.csv(data.frame(stay_id = lm6$stay_id, onset_within_h = y,
                     risk_en = oos_en, risk_baseline_traj = p_traj),
          file.path(OUT, "tier1_oos_pred.csv"), row.names = FALSE)
```

## Calibration Curve

We produce a decile-binned calibration plot comparing predicted risk against observed onset rate. This plot is saved to `figure/calibration_h6.png`.

```{r calibration-curve, fig.width=6, fig.height=6}
png(file.path(FIG, "calibration_h6.png"), width = 1000, height = 1000, res = 200)
bins <- cut(oos_en, quantile(oos_en, seq(0, 1, 0.1)), include.lowest = TRUE)
cal <- aggregate(cbind(pred = oos_en, obs = y), list(bin = bins), mean)
plot(cal$pred, cal$obs, pch = 19, xlab = "Predicted risk",
     ylab = "Observed onset rate", main = "Calibration (out-of-sample, h6)",
     xlim = range(cal$pred, cal$obs), ylim = range(cal$pred, cal$obs))
abline(0, 1, lty = 2, col = "grey50"); lines(cal$pred, cal$obs, col = "steelblue")
dev.off()
```

## Nomogram

We refit an unpenalized logistic regression model (`lrm`) on the features selected by elastic net, then render a clinician-facing nomogram. The figure is saved to `figure/nomogram_h6.png`.

```{r nomogram, fig.width=10, fig.height=8}
sel <- setdiff(coefs$term, "(Intercept)")
nomo_df <- df[, intersect(sel, names(df)), drop = FALSE]; nomo_df$y <- y
dd <- datadist(nomo_df); options(datadist = "dd")
frm <- as.formula(paste("y ~", paste(names(nomo_df)[names(nomo_df) != "y"],
                                     collapse = " + ")))
fit_lrm <- lrm(frm, data = nomo_df, maxit = 1000)
png(file.path(FIG, "nomogram_h6.png"), width = 1800, height = 1400, res = 180)
plot(nomogram(fit_lrm, fun = plogis, funlabel = "Risk of sepsis within 12h"),
     cex.axis = 0.6, cex.var = 0.7)
dev.off()
```

## Done

```{r done}
cat("\nTier 1 complete. Outputs in", OUT, "and", FIG, "\n")
```
