---
title: "Tier 2: Dynamic Landmark Supermodel"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.width=8, fig.height=5)
```

## Overview

This script fits a dynamic-prediction landmark supermodel (van Houwelingen 2007). A single penalized-logistic model is fit across the full stack of landmarks at {6, 12, 18, 24, 36, 48} hours, with the landmark time *s* and feature-by-*s* interactions as covariates. Evaluating it at the current hour yields an hourly-updating sepsis risk, a genuine real-time model built entirely from classical statistics. Leakage-safe evaluation uses patient-grouped CV folds and only past-information features.

## Load Libraries

We load the required packages for elastic-net regression and ROC analysis.

```{r load-libraries, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({ library(glmnet); library(pROC) })
set.seed(486649)
```

## Locate Data and Set Output Directories

This helper function searches candidate paths for data files, and then sets the output and figure directories.

```{r locate-data}
find_data <- function(f) {
  cands <- c(file.path("../../processed_data", f),
             file.path("processed_data", f), f)
  hit <- cands[file.exists(cands)][1]
  if (is.na(hit)) stop("cannot find ", f); hit
}
OUT <- dirname(find_data("landmark_stack.csv"))
FIG <- if (dir.exists("figure")) "figure" else { dir.create("figure"); "figure" }
```

## Load Landmark Stack

We read in the stacked landmark dataset and print a summary of its dimensions and available landmark times.

```{r load-data}
st <- read.csv(find_data("landmark_stack.csv"))
cat(sprintf("Loaded stack: %d rows, %d stays, landmarks {%s}\n",
            nrow(st), length(unique(st$stay_id)),
            paste(sort(unique(st$landmark_time)), collapse = ",")))
```

## Feature Design

Features follow the same informative-missingness scheme as Tier 1: vitals and scores get simple median imputation, labs and slopes receive median/zero imputation plus a binary "measured" indicator. Age and sex are also included.

```{r feature-design}
# Feature design (same informative-missingness scheme as Tier 1)
vitals <- c("hr","resp_rate","map","spo2","temp_c","gcs")
labs   <- c("lactate","creatinine","wbc","pf_ratio")
slopes <- c("hr_slope6","resp_rate_slope6","map_slope6","spo2_slope6")
scores <- c("sofa_total","qsofa","sirs")

df <- data.frame(row.names = seq_len(nrow(st)))
for (v in c(vitals, scores)) { x <- st[[v]]; x[is.na(x)] <- median(x, na.rm=TRUE); df[[v]] <- x }
for (v in labs)   { x <- st[[v]]; df[[paste0(v,"_msg")]] <- as.integer(is.na(x));
                    x[is.na(x)] <- median(x, na.rm=TRUE); df[[v]] <- x }
for (v in slopes) { x <- st[[v]]; df[[paste0(v,"_msg")]] <- as.integer(is.na(x));
                    x[is.na(x)] <- 0; df[[v]] <- x }
df$age <- st$age; df$female <- as.integer(st$gender == "F")
base_feats <- names(df)
```

## Landmark Time and Interactions

We add the standardized landmark time *s*, its quadratic term, and interactions of each base feature with *s*. Standardizing *s* ensures that the elastic-net penalty treats the interaction terms on a comparable scale to the main effects.

```{r landmark-interactions}
# landmark time and its interactions: s, s^2, and feature x s (standardized s so
# penalization treats the interactions on a comparable scale)
s   <- st$landmark_time
s_z <- as.numeric(scale(s))
df$s   <- s_z
df$s2  <- s_z^2
for (v in base_feats) df[[paste0(v, "_xS")]] <- df[[v]] * s_z   # smooth s-modulation
X <- as.matrix(df)
y <- st$onset_within_h
cat(sprintf("Supermodel design: %d features (incl. s, s^2, %d s-interactions)\n",
            ncol(X), length(base_feats)))
```

## Patient-Grouped 5-Fold CV

Cross-validation folds are assigned at the stay (patient) level rather than the row level, because each stay recurs across multiple landmarks. This prevents information leakage.

```{r patient-grouped-cv}
# Patient-grouped 5-fold CV (split on stay_id, NOT rows)
ids   <- unique(st$stay_id)
idf   <- setNames(sample(rep(1:5, length.out = length(ids))), ids)
fold  <- idf[as.character(st$stay_id)]

oos <- numeric(nrow(X))
for (k in 1:5) {
  tr <- fold != k
  cv <- cv.glmnet(X[tr,], y[tr], family="binomial", alpha=0.5,
                  nfolds=5, standardize=TRUE)
  oos[!tr] <- as.numeric(predict(cv, X[!tr,], s="lambda.min", type="response"))
}
```

## Separate Per-Landmark Comparator

As a comparator, we fit a separate penalized model at each landmark time independently (no pooling across landmarks, no *s* terms). This lets us assess whether the supermodel's pooling-plus-*s* approach improves discrimination.

```{r separate-per-landmark}
# Comparator: a SEPARATE penalized model per landmark (no pooling)
Xb <- as.matrix(df[, base_feats])        # base features only, no s terms
oos_sep <- numeric(nrow(X))
for (sl in sort(unique(s))) {
  idx <- which(s == sl)
  fk  <- fold[idx]
  for (k in 1:5) {
    tr <- fk != k
    if (sum(y[idx][tr]) < 10) { oos_sep[idx][!tr] <- mean(y[idx][tr]); next }
    cv <- cv.glmnet(Xb[idx,][tr,], y[idx][tr], family="binomial",
                    alpha=0.5, nfolds=5, standardize=TRUE)
    oos_sep[idx][!tr] <- as.numeric(predict(cv, Xb[idx,][!tr,],
                                            s="lambda.min", type="response"))
  }
}
```

## AUC by Landmark

We compute the out-of-sample AUROC at each prediction hour for both the supermodel and the separate per-landmark models. Results are saved to `tier2_auc_by_landmark.csv` and `tier2_supermodel_oos.csv`.

```{r auc-by-landmark}
# AUC(s): discrimination at each prediction hour
auc_at <- function(p, sl) {
  i <- s == sl
  as.numeric(auc(roc(y[i], p[i], quiet=TRUE, direction="<")))
}
grid <- sort(unique(s))
tab <- data.frame(
  landmark_h = grid,
  n_at_risk  = as.integer(table(s)[as.character(grid)]),
  n_onset    = as.integer(tapply(y, s, sum)[as.character(grid)]),
  auc_supermodel   = sapply(grid, function(sl) auc_at(oos, sl)),
  auc_separate     = sapply(grid, function(sl) auc_at(oos_sep, sl)))
tab$auc_supermodel <- round(tab$auc_supermodel, 4)
tab$auc_separate   <- round(tab$auc_separate, 4)
overall <- as.numeric(auc(roc(y, oos, quiet=TRUE, direction="<")))
cat(sprintf("\nOverall supermodel out-of-sample AUROC (pooled): %.4f\n", overall))
print(tab, row.names = FALSE)
write.csv(tab, file.path(OUT, "tier2_auc_by_landmark.csv"), row.names=FALSE)
write.csv(data.frame(stay_id=st$stay_id, landmark_time=st$landmark_time,
                     onset_within_h=y, risk=oos),
          file.path(OUT, "tier2_supermodel_oos.csv"), row.names=FALSE)
```

## AUC by Landmark Figure

This figure plots AUC as a function of prediction hour, comparing the supermodel against the separate per-landmark models. The plot is saved to `figure/auc_by_landmark.png`.

```{r auc-figure, fig.width=7, fig.height=5}
png(file.path(FIG, "auc_by_landmark.png"), width=1100, height=850, res=200)
plot(tab$landmark_h, tab$auc_supermodel, type="b", pch=19, col="steelblue",
     ylim=range(c(tab$auc_supermodel, tab$auc_separate, 0.5)),
     xlab="Prediction hour s (landmark)", ylab="Out-of-sample AUROC (12h horizon)",
     main="Dynamic AUC(s): landmark supermodel")
lines(tab$landmark_h, tab$auc_separate, type="b", pch=1, lty=2, col="grey40")
abline(h=0.5, lty=3, col="grey70")
legend("topright", c("Supermodel (pooled + s)","Separate per-landmark"),
       col=c("steelblue","grey40"), pch=c(19,1), lty=c(1,2), bty="n", cex=0.8)
dev.off()
```

## Done

```{r done}
cat("\nTier 2 complete.\n")
```
