---
title: "Tier 3: Within-Patient Correlation (Repeated Measures)"
subtitle: "GEE and Penalized Time-Varying Cox Models on the Hourly Panel"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.width=8, fig.height=5)
```

## Overview

The longitudinal report (step 09) never modelled the within-patient correlation
of repeated hourly measurements. This script addresses that gap in two classical
ways: (1) GEE with an exchangeable working correlation, and (2) a penalized
(ridge) time-varying-covariate Cox model on a counting-process formulation. Both
operate on the sampled long panel (2,500 stays) for tractability.

## Load Libraries and Set Seed

```{r load-libraries, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({ library(geepack); library(sandwich); library(survival) })
set.seed(486649)
```

## Locate and Read Data

```{r load-data}
find_data <- function(f) {
  cands <- c(file.path("../../processed_data", f),
             file.path("processed_data", f), f)
  hit <- cands[file.exists(cands)][1]
  if (is.na(hit)) stop("cannot find ", f); hit
}
OUT <- dirname(find_data("08_panel_sample_long.parquet"))
d0 <- as.data.frame(arrow::read_parquet(find_data("08_panel_sample_long.parquet")))   # full copy (for the Cox)
d <- d0
```

## Define the Dynamic Incident Outcome

The stored per-hour `label` is nearly constant within a patient, so a random intercept would suffer complete separation. A proper early-warning target restricts to at-risk hours and defines the outcome as whether sepsis onset occurs within the next 12 hours. This outcome varies within septic patients, making it a well-posed repeated-measures formulation.

```{r define-outcome}
HORIZON <- 12
d <- d[is.na(d$onset_hour) | d$hour < d$onset_hour, ]
d$y <- as.integer(!is.na(d$onset_hour) &
                    (d$onset_hour - d$hour) <= HORIZON &
                    (d$onset_hour - d$hour) > 0)
cat(sprintf("Loaded sampled panel (at-risk hours): %d stay-hours, %d stays, incident-within-12h rate %.3f\n",
            nrow(d), length(unique(d$stay_id)), mean(d$y)))
```

## Standardize Predictors

Median imputation and z-scoring yield odds ratios on a comparable scale across predictors.

```{r standardize-predictors}
zt <- function(x) { x[is.na(x)] <- median(x, na.rm=TRUE); as.numeric(scale(x)) }
d$z_hr   <- zt(d$hr);   d$z_map  <- zt(d$map);   d$z_rr <- zt(d$resp_rate)
d$z_temp <- zt(d$temp_c); d$z_lac <- zt(d$lactate); d$z_hour <- zt(d$hour)
preds <- c("z_hr","z_map","z_rr","z_temp","z_lac","z_hour")
form  <- as.formula(paste("y ~", paste(preds, collapse=" + ")))
```

## GEE vs Naive Pooled GLM

A per-patient random-intercept GLMM is degenerate here: most patients never experience an event, so the per-cluster intercepts diverge and the variance component explodes. GEE avoids this by estimating population-averaged effects with an exchangeable within-patient working correlation and cluster-robust sandwich standard errors. Contrasting with a naive pooled GLM that treats the approximately 76,000 stay-hours as independent quantifies how much ignoring within-patient correlation understates the standard errors.

```{r fit-gee}
d <- d[order(d$stay_id, d$hour), ]
g_pool <- glm(form, data=d, family=binomial)

se_naive  <- sqrt(diag(vcov(g_pool)))
se_robust <- sqrt(diag(vcovCL(g_pool, cluster = d$stay_id, type = "HC0")))
cat("fitting GEE (exchangeable within-patient correlation) ...\n")
g_gee <- geeglm(form, id=stay_id, data=d, family=binomial, corstr="exchangeable")
alpha <- summary(g_gee)$corr[["Estimate"]]
cat(sprintf("exchangeable within-patient correlation alpha = %.3f\n", alpha))

nm <- names(coef(g_pool))
comp <- data.frame(
  term            = nm,
  OR_glm          = round(exp(coef(g_pool)), 3),
  SE_naive        = round(se_naive[nm], 3),
  SE_cluster_robust = round(se_robust[nm], 3))
comp$SE_inflation <- round(comp$SE_cluster_robust / comp$SE_naive, 2)
rownames(comp) <- NULL
cat(sprintf("\nCluster-robust vs naive SEs on the pooled GLM (within-patient alpha=%.3f):\n", alpha))
print(comp, row.names=FALSE)
write.csv(rbind(comp,
                data.frame(term="__alpha_within_patient__", OR_glm=round(alpha,4),
                           SE_naive=NA, SE_cluster_robust=NA, SE_inflation=NA)),
          file.path(OUT, "13_tier3_gee.csv"), row.names=FALSE)
```

## Penalized Time-Varying-Covariate Cox Model

A counting-process (start, stop] formulation of the hourly data supports a ridge-penalized Cox model that captures how time-varying physiology predicts the hazard of sepsis onset. Cluster-robust variance accounts for within-patient correlation. The full sampled panel (not the pre-onset-only GEE frame) is used here.

```{r fit-tvcox}
cat("\nbuilding counting-process time-varying Cox (ridge) ...\n")
dc <- d0[order(d0$stay_id, d0$hour), ]
dc$z_hr <- zt(dc$hr); dc$z_map <- zt(dc$map); dc$z_rr <- zt(dc$resp_rate)
dc$z_temp <- zt(dc$temp_c); dc$z_lac <- zt(dc$lactate)
dc$start <- dc$hour
dc$stop  <- dc$hour + 1
dc$ev <- as.integer(!is.na(dc$onset_hour) & dc$hour == dc$onset_hour)
dc <- dc[is.na(dc$onset_hour) | dc$hour <= dc$onset_hour, ]
tv <- coxph(Surv(start, stop, ev) ~
              ridge(z_hr, z_map, z_rr, z_temp, z_lac, theta=1) + cluster(stay_id),
            data=dc)
cox_tab <- data.frame(
  term = c("z_hr","z_map","z_rr","z_temp","z_lac"),
  HR   = round(exp(coef(tv)), 3))
cat("\nTime-varying Cox (ridge) hazard ratios for incident onset:\n")
print(cox_tab, row.names=FALSE)
write.csv(cox_tab, file.path(OUT, "13_tier3_tvcox.csv"), row.names=FALSE)
cat("\nTier 3 complete.\n")
```
