---
title: "Analysis 5: Subgroup Fairness Analysis"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.width=8, fig.height=5)
```

## Overview

Subgroup discrimination and calibration of the Analysis-1 nomogram's out-of-sample risk are assessed here, stratified by sex and age band. For each subgroup the sample size, event rate, AUROC, and calibration-in-the-large (mean predicted risk versus observed rate) are reported. External validation on eICU is deferred (no local database available).

## Load Libraries

```{r load-libraries, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({ library(pROC) })
```

## Locate and Read Data

```{r load-data}
find_data <- function(f) {
  cands <- c(file.path("../../processed_data", f),
             file.path("processed_data", f), f)
  hit <- cands[file.exists(cands)][1]; if (is.na(hit)) stop("cannot find ", f); hit
}
OUT <- dirname(find_data("10_analysis1_oos_pred.parquet"))

pred <- as.data.frame(arrow::read_parquet(find_data("10_analysis1_oos_pred.parquet")))
lm6  <- as.data.frame(arrow::read_parquet(find_data("08_landmark_h6.parquet")))[, c("stay_id","age","gender")]
d <- merge(pred, lm6, by="stay_id")
d$age_band <- cut(d$age, c(-Inf,50,65,80,Inf),
                  labels=c("<50","50-64","65-79","80+"))
```

## Define Subgroup Metrics Function

For each subgroup, AUROC, mean predicted risk, observed event rate, and their ratio (calibration-in-the-large) are computed. A calibration ratio of 1.0 indicates that the average prediction matches the true event rate; values above 1 indicate over-prediction, values below 1 under-prediction.

```{r define-metrics}
grp_metrics <- function(sub, label) {
  y <- sub$onset_within_h; p <- sub$risk_en
  au <- tryCatch(as.numeric(auc(roc(y, p, quiet=TRUE, direction="<"))), error=function(e) NA)
  data.frame(subgroup=label, n=nrow(sub), event_rate=round(mean(y),4),
             auroc=round(au,4), mean_pred=round(mean(p),4),
             obs_rate=round(mean(y),4),
             calib_ratio=round(mean(p)/mean(y),3))
}
```

## Compute Subgroup Metrics and Write Results

Fairness metrics are computed for the full cohort and for each sex and age-band subgroup, then saved to `analysis5_subgroups.csv`.

```{r compute-metrics}
rows <- list(grp_metrics(d, "ALL"))
for (g in sort(unique(d$gender)))
  rows[[length(rows)+1]] <- grp_metrics(d[d$gender==g,], paste0("sex=",g))
for (b in levels(d$age_band))
  rows[[length(rows)+1]] <- grp_metrics(d[d$age_band==b,], paste0("age=",b))
tab <- do.call(rbind, rows)
print(tab, row.names=FALSE)
write.csv(tab, file.path(OUT, "16_analysis5_subgroups.csv"), row.names=FALSE)
cat("\nAnalysis 5 (fairness) complete. External eICU validation deferred (no data).\n")
```
