---
title: "01 Setup - Environment and Data Verification"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook verifies the conditions that every later stage assumes: the
required R packages are installed, the MIMIC-IV and eICU-CRD directories and
their source tables are at the locations set in `config.R`, the patient table
contains the variable that defines the temporal split, and the output
directory accepts writes. It inspects file metadata and the first rows of two
tables only. It does not validate the contents of the data.

```{r load-config}
source("config.R")
source("utils.R")

step("=== Phase 1: Environment Setup ===")
```

## Required Packages

Each required package is tested for whether it can be loaded, and any package
that cannot is installed from CRAN. The table reports the status found before
that installation step: a package marked MISSING was absent when the run began,
and an installation was attempted for it. A package installed here takes the
current CRAN version. The versions under which the reported results were
produced are pinned in `requirements.txt` and installed by
`install_requirements.R`.

```{r check-packages}
REQUIRED_PKGS <- c(
  "data.table", "arrow", "dplyr", "tidyr", "ggplot2", "lubridate",
  "survival", "nnet", "VGAM", "discSurv", "rms", "cmprsk",
  "xgboost", "pmsampsize", "dcurves", "timeROC", "CalibrationCurves",
  "purrr", "tibble", "readr"
)

step("Checking required packages...")
pkg_status <- vapply(REQUIRED_PKGS, requireNamespace, logical(1), quietly = TRUE)
missing_pkgs <- names(pkg_status)[!pkg_status]

if (length(missing_pkgs) > 0) {
  v2_log("Installing missing packages: ", paste(missing_pkgs, collapse = ", "))
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org", quiet = TRUE)
}

pkg_report <- data.frame(
  package = REQUIRED_PKGS,
  status  = ifelse(pkg_status, "OK", "MISSING"),
  stringsAsFactors = FALSE
)
knitr::kable(pkg_report, caption = "Package availability")
```

## Data Directories

The notebook checks that the MIMIC-IV v3.1 root directory (Johnson et al.,
2023) and its `hosp` and `icu` modules exist, together with the eICU-CRD v2.0
root directory (Pollard et al., 2018). Both datasets are located under the
project root, which `config.R` takes from the `MIMIC_RESEARCH_ROOT`
environment variable or, if that is unset, finds by searching upward from the
pipeline directory for `data/mimic-iv-3.1`. The project root can therefore sit
above the pipeline directory, whereas outputs are always written to `output/`
inside the pipeline directory.

```{r check-dirs}
step("Verifying data directories...")

check_dir <- function(label, path) {
  exists <- dir.exists(path)
  v2_log(sprintf("  %-30s %s  (%s)", label,
                 if (exists) "[OK]" else "[MISSING]", path))
  exists
}

mimic_ok <- check_dir("MIMIC-IV root",  MIMIC_DIR)
hosp_ok  <- check_dir("MIMIC-IV/hosp", HOSP_DIR)
icu_ok   <- check_dir("MIMIC-IV/icu",  ICU_DIR)
eicu_ok  <- check_dir("eICU root",      EICU_DIR)

dir_report <- data.frame(
  directory = c("MIMIC-IV root", "MIMIC-IV/hosp", "MIMIC-IV/icu", "eICU root"),
  path      = c(MIMIC_DIR, HOSP_DIR, ICU_DIR, EICU_DIR),
  status    = c(mimic_ok, hosp_ok, icu_ok, eicu_ok)
)
knitr::kable(dir_report, caption = "Data directory status")
```

## Source Table Availability

For the principal MIMIC-IV and eICU-CRD source tables read by later stages,
the notebook records whether each file exists and its size on disk. The files
are not opened at this point, so a file that is present may still be
truncated or unreadable. Such a fault appears when a later stage first reads
the file.

```{r check-mimic-files}
step("Spot-checking key MIMIC-IV files...")

KEY_FILES <- list(
  admissions         = hosp_file("admissions.csv"),
  patients           = hosp_file("patients.csv"),
  icustays           = icu_file("icustays.csv"),
  chartevents        = icu_file("chartevents.csv"),
  labevents          = hosp_file("labevents.csv"),
  prescriptions      = hosp_file("prescriptions.csv"),
  microbiologyevents = hosp_file("microbiologyevents.csv"),
  inputevents        = icu_file("inputevents.csv")
)

file_report <- lapply(names(KEY_FILES), function(nm) {
  path   <- KEY_FILES[[nm]]
  exists <- file.exists(path)
  sz     <- if (exists) file.size(path) else 0
  v2_log(sprintf("  %-25s %s  (%.1f MB)", nm,
                 if (exists) "[OK]" else "[MISSING]", sz / 1e6))
  data.frame(table = nm, exists = exists, size_mb = round(sz / 1e6, 1))
})
file_df <- do.call(rbind, file_report)
knitr::kable(file_df, caption = "MIMIC-IV key files")
```

```{r check-eicu-files}
step("Spot-checking eICU files...")

EICU_KEY_FILES <- list(
  patient       = file.path(EICU_DIR, "patient.csv.gz"),
  lab           = file.path(EICU_DIR, "lab.csv.gz"),
  medication    = file.path(EICU_DIR, "medication.csv.gz"),
  vitalPeriodic = file.path(EICU_DIR, "vitalPeriodic.csv.gz"),
  microLab      = file.path(EICU_DIR, "microLab.csv.gz"),
  diagnosis     = file.path(EICU_DIR, "diagnosis.csv.gz")
)

eicu_report <- lapply(names(EICU_KEY_FILES), function(nm) {
  path   <- EICU_KEY_FILES[[nm]]
  exists <- file.exists(path)
  sz     <- if (exists) file.size(path) else 0
  v2_log(sprintf("  %-25s %s  (%.1f MB)", nm,
                 if (exists) "[OK]" else "[MISSING]", sz / 1e6))
  data.frame(table = nm, exists = exists, size_mb = round(sz / 1e6, 1))
})
eicu_df <- do.call(rbind, eicu_report)
knitr::kable(eicu_df, caption = "eICU key files")
```

## Column Check

Two tables are read for their first five rows only. The column names of
`icustays` are printed for inspection and are not tested against an expected
list. The `patients` table is tested for `anchor_year_group`, the variable
from which stage 03 builds the temporal train/test split. Column types and
value ranges are not checked here.

```{r check-schema}
step("Peeking at icustays to verify schema...")

icustays_path <- icu_file("icustays.csv")
if (file.exists(icustays_path)) {
  stays_sample <- data.table::fread(icustays_path, nrows = 5, showProgress = FALSE)
  cat("icustays columns:\n")
  print(names(stays_sample))

  patients_path <- hosp_file("patients.csv")
  if (file.exists(patients_path)) {
    pts_sample      <- data.table::fread(patients_path, nrows = 5, showProgress = FALSE)
    has_anchor      <- "anchor_year_group" %in% names(pts_sample)
    v2_log("anchor_year_group present: ", has_anchor)
    cat(sprintf("anchor_year_group present: %s\n", has_anchor))
  }
} else {
  cat("icustays.csv not found - check MIMIC_DIR path.\n")
}
```

## Output Directory Write Check

The notebook writes a small test file to the output directory and then
deletes it. `config.R` creates the directory when it is absent, so this step
tests write permission only. A failure is printed below but does not stop the
run.

```{r check-output-dir}
step("Checking output directory...")

test_file <- file.path(OUTPUT_DIR, ".write_test")
tryCatch({
  writeLines("test", test_file)
  file.remove(test_file)
  cat(sprintf("Output dir writable: OK\n%s\n", OUTPUT_DIR))
}, error = function(e) {
  cat(sprintf("Output dir NOT writable: %s\n", e$message))
})
```

## Summary

The summary table lists the resolved project root, the availability of each
database, the output directory, the R version, and the label variants. MIMIC-IV
is reported as available only when both its `hosp` and `icu` modules are
present. The label-variant row lists the three pre-registered variants on
which the confirmatory analyses are defined; their names and Sepsis-3
operationalisations are set in `LABEL_VARIANTS` in `config.R`, and stage 02
builds a cohort for each. The alternative variants D to F are defined
separately in `config.R` and are not listed here. The MIMIC-IV source table
report is also saved as `01_setup_file_report.parquet` in
`output/processed_data/`.

```{r summary}
step("=== Setup Summary ===")

summary_df <- data.frame(
  item    = c("Project root", "MIMIC-IV", "eICU", "Output dir", "R version", "Label variants"),
  value   = c(
    PROJECT_ROOT,
    if (hosp_ok && icu_ok) "AVAILABLE" else "MISSING",
    if (eicu_ok) "AVAILABLE" else "MISSING",
    OUTPUT_DIR,
    as.character(getRversion()),
    "A (Narrow), B (Seymour-Standard), C (Liberal)"
  )
)
knitr::kable(summary_df, caption = "Setup summary")

save_parquet(file_df, "01_setup_file_report", DATA_DIR)
cat("\nSaved: 01_setup_file_report.parquet\n")
```
