---
title: "02 Cohort Labels - Base Cohort and Sepsis Label Variants"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook builds the study's base cohort from MIMIC-IV and assigns a
sepsis label to every stay under each label variant. The base cohort is fixed:
every variant is applied to the same ICU stays, so the variants differ only in
which stays receive a sepsis onset and when that onset is placed.

The three pre-registered variants, A (Narrow), B (Seymour-Standard) and C
(Liberal), are alternative operationalisations of Sepsis-3 (Singer et al.,
2016), following Cohen et al. (2024). They differ in the antibiotic-culture
windows that define suspected infection, in the window over which the SOFA
(Sequential Organ Failure Assessment) score is evaluated, and in the SOFA
baseline. These three variants form the comparison for the label-dominance
hypothesis (H1), which states that the spread in AUROC across label variants is
at least as large as the spread across model classes. The label-variance table
in this notebook describes how far the three labels differ before any model is
fitted.

Three further variants are built for analyses that lie outside H1. Protocol
Amendment 2 labels the two parts of the Sepsis-3 definition separately.
Variant E (Anchor-Only) takes the suspected-infection time as the onset and
does not require organ dysfunction. Variant D (Deterioration) requires acute
organ dysfunction but uses no infection criterion and no treatment timestamp.

Protocol Amendment 4 adds Variant F (Sepsis-3 Early-Onset). It uses Variant
B's suspected-infection windows and SOFA criterion, so it labels exactly the
same stays as septic. Only its onset time differs. That time is assigned in
`03_person_hours.Rmd` as $\min(t_{\text{susp}}, t_{\text{SOFA}})$, the
earlier of the suspected-infection time and the first hour at which the SOFA
rise is present. This is the onset convention used by Seymour et al. (2016) and
the PhysioNet/CinC 2019 Challenge (Reyna et al., 2020). Variants D, E and F are
reported separately and never enter the H1 comparison.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(dplyr)
})

source("config.R")
source("utils.R")

step(sprintf("=== Phase 2a: Cohort extraction under %d label variants ===",
             length(active_variants())))
```

## Base Tables

The `patients`, `admissions` and `icustays` tables supply patient
demographics and anchor year group, hospital-admission details including
in-hospital death, and the ICU admission and discharge times that bound each
stay.

```{r load-base-tables}
step("Loading base tables...")

patients <- read_mimic(hosp_file("patients.csv"),
                       cols = c("subject_id", "gender", "anchor_age",
                                "anchor_year", "anchor_year_group"))

admissions <- read_mimic(hosp_file("admissions.csv"),
                         cols = c("subject_id", "hadm_id", "admittime",
                                  "dischtime", "race", "language",
                                  "insurance", "hospital_expire_flag"))
admissions[, admittime := parse_dt(admittime)]
admissions[, dischtime := parse_dt(dischtime)]

icustays <- read_mimic(icu_file("icustays.csv"),
                       cols = c("subject_id", "hadm_id", "stay_id",
                                "intime", "outtime", "los"))
icustays[, intime    := parse_dt(intime)]
icustays[, outtime   := parse_dt(outtime)]
icustays[, los_hours := hours_diff(intime, outtime)]
```

## Base Cohort Criteria

Three inclusion criteria are applied in order: age of at least 18 years, the
earliest ICU stay of each patient (ordered by ICU admission time), and an ICU
length of stay (LOS) of at least four hours. Age is MIMIC-IV's `anchor_age`,
the patient's age in their anchor year. MIMIC-IV's ICU module records adult
patients only, so the age criterion is expected to exclude no stays; it is
still applied so that its count is reported. Because one stay is kept per
patient, the final cohort has as many stays as patients.

The table below gives, for each criterion, the number of stays remaining and
the number removed by that criterion alone. The same counts are written to
`02_cohort_flow`, from which the thesis cohort flow diagram is generated.

```{r build-base-cohort}
step("Building base cohort (adult, first ICU stay, min LOS)...")

stays <- merge(icustays, patients, by = "subject_id", all.x = TRUE)
stays <- merge(stays, admissions[, .(hadm_id, race, language, insurance,
                                      hospital_expire_flag)],
               by = "hadm_id", all.x = TRUE)

# The counts at each rung are captured as they are applied. The CONSORT figure
# in methodology.tex used to hardcode them and its caption said so ("from the
# pipeline's cohort extraction log"), which put five numbers in the thesis that
# no re-run could correct. Written to a result table so the figure can be
# driven from macros like every other quantity.
consort <- list(data.table(step = "icu_stays_total",
                           label = "MIMIC-IV v3.1 ICU stays",
                           n = nrow(stays), n_excluded = NA_integer_))

n_before <- nrow(stays)
stays    <- stays[anchor_age >= MIN_AGE]
consort  <- c(consort, list(data.table(
  step = "adult", label = sprintf("Age >= %s at admission", MIN_AGE),
  n = nrow(stays), n_excluded = n_before - nrow(stays))))

setorderv(stays, c("subject_id", "intime"))
stays[, icu_rank := seq_len(.N), by = subject_id]
n_before <- nrow(stays)
if (FIRST_ICU_STAY_ONLY) stays <- stays[icu_rank == 1]
consort  <- c(consort, list(data.table(
  step = "first_icu_stay", label = "First ICU stay per patient",
  n = nrow(stays), n_excluded = n_before - nrow(stays))))

n_before <- nrow(stays)
stays    <- stays[los_hours >= MIN_ICU_LOS_HOURS]
consort  <- c(consort, list(data.table(
  step = "min_los", label = sprintf("ICU LOS >= %s hours", MIN_ICU_LOS_HOURS),
  n = nrow(stays), n_excluded = n_before - nrow(stays))))

consort_dt <- rbindlist(consort)
consort_dt[, n_patients := NA_integer_]
consort_dt[.N, n_patients := stays[, uniqueN(subject_id)]]
save_parquet(consort_dt, "02_cohort_flow", DATA_DIR)
save_csv(consort_dt,     "02_cohort_flow", DATA_DIR)
print(consort_dt)

cat(sprintf("Base cohort: %s stays, %s unique patients\n",
            format(nrow(stays), big.mark = ","),
            format(stays[, uniqueN(subject_id)], big.mark = ",")))
```

## Suspected Infection

Sepsis-3 (Singer et al., 2016) defines sepsis as life-threatening organ
dysfunction caused by a dysregulated host response to infection. Following
Seymour et al. (2016), suspected infection is identified from an antibiotic
order and a microbiology culture placed within a set interval of each other.
Antibiotic orders come from `prescriptions`: systemic antibiotics matched by
drug name, timed by the order start. Cultures come from `microbiologyevents`.
Screening specimens listed in `SURVEILLANCE_SPEC_TYPES` in `config.R`, such as
MRSA and CRE screens, are excluded, and a culture recorded with a date but no
time is placed at midnight on that date.

Every antibiotic order is paired with every culture from the same hospital
admission. The number of pairs is therefore much larger than the number of
admissions and is not an event count. Each variant's time windows are applied
to these pairs when its labels are constructed.

```{r suspected-infection}
step("Identifying suspected infection episodes...")

prescriptions <- read_mimic(hosp_file("prescriptions.csv"),
                             cols = c("hadm_id", "starttime", "stoptime",
                                      "drug", "route"))
prescriptions[, starttime := parse_dt(starttime)]
prescriptions[, is_abx := match_antibiotics(drug, ANTIBIOTICS)]
abx <- prescriptions[is_abx == TRUE & !is.na(starttime),
                     .(hadm_id, abx_time = starttime)]
abx <- unique(abx)
cat(sprintf("Antibiotic orders: %s\n", format(nrow(abx), big.mark = ",")))

microbio <- read_mimic(hosp_file("microbiologyevents.csv"),
                       cols = c("hadm_id", "charttime", "chartdate",
                                "spec_type_desc"))
microbio[, charttime := parse_dt(charttime)]
if (nrow(microbio) > 0)
  microbio <- microbio[!(spec_type_desc %in% SURVEILLANCE_SPEC_TYPES)]
microbio[is.na(charttime) & !is.na(chartdate),
         charttime := as.POSIXct(chartdate, format = "%Y-%m-%d", tz = "UTC")]
cultures <- microbio[!is.na(charttime), .(hadm_id, culture_time = charttime)]
cultures <- unique(cultures)
cat(sprintf("Culture samples: %s\n", format(nrow(cultures), big.mark = ",")))

abx[, abx_id := .I]
cultures[, cul_id := .I]

suspicion_raw <- merge(abx, cultures, by = "hadm_id", allow.cartesian = TRUE)
suspicion_raw[, dt_abx_before_cul := hours_diff(abx_time, culture_time)]
suspicion_raw[, dt_cul_before_abx := hours_diff(culture_time, abx_time)]
cat(sprintf("ABX-culture pairs (raw): %s\n",
            format(nrow(suspicion_raw), big.mark = ",")))
```

## SOFA Components

The SOFA score (Vincent et al., 1996) grades dysfunction in six organ
systems. At this stage it is computed from laboratory values (PaO$_2$/FiO$_2$
ratio, platelets, bilirubin and creatinine) and from charted mean arterial
pressure (MAP) and Glasgow Coma Scale (GCS) scores. The large `labevents` and
`chartevents` tables are read only for hospital admissions in the base cohort.
Vasopressor infusions are also read, but they are not passed to the score at
this stage, so the cardiovascular component is graded from MAP alone and the
renal component from creatinine alone.

```{r sofa-components}
step("Computing SOFA components for all stays...")

cohort_hadm_ids <- stays[, unique(hadm_id)]
cat(sprintf("Pre-filtering large tables to %d cohort hadm_ids...\n",
            length(cohort_hadm_ids)))

lab_sofa_items <- c(LAB_ITEMIDS$platelets, LAB_ITEMIDS$bilirubin_total,
                    LAB_ITEMIDS$creatinine, LAB_ITEMIDS$pao2, LAB_ITEMIDS$fio2_lab)

v2_log("Loading labevents for SOFA components via DuckDB...")
labevents <- read_large_csv_filtered(
  path     = hosp_file("labevents.csv"),
  cols     = c("hadm_id", "itemid", "charttime", "valuenum"),
  itemids  = lab_sofa_items,
  hadm_ids = cohort_hadm_ids
)
labevents[, charttime := parse_dt(charttime)]
setkeyv(labevents, c("hadm_id", "charttime"))

chart_sofa_items <- c(CHART_ITEMIDS$map_invasive, CHART_ITEMIDS$map_noninvasive,
                      CHART_ITEMIDS$gcs_eye, CHART_ITEMIDS$gcs_verbal,
                      CHART_ITEMIDS$gcs_motor)

v2_log("Loading chartevents for SOFA components (MAP, GCS) via DuckDB...")
chartevents_sofa <- read_large_csv_filtered(
  path     = icu_file("chartevents.csv"),
  cols     = c("stay_id", "hadm_id", "itemid", "charttime", "valuenum"),
  itemids  = chart_sofa_items,
  hadm_ids = cohort_hadm_ids
)
chartevents_sofa[, charttime := parse_dt(charttime)]
setkeyv(chartevents_sofa, c("hadm_id", "charttime"))

v2_log("Loading inputevents for vasopressors...")
inputevents <- read_mimic(icu_file("inputevents.csv"),
                          cols = c("stay_id", "hadm_id", "itemid",
                                   "starttime", "endtime", "amount",
                                   "amountuom", "rate", "rateuom"))
inputevents <- inputevents[itemid %in% unlist(VASOPRESSOR_ITEMIDS) &
                             hadm_id %in% cohort_hadm_ids]
inputevents[, starttime := parse_dt(starttime)]
inputevents[, endtime   := parse_dt(endtime)]
setkeyv(inputevents, c("hadm_id", "starttime"))
```

## SOFA Change from Baseline

A stay meets the organ-dysfunction criterion when its SOFA score in the
evaluation window is at least two points above its baseline score. The
evaluation window runs from a set number of hours before the suspected-infection
time to a set number after it, with both bounds defined per variant in
`config.R`. The baseline window is either the first four hours of the ICU stay
(admission baseline, Variants A and B) or the 24 hours before the evaluation
window opens (Variant C). Within each window, the measurements of each item are
averaged and the averages are scored. Missing laboratory and MAP values are
scored as normal, and a window with no measurements at all scores zero.
Measurements are matched to windows with interval joins
(`data.table::foverlaps`).

```{r sofa-delta, code_folding="hide"}
source("clinical_scores.R")

lab_pt <- labevents[, .(hadm_id, t_start = charttime, t_end = charttime,
                         itemid, valuenum)]
setkeyv(lab_pt, c("hadm_id", "t_start", "t_end"))

chart_pt <- chartevents_sofa[, .(hadm_id, t_start = charttime, t_end = charttime,
                                  itemid, valuenum)]
setkeyv(chart_pt, c("hadm_id", "t_start", "t_end"))

vaso_iv <- inputevents[, .(hadm_id, t_start = starttime,
                             t_end = fifelse(is.na(endtime), starttime + 3600, endtime))]
setkeyv(vaso_iv, c("hadm_id", "t_start", "t_end"))

sofa_from_agg <- function(agg_dt) {
  w   <- dcast(agg_dt, stay_id ~ itemid, value.var = "valuenum",
               fun.aggregate = mean, na.rm = TRUE)
  ids <- names(w)[names(w) != "stay_id"]

  get_v <- function(iids) {
    cs <- intersect(as.character(as.integer(iids)), ids)
    if (length(cs) == 0) return(rep(NA_real_, nrow(w)))
    if (length(cs) == 1) return(as.numeric(w[[cs]]))
    rowMeans(w[, ..cs], na.rm = TRUE)
  }

  pao2_v <- get_v(LAB_ITEMIDS$pao2)
  fio2_v <- get_v(LAB_ITEMIDS$fio2_lab)
  fio2_f <- ifelse(!is.na(fio2_v) & fio2_v > 1, fio2_v / 100, fio2_v)
  pf     <- ifelse(!is.na(pao2_v) & !is.na(fio2_f) & fio2_f > 0,
                   pao2_v / fio2_f, NA_real_)

  panel <- data.frame(
    pf_ratio   = pf,
    platelets  = get_v(LAB_ITEMIDS$platelets),
    bilirubin  = get_v(LAB_ITEMIDS$bilirubin_total),
    creatinine = get_v(LAB_ITEMIDS$creatinine),
    map = {
      mi <- get_v(CHART_ITEMIDS$map_invasive)
      mn <- get_v(CHART_ITEMIDS$map_noninvasive)
      pmin(mi, mn, na.rm = TRUE)
    },
    gcs = {
      e  <- get_v(CHART_ITEMIDS$gcs_eye)
      v2 <- get_v(CHART_ITEMIDS$gcs_verbal)
      m  <- get_v(CHART_ITEMIDS$gcs_motor)
      rowSums(cbind(e, v2, m), na.rm = TRUE)
    },
    vaso_any       = FALSE,
    norepi_epi_any = FALSE
  )
  s <- score_sofa(panel)
  data.table(stay_id = w$stay_id, sofa_total = s$sofa_total)
}

compute_sofa_delta_vectorized <- function(rows, v) {
  if (nrow(rows) == 0) return(list(sofa_window = numeric(0), sofa_base = numeric(0)))

  eval_wins <- rows[, .(hadm_id, stay_id,
                         t_start = sofa_window_start, t_end = sofa_window_end,
                         wtype = "eval")]
  if (v$sofa_baseline == "admission") {
    base_wins <- rows[, .(hadm_id, stay_id,
                           t_start = intime, t_end = intime + 3600 * 4,
                           wtype = "base")]
  } else {
    base_wins <- rows[, .(hadm_id, stay_id,
                           t_start = sofa_window_start - 3600 * 24,
                           t_end   = sofa_window_start,
                           wtype = "base")]
  }
  all_wins <- rbindlist(list(eval_wins, base_wins))
  all_wins <- all_wins[!is.na(t_start) & !is.na(t_end) & t_end >= t_start]
  setkeyv(all_wins, c("hadm_id", "t_start", "t_end"))

  lab_m   <- foverlaps(lab_pt,   all_wins, type = "within", nomatch = 0L)
  chart_m <- foverlaps(chart_pt, all_wins, type = "within", nomatch = 0L)
  vaso_m  <- foverlaps(vaso_iv,  all_wins, type = "any",    nomatch = 0L)

  all_m <- rbindlist(list(
    lab_m[,   .(stay_id, wtype, itemid, valuenum)],
    chart_m[, .(stay_id, wtype, itemid, valuenum)]
  ))
  if (nrow(all_m) == 0) {
    zeros <- data.table(stay_id = rows$stay_id, sofa_total = 0)
    return(list(sofa_window = zeros$sofa_total, sofa_base = zeros$sofa_total))
  }

  agg <- all_m[, .(valuenum = mean(valuenum, na.rm = TRUE)),
               by = .(stay_id, wtype, itemid)]
  agg[, itemid := as.character(itemid)]

  if (nrow(vaso_m) > 0)
    vaso_flag <- unique(vaso_m[, .(stay_id, wtype)])[, vaso_any := TRUE]

  compute_for_type <- function(wt) {
    sub <- agg[wtype == wt]
    if (nrow(sub) == 0) return(data.table(stay_id = rows$stay_id, sofa_total = 0))
    res   <- sofa_from_agg(sub)
    all_s <- data.table(stay_id = rows$stay_id)
    res   <- merge(all_s, res, by = "stay_id", all.x = TRUE)
    res[is.na(sofa_total), sofa_total := 0]
    res
  }

  eval_sofa <- compute_for_type("eval")
  base_sofa <- compute_for_type("base")
  list(sofa_window = eval_sofa$sofa_total, sofa_base = base_sofa$sofa_total)
}
```

## Variant-Specific Label Construction

For each variant, an antibiotic-culture pair is kept when the antibiotic was
ordered no more than the variant's limit before the culture, or the culture
was drawn no more than its limit before the antibiotic. The suspected-infection
time of a pair is the later of its two timestamps, and only the earliest
suspected-infection time in each admission is evaluated. For Variants A, B, C
and F, that time becomes the onset when the SOFA criterion is met. For Variant
E, it becomes the onset without the SOFA criterion. Variant D receives no
onset here, because its onsets are derived hour by hour in
`03_person_hours.Rmd`.

Two rules then discard onsets that cannot be used: an onset before ICU
admission or after ICU discharge, and an onset in hour 0 or hour 1 of the stay
(less than two hours after ICU admission). A stay whose onset is discarded
remains in the cohort without an onset; it is never removed. Every variant
therefore keeps the full base cohort, and each event rate below uses all
base-cohort stays as its denominator. Variants B and F report identical counts
because they label the same stays.

```{r build-cohorts}
build_cohort_for_variant <- function(variant_id) {
  v    <- variant_def(variant_id)
  ty   <- variant_type(variant_id)
  step(sprintf("Building cohort for Variant %s (%s)...", variant_id, v$name))

  # Variant D carries no suspicion anchor at all: onset is derived per hour from
  # physiology in 03_person_hours.Rmd, where the hourly panel exists. The cohort
  # is the same base cohort with an empty onset column.
  if (ty == "deterioration") {
    cohort <- copy(stays)
    cohort[, `:=`(t_susp = as.POSIXct(NA), onset_time = as.POSIXct(NA),
                  onset_hour = NA_integer_,
                  died_in_hosp = hospital_expire_flag == 1L,
                  label_variant = variant_id, label_name = v$name)]
    cat(sprintf("Variant %s: %d total stays, onsets derived in stage 03\n",
                variant_id, nrow(cohort)))
    return(cohort)
  }

  susp <- suspicion_raw[
    (dt_abx_before_cul >= 0 & dt_abx_before_cul <= v$abx_before_culture_h) |
    (dt_cul_before_abx >= 0 & dt_cul_before_abx <= v$culture_before_abx_h)
  ]
  susp[dt_abx_before_cul >= 0,                         t_susp := pmax(abx_time, culture_time)]
  susp[dt_cul_before_abx >= 0 & is.na(t_susp),         t_susp := pmax(abx_time, culture_time)]
  susp <- susp[!is.na(t_susp)]
  setorderv(susp, c("hadm_id", "t_susp"))
  susp_first <- susp[, .SD[1], by = hadm_id]

  cohort <- merge(stays, susp_first[, .(hadm_id, t_susp)],
                  by = "hadm_id", all.x = TRUE)

  v2_log(sprintf("  Stays with suspected infection: %s / %s",
                 format(cohort[!is.na(t_susp), .N], big.mark = ","),
                 format(nrow(cohort), big.mark = ",")))

  # Variant E is the anchor limb on its own: the suspicion time *is* the onset,
  # with no organ-dysfunction requirement.
  if (ty == "anchor_only") {
    cohort[, onset_time := t_susp]
  } else {

  cohort[, sofa_window_start := t_susp - lubridate::hours(v$sofa_window_before_h)]
  cohort[, sofa_window_end   := t_susp + lubridate::hours(v$sofa_window_after_h)]

  susp_idx <- which(!is.na(cohort$t_susp))
  if (length(susp_idx) > 0) {
    v2_log(sprintf("  Computing SOFA delta (vectorized) for %d stays...",
                   length(susp_idx)))
    sofa_res <- compute_sofa_delta_vectorized(cohort[susp_idx], v)
    cohort[susp_idx, sofa_in_window := sofa_res$sofa_window]
    cohort[susp_idx, sofa_baseline  := sofa_res$sofa_base]
    cohort[susp_idx, sofa_delta     := sofa_in_window - sofa_baseline]
    cohort[susp_idx,
           onset_time := fifelse(sofa_delta >= SOFA_INCREASE_THRESHOLD,
                                 t_susp, as.POSIXct(NA))]
  }

  }

  cohort[!is.na(onset_time) & (onset_time < intime | onset_time > outtime),
         onset_time := NA_real_]
  cohort[!is.na(onset_time),
         onset_hour := as.integer(hours_diff(intime, onset_time))]
  cohort[!is.na(onset_hour) & onset_hour <= 1, onset_time := NA_real_]
  cohort[!is.na(onset_hour) & onset_hour <= 1, onset_hour := NA_integer_]

  cohort[, died_in_hosp   := hospital_expire_flag == 1L]
  cohort[, label_variant  := variant_id]
  cohort[, label_name     := v$name]

  n_events <- cohort[!is.na(onset_time), .N]
  cat(sprintf("Variant %s: %d total stays, %d sepsis onsets (%.1f%%)\n",
              variant_id, nrow(cohort), n_events,
              100 * n_events / nrow(cohort)))
  cohort
}

run_variants   <- active_variants()
cohorts        <- lapply(run_variants, build_cohort_for_variant)
names(cohorts) <- run_variants

for (vid in names(cohorts))
  save_parquet(cohorts[[vid]], paste0("02_cohort_", vid), DATA_DIR)
```

## Label-Variance Table (Variants A--C)

The label-variance table reports, for each pre-registered variant, the number
of stays, the number and percentage with a sepsis onset, the number without
one, in-hospital deaths among stays without an onset, and the median and
interquartile range of the onset hour. Because the three variants share the
same stays, every difference in the table comes from the label definition. A
large spread in event rate or onset timing across A--C shows that the labels
disagree. It does not test H1, which concerns discrimination and is evaluated
in later stages. The table is saved as `02_label_variance_table.parquet` in
`output/processed_data/`.

```{r label-variance-table}
step("Building M1 label-variance table...")

summarise_variant <- function(vid) {
  c_dt <- cohorts[[vid]]
  data.frame(
    variant         = vid,
    name            = variant_name(vid),
    n_stays         = nrow(c_dt),
    n_sepsis_onsets = c_dt[!is.na(onset_time), .N],
    pct_sepsis      = round(100 * c_dt[!is.na(onset_time), .N] / nrow(c_dt), 1),
    n_non_sepsis    = c_dt[is.na(onset_time), .N],
    n_deaths_no_sep = c_dt[is.na(onset_time) & died_in_hosp == TRUE, .N],
    median_onset_h  = round(median(c_dt[!is.na(onset_hour), onset_hour], na.rm = TRUE), 1),
    iqr_onset_h_lo  = round(quantile(c_dt[!is.na(onset_hour), onset_hour], 0.25, na.rm = TRUE), 1),
    iqr_onset_h_hi  = round(quantile(c_dt[!is.na(onset_hour), onset_hour], 0.75, na.rm = TRUE), 1)
  )
}

prereg_run <- intersect(names(cohorts), names(LABEL_VARIANTS))
if (length(prereg_run) > 0) {
  label_variance <- do.call(rbind, lapply(prereg_run, summarise_variant))
  knitr::kable(label_variance, caption = "M1 Label-Variance Table (cohort size × event count × onset timing)")
  save_parquet(label_variance, "02_label_variance_table", DATA_DIR)
  cat("\nSaved: 02_label_variance_table.parquet\n")
}
```

## Alternative-Label Variance Table (Variants D, E and F)

Variants A--C anchor suspected infection in the same way, on a clinician's
decision to culture and treat, and differ only in the windows around that
anchor and in the SOFA baseline. Part of their spread therefore follows from
the definition itself. Variants E and D separate the two parts of the
definition: Variant E keeps the suspected-infection anchor and drops the
organ-dysfunction requirement, while Variant D keeps organ dysfunction and
drops the anchor. The table has the same columns as the label-variance table,
plus the label family of each variant.

Two rows need care. Variant D's onset count, percentage and non-onset count are
left empty, because its onsets are assigned hourly in `03_person_hours.Rmd` and
none exist at this stage. Variant F's counts and onset hours match Variant B's,
because its earlier onset time is also assigned in `03_person_hours.Rmd`. The
table is saved as `02_alt_label_variance_table.parquet` in
`output/processed_data/`.

```{r alt-label-variance-table}
alt_run <- intersect(names(cohorts), names(ALT_LABEL_VARIANTS))
if (length(alt_run) > 0) {
  alt_variance <- do.call(rbind, lapply(alt_run, summarise_variant))
  alt_variance$label_family <- vapply(alt_run, variant_type, character(1))
  # Variant D's onsets are assigned per hour in stage 03, so summarise_variant()
  # counts zero of them here. A literal 0 in a column headed "sepsis onsets" is
  # indistinguishable from a real finding to anything that reads this file, and
  # the true panel count is 27,209. Blank it so a consumer gets a visible
  # missing marker rather than a plausible wrong number.
  d_row <- alt_variance$variant %in% "D"
  if (any(d_row))
    alt_variance[d_row, c("n_sepsis_onsets", "pct_sepsis", "n_non_sepsis")] <- NA
  knitr::kable(alt_variance,
               caption = "Alternative (non-Sepsis-3) label variants; Variant D onsets are assigned in stage 03")
  save_parquet(alt_variance, "02_alt_label_variance_table", DATA_DIR)
  cat("\nSaved: 02_alt_label_variance_table.parquet\n")
} else {
  cat("No alternative-label variants selected in this run.\n")
}
```
