---
title: "04 Sample Size - Minimum Sample Size for Model Development"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook checks whether the training data are large enough to develop
the primary model. It applies the minimum sample-size criteria of Riley et al.
(2019) for clinical prediction models, as implemented in the `pmsampsize`
package. These criteria are not a power calculation for a hypothesis test. For a
binary outcome, they give the smallest sample that meets three conditions: an
expected global shrinkage factor of at least 0.90; a difference of no more than
0.05 between the apparent and the optimism-adjusted Nagelkerke $R^2$; and
estimation of the overall outcome proportion within a margin of error of 0.05.

The criteria assume independent observations. Person-hours are not
independent, because each ICU stay contributes many correlated rows. The
primary assessment therefore treats the ICU stay as the unit of analysis, and
because the cohort keeps one stay per patient, this is also the patient level.
For each variant, the outcome proportion passed to `pmsampsize` is the
proportion of training stays with an onset, and the resulting minimum is
compared with the number of training stays.

A second calculation uses the person-hour event proportion instead. It shows
how much larger the nominal minimum becomes at the much lower hourly
proportion. It does not account for the correlation between rows of the same
stay, so its result is not an effective sample size and is not comparable with
the number of person-hour rows. It is not used for the adequacy conclusion.

The calculation is repeated for each pre-registered variant (A--C), because
their outcome proportions differ. The alternative variants D--F are not
assessed here.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(pmsampsize)
})

source("config.R")
source("utils.R")

step("=== Sample Size Calculation (Riley et al., 2019) ===")
```

## Label-Variance Recap

The label-variance table from `02_cohort_labels.Rmd` lists the variants to be
assessed, with their cohort-wide counts, and is shown here for context. The
calculation below does not use these counts directly. It counts training stays
of the temporal split only, and only onsets that fall inside the
`r MAX_PANEL_HOURS`-hour panel, so its outcome proportions are lower than the
cohort-wide percentages in this table.

```{r load-label-variance}
lv_path <- file.path(DATA_DIR, "02_label_variance_table.parquet")
if (!file.exists(lv_path))
  stop("02_label_variance_table.parquet not found. Run 02_cohort_labels.Rmd first.")

lv <- as.data.table(arrow::read_parquet(lv_path))
knitr::kable(lv, caption = "Label-variance table (from 02\\_cohort\\_labels.Rmd)")
```

## Sample Size by Variant

For each variant, the training rows of the temporal split, the primary split
built in `03_person_hours.Rmd`, are taken from the person-hour table. Positive
training stays are counted from onset rows. Each stay has at most one onset
row, so this count equals the number of training stays whose onset falls within
the panel.

Two further inputs are the same for every variant. The anticipated C-statistic
is `r ANTICIPATED_AUC`, the internal AUROC reported by Moor et al. (2023). The
number of candidate parameters is `r P_CANDIDATE_PREDICTORS`: the terms the
primary model estimates for one outcome, namely its covariates
(`BASE_FEATURES` and `MISSINGNESS_FEATURES` in `config.R`), a natural spline in
the hour of stay, and the intercept. Because the primary model is multinomial,
it estimates this set of terms for each of its three non-reference outcomes.
Using the per-outcome count treats each cause-specific sub-model as the unit; a
count for the whole fit would be three times larger. The required sample size
grows roughly in proportion to the parameter count, so any variant whose ratio
of available to required stays exceeds three would still meet the criterion
under that larger count.

For each variant, the output gives the training person-hours, training stays,
positive training stays, the hourly event proportion and the stay-level outcome
proportion. It then reports the stay-level minimum (primary), the ratio of
training stays to that minimum (printed as the margin), and the minimum at the
hourly proportion (reference only). The events per candidate parameter (printed
as EPV) is the number of positive training stays divided by
`r P_CANDIDATE_PREDICTORS`. It is reported for reference and is not an adequacy
criterion.

```{r sample-size-calc}
results <- list()

for (i in seq_len(nrow(lv))) {
  row <- lv[i]
  vid <- row$variant

  ph_path <- file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
  if (!file.exists(ph_path)) {
    v2_log(sprintf("  Variant %s: person_hours not found, skipping", vid), level = "WARN")
    next
  }

  ph       <- arrow::read_parquet(ph_path, col_select = c("stay_id", "event", "temporal_split"))
  ph_train <- ph[ph$temporal_split == "train", ]

  n_person_hours <- nrow(ph_train)
  n_events       <- sum(ph_train$event == 1, na.rm = TRUE)
  n_stays        <- length(unique(ph_train$stay_id))
  event_rate_ph  <- n_events / n_person_hours

  cat(sprintf("\n--- Variant %s (%s) ---\n", vid, row$name))
  cat(sprintf("  Training person-hours: %s\n", format(n_person_hours, big.mark = ",")))
  cat(sprintf("  Training stays:        %s\n", format(n_stays,         big.mark = ",")))
  cat(sprintf("  Onset events (train):  %s\n", format(n_events,        big.mark = ",")))
  cat(sprintf("  Per-hour event rate:   %.4f%%\n", 100 * event_rate_ph))

  # Stay-level prevalence: the fraction of INDEPENDENT units carrying the
  # outcome. This is the parameterisation Riley et al.'s derivation assumes.
  stay_prevalence <- n_events / n_stays
  cat(sprintf("  Stay-level sepsis prevalence: %.4f%%\n", 100 * stay_prevalence))

  riley <- function(prev) {
    tryCatch(
      pmsampsize::pmsampsize(
        type       = "b",
        rsquared   = NULL,
        parameters = P_CANDIDATE_PREDICTORS,
        shrinkage  = 0.9,
        prevalence = prev,
        cstatistic = ANTICIPATED_AUC,
        seed       = 42
      ),
      error = function(e) {
        v2_log(sprintf("  pmsampsize error at prevalence %.6f: %s", prev, e$message),
               level = "WARN")
        NULL
      }
    )
  }

  # PRIMARY: the independent unit is the stay. This is the adequacy verdict.
  ss_stay <- riley(stay_prevalence)
  # SECONDARY, reference only: the per-hour parameterisation. Retained so the
  # two can be seen side by side and never confused. It is NOT an effective
  # sample size and must not be compared against the person-hour row count.
  ss_ph   <- riley(event_rate_ph)

  if (!is.null(ss_stay)) {
    min_n_stay <- ss_stay$sample_size
    cat(sprintf("  [PRIMARY] Minimum independent observations (Riley, stay-level): %s\n",
                format(min_n_stay, big.mark = ",")))
    cat(sprintf("            Available independent units (training stays):        %s\n",
                format(n_stays, big.mark = ",")))
    cat(sprintf("            Sufficient: %s (margin %.2fx)\n",
                if (n_stays >= min_n_stay) "YES" else "NO - variant may be underpowered",
                n_stays / min_n_stay))
  }
  if (!is.null(ss_ph)) {
    cat(sprintf("  [reference] Minimum at the per-hour prevalence: %s\n",
                format(ss_ph$sample_size, big.mark = ",")))
    cat("             Rows are clustered within stays, so the person-hour count\n")
    cat("             does not supply that many independent observations.\n")
  }

  epv <- if (P_CANDIDATE_PREDICTORS > 0) n_events / P_CANDIDATE_PREDICTORS else NA
  cat(sprintf("  Events-per-variable (EPV): %.1f\n", epv))

  results[[vid]] <- data.frame(
    variant             = vid,
    name                = row$name,
    # --- independent unit = the stay; this is the adequacy verdict ---
    n_train_stays       = n_stays,
    n_onset_events      = n_events,
    stay_prevalence     = round(stay_prevalence, 5),
    min_n_riley_stay    = if (!is.null(ss_stay)) ss_stay$sample_size else NA,
    sufficient_stay     = if (!is.null(ss_stay)) n_stays >= ss_stay$sample_size else NA,
    margin_stay         = if (!is.null(ss_stay)) round(n_stays / ss_stay$sample_size, 2) else NA,
    epv                 = round(epv, 1),
    # --- design-matrix size, reported alongside; NOT a sample size ---
    n_train_person_h    = n_person_hours,
    per_hour_event_rate = round(event_rate_ph, 5),
    min_n_riley_perhour = if (!is.null(ss_ph)) ss_ph$sample_size else NA,
    anticipated_auc     = ANTICIPATED_AUC,
    n_candidate_preds   = P_CANDIDATE_PREDICTORS
  )
}
```

## Summary and Interpretation

The summary table gives, for each variant, the training stays, positive
training stays, stay-level outcome proportion, minimum required number of
stays, whether that minimum is met, the ratio of available to required stays,
and the events per candidate parameter. A ratio above 1 means the stay-level
minimum is met. The remaining columns (training person-hours, the hourly event
proportion and the minimum at that proportion) describe the size of the design
matrix and are not used in the adequacy conclusion.

Meeting the minimum shows that the sample is large enough under the stated
assumptions: the anticipated C-statistic, the parameter count, the outcome
proportion and the shrinkage target. It does not establish that the model will
be well calibrated or discriminate well, and it does not address performance in
other settings, which stage 08 examines. The results are saved as
`04_sample_size.parquet` in `output/processed_data/`.

```{r summary-table}
if (length(results) > 0) {
  result_df <- do.call(rbind, results)
  knitr::kable(result_df,
               caption = "Sample size adequacy (Riley et al., 2019) per label variant")
  save_parquet(result_df, "04_sample_size", DATA_DIR)
  cat("\nSaved: 04_sample_size.parquet\n")
}
```
