---
title: "05 Primary Model - Dynamic Discrete-Time Competing-Risks Hazard Model"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook fits the primary model of the study, a multinomial pooled
logistic regression for discrete-time competing events (D'Agostino et al.,
1990; Heyard et al., 2019). It is the pre-registered main model: the
coefficient-stability hypothesis (H2), the split hypothesis (H4) and the
external-validation hypothesis (H5) are all evaluated on it. It is also the
interpretable benchmark for H6, which compares its temporal-test AUROC (area
under the receiver operating characteristic curve) with that of the
gradient-boosted trees (GBT) fitted in `06_comparators.Rmd`. H6's
pre-registered criterion is a difference of no more than 0.02 in AUROC. If the
criterion is met, the boosted trees give no material gain in discrimination
over the regression in this setting. That would not show that the more complex
model has no value for other outcomes or other data.

For patient $i$ at hour $t$, with competing events
$r \in \{1\text{=onset}, 2\text{=discharge}, 3\text{=death}\}$:

$$\lambda_r(t \mid \mathbf{X}_{it}) = \frac{\exp\bigl(\alpha_r(t) + \boldsymbol{\beta}_r^\top \mathbf{X}_{it}\bigr)}{1 + \sum_{s=1}^{3} \exp\bigl(\alpha_s(t) + \boldsymbol{\beta}_s^\top \mathbf{X}_{it}\bigr)}$$

The reference outcome, which gives the 1 in the denominator, is that the stay
remains at risk with no event in hour $t$ (state 0 in `03_person_hours.Rmd`).
The three events are mutually exclusive within an hour, and a stay leaves the
risk set at its first event. Each $\lambda_r$ is therefore a discrete-time
cause-specific hazard: the probability of event $r$ in hour $t$, given that no
event has occurred before $t$. The three hazards are estimated jointly in one
multinomial fit rather than as three separate models. Event 1 is the onset
defined by the label variant, which is Sepsis-3 onset for Variants A--C and F.

$\alpha_r(t)$ is the baseline log-odds of event $r$ against remaining at risk.
It is modelled as a natural cubic spline in hours since ICU admission with five
interior knots, so each event has an intercept and six spline coefficients.
$\boldsymbol{\beta}_r$ holds the covariate coefficients for event $r$, so each
covariate has a separate coefficient for each event. A small ridge penalty is
applied to the coefficients (Protocol Amendment 3), because without it the
likelihood is separated and the coefficients are not identified. Standard
errors are cluster-robust sandwich estimates, clustered by ICU stay
(`stay_id`). Because the cohort keeps one stay per patient, this is also
clustering by patient. With the penalty in place, these are the standard
errors of the penalised estimator.

The model is fitted separately for each label variant. Variants A (Narrow),
B (Seymour-Standard) and C (Liberal) are the pre-registered Sepsis-3
definitions. They differ in the suspicion-of-infection windows, the SOFA
evaluation window and the SOFA baseline, as defined in `02_cohort_labels.Rmd`.
Variant B is the primary analysis, and H2 compares its coefficients with those
of A and C. Three post-hoc variants are fitted by the same procedure. Variants
D (Deterioration) and E (Anchor-Only), from Protocol Amendment 2, each keep one
limb of the Sepsis-3 definition. Variant F (Sepsis-3 Early-Onset), from
Protocol Amendment 4, keeps Variant B's patients but times the onset at the
earlier of suspicion of infection and the SOFA rise. The coefficients of D--F
are written to a separate file and are excluded from H2.

The procedure and the covariates are the same for every variant, but the
fitted rows are not. Because a stay leaves the risk set at its onset, the label
determines how many person-hours each stay contributes and how many onsets
there are. Through the training hours, it also determines where the spline
knots fall. Differences between the variants' fits therefore reflect the label
together with these consequences of it.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(nnet)
  library(rms)
  library(survival)
  library(sandwich)
  library(lmtest)
  library(dplyr)
  library(splines)
})

source("config.R")
source("utils.R")

step("=== Phase 3a: Primary discrete-time competing-risks model ===")
```

## Feature Set

The 22 covariates are declared once in `config.R`, as `BASE_FEATURES` and
`MISSINGNESS_FEATURES`, and are printed below under their column names. They
enter the model on their recorded scales, without centring or
standardisation, so each coefficient is a change in log-odds per unit of its
covariate. They are:

- vital signs: heart rate (`hr`, beats/min), respiratory rate (`resp_rate`,
  breaths/min), oxygen saturation (`spo2`, %), mean arterial pressure (`map`,
  mmHg), temperature (`temp_c`, °C) and the Glasgow Coma Scale total (`gcs`,
  3--15);
- laboratory values: lactate (mmol/L), creatinine (mg/dL), total bilirubin
  (`bilirubin_total`, mg/dL), platelets ($\times 10^9$/L) and white blood cell
  count (`wbc`, $\times 10^9$/L);
- vasopressor exposure: `vaso_any` is true in any hour in which a vasopressor
  infusion is running, and `norepi_epi_any` is true when that infusion is
  norepinephrine or epinephrine. The second is a subset of the first, so its
  coefficient is the additional association of these two agents beyond that of
  vasopressor exposure in general;
- six-hour means: `hr_slope6h`, `resp_rate_slope6h` and `map_slope6h` hold the
  mean of the forward-filled value over the current and five preceding hours.
  Despite their names they are trailing means, not regression slopes
  (`03_person_hours.Rmd`);
- six indicators (`*_measured`), for the five laboratory values and the
  PaO$_2$/FiO$_2$ ratio. Each is 1 from the first measurement of that test in
  the stay onwards and 0 before it.

Vital signs and laboratory values are forward-filled within the stay in
`03_person_hours.Rmd`, so a value is missing only in the hours before it is
first recorded. These remaining missing values are set to zero before fitting.
For each laboratory value, its indicator is 0 in exactly those hours. The
indicator's coefficient therefore absorbs the offset that the zero would
otherwise create, and the value's coefficient is estimated from measured hours.
The indicators record that a test was ordered, which reflects clinical
attention as well as the patient's state, so they are not physiological
measurements. The PaO$_2$/FiO$_2$ ratio enters only through its indicator. The
ratio itself is not a covariate, because it is recorded for a minority of
person-hours and its forward-filled value reflects ventilation practice more
than the current hour.

The vital signs and the six-hour means have no indicators. An hour before a
vital sign is first recorded therefore enters with a value of zero, which lies
outside every plausible range; this affects mainly the first hours of a stay.
The six-hour means are computed only over a complete six-hour window, so they
are missing, and are set to zero, in hours 0--4 of every stay. In those hours
the zero acts as a shared offset common to all stays, which the baseline spline
in hour of stay can partly absorb. It is not a measured value, and the
six-hour-mean coefficients should be read with this in mind.

```{r feature-set}
# BASE_FEATURES and MISSINGNESS_FEATURES are declared once, in config.R, and
# read from there by every stage that fits or transports a model. They used to
# be re-declared here, in 06 and in 08; see the comment at the declaration for
# why three copies of the covariate vector was the wrong arrangement.
cat("Base features:", paste(BASE_FEATURES, collapse = ", "), "\n")
cat("Missingness indicators:", paste(MISSINGNESS_FEATURES, collapse = ", "), "\n")
```

## Fit Primary Model by Variant

For each variant, the model is fitted to the training person-hours of the
temporal split. That split trains on the anchor-year groups up to 2014--2016
and tests on 2017--2019 (`03_person_hours.Rmd`). The outcome is coded as a
four-level factor with the at-risk state as the reference (`inprogress`).
Remaining missing values are set to zero, or to false for logical columns, as
described above. The model is fitted with `nnet::multinom`, with ridge decay
`r PRIMARY_MODEL_DECAY` and at most `r PRIMARY_MODEL_MAXIT` iterations.

The baseline spline has interior knots at the 10th, 30th, 50th, 70th and 90th
percentiles of the training hours. Its boundary knots are hour 0 and the last
training hour. The knots are recomputed for every fit from that fit's own
training rows, so they differ slightly between variants, as the printed
positions show, while the rule stays the same. The knots and the zero
imputation are determined from training rows only. The test rows receive the
same imputation and are scored with the fitted spline basis.

Cluster-robust standard errors are computed with `sandwich::vcovCL`, clustered
by `stay_id`. The `sandwich` package has no score function for `multinom`
fits, so `utils.R` supplies one (`estfun.multinom`). Predicted onset
probabilities for the temporal test set are saved, and `07_metrics_suite.Rmd`
turns these hourly predictions into performance measures.

If `nnet::multinom` stops with an error, the notebook falls back to a binary
logistic regression of onset against all other states, with the same
covariates and spline. That model has no competing events, so its estimates
are not comparable with the multinomial ones. It exists so that the pipeline
can complete, and a fallback fit is marked as such (`is_binary_fallback`).
Reaching the iteration limit is not an error: it is logged as a warning and the
multinomial fit is kept. Only multinomial fits print the fit diagnostics
described below, so a fit printed without them would be a fallback.

For each fit, the output prints the following:

- the number of training person-hours;
- the number of onset rows, printed as `events (sepsis)`, although for
  Variants D and E the event is not Sepsis-3 onset. Each stay has at most one
  onset row, so this is also the number of training stays with an onset in the
  panel;
- a check that all 22 declared covariates are present;
- the knot positions;
- three fit diagnostics.

`convergence=0` means only that the optimiser met its stopping rule before the
iteration limit. On its own, it does not show that the coefficients are
identified, because `nnet` reports the same code on a separated likelihood.
Identification comes from the ridge penalty, which makes the penalised
likelihood strictly concave. It is checked through `max|coef|`, the largest
absolute coefficient in the fit, with intercepts and spline terms included. A
value above 20 is logged as a warning of possible separation. In these fits the
largest coefficient is a spline-basis term of the death outcome, not a
covariate effect. `deviance` is twice the penalised negative log-likelihood at
the fitted values. The variants are fitted to different rows, so their
deviances are not comparable with one another.

When the ablated specification of Protocol Amendment 5 is run, its fits are
printed after the full ones, labelled `arm = ablated`. The printed output uses
"arm" for what this notebook calls a specification. The procedure is the same
except that it omits the eight action-derived covariates: `vaso_any`,
`norepi_epi_any` and the six `*_measured` indicators. These record clinician
actions, namely a treatment given or a test ordered, rather than the patient's
physiology. The laboratory values themselves stay in the model, so 14 of the
22 covariates remain. Only Variants A--C are refitted. The ablation asks
whether H2's cross-label instability, which is defined on those three
variants, depends on these covariates.

```{r fit-primary-model, code_folding="hide"}
fit_primary_model <- function(vid, arm = "full") {
  step(sprintf("--- Fitting primary model: Variant %s (%s), arm = %s ---",
               vid, variant_name(vid), arm))

  ph <- as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
  ))

  train <- ph[temporal_split == "train"]
  cat(sprintf("  Training rows: %s, events (sepsis): %d\n",
              format(nrow(train), big.mark = ","), sum(train$event == 1)))

  train[, outcome := factor(event, levels = c(0, 1, 2, 3),
                             labels = c("inprogress", "sepsis", "discharge", "death"))]

  avail_base  <- intersect(BASE_FEATURES,        names(train))
  avail_miss  <- intersect(MISSINGNESS_FEATURES, names(train))
  require_features(BASE_FEATURES, MISSINGNESS_FEATURES, available = names(train))

  # Protocol Amendment 5: the ablated arm drops the action-derived features
  # (vasopressor exposure and the order indicators) and nothing else. The
  # required-feature check above still runs against the full declared set, so a
  # column vanishing from the panel is caught in both arms.
  all_features <- arm_features(c(avail_base, avail_miss), arm)
  if (identical(arm, "ablated"))
    cat(sprintf("  Ablated arm: %d action-derived features dropped, %d remain\n",
                length(intersect(ACTION_DERIVED_FEATURES, c(avail_base, avail_miss))),
                length(all_features)))

  for (f in all_features) {
    if (is.numeric(train[[f]])) train[is.na(get(f)), (f) := 0]
    if (is.logical(train[[f]])) train[is.na(get(f)), (f) := FALSE]
  }

  n_knots  <- 5
  hour_max <- max(train$hour, na.rm = TRUE)
  knots    <- quantile(train$hour, seq(0.1, 0.9, length.out = n_knots), na.rm = TRUE)
  cat(sprintf("  Spline knots at hours: %s\n", paste(round(knots, 1), collapse = ", ")))

  feat_formula <- paste(all_features, collapse = " + ")
  formula_str  <- sprintf(
    "outcome ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
    paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
  )
  mod_formula <- as.formula(formula_str)

  v2_log(sprintf("  Fitting multinomial logistic regression (ridge decay = %g)...",
                 PRIMARY_MODEL_DECAY))
  fit <- tryCatch(
    nnet::multinom(mod_formula, data = as.data.frame(train),
                   trace = FALSE, MaxNWts = 50000, maxit = PRIMARY_MODEL_MAXIT,
                   decay = PRIMARY_MODEL_DECAY,
                   # Hess = TRUE only computes the Hessian after the optimiser
                   # has stopped; it does not change the fit. It is the bread of
                   # the cluster-robust sandwich below, which cannot be formed
                   # without it.
                   Hess = TRUE),
    error = function(e) {
      v2_log("  nnet::multinom failed: ", e$message, level = "ERROR")
      NULL
    }
  )

  check_multinom_fit(fit, vid)

  if (is.null(fit)) {
    v2_log("  Fallback: binary logistic for sepsis onset only")
    train[, outcome_bin := as.integer(event == 1)]
    formula_bin <- as.formula(sprintf(
      "outcome_bin ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
      paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
    ))
    fit <- glm(formula_bin, data = as.data.frame(train),
               family = binomial(link = "logit"))
    fit$is_binary_fallback <- TRUE
  }

  # Cluster-robust SEs, clustered on stay_id: person-hours within a stay are
  # ~45 correlated rows, so an independence-assuming variance is wrong by
  # construction. `sandwich` has no estfun method for multinom, so utils.R
  # supplies one (see `estfun.multinom`); without it this call fails, and it
  # previously failed inside a tryCatch that only warned, leaving vcov_robust
  # NULL while the methodology claimed the quantity.
  v2_log("  Computing cluster-robust SEs (clustered on stay_id)...")
  vcov_robust <- if (inherits(fit, "multinom")) {
    multinom_vcov_cluster(fit, cluster = train$stay_id, tag = vid)
  } else {
    tryCatch(sandwich::vcovCL(fit, cluster = ~stay_id,
                              data = as.data.frame(train)),
             error = function(e) {
               v2_log("  vcov_robust failed: ", e$message, level = "WARN"); NULL
             })
  }
  if (!is.null(vcov_robust)) {
    se_robust <- sqrt(diag(vcov_robust))
    v2_log(sprintf("  Cluster-robust SEs computed for %d parameters (median SE %.4f)",
                   length(se_robust), stats::median(se_robust)))
    saveRDS(vcov_robust,
            file.path(DATA_DIR,
                      paste0("05_vcov_cluster", arm_suffix(arm), "_", vid, ".rds")))
  }

  # Coefficient table
  coefs_raw <- if (inherits(fit, "multinom")) coef(fit) else coef(summary(fit))
  if (is.matrix(coefs_raw)) {
    coef_df    <- as.data.table(as.data.frame(coefs_raw), keep.rownames = "outcome")
    coef_table <- melt(coef_df, id.vars = "outcome",
                       variable.name = "term", value.name = "coef")
    setDF(coef_table)
  } else {
    coef_table         <- as.data.frame(coefs_raw)
    coef_table$term    <- rownames(coef_table)
    coef_table$outcome <- NA_character_
  }
  coef_table$variant <- vid
  coef_table$model   <- "discrete_time_competing_risks"
  coef_table$arm     <- arm

  # The cluster-robust standard error travels with the coefficient from the
  # point it is produced, rather than being rejoined from the saved covariance
  # later. H2 asks whether a coefficient changes across label variants, and
  # that question is unanswerable from point estimates alone: a threshold rule
  # cannot tell a coefficient that moved from one that was never estimated
  # precisely enough to say. The join is by "<outcome>:<term>", never by
  # position.
  coef_table <- attach_cluster_se(coef_table, vcov_robust)

  # Test-set predictions
  test <- ph[temporal_split == "test"]
  for (f in all_features) {
    if (is.numeric(test[[f]])) test[is.na(get(f)), (f) := 0]
    if (is.logical(test[[f]])) test[is.na(get(f)), (f) := FALSE]
  }
  v2_log(sprintf("  Predicting on test set (%s rows)...", format(nrow(test), big.mark = ",")))
  if (inherits(fit, "multinom")) {
    pred_probs <- predict(fit, newdata = as.data.frame(test), type = "probs")
    test[, pred_sepsis := if (is.matrix(pred_probs)) pred_probs[, "sepsis"]
                           else as.numeric(pred_probs)]
  } else {
    test[, pred_sepsis := predict(fit, newdata = as.data.frame(test), type = "response")]
  }

  save_cols <- c("stay_id", "hour", "event", "pred_sepsis", "temporal_split", "random_split",
                 "anchor_year_group", "onset_hour",
                 intersect(c("sofa_total", "news2_total", "qsofa_score"), names(test)))
  sfx <- arm_suffix(arm)
  save_parquet(test[, ..save_cols],
               paste0("05_predictions_primary", sfx, "_", vid), DATA_DIR)
  save_parquet(coef_table, paste0("05_coef_table_primary", sfx, "_", vid), DATA_DIR)
  saveRDS(fit, file.path(DATA_DIR, paste0("05_model_primary", sfx, "_", vid, ".rds")))

  cat(sprintf("  Variant %s model saved (arm = %s).\n", vid, arm))

  # Deliberately do NOT return the fit. A multinom object on 2.3 M person-hours
  # carries its fitted values and residuals (~1.5 GB live, 450 MB on disk), and
  # the caller keeps one per variant for the whole chunk. With five variants
  # that retained ~7 GB and made the random-split re-fit below hit R's 16 GB
  # vector limit. Only `coef_table` is used downstream (the H2 chunk); the fit
  # itself is already persisted to `05_model_primary_<vid>.rds`.
  out <- list(coef_table = coef_table, vcov_robust = vcov_robust)
  rm(fit, train, test, ph); invisible(gc(FALSE))
  out
}
```

```{r run-all-variants}
run_variants <- active_variants()
run_arms     <- active_arms()
v2_log(sprintf("Arms in this run: %s", paste(run_arms, collapse = ", ")))

models <- NULL
if ("full" %in% run_arms) {
  models        <- lapply(run_variants, fit_primary_model, arm = "full")
  names(models) <- run_variants
}

# The ablated arm is Protocol Amendment 5 and post-hoc. It writes to its own
# files and never overwrites the pre-registered fit, so `V2_ARMS=ablated`
# refreshes it while leaving every headline number on disk untouched.
models_ablated <- NULL
if ("ablated" %in% run_arms) {
  ablate_variants        <- intersect(run_variants, names(LABEL_VARIANTS))
  models_ablated         <- lapply(ablate_variants, fit_primary_model, arm = "ablated")
  names(models_ablated)  <- ablate_variants
}
```

## Random-Split Refit (H4: Split Optimism)

H4 asks whether a random train/test split overstates performance relative to
the temporal split (Guo et al., 2022). The temporal split tests the model on
later admissions, so its test set includes any change in case mix or care
practice over time. A random split mixes admission years, so its test set does
not, and the pre-registered expectation was that it would give the higher
AUROC. To measure the difference, Variant B's model is refitted on the random
split built in `03_person_hours.Rmd`, with the same covariates, spline rule,
penalty and outcome definition. That split assigns
`r round(100 * RANDOM_SPLIT_TEST_FRAC)`% of stays to the test set with seed
`r RANDOM_SPLIT_SEED`. It is made on stays, so all hours of a stay fall on the
same side. With one stay per patient, it is also a patient-level split. Only
Variant B is refitted, because H4 holds the label and the model class fixed.
H4 is defined on AUROC, and the contrast is computed in later stages.

The two training sets differ in size. The temporal split's proportion is set by
the anchor-year groups rather than fixed, and the printed row counts show its
training set to be the larger. Part of any AUROC difference between the two
splits may therefore come from training-set size rather than from the split
itself.

```{r fit-random-split}
step("--- Re-fitting primary model on random split (H4) ---")

fit_primary_random <- function(vid = "B") {
  ph <- as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
  ))

  if (!"random_split" %in% names(ph)) {
    v2_log("  random_split column not found: skipping H4 re-fit.", level = "WARN")
    return(NULL)
  }

  train <- ph[random_split == "train"]
  cat(sprintf("  Random-split training rows: %s, events (sepsis): %d\n",
              format(nrow(train), big.mark = ","), sum(train$event == 1)))

  train[, outcome := factor(event, levels = c(0, 1, 2, 3),
                             labels = c("inprogress", "sepsis", "discharge", "death"))]

  avail_base  <- intersect(BASE_FEATURES,        names(train))
  avail_miss  <- intersect(MISSINGNESS_FEATURES, names(train))
  all_features <- c(avail_base, avail_miss)
  require_features(BASE_FEATURES, MISSINGNESS_FEATURES, available = names(train))

  for (f in all_features) {
    if (is.numeric(train[[f]])) train[is.na(get(f)), (f) := 0]
    if (is.logical(train[[f]])) train[is.na(get(f)), (f) := FALSE]
  }

  n_knots  <- 5
  hour_max <- max(train$hour, na.rm = TRUE)
  knots    <- quantile(train$hour, seq(0.1, 0.9, length.out = n_knots), na.rm = TRUE)

  feat_formula <- paste(all_features, collapse = " + ")
  formula_str  <- sprintf(
    "outcome ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
    paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
  )
  mod_formula <- as.formula(formula_str)

  fit <- tryCatch(
    nnet::multinom(mod_formula, data = as.data.frame(train),
                   trace = FALSE, MaxNWts = 50000, maxit = PRIMARY_MODEL_MAXIT,
                   decay = PRIMARY_MODEL_DECAY),
    error = function(e) {
      v2_log("  Random-split multinom failed: ", e$message, level = "ERROR")
      NULL
    }
  )

  check_multinom_fit(fit, paste0(vid, " (random split)"))

  if (is.null(fit)) {
    v2_log("  Fallback: binary logistic for random split")
    train[, outcome_bin := as.integer(event == 1)]
    formula_bin <- as.formula(sprintf(
      "outcome_bin ~ ns(hour, knots = c(%s), Boundary.knots = c(0, %d)) + %s",
      paste(round(knots, 2), collapse = ", "), hour_max, feat_formula
    ))
    fit <- glm(formula_bin, data = as.data.frame(train),
               family = binomial(link = "logit"))
    fit$is_binary_fallback <- TRUE
  }

  test <- ph[random_split == "test"]
  for (f in all_features) {
    if (is.numeric(test[[f]])) test[is.na(get(f)), (f) := 0]
    if (is.logical(test[[f]])) test[is.na(get(f)), (f) := FALSE]
  }
  v2_log(sprintf("  Predicting on random-split test set (%s rows)...",
                 format(nrow(test), big.mark = ",")))
  if (inherits(fit, "multinom")) {
    pred_probs <- predict(fit, newdata = as.data.frame(test), type = "probs")
    test[, pred_sepsis := if (is.matrix(pred_probs)) pred_probs[, "sepsis"]
                           else as.numeric(pred_probs)]
  } else {
    test[, pred_sepsis := predict(fit, newdata = as.data.frame(test), type = "response")]
  }

  save_cols <- c("stay_id", "hour", "event", "pred_sepsis", "random_split",
                 "anchor_year_group", "onset_hour",
                 intersect(c("sofa_total", "news2_total", "qsofa_score"), names(test)))
  save_parquet(test[, ..save_cols],
               paste0("05_predictions_primary_random_", vid), DATA_DIR)
  saveRDS(fit, file.path(DATA_DIR, paste0("05_model_primary_random_", vid, ".rds")))
  cat(sprintf("  Random-split model for Variant %s saved.\n", vid))
  rm(fit, train, test, ph); invisible(gc(FALSE))
  invisible(TRUE)
}

# Release whatever the per-variant loop still holds before the random-split fit
# allocates its own design matrix.
invisible(gc(FALSE))

# The random-split arm exists only to quantify optimism under the pre-registered
# specification (H4), so it is tied to the full arm. Under V2_ARMS=ablated it is
# skipped rather than refitted, which is what keeps an ablation-only pass from
# touching any H4 input.
random_split_done <- if (!"full" %in% run_arms) {
  cat("Full arm not selected; keeping the existing random-split fit.\n")
  NULL
} else if ("B" %in% run_variants) fit_primary_random("B") else {
  cat("Variant B not selected in this run; keeping the existing random-split fit.\n")
  NULL
}
```

## Cross-Label Coefficient Stability

H2 asks whether the choice of label changes the fitted covariate effects. Its
pre-registered rule compares each covariate's onset-outcome coefficient across
Variants A, B and C. A covariate is flagged for a sign reversal if its
coefficient is positive under one variant and negative under another. It is
flagged for a magnitude change if its largest absolute coefficient is at least
twice its smallest. Intercepts and spline terms are excluded. The rule uses
point estimates and has no minimum magnitude, so a coefficient close to zero
can be flagged after a small absolute change.

A sign reversal shows that the estimated direction of association, conditional
on the other covariates, depends on the label. It does not by itself show a
change in an underlying clinical relationship. Collinearity among the
covariates, the different at-risk rows and the different numbers of onsets can
all move a coefficient. The comparison is like for like in coding, reference
category and covariates. The knots differ slightly between the three fits,
which acts mainly on the baseline terms that the rule excludes.

This notebook writes the combined coefficient table for A--C. The H2 count for
the full specification is computed in `12_inference.Rmd`. That notebook also
checks each flagged change against its cluster-robust standard errors
(Protocol Amendment 7). The coefficients of D--F go to a separate file for the
analyses of Protocol Amendments 2 and 4.

Table 1 lists the fifteen onset-outcome covariate coefficients of Variant B's
full fit with the largest absolute values, with their cluster-robust standard
errors (`se_cluster`). Each coefficient is the change in the log-odds of onset,
relative to remaining at risk, per unit of the covariate, holding the other
covariates and the hour fixed. The covariates are unstandardised, so the
ordering compares unlike units. A binary indicator moves by 1 between its two
states, while heart rate moves by 1 per beat per minute, so the table is not a
ranking of importance. The `*_measured` indicators record testing rather than
physiology. A laboratory value's coefficient is conditional on its indicator
and on the other covariates. Neither kind should be read as a clinical effect.
Factor terms carry R's level suffix, so `vaso_anyTRUE` denotes vasopressor
exposure.

```{r cross-label-coefs}
step("Computing cross-label coefficient stability (→ H2)...")

bind_coefs <- function(vids, src = models) data.table::rbindlist(
  lapply(vids, function(vid) as.data.table(src[[vid]]$coef_table)), fill = TRUE
)

# H2 is defined on the pre-registered family, so the alternative-label
# coefficients are written to their own file and never mixed in.
prereg_run <- if (is.null(models)) {
  character(0)
} else {
  intersect(run_variants, names(LABEL_VARIANTS))
}
if (length(prereg_run) > 0) {
  all_coefs <- bind_coefs(prereg_run)
  save_parquet(all_coefs, "05_coef_all_variants", DATA_DIR)
  cat("Cross-label coefficient tables saved for H2 testing.\n")

  # Show sepsis-outcome coefficients for Variant B as a preview
  coef_b_sepsis <- all_coefs[outcome == "sepsis" & variant == "B" &
                                !grepl("^ns\\(|^\\(Intercept\\)", as.character(term))]
  if (nrow(coef_b_sepsis) > 0) {
    prev_cols <- c("term", "coef", "se_cluster", "variant")
    prev_cols <- intersect(prev_cols, names(coef_b_sepsis))
    knitr::kable(head(coef_b_sepsis[order(-abs(coef)), ..prev_cols], 15),
                 caption = paste("Top 15 sepsis-outcome coefficients (Variant B,",
                                 "by |coef|), with cluster-robust standard errors"))
  }
} else {
  cat("No pre-registered variants in this run; 05_coef_all_variants left unchanged.\n")
}

alt_run <- if (is.null(models)) {
  character(0)
} else {
  intersect(run_variants, names(ALT_LABEL_VARIANTS))
}
if (length(alt_run) > 0) {
  save_parquet(bind_coefs(alt_run), "05_coef_alt_variants", DATA_DIR)
  cat("Alternative-label coefficient tables saved (Protocol Amendment 2).\n")
}
```

### Stability of the Ablated Specification (Protocol Amendment 5)

The same rule, implemented by the same function (`coef_stability_flags()` in
`utils.R`), is applied to the ablated fits of Variants A--C. Comparing the
result with the full specification shows whether the cross-label instability
is concentrated in the eight action-derived covariates or spread across the
physiological ones. If it is concentrated in them, a smaller proportion of the
ablated covariates should be flagged than of the full set. The full set has 22
covariates and the ablated set 14, so the counts are compared as proportions.
No threshold for a meaningful reduction was set in advance and no formal test
is made, so the comparison is descriptive. The full-specification count comes
from `12_inference.Rmd`.

The printed line gives the number of covariates shared by the three ablated
fits. It then gives how many of them reverse sign, how many change magnitude at
least twofold, and how many meet either criterion. Table 2 lists the flagged
covariates, with `TRUE` marking each criterion met. Because the flags are based
on point estimates only, a flagged covariate is not necessarily one whose
change exceeds its sampling uncertainty.

```{r ablated-coef-stability}
if (!is.null(models_ablated) && length(models_ablated) >= 2) {
  abl_coefs <- bind_coefs(names(models_ablated), src = models_ablated)
  save_parquet(abl_coefs, "05_coef_all_variants_ablated", DATA_DIR)

  # coef_stability_flags() is the same function 12_inference.Rmd applies to the
  # full arm, so the two counts differ only by the feature set.
  abl_flags <- coef_stability_flags(abl_coefs, variants = names(models_ablated))
  save_parquet(abl_flags, "05_h2_stability_ablated", DATA_DIR)
  save_csv(abl_flags,     "05_h2_stability_ablated", DATA_DIR)

  cat(sprintf("Ablated arm: of %d shared covariates, %d reverse sign, %d change magnitude >=2x, %d do either\n",
              nrow(abl_flags), sum(abl_flags$sign_flip), sum(abl_flags$mag_flip),
              sum(abl_flags$sign_flip | abl_flags$mag_flip)))
  knitr::kable(abl_flags[sign_flip | mag_flip],
               caption = "Unstable covariates remaining after ablation")
} else {
  cat("Ablated arm not run (or fewer than two variants); ablated stability unchanged.\n")
}
```
