---
title: "06 Comparators - Rule-Based, Cox, and GBT Comparator Models"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook produces the comparator models against which the primary model
of `05_primary_model.Rmd` is judged. Every comparator is evaluated on the same
cohort, label variants and temporal split as the primary model, and on the
same test person-hours. `07_metrics_suite.Rmd` then scores all of them with the
same metrics. The hourly prediction task follows the framing of the
PhysioNet/Computing in Cardiology Challenge 2019 (Reyna et al., 2020). The
primary model estimates onset, discharge and death jointly, whereas every
comparator here predicts onset only.

| Comparator | Role |
|---|---|
| **NEWS2** | The National Early Warning Score 2, a bedside deterioration score. It is the principal rule-based comparator, because the Surviving Sepsis Campaign guidelines advise against using qSOFA as a sole screening tool (Evans et al., 2021). It is computed in `03_person_hours.Rmd` and needs no fitting. |
| **qSOFA, SIRS** | The quick SOFA score and the Systemic Inflammatory Response Syndrome criteria, also computed in `03_person_hours.Rmd`. They are reported for descriptive comparison only. |
| **Cox** | A Cox proportional-hazards (PH) model with time-varying covariates and time-constant coefficients, which the pre-registration calls the static comparator. It is a conventional survival-model baseline. Its PH assumption is tested with Schoenfeld residuals and was not expected to hold. |
| **GBT** | Gradient-boosted trees (GBT), fitted with the XGBoost library. This is the flexible machine-learning comparator, and its contrast with the primary model is H6. |

All comparators are produced for every variant in the run. Variant B is the
primary analysis and the variant on which H6 is evaluated. Variants A--C
together provide the label component of the variance decomposition. Variants
D--F are the post-hoc variants of Protocol Amendments 2 and 4.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(survival)
  library(xgboost)
  library(rms)
  library(dplyr)
})

source("config.R")
source("utils.R")
source("clinical_scores.R")

step("=== Phase 3b: Comparator models ===")
```

## Feature Set

The GBT uses the same 22 covariates as the primary model, in the same order
(`model_features()` in `config.R`). H6 asks whether performance differs by
model class when the other conditions are held fixed. If the GBT received
covariates that the primary model lacked, a difference in AUROC could come
from the covariates rather than from the model class. The two inputs still
differ in one respect. The primary model also includes a spline in hours since
ICU admission, whereas the GBT has no time input beyond the 22 covariates.

The Cox model uses 12 of the 22 covariates: the six vital signs, the five
laboratory values and `vaso_any`. It omits the six-hour means,
`norepi_epi_any` and the six `*_measured` indicators. Hours since admission
form its time scale, and it treats discharge and death as censoring. Its
contrast with the primary model therefore compares complete specifications
rather than model form alone. NEWS2, qSOFA and SIRS apply their own fixed
scoring rules.

Missing covariate values are set to zero, or to false for logical columns, as
in `05_primary_model.Rmd`.

```{r feature-set}
# The comparators are fitted on the same covariate vector as the primary model,
# in the same order. It is declared once in config.R; `model_features()`
# returns BASE_FEATURES followed by MISSINGNESS_FEATURES, which is the order
# the stored xgboost design matrix was built in and the order stage 08 must
# reconstruct.
MODEL_FEATURES <- model_features()
```

## Fit Comparators by Variant

For each variant, the Cox model and the GBT are fitted to the training
person-hours of the temporal split. All comparators are then scored on the
test person-hours of that split.

**Rule-based scores.** NEWS2, qSOFA and SIRS are read from the person-hour
table. Each score is divided by its maximum (20, 3 and 4 respectively) to give
a value between 0 and 1. This rescaling preserves the ranking and so leaves the
AUROC unchanged, but the rescaled values are not probabilities. An NEWS2 alert
is a total of 5 or more. The printed alert rate is the percentage of test
person-hours at or above that threshold.

**Cox model.** The Cox model is fitted in counting-process form. Each
person-hour contributes the interval from hour $t$ to $t + 1$, with its onset
indicator as the event. Ties are handled by Efron's method, and standard errors
are clustered by stay. The printed line gives the p-value of the global
Schoenfeld test of the PH assumption. It is printed to three decimal places,
so `p = 0.000` means $p < 0.0005$. With more than a million training
person-hours in every variant, the test detects even small departures from proportionality. A
small p-value therefore shows that the assumption fails, not how far it fails.
The test-set score is the linear predictor passed through the logistic
function. It ranks person-hours by relative hazard and so supports measures of
discrimination, but it is not a calibrated probability.

**GBT.** The GBT uses a binary logistic objective, onset against all other
states, with maximum tree depth 6, learning rate 0.05, row and column
subsampling of 0.8 and minimum child weight 5. These settings are fixed rather
than tuned. They are the same for every variant, split and specification. The
onset rows are up-weighted by the ratio of non-onset to onset rows in the
fitting part of the training data. This offsets the rarity of onsets, but it
also inflates the predicted probabilities, so the GBT's scores are not
calibrated probabilities either.

The number of boosting rounds is chosen by early stopping (Protocol Amendment
8). `r 100 * GBT_VALID_FRACTION`% of training stays, drawn with seed
`r GBT_VALID_SEED`, form a validation subset, and the model is fitted on the
remaining training stays. The area under the precision-recall curve is
monitored on the validation subset. Boosting stops after `r GBT_EARLY_STOP`
rounds without improvement, or at `r GBT_NROUNDS_MAX` rounds. The subset is
drawn by stay, so all hours of a stay fall on the same side. The test set is
not used for fitting, for early stopping or for any other modelling choice.

For each variant, the output prints the following:

- the covariate check;
- the NEWS2 alert rate;
- the Schoenfeld result;
- the early-stopping split, giving the number of training stays used for
  fitting, the number in the validation subset and the onset rows on each side.

Every variant draws the split from the same training stays with the same seed,
so the stay counts are identical across variants and only the onset counts
differ.

```{r fit-comparators-fn, code_folding="hide"}
fit_comparators <- function(vid) {
  step(sprintf("--- Comparators for Variant %s ---", vid))

  ph <- as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
  ))

  train <- ph[temporal_split == "train"]
  test  <- ph[temporal_split == "test"]

  avail_feats <- intersect(MODEL_FEATURES, names(ph))
  require_features(MODEL_FEATURES, available = names(ph))
  for (f in avail_feats) {
    if (is.numeric(train[[f]])) train[is.na(get(f)), (f) := 0]
    if (is.logical(train[[f]])) train[is.na(get(f)), (f) := FALSE]
    if (is.numeric(test[[f]]))  test[is.na(get(f)),  (f) := 0]
    if (is.logical(test[[f]]))  test[is.na(get(f)),  (f) := FALSE]
  }

  train[, event_bin := as.integer(event == 1)]
  test[,  event_bin := as.integer(event == 1)]

  results <- list()

  # --- NEWS2 (pre-computed in 03_person_hours.Rmd) ---
  if ("news2_total" %in% names(test)) {
    step("  Comparator: NEWS2")
    test[, pred_news2  := as.numeric(news2_total) / 20]
    test[, alert_news2 := as.integer(news2_total >= 5)]
    results[["news2"]] <- test[, .(stay_id, hour, event, event_bin,
                                    pred_news2, alert_news2)]
    cat(sprintf("    NEWS2 alert rate: %.1f%%\n",
                100 * mean(test$alert_news2, na.rm = TRUE)))
  }

  if ("qsofa_score" %in% names(test)) {
    test[, pred_qsofa := qsofa_score / 3]
    results[["qsofa"]] <- test[, .(stay_id, hour, event, event_bin, pred_qsofa)]
  }

  if ("sirs_score" %in% names(test)) {
    test[, pred_sirs := sirs_score / 4]
    results[["sirs"]] <- test[, .(stay_id, hour, event, event_bin, pred_sirs)]
  }

  # --- Static Cox PH ---
  step("  Comparator: Static Cox PH")
  cox_feats   <- intersect(c("hr", "resp_rate", "spo2", "map", "temp_c", "gcs",
                               "lactate", "creatinine", "bilirubin_total",
                               "platelets", "wbc", "vaso_any"), names(train))
  cox_fit <- tryCatch({
    train_cox <- train[, .SD, .SDcols = c("stay_id", "hour", "event_bin", cox_feats)]
    train_cox[, stop := hour + 1L]
    cox_formula_tv <- as.formula(paste(
      "Surv(hour, stop, event_bin) ~", paste(cox_feats, collapse = " + ")
    ))
    survival::coxph(cox_formula_tv, data = as.data.frame(train_cox),
                    cluster = stay_id, ties = "efron")
  }, error = function(e) {
    v2_log("  Cox fit error: ", e$message, level = "WARN"); NULL
  })

  if (!is.null(cox_fit)) {
    schoen_test <- tryCatch(survival::cox.zph(cox_fit, transform = "log"),
                            error = function(e) NULL)
    if (!is.null(schoen_test)) {
      schoen_df            <- as.data.frame(schoen_test$table)
      schoen_df$covariate  <- rownames(schoen_df)
      schoen_df$variant    <- vid
      save_parquet(schoen_df, paste0("06_cox_schoenfeld_", vid), DATA_DIR)
      cat(sprintf("  Cox PH test: global p = %.3f (p < 0.05 = PH violated, expected)\n",
                  schoen_test$table["GLOBAL", "p"]))
    }
    test_cox <- test[, .SD, .SDcols = c("stay_id", "hour", "event_bin", cox_feats)]
    test_cox[, stop := hour + 1L]
    lp <- tryCatch(
      predict(cox_fit, newdata = as.data.frame(test_cox), type = "lp"),
      error = function(e) rep(NA_real_, nrow(test_cox))
    )
    test[, pred_cox := plogis(lp)]
    results[["cox"]] <- test[, .(stay_id, hour, event, event_bin, pred_cox)]
    saveRDS(cox_fit, file.path(DATA_DIR, paste0("06_model_cox_", vid, ".rds")))
    cox_coef           <- as.data.frame(coef(summary(cox_fit)))
    cox_coef$covariate <- rownames(cox_coef)
    cox_coef$variant   <- vid
    save_parquet(cox_coef, paste0("06_coef_cox_", vid), DATA_DIR)
  }

  # --- Gradient-Boosted Trees ---
  step("  Comparator: Gradient-Boosted Trees (XGBoost)")
  gbt_feats <- avail_feats
  X_train   <- as.matrix(train[, ..gbt_feats])
  y_train   <- train$event_bin
  X_test    <- as.matrix(test[, ..gbt_feats])
  y_test    <- test$event_bin

  X_train[is.na(X_train)] <- 0
  X_test[is.na(X_test)]   <- 0
  X_train <- apply(X_train, 2, as.numeric)
  X_test  <- apply(X_test,  2, as.numeric)

  # Protocol Amendment 8: the early-stopping set is carved out of the TRAINING
  # stays. `xgb.train` selects `best_iteration` against the last element of
  # `evals`, so passing the test set there would choose the number of boosting
  # rounds on the rows the model is then scored on.
  is_valid <- gbt_valid_split(train$stay_id, y_train, tag = paste0("full/", vid))
  dfit     <- xgboost::xgb.DMatrix(data = X_train[!is_valid, , drop = FALSE],
                                   label = y_train[!is_valid])
  dvalid   <- xgboost::xgb.DMatrix(data = X_train[is_valid, , drop = FALSE],
                                   label = y_train[is_valid])
  dtest    <- xgboost::xgb.DMatrix(data = X_test,  label = y_test)

  # Class-imbalance weighting is computed on the fitting side only, for the same
  # reason: it is a property of the data the model sees.
  n_pos      <- sum(y_train[!is_valid], na.rm = TRUE)
  n_neg      <- sum(!y_train[!is_valid], na.rm = TRUE)
  scale_pos  <- n_neg / max(n_pos, 1)

  gbt_params <- list(
    objective        = "binary:logistic",
    eval_metric      = "aucpr",
    max_depth        = 6,
    eta              = 0.05,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    min_child_weight = 5,
    scale_pos_weight = scale_pos,
    nthread          = 4
  )

  gbt_fit <- tryCatch({
    xgboost::xgb.train(
      params                = gbt_params,
      data                  = dfit,
      nrounds               = GBT_NROUNDS_MAX,
      evals                 = list(fit = dfit, valid = dvalid),
      early_stopping_rounds = GBT_EARLY_STOP,
      verbose               = 0
    )
  }, error = function(e) {
    v2_log("  XGBoost error: ", e$message, level = "WARN"); NULL
  })

  if (!is.null(gbt_fit)) {
    pred_gbt <- predict(gbt_fit, dtest)
    test[, pred_gbt := pred_gbt]
    results[["gbt"]] <- test[, .(stay_id, hour, event, event_bin, pred_gbt)]
    saveRDS(gbt_fit, file.path(DATA_DIR, paste0("06_model_gbt_", vid, ".rds")))
    feat_imp <- xgboost::xgb.importance(feature_names = gbt_feats, model = gbt_fit)
    if (!is.null(feat_imp)) {
      feat_imp$variant <- vid
      save_parquet(as.data.frame(feat_imp),
               paste0("06_gbt_feature_importance_", vid), DATA_DIR)
    }
    cat(sprintf("  GBT fitted: best iteration %d\n", gbt_fit$best_iteration))
  }

  # Combine comparator predictions
  all_preds <- test[, .(stay_id, hour, event, event_bin, onset_hour,
                          anchor_year_group, temporal_split,
                          first_abx_hour = if ("first_abx_hour" %in% names(test))
                                             first_abx_hour else NA,
                          first_culture_hour = if ("first_culture_hour" %in% names(test))
                                                first_culture_hour else NA)]
  for (nm in names(results)) {
    pred_col <- grep("^pred_", names(results[[nm]]), value = TRUE)
    if (length(pred_col) > 0)
      all_preds <- merge(all_preds, results[[nm]][, c("stay_id", "hour", pred_col),
                                                   with = FALSE],
                         by = c("stay_id", "hour"), all.x = TRUE)
  }
  all_preds[, variant := vid]
  save_parquet(all_preds, paste0("06_predictions_comparators_", vid), DATA_DIR)
  cat(sprintf("  Comparator predictions saved for Variant %s\n", vid))
  all_preds
}
```

```{r run-comparators}
run_variants <- active_variants()
run_arms     <- active_arms()
v2_log(sprintf("Arms in this run: %s", paste(run_arms, collapse = ", ")))

all_comparator_preds <- NULL
if ("full" %in% run_arms) {
  all_comparator_preds        <- lapply(run_variants, fit_comparators)
  names(all_comparator_preds) <- run_variants
} else {
  cat("Full arm not selected; keeping the existing comparator fits.\n")
}
```

## Ablated GBT (Protocol Amendment 5)

The GBT is refitted for Variants A--C without the eight action-derived
covariates: `vaso_any`, `norepi_epi_any` and the six `*_measured` indicators.
The primary model is refitted in the same way in `05_primary_model.Rmd`, so the
ablation is measured on both the regression and the tree-based model. The
hyperparameters, the early-stopping rule and the validation stays are the same
as in the full GBT, so the two GBT fits differ only in their covariates. The
printed output calls the ablated specification the "ablated arm".

No other comparator is ablated. NEWS2, qSOFA and SIRS do not use these
covariates. The Cox model already omits all of them except `vaso_any`, and its
main purpose in the study is the test of the PH assumption rather than
discrimination.

```{r fit-gbt-ablated}
fit_gbt_ablated <- function(vid) {
  step(sprintf("--- Ablated GBT: Variant %s ---", vid))

  ph <- as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
  ))
  train <- ph[temporal_split == "train"]
  test  <- ph[temporal_split == "test"]

  require_features(MODEL_FEATURES, available = names(ph))
  gbt_feats <- arm_features(intersect(MODEL_FEATURES, names(ph)), "ablated")
  cat(sprintf("  Ablated arm: %d features (from %d)\n",
              length(gbt_feats), length(intersect(MODEL_FEATURES, names(ph)))))

  for (f in gbt_feats) {
    if (is.numeric(train[[f]])) train[is.na(get(f)), (f) := 0]
    if (is.logical(train[[f]])) train[is.na(get(f)), (f) := FALSE]
    if (is.numeric(test[[f]]))  test[is.na(get(f)),  (f) := 0]
    if (is.logical(test[[f]]))  test[is.na(get(f)),  (f) := FALSE]
  }
  train[, event_bin := as.integer(event == 1)]
  test[,  event_bin := as.integer(event == 1)]

  X_train <- apply(as.matrix(train[, ..gbt_feats]), 2, as.numeric)
  X_test  <- apply(as.matrix(test[,  ..gbt_feats]), 2, as.numeric)
  X_train[is.na(X_train)] <- 0
  X_test[is.na(X_test)]   <- 0
  y_train <- train$event_bin
  y_test  <- test$event_bin

  # Protocol Amendment 8, as in the full arm: early stopping on TRAINING stays.
  # The split uses the same seed and fraction, so the two arms hold out the same
  # stays and the ablation contrast is not confounded by a different held-out
  # set.
  is_valid <- gbt_valid_split(train$stay_id, y_train, tag = paste0("ablated/", vid))
  dfit     <- xgboost::xgb.DMatrix(data = X_train[!is_valid, , drop = FALSE],
                                   label = y_train[!is_valid])
  dvalid   <- xgboost::xgb.DMatrix(data = X_train[is_valid, , drop = FALSE],
                                   label = y_train[is_valid])
  dtest    <- xgboost::xgb.DMatrix(data = X_test,  label = y_test)

  # Identical hyperparameters to the full arm. The ablation must isolate the
  # feature set; re-tuning would confound it with a search-budget difference.
  gbt_params <- list(
    objective        = "binary:logistic",
    eval_metric      = "aucpr",
    max_depth        = 6,
    eta              = 0.05,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    min_child_weight = 5,
    scale_pos_weight = sum(!y_train[!is_valid]) /
                       max(sum(y_train[!is_valid], na.rm = TRUE), 1),
    nthread          = 4
  )

  gbt_fit <- tryCatch(
    xgboost::xgb.train(params = gbt_params, data = dfit, nrounds = GBT_NROUNDS_MAX,
                       evals = list(fit = dfit, valid = dvalid),
                       early_stopping_rounds = GBT_EARLY_STOP, verbose = 0),
    error = function(e) { v2_log("  XGBoost error: ", e$message, level = "WARN"); NULL }
  )
  if (is.null(gbt_fit)) return(NULL)

  test[, pred_gbt := predict(gbt_fit, dtest)]
  saveRDS(gbt_fit, file.path(DATA_DIR, paste0("06_model_gbt_ablated_", vid, ".rds")))
  out <- test[, .(stay_id, hour, event, event_bin, onset_hour,
                   anchor_year_group, temporal_split, pred_gbt)]
  out[, variant := vid]
  save_parquet(out, paste0("06_predictions_gbt_ablated_", vid), DATA_DIR)
  cat(sprintf("  Ablated GBT fitted: best iteration %d\n", gbt_fit$best_iteration))
  rm(X_train, X_test, dfit, dvalid, dtest, gbt_fit, train, test, ph); invisible(gc(FALSE))
  invisible(TRUE)
}

if ("ablated" %in% run_arms) {
  invisible(lapply(intersect(run_variants, names(LABEL_VARIANTS)), fit_gbt_ablated))
} else {
  cat("Ablated arm not selected.\n")
}
```

## Random-Split Comparators (H4)

For H4, which compares AUROC under the temporal and random splits, the Cox
model and the GBT are refitted for Variant B on the random patient-level split
described in `03_person_hours.Rmd`. Their settings, including the
early-stopping rule, are those of the temporal fits. NEWS2 needs no fitting.
Its scores are taken from the random-split test rows. The predictions saved
here are those of NEWS2, the Cox model and the GBT. qSOFA and SIRS are not
included, so the random-split comparison in `07_metrics_suite.Rmd` covers
these three models and the primary model refitted in `05_primary_model.Rmd`.
No Schoenfeld test is run on the random-split Cox fit.

```{r fit-random-comparators}
step("--- Random-split comparators for Variant B (H4) ---")

vid <- "B"
# Tied to the full arm, exactly as the random-split primary fit in stage 05 is:
# H4 is defined on the pre-registered specification, so an ablation-only pass
# must leave every random-split input untouched.
ph <- if (!"full" %in% run_arms) {
  cat("Full arm not selected; keeping the existing random-split comparators.\n")
  data.table()
} else if ("B" %in% run_variants) {
  as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, paste0("03_person_hours_", vid, ".parquet"))
  ))
} else {
  cat("Variant B not selected in this run; keeping the existing random-split comparators.\n")
  data.table()
}

if ("random_split" %in% names(ph)) {
  train_r <- ph[random_split == "train"]
  test_r  <- ph[random_split == "test"]

  avail_feats <- intersect(MODEL_FEATURES, names(ph))
  require_features(MODEL_FEATURES, available = names(ph))
  for (f in avail_feats) {
    if (is.numeric(train_r[[f]])) train_r[is.na(get(f)), (f) := 0]
    if (is.logical(train_r[[f]])) train_r[is.na(get(f)), (f) := FALSE]
    if (is.numeric(test_r[[f]]))  test_r[is.na(get(f)),  (f) := 0]
    if (is.logical(test_r[[f]]))  test_r[is.na(get(f)),  (f) := FALSE]
  }
  train_r[, event_bin := as.integer(event == 1)]
  test_r[,  event_bin := as.integer(event == 1)]

  results_r <- list()

  # NEWS2/qSOFA/SIRS (no re-fit needed)
  if ("news2_total" %in% names(test_r)) {
    test_r[, pred_news2 := as.numeric(news2_total) / 20]
    results_r[["news2"]] <- test_r[, .(stay_id, hour, event, event_bin, pred_news2)]
  }
  if ("qsofa_score" %in% names(test_r))
    test_r[, pred_qsofa := qsofa_score / 3]
  if ("sirs_score" %in% names(test_r))
    test_r[, pred_sirs := sirs_score / 4]

  # Cox PH on random split
  cox_feats_r <- intersect(c("hr", "resp_rate", "spo2", "map", "temp_c", "gcs",
                              "lactate", "creatinine", "bilirubin_total",
                              "platelets", "wbc", "vaso_any"), names(train_r))
  cox_fit_r <- tryCatch({
    train_cox_r <- train_r[, .SD, .SDcols = c("stay_id", "hour", "event_bin", cox_feats_r)]
    train_cox_r[, stop := hour + 1L]
    cox_formula_r <- as.formula(paste(
      "Surv(hour, stop, event_bin) ~", paste(cox_feats_r, collapse = " + ")
    ))
    survival::coxph(cox_formula_r, data = as.data.frame(train_cox_r),
                    cluster = stay_id, ties = "efron")
  }, error = function(e) { v2_log("  Random Cox error: ", e$message, level = "WARN"); NULL })

  if (!is.null(cox_fit_r)) {
    test_cox_r <- test_r[, .SD, .SDcols = c("stay_id", "hour", "event_bin", cox_feats_r)]
    test_cox_r[, stop := hour + 1L]
    lp_r <- tryCatch(
      predict(cox_fit_r, newdata = as.data.frame(test_cox_r), type = "lp"),
      error = function(e) rep(NA_real_, nrow(test_cox_r))
    )
    test_r[, pred_cox := plogis(lp_r)]
    results_r[["cox"]] <- test_r[, .(stay_id, hour, event, event_bin, pred_cox)]
  }

  # GBT on random split
  gbt_feats_r <- avail_feats
  X_train_r   <- as.matrix(train_r[, ..gbt_feats_r])
  y_train_r   <- train_r$event_bin
  X_test_r    <- as.matrix(test_r[, ..gbt_feats_r])
  y_test_r    <- test_r$event_bin
  X_train_r[is.na(X_train_r)] <- 0
  X_test_r[is.na(X_test_r)]   <- 0
  X_train_r <- apply(X_train_r, 2, as.numeric)
  X_test_r  <- apply(X_test_r,  2, as.numeric)

  # Protocol Amendment 8, as in the temporal arm. H4 contrasts the two split
  # designs, so an early-stopping set drawn from the test rows would have made
  # the split contrast partly a contrast in how much leakage each arm carried.
  is_valid_r <- gbt_valid_split(train_r$stay_id, y_train_r, tag = "random/B")
  dfit_r     <- xgboost::xgb.DMatrix(data = X_train_r[!is_valid_r, , drop = FALSE],
                                     label = y_train_r[!is_valid_r])
  dvalid_r   <- xgboost::xgb.DMatrix(data = X_train_r[is_valid_r, , drop = FALSE],
                                     label = y_train_r[is_valid_r])
  dtest_r    <- xgboost::xgb.DMatrix(data = X_test_r,  label = y_test_r)

  n_pos_r     <- sum(y_train_r[!is_valid_r], na.rm = TRUE)
  n_neg_r     <- sum(!y_train_r[!is_valid_r], na.rm = TRUE)
  scale_pos_r <- n_neg_r / max(n_pos_r, 1)

  gbt_params_r <- list(
    objective        = "binary:logistic",
    eval_metric      = "aucpr",
    max_depth        = 6,
    eta              = 0.05,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    min_child_weight = 5,
    scale_pos_weight = scale_pos_r,
    nthread          = 4
  )

  gbt_fit_r <- tryCatch({
    xgboost::xgb.train(
      params                = gbt_params_r,
      data                  = dfit_r,
      nrounds               = GBT_NROUNDS_MAX,
      evals                 = list(fit = dfit_r, valid = dvalid_r),
      early_stopping_rounds = GBT_EARLY_STOP,
      verbose               = 0
    )
  }, error = function(e) { v2_log("  Random GBT error: ", e$message, level = "WARN"); NULL })

  if (!is.null(gbt_fit_r)) {
    test_r[, pred_gbt := predict(gbt_fit_r, dtest_r)]
    results_r[["gbt"]] <- test_r[, .(stay_id, hour, event, event_bin, pred_gbt)]
    cat(sprintf("  Random-split GBT: best iteration %d\n", gbt_fit_r$best_iteration))
  }

  # Combine random-split comparator predictions
  all_preds_r <- test_r[, .(stay_id, hour, event, event_bin, onset_hour,
                             anchor_year_group, random_split,
                             first_abx_hour = if ("first_abx_hour" %in% names(test_r))
                                                first_abx_hour else NA,
                             first_culture_hour = if ("first_culture_hour" %in% names(test_r))
                                                   first_culture_hour else NA)]
  for (nm in names(results_r)) {
    pred_col <- grep("^pred_", names(results_r[[nm]]), value = TRUE)
    if (length(pred_col) > 0)
      all_preds_r <- merge(all_preds_r, results_r[[nm]][, c("stay_id", "hour", pred_col),
                                                         with = FALSE],
                           by = c("stay_id", "hour"), all.x = TRUE)
  }
  all_preds_r[, variant := vid]
  save_parquet(all_preds_r, paste0("06_predictions_comparators_random_", vid), DATA_DIR)
  cat(sprintf("  Random-split comparator predictions saved for Variant %s\n", vid))
} else {
  cat("random_split column not found in person-hours; skipping H4 comparators.\n")
}
```

## Feature Importance (GBT)

Table 1 lists the 20 covariates with the highest gain in Variant B's full GBT.
The three measures are those reported by XGBoost, and each is expressed as a
share of its total across all covariates, so it sums to 1:

- **gain** is the covariate's share of the total improvement in the training
  objective achieved by the splits that use it;
- **cover** is its share of the training observations affected by those
  splits;
- **frequency** is its share of all splits.

These measures describe how the trees use each covariate. They are not effect
sizes and carry no direction.

The table gives a complementary view of the primary multinomial regression of
`05_primary_model.Rmd`. Broad agreement between the covariates that rank high
here and those with large effects in the regression would suggest that the two
model classes draw on similar predictive information. The two quantities are
not directly comparable, however. Gain measures contribution to tree splits,
while a regression coefficient is a conditional association per unit of an
unstandardised covariate, so close agreement in ranking is not expected.

One feature of the input bears on the ranking. The GBT has no time input, but
the six-hour means are set to zero in hours 0--4 of every stay (see
`05_primary_model.Rmd`). A split on these covariates can therefore separate the
first hours of a stay from later ones, and part of their gain may reflect
elapsed time rather than physiology.

```{r gbt-importance}
imp_files <- file.path(DATA_DIR, sprintf("06_gbt_feature_importance_%s.parquet",
                                          names(LABEL_VARIANTS)))
imp_files <- imp_files[file.exists(imp_files)]
if (length(imp_files) > 0) {
  imp_all <- rbindlist(lapply(imp_files, function(f) as.data.table(arrow::read_parquet(f))))
  if (nrow(imp_all) > 0) {
    top_imp <- imp_all[variant == "B"][order(-Gain)][1:min(20, .N)]
    knitr::kable(top_imp[, .(Feature, Gain = round(Gain, 4), Cover = round(Cover, 4),
                              Frequency = round(Frequency, 4))],
                 caption = "Top 20 GBT features by Gain (Variant B)")
  }
}
```
