---
title: "07 Metrics Suite - Evaluation Metrics and Operating Points"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook scores the test-set predictions of every model with the
pre-specified evaluation metrics, following Wang et al. (2025) and Van Calster
et al. (2019).

The primary metrics are an adapted PhysioNet-style Utility Score and the
calibration intercept and slope. The Utility Score keeps the normalisation of
the PhysioNet/Computing in Cardiology Challenge 2019 score (Reyna et al., 2020)
but departs from that score in other respects, described under Metric
Functions. Its values are therefore not comparable with published Challenge
scores. The secondary metrics are AUROC, AUPRC, decision-curve net benefit,
alert burden and median lead time. Alert burden is the number of false alerts
per true alert, and the benchmark for it is the 1.4 reported by Moor et al.
(2023).

Every metric is computed in two evaluation windows. The unrestricted window
contains all test person-hours. The treatment-anchored window contains only the
hours before a stay's first antibiotic or first culture, whichever comes first
(Kamran et al., 2024). All results come from the temporal test set, except in
the final section, which scores the random-split predictions used for H4.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(PRROC)
  library(dcurves)
  library(CalibrationCurves)
  library(dplyr)
  library(ggplot2)
})

source("config.R")
source("utils.R")

step("=== Full Evaluation Metric Suite ===")
```

## Metric Functions

The metric functions are defined in the chunk below and used throughout this
notebook. Stages 08 (external validation) and 10 (subgroups) compute AUROC and
AUPRC with the same `PRROC` functions, and stage 08 reads its internal AUROC
from this notebook's output. Both sides of the H5 contrast between internal and
external AUROC therefore come from the same estimator, so a difference between
them cannot arise from a difference in implementation.

The Utility Score is computed per stay. A septic stay that receives an alert
between six hours before onset and onset earns a reward equal to the warning
time of its earliest such alert divided by six hours. A six-hour warning
therefore earns 1, and an alert at onset earns 0. A septic stay whose only
alerts fall in the three hours after onset is charged 0.05 for each of those
alerts, and a septic stay with no alert in either window is charged 1. Every
alert hour outside these two windows, on any stay, is charged 0.05. The raw
total $U$ is then normalised against two reference strategies, issuing no
alerts ($U_{\text{inaction}}$) and alerting perfectly with no false alarms
($U_{\text{optimal}}$):

$$U_{\text{norm}} = \frac{U - U_{\text{inaction}}}{U_{\text{optimal}} - U_{\text{inaction}}}.$$

A score of 1 is a perfect forecaster, 0 is exactly the no-alert strategy, and a
negative score is worse than issuing no alerts. This normalisation follows
Reyna et al. (2020). The reward window, the single penalty for a missed stay
and the accumulation per stay rather than per hour do not, which is why the
score is described as adapted. The constants are the same for every model and
variant, so comparisons within this study are unaffected.

The calibration intercept and slope come from logistic recalibration of the
outcome on the logit of the predicted probability. Median lead time is taken
over septic stays with at least one alert at the 0.5 threshold, and is the
number of hours from the first alert to onset.

```{r metric-fns}
# --------------------------------------------------------------------------- #
# PhysioNet Utility Score (Reyna et al., 2020), stay-level operationalisation
# --------------------------------------------------------------------------- #
# Raw utility, per stay:
#   sepsis stay alerted in the reward window [onset + UTILITY_TP_MIN, onset]
#       + lead / PREDICTION_HORIZON_H, where lead is the warning time of the
#       *earliest* alert in that window. A full-horizon warning scores 1; an
#       alert that only coincides with onset scores 0.
#   sepsis stay alerted only in (onset, onset + UTILITY_TP_MAX]
#       UTILITY_FP per late alert hour.
#   sepsis stay never alerted in either window
#       UTILITY_FN.
#   any alert hour outside a reward window, on any stay
#       UTILITY_FP.
#
# The raw total is referenced to the two strategies of Reyna et al. (2020):
#   U_inaction = UTILITY_FN            x n_sepsis_stays   (never alert)
#   U_optimal  = UTILITY_TP_MAX_REWARD x n_sepsis_stays   (perfect, no false alarms)
#   normalised = (U - U_inaction) / (U_optimal - U_inaction)
# so that 1 is a perfect forecaster, 0 is exactly the no-alert strategy, and a
# negative score is worse than issuing no alerts at all. Without the inaction
# baseline in both numerator and denominator a model that raises no alert scores
# -1 rather than 0, which contradicts the definition of the metric.
#
# `utility_curve()` evaluates the whole threshold grid in one pass: the per-stay
# reward is monotone in a falling threshold, so the sweep costs no more than a
# single sort plus one traversal of the reward-window rows.
utility_curve <- function(ph_pred, pred_col, thresholds) {
  dt <- as.data.table(ph_pred)
  p  <- as.numeric(dt[[pred_col]])
  ok <- is.finite(p)

  p     <- p[ok]
  stay  <- dt$stay_id[ok]
  hour  <- as.numeric(dt$hour)[ok]
  onset <- as.numeric(dt$onset_hour)[ok]
  y_hr  <- as.integer(dt$event)[ok] == 1L

  thresholds <- sort(unique(as.numeric(thresholds)), decreasing = TRUE)
  sep_stays  <- unique(stay[!is.na(onset)])
  n_sep      <- length(sep_stays)

  if (n_sep == 0L || length(thresholds) == 0L || length(p) == 0L)
    return(data.table(threshold = thresholds, n_alerts = NA_integer_,
                      alert_rate = NA_real_, n_sepsis_stays = n_sep,
                      n_detected_stays = NA_integer_, detection_rate = NA_real_,
                      utility_total = NA_real_, utility_normalised = NA_real_,
                      alert_burden = NA_real_))

  in_early <- !is.na(onset) & hour >= (onset + UTILITY_TP_MIN) & hour <= onset
  in_late  <- !is.na(onset) & hour >  onset & hour <= (onset + UTILITY_TP_MAX)
  n_early  <- sum(in_early)
  n_late   <- sum(in_late)

  # Reward-window rows only, sorted by descending score: these are the rows
  # whose status changes as the threshold falls.
  ev <- data.table(
    si   = c(match(stay[in_early], sep_stays), match(stay[in_late], sep_stays)),
    p    = c(p[in_early], p[in_late]),
    v    = c((onset[in_early] - hour[in_early]) / PREDICTION_HORIZON_H,
             rep(NA_real_, n_late)),
    late = c(rep(FALSE, n_early), rep(TRUE, n_late))
  )
  setorder(ev, -p)

  p_all <- sort(p)
  p_tp  <- sort(p[y_hr])
  count_ge <- function(sorted_x, tau)
    length(sorted_x) - findInterval(tau, sorted_x, left.open = TRUE)

  best       <- rep(NA_real_, n_sep)  # best TP reward so far, per sepsis stay
  late_cnt   <- integer(n_sep)        # late alerts so far, per sepsis stay
  sum_best   <- 0                     # sum of best[] over detected stays
  n_det      <- 0L                    # sepsis stays with a reward-window alert
  late_undet <- 0L                    # late alerts on stays not (yet) detected
  n_undet_late <- 0L                  # undetected stays carrying >= 1 late alert
  n_reward_alerts <- 0L               # alert hours inside a reward window

  j   <- 1L
  res <- vector("list", length(thresholds))

  for (k in seq_along(thresholds)) {
    tau <- thresholds[k]
    while (j <= nrow(ev) && ev$p[j] >= tau) {
      si <- ev$si[j]
      if (ev$late[j]) {
        late_cnt[si] <- late_cnt[si] + 1L
        if (is.na(best[si])) {
          late_undet <- late_undet + 1L
          if (late_cnt[si] == 1L) n_undet_late <- n_undet_late + 1L
        }
      } else {
        v <- ev$v[j]
        if (is.na(best[si])) {
          best[si] <- v
          sum_best <- sum_best + v
          n_det    <- n_det + 1L
          if (late_cnt[si] > 0L) {
            late_undet   <- late_undet - late_cnt[si]
            n_undet_late <- n_undet_late - 1L
          }
        } else if (v > best[si]) {
          sum_best <- sum_best + (v - best[si])
          best[si] <- v
        }
      }
      n_reward_alerts <- n_reward_alerts + 1L
      j <- j + 1L
    }

    n_alerts <- count_ge(p_all, tau)
    n_tp_hr  <- count_ge(p_tp,  tau)

    u_raw <- sum_best +
             UTILITY_FP * late_undet +
             UTILITY_FN * (n_sep - n_det - n_undet_late) +
             UTILITY_FP * (n_alerts - n_reward_alerts)

    u_inaction <- UTILITY_FN            * n_sep
    u_optimal  <- UTILITY_TP_MAX_REWARD * n_sep

    res[[k]] <- data.table(
      threshold          = tau,
      n_alerts           = n_alerts,
      alert_rate         = n_alerts / length(p),
      n_sepsis_stays     = n_sep,
      n_detected_stays   = n_det,
      detection_rate     = n_det / n_sep,
      utility_total      = u_raw,
      utility_normalised = (u_raw - u_inaction) / (u_optimal - u_inaction),
      alert_burden       = if (n_tp_hr == 0L) Inf else (n_alerts - n_tp_hr) / n_tp_hr
    )
  }

  out <- rbindlist(res)
  setorder(out, threshold)
  out[]
}

# Threshold grid, defined on the alert rate so that it adapts to each model's
# calibration. At an hourly event rate below 0.5 per cent no model's predicted
# probability approaches 0.5, so a grid on the probability scale would evaluate
# nothing but the no-alert strategy. The default 0.5 cut-off is appended so the
# sweep table always contains the operating point used elsewhere.
utility_threshold_grid <- function(p) {
  p <- p[is.finite(p)]
  if (length(p) == 0L) return(numeric(0))
  rates <- exp(seq(log(UTILITY_SWEEP_MIN_RATE), log(UTILITY_SWEEP_MAX_RATE),
                   length.out = UTILITY_SWEEP_N_GRID))
  taus  <- as.numeric(quantile(p, probs = 1 - rates, names = FALSE, type = 1))
  sort(unique(c(taus, 0.5)))
}

# Single-threshold wrapper, used for the default 0.5 operating point.
compute_utility_score <- function(ph_pred, pred_col, threshold = 0.5) {
  row <- utility_curve(ph_pred, pred_col, threshold)
  list(utility_total      = row$utility_total[1],
       utility_max        = UTILITY_TP_MAX_REWARD * row$n_sepsis_stays[1],
       utility_normalised = row$utility_normalised[1])
}

# Calibration
compute_calibration <- function(y_true, pred_prob) {
  df <- data.frame(y = y_true, p = pred_prob)
  df <- df[!is.na(df$p) & !is.na(df$y), ]
  if (nrow(df) < 10) return(list(intercept = NA, slope = NA))

  calib_model <- tryCatch(
    glm(y ~ offset(qlogis(pmax(pmin(p, 0.999), 0.001))),
        data = df, family = binomial),
    error = function(e) NULL
  )
  intercept <- if (!is.null(calib_model)) coef(calib_model)[1] else NA_real_
  calib_slope <- tryCatch({
    mod <- glm(y ~ qlogis(pmax(pmin(p, 0.999), 0.001)), data = df, family = binomial)
    coef(mod)[2]
  }, error = function(e) NA_real_)

  list(intercept = intercept, slope = calib_slope)
}

# AUROC + AUPRC
compute_roc_pr <- function(y_true, pred_prob) {
  if (sum(y_true, na.rm = TRUE) == 0 || sum(!y_true, na.rm = TRUE) == 0)
    return(list(auroc = NA, auprc = NA))
  valid <- !is.na(pred_prob) & !is.na(y_true)
  if (sum(valid) < 10) return(list(auroc = NA, auprc = NA))
  roc_obj <- tryCatch(
    PRROC::roc.curve(scores.class0 = pred_prob[valid], weights.class0 = y_true[valid]),
    error = function(e) NULL
  )
  pr_obj <- tryCatch(
    PRROC::pr.curve(scores.class0 = pred_prob[valid], weights.class0 = y_true[valid]),
    error = function(e) NULL
  )
  list(auroc = if (!is.null(roc_obj)) roc_obj$auc else NA_real_,
       auprc = if (!is.null(pr_obj))  pr_obj$auc.integral else NA_real_)
}

# Alert burden
compute_alert_burden <- function(y_true, alerts) {
  tp <- sum(alerts == 1 & y_true == 1, na.rm = TRUE)
  fp <- sum(alerts == 1 & y_true == 0, na.rm = TRUE)
  if (tp == 0) return(Inf)
  fp / tp
}

# Lead time to clinical recognition
compute_lead_time <- function(ph_pred, pred_col, threshold = 0.5) {
  sepsis_stays <- ph_pred[!is.na(onset_hour)]
  if (nrow(sepsis_stays) == 0) return(NA_real_)
  lead_times <- sepsis_stays[, {
    rows_ord    <- .SD[order(hour)]
    first_alert <- rows_ord[get(pred_col) >= threshold, min(hour)]
    if (is.infinite(first_alert) || is.na(first_alert)) NA_real_
    else as.numeric(onset_hour[1]) - as.numeric(first_alert)
  }, by = stay_id]
  median(lead_times$V1, na.rm = TRUE)
}
```

## Evaluation Loop

Every model is scored under every label variant in both evaluation windows.
The models are the primary model, a multinomial discrete-time competing-risks
hazard model (`05_primary_model.Rmd`), and the comparators of
`06_comparators.Rmd`: gradient-boosted trees fitted with XGBoost (GBT), the Cox
model, and the NEWS2, qSOFA and SIRS scores. A model is skipped in a window
with fewer than 100 scored person-hours. AUROC, AUPRC and calibration use every
scored person-hour. Alert burden, lead time and the Utility Score are first
computed at the default threshold of 0.5, and the Utility Score is also
computed across the threshold grid described below.

For each variant, a decision-curve analysis of the primary model and the GBT is
also run with the `dcurves` package over thresholds from 0.01 to 0.20, and
saved as `07_dca_<variant>.parquet`. It serves only as a cross-check on the
net-benefit analysis in `11_clinical_metrics.Rmd`, which uses a threshold grid
scaled to the event rate.

```{r evaluate-all}
MODEL_NAMES <- c(primary = "pred_sepsis", news2 = "pred_news2",
                  cox = "pred_cox", gbt = "pred_gbt",
                  qsofa = "pred_qsofa", sirs = "pred_sirs")

all_results <- list()
all_sweeps  <- list()

run_variants <- active_variants()

for (vid in run_variants) {
  step(sprintf("--- Evaluating Variant %s (%s) ---", vid, variant_name(vid)))

  prim_pred_path <- file.path(DATA_DIR, paste0("05_predictions_primary_",     vid, ".parquet"))
  comp_pred_path <- file.path(DATA_DIR, paste0("06_predictions_comparators_", vid, ".parquet"))

  if (!file.exists(prim_pred_path)) {
    v2_log(sprintf("  Skipping Variant %s: primary predictions not found", vid))
    next
  }

  ph_prim <- as.data.table(arrow::read_parquet(prim_pred_path))
  ph_comp <- if (file.exists(comp_pred_path))
               as.data.table(arrow::read_parquet(comp_pred_path)) else NULL

  ph_all <- ph_prim
  if (!is.null(ph_comp)) {
    merge_cols <- c("stay_id", "hour", setdiff(names(ph_comp), names(ph_prim)))
    ph_all <- merge(ph_all, ph_comp[, ..merge_cols], by = c("stay_id", "hour"), all.x = TRUE)
  }

  eval_windows <- list(
    unrestricted = ph_all,
    treatment_anchored = {
      if ("first_abx_hour" %in% names(ph_all)) {
        anchor_h <- if ("first_culture_hour" %in% names(ph_all))
                      pmin(ph_all$first_abx_hour, ph_all$first_culture_hour, na.rm = TRUE)
                    else ph_all$first_abx_hour
        ph_all[is.na(anchor_h) | hour < anchor_h]
      } else ph_all
    }
  )

  for (win_name in names(eval_windows)) {
    data_win <- eval_windows[[win_name]]
    if (nrow(data_win) == 0) next

    for (mod_name in names(MODEL_NAMES)) {
      pred_col <- MODEL_NAMES[[mod_name]]
      if (!pred_col %in% names(data_win)) next
      valid <- !is.na(data_win[[pred_col]])
      if (sum(valid) < 100) next

      y    <- as.integer(data_win$event == 1)
      pred <- as.numeric(data_win[[pred_col]])

      roc_pr  <- compute_roc_pr(y[valid], pred[valid])
      calib   <- compute_calibration(y[valid], pred[valid])
      alerts  <- as.integer(pred >= 0.5)
      burden  <- compute_alert_burden(y[valid], alerts[valid])
      utility <- tryCatch(
        compute_utility_score(data_win[valid], pred_col, threshold = 0.5),
        error = function(e) list(utility_normalised = NA, utility_total = NA)
      )
      lead_t  <- tryCatch(
        compute_lead_time(data_win[valid], pred_col),
        error = function(e) NA_real_
      )

      # Threshold sweep. Every Utility Score above is read off the default 0.5
      # cut-off, which no deployment would use; the sweep reports what each
      # model can achieve at its best operating point.
      sweep <- tryCatch(
        utility_curve(data_win[valid], pred_col,
                      utility_threshold_grid(pred[valid])),
        error = function(e) NULL
      )
      best <- if (!is.null(sweep)) {
        cand <- sweep[n_alerts > 0 & is.finite(utility_normalised)]
        if (nrow(cand) > 0) cand[which.max(utility_normalised)] else NULL
      } else NULL

      if (!is.null(sweep)) {
        sweep[, `:=`(variant = vid, model = mod_name, eval_window = win_name,
                     split_type = "temporal")]
        all_sweeps[[paste(vid, mod_name, win_name, sep = "-")]] <- sweep
      }

      result_row <- data.frame(
        variant            = vid,
        variant_name       = variant_name(vid),
        model              = mod_name,
        eval_window        = win_name,
        split_type         = "temporal",
        n_rows             = sum(valid),
        n_sepsis_events    = sum(y[valid]),
        auroc              = round(roc_pr$auroc, 4),
        auprc              = round(roc_pr$auprc, 4),
        calib_intercept    = round(calib$intercept, 4),
        calib_slope        = round(calib$slope, 4),
        utility_normalised = round(utility$utility_normalised, 4),
        alert_burden       = round(burden, 2),
        lead_time_h        = round(lead_t, 2),
        utility_max        = if (is.null(best)) NA_real_ else round(best$utility_normalised, 4),
        utility_opt_threshold = if (is.null(best)) NA_real_ else signif(best$threshold, 4),
        utility_opt_alert_rate = if (is.null(best)) NA_real_ else signif(best$alert_rate, 4),
        utility_opt_burden  = if (is.null(best)) NA_real_ else round(best$alert_burden, 2),
        utility_opt_detection = if (is.null(best)) NA_real_ else round(best$detection_rate, 4)
      )
      all_results[[paste(vid, mod_name, win_name, sep = "-")]] <- result_row
    }
  }

  # Decision-curve analysis
  ph_test_for_dca <- ph_all[!is.na(pred_sepsis)]
  if (nrow(ph_test_for_dca) > 0 && "pred_gbt" %in% names(ph_test_for_dca)) {
    ph_test_for_dca[, outcome := as.integer(event == 1)]
    dca_result <- tryCatch(
      dcurves::dca(outcome ~ pred_sepsis + pred_gbt,
                   data       = as.data.frame(ph_test_for_dca),
                   thresholds = seq(0.01, 0.20, 0.01)),
      error = function(e) NULL
    )
    if (!is.null(dca_result)) {
      dca_df <- tryCatch(dca_result$dca, error = function(e) as.data.frame(dca_result))
      if (!is.null(dca_df) && nrow(dca_df) > 0) {
        dca_df$variant <- vid
        save_parquet(as.data.frame(dca_df), paste0("07_dca_", vid), DATA_DIR)
      }
    }
  }
}
```

## Results and Interpretation

The table below gives the results for the pre-registered Variants A--C on the
temporal test set. It is saved as `07_metric_results.parquet`, the input to
the variance decomposition. Variants D--F are written to a separate file (see
below), so they cannot enter H1 or the decomposition. The table supports three
comparisons.

The first is between the primary model and the GBT. H6 compares the two on
AUROC under Variant B. The table also gives their Utility Scores, and the two
metrics rank the models differently: the primary model has the higher AUROC
under every variant, and the GBT the higher maximum Utility Score. AUROC
measures how well a model ranks person-hours by risk, whereas the Utility Score
rewards early alerts and penalises false ones at a given threshold, so the two
need not agree. Both rankings are reported.

The second is between the two evaluation windows, and it is the comparison H3
was defined on. H3 predicted that restricting evaluation to pre-treatment hours
would reduce AUROC by at least 0.03. Such a reduction would suggest that the
unrestricted AUROC draws on information reflecting clinicians' suspicion of
infection. Under Variants A--C the comparison cannot be made. Their onset is
the time at which suspected infection is established from an antibiotic and a
culture together, and that time cannot precede the first of the two. The
pre-treatment window therefore contains no onsets by construction, and its
AUROC is undefined (shown as NA). The empirical counterpart is Variant F of
Protocol Amendment 4, whose onset can precede treatment. Its treatment-anchored
results appear in the alternative-label table.

The third concerns the threshold. The Utility Score, alert burden and lead time
in this table are read at 0.5. The primary model raises no alert at that
threshold, so its Utility Score is exactly 0, its alert burden is infinite and
its lead time is missing. These values describe the threshold rather than the
model, and the next section evaluates the Utility Score across a grid of
thresholds.

The calibration intercept and slope are saved in `07_metric_results.parquet`
but not displayed. They are estimated with predictions clipped to the interval
[0.001, 0.999]. At an hourly event rate of about 0.2 per cent, that clip alters
between about 9 and 47 per cent of the primary model's predictions, depending
on the variant. The calibration analysis of record is therefore the one in
`11_clinical_metrics.Rmd`, which uses a floor of $10^{-6}$, and the clipped
estimates here serve only to show the effect of the clip. In either analysis
the ideal slope is 1. A slope below 1 means the predictions are too extreme,
and a slope above 1 means they are too moderate. Discrimination metrics such
as AUROC do not detect either departure, but both matter when a threshold is
chosen.

```{r show-results}
display_cols <- c("variant", "model", "eval_window", "auroc", "auprc",
                  "utility_normalised", "utility_max", "alert_burden", "lead_time_h")

results_df <- if (length(all_results) > 0) do.call(rbind, all_results) else NULL

# The pre-registered variance decomposition reads 07_metric_results.parquet, so
# the alternative-label variants of Protocol Amendment 2 are written to their
# own file and cannot leak into H1 or the decomposition table.
prereg_df <- if (!is.null(results_df))
               results_df[results_df$variant %in% names(LABEL_VARIANTS), ] else NULL

if (!is.null(prereg_df) && nrow(prereg_df) > 0) {
  save_parquet(prereg_df, "07_metric_results", DATA_DIR)
  knitr::kable(
    prereg_df[order(prereg_df$variant, prereg_df$eval_window, prereg_df$model),
              display_cols],
    caption = "Evaluation results: all models x variants x windows (temporal split)"
  )
  cat("\nSaved: 07_metric_results.parquet\n")
} else {
  cat("No pre-registered-variant results in this run; 07_metric_results.parquet left unchanged.\n")
}
```

## Utility Score Across the Threshold Grid

The Utility Scores in the table above are read at the default threshold of
0.5, which is not informative here. The primary model's largest hourly
predicted probability is well below 0.5, so it raises no alert at that
threshold. The other models alert on a large share of person-hours at 0.5,
because their outputs are not calibrated probabilities. The GBT is trained with
up-weighted onset rows, the Cox score is a transformed linear predictor, and
the rule-based scores are rescaled to lie between 0 and 1. The 0.5 threshold
therefore gives the no-alert strategy for the primary model and an arbitrary
operating point for the others.

The Utility Score is therefore also computed over a grid of thresholds defined
on the alert rate rather than on the probability scale, so that the grid
adapts to each model's calibration. For each model, up to 60 thresholds are
placed at the quantiles of its predictions that give alert rates evenly spaced
on a log scale from 0.001 to 50 per cent of person-hours, and 0.5 is added. The
full grid is saved to `07_utility_sweep.csv`. The maximum Utility Score over
the thresholds that raise at least one alert, with the operating point that
attains it, is saved to `07_utility_max.csv` and shown below for the
unrestricted window. Both the threshold and the score come from the test set,
so this maximum is an optimistic ceiling on the performance observed here. It
is not an estimate of the performance of a threshold chosen in advance.

Two properties of the normalisation are enforced at run time, and the notebook
stops with an error if either fails. A threshold above every predicted
probability raises no alerts and must score exactly 0, and no operating point
may score above the upper bound of 1. The two lines printed below confirm that
both held.

```{r utility-sweep}
sweep_df <- if (length(all_sweeps) > 0) rbindlist(all_sweeps, fill = TRUE) else NULL

if (!is.null(sweep_df) && nrow(sweep_df) > 0) {
  setcolorder(sweep_df, c("variant", "model", "eval_window", "split_type",
                          "threshold"))
  save_csv(as.data.frame(sweep_df), "07_utility_sweep", DATA_DIR)

  # Invariant 1: no alerts => exactly the inaction strategy => score 0.
  no_alert <- sweep_df[n_alerts == 0]
  if (nrow(no_alert) > 0) {
    worst <- max(abs(no_alert$utility_normalised), na.rm = TRUE)
    if (worst > 1e-12)
      stop(sprintf("Utility normalisation broken: no-alert score is %.6f, not 0", worst))
    cat(sprintf("Check: %d no-alert operating points all score exactly 0.\n",
                nrow(no_alert)))
  }
  # Invariant 2: the optimal strategy is the upper bound.
  if (any(sweep_df$utility_normalised > 1 + 1e-12, na.rm = TRUE))
    stop("Utility normalisation broken: a score exceeds the optimal strategy")
  cat("Check: no operating point exceeds the optimal-strategy bound of 1.\n\n")

  best_df <- sweep_df[n_alerts > 0 & is.finite(utility_normalised),
                      .SD[which.max(utility_normalised)],
                      by = .(variant, model, eval_window, split_type)]
  save_csv(as.data.frame(best_df), "07_utility_max", DATA_DIR)

  knitr::kable(
    best_df[eval_window == "unrestricted"][order(variant, model),
            .(variant, model, threshold = signif(threshold, 3),
              alert_rate = signif(alert_rate, 3),
              detection_rate = round(detection_rate, 3),
              alert_burden = round(alert_burden, 1),
              utility_max = round(utility_normalised, 4))],
    caption = "Maximum achievable Utility Score over the threshold grid (unrestricted window)"
  )
  cat("\nSaved: 07_utility_sweep.csv, 07_utility_max.csv\n")
} else {
  cat("No threshold sweeps computed in this run.\n")
}
```

```{r utility-sweep-figure, fig.width=7, fig.height=4.2, out.width="90%", fig.cap="Normalised Utility Score against alert rate for every model and pre-registered label variant on the temporal test set. The dashed line at zero is the no-alert strategy: a curve below it describes a configuration that would be better switched off. Points mark each model's best operating point. The vertical axis is clipped, and every curve continues monotonically downwards to the right."}
if (!is.null(sweep_df) && nrow(sweep_df) > 0) {
  plot_dt <- sweep_df[eval_window == "unrestricted" &
                        variant %in% names(LABEL_VARIANTS) &
                        n_alerts > 0 & is.finite(utility_normalised)]
  if (nrow(plot_dt) > 0) {
    # The curves plunge to below -5 at high alert rates, which would compress
    # everything of interest onto the zero line. The axis is clipped to a window
    # around the decision boundary; the omitted tails are monotone downwards.
    y_hi   <- max(plot_dt$utility_normalised, na.rm = TRUE)
    y_span <- c(-4 * max(y_hi, 0.02), y_hi + 0.02)

    best_pts <- plot_dt[, .SD[which.max(utility_normalised)], by = .(variant, model)]

    p_sweep <- ggplot(plot_dt, aes(x = alert_rate, y = utility_normalised,
                                   colour = model)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
      geom_line(linewidth = 0.6) +
      geom_point(data = best_pts, size = 1.4) +
      facet_wrap(~ variant, nrow = 1) +
      scale_x_log10(labels = function(x) paste0(signif(100 * x, 2), "%")) +
      coord_cartesian(ylim = y_span) +
      labs(x = "Alert rate (per cent of person-hours alerted, log scale)",
           y = "Normalised Utility Score",
           colour = "Model",
           title = "Utility Score across the threshold grid",
           subtitle = paste("0 is the no-alert strategy; points mark each model's best",
                            "operating point.\nThe axis is clipped -- every curve",
                            "continues monotonically downwards to the right.")) +
      theme_minimal(base_size = 10) +
      theme(legend.position = "bottom",
            plot.subtitle = element_text(size = 8))
    ggsave(file.path(FIGURES_DIR, "07_utility_threshold_sweep.png"), p_sweep,
           width = 9, height = 4.6, dpi = 150)
    print(p_sweep)
  }
}
```

## Alternative-Label Evaluation (Protocol Amendments 2 and 4)

The post-hoc Variants D and E (Protocol Amendment 2) and F (Protocol Amendment
4) are scored with the same metric suite. Their results are written to
`07_metric_results_altlabel.parquet` and `07_metric_results_altlabel.csv`,
separately from the pre-registered results, so they cannot enter H1 or the
variance decomposition. Variants D and E define the event differently from the
Sepsis-3 variants, and Variant F re-times onset, so their event counts and
prevalences differ from those of Variants A--C. The table therefore adds the
number of events to each row. The comparisons between these variants and
Variant B are made in `09_variance_decomposition.Rmd`. Variant F's
treatment-anchored rows give the empirical counterpart to H3 described above.

When only some of these variants are run, the rows of the variants not in the
run are carried forward from the existing file rather than deleted.

```{r show-alt-results}
alt_df <- if (!is.null(results_df))
            results_df[results_df$variant %in% names(ALT_LABEL_VARIANTS), ] else NULL

if (!is.null(alt_df) && nrow(alt_df) > 0) {
  alt_df$label_family <- vapply(alt_df$variant, variant_type, character(1))

  # This write used to be a plain overwrite, so a partial pass silently
  # truncated the file to whatever ran: V2_VARIANTS=F deleted D's and E's rows,
  # and four consumers read it (09, 12 and thesis_constants.R twice), so the
  # alt-label table, family E3 and the phi_anchor macros all lost two variants
  # with no error anywhere. Merge instead: rows for the variants in THIS run
  # replace their previous selves, every other variant is carried forward.
  alt_prev <- tryCatch({
    p <- file.path(DATA_DIR, "07_metric_results_altlabel.parquet")
    if (file.exists(p)) as.data.frame(arrow::read_parquet(p)) else NULL
  }, error = function(e) NULL)
  if (!is.null(alt_prev) && nrow(alt_prev) > 0 &&
      identical(sort(names(alt_prev)), sort(names(alt_df)))) {
    carried <- alt_prev[!(alt_prev$variant %in% unique(alt_df$variant)), , drop = FALSE]
    if (nrow(carried) > 0) {
      cat(sprintf("Carrying forward %d alt-label rows for variants not in this run: %s\n",
                  nrow(carried), paste(sort(unique(carried$variant)), collapse = ", ")))
      alt_df <- rbind(alt_df, carried[, names(alt_df), drop = FALSE])
    }
  }
  alt_df <- alt_df[order(alt_df$variant, alt_df$eval_window, alt_df$model), , drop = FALSE]

  save_parquet(alt_df, "07_metric_results_altlabel", DATA_DIR)
  save_csv(alt_df,     "07_metric_results_altlabel", DATA_DIR)
  knitr::kable(
    alt_df[order(alt_df$variant, alt_df$eval_window, alt_df$model),
           c(display_cols, "n_sepsis_events")],
    caption = "Evaluation results under the alternative (non-Sepsis-3) labels"
  )
  cat("\nSaved: 07_metric_results_altlabel.parquet\n")
} else {
  cat("No alternative-label variants selected in this run.\n")
}
```

## Feature-Ablation Evaluation (Protocol Amendment 5)

Protocol Amendment 5 is a post-hoc, exploratory analysis. It enters no
hypothesis test and no part of the variance decomposition. The ablated primary
model and GBT from stages 05 and 06 are fitted without the eight
action-derived covariates, which record clinicians' actions rather than the
patient's state. These are the two vasopressor indicators (`vaso_any` and
`norepi_epi_any`) and the six indicators of whether a laboratory test has been
measured (lactate, creatinine, total bilirubin, platelets, white cell count and
P/F ratio). The remaining 14 covariates, all physiological, are retained,
including the forward-filled laboratory values. A laboratory value exists only
because the test was ordered, so these values also carry some information about
clinicians' actions. They are retained because removing them would also remove
physiological signal, and the ablation would then no longer isolate the
action-derived covariates.

Everything else in the fit is held fixed, including, for the GBT, the
hyperparameters and the stay-level validation subset. The ablated models are
scored with the same metric suite on the same temporal test person-hours as
the full models, in the unrestricted window. Each ablated result is paired with
the full model's result for the same variant and model, and the difference
(ablated minus full) is the quantity of interest. The difference in maximum
Utility Score compares two test-set maxima and carries the caveat given above.
The output is saved to `07_ablation_metrics.parquet` and
`07_ablation_metrics.csv`.

```{r evaluate-ablation}
step("--- Evaluating the ablated arm (Protocol Amendment 5) ---")

ablation_rows <- list()

for (vid in intersect(active_variants(), names(LABEL_VARIANTS))) {
  abl_paths <- c(primary = file.path(DATA_DIR,
                    paste0("05_predictions_primary_ablated_", vid, ".parquet")),
                 gbt     = file.path(DATA_DIR,
                    paste0("06_predictions_gbt_ablated_", vid, ".parquet")))
  abl_cols  <- c(primary = "pred_sepsis", gbt = "pred_gbt")

  for (mod_name in names(abl_paths)) {
    if (!file.exists(abl_paths[[mod_name]])) next
    d        <- as.data.table(arrow::read_parquet(abl_paths[[mod_name]]))
    pred_col <- abl_cols[[mod_name]]
    if (!pred_col %in% names(d)) next

    y     <- as.integer(d$event == 1)
    pred  <- as.numeric(d[[pred_col]])
    valid <- !is.na(pred) & !is.na(y)
    if (sum(valid) < 10) next

    roc_pr <- compute_roc_pr(y[valid], pred[valid])
    calib  <- tryCatch(compute_calibration(y[valid], pred[valid]),
                       error = function(e) list(intercept = NA, slope = NA))
    sweep  <- tryCatch(utility_curve(d[valid], pred_col,
                                     utility_threshold_grid(pred[valid])),
                       error = function(e) NULL)
    best   <- if (!is.null(sweep)) {
      cand <- sweep[n_alerts > 0 & is.finite(utility_normalised)]
      if (nrow(cand) > 0) cand[which.max(utility_normalised)] else NULL
    } else NULL

    ablation_rows[[paste(vid, mod_name, sep = "-")]] <- data.frame(
      variant         = vid,
      variant_name    = variant_name(vid),
      model           = mod_name,
      arm             = "ablated",
      eval_window     = "unrestricted",
      split_type      = "temporal",
      # The ablated arm drops ACTION_DERIVED_FEATURES from the declared
      # covariate vector; both counts come from config.R rather than being
      # asserted, so the "8 of 22" in the thesis is checkable against the file.
      n_features      = length(model_features()) - length(ACTION_DERIVED_FEATURES),
      n_features_full = length(model_features()),
      n_rows          = sum(valid),
      n_sepsis_events = sum(y[valid]),
      auroc           = round(roc_pr$auroc, 4),
      auprc           = round(roc_pr$auprc, 4),
      calib_slope     = round(calib$slope, 4),
      utility_max     = if (is.null(best)) NA_real_ else round(best$utility_normalised, 4)
    )
    rm(d); invisible(gc(FALSE))
  }
}

if (length(ablation_rows) > 0) {
  abl_res <- as.data.table(do.call(rbind, ablation_rows))

  # Paired against the full arm on the same variant, model and window. The
  # delta is the quantity of interest; the two AUROCs on their own are not.
  full_res <- as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, "07_metric_results.parquet")))
  full_res <- full_res[eval_window == "unrestricted" & split_type == "temporal",
                       .(variant, model,
                         auroc_full = auroc, auprc_full = auprc,
                         calib_slope_full = calib_slope,
                         utility_max_full = utility_max)]

  abl_res <- merge(abl_res, full_res, by = c("variant", "model"), all.x = TRUE)
  abl_res[, `:=`(
    delta_auroc       = round(auroc       - auroc_full,       4),
    delta_auprc       = round(auprc       - auprc_full,       4),
    delta_utility_max = round(utility_max - utility_max_full, 4)
  )]
  setorder(abl_res, variant, model)

  save_parquet(as.data.frame(abl_res), "07_ablation_metrics", DATA_DIR)
  save_csv(as.data.frame(abl_res),     "07_ablation_metrics", DATA_DIR)
  knitr::kable(
    abl_res[, .(variant, model, n_sepsis_events,
                auroc_full, auroc, delta_auroc,
                utility_max_full, utility_max, delta_utility_max)],
    caption = "Protocol Amendment 5: performance with and without the action-derived features"
  )
  cat("\nSaved: 07_ablation_metrics.parquet / .csv\n")
} else {
  cat("No ablated predictions found; the ablation arm has not been run.\n")
}
```

## Random-Split Evaluation (H4: Split Optimism)

H4 asks whether a random train/test split overstates performance relative to
the temporal split (Guo et al., 2022). The same metric suite is applied to the
predictions of the models refitted on the random split in stages 05 and 06.
That split, built in `03_person_hours.Rmd`, assigns 25 per cent of stays to the
test set, so all hours of a stay fall on the same side. With one stay per
patient, it is also a patient-level split. Only Variant B is evaluated, because
H4 holds the label fixed. The results are saved to
`07_metric_results_random.parquet`, from which `09_variance_decomposition.Rmd`
computes the split component of the decomposition.

```{r evaluate-random-split}
step("--- Random-split evaluation for H4 ---")

random_results <- list()
vid <- "B"

prim_random_path <- file.path(DATA_DIR, paste0("05_predictions_primary_random_", vid, ".parquet"))
comp_random_path <- file.path(DATA_DIR, paste0("06_predictions_comparators_random_", vid, ".parquet"))

if (file.exists(prim_random_path)) {
  ph_prim_r <- as.data.table(arrow::read_parquet(prim_random_path))
  ph_comp_r <- if (file.exists(comp_random_path))
                 as.data.table(arrow::read_parquet(comp_random_path)) else NULL

  ph_all_r <- ph_prim_r
  if (!is.null(ph_comp_r)) {
    merge_cols_r <- c("stay_id", "hour", setdiff(names(ph_comp_r), names(ph_prim_r)))
    ph_all_r <- merge(ph_all_r, ph_comp_r[, ..merge_cols_r],
                      by = c("stay_id", "hour"), all.x = TRUE)
  }

  eval_windows_r <- list(
    unrestricted = ph_all_r,
    treatment_anchored = {
      if ("first_abx_hour" %in% names(ph_all_r)) {
        anchor_h_r <- if ("first_culture_hour" %in% names(ph_all_r))
                        pmin(ph_all_r$first_abx_hour, ph_all_r$first_culture_hour, na.rm = TRUE)
                      else ph_all_r$first_abx_hour
        ph_all_r[is.na(anchor_h_r) | hour < anchor_h_r]
      } else ph_all_r
    }
  )

  for (win_name in names(eval_windows_r)) {
    data_win_r <- eval_windows_r[[win_name]]
    if (nrow(data_win_r) == 0) next

    for (mod_name in names(MODEL_NAMES)) {
      pred_col <- MODEL_NAMES[[mod_name]]
      if (!pred_col %in% names(data_win_r)) next
      valid <- !is.na(data_win_r[[pred_col]])
      if (sum(valid) < 100) next

      y    <- as.integer(data_win_r$event == 1)
      pred <- as.numeric(data_win_r[[pred_col]])

      roc_pr  <- compute_roc_pr(y[valid], pred[valid])
      calib   <- compute_calibration(y[valid], pred[valid])
      alerts  <- as.integer(pred >= 0.5)
      burden  <- compute_alert_burden(y[valid], alerts[valid])
      utility <- tryCatch(
        compute_utility_score(data_win_r[valid], pred_col, threshold = 0.5),
        error = function(e) list(utility_normalised = NA, utility_total = NA)
      )
      lead_t  <- tryCatch(
        compute_lead_time(data_win_r[valid], pred_col),
        error = function(e) NA_real_
      )

      sweep_r <- tryCatch(
        utility_curve(data_win_r[valid], pred_col,
                      utility_threshold_grid(pred[valid])),
        error = function(e) NULL
      )
      best_r <- if (!is.null(sweep_r)) {
        cand_r <- sweep_r[n_alerts > 0 & is.finite(utility_normalised)]
        if (nrow(cand_r) > 0) cand_r[which.max(utility_normalised)] else NULL
      } else NULL

      result_row <- data.frame(
        variant            = vid,
        variant_name       = LABEL_VARIANTS[[vid]]$name,
        model              = mod_name,
        eval_window        = win_name,
        split_type         = "random",
        n_rows             = sum(valid),
        n_sepsis_events    = sum(y[valid]),
        auroc              = round(roc_pr$auroc, 4),
        auprc              = round(roc_pr$auprc, 4),
        calib_intercept    = round(calib$intercept, 4),
        calib_slope        = round(calib$slope, 4),
        utility_normalised = round(utility$utility_normalised, 4),
        alert_burden       = round(burden, 2),
        lead_time_h        = round(lead_t, 2),
        utility_max        = if (is.null(best_r)) NA_real_ else round(best_r$utility_normalised, 4),
        utility_opt_threshold = if (is.null(best_r)) NA_real_ else signif(best_r$threshold, 4),
        utility_opt_alert_rate = if (is.null(best_r)) NA_real_ else signif(best_r$alert_rate, 4),
        utility_opt_burden  = if (is.null(best_r)) NA_real_ else round(best_r$alert_burden, 2),
        utility_opt_detection = if (is.null(best_r)) NA_real_ else round(best_r$detection_rate, 4)
      )
      random_results[[paste(vid, mod_name, win_name, "random", sep = "-")]] <- result_row
    }
  }

  if (length(random_results) > 0) {
    random_results_df <- do.call(rbind, random_results)
    save_parquet(random_results_df, "07_metric_results_random", DATA_DIR)

    knitr::kable(
      random_results_df[order(random_results_df$eval_window, random_results_df$model),
                        c("variant", "model", "eval_window", "auroc", "auprc",
                          "utility_normalised", "utility_max", "alert_burden",
                          "lead_time_h")],
      caption = "Evaluation results: random split (Variant B only, H4)"
    )
    cat("\nSaved: 07_metric_results_random.parquet\n")
  }
} else {
  cat("Random-split primary predictions not found; skipping H4 evaluation.\n")
  cat("Run 05_primary_model.Rmd first to generate random-split predictions.\n")
}
```
