---
title: "07 Metrics Suite - Full Evaluation Metric Suite"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook applies a full metric suite following Wang et al. (2025) and Van Calster et al. (2019).

**Primary:** PhysioNet Utility Score (Reyna et al., 2020), calibration (intercept + slope).

**Secondary:** AUPRC, AUROC, decision-curve net benefit, alert burden ($\leq 1.4$ benchmark; Moor et al., 2023), median lead time.

All metrics are computed in two windows: **unrestricted** (all test hours) and **treatment-anchored** (pre-treatment hours only; Kamran et al., 2024).

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(PRROC)
  library(dcurves)
  library(CalibrationCurves)
  library(dplyr)
  library(ggplot2)
})

source("config.R")
source("utils.R")

step("=== Full Evaluation Metric Suite ===")
```

## Metric Functions

Standalone metric functions, reused across internal, external, and subgroup
evaluations. Keeping one implementation matters more than it looks: internal
and external AUROC are differenced in H5, and per-hour and per-patient
operating points are compared in stage 11. Two scoring paths that disagreed
even slightly would put that disagreement into the contrast.

```{r metric-fns}
# --------------------------------------------------------------------------- #
# PhysioNet Utility Score (Reyna et al., 2020), stay-level operationalisation
# --------------------------------------------------------------------------- #
# Raw utility, per stay:
#   sepsis stay alerted in the reward window [onset + UTILITY_TP_MIN, onset]
#       + lead / PREDICTION_HORIZON_H, where lead is the warning time of the
#       *earliest* alert in that window. A full-horizon warning scores 1; an
#       alert that only coincides with onset scores 0.
#   sepsis stay alerted only in (onset, onset + UTILITY_TP_MAX]
#       UTILITY_FP per late alert hour.
#   sepsis stay never alerted in either window
#       UTILITY_FN.
#   any alert hour outside a reward window, on any stay
#       UTILITY_FP.
#
# The raw total is referenced to the two strategies of Reyna et al. (2020):
#   U_inaction = UTILITY_FN            x n_sepsis_stays   (never alert)
#   U_optimal  = UTILITY_TP_MAX_REWARD x n_sepsis_stays   (perfect, no false alarms)
#   normalised = (U - U_inaction) / (U_optimal - U_inaction)
# so that 1 is a perfect forecaster, 0 is exactly the no-alert strategy, and a
# negative score is worse than issuing no alerts at all. Without the inaction
# baseline in both numerator and denominator a model that raises no alert scores
# -1 rather than 0, which contradicts the definition of the metric.
#
# `utility_curve()` evaluates the whole threshold grid in one pass: the per-stay
# reward is monotone in a falling threshold, so the sweep costs no more than a
# single sort plus one traversal of the reward-window rows.
utility_curve <- function(ph_pred, pred_col, thresholds) {
  dt <- as.data.table(ph_pred)
  p  <- as.numeric(dt[[pred_col]])
  ok <- is.finite(p)

  p     <- p[ok]
  stay  <- dt$stay_id[ok]
  hour  <- as.numeric(dt$hour)[ok]
  onset <- as.numeric(dt$onset_hour)[ok]
  y_hr  <- as.integer(dt$event)[ok] == 1L

  thresholds <- sort(unique(as.numeric(thresholds)), decreasing = TRUE)
  sep_stays  <- unique(stay[!is.na(onset)])
  n_sep      <- length(sep_stays)

  if (n_sep == 0L || length(thresholds) == 0L || length(p) == 0L)
    return(data.table(threshold = thresholds, n_alerts = NA_integer_,
                      alert_rate = NA_real_, n_sepsis_stays = n_sep,
                      n_detected_stays = NA_integer_, detection_rate = NA_real_,
                      utility_total = NA_real_, utility_normalised = NA_real_,
                      alert_burden = NA_real_))

  in_early <- !is.na(onset) & hour >= (onset + UTILITY_TP_MIN) & hour <= onset
  in_late  <- !is.na(onset) & hour >  onset & hour <= (onset + UTILITY_TP_MAX)
  n_early  <- sum(in_early)
  n_late   <- sum(in_late)

  # Reward-window rows only, sorted by descending score: these are the rows
  # whose status changes as the threshold falls.
  ev <- data.table(
    si   = c(match(stay[in_early], sep_stays), match(stay[in_late], sep_stays)),
    p    = c(p[in_early], p[in_late]),
    v    = c((onset[in_early] - hour[in_early]) / PREDICTION_HORIZON_H,
             rep(NA_real_, n_late)),
    late = c(rep(FALSE, n_early), rep(TRUE, n_late))
  )
  setorder(ev, -p)

  p_all <- sort(p)
  p_tp  <- sort(p[y_hr])
  count_ge <- function(sorted_x, tau)
    length(sorted_x) - findInterval(tau, sorted_x, left.open = TRUE)

  best       <- rep(NA_real_, n_sep)  # best TP reward so far, per sepsis stay
  late_cnt   <- integer(n_sep)        # late alerts so far, per sepsis stay
  sum_best   <- 0                     # sum of best[] over detected stays
  n_det      <- 0L                    # sepsis stays with a reward-window alert
  late_undet <- 0L                    # late alerts on stays not (yet) detected
  n_undet_late <- 0L                  # undetected stays carrying >= 1 late alert
  n_reward_alerts <- 0L               # alert hours inside a reward window

  j   <- 1L
  res <- vector("list", length(thresholds))

  for (k in seq_along(thresholds)) {
    tau <- thresholds[k]
    while (j <= nrow(ev) && ev$p[j] >= tau) {
      si <- ev$si[j]
      if (ev$late[j]) {
        late_cnt[si] <- late_cnt[si] + 1L
        if (is.na(best[si])) {
          late_undet <- late_undet + 1L
          if (late_cnt[si] == 1L) n_undet_late <- n_undet_late + 1L
        }
      } else {
        v <- ev$v[j]
        if (is.na(best[si])) {
          best[si] <- v
          sum_best <- sum_best + v
          n_det    <- n_det + 1L
          if (late_cnt[si] > 0L) {
            late_undet   <- late_undet - late_cnt[si]
            n_undet_late <- n_undet_late - 1L
          }
        } else if (v > best[si]) {
          sum_best <- sum_best + (v - best[si])
          best[si] <- v
        }
      }
      n_reward_alerts <- n_reward_alerts + 1L
      j <- j + 1L
    }

    n_alerts <- count_ge(p_all, tau)
    n_tp_hr  <- count_ge(p_tp,  tau)

    u_raw <- sum_best +
             UTILITY_FP * late_undet +
             UTILITY_FN * (n_sep - n_det - n_undet_late) +
             UTILITY_FP * (n_alerts - n_reward_alerts)

    u_inaction <- UTILITY_FN            * n_sep
    u_optimal  <- UTILITY_TP_MAX_REWARD * n_sep

    res[[k]] <- data.table(
      threshold          = tau,
      n_alerts           = n_alerts,
      alert_rate         = n_alerts / length(p),
      n_sepsis_stays     = n_sep,
      n_detected_stays   = n_det,
      detection_rate     = n_det / n_sep,
      utility_total      = u_raw,
      utility_normalised = (u_raw - u_inaction) / (u_optimal - u_inaction),
      alert_burden       = if (n_tp_hr == 0L) Inf else (n_alerts - n_tp_hr) / n_tp_hr
    )
  }

  out <- rbindlist(res)
  setorder(out, threshold)
  out[]
}

# Threshold grid, defined on the alert rate so that it adapts to each model's
# calibration. At an hourly event rate below 0.5 per cent no model's predicted
# probability approaches 0.5, so a grid on the probability scale would evaluate
# nothing but the no-alert strategy. The default 0.5 cut-off is appended so the
# sweep table always contains the operating point used elsewhere.
utility_threshold_grid <- function(p) {
  p <- p[is.finite(p)]
  if (length(p) == 0L) return(numeric(0))
  rates <- exp(seq(log(UTILITY_SWEEP_MIN_RATE), log(UTILITY_SWEEP_MAX_RATE),
                   length.out = UTILITY_SWEEP_N_GRID))
  taus  <- as.numeric(quantile(p, probs = 1 - rates, names = FALSE, type = 1))
  sort(unique(c(taus, 0.5)))
}

# Single-threshold wrapper, used for the default 0.5 operating point.
compute_utility_score <- function(ph_pred, pred_col, threshold = 0.5) {
  row <- utility_curve(ph_pred, pred_col, threshold)
  list(utility_total      = row$utility_total[1],
       utility_max        = UTILITY_TP_MAX_REWARD * row$n_sepsis_stays[1],
       utility_normalised = row$utility_normalised[1])
}

# Calibration
compute_calibration <- function(y_true, pred_prob) {
  df <- data.frame(y = y_true, p = pred_prob)
  df <- df[!is.na(df$p) & !is.na(df$y), ]
  if (nrow(df) < 10) return(list(intercept = NA, slope = NA))

  calib_model <- tryCatch(
    glm(y ~ offset(qlogis(pmax(pmin(p, 0.999), 0.001))),
        data = df, family = binomial),
    error = function(e) NULL
  )
  intercept <- if (!is.null(calib_model)) coef(calib_model)[1] else NA_real_
  calib_slope <- tryCatch({
    mod <- glm(y ~ qlogis(pmax(pmin(p, 0.999), 0.001)), data = df, family = binomial)
    coef(mod)[2]
  }, error = function(e) NA_real_)

  list(intercept = intercept, slope = calib_slope)
}

# AUROC + AUPRC
compute_roc_pr <- function(y_true, pred_prob) {
  if (sum(y_true, na.rm = TRUE) == 0 || sum(!y_true, na.rm = TRUE) == 0)
    return(list(auroc = NA, auprc = NA))
  valid <- !is.na(pred_prob) & !is.na(y_true)
  if (sum(valid) < 10) return(list(auroc = NA, auprc = NA))
  roc_obj <- tryCatch(
    PRROC::roc.curve(scores.class0 = pred_prob[valid], weights.class0 = y_true[valid]),
    error = function(e) NULL
  )
  pr_obj <- tryCatch(
    PRROC::pr.curve(scores.class0 = pred_prob[valid], weights.class0 = y_true[valid]),
    error = function(e) NULL
  )
  list(auroc = if (!is.null(roc_obj)) roc_obj$auc else NA_real_,
       auprc = if (!is.null(pr_obj))  pr_obj$auc.integral else NA_real_)
}

# Alert burden
compute_alert_burden <- function(y_true, alerts) {
  tp <- sum(alerts == 1 & y_true == 1, na.rm = TRUE)
  fp <- sum(alerts == 1 & y_true == 0, na.rm = TRUE)
  if (tp == 0) return(Inf)
  fp / tp
}

# Lead time to clinical recognition
compute_lead_time <- function(ph_pred, pred_col, threshold = 0.5) {
  sepsis_stays <- ph_pred[!is.na(onset_hour)]
  if (nrow(sepsis_stays) == 0) return(NA_real_)
  lead_times <- sepsis_stays[, {
    rows_ord    <- .SD[order(hour)]
    first_alert <- rows_ord[get(pred_col) >= threshold, min(hour)]
    if (is.infinite(first_alert) || is.na(first_alert)) NA_real_
    else as.numeric(onset_hour[1]) - as.numeric(first_alert)
  }, by = stay_id]
  median(lead_times$V1, na.rm = TRUE)
}
```

## Evaluation Loop

The full models x variants x windows grid is evaluated. Decision-curve analysis runs where both primary and GBT predictions are available.

```{r evaluate-all}
MODEL_NAMES <- c(primary = "pred_sepsis", news2 = "pred_news2",
                  cox = "pred_cox", gbt = "pred_gbt",
                  qsofa = "pred_qsofa", sirs = "pred_sirs")

all_results <- list()
all_sweeps  <- list()

run_variants <- active_variants()

for (vid in run_variants) {
  step(sprintf("--- Evaluating Variant %s (%s) ---", vid, variant_name(vid)))

  prim_pred_path <- file.path(DATA_DIR, paste0("05_predictions_primary_",     vid, ".parquet"))
  comp_pred_path <- file.path(DATA_DIR, paste0("06_predictions_comparators_", vid, ".parquet"))

  if (!file.exists(prim_pred_path)) {
    v2_log(sprintf("  Skipping Variant %s: primary predictions not found", vid))
    next
  }

  ph_prim <- as.data.table(arrow::read_parquet(prim_pred_path))
  ph_comp <- if (file.exists(comp_pred_path))
               as.data.table(arrow::read_parquet(comp_pred_path)) else NULL

  ph_all <- ph_prim
  if (!is.null(ph_comp)) {
    merge_cols <- c("stay_id", "hour", setdiff(names(ph_comp), names(ph_prim)))
    ph_all <- merge(ph_all, ph_comp[, ..merge_cols], by = c("stay_id", "hour"), all.x = TRUE)
  }

  eval_windows <- list(
    unrestricted = ph_all,
    treatment_anchored = {
      if ("first_abx_hour" %in% names(ph_all)) {
        anchor_h <- if ("first_culture_hour" %in% names(ph_all))
                      pmin(ph_all$first_abx_hour, ph_all$first_culture_hour, na.rm = TRUE)
                    else ph_all$first_abx_hour
        ph_all[is.na(anchor_h) | hour < anchor_h]
      } else ph_all
    }
  )

  for (win_name in names(eval_windows)) {
    data_win <- eval_windows[[win_name]]
    if (nrow(data_win) == 0) next

    for (mod_name in names(MODEL_NAMES)) {
      pred_col <- MODEL_NAMES[[mod_name]]
      if (!pred_col %in% names(data_win)) next
      valid <- !is.na(data_win[[pred_col]])
      if (sum(valid) < 100) next

      y    <- as.integer(data_win$event == 1)
      pred <- as.numeric(data_win[[pred_col]])

      roc_pr  <- compute_roc_pr(y[valid], pred[valid])
      calib   <- compute_calibration(y[valid], pred[valid])
      alerts  <- as.integer(pred >= 0.5)
      burden  <- compute_alert_burden(y[valid], alerts[valid])
      utility <- tryCatch(
        compute_utility_score(data_win[valid], pred_col, threshold = 0.5),
        error = function(e) list(utility_normalised = NA, utility_total = NA)
      )
      lead_t  <- tryCatch(
        compute_lead_time(data_win[valid], pred_col),
        error = function(e) NA_real_
      )

      # Threshold sweep. Every Utility Score above is read off the default 0.5
      # cut-off, which no deployment would use; the sweep reports what each
      # model can achieve at its best operating point.
      sweep <- tryCatch(
        utility_curve(data_win[valid], pred_col,
                      utility_threshold_grid(pred[valid])),
        error = function(e) NULL
      )
      best <- if (!is.null(sweep)) {
        cand <- sweep[n_alerts > 0 & is.finite(utility_normalised)]
        if (nrow(cand) > 0) cand[which.max(utility_normalised)] else NULL
      } else NULL

      if (!is.null(sweep)) {
        sweep[, `:=`(variant = vid, model = mod_name, eval_window = win_name,
                     split_type = "temporal")]
        all_sweeps[[paste(vid, mod_name, win_name, sep = "-")]] <- sweep
      }

      result_row <- data.frame(
        variant            = vid,
        variant_name       = variant_name(vid),
        model              = mod_name,
        eval_window        = win_name,
        split_type         = "temporal",
        n_rows             = sum(valid),
        n_sepsis_events    = sum(y[valid]),
        auroc              = round(roc_pr$auroc, 4),
        auprc              = round(roc_pr$auprc, 4),
        calib_intercept    = round(calib$intercept, 4),
        calib_slope        = round(calib$slope, 4),
        utility_normalised = round(utility$utility_normalised, 4),
        alert_burden       = round(burden, 2),
        lead_time_h        = round(lead_t, 2),
        utility_max        = if (is.null(best)) NA_real_ else round(best$utility_normalised, 4),
        utility_opt_threshold = if (is.null(best)) NA_real_ else signif(best$threshold, 4),
        utility_opt_alert_rate = if (is.null(best)) NA_real_ else signif(best$alert_rate, 4),
        utility_opt_burden  = if (is.null(best)) NA_real_ else round(best$alert_burden, 2),
        utility_opt_detection = if (is.null(best)) NA_real_ else round(best$detection_rate, 4)
      )
      all_results[[paste(vid, mod_name, win_name, sep = "-")]] <- result_row
    }
  }

  # Decision-curve analysis
  ph_test_for_dca <- ph_all[!is.na(pred_sepsis)]
  if (nrow(ph_test_for_dca) > 0 && "pred_gbt" %in% names(ph_test_for_dca)) {
    ph_test_for_dca[, outcome := as.integer(event == 1)]
    dca_result <- tryCatch(
      dcurves::dca(outcome ~ pred_sepsis + pred_gbt,
                   data       = as.data.frame(ph_test_for_dca),
                   thresholds = seq(0.01, 0.20, 0.01)),
      error = function(e) NULL
    )
    if (!is.null(dca_result)) {
      dca_df <- tryCatch(dca_result$dca, error = function(e) as.data.frame(dca_result))
      if (!is.null(dca_df) && nrow(dca_df) > 0) {
        dca_df$variant <- vid
        save_parquet(as.data.frame(dca_df), paste0("07_dca_", vid), DATA_DIR)
      }
    }
  }
}
```

## Results and Interpretation

The central performance comparison. Three readings are worth making from the
table, and each answers a different question.

Primary against GBT on the Utility Score speaks to H6, though the two models
can order differently on utility than on AUROC; where they do, the disagreement
is itself informative and the thesis reports both orderings rather than
choosing one.

Treatment-anchored against unrestricted speaks to H3. A gap here would be
leakage from clinician suspicion. No gap is harder to interpret than it looks,
because under a conjunction onset rule the pre-treatment window may hold no
events at all, in which case the comparison has nothing to measure.

Calibration slope is read against 1.0. A slope well below 1 means predicted
risks are too extreme at both ends, which a discrimination metric cannot see
and which matters for any threshold a clinician would actually deploy.

```{r show-results}
display_cols <- c("variant", "model", "eval_window", "auroc", "auprc",
                  "utility_normalised", "utility_max", "alert_burden", "lead_time_h")

results_df <- if (length(all_results) > 0) do.call(rbind, all_results) else NULL

# The pre-registered variance decomposition reads 07_metric_results.parquet, so
# the alternative-label variants of Protocol Amendment 2 are written to their
# own file and cannot leak into H1 or the decomposition table.
prereg_df <- if (!is.null(results_df))
               results_df[results_df$variant %in% names(LABEL_VARIANTS), ] else NULL

if (!is.null(prereg_df) && nrow(prereg_df) > 0) {
  save_parquet(prereg_df, "07_metric_results", DATA_DIR)
  knitr::kable(
    prereg_df[order(prereg_df$variant, prereg_df$eval_window, prereg_df$model),
              display_cols],
    caption = "Evaluation results: all models x variants x windows (temporal split)"
  )
  cat("\nSaved: 07_metric_results.parquet\n")
} else {
  cat("No pre-registered-variant results in this run; 07_metric_results.parquet left unchanged.\n")
}
```

## Utility Score Across the Threshold Grid

Every Utility Score in the table above is read at the default 0.5 cut-off. At an
hourly event rate below 0.5 per cent no model's predicted probability comes near
0.5, so that cut-off measures the no-alert strategy for some models and an
arbitrary point for the others. The full grid is written to
`07_utility_sweep.csv`; the maximum over all thresholds that raise at least one
alert is the honest statement of what each model can achieve.

Two invariants are checked here rather than asserted: a threshold above every
predicted probability must score exactly 0 (it *is* the no-alert strategy), and
the normalised score must never exceed 1.

```{r utility-sweep}
sweep_df <- if (length(all_sweeps) > 0) rbindlist(all_sweeps, fill = TRUE) else NULL

if (!is.null(sweep_df) && nrow(sweep_df) > 0) {
  setcolorder(sweep_df, c("variant", "model", "eval_window", "split_type",
                          "threshold"))
  save_csv(as.data.frame(sweep_df), "07_utility_sweep", DATA_DIR)

  # Invariant 1: no alerts => exactly the inaction strategy => score 0.
  no_alert <- sweep_df[n_alerts == 0]
  if (nrow(no_alert) > 0) {
    worst <- max(abs(no_alert$utility_normalised), na.rm = TRUE)
    if (worst > 1e-12)
      stop(sprintf("Utility normalisation broken: no-alert score is %.6f, not 0", worst))
    cat(sprintf("Check: %d no-alert operating points all score exactly 0.\n",
                nrow(no_alert)))
  }
  # Invariant 2: the optimal strategy is the upper bound.
  if (any(sweep_df$utility_normalised > 1 + 1e-12, na.rm = TRUE))
    stop("Utility normalisation broken: a score exceeds the optimal strategy")
  cat("Check: no operating point exceeds the optimal-strategy bound of 1.\n\n")

  best_df <- sweep_df[n_alerts > 0 & is.finite(utility_normalised),
                      .SD[which.max(utility_normalised)],
                      by = .(variant, model, eval_window, split_type)]
  save_csv(as.data.frame(best_df), "07_utility_max", DATA_DIR)

  knitr::kable(
    best_df[eval_window == "unrestricted"][order(variant, model),
            .(variant, model, threshold = signif(threshold, 3),
              alert_rate = signif(alert_rate, 3),
              detection_rate = round(detection_rate, 3),
              alert_burden = round(alert_burden, 1),
              utility_max = round(utility_normalised, 4))],
    caption = "Maximum achievable Utility Score over the threshold grid (unrestricted window)"
  )
  cat("\nSaved: 07_utility_sweep.csv, 07_utility_max.csv\n")
} else {
  cat("No threshold sweeps computed in this run.\n")
}
```

```{r utility-sweep-figure, fig.width=7, fig.height=4.2, out.width="90%", fig.cap="Normalised Utility Score against alert rate for every model and pre-registered label variant on the temporal test set. The dashed line at zero is the no-alert strategy: a curve below it describes a configuration that would be better switched off. Points mark each model's best operating point. The vertical axis is clipped, and every curve continues monotonically downwards to the right."}
if (!is.null(sweep_df) && nrow(sweep_df) > 0) {
  plot_dt <- sweep_df[eval_window == "unrestricted" &
                        variant %in% names(LABEL_VARIANTS) &
                        n_alerts > 0 & is.finite(utility_normalised)]
  if (nrow(plot_dt) > 0) {
    # The curves plunge to below -5 at high alert rates, which would compress
    # everything of interest onto the zero line. The axis is clipped to a window
    # around the decision boundary; the omitted tails are monotone downwards.
    y_hi   <- max(plot_dt$utility_normalised, na.rm = TRUE)
    y_span <- c(-4 * max(y_hi, 0.02), y_hi + 0.02)

    best_pts <- plot_dt[, .SD[which.max(utility_normalised)], by = .(variant, model)]

    p_sweep <- ggplot(plot_dt, aes(x = alert_rate, y = utility_normalised,
                                   colour = model)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
      geom_line(linewidth = 0.6) +
      geom_point(data = best_pts, size = 1.4) +
      facet_wrap(~ variant, nrow = 1) +
      scale_x_log10(labels = function(x) paste0(signif(100 * x, 2), "%")) +
      coord_cartesian(ylim = y_span) +
      labs(x = "Alert rate (per cent of person-hours alerted, log scale)",
           y = "Normalised Utility Score",
           colour = "Model",
           title = "Utility Score across the threshold grid",
           subtitle = paste("0 is the no-alert strategy; points mark each model's best",
                            "operating point.\nThe axis is clipped -- every curve",
                            "continues monotonically downwards to the right.")) +
      theme_minimal(base_size = 10) +
      theme(legend.position = "bottom",
            plot.subtitle = element_text(size = 8))
    ggsave(file.path(FIGURES_DIR, "07_utility_threshold_sweep.png"), p_sweep,
           width = 9, height = 4.6, dpi = 150)
    print(p_sweep)
  }
}
```

## Alternative-Label Evaluation (Protocol Amendment 2)

Variants D and E are evaluated with the identical metric suite. Because a
label change alters the event definition, the event counts and prevalences
differ from the Sepsis-3 variants and are reported alongside the metrics; the
comparison of interest is made in `09_variance_decomposition.Rmd`.

```{r show-alt-results}
alt_df <- if (!is.null(results_df))
            results_df[results_df$variant %in% names(ALT_LABEL_VARIANTS), ] else NULL

if (!is.null(alt_df) && nrow(alt_df) > 0) {
  alt_df$label_family <- vapply(alt_df$variant, variant_type, character(1))

  # This write used to be a plain overwrite, so a partial pass silently
  # truncated the file to whatever ran: V2_VARIANTS=F deleted D's and E's rows,
  # and four consumers read it (09, 12 and thesis_constants.R twice), so the
  # alt-label table, family E3 and the phi_anchor macros all lost two variants
  # with no error anywhere. Merge instead: rows for the variants in THIS run
  # replace their previous selves, every other variant is carried forward.
  alt_prev <- tryCatch({
    p <- file.path(DATA_DIR, "07_metric_results_altlabel.parquet")
    if (file.exists(p)) as.data.frame(arrow::read_parquet(p)) else NULL
  }, error = function(e) NULL)
  if (!is.null(alt_prev) && nrow(alt_prev) > 0 &&
      identical(sort(names(alt_prev)), sort(names(alt_df)))) {
    carried <- alt_prev[!(alt_prev$variant %in% unique(alt_df$variant)), , drop = FALSE]
    if (nrow(carried) > 0) {
      cat(sprintf("Carrying forward %d alt-label rows for variants not in this run: %s\n",
                  nrow(carried), paste(sort(unique(carried$variant)), collapse = ", ")))
      alt_df <- rbind(alt_df, carried[, names(alt_df), drop = FALSE])
    }
  }
  alt_df <- alt_df[order(alt_df$variant, alt_df$eval_window, alt_df$model), , drop = FALSE]

  save_parquet(alt_df, "07_metric_results_altlabel", DATA_DIR)
  save_csv(alt_df,     "07_metric_results_altlabel", DATA_DIR)
  knitr::kable(
    alt_df[order(alt_df$variant, alt_df$eval_window, alt_df$model),
           c(display_cols, "n_sepsis_events")],
    caption = "Evaluation results under the alternative (non-Sepsis-3) labels"
  )
  cat("\nSaved: 07_metric_results_altlabel.parquet\n")
} else {
  cat("No alternative-label variants selected in this run.\n")
}
```

## Feature-Ablation Evaluation (Protocol Amendment 5)

The ablated fits from stages 05 and 06 are scored with the identical metric
suite on the identical test rows, so the difference between the two arms is
attributable to the feature set alone. The ablation removes the action-derived
features --- vasopressor exposure and the per-test order indicators --- and
retains all physiological measurements, including forward-filled laboratory
values.

```{r evaluate-ablation}
step("--- Evaluating the ablated arm (Protocol Amendment 5) ---")

ablation_rows <- list()

for (vid in intersect(active_variants(), names(LABEL_VARIANTS))) {
  abl_paths <- c(primary = file.path(DATA_DIR,
                    paste0("05_predictions_primary_ablated_", vid, ".parquet")),
                 gbt     = file.path(DATA_DIR,
                    paste0("06_predictions_gbt_ablated_", vid, ".parquet")))
  abl_cols  <- c(primary = "pred_sepsis", gbt = "pred_gbt")

  for (mod_name in names(abl_paths)) {
    if (!file.exists(abl_paths[[mod_name]])) next
    d        <- as.data.table(arrow::read_parquet(abl_paths[[mod_name]]))
    pred_col <- abl_cols[[mod_name]]
    if (!pred_col %in% names(d)) next

    y     <- as.integer(d$event == 1)
    pred  <- as.numeric(d[[pred_col]])
    valid <- !is.na(pred) & !is.na(y)
    if (sum(valid) < 10) next

    roc_pr <- compute_roc_pr(y[valid], pred[valid])
    calib  <- tryCatch(compute_calibration(y[valid], pred[valid]),
                       error = function(e) list(intercept = NA, slope = NA))
    sweep  <- tryCatch(utility_curve(d[valid], pred_col,
                                     utility_threshold_grid(pred[valid])),
                       error = function(e) NULL)
    best   <- if (!is.null(sweep)) {
      cand <- sweep[n_alerts > 0 & is.finite(utility_normalised)]
      if (nrow(cand) > 0) cand[which.max(utility_normalised)] else NULL
    } else NULL

    ablation_rows[[paste(vid, mod_name, sep = "-")]] <- data.frame(
      variant         = vid,
      variant_name    = variant_name(vid),
      model           = mod_name,
      arm             = "ablated",
      eval_window     = "unrestricted",
      split_type      = "temporal",
      # The ablated arm drops ACTION_DERIVED_FEATURES from the declared
      # covariate vector; both counts come from config.R rather than being
      # asserted, so the "8 of 22" in the thesis is checkable against the file.
      n_features      = length(model_features()) - length(ACTION_DERIVED_FEATURES),
      n_features_full = length(model_features()),
      n_rows          = sum(valid),
      n_sepsis_events = sum(y[valid]),
      auroc           = round(roc_pr$auroc, 4),
      auprc           = round(roc_pr$auprc, 4),
      calib_slope     = round(calib$slope, 4),
      utility_max     = if (is.null(best)) NA_real_ else round(best$utility_normalised, 4)
    )
    rm(d); invisible(gc(FALSE))
  }
}

if (length(ablation_rows) > 0) {
  abl_res <- as.data.table(do.call(rbind, ablation_rows))

  # Paired against the full arm on the same variant, model and window. The
  # delta is the quantity of interest; the two AUROCs on their own are not.
  full_res <- as.data.table(arrow::read_parquet(
    file.path(DATA_DIR, "07_metric_results.parquet")))
  full_res <- full_res[eval_window == "unrestricted" & split_type == "temporal",
                       .(variant, model,
                         auroc_full = auroc, auprc_full = auprc,
                         calib_slope_full = calib_slope,
                         utility_max_full = utility_max)]

  abl_res <- merge(abl_res, full_res, by = c("variant", "model"), all.x = TRUE)
  abl_res[, `:=`(
    delta_auroc       = round(auroc       - auroc_full,       4),
    delta_auprc       = round(auprc       - auprc_full,       4),
    delta_utility_max = round(utility_max - utility_max_full, 4)
  )]
  setorder(abl_res, variant, model)

  save_parquet(as.data.frame(abl_res), "07_ablation_metrics", DATA_DIR)
  save_csv(as.data.frame(abl_res),     "07_ablation_metrics", DATA_DIR)
  knitr::kable(
    abl_res[, .(variant, model, n_sepsis_events,
                auroc_full, auroc, delta_auroc,
                utility_max_full, utility_max, delta_utility_max)],
    caption = "Protocol Amendment 5: performance with and without the action-derived features"
  )
  cat("\nSaved: 07_ablation_metrics.parquet / .csv\n")
} else {
  cat("No ablated predictions found; the ablation arm has not been run.\n")
}
```

## Random-Split Evaluation (H4: Split Optimism)

The same metric suite is applied to random-split predictions (Variant B only). The output `07_metric_results_random.parquet` feeds the variance decomposition for H4 (Guo et al., 2022).

```{r evaluate-random-split}
step("--- Random-split evaluation for H4 ---")

random_results <- list()
vid <- "B"

prim_random_path <- file.path(DATA_DIR, paste0("05_predictions_primary_random_", vid, ".parquet"))
comp_random_path <- file.path(DATA_DIR, paste0("06_predictions_comparators_random_", vid, ".parquet"))

if (file.exists(prim_random_path)) {
  ph_prim_r <- as.data.table(arrow::read_parquet(prim_random_path))
  ph_comp_r <- if (file.exists(comp_random_path))
                 as.data.table(arrow::read_parquet(comp_random_path)) else NULL

  ph_all_r <- ph_prim_r
  if (!is.null(ph_comp_r)) {
    merge_cols_r <- c("stay_id", "hour", setdiff(names(ph_comp_r), names(ph_prim_r)))
    ph_all_r <- merge(ph_all_r, ph_comp_r[, ..merge_cols_r],
                      by = c("stay_id", "hour"), all.x = TRUE)
  }

  eval_windows_r <- list(
    unrestricted = ph_all_r,
    treatment_anchored = {
      if ("first_abx_hour" %in% names(ph_all_r)) {
        anchor_h_r <- if ("first_culture_hour" %in% names(ph_all_r))
                        pmin(ph_all_r$first_abx_hour, ph_all_r$first_culture_hour, na.rm = TRUE)
                      else ph_all_r$first_abx_hour
        ph_all_r[is.na(anchor_h_r) | hour < anchor_h_r]
      } else ph_all_r
    }
  )

  for (win_name in names(eval_windows_r)) {
    data_win_r <- eval_windows_r[[win_name]]
    if (nrow(data_win_r) == 0) next

    for (mod_name in names(MODEL_NAMES)) {
      pred_col <- MODEL_NAMES[[mod_name]]
      if (!pred_col %in% names(data_win_r)) next
      valid <- !is.na(data_win_r[[pred_col]])
      if (sum(valid) < 100) next

      y    <- as.integer(data_win_r$event == 1)
      pred <- as.numeric(data_win_r[[pred_col]])

      roc_pr  <- compute_roc_pr(y[valid], pred[valid])
      calib   <- compute_calibration(y[valid], pred[valid])
      alerts  <- as.integer(pred >= 0.5)
      burden  <- compute_alert_burden(y[valid], alerts[valid])
      utility <- tryCatch(
        compute_utility_score(data_win_r[valid], pred_col, threshold = 0.5),
        error = function(e) list(utility_normalised = NA, utility_total = NA)
      )
      lead_t  <- tryCatch(
        compute_lead_time(data_win_r[valid], pred_col),
        error = function(e) NA_real_
      )

      sweep_r <- tryCatch(
        utility_curve(data_win_r[valid], pred_col,
                      utility_threshold_grid(pred[valid])),
        error = function(e) NULL
      )
      best_r <- if (!is.null(sweep_r)) {
        cand_r <- sweep_r[n_alerts > 0 & is.finite(utility_normalised)]
        if (nrow(cand_r) > 0) cand_r[which.max(utility_normalised)] else NULL
      } else NULL

      result_row <- data.frame(
        variant            = vid,
        variant_name       = LABEL_VARIANTS[[vid]]$name,
        model              = mod_name,
        eval_window        = win_name,
        split_type         = "random",
        n_rows             = sum(valid),
        n_sepsis_events    = sum(y[valid]),
        auroc              = round(roc_pr$auroc, 4),
        auprc              = round(roc_pr$auprc, 4),
        calib_intercept    = round(calib$intercept, 4),
        calib_slope        = round(calib$slope, 4),
        utility_normalised = round(utility$utility_normalised, 4),
        alert_burden       = round(burden, 2),
        lead_time_h        = round(lead_t, 2),
        utility_max        = if (is.null(best_r)) NA_real_ else round(best_r$utility_normalised, 4),
        utility_opt_threshold = if (is.null(best_r)) NA_real_ else signif(best_r$threshold, 4),
        utility_opt_alert_rate = if (is.null(best_r)) NA_real_ else signif(best_r$alert_rate, 4),
        utility_opt_burden  = if (is.null(best_r)) NA_real_ else round(best_r$alert_burden, 2),
        utility_opt_detection = if (is.null(best_r)) NA_real_ else round(best_r$detection_rate, 4)
      )
      random_results[[paste(vid, mod_name, win_name, "random", sep = "-")]] <- result_row
    }
  }

  if (length(random_results) > 0) {
    random_results_df <- do.call(rbind, random_results)
    save_parquet(random_results_df, "07_metric_results_random", DATA_DIR)

    knitr::kable(
      random_results_df[order(random_results_df$eval_window, random_results_df$model),
                        c("variant", "model", "eval_window", "auroc", "auprc",
                          "utility_normalised", "utility_max", "alert_burden",
                          "lead_time_h")],
      caption = "Evaluation results: random split (Variant B only, H4)"
    )
    cat("\nSaved: 07_metric_results_random.parquet\n")
  }
} else {
  cat("Random-split primary predictions not found; skipping H4 evaluation.\n")
  cat("Run 05_primary_model.Rmd first to generate random-split predictions.\n")
}
```
