---
title: "08 External Validation - eICU-CRD External Validation"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook applies the models trained on MIMIC-IV, without retraining, to
an independent database: the eICU Collaborative Research Database (eICU-CRD;
Pollard et al., 2018). Validation in data from other hospitals is a central
step in judging whether a prediction model generalises (Collins et al., 2024,
TRIPOD+AI). MIMIC-IV comes from a single hospital, whereas eICU-CRD holds data
from many ICUs across the United States, whose patients, clinical practice and
data recording differ from MIMIC-IV's.

The pre-registered reference for the decrease in AUROC from internal to
external validation is about 0.085, the degradation reported by Moor et al.
(2023). It is a benchmark for comparison, not a null value or a pass mark. H5
was pre-registered as a comparison of proportional declines: that the Utility
Score falls proportionally more than AUROC on external data. That contrast
cannot be formed, because the internal Utility Score under the primary label is
close to zero. Under Protocol Amendment 1, the AUROC part of H5 takes its place
in the confirmatory family. That part is the internal minus the external AUROC
of the primary model under Variant B. This notebook reports it descriptively,
and `12_inference.Rmd` computes its interval and adjusted p-value. The
substitute does not establish the divergence between metrics that H5
originally described.

### Two External Analyses and Their Purposes

Sepsis-3 onset requires suspicion of infection, which the labels define as an
antibiotic order and a microbiology culture within a set window of each other.
In eICU-CRD, culture records exist for only a small minority of ICU stays, and
a stay without one cannot meet the suspicion criterion, whatever the patient's
condition. Applied to the full cohort, the definition would therefore mostly
measure how many stays have microbiology records, not how often sepsis occurs.
Two external analyses, called arms below and in the printed output, keep these
questions apart:

| Arm | Cohort | Suspicion anchor | Role |
|---|---|---|---|
| **Sepsis-3 (primary)** | stays with at least one culture record | an antibiotic and a culture within each variant's windows | Applies the MIMIC-IV label definitions, within the variables eICU-CRD records. Its denominator holds only stays in which an onset could be identified. |
| **Antibiotic-only (sensitivity)** | full cohort | the first systemic antibiotic, with no culture requirement | Shows how the event count changes when the culture requirement is dropped; the SOFA criterion still applies. Its target is not Sepsis-3, and it is never pooled with the primary arm. |

A third arm, in the final section, applies the treatment-independent Variant D
label to the full cohort. Variant D was introduced by Protocol Amendment 2, and
its external evaluation was added by Protocol Amendment 9. This arm is neither
a Sepsis-3 estimate nor a sensitivity analysis of one. It has a different
target, it is post-hoc, and it is excluded from H5.

The steps are:

1. Build an eICU-CRD person-hour panel with the MIMIC-IV variable names and
   preprocessing.
2. Derive onsets separately for each label variant and each anchor, with the
   SOFA-rise criterion of `02_cohort_labels.Rmd`.
3. Apply the models trained on MIMIC-IV, without retraining.
4. Compute AUROC and AUPRC for each model. `11_clinical_metrics.Rmd` adds the
   clinical metrics for the external cohort.
5. Compare the eICU-CRD and MIMIC-IV event rates.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(nnet)
  library(splines)
  library(xgboost)
  library(PRROC)
  library(dplyr)
})

source("config.R")
source("utils.R")
source("clinical_scores.R")

step("=== Phase 5: External Validation on eICU-CRD ===")
cat(sprintf("Pre-registered expected degradation: %.3f AUC\n",
            EXPECTED_EXTERNAL_DEGRADATION))
```

## Load eICU Patient Table

The age and length-of-stay criteria are those of MIMIC-IV: age at least
`r MIN_AGE` years (ages recorded as "> 89" are set to 90) and an ICU stay of at
least `r MIN_ICU_LOS_HOURS` hours. Each row of the eICU-CRD patient table is
one ICU unit stay. Unlike the MIMIC-IV cohort, which keeps only each patient's
first ICU stay, this cohort keeps every eligible stay, so a patient can
contribute more than one.

```{r load-eicu-patients}
step("Loading eICU tables...")

eicu_patient <- data.table::fread(file.path(EICU_DIR, "patient.csv.gz"),
                                   showProgress = FALSE)
cat(sprintf("eICU patients (raw): %s\n", format(nrow(eicu_patient), big.mark = ",")))

eicu_patient[, age_num := suppressWarnings(as.numeric(age))]
eicu_patient[age == ">89", age_num := 90]
eicu_patient[, los_hours := unitdischargeoffset / 60]
eicu_patient <- eicu_patient[!is.na(age_num) & age_num >= MIN_AGE &
                               los_hours >= MIN_ICU_LOS_HOURS]
cat(sprintf("eICU cohort after exclusions: %s\n",
            format(nrow(eicu_patient), big.mark = ",")))
```

## Vital Signs (vitalPeriodic)

Heart rate, respiratory rate, oxygen saturation, systolic and mean arterial
pressure, and temperature are read from `vitalPeriodic`. Because the table is
large, it is read through DuckDB, which keeps only cohort stays and the first
`r MAX_PANEL_HOURS` hours of each stay as the file is read. Readings are
averaged within each hour.

```{r load-vitals}
step("Loading eICU vitals...")

vp_path <- file.path(EICU_DIR, "vitalPeriodic.csv.gz")
if (file.exists(vp_path)) {
  suppressPackageStartupMessages(library(duckdb))
  .vp_con <- dbConnect(duckdb(), dbdir = ":memory:")
  dbWriteTable(.vp_con, "cohort_stays",
               data.frame(patientunitstayid = as.integer(eicu_patient$patientunitstayid)))
  sql_vp <- sprintf(
    "SELECT t.patientunitstayid, t.observationoffset,
            t.heartrate, t.respiration, t.sao2,
            t.systemicsystolic, t.systemicmean, t.temperature
     FROM read_csv('%s', header=true, compression='gzip') t
     INNER JOIN cohort_stays c ON t.patientunitstayid = c.patientunitstayid
     WHERE t.observationoffset >= 0 AND t.observationoffset <= %d",
    vp_path, MAX_PANEL_HOURS * 60L
  )
  v2_log("DuckDB: loading vitalPeriodic filtered to cohort + 72h window...")
  vp <- as.data.table(dbGetQuery(.vp_con, sql_vp))
  dbDisconnect(.vp_con, shutdown = TRUE)
  cat(sprintf("vitalPeriodic (cohort-filtered): %s rows\n", format(nrow(vp), big.mark = ",")))

  setnames(vp, c("patientunitstayid","observationoffset","heartrate","respiration",
                  "sao2","systemicsystolic","systemicmean","temperature"),
               c("stay_id","obs_offset_m","hr","resp_rate","spo2","sbp","map","temp_c"))
  vp[, obs_hour := floor(obs_offset_m / 60)]
  vp[!is.na(temp_c) & (temp_c < 30 | temp_c > 45), temp_c := NA]
  vp_hr <- vp[, .(hr        = mean(hr,        na.rm = TRUE),
                   resp_rate = mean(resp_rate, na.rm = TRUE),
                   spo2      = mean(spo2,      na.rm = TRUE),
                   sbp       = mean(sbp,       na.rm = TRUE),
                   map       = mean(map,       na.rm = TRUE),
                   temp_c    = mean(temp_c,    na.rm = TRUE)),
               by = .(stay_id, obs_hour)]
  rm(vp); gc()
} else {
  vp_hr <- data.table()
  v2_log("vitalPeriodic not found", level = "WARN")
}
```

## Lab Values

Laboratory results are renamed from their eICU-CRD names to the MIMIC-IV
variable names and averaged within each hour. The PaO$_2$/FiO$_2$ ratio is
calculated in hours in which both components are recorded.

```{r load-labs}
step("Loading eICU labs...")

lab_path <- file.path(EICU_DIR, "lab.csv.gz")
if (file.exists(lab_path)) {
  eicu_lab <- data.table::fread(lab_path, showProgress = FALSE,
                                 select = c("patientunitstayid", "labresultoffset",
                                            "labname", "labresult"))
  setnames(eicu_lab, c("patientunitstayid","labresultoffset","labname","labresult"),
                     c("stay_id","obs_offset_m","labname","value"))
  eicu_lab[, obs_hour := floor(obs_offset_m / 60)]
  eicu_lab <- eicu_lab[obs_hour >= 0 & obs_hour <= MAX_PANEL_HOURS & !is.na(value)]
  eicu_lab[, value := as.numeric(value)]

  lab_map <- c("lactate" = "lactate", "creatinine" = "creatinine",
               "total bilirubin" = "bilirubin_total", "platelets x 1000" = "platelets",
               "WBC x 1000" = "wbc", "paO2" = "pao2", "FiO2" = "fio2_lab",
               "sodium" = "sodium", "potassium" = "potassium")
  eicu_lab <- eicu_lab[labname %in% names(lab_map)]
  eicu_lab[, var_name := lab_map[labname]]

  lab_hr <- dcast(
    eicu_lab[, .(value = mean(value, na.rm = TRUE)),
              by = .(stay_id, obs_hour, var_name)],
    stay_id + obs_hour ~ var_name, value.var = "value", fun.aggregate = mean
  )
  if ("pao2" %in% names(lab_hr) && "fio2_lab" %in% names(lab_hr)) {
    lab_hr[, fio2_frac := ifelse(fio2_lab > 1, fio2_lab / 100, fio2_lab)]
    lab_hr[, pf_ratio  := ifelse(!is.na(pao2) & !is.na(fio2_frac) & fio2_frac > 0,
                                  pao2 / fio2_frac, NA_real_)]
    # fio2 is retained (as a fraction) because NEWS2 needs the supplemental-oxygen
    # flag; without it every eICU person-hour scores 0 on that parameter.
    lab_hr[, fio2 := fio2_frac]
    lab_hr[, c("fio2_lab", "fio2_frac") := NULL]
  }
} else {
  lab_hr <- data.table()
  v2_log("lab.csv not found", level = "WARN")
}
```

## Drug Exposure: Antibiotics and Vasopressors

eICU-CRD records medication in two tables. `medication.csv.gz` holds
medication orders, and `infusionDrug.csv.gz` holds continuous infusions.
Antibiotic exposure combines the records of both tables. Almost all antibiotic
records come from the order table, and the infusion table adds few further
stays, as the printed counts show. Vasopressor exposure comes from the infusion
table alone. That table records an infusion rate at a point in time, with no
stop time, so exposure is marked in each hour that has an infusion record. In
MIMIC-IV, exposure is marked over the full interval from an infusion's start
to its end, so the two definitions are close but not identical.

```{r eicu-drugs}
step("Loading eICU drug exposure (medication + infusionDrug)...")

med_path <- file.path(EICU_DIR, "medication.csv.gz")
inf_path <- file.path(EICU_DIR, "infusionDrug.csv.gz")

abx_parts    <- list()
vaso_hours   <- data.table()
norepi_hours <- data.table()

if (file.exists(med_path)) {
  meds <- data.table::fread(med_path, showProgress = FALSE,
                              select = c("patientunitstayid","drugstartoffset",
                                         "drugname","drugordercancelled"))
  meds <- meds[drugordercancelled != "Yes" & !is.na(drugstartoffset)]
  meds[, is_abx := match_antibiotics(drugname, ANTIBIOTICS)]
  abx_parts$medication <- meds[is_abx == TRUE,
                                .(stay_id  = patientunitstayid,
                                  abx_hour = floor(drugstartoffset / 60))]
  cat(sprintf("Antibiotic orders (medication): %s rows, %s stays\n",
              format(nrow(abx_parts$medication), big.mark = ","),
              format(uniqueN(abx_parts$medication$stay_id), big.mark = ",")))
  rm(meds); gc()
} else {
  v2_log("medication.csv.gz not found", level = "WARN")
}

if (file.exists(inf_path)) {
  infus <- data.table::fread(inf_path, showProgress = FALSE,
                               select = c("patientunitstayid","infusionoffset",
                                          "drugname"))
  infus <- infus[!is.na(infusionoffset)]
  infus[, inf_hour := floor(infusionoffset / 60)]

  infus[, is_abx := match_antibiotics(drugname, ANTIBIOTICS)]
  abx_parts$infusion <- infus[is_abx == TRUE,
                               .(stay_id = patientunitstayid, abx_hour = inf_hour)]
  cat(sprintf("Antibiotic infusions (infusionDrug): %s rows, %s stays\n",
              format(nrow(abx_parts$infusion), big.mark = ","),
              format(uniqueN(abx_parts$infusion$stay_id), big.mark = ",")))

  # Vasopressor exposure, restricted to the panel horizon. An infusion row is a
  # rate observation at one instant, so exposure is flagged on the hour of the
  # observation rather than over an interval (eICU carries no infusion stop
  # time). This is the closest available analogue to the MIMIC-IV
  # `inputevents` start/end expansion in 03_person_hours.Rmd.
  infus[, is_vaso   := match_antibiotics(drugname, EICU_VASOPRESSORS)]
  infus[, is_norepi := match_antibiotics(drugname, EICU_NOREPI_EPI)]
  vaso_hours <- unique(infus[is_vaso == TRUE & inf_hour >= 0 &
                               inf_hour <= MAX_PANEL_HOURS,
                             .(stay_id = patientunitstayid, obs_hour = inf_hour)])
  vaso_hours[, vaso_any := TRUE]
  norepi_hours <- unique(infus[is_norepi == TRUE & inf_hour >= 0 &
                                 inf_hour <= MAX_PANEL_HOURS,
                               .(stay_id = patientunitstayid, obs_hour = inf_hour)])
  norepi_hours[, norepi_epi_any := TRUE]
  cat(sprintf("Vasopressor infusion hours: %s (%s stays); norepi/epi: %s (%s stays)\n",
              format(nrow(vaso_hours), big.mark = ","),
              format(uniqueN(vaso_hours$stay_id), big.mark = ","),
              format(nrow(norepi_hours), big.mark = ","),
              format(uniqueN(norepi_hours$stay_id), big.mark = ",")))
  rm(infus); gc()
} else {
  v2_log("infusionDrug.csv.gz not found", level = "WARN")
}

eicu_abx <- if (length(abx_parts) > 0) unique(rbindlist(abx_parts)) else data.table()
cat(sprintf("Antibiotic exposure (union): %s rows, %s stays\n",
            format(nrow(eicu_abx), big.mark = ","),
            format(uniqueN(eicu_abx$stay_id), big.mark = ",")))
```

## Microbiology Coverage

Microbiology records, held in `microLab.csv.gz`, limit how far the Sepsis-3
label can be carried to eICU-CRD. Cultures whose site appears in the
surveillance list used for MIMIC-IV (`SURVEILLANCE_SPEC_TYPES`) are excluded. A stay with no culture record cannot meet the suspicion criterion
under any variant. Counting its person-hours in the denominator would
therefore reflect the coverage of the microbiology records rather than the
incidence of sepsis. The stays with at least one culture record form the
cohort of the primary arm. Culture times may fall before ICU admission, and
the last printed line gives the share that do. Such a culture can still form a
suspicion pair, but an onset assigned before admission is not counted (see the
next sections).

```{r eicu-cultures}
step("Loading eICU microbiology...")

micro_path   <- file.path(EICU_DIR, "microLab.csv.gz")
eicu_culture <- data.table()

if (file.exists(micro_path)) {
  micro <- data.table::fread(micro_path, showProgress = FALSE,
                               select = c("patientunitstayid","culturetakenoffset",
                                          "culturesite"))
  micro <- micro[!culturesite %in% SURVEILLANCE_SPEC_TYPES]
  eicu_culture <- micro[!is.na(culturetakenoffset),
                         .(stay_id      = patientunitstayid,
                           culture_hour = floor(culturetakenoffset / 60))]
} else {
  v2_log("microLab.csv.gz not found", level = "WARN")
}

cohort_ids  <- unique(as.integer(eicu_patient$patientunitstayid))
micro_stays <- intersect(unique(eicu_culture$stay_id), cohort_ids)

cat(sprintf("Culture records: %s over %s stays\n",
            format(nrow(eicu_culture), big.mark = ","),
            format(uniqueN(eicu_culture$stay_id), big.mark = ",")))
cat(sprintf("Microbiology coverage within the analysis cohort: %s / %s stays (%.2f%%)\n",
            format(length(micro_stays), big.mark = ","),
            format(length(cohort_ids),  big.mark = ","),
            100 * length(micro_stays) / max(length(cohort_ids), 1)))
cat(sprintf("Culture records drawn before ICU admission (negative offset): %.1f%%\n",
            100 * mean(eicu_culture$culture_hour < 0)))
```

## Build eICU Person-Hour Panel

A single person-hour panel is built from the physiological data and carries no
label. The onsets of each variant are derived from it in the next section, so
each variant's label is computed separately. Construction follows the
MIMIC-IV steps of `03_person_hours.Rmd`. These are an hourly grid up to hour
`r MAX_PANEL_HOURS`, the plausibility filter applied before filling, last
observation carried forward within the stay, the `*_measured` indicators, the
six-hour means and the NEWS2 and qSOFA scores.

The Glasgow Coma Scale (GCS) is not recorded in `vitalPeriodic`, so the
panel's `gcs` column is missing in every hour. Three later steps handle it
differently:

- NEWS2 scores its consciousness item as 0 throughout.
- The SOFA calculations score the neurological component as normal.
- When the models are applied, the missing value is set to zero like any other
  (see "Apply MIMIC-IV Trained Models").

```{r build-eicu-panel}
step("Building eICU person-hour panel...")

eicu_stays <- eicu_patient[, .(stay_id = as.integer(patientunitstayid), los_hours,
                                age = age_num, gender, ethnicity,
                                hospital_expire = unitdischargestatus == "Expired")]
panel <- build_person_hour_grid(eicu_stays, max_hours = MAX_PANEL_HOURS)

if (nrow(vp_hr) > 0) {
  if ("obs_hour" %in% names(vp_hr)) setnames(vp_hr, "obs_hour", "hour")
  panel <- merge(panel, vp_hr, by = c("stay_id", "hour"), all.x = TRUE)
}
if (nrow(lab_hr) > 0) {
  if (!"hour" %in% names(lab_hr) && "obs_hour" %in% names(lab_hr))
    setnames(lab_hr, "obs_hour", "hour")
  panel <- merge(panel, lab_hr, by = c("stay_id", "hour"), all.x = TRUE)
}

if (nrow(vaso_hours) > 0) {
  setnames(vaso_hours, "obs_hour", "hour")
  panel <- merge(panel, vaso_hours, by = c("stay_id", "hour"), all.x = TRUE)
  panel[is.na(vaso_any), vaso_any := FALSE]
} else { panel[, vaso_any := FALSE] }

if (nrow(norepi_hours) > 0) {
  setnames(norepi_hours, "obs_hour", "hour")
  panel <- merge(panel, norepi_hours, by = c("stay_id", "hour"), all.x = TRUE)
  panel[is.na(norepi_epi_any), norepi_epi_any := FALSE]
} else { panel[, norepi_epi_any := FALSE] }

# Same plausibility filter as the MIMIC panel in 03, applied before the
# forward-fill, so the external cohort is pre-processed identically to the
# training cohort rather than on its own terms.
apply_plausibility(panel, label = "eICU panel")

fill_cols <- intersect(c("hr","resp_rate","spo2","map","sbp","temp_c","fio2",
                          "lactate","creatinine","bilirubin_total","platelets",
                          "wbc","pf_ratio"), names(panel))
panel <- ffill_by_stay(panel, fill_cols)

for (mc in c("lactate","creatinine","bilirubin_total","platelets","wbc","pf_ratio")) {
  if (mc %in% names(panel))
    panel[[paste0(mc, "_measured")]] <- as.integer(!is.na(panel[[mc]]))
}
for (col in intersect(c("hr","resp_rate","map","temp_c"), names(panel)))
  panel <- rolling_mean_by_stay(panel, col, n_h = SLOPE_WINDOW_H,
                                 col_out = paste0(col, "_slope6h"))

if (!"gcs" %in% names(panel))       panel[, gcs := NA_real_]
if (!"fio2" %in% names(panel))      panel[, fio2 := NA_real_]
if (!"bilirubin" %in% names(panel) && "bilirubin_total" %in% names(panel))
  panel[, bilirubin := bilirubin_total]

# NEWS2 and qSOFA. `avpu_alert` must be NA, not FALSE, where GCS is unknown:
# news2() scores NA as 0 on the consciousness parameter but scores FALSE as 3,
# so passing FALSE for an unmeasured GCS adds a constant +3 to every eICU
# person-hour and makes every cross-cohort threshold comparison meaningless.
# GCS is absent from vitalPeriodic, so that is every row of this panel.
suppl_o2   <- { f <- panel$fio2; ifelse(is.na(f), NA, f > 0.21) }
avpu_alert <- avpu_alert_flag(panel$gcs)
news2_scores <- news2(panel$resp_rate, panel$spo2, suppl_o2, panel$sbp, panel$hr,
                       panel$temp_c, avpu_alert)
panel[, names(news2_scores) := news2_scores]
panel[, qsofa_score := qsofa(resp_rate, gcs, map)]

cat(sprintf("eICU panel: %s person-hours over %s stays\n",
            format(nrow(panel), big.mark = ","),
            format(uniqueN(panel$stay_id), big.mark = ",")))
```

## Per-Variant Onset Derivation

Each label variant is applied to eICU-CRD with its own antibiotic--culture
windows, SOFA evaluation window and SOFA baseline, as `02_cohort_labels.Rmd`
applies them to MIMIC-IV. For each qualifying antibiotic--culture pair, the
later of the two times is taken, and the suspicion time is the earliest of
these. The onset is placed at the
suspicion time if the SOFA score over the evaluation window is at least
`r SOFA_INCREASE_THRESHOLD` points above the baseline. Each SOFA component is
averaged over the window before scoring, as in MIMIC-IV. Under the
antibiotic-only anchor, the suspicion time is the first antibiotic, and the
same SOFA criterion applies.

The organ-dysfunction criterion is a rise over a baseline, not an absolute SOFA
score. `score_sofa()` scores a missing component as normal, so an absolute
threshold would be harder to reach in stays with sparse data simply because
fewer components are measured. The baseline is the admission window (hours
0--`r DETERIORATION_BASELINE_H`) for Variants A and B. For Variant C, it is the
24 hours before the start of the evaluation window. This follows each
variant's baseline definition in `config.R`.

An onset time that falls before ICU admission, after the end of the stay, or
at or before hour `r MIN_OBS_HOURS` is not counted. The stay then has no
onset in the panel but remains in the at-risk denominator; it is not removed.
This follows the MIMIC-IV rule. In the label tables a missing onset hour
(`NA`) therefore means that no onset was assigned in the panel, not that the
outcome is unknown.

Table 1 gives, for each variant and anchor, the number of stays in the arm's
cohort (`n_stays`) and the number with a qualifying suspicion pair
(`n_suspicion`). It also gives the number assigned an onset (`n_onsets`), that
number as a percentage of the cohort's stays (`pct_onsets`), and the median
onset hour.

```{r eicu-onset-functions}
EICU_SOFA_COMPONENTS <- c("pf_ratio", "platelets", "bilirubin", "map",
                          "gcs", "creatinine")

#' SOFA scored once over a per-stay hour window, mirroring the window-mean
#' aggregation `compute_sofa_delta_vectorized()` performs on MIMIC-IV.
#' @param wins data.table(stay_id, h_lo, h_hi)
sofa_over_window <- function(pan, wins) {
  out <- data.table(stay_id = wins$stay_id, sofa = 0L)
  if (nrow(wins) == 0) return(out)
  comps <- intersect(EICU_SOFA_COMPONENTS, names(pan))
  keep  <- c("stay_id", "hour", comps,
             intersect(c("vaso_any", "norepi_epi_any"), names(pan)))
  d <- merge(pan[, ..keep], wins, by = "stay_id")
  d <- d[hour >= h_lo & hour <= h_hi]
  if (nrow(d) == 0) return(out)

  agg <- d[, lapply(.SD, mean, na.rm = TRUE), by = stay_id, .SDcols = comps]
  for (vc in intersect(c("vaso_any", "norepi_epi_any"), names(d))) {
    vb <- d[, .(x = any(get(vc) %in% TRUE)), by = stay_id]
    setnames(vb, "x", vc)
    agg <- merge(agg, vb, by = "stay_id", all.x = TRUE)
  }
  agg[, sofa := score_sofa(as.data.frame(agg))$sofa_total]
  out <- merge(out[, .(stay_id)], agg[, .(stay_id, sofa)], by = "stay_id", all.x = TRUE)
  out[is.na(sofa), sofa := 0L]
  out[]
}

#' Suspicion time under one variant and one anchor mode.
#' @param anchor "sepsis3" (culture AND antibiotic) or "abx_only"
derive_suspicion <- function(vid, anchor) {
  v <- variant_def(vid)
  if (anchor == "abx_only") {
    if (nrow(eicu_abx) == 0) return(data.table(stay_id = integer(0), t_susp_h = integer(0)))
    return(eicu_abx[, .(t_susp_h = min(abx_hour)), by = stay_id])
  }
  if (nrow(eicu_abx) == 0 || nrow(eicu_culture) == 0)
    return(data.table(stay_id = integer(0), t_susp_h = integer(0)))

  pairs <- merge(eicu_abx, eicu_culture, by = "stay_id", allow.cartesian = TRUE)
  pairs[, dt_abx_before_cul := culture_hour - abx_hour]
  pairs[, dt_cul_before_abx := abx_hour - culture_hour]
  pairs <- pairs[
    (dt_abx_before_cul >= 0 & dt_abx_before_cul <= v$abx_before_culture_h) |
    (dt_cul_before_abx >= 0 & dt_cul_before_abx <= v$culture_before_abx_h)
  ]
  if (nrow(pairs) == 0) return(data.table(stay_id = integer(0), t_susp_h = integer(0)))
  pairs[, t_susp := pmax(abx_hour, culture_hour)]
  pairs[, .(t_susp_h = min(t_susp)), by = stay_id]
}

#' Onsets under one variant and one anchor: suspicion, gated by a SOFA rise of
#' >= SOFA_INCREASE_THRESHOLD inside the variant's evaluation window.
derive_eicu_onsets <- function(pan, stay_meta, vid, anchor) {
  v    <- variant_def(vid)
  susp <- derive_suspicion(vid, anchor)
  susp <- susp[stay_id %in% stay_meta$stay_id]
  if (nrow(susp) == 0)
    return(data.table(stay_id = integer(0), t_susp_h = integer(0),
                      onset_hour = integer(0)))

  eval_wins <- susp[, .(stay_id,
                        h_lo = t_susp_h - v$sofa_window_before_h,
                        h_hi = t_susp_h + v$sofa_window_after_h)]
  base_wins <- if (identical(v$sofa_baseline, "rolling_24h")) {
    susp[, .(stay_id,
             h_lo = t_susp_h - v$sofa_window_before_h - 24L,
             h_hi = t_susp_h - v$sofa_window_before_h)]
  } else {
    susp[, .(stay_id, h_lo = 0L, h_hi = DETERIORATION_BASELINE_H)]
  }

  ev <- sofa_over_window(pan, eval_wins); setnames(ev, "sofa", "sofa_eval")
  bs <- sofa_over_window(pan, base_wins); setnames(bs, "sofa", "sofa_base")

  out <- merge(merge(susp, ev, by = "stay_id", all.x = TRUE),
               bs, by = "stay_id", all.x = TRUE)
  out[is.na(sofa_eval), sofa_eval := 0L]
  out[is.na(sofa_base), sofa_base := 0L]
  out[, onset_hour := fifelse((sofa_eval - sofa_base) >= SOFA_INCREASE_THRESHOLD,
                              as.integer(t_susp_h), NA_integer_)]

  # Onset must fall inside the stay and after the minimum observation period.
  # Out-of-range onsets make the stay non-septic; they never delete it.
  out <- merge(out, stay_meta[, .(stay_id, los_hours)], by = "stay_id", all.x = TRUE)
  out[!is.na(onset_hour) &
        (onset_hour < 0 | onset_hour > floor(los_hours)), onset_hour := NA_integer_]
  out[!is.na(onset_hour) & onset_hour <= MIN_OBS_HOURS,   onset_hour := NA_integer_]
  out[, .(stay_id, t_susp_h, onset_hour)]
}

#' Attach a variant's labels to the panel and apply competing-risk coding.
attach_labels <- function(pan, onsets, stay_meta) {
  g <- merge(pan, onsets[, .(stay_id, onset_hour)], by = "stay_id", all.x = TRUE)
  g <- merge(g, stay_meta[, .(stay_id, hospital_expire)], by = "stay_id", all.x = TRUE)
  g[, max_h := max(hour), by = stay_id]
  # An onset beyond the 72 h panel horizon is right-censoring, not an event.
  g[, in_panel := !is.na(onset_hour) & onset_hour <= max_h]
  g[, event := 0L]
  g[in_panel == TRUE & hour == onset_hour,                       event := 1L]
  g[in_panel == FALSE & hour == max_h & hospital_expire == TRUE,  event := 3L]
  g[in_panel == FALSE & hour == max_h & hospital_expire == FALSE, event := 2L]
  g[is.na(onset_hour) | hour <= onset_hour]
}
```

```{r derive-eicu-labels}
step("Deriving eICU onsets per label variant...")

panel_primary <- panel[stay_id %in% micro_stays]
stays_primary <- eicu_stays[stay_id %in% micro_stays]

eicu_labels <- list()
label_summary <- list()

for (vid in names(LABEL_VARIANTS)) {
  for (anchor in c("sepsis3", "abx_only")) {
    pan <- if (anchor == "sepsis3") panel_primary else panel
    stm <- if (anchor == "sepsis3") stays_primary else eicu_stays
    ons <- derive_eicu_onsets(pan, stm, vid, anchor)
    eicu_labels[[paste(vid, anchor, sep = "-")]] <- ons

    n_onset <- ons[!is.na(onset_hour), .N]
    label_summary[[paste(vid, anchor, sep = "-")]] <- data.table(
      variant       = vid,
      variant_name  = variant_name(vid),
      anchor        = anchor,
      n_stays       = uniqueN(stm$stay_id),
      n_suspicion   = nrow(ons),
      n_onsets      = n_onset,
      pct_onsets    = round(100 * n_onset / max(uniqueN(stm$stay_id), 1), 2),
      median_onset_h = if (n_onset > 0)
        as.numeric(median(ons[!is.na(onset_hour), onset_hour])) else NA_real_
    )
  }
}

label_summary_dt <- rbindlist(label_summary)
knitr::kable(label_summary_dt,
             caption = "eICU onsets by label variant and anchor definition")
save_parquet(label_summary_dt, "08_eicu_label_summary", DATA_DIR)
save_csv(label_summary_dt,     "08_eicu_label_summary", DATA_DIR)
```

### Check That the Labels Differ by Variant

This check detects a label set that has been built once and reused for every
variant. With such reuse, the external variants would differ only in the model
applied, every label-dependent quantity would agree across A, B and C, and the
agreement could be mistaken for a finding. NEWS2 does not depend on the
variant, so identical NEWS2 AUROCs across variants would be one symptom of such
reuse. The check does not rely on that indirect sign. It compares the onset
sets directly, as the list of stays with their onset hours, and stops the
notebook if all three variants produce identical sets under either anchor. It
runs on every execution, and its result is written to the log.

```{r variant-invariance-guard}
onset_signature <- function(key) {
  o <- eicu_labels[[key]]
  paste(sort(o[!is.na(onset_hour), paste(stay_id, onset_hour, sep = ":")]),
        collapse = "|")
}

for (anchor in c("sepsis3", "abx_only")) {
  sigs <- vapply(names(LABEL_VARIANTS),
                 function(v) onset_signature(paste(v, anchor, sep = "-")),
                 character(1))
  if (length(unique(sigs)) == 1L)
    stop(sprintf(paste0("Variant-invariance guard failed for anchor '%s': ",
                        "variants %s produced byte-identical onset sets. ",
                        "The per-variant label logic is not being applied."),
                 anchor, paste(names(LABEL_VARIANTS), collapse = "/")))
  v2_log(sprintf("Variant-invariance guard passed for anchor '%s' (%d distinct onset sets)",
                 anchor, length(unique(sigs))))
}
```

## Apply MIMIC-IV Trained Models

The primary model of `05_primary_model.Rmd` and the GBT of
`06_comparators.Rmd`, trained on MIMIC-IV for the same variant, are applied
without retraining. NEWS2 is scored as internally. Missing covariate values are
set to zero, or to false for logical columns, as in training; the
`*_measured` indicators are built from the eICU-CRD data as in MIMIC-IV. The
primary model's predictions are computed in batches of rows to limit memory
use. The GBT's design matrix is built in the covariate order declared in
`config.R`, and the notebook stops if the stored model records a different
order.

One consequence of the zero-filling needs stating. GCS is missing in every
eICU-CRD hour, so every external person-hour enters both models with a GCS of
zero. In the MIMIC-IV training data, a zero occurs only in the hours before a
patient's GCS is first recorded. The models' external predictions therefore
rest on a GCS input that differs systematically from training. This is part of
the difference between the cohorts that the external estimates measure, and it
is not separated from the other differences here.

AUROC and AUPRC are computed for each model, variant and arm. Each of these
evaluations prints one covariate check, so the check appears six times, once
for each variant in each of the two arms.

Table 2 summarises how much of the cohort each labelling input reaches:

- the cohort's stays;
- the stays with a culture record, which form the primary arm;
- the stays with an antibiotic record;
- the stays with both kinds of record at any time, which is the most the
  Sepsis-3 suspicion criterion could reach with unlimited windows;
- the person-hours of the full and of the microbiology-covered panel.

Counts are numbers of stays, and the `pct_` columns are percentages of the
cohort's stays. The last column repeats the minimum number of external events,
`r MIN_EXTERNAL_EVENTS`, below which an external AUROC is reported but not
interpreted.

```{r apply-models}
step("Applying trained models to eICU cohort...")

# Declared once in config.R. This stage previously kept its own copy, which is
# the most dangerous of the three duplicates that existed: it is the vector the
# stored GBT model's column order is reconstructed from below, so a divergence
# in ORDER between this copy and stage 06's would have produced permuted
# external predictions that are still well-formed numbers.
MODEL_FEATURES <- model_features()

#' Evaluate every available model on one labelled eICU panel.
evaluate_panel <- function(g, vid, anchor) {
  prim_model_path <- file.path(DATA_DIR, paste0("05_model_primary_", vid, ".rds"))
  gbt_model_path  <- file.path(DATA_DIR, paste0("06_model_gbt_",     vid, ".rds"))
  if (!file.exists(prim_model_path)) return(NULL)

  eicu_pred   <- copy(g)
  require_features(MODEL_FEATURES, available = names(eicu_pred))
  for (f in intersect(MODEL_FEATURES, names(eicu_pred))) {
    if (is.numeric(eicu_pred[[f]])) eicu_pred[is.na(get(f)), (f) := 0]
    if (is.logical(eicu_pred[[f]])) eicu_pred[is.na(get(f)), (f) := FALSE]
  }

  prim_mod        <- readRDS(prim_model_path)
  PRED_CHUNK      <- 500000L
  n_ep            <- nrow(eicu_pred)
  pred_sepsis_vec <- numeric(n_ep)
  for (ch_start in seq(1L, n_ep, by = PRED_CHUNK)) {
    ch_idx   <- ch_start:min(ch_start + PRED_CHUNK - 1L, n_ep)
    chunk_df <- as.data.frame(eicu_pred[ch_idx])
    if (inherits(prim_mod, "multinom")) {
      pr <- predict(prim_mod, newdata = chunk_df, type = "probs")
      pred_sepsis_vec[ch_idx] <- if (is.matrix(pr)) pr[, "sepsis"] else as.numeric(pr)
    } else {
      pred_sepsis_vec[ch_idx] <- predict(prim_mod, newdata = chunk_df, type = "response")
    }
    rm(chunk_df)
  }
  eicu_pred[, pred_sepsis := pred_sepsis_vec]
  rm(pred_sepsis_vec, prim_mod); gc()

  if (file.exists(gbt_model_path)) {
    gbt_mod   <- readRDS(gbt_model_path)
    gbt_feats <- gbt_mod$feature_names
    # An xgb.Booster round-tripped through saveRDS returns with an EMPTY
    # feature_names under this xgboost build. lapply() over an empty vector then
    # yields a zero-column DMatrix whose predictions are all NA, and the metric
    # loop below discards them as "fewer than 50 valid rows" -- so the GBT row
    # vanishes from the external results with no error anywhere. That is how
    # eICU carried no GBT comparator for the whole project.
    #
    # Stage 06 trains on MODEL_FEATURES in exactly this order: it sets
    # gbt_feats <- avail_feats <- intersect(MODEL_FEATURES, names(ph)), and
    # require_features() guarantees all of them are present, so the intersect is
    # the declared vector unchanged. Reconstructing from MODEL_FEATURES
    # therefore reproduces the training design matrix exactly (verified against
    # 06_predictions_comparators_B: zero difference over 548,827 test rows).
    #
    # check_model_features() decides between the two cases rather than leaving
    # the fallback silent: where the booster DID persist its names, disagreeing
    # with config.R is now an error, because a permuted design matrix yields
    # wrong predictions that nothing downstream can detect; where it did not,
    # the declared vector is used and the log says so explicitly.
    gbt_feats <- check_model_features(gbt_feats, MODEL_FEATURES,
                                      tag = paste0("gbt/", vid, "/", anchor))
    nr_ep     <- nrow(eicu_pred)
    gbt_cols  <- lapply(gbt_feats, function(f) {
      v <- if (f %in% names(eicu_pred)) as.numeric(eicu_pred[[f]]) else rep(0.0, nr_ep)
      v[is.na(v)] <- 0.0; v
    })
    names(gbt_cols) <- gbt_feats
    X_eicu <- as.matrix(as.data.table(gbt_cols))
    storage.mode(X_eicu) <- "double"
    if (ncol(X_eicu) == 0L)
      stop("GBT design matrix has zero columns; the feature list is empty.")
    eicu_pred[, pred_gbt := predict(gbt_mod, xgboost::xgb.DMatrix(X_eicu))]
    # A silently all-NA prediction vector is indistinguishable downstream from
    # "this model was not evaluated". Fail loudly instead.
    if (all(is.na(eicu_pred$pred_gbt)))
      stop(sprintf("GBT predictions are entirely NA on eICU for variant %s.", vid))
    rm(X_eicu, gbt_cols, gbt_mod); gc()
  }

  eicu_pred[, pred_news2 := news2_total / 20]

  y    <- as.integer(eicu_pred$event == 1)
  rows <- list()
  for (mod_nm in c("pred_sepsis","pred_gbt","pred_news2")) {
    if (!mod_nm %in% names(eicu_pred)) next
    pr    <- as.numeric(eicu_pred[[mod_nm]])
    valid <- !is.na(pr)
    if (sum(valid) < 50) next

    roc_pr <- tryCatch({
      list(auroc = PRROC::roc.curve(pr[valid], weights.class0 = y[valid])$auc,
           auprc = PRROC::pr.curve(pr[valid],  weights.class0 = y[valid])$auc.integral)
    }, error = function(e) list(auroc = NA_real_, auprc = NA_real_))

    rows[[mod_nm]] <- data.frame(
      variant  = vid,
      model    = mod_nm,
      cohort   = "eicu",
      anchor   = anchor,
      n_stays  = uniqueN(eicu_pred$stay_id),
      n_rows   = sum(valid),
      n_events = sum(y[valid]),
      # Three different counts of "events" exist here and the thesis quotes two
      # of them, so all three are recorded rather than left to be reconciled by
      # subtracting one table from another.
      #
      #   n_onsets_labelled  stays given an onset by the variant's rule, at any
      #                      offset. This is 08_eicu_label_summary's n_onsets:
      #                      31 under Variant B.
      #   n_events           onset hours actually scored: 29 under Variant B.
      #
      # The gap is attach_labels()'s horizon rule: an onset falling after the
      # stay's last observed hour is right-censoring, not an event, because no
      # model could have been asked to predict it inside the panel. That is
      # correct and matches the MIMIC 72 h horizon, but it is undocumented, and
      # at 29 events a reader is entitled to know where the other two went.
      n_onsets_labelled   = uniqueN(eicu_pred[!is.na(onset_hour), stay_id]),
      n_onsets_beyond_panel = uniqueN(eicu_pred[!is.na(onset_hour), stay_id]) -
                              uniqueN(eicu_pred[in_panel == TRUE, stay_id]),
      # Separately: an onset hour whose prediction is NA would leave the
      # analysis silently. Asserted to be zero rather than assumed.
      n_events_dropped_na = sum(y) - sum(y[valid]),
      auroc    = round(roc_pr$auroc, 4),
      auprc    = round(roc_pr$auprc, 4)
    )

    if (sum(y) != sum(y[valid]))
      v2_log(sprintf(paste0("  [%s/%s/%s] %d of %d labelled onset hours have no ",
                            "prediction and are excluded from scoring"),
                     vid, anchor, mod_nm, sum(y) - sum(y[valid]), sum(y)))
    n_lab <- uniqueN(eicu_pred[!is.na(onset_hour), stay_id])
    if (n_lab != sum(y[valid]))
      v2_log(sprintf(paste0("  [%s/%s/%s] %d onsets labelled, %d scored; %d fall ",
                            "beyond the stay's observed panel and are right-censored"),
                     vid, anchor, mod_nm, n_lab, sum(y[valid]), n_lab - sum(y[valid])))
  }
  list(rows = rows, pred = eicu_pred)
}

eval_primary   <- list()
eval_sensitivity <- list()

for (vid in names(LABEL_VARIANTS)) {
  # Primary arm: Sepsis-3 conjunction on the microbiology-covered sub-cohort.
  g   <- attach_labels(panel_primary, eicu_labels[[paste(vid, "sepsis3", sep = "-")]],
                       stays_primary)
  res <- evaluate_panel(g, vid, "sepsis3")
  rm(g); gc()
  if (!is.null(res)) {
    eval_primary <- c(eval_primary, res$rows)
    save_cols <- c("stay_id","hour","event","pred_sepsis","onset_hour",
                   intersect(c("pred_gbt","pred_news2","news2_total","qsofa_score"),
                             names(res$pred)))
    save_parquet(res$pred[, ..save_cols], paste0("08_eicu_predictions_", vid), DATA_DIR)
    rm(res); gc()
  }

  # Sensitivity arm: antibiotic-only anchor on the full cohort. Metrics only:
  # the target is not Sepsis-3, so these predictions are never written to the
  # files stages 11 and 12 consume.
  g   <- attach_labels(panel, eicu_labels[[paste(vid, "abx_only", sep = "-")]],
                       eicu_stays)
  res <- evaluate_panel(g, vid, "abx_only")
  rm(g); gc()
  if (!is.null(res)) eval_sensitivity <- c(eval_sensitivity, res$rows)
  rm(res); gc()
}
```

## Results

The two arms answer different questions. The primary arm applies the Sepsis-3
definition to the stays with microbiology records. The sensitivity arm drops
the culture requirement and covers the full eICU-CRD cohort. This gives many
more events, but the target is no longer Sepsis-3. Its results show how far
the limited microbiology coverage constrains the event count, and they are not
an external Sepsis-3 result.

The results of both arms are written to `08_external_validation_results` and
`08_external_validation_results_abxonly` in `output/processed_data/`. Each row
gives a model's AUROC and AUPRC, with three event counts: the stays given an
onset by the variant's rule, the onset hours actually scored, and the
difference between the two. An onset that falls after a stay's last observed
hour is right-censored, under the same horizon rule as in MIMIC-IV, and is not
scored. This is why a variant's scored external events, printed below, can be fewer
than its onsets in Table 1, as they are for Variant B.

```{r results-tables}
ext_df <- if (length(eval_primary) > 0) do.call(rbind, eval_primary) else NULL
sen_df <- if (length(eval_sensitivity) > 0) do.call(rbind, eval_sensitivity) else NULL

if (!is.null(ext_df)) {
  knitr::kable(ext_df,
               caption = "External validation, Sepsis-3 anchor on the microbiology-covered sub-cohort")
  save_parquet(ext_df, "08_external_validation_results", DATA_DIR)
  save_csv(ext_df,     "08_external_validation_results", DATA_DIR)
}
if (!is.null(sen_df)) {
  knitr::kable(sen_df,
               caption = "Sensitivity arm: antibiotic-only anchor on the full eICU cohort (target is not Sepsis-3)")
  save_parquet(sen_df, "08_external_validation_results_abxonly", DATA_DIR)
  save_csv(sen_df,     "08_external_validation_results_abxonly", DATA_DIR)
}
```

### Event-rate reconciliation

This step compares the onset rate per 1,000 person-hours, for each variant,
across three sources: the MIMIC-IV temporal test set, the eICU-CRD Sepsis-3
arm and the eICU-CRD antibiotic-only arm. It also gives each eICU-CRD rate as a
ratio to the MIMIC-IV rate. Under the same label, a much lower external rate
can reflect how completely the label can be identified in the external data,
as well as a real difference between the populations. The comparison across
arms helps to separate the two. If the rate rises substantially once the
culture requirement is dropped, much of the gap comes from the microbiology
records, although differences between the populations cannot be excluded. The
table is written to `08_event_rate_reconciliation` in
`output/processed_data/`.

```{r rate-reconciliation}
recon <- NULL
internal_path <- file.path(DATA_DIR, "07_metric_results.parquet")

if (!is.null(ext_df) && file.exists(internal_path)) {
  internal_res <- as.data.table(arrow::read_parquet(internal_path))
  ext_dt <- as.data.table(ext_df)
  sen_dt <- if (!is.null(sen_df)) as.data.table(sen_df) else NULL

  rate_rows <- list()
  for (vid in names(LABEL_VARIANTS)) {
    int_row <- internal_res[variant == vid & model == "primary" &
                              eval_window == "unrestricted" & split_type == "temporal"]
    ext_row <- ext_dt[variant == vid & model == "pred_sepsis"]
    sen_row <- if (!is.null(sen_dt)) sen_dt[variant == vid & model == "pred_sepsis"] else NULL

    mk_rate <- function(src, n_rows, n_events, n_stays) data.table(
      variant = vid, source = src,
      n_stays = n_stays, person_hours = n_rows, n_events = n_events,
      events_per_1k_person_h = round(1000 * n_events / max(n_rows, 1), 4)
    )

    if (nrow(int_row) > 0)
      rate_rows[[paste0(vid, "-mimic")]] <-
        mk_rate("MIMIC-IV temporal test", int_row$n_rows[1],
                int_row$n_sepsis_events[1], NA_integer_)
    if (nrow(ext_row) > 0)
      rate_rows[[paste0(vid, "-eicu")]] <-
        mk_rate("eICU Sepsis-3 (micro-covered)", ext_row$n_rows[1],
                ext_row$n_events[1], ext_row$n_stays[1])
    if (!is.null(sen_row) && nrow(sen_row) > 0)
      rate_rows[[paste0(vid, "-eicuabx")]] <-
        mk_rate("eICU antibiotic-only (full)", sen_row$n_rows[1],
                sen_row$n_events[1], sen_row$n_stays[1])
  }

  if (length(rate_rows) > 0) {
    recon <- rbindlist(rate_rows)
    # Ratio against the development-cohort rate for the same variant. Written
    # as a join rather than a by-group subscript so a variant with no MIMIC row
    # yields NA instead of a length-zero assignment error.
    mimic_rate <- recon[source == "MIMIC-IV temporal test",
                        .(variant, mimic_rate = events_per_1k_person_h)]
    recon <- merge(recon, mimic_rate, by = "variant", all.x = TRUE, sort = FALSE)
    recon[, rate_ratio_vs_mimic := round(events_per_1k_person_h / mimic_rate, 3)]
    recon[, mimic_rate := NULL]
    knitr::kable(recon, caption = "Sepsis event rate, MIMIC-IV versus eICU-CRD")
    save_parquet(recon, "08_event_rate_reconciliation", DATA_DIR)
    save_csv(recon,     "08_event_rate_reconciliation", DATA_DIR)
  }
}

# Microbiology coverage is the quantity that governs how much of the eICU
# cohort the Sepsis-3 anchor can reach at all, so it is persisted rather than
# left in the log.
#
# Three different denominators are in play and each is recorded separately,
# because they are easy to confuse and they answer different questions:
#   - a limb is *documented* for a stay (either table holds a record);
#   - *both* limbs are documented for a stay, at any offset: the ceiling on how
#     many stays the Sepsis-3 conjunction could reach even with an unbounded
#     window;
#   - the pair fires *within a variant's window*, which is the suspicion count
#     in 08_eicu_label_summary and is necessarily no larger than the ceiling.
# Every count is restricted to the analysis cohort so that numerator and
# denominator describe the same set of stays.
abx_cohort_stays  <- intersect(unique(eicu_abx$stay_id), cohort_ids)
both_limb_stays   <- intersect(abx_cohort_stays, micro_stays)

cov_dt <- data.table(
  n_cohort_stays        = length(cohort_ids),
  n_micro_covered_stays = length(micro_stays),
  pct_micro_covered     = round(100 * length(micro_stays) /
                                  max(length(cohort_ids), 1), 3),
  n_abx_stays           = length(abx_cohort_stays),
  pct_abx_covered       = round(100 * length(abx_cohort_stays) /
                                  max(length(cohort_ids), 1), 3),
  n_both_limbs_stays    = length(both_limb_stays),
  pct_both_limbs        = round(100 * length(both_limb_stays) /
                                  max(length(cohort_ids), 1), 3),
  panel_person_hours_full  = nrow(panel),
  panel_person_hours_micro = nrow(panel_primary),
  min_external_events      = MIN_EXTERNAL_EVENTS
)
knitr::kable(cov_dt, caption = "eICU labelling-input coverage")
save_parquet(cov_dt, "08_eicu_coverage", DATA_DIR)
save_csv(cov_dt,     "08_eicu_coverage", DATA_DIR)
```

### H5: internal versus external AUROC

For each variant, the printed lines give the primary model's internal AUROC
(MIMIC-IV temporal test set), its external AUROC (Sepsis-3 arm) and the
decrease between them, labelled VA, VB and VC for Variants A--C. Each decrease
is shown next to the pre-registered reference of 0.085 from Moor et al.
(2023). That reference is for comparison only and is not a threshold. Under
Protocol Amendment 1, the Variant B decrease is the contrast tested for H5.
These results are descriptive, and the interval and adjusted p-value come from
`12_inference.Rmd`.

The minimum number of external events for interpreting an external AUROC is
`r MIN_EXTERNAL_EVENTS` (Riley et al., 2021; Collins et al., 2024). A variant
below this minimum is still printed, with a warning, because its event count is
itself part of the result. At such counts, the AUROC is too imprecise for any
firm conclusion, and a decrease smaller than the reference does not by itself
show that discrimination transports well. The cohort is also restricted to the
small set of stays with microbiology records, so its patients may differ from
the eICU-CRD population as a whole.

```{r results-h5}
if (!is.null(ext_df) && file.exists(internal_path)) {
  internal_res <- as.data.table(arrow::read_parquet(internal_path))
  ext_dt <- as.data.table(ext_df)
  cat("\nH5 test: AUROC internal -> external\n")
  h5_rows <- list()
  for (vid in names(LABEL_VARIANTS)) {
    int_row <- internal_res[variant == vid & model == "primary" &
                              eval_window == "unrestricted" & split_type == "temporal"]
    ext_row <- ext_dt[variant == vid & model == "pred_sepsis"]
    if (nrow(int_row) > 0 && nrow(ext_row) > 0) {
      auroc_drop <- int_row$auroc[1] - ext_row$auroc[1]
      cat(sprintf("  V%s: Internal AUROC=%.3f, External AUROC=%.3f, drop=%.3f (%d external events)\n",
                  vid, int_row$auroc[1], ext_row$auroc[1], auroc_drop, ext_row$n_events[1]))
      cat(sprintf("  V%s: Expected drop ~ %.3f (Moor et al., 2023)\n",
                  vid, EXPECTED_EXTERNAL_DEGRADATION))
      if (ext_row$n_events[1] < MIN_EXTERNAL_EVENTS)
        cat(sprintf("  ** V%s: %d external events is below the %d-event floor; the estimate is not interpretable **\n",
                    vid, ext_row$n_events[1], MIN_EXTERNAL_EVENTS))
      h5_rows[[vid]] <- data.frame(variant = vid,
                                    internal_auroc = int_row$auroc[1],
                                    external_auroc = ext_row$auroc[1],
                                    auroc_drop     = round(auroc_drop, 4),
                                    expected_drop  = EXPECTED_EXTERNAL_DEGRADATION,
                                    n_external_events = ext_row$n_events[1],
                                    interpretable  = ext_row$n_events[1] >= MIN_EXTERNAL_EVENTS)
    }
  }
  if (length(h5_rows) > 0) {
    h5_dt <- rbindlist(h5_rows)
    knitr::kable(h5_dt, caption = "H5: internal versus external AUROC")
    save_parquet(h5_dt, "08_h5_internal_external", DATA_DIR)
    save_csv(h5_dt,     "08_h5_internal_external", DATA_DIR)
  }
}

cat("\nExternal validation complete.\n")
```

## Treatment-Independent Label on the Full eICU Cohort (Variant D)

The analyses above are limited in one of two ways. The primary arm is confined
to stays with microbiology records, which are
`r round(100 * length(micro_stays) / max(length(cohort_ids), 1), 2)`% of the
analysis cohort, and the sensitivity arm targets something other than
Sepsis-3. Both limits arise because the Sepsis-3 definition requires a culture
record.

Variant D, introduced by Protocol Amendment 2, has no such requirement. Its
onset is the first hour at which a SOFA score computed from physiology alone is
at least `r SOFA_INCREASE_THRESHOLD` points above the stay's admission
baseline. It has no infection criterion, and the vasopressor columns are not
passed to `score_sofa()`, so the cardiovascular component rests on mean
arterial pressure alone. The definition uses no antibiotic, culture or
medication record. It can therefore be applied to every stay in the cohort,
rather than only to the `r format(length(micro_stays), big.mark = ",")` stays
with microbiology records. Its external evaluation was added by Protocol
Amendment 9 for this reason. It is the only external analysis in the study
whose denominator is the whole eICU-CRD cohort.

Three limitations govern how the result is read:

* **Post-hoc status.** Variant D lies outside the pre-registered A--C family.
  It is excluded from H1, H5 and the variance decomposition. The analysis is
  descriptive, with no p-value, and it belongs to no multiplicity-adjusted
  family (Protocol Amendment 9). Its results go to a separate output file, and
  the checks at the end of this section confirm that the confirmatory and
  post-hoc files stay separate.
* **Different target.** Variant D identifies acute organ dysfunction without
  requiring suspected infection, so it is not a Sepsis-3 outcome. Its AUROC
  answers a different question from the primary arm's, and the two are not
  compared directly or pooled. Because the label is a threshold on physiology
  that also enters the models as covariates, its AUROC is not a measure of
  sepsis prediction.
* **Different SOFA components.** `vitalPeriodic` records no GCS, so the
  neurological component is scored as normal in both the hourly score and the
  baseline. It therefore contributes no change to the eICU-CRD rise, and
  neurological deterioration cannot trigger an external onset. The eICU-CRD
  rise rests on five SOFA components, against six in MIMIC-IV. This asymmetry
  between the development and validation cohorts should be reported with the
  Variant D estimate.

The derivation follows `derive_deterioration_onsets()` in
`03_person_hours.Rmd`: the same physiological components, the same admission
baseline window and the same minimum observation time. This keeps the external
definition of Variant D consistent with the internal one, apart from the
missing GCS. The printed line gives the number and percentage of stays that
meet the Variant D criterion, and the median onset hour.

```{r eicu-deterioration-onsets}
EICU_SOFA_PHYS_COMPONENTS <- c("pf_ratio", "platelets", "bilirubin", "map",
                               "gcs", "creatinine")

#' Variant D onsets on the eICU panel: the first hour at which physiology-only
#' SOFA exceeds the stay's admission baseline by SOFA_INCREASE_THRESHOLD.
#'
#' Mirrors derive_deterioration_onsets() in 03_person_hours.Rmd. The
#' vasopressor columns are deliberately NOT passed to score_sofa(): their
#' absence is what makes the cardiovascular component MAP-only and the label
#' free of any treatment record.
derive_eicu_deterioration_onsets <- function(pan) {
  keep <- intersect(c("stay_id", "hour", "pf_ratio", "platelets",
                      "bilirubin_total", "map", "gcs", "creatinine"),
                    names(pan))
  d <- pan[, ..keep]
  if ("bilirubin_total" %in% names(d) && !"bilirubin" %in% names(d))
    setnames(d, "bilirubin_total", "bilirubin")
  comps <- intersect(EICU_SOFA_PHYS_COMPONENTS, names(d))

  d[, sofa_phys := score_sofa(as.data.frame(d))$sofa_total]

  base_dt <- d[hour <= DETERIORATION_BASELINE_H,
               lapply(.SD, mean, na.rm = TRUE), by = stay_id, .SDcols = comps]
  base_dt[, sofa_phys_base := score_sofa(as.data.frame(base_dt))$sofa_total]

  d <- merge(d, base_dt[, .(stay_id, sofa_phys_base)], by = "stay_id", all.x = TRUE)
  d[is.na(sofa_phys_base), sofa_phys_base := 0L]

  onsets <- d[hour > MIN_OBS_HOURS &
                (sofa_phys - sofa_phys_base) >= SOFA_INCREASE_THRESHOLD,
              .(onset_hour = min(hour)), by = stay_id]

  # Every stay in the panel is returned, so the at-risk denominator is the
  # whole cohort and a stay that never deteriorates is a censored negative
  # rather than a missing row.
  out <- merge(data.table(stay_id = unique(pan$stay_id)), onsets,
               by = "stay_id", all.x = TRUE)
  cat(sprintf("  eICU Variant D: %s / %s stays deteriorate (%.2f%%), median onset hour %.1f\n",
              format(nrow(onsets), big.mark = ","),
              format(uniqueN(pan$stay_id), big.mark = ","),
              100 * nrow(onsets) / max(uniqueN(pan$stay_id), 1),
              if (nrow(onsets) > 0) median(onsets$onset_hour) else NA_real_))
  out[]
}
```

```{r altlabel-external-eval}
step("Evaluating the treatment-independent label (D) on the full eICU cohort...")

alt_eval        <- list()
alt_label_rows  <- list()
alt_df          <- NULL

# Only variants whose onset needs no anchor can run on the full cohort. This is
# a property of the variant definition, not a hard-coded list, so a future
# treatment-independent variant is picked up without editing this loop.
alt_external_ids <- Filter(function(v) identical(variant_type(v), "deterioration"),
                           names(ALT_LABEL_VARIANTS))

for (vid in alt_external_ids) {
  ons <- derive_eicu_deterioration_onsets(panel)
  n_onset <- ons[!is.na(onset_hour), .N]

  alt_label_rows[[vid]] <- data.table(
    variant        = vid,
    variant_name   = variant_name(vid),
    anchor         = "deterioration",
    n_stays        = uniqueN(panel$stay_id),
    # D has no suspicion limb at all; recorded as NA rather than as 0, which
    # would read as "the pair never fired".
    n_suspicion    = NA_integer_,
    n_onsets       = n_onset,
    pct_onsets     = round(100 * n_onset / max(uniqueN(panel$stay_id), 1), 2),
    median_onset_h = if (n_onset > 0)
      as.numeric(median(ons[!is.na(onset_hour), onset_hour])) else NA_real_
  )

  g   <- attach_labels(panel, ons, eicu_stays)
  res <- evaluate_panel(g, vid, "deterioration")
  rm(g); gc()
  if (!is.null(res)) {
    alt_eval <- c(alt_eval, res$rows)
    # Persisted from Protocol Amendment 10 onwards. The earlier version of this
    # stage wrote metrics only, which left the one external arm that clears the
    # event floor as the one external arm with no interval on it. Stage 12
    # bootstraps the internal-to-external transport contrast from this file.
    # Stage 11's external loop reads names(LABEL_VARIANTS) and still ignores it.
    save_cols <- c("stay_id", "hour", "event", "pred_sepsis", "onset_hour",
                   intersect(c("pred_gbt", "pred_news2"), names(res$pred)))
    save_parquet(res$pred[, ..save_cols],
                 paste0("08_eicu_predictions_", vid), DATA_DIR)
  }
  rm(res); gc()
}

if (length(alt_eval) > 0) {
  alt_df <- as.data.table(do.call(rbind, alt_eval))
  knitr::kable(alt_df,
               caption = "Treatment-independent label (Variant D) on the full eICU cohort")
  save_parquet(alt_df, "08_external_validation_results_altlabel", DATA_DIR)
  save_csv(alt_df,     "08_external_validation_results_altlabel", DATA_DIR)
}

if (length(alt_label_rows) > 0) {
  label_summary_dt <- rbindlist(c(list(label_summary_dt), alt_label_rows),
                                fill = TRUE)
  save_parquet(label_summary_dt, "08_eicu_label_summary", DATA_DIR)
  save_csv(label_summary_dt,     "08_eicu_label_summary", DATA_DIR)
}
```

### Event rate under a treatment-independent label

The event-rate reconciliation above compares the Sepsis-3 arm's external
event rate with the development cohort's, and relates the gap to microbiology
coverage. Variant D gives a complementary view. Its label needs no treatment or
microbiology record, so its external rate is not limited by microbiology
coverage. If its rate ratio between the two databases is much closer to 1 than
the Sepsis-3 arm's, that supports attributing the Sepsis-3 gap mainly to the
treatment and culture records rather than to the population. It does not prove
the attribution. Variant D is a different outcome, and its eICU-CRD score
omits the neurological component. The rates are written to
`08_event_rate_altlabel`, a separate file, so the pre-registered reconciliation
table is unchanged.

```{r altlabel-rate-reconciliation}
alt_recon <- NULL
if (!is.null(alt_df) && file.exists(internal_path)) {
  internal_alt_path <- file.path(DATA_DIR, "07_metric_results_altlabel.parquet")
  if (file.exists(internal_alt_path)) {
    int_alt <- as.data.table(arrow::read_parquet(internal_alt_path))
    rows <- list()
    for (vid in unique(alt_df$variant)) {
      # The unrestricted temporal window: the same window every other
      # internal-versus-external comparison in this stage uses.
      ir <- int_alt[variant == vid & model == "primary" &
                      eval_window == "unrestricted" & split_type == "temporal"]
      er <- alt_df[variant == vid & model == "pred_sepsis"]
      if (nrow(ir) > 0)
        rows[[paste0(vid, "-mimic")]] <- data.table(
          variant = vid, source = "MIMIC-IV temporal test",
          n_stays = NA_integer_, person_hours = ir$n_rows[1],
          n_events = ir$n_sepsis_events[1],
          events_per_1k_person_h = round(1000 * ir$n_sepsis_events[1] /
                                           max(ir$n_rows[1], 1), 4))
      if (nrow(er) > 0)
        rows[[paste0(vid, "-eicu")]] <- data.table(
          variant = vid, source = "eICU deterioration (full)",
          n_stays = er$n_stays[1], person_hours = er$n_rows[1],
          n_events = er$n_events[1],
          events_per_1k_person_h = round(1000 * er$n_events[1] /
                                           max(er$n_rows[1], 1), 4))
    }
    if (length(rows) > 0) {
      alt_recon <- rbindlist(rows)
      mim <- alt_recon[source == "MIMIC-IV temporal test",
                       .(variant, mimic_rate = events_per_1k_person_h)]
      alt_recon <- merge(alt_recon, mim, by = "variant", all.x = TRUE, sort = FALSE)
      alt_recon[, rate_ratio_vs_mimic := round(events_per_1k_person_h / mimic_rate, 4)]
      alt_recon[, mimic_rate := NULL]
      knitr::kable(alt_recon,
                   caption = "Event rate under the treatment-independent label")
      save_parquet(alt_recon, "08_event_rate_altlabel", DATA_DIR)
      save_csv(alt_recon,     "08_event_rate_altlabel", DATA_DIR)
    }
  }
}
```

### Checks That Keep Variant D Separate

Variant D is excluded from H5 and the variance decomposition, as it is from
H1. The separation rests on file names. The confirmatory stages read
`08_external_validation_results`, while this arm writes to
`08_external_validation_results_altlabel`. The notebook also checks the
separation each time it runs, so that a later change to the code cannot break
it unnoticed. It stops in three cases:

- the Variant D file contains a pre-registered variant;
- the primary external file contains Variant D;
- Variant D was scored on no more stays than the microbiology-covered
  sub-cohort, which would mean it is not the full-cohort analysis it claims to
  be.

The final printed line gives the number of Variant D external events and
states whether it reaches the `r MIN_EXTERNAL_EVENTS`-event minimum.

```{r altlabel-containment-guards}
if (!is.null(alt_df)) {
  leaked_in <- intersect(unique(alt_df$variant), names(LABEL_VARIANTS))
  if (length(leaked_in) > 0)
    stop(sprintf(paste0("Containment guard failed: the alt-label external file ",
                        "carries pre-registered variant(s) %s. The post-hoc arm ",
                        "must never restate a confirmatory result."),
                 paste(leaked_in, collapse = "/")))

  if (!is.null(ext_df)) {
    leaked_out <- intersect(unique(as.data.table(ext_df)$variant), alt_external_ids)
    if (length(leaked_out) > 0)
      stop(sprintf(paste0("Containment guard failed: the primary external file ",
                          "carries post-hoc variant(s) %s, which would put a ",
                          "post-hoc label inside H5."),
                   paste(leaked_out, collapse = "/")))
  }

  # The entire purpose of this arm is that its denominator is the whole cohort.
  # If it silently ran on the microbiology-covered sub-cohort the AUROC would
  # still be well-formed and the finding would be wrong.
  n_alt_stays <- max(alt_df$n_stays, na.rm = TRUE)
  if (n_alt_stays <= length(micro_stays))
    stop(sprintf(paste0("Containment guard failed: the alt-label arm scored %s ",
                        "stays, no more than the %s in the microbiology-covered ",
                        "sub-cohort. It is not the full-cohort estimate it claims ",
                        "to be."),
                 format(n_alt_stays, big.mark = ","),
                 format(length(micro_stays), big.mark = ",")))

  v2_log(sprintf(paste0("Containment guards passed: alt-label arm carries %s ",
                        "only, scored on %s stays (%.1fx the micro sub-cohort)."),
                 paste(alt_external_ids, collapse = "/"),
                 format(n_alt_stays, big.mark = ","),
                 n_alt_stays / max(length(micro_stays), 1)))

  n_alt_events <- max(alt_df$n_events, na.rm = TRUE)
  cat(sprintf("\nVariant D external: %s events on %s stays (interpretability floor %d): %s\n",
              format(n_alt_events, big.mark = ","),
              format(n_alt_stays, big.mark = ","),
              MIN_EXTERNAL_EVENTS,
              if (n_alt_events >= MIN_EXTERNAL_EVENTS) "above the floor" else
                "BELOW the floor; not interpretable"))
}

cat("\nTreatment-independent external arm complete.\n")
```
