---
title: "08 External Validation - eICU-CRD External Validation"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

External validation on a distinct dataset is the minimum standard for generalisability (Collins et al., 2024, TRIPOD+AI). The eICU-CRD (Pollard et al., 2018) provides multi-centre US ICU data with different demographics, protocols, and data infrastructure.

Expected AUC degradation: ~0.085 (Moor et al., 2023). This analysis tests **H5**: Utility Score degrades proportionally more than AUROC on external data.

### Two anchor arms, and why both are needed

Sepsis-3 onset is anchored on the co-occurrence of an antibiotic order and a
microbiology culture. In eICU-CRD that conjunction is only *recoverable* where
microbiology exists, and `microLab.csv.gz` covers a small minority of unit
stays. Applying the conjunction to the full cohort therefore does not measure a
low sepsis incidence; it measures the coverage of one source table, because the
overwhelming majority of person-hours sit in stays where the anchor cannot fire
by construction. The two arms separate those questions:

| Arm | Cohort | Anchor | Role |
|---|---|---|---|
| **Sepsis-3 (primary)** | stays with $\geq 1$ culture record | culture $\wedge$ antibiotic, per-variant windows | label matches MIMIC-IV exactly; denominator contains only person-hours in which an onset *could* be observed |
| **Antibiotic-only (sensitivity)** | full cohort | first systemic antibiotic, no culture requirement | quantifies how much of the event count the culture requirement removes; the target is **not** Sepsis-3 and is never pooled with the primary arm |

Steps:

1. Build the eICU person-hour panel harmonised to MIMIC-IV naming
2. Derive onsets **per label variant** under each anchor, using the same
   SOFA-delta criterion as `02_cohort_labels.Rmd`
3. Load MIMIC-IV trained models (no retraining)
4. Evaluate with the same metric suite
5. Reconcile the eICU and MIMIC-IV event rates explicitly

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(nnet)
  library(splines)
  library(xgboost)
  library(PRROC)
  library(dplyr)
})

source("config.R")
source("utils.R")
source("clinical_scores.R")

step("=== Phase 5: External Validation on eICU-CRD ===")
cat(sprintf("Pre-registered expected degradation: %.3f AUC\n",
            EXPECTED_EXTERNAL_DEGRADATION))
```

## Load eICU Patient Table

Same inclusion filters as MIMIC-IV: age $\geq$ 18, LOS $\geq$ 4 hours.

```{r load-eicu-patients}
step("Loading eICU tables...")

eicu_patient <- data.table::fread(file.path(EICU_DIR, "patient.csv.gz"),
                                   showProgress = FALSE)
cat(sprintf("eICU patients (raw): %s\n", format(nrow(eicu_patient), big.mark = ",")))

eicu_patient[, age_num := suppressWarnings(as.numeric(age))]
eicu_patient[age == ">89", age_num := 90]
eicu_patient[, los_hours := unitdischargeoffset / 60]
eicu_patient <- eicu_patient[!is.na(age_num) & age_num >= MIN_AGE &
                               los_hours >= MIN_ICU_LOS_HOURS]
cat(sprintf("eICU cohort after exclusions: %s\n",
            format(nrow(eicu_patient), big.mark = ",")))
```

## Vital Signs (vitalPeriodic)

Loaded via DuckDB with pushdown filters on cohort membership and time window.

```{r load-vitals}
step("Loading eICU vitals...")

vp_path <- file.path(EICU_DIR, "vitalPeriodic.csv.gz")
if (file.exists(vp_path)) {
  suppressPackageStartupMessages(library(duckdb))
  .vp_con <- dbConnect(duckdb(), dbdir = ":memory:")
  dbWriteTable(.vp_con, "cohort_stays",
               data.frame(patientunitstayid = as.integer(eicu_patient$patientunitstayid)))
  sql_vp <- sprintf(
    "SELECT t.patientunitstayid, t.observationoffset,
            t.heartrate, t.respiration, t.sao2,
            t.systemicsystolic, t.systemicmean, t.temperature
     FROM read_csv('%s', header=true, compression='gzip') t
     INNER JOIN cohort_stays c ON t.patientunitstayid = c.patientunitstayid
     WHERE t.observationoffset >= 0 AND t.observationoffset <= %d",
    vp_path, MAX_PANEL_HOURS * 60L
  )
  v2_log("DuckDB: loading vitalPeriodic filtered to cohort + 72h window...")
  vp <- as.data.table(dbGetQuery(.vp_con, sql_vp))
  dbDisconnect(.vp_con, shutdown = TRUE)
  cat(sprintf("vitalPeriodic (cohort-filtered): %s rows\n", format(nrow(vp), big.mark = ",")))

  setnames(vp, c("patientunitstayid","observationoffset","heartrate","respiration",
                  "sao2","systemicsystolic","systemicmean","temperature"),
               c("stay_id","obs_offset_m","hr","resp_rate","spo2","sbp","map","temp_c"))
  vp[, obs_hour := floor(obs_offset_m / 60)]
  vp[!is.na(temp_c) & (temp_c < 30 | temp_c > 45), temp_c := NA]
  vp_hr <- vp[, .(hr        = mean(hr,        na.rm = TRUE),
                   resp_rate = mean(resp_rate, na.rm = TRUE),
                   spo2      = mean(spo2,      na.rm = TRUE),
                   sbp       = mean(sbp,       na.rm = TRUE),
                   map       = mean(map,       na.rm = TRUE),
                   temp_c    = mean(temp_c,    na.rm = TRUE)),
               by = .(stay_id, obs_hour)]
  rm(vp); gc()
} else {
  vp_hr <- data.table()
  v2_log("vitalPeriodic not found", level = "WARN")
}
```

## Lab Values

Lab values are mapped from eICU names to MIMIC-IV conventions. PaO$_2$/FiO$_2$ ratios computed where both components exist.

```{r load-labs}
step("Loading eICU labs...")

lab_path <- file.path(EICU_DIR, "lab.csv.gz")
if (file.exists(lab_path)) {
  eicu_lab <- data.table::fread(lab_path, showProgress = FALSE,
                                 select = c("patientunitstayid", "labresultoffset",
                                            "labname", "labresult"))
  setnames(eicu_lab, c("patientunitstayid","labresultoffset","labname","labresult"),
                     c("stay_id","obs_offset_m","labname","value"))
  eicu_lab[, obs_hour := floor(obs_offset_m / 60)]
  eicu_lab <- eicu_lab[obs_hour >= 0 & obs_hour <= MAX_PANEL_HOURS & !is.na(value)]
  eicu_lab[, value := as.numeric(value)]

  lab_map <- c("lactate" = "lactate", "creatinine" = "creatinine",
               "total bilirubin" = "bilirubin_total", "platelets x 1000" = "platelets",
               "WBC x 1000" = "wbc", "paO2" = "pao2", "FiO2" = "fio2_lab",
               "sodium" = "sodium", "potassium" = "potassium")
  eicu_lab <- eicu_lab[labname %in% names(lab_map)]
  eicu_lab[, var_name := lab_map[labname]]

  lab_hr <- dcast(
    eicu_lab[, .(value = mean(value, na.rm = TRUE)),
              by = .(stay_id, obs_hour, var_name)],
    stay_id + obs_hour ~ var_name, value.var = "value", fun.aggregate = mean
  )
  if ("pao2" %in% names(lab_hr) && "fio2_lab" %in% names(lab_hr)) {
    lab_hr[, fio2_frac := ifelse(fio2_lab > 1, fio2_lab / 100, fio2_lab)]
    lab_hr[, pf_ratio  := ifelse(!is.na(pao2) & !is.na(fio2_frac) & fio2_frac > 0,
                                  pao2 / fio2_frac, NA_real_)]
    # fio2 is retained (as a fraction) because NEWS2 needs the supplemental-oxygen
    # flag; without it every eICU person-hour scores 0 on that parameter.
    lab_hr[, fio2 := fio2_frac]
    lab_hr[, c("fio2_lab", "fio2_frac") := NULL]
  }
} else {
  lab_hr <- data.table()
  v2_log("lab.csv not found", level = "WARN")
}
```

## Drug Exposure: Antibiotics and Vasopressors

eICU records drugs across two tables. `medication.csv.gz` holds physician
orders and `infusionDrug.csv.gz` holds continuous infusions; vasopressors and a
large share of the intravenous antibiotics appear only in the latter. Both are
read, and the antibiotic exposure is the union.

```{r eicu-drugs}
step("Loading eICU drug exposure (medication + infusionDrug)...")

med_path <- file.path(EICU_DIR, "medication.csv.gz")
inf_path <- file.path(EICU_DIR, "infusionDrug.csv.gz")

abx_parts    <- list()
vaso_hours   <- data.table()
norepi_hours <- data.table()

if (file.exists(med_path)) {
  meds <- data.table::fread(med_path, showProgress = FALSE,
                              select = c("patientunitstayid","drugstartoffset",
                                         "drugname","drugordercancelled"))
  meds <- meds[drugordercancelled != "Yes" & !is.na(drugstartoffset)]
  meds[, is_abx := match_antibiotics(drugname, ANTIBIOTICS)]
  abx_parts$medication <- meds[is_abx == TRUE,
                                .(stay_id  = patientunitstayid,
                                  abx_hour = floor(drugstartoffset / 60))]
  cat(sprintf("Antibiotic orders (medication): %s rows, %s stays\n",
              format(nrow(abx_parts$medication), big.mark = ","),
              format(uniqueN(abx_parts$medication$stay_id), big.mark = ",")))
  rm(meds); gc()
} else {
  v2_log("medication.csv.gz not found", level = "WARN")
}

if (file.exists(inf_path)) {
  infus <- data.table::fread(inf_path, showProgress = FALSE,
                               select = c("patientunitstayid","infusionoffset",
                                          "drugname"))
  infus <- infus[!is.na(infusionoffset)]
  infus[, inf_hour := floor(infusionoffset / 60)]

  infus[, is_abx := match_antibiotics(drugname, ANTIBIOTICS)]
  abx_parts$infusion <- infus[is_abx == TRUE,
                               .(stay_id = patientunitstayid, abx_hour = inf_hour)]
  cat(sprintf("Antibiotic infusions (infusionDrug): %s rows, %s stays\n",
              format(nrow(abx_parts$infusion), big.mark = ","),
              format(uniqueN(abx_parts$infusion$stay_id), big.mark = ",")))

  # Vasopressor exposure, restricted to the panel horizon. An infusion row is a
  # rate observation at one instant, so exposure is flagged on the hour of the
  # observation rather than over an interval (eICU carries no infusion stop
  # time). This is the closest available analogue to the MIMIC-IV
  # `inputevents` start/end expansion in 03_person_hours.Rmd.
  infus[, is_vaso   := match_antibiotics(drugname, EICU_VASOPRESSORS)]
  infus[, is_norepi := match_antibiotics(drugname, EICU_NOREPI_EPI)]
  vaso_hours <- unique(infus[is_vaso == TRUE & inf_hour >= 0 &
                               inf_hour <= MAX_PANEL_HOURS,
                             .(stay_id = patientunitstayid, obs_hour = inf_hour)])
  vaso_hours[, vaso_any := TRUE]
  norepi_hours <- unique(infus[is_norepi == TRUE & inf_hour >= 0 &
                                 inf_hour <= MAX_PANEL_HOURS,
                               .(stay_id = patientunitstayid, obs_hour = inf_hour)])
  norepi_hours[, norepi_epi_any := TRUE]
  cat(sprintf("Vasopressor infusion hours: %s (%s stays); norepi/epi: %s (%s stays)\n",
              format(nrow(vaso_hours), big.mark = ","),
              format(uniqueN(vaso_hours$stay_id), big.mark = ","),
              format(nrow(norepi_hours), big.mark = ","),
              format(uniqueN(norepi_hours$stay_id), big.mark = ",")))
  rm(infus); gc()
} else {
  v2_log("infusionDrug.csv.gz not found", level = "WARN")
}

eicu_abx <- if (length(abx_parts) > 0) unique(rbindlist(abx_parts)) else data.table()
cat(sprintf("Antibiotic exposure (union): %s rows, %s stays\n",
            format(nrow(eicu_abx), big.mark = ","),
            format(uniqueN(eicu_abx$stay_id), big.mark = ",")))
```

## Microbiology Coverage

`microLab.csv.gz` is the binding constraint on Sepsis-3 label transport. The
coverage figure below defines the primary arm's cohort: a stay with no culture
record cannot satisfy the Sepsis-3 conjunction under any variant, so counting
its person-hours in the denominator measures table coverage rather than
incidence.

```{r eicu-cultures}
step("Loading eICU microbiology...")

micro_path   <- file.path(EICU_DIR, "microLab.csv.gz")
eicu_culture <- data.table()

if (file.exists(micro_path)) {
  micro <- data.table::fread(micro_path, showProgress = FALSE,
                               select = c("patientunitstayid","culturetakenoffset",
                                          "culturesite"))
  micro <- micro[!culturesite %in% SURVEILLANCE_SPEC_TYPES]
  eicu_culture <- micro[!is.na(culturetakenoffset),
                         .(stay_id      = patientunitstayid,
                           culture_hour = floor(culturetakenoffset / 60))]
} else {
  v2_log("microLab.csv.gz not found", level = "WARN")
}

cohort_ids  <- unique(as.integer(eicu_patient$patientunitstayid))
micro_stays <- intersect(unique(eicu_culture$stay_id), cohort_ids)

cat(sprintf("Culture records: %s over %s stays\n",
            format(nrow(eicu_culture), big.mark = ","),
            format(uniqueN(eicu_culture$stay_id), big.mark = ",")))
cat(sprintf("Microbiology coverage within the analysis cohort: %s / %s stays (%.2f%%)\n",
            format(length(micro_stays), big.mark = ","),
            format(length(cohort_ids),  big.mark = ","),
            100 * length(micro_stays) / max(length(cohort_ids), 1)))
cat(sprintf("Culture records drawn before ICU admission (negative offset): %.1f%%\n",
            100 * mean(eicu_culture$culture_hour < 0)))
```

## Build eICU Person-Hour Panel

The panel is built **once**, from physiology alone, and carries no label. Onsets
are derived from it per variant in the next section, which is what makes the
three external label sets genuinely distinct. Construction mirrors MIMIC-IV
(hourly grid, plausibility filter, LOCF, missingness indicators, rolling
slopes, clinical scores). GCS is unavailable in `vitalPeriodic` and is left as
`NA` rather than imputed.

```{r build-eicu-panel}
step("Building eICU person-hour panel...")

eicu_stays <- eicu_patient[, .(stay_id = as.integer(patientunitstayid), los_hours,
                                age = age_num, gender, ethnicity,
                                hospital_expire = unitdischargestatus == "Expired")]
panel <- build_person_hour_grid(eicu_stays, max_hours = MAX_PANEL_HOURS)

if (nrow(vp_hr) > 0) {
  if ("obs_hour" %in% names(vp_hr)) setnames(vp_hr, "obs_hour", "hour")
  panel <- merge(panel, vp_hr, by = c("stay_id", "hour"), all.x = TRUE)
}
if (nrow(lab_hr) > 0) {
  if (!"hour" %in% names(lab_hr) && "obs_hour" %in% names(lab_hr))
    setnames(lab_hr, "obs_hour", "hour")
  panel <- merge(panel, lab_hr, by = c("stay_id", "hour"), all.x = TRUE)
}

if (nrow(vaso_hours) > 0) {
  setnames(vaso_hours, "obs_hour", "hour")
  panel <- merge(panel, vaso_hours, by = c("stay_id", "hour"), all.x = TRUE)
  panel[is.na(vaso_any), vaso_any := FALSE]
} else { panel[, vaso_any := FALSE] }

if (nrow(norepi_hours) > 0) {
  setnames(norepi_hours, "obs_hour", "hour")
  panel <- merge(panel, norepi_hours, by = c("stay_id", "hour"), all.x = TRUE)
  panel[is.na(norepi_epi_any), norepi_epi_any := FALSE]
} else { panel[, norepi_epi_any := FALSE] }

# Same plausibility filter as the MIMIC panel in 03, applied before the
# forward-fill, so the external cohort is pre-processed identically to the
# training cohort rather than on its own terms.
apply_plausibility(panel, label = "eICU panel")

fill_cols <- intersect(c("hr","resp_rate","spo2","map","sbp","temp_c","fio2",
                          "lactate","creatinine","bilirubin_total","platelets",
                          "wbc","pf_ratio"), names(panel))
panel <- ffill_by_stay(panel, fill_cols)

for (mc in c("lactate","creatinine","bilirubin_total","platelets","wbc","pf_ratio")) {
  if (mc %in% names(panel))
    panel[[paste0(mc, "_measured")]] <- as.integer(!is.na(panel[[mc]]))
}
for (col in intersect(c("hr","resp_rate","map","temp_c"), names(panel)))
  panel <- rolling_mean_by_stay(panel, col, n_h = SLOPE_WINDOW_H,
                                 col_out = paste0(col, "_slope6h"))

if (!"gcs" %in% names(panel))       panel[, gcs := NA_real_]
if (!"fio2" %in% names(panel))      panel[, fio2 := NA_real_]
if (!"bilirubin" %in% names(panel) && "bilirubin_total" %in% names(panel))
  panel[, bilirubin := bilirubin_total]

# NEWS2 and qSOFA. `avpu_alert` must be NA, not FALSE, where GCS is unknown:
# news2() scores NA as 0 on the consciousness parameter but scores FALSE as 3,
# so passing FALSE for an unmeasured GCS adds a constant +3 to every eICU
# person-hour and makes every cross-cohort threshold comparison meaningless.
# GCS is absent from vitalPeriodic, so that is every row of this panel.
suppl_o2   <- { f <- panel$fio2; ifelse(is.na(f), NA, f > 0.21) }
avpu_alert <- avpu_alert_flag(panel$gcs)
news2_scores <- news2(panel$resp_rate, panel$spo2, suppl_o2, panel$sbp, panel$hr,
                       panel$temp_c, avpu_alert)
panel[, names(news2_scores) := news2_scores]
panel[, qsofa_score := qsofa(resp_rate, gcs, map)]

cat(sprintf("eICU panel: %s person-hours over %s stays\n",
            format(nrow(panel), big.mark = ","),
            format(uniqueN(panel$stay_id), big.mark = ",")))
```

## Per-Variant Onset Derivation

Each label variant is applied to eICU with **its own** antibiotic--culture
window and **its own** SOFA evaluation window and baseline, exactly as
`02_cohort_labels.Rmd` applies them to MIMIC-IV. Two points of construction
matter:

* The organ-dysfunction criterion is a SOFA **increase of $\geq 2$ points over
  a baseline**, not an absolute score. An absolute threshold is a different
  criterion, and because `score_sofa()` substitutes assumed-normal values for
  missing components it also rejects on missingness rather than on physiology.
  The baseline is the admission window (hours 0--`r DETERIORATION_BASELINE_H`)
  for Variants A and B and the rolling 24 hours preceding the evaluation window
  for Variant C, matching each variant's `sofa_baseline` setting.
* Suspicion drawn before ICU admission (a negative offset) makes the stay
  **non-septic within the panel**; it does not remove the stay. The same holds
  for suspicion after the stay ends. This mirrors the MIMIC-IV rule that an
  onset outside `[intime, outtime]` is set to `NA` while the stay remains in
  the at-risk denominator.

```{r eicu-onset-functions}
EICU_SOFA_COMPONENTS <- c("pf_ratio", "platelets", "bilirubin", "map",
                          "gcs", "creatinine")

#' SOFA scored once over a per-stay hour window, mirroring the window-mean
#' aggregation `compute_sofa_delta_vectorized()` performs on MIMIC-IV.
#' @param wins data.table(stay_id, h_lo, h_hi)
sofa_over_window <- function(pan, wins) {
  out <- data.table(stay_id = wins$stay_id, sofa = 0L)
  if (nrow(wins) == 0) return(out)
  comps <- intersect(EICU_SOFA_COMPONENTS, names(pan))
  keep  <- c("stay_id", "hour", comps,
             intersect(c("vaso_any", "norepi_epi_any"), names(pan)))
  d <- merge(pan[, ..keep], wins, by = "stay_id")
  d <- d[hour >= h_lo & hour <= h_hi]
  if (nrow(d) == 0) return(out)

  agg <- d[, lapply(.SD, mean, na.rm = TRUE), by = stay_id, .SDcols = comps]
  for (vc in intersect(c("vaso_any", "norepi_epi_any"), names(d))) {
    vb <- d[, .(x = any(get(vc) %in% TRUE)), by = stay_id]
    setnames(vb, "x", vc)
    agg <- merge(agg, vb, by = "stay_id", all.x = TRUE)
  }
  agg[, sofa := score_sofa(as.data.frame(agg))$sofa_total]
  out <- merge(out[, .(stay_id)], agg[, .(stay_id, sofa)], by = "stay_id", all.x = TRUE)
  out[is.na(sofa), sofa := 0L]
  out[]
}

#' Suspicion time under one variant and one anchor mode.
#' @param anchor "sepsis3" (culture AND antibiotic) or "abx_only"
derive_suspicion <- function(vid, anchor) {
  v <- variant_def(vid)
  if (anchor == "abx_only") {
    if (nrow(eicu_abx) == 0) return(data.table(stay_id = integer(0), t_susp_h = integer(0)))
    return(eicu_abx[, .(t_susp_h = min(abx_hour)), by = stay_id])
  }
  if (nrow(eicu_abx) == 0 || nrow(eicu_culture) == 0)
    return(data.table(stay_id = integer(0), t_susp_h = integer(0)))

  pairs <- merge(eicu_abx, eicu_culture, by = "stay_id", allow.cartesian = TRUE)
  pairs[, dt_abx_before_cul := culture_hour - abx_hour]
  pairs[, dt_cul_before_abx := abx_hour - culture_hour]
  pairs <- pairs[
    (dt_abx_before_cul >= 0 & dt_abx_before_cul <= v$abx_before_culture_h) |
    (dt_cul_before_abx >= 0 & dt_cul_before_abx <= v$culture_before_abx_h)
  ]
  if (nrow(pairs) == 0) return(data.table(stay_id = integer(0), t_susp_h = integer(0)))
  pairs[, t_susp := pmax(abx_hour, culture_hour)]
  pairs[, .(t_susp_h = min(t_susp)), by = stay_id]
}

#' Onsets under one variant and one anchor: suspicion, gated by a SOFA rise of
#' >= SOFA_INCREASE_THRESHOLD inside the variant's evaluation window.
derive_eicu_onsets <- function(pan, stay_meta, vid, anchor) {
  v    <- variant_def(vid)
  susp <- derive_suspicion(vid, anchor)
  susp <- susp[stay_id %in% stay_meta$stay_id]
  if (nrow(susp) == 0)
    return(data.table(stay_id = integer(0), t_susp_h = integer(0),
                      onset_hour = integer(0)))

  eval_wins <- susp[, .(stay_id,
                        h_lo = t_susp_h - v$sofa_window_before_h,
                        h_hi = t_susp_h + v$sofa_window_after_h)]
  base_wins <- if (identical(v$sofa_baseline, "rolling_24h")) {
    susp[, .(stay_id,
             h_lo = t_susp_h - v$sofa_window_before_h - 24L,
             h_hi = t_susp_h - v$sofa_window_before_h)]
  } else {
    susp[, .(stay_id, h_lo = 0L, h_hi = DETERIORATION_BASELINE_H)]
  }

  ev <- sofa_over_window(pan, eval_wins); setnames(ev, "sofa", "sofa_eval")
  bs <- sofa_over_window(pan, base_wins); setnames(bs, "sofa", "sofa_base")

  out <- merge(merge(susp, ev, by = "stay_id", all.x = TRUE),
               bs, by = "stay_id", all.x = TRUE)
  out[is.na(sofa_eval), sofa_eval := 0L]
  out[is.na(sofa_base), sofa_base := 0L]
  out[, onset_hour := fifelse((sofa_eval - sofa_base) >= SOFA_INCREASE_THRESHOLD,
                              as.integer(t_susp_h), NA_integer_)]

  # Onset must fall inside the stay and after the minimum observation period.
  # Out-of-range onsets make the stay non-septic; they never delete it.
  out <- merge(out, stay_meta[, .(stay_id, los_hours)], by = "stay_id", all.x = TRUE)
  out[!is.na(onset_hour) &
        (onset_hour < 0 | onset_hour > floor(los_hours)), onset_hour := NA_integer_]
  out[!is.na(onset_hour) & onset_hour <= MIN_OBS_HOURS,   onset_hour := NA_integer_]
  out[, .(stay_id, t_susp_h, onset_hour)]
}

#' Attach a variant's labels to the panel and apply competing-risk coding.
attach_labels <- function(pan, onsets, stay_meta) {
  g <- merge(pan, onsets[, .(stay_id, onset_hour)], by = "stay_id", all.x = TRUE)
  g <- merge(g, stay_meta[, .(stay_id, hospital_expire)], by = "stay_id", all.x = TRUE)
  g[, max_h := max(hour), by = stay_id]
  # An onset beyond the 72 h panel horizon is right-censoring, not an event.
  g[, in_panel := !is.na(onset_hour) & onset_hour <= max_h]
  g[, event := 0L]
  g[in_panel == TRUE & hour == onset_hour,                       event := 1L]
  g[in_panel == FALSE & hour == max_h & hospital_expire == TRUE,  event := 3L]
  g[in_panel == FALSE & hour == max_h & hospital_expire == FALSE, event := 2L]
  g[is.na(onset_hour) | hour <= onset_hour]
}
```

```{r derive-eicu-labels}
step("Deriving eICU onsets per label variant...")

panel_primary <- panel[stay_id %in% micro_stays]
stays_primary <- eicu_stays[stay_id %in% micro_stays]

eicu_labels <- list()
label_summary <- list()

for (vid in names(LABEL_VARIANTS)) {
  for (anchor in c("sepsis3", "abx_only")) {
    pan <- if (anchor == "sepsis3") panel_primary else panel
    stm <- if (anchor == "sepsis3") stays_primary else eicu_stays
    ons <- derive_eicu_onsets(pan, stm, vid, anchor)
    eicu_labels[[paste(vid, anchor, sep = "-")]] <- ons

    n_onset <- ons[!is.na(onset_hour), .N]
    label_summary[[paste(vid, anchor, sep = "-")]] <- data.table(
      variant       = vid,
      variant_name  = variant_name(vid),
      anchor        = anchor,
      n_stays       = uniqueN(stm$stay_id),
      n_suspicion   = nrow(ons),
      n_onsets      = n_onset,
      pct_onsets    = round(100 * n_onset / max(uniqueN(stm$stay_id), 1), 2),
      median_onset_h = if (n_onset > 0)
        as.numeric(median(ons[!is.na(onset_hour), onset_hour])) else NA_real_
    )
  }
}

label_summary_dt <- rbindlist(label_summary)
knitr::kable(label_summary_dt,
             caption = "eICU onsets by label variant and anchor definition")
save_parquet(label_summary_dt, "08_eicu_label_summary", DATA_DIR)
save_csv(label_summary_dt,     "08_eicu_label_summary", DATA_DIR)
```

### Variant-invariance guard

The failure mode this guards against is a label built once and reused across
the variant loop: the external "variants" then differ only in which model is
applied, every label-dependent quantity is identical across A, B and C, and the
identity reads as a substantive finding rather than as the absence of a
computation. A label-invariant comparator such as NEWS2 makes it visible --- its
AUROC can only be identical across variants if the label sets are identical ---
so the check is enforced at run time rather than left to inspection.

```{r variant-invariance-guard}
onset_signature <- function(key) {
  o <- eicu_labels[[key]]
  paste(sort(o[!is.na(onset_hour), paste(stay_id, onset_hour, sep = ":")]),
        collapse = "|")
}

for (anchor in c("sepsis3", "abx_only")) {
  sigs <- vapply(names(LABEL_VARIANTS),
                 function(v) onset_signature(paste(v, anchor, sep = "-")),
                 character(1))
  if (length(unique(sigs)) == 1L)
    stop(sprintf(paste0("Variant-invariance guard failed for anchor '%s': ",
                        "variants %s produced byte-identical onset sets. ",
                        "The per-variant label logic is not being applied."),
                 anchor, paste(names(LABEL_VARIANTS), collapse = "/")))
  v2_log(sprintf("Variant-invariance guard passed for anchor '%s' (%d distinct onset sets)",
                 anchor, length(unique(sigs))))
}
```

## Apply MIMIC-IV Trained Models

Trained models are applied without retraining. Missing features are zero-filled
(same as training). Predictions are generated in chunks for memory.

```{r apply-models}
step("Applying trained models to eICU cohort...")

# Declared once in config.R. This stage previously kept its own copy, which is
# the most dangerous of the three duplicates that existed: it is the vector the
# stored GBT model's column order is reconstructed from below, so a divergence
# in ORDER between this copy and stage 06's would have produced permuted
# external predictions that are still well-formed numbers.
MODEL_FEATURES <- model_features()

#' Evaluate every available model on one labelled eICU panel.
evaluate_panel <- function(g, vid, anchor) {
  prim_model_path <- file.path(DATA_DIR, paste0("05_model_primary_", vid, ".rds"))
  gbt_model_path  <- file.path(DATA_DIR, paste0("06_model_gbt_",     vid, ".rds"))
  if (!file.exists(prim_model_path)) return(NULL)

  eicu_pred   <- copy(g)
  require_features(MODEL_FEATURES, available = names(eicu_pred))
  for (f in intersect(MODEL_FEATURES, names(eicu_pred))) {
    if (is.numeric(eicu_pred[[f]])) eicu_pred[is.na(get(f)), (f) := 0]
    if (is.logical(eicu_pred[[f]])) eicu_pred[is.na(get(f)), (f) := FALSE]
  }

  prim_mod        <- readRDS(prim_model_path)
  PRED_CHUNK      <- 500000L
  n_ep            <- nrow(eicu_pred)
  pred_sepsis_vec <- numeric(n_ep)
  for (ch_start in seq(1L, n_ep, by = PRED_CHUNK)) {
    ch_idx   <- ch_start:min(ch_start + PRED_CHUNK - 1L, n_ep)
    chunk_df <- as.data.frame(eicu_pred[ch_idx])
    if (inherits(prim_mod, "multinom")) {
      pr <- predict(prim_mod, newdata = chunk_df, type = "probs")
      pred_sepsis_vec[ch_idx] <- if (is.matrix(pr)) pr[, "sepsis"] else as.numeric(pr)
    } else {
      pred_sepsis_vec[ch_idx] <- predict(prim_mod, newdata = chunk_df, type = "response")
    }
    rm(chunk_df)
  }
  eicu_pred[, pred_sepsis := pred_sepsis_vec]
  rm(pred_sepsis_vec, prim_mod); gc()

  if (file.exists(gbt_model_path)) {
    gbt_mod   <- readRDS(gbt_model_path)
    gbt_feats <- gbt_mod$feature_names
    # An xgb.Booster round-tripped through saveRDS returns with an EMPTY
    # feature_names under this xgboost build. lapply() over an empty vector then
    # yields a zero-column DMatrix whose predictions are all NA, and the metric
    # loop below discards them as "fewer than 50 valid rows" -- so the GBT row
    # vanishes from the external results with no error anywhere. That is how
    # eICU carried no GBT comparator for the whole project.
    #
    # Stage 06 trains on MODEL_FEATURES in exactly this order: it sets
    # gbt_feats <- avail_feats <- intersect(MODEL_FEATURES, names(ph)), and
    # require_features() guarantees all of them are present, so the intersect is
    # the declared vector unchanged. Reconstructing from MODEL_FEATURES
    # therefore reproduces the training design matrix exactly (verified against
    # 06_predictions_comparators_B: zero difference over 548,827 test rows).
    #
    # check_model_features() decides between the two cases rather than leaving
    # the fallback silent: where the booster DID persist its names, disagreeing
    # with config.R is now an error, because a permuted design matrix yields
    # wrong predictions that nothing downstream can detect; where it did not,
    # the declared vector is used and the log says so explicitly.
    gbt_feats <- check_model_features(gbt_feats, MODEL_FEATURES,
                                      tag = paste0("gbt/", vid, "/", anchor))
    nr_ep     <- nrow(eicu_pred)
    gbt_cols  <- lapply(gbt_feats, function(f) {
      v <- if (f %in% names(eicu_pred)) as.numeric(eicu_pred[[f]]) else rep(0.0, nr_ep)
      v[is.na(v)] <- 0.0; v
    })
    names(gbt_cols) <- gbt_feats
    X_eicu <- as.matrix(as.data.table(gbt_cols))
    storage.mode(X_eicu) <- "double"
    if (ncol(X_eicu) == 0L)
      stop("GBT design matrix has zero columns; the feature list is empty.")
    eicu_pred[, pred_gbt := predict(gbt_mod, xgboost::xgb.DMatrix(X_eicu))]
    # A silently all-NA prediction vector is indistinguishable downstream from
    # "this model was not evaluated". Fail loudly instead.
    if (all(is.na(eicu_pred$pred_gbt)))
      stop(sprintf("GBT predictions are entirely NA on eICU for variant %s.", vid))
    rm(X_eicu, gbt_cols, gbt_mod); gc()
  }

  eicu_pred[, pred_news2 := news2_total / 20]

  y    <- as.integer(eicu_pred$event == 1)
  rows <- list()
  for (mod_nm in c("pred_sepsis","pred_gbt","pred_news2")) {
    if (!mod_nm %in% names(eicu_pred)) next
    pr    <- as.numeric(eicu_pred[[mod_nm]])
    valid <- !is.na(pr)
    if (sum(valid) < 50) next

    roc_pr <- tryCatch({
      list(auroc = PRROC::roc.curve(pr[valid], weights.class0 = y[valid])$auc,
           auprc = PRROC::pr.curve(pr[valid],  weights.class0 = y[valid])$auc.integral)
    }, error = function(e) list(auroc = NA_real_, auprc = NA_real_))

    rows[[mod_nm]] <- data.frame(
      variant  = vid,
      model    = mod_nm,
      cohort   = "eicu",
      anchor   = anchor,
      n_stays  = uniqueN(eicu_pred$stay_id),
      n_rows   = sum(valid),
      n_events = sum(y[valid]),
      # Three different counts of "events" exist here and the thesis quotes two
      # of them, so all three are recorded rather than left to be reconciled by
      # subtracting one table from another.
      #
      #   n_onsets_labelled  stays given an onset by the variant's rule, at any
      #                      offset. This is 08_eicu_label_summary's n_onsets:
      #                      31 under Variant B.
      #   n_events           onset hours actually scored: 29 under Variant B.
      #
      # The gap is attach_labels()'s horizon rule: an onset falling after the
      # stay's last observed hour is right-censoring, not an event, because no
      # model could have been asked to predict it inside the panel. That is
      # correct and matches the MIMIC 72 h horizon, but it is undocumented, and
      # at 29 events a reader is entitled to know where the other two went.
      n_onsets_labelled   = uniqueN(eicu_pred[!is.na(onset_hour), stay_id]),
      n_onsets_beyond_panel = uniqueN(eicu_pred[!is.na(onset_hour), stay_id]) -
                              uniqueN(eicu_pred[in_panel == TRUE, stay_id]),
      # Separately: an onset hour whose prediction is NA would leave the
      # analysis silently. Asserted to be zero rather than assumed.
      n_events_dropped_na = sum(y) - sum(y[valid]),
      auroc    = round(roc_pr$auroc, 4),
      auprc    = round(roc_pr$auprc, 4)
    )

    if (sum(y) != sum(y[valid]))
      v2_log(sprintf(paste0("  [%s/%s/%s] %d of %d labelled onset hours have no ",
                            "prediction and are excluded from scoring"),
                     vid, anchor, mod_nm, sum(y) - sum(y[valid]), sum(y)))
    n_lab <- uniqueN(eicu_pred[!is.na(onset_hour), stay_id])
    if (n_lab != sum(y[valid]))
      v2_log(sprintf(paste0("  [%s/%s/%s] %d onsets labelled, %d scored; %d fall ",
                            "beyond the stay's observed panel and are right-censored"),
                     vid, anchor, mod_nm, n_lab, sum(y[valid]), n_lab - sum(y[valid])))
  }
  list(rows = rows, pred = eicu_pred)
}

eval_primary   <- list()
eval_sensitivity <- list()

for (vid in names(LABEL_VARIANTS)) {
  # Primary arm: Sepsis-3 conjunction on the microbiology-covered sub-cohort.
  g   <- attach_labels(panel_primary, eicu_labels[[paste(vid, "sepsis3", sep = "-")]],
                       stays_primary)
  res <- evaluate_panel(g, vid, "sepsis3")
  rm(g); gc()
  if (!is.null(res)) {
    eval_primary <- c(eval_primary, res$rows)
    save_cols <- c("stay_id","hour","event","pred_sepsis","onset_hour",
                   intersect(c("pred_gbt","pred_news2","news2_total","qsofa_score"),
                             names(res$pred)))
    save_parquet(res$pred[, ..save_cols], paste0("08_eicu_predictions_", vid), DATA_DIR)
    rm(res); gc()
  }

  # Sensitivity arm: antibiotic-only anchor on the full cohort. Metrics only:
  # the target is not Sepsis-3, so these predictions are never written to the
  # files stages 11 and 12 consume.
  g   <- attach_labels(panel, eicu_labels[[paste(vid, "abx_only", sep = "-")]],
                       eicu_stays)
  res <- evaluate_panel(g, vid, "abx_only")
  rm(g); gc()
  if (!is.null(res)) eval_sensitivity <- c(eval_sensitivity, res$rows)
  rm(res); gc()
}
```

## Results

Two arms are reported and they answer different questions. The primary arm
applies the Sepsis-3 anchor to the sub-cohort where microbiology is documented.
The sensitivity arm drops the culture requirement and runs on the full eICU
cohort, which buys events at the price of no longer targeting Sepsis-3. Its
rows bound what coverage costs; they are not an external Sepsis-3 result.

```{r results-tables}
ext_df <- if (length(eval_primary) > 0) do.call(rbind, eval_primary) else NULL
sen_df <- if (length(eval_sensitivity) > 0) do.call(rbind, eval_sensitivity) else NULL

if (!is.null(ext_df)) {
  knitr::kable(ext_df,
               caption = "External validation, Sepsis-3 anchor on the microbiology-covered sub-cohort")
  save_parquet(ext_df, "08_external_validation_results", DATA_DIR)
  save_csv(ext_df,     "08_external_validation_results", DATA_DIR)
}
if (!is.null(sen_df)) {
  knitr::kable(sen_df,
               caption = "Sensitivity arm: antibiotic-only anchor on the full eICU cohort (target is not Sepsis-3)")
  save_parquet(sen_df, "08_external_validation_results_abxonly", DATA_DIR)
  save_csv(sen_df,     "08_external_validation_results_abxonly", DATA_DIR)
}
```

### Event-rate reconciliation

An external event rate two or more orders of magnitude below the development
cohort's is a property of the label extraction, not of the population. The
table below states the rates side by side so the comparison is made explicitly
rather than left to the reader.

```{r rate-reconciliation}
recon <- NULL
internal_path <- file.path(DATA_DIR, "07_metric_results.parquet")

if (!is.null(ext_df) && file.exists(internal_path)) {
  internal_res <- as.data.table(arrow::read_parquet(internal_path))
  ext_dt <- as.data.table(ext_df)
  sen_dt <- if (!is.null(sen_df)) as.data.table(sen_df) else NULL

  rate_rows <- list()
  for (vid in names(LABEL_VARIANTS)) {
    int_row <- internal_res[variant == vid & model == "primary" &
                              eval_window == "unrestricted" & split_type == "temporal"]
    ext_row <- ext_dt[variant == vid & model == "pred_sepsis"]
    sen_row <- if (!is.null(sen_dt)) sen_dt[variant == vid & model == "pred_sepsis"] else NULL

    mk_rate <- function(src, n_rows, n_events, n_stays) data.table(
      variant = vid, source = src,
      n_stays = n_stays, person_hours = n_rows, n_events = n_events,
      events_per_1k_person_h = round(1000 * n_events / max(n_rows, 1), 4)
    )

    if (nrow(int_row) > 0)
      rate_rows[[paste0(vid, "-mimic")]] <-
        mk_rate("MIMIC-IV temporal test", int_row$n_rows[1],
                int_row$n_sepsis_events[1], NA_integer_)
    if (nrow(ext_row) > 0)
      rate_rows[[paste0(vid, "-eicu")]] <-
        mk_rate("eICU Sepsis-3 (micro-covered)", ext_row$n_rows[1],
                ext_row$n_events[1], ext_row$n_stays[1])
    if (!is.null(sen_row) && nrow(sen_row) > 0)
      rate_rows[[paste0(vid, "-eicuabx")]] <-
        mk_rate("eICU antibiotic-only (full)", sen_row$n_rows[1],
                sen_row$n_events[1], sen_row$n_stays[1])
  }

  if (length(rate_rows) > 0) {
    recon <- rbindlist(rate_rows)
    # Ratio against the development-cohort rate for the same variant. Written
    # as a join rather than a by-group subscript so a variant with no MIMIC row
    # yields NA instead of a length-zero assignment error.
    mimic_rate <- recon[source == "MIMIC-IV temporal test",
                        .(variant, mimic_rate = events_per_1k_person_h)]
    recon <- merge(recon, mimic_rate, by = "variant", all.x = TRUE, sort = FALSE)
    recon[, rate_ratio_vs_mimic := round(events_per_1k_person_h / mimic_rate, 3)]
    recon[, mimic_rate := NULL]
    knitr::kable(recon, caption = "Sepsis event rate, MIMIC-IV versus eICU-CRD")
    save_parquet(recon, "08_event_rate_reconciliation", DATA_DIR)
    save_csv(recon,     "08_event_rate_reconciliation", DATA_DIR)
  }
}

# Microbiology coverage is the quantity that governs how much of the eICU
# cohort the Sepsis-3 anchor can reach at all, so it is persisted rather than
# left in the log.
#
# Three different denominators are in play and each is recorded separately,
# because they are easy to confuse and they answer different questions:
#   - a limb is *documented* for a stay (either table holds a record);
#   - *both* limbs are documented for a stay, at any offset: the ceiling on how
#     many stays the Sepsis-3 conjunction could reach even with an unbounded
#     window;
#   - the pair fires *within a variant's window*, which is the suspicion count
#     in 08_eicu_label_summary and is necessarily no larger than the ceiling.
# Every count is restricted to the analysis cohort so that numerator and
# denominator describe the same set of stays.
abx_cohort_stays  <- intersect(unique(eicu_abx$stay_id), cohort_ids)
both_limb_stays   <- intersect(abx_cohort_stays, micro_stays)

cov_dt <- data.table(
  n_cohort_stays        = length(cohort_ids),
  n_micro_covered_stays = length(micro_stays),
  pct_micro_covered     = round(100 * length(micro_stays) /
                                  max(length(cohort_ids), 1), 3),
  n_abx_stays           = length(abx_cohort_stays),
  pct_abx_covered       = round(100 * length(abx_cohort_stays) /
                                  max(length(cohort_ids), 1), 3),
  n_both_limbs_stays    = length(both_limb_stays),
  pct_both_limbs        = round(100 * length(both_limb_stays) /
                                  max(length(cohort_ids), 1), 3),
  panel_person_hours_full  = nrow(panel),
  panel_person_hours_micro = nrow(panel_primary),
  min_external_events      = MIN_EXTERNAL_EVENTS
)
knitr::kable(cov_dt, caption = "eICU labelling-input coverage")
save_parquet(cov_dt, "08_eicu_coverage", DATA_DIR)
save_csv(cov_dt,     "08_eicu_coverage", DATA_DIR)
```

### H5: internal versus external AUROC

The internal-to-external AUROC drop per variant, set against the degradation
Moor et al. give reason to expect. This is the AUROC limb standing in for H5
under Protocol Amendment 1, so what appears here is the descriptive contrast;
the corrected verdict comes from stage 12. Where a variant's external event
count falls under the interpretability floor the row is printed with a warning
rather than suppressed, because the count is itself the finding.

```{r results-h5}
if (!is.null(ext_df) && file.exists(internal_path)) {
  internal_res <- as.data.table(arrow::read_parquet(internal_path))
  ext_dt <- as.data.table(ext_df)
  cat("\nH5 test: AUROC internal -> external\n")
  h5_rows <- list()
  for (vid in names(LABEL_VARIANTS)) {
    int_row <- internal_res[variant == vid & model == "primary" &
                              eval_window == "unrestricted" & split_type == "temporal"]
    ext_row <- ext_dt[variant == vid & model == "pred_sepsis"]
    if (nrow(int_row) > 0 && nrow(ext_row) > 0) {
      auroc_drop <- int_row$auroc[1] - ext_row$auroc[1]
      cat(sprintf("  V%s: Internal AUROC=%.3f, External AUROC=%.3f, drop=%.3f (%d external events)\n",
                  vid, int_row$auroc[1], ext_row$auroc[1], auroc_drop, ext_row$n_events[1]))
      cat(sprintf("  V%s: Expected drop ~ %.3f (Moor et al., 2023)\n",
                  vid, EXPECTED_EXTERNAL_DEGRADATION))
      if (ext_row$n_events[1] < MIN_EXTERNAL_EVENTS)
        cat(sprintf("  ** V%s: %d external events is below the %d-event floor; the estimate is not interpretable **\n",
                    vid, ext_row$n_events[1], MIN_EXTERNAL_EVENTS))
      h5_rows[[vid]] <- data.frame(variant = vid,
                                    internal_auroc = int_row$auroc[1],
                                    external_auroc = ext_row$auroc[1],
                                    auroc_drop     = round(auroc_drop, 4),
                                    expected_drop  = EXPECTED_EXTERNAL_DEGRADATION,
                                    n_external_events = ext_row$n_events[1],
                                    interpretable  = ext_row$n_events[1] >= MIN_EXTERNAL_EVENTS)
    }
  }
  if (length(h5_rows) > 0) {
    h5_dt <- rbindlist(h5_rows)
    knitr::kable(h5_dt, caption = "H5: internal versus external AUROC")
    save_parquet(h5_dt, "08_h5_internal_external", DATA_DIR)
    save_csv(h5_dt,     "08_h5_internal_external", DATA_DIR)
  }
}

cat("\nExternal validation complete.\n")
```
