---
title: "09 Variance Decomposition - Attribution of Performance Variation"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

This notebook produces the variance decomposition, the study's summary of how
far reported performance moves with each analytical decision. Published
sepsis-prediction studies differ in several such decisions at once (Cohen et
al., 2024), so a difference in their reported performance cannot be traced to
any one decision. Here one factor is varied at a time while the others are held
at reference levels: Variant B, the primary model, the temporal split and the
unrestricted window. "Variance decomposition" is the pre-registered name of the
analysis. The quantities it reports are ranges and differences in a metric,
not variance components, so they do not partition a variance and cannot be
summed.

The decomposition considers five factors.

| Factor | Levels compared |
|--------|-----------------|
| Model class | Primary model (multinomial discrete-time competing-risks hazard model), gradient-boosted trees fitted with XGBoost (GBT), Cox model, NEWS2, qSOFA, SIRS |
| Label operationalisation | Variants A (Narrow), B (Seymour-Standard) and C (Liberal) |
| Validation split | Temporal, random |
| Prediction-time anchoring | Unrestricted window (all test hours), treatment-anchored window (hours before the first antibiotic or culture) |
| Evaluation metric | Pre-registered as the internal-to-external decline in the Utility Score against the decline in AUROC. The row computed here is the AUROC part only: internal (MIMIC-IV) against external (eICU) AUROC |

Six pre-registered hypotheses structure the interpretation. Where a hypothesis
is a difference, the direction of subtraction is stated.

| Hypothesis | Claim | Threshold |
|----|-------|-----------|
| H1 | The AUROC spread across label variants is at least the AUROC spread across model classes | label spread $\geq$ model spread |
| H2 | At least one covariate changes sign, or changes in magnitude by a factor of 2 or more, across label variants | none |
| H3 | Unrestricted AUROC minus treatment-anchored AUROC is at least 0.03 | $\Delta \geq 0.03$ |
| H4 | Random-split AUROC minus temporal-split AUROC is at least 0.02 | $\Delta \geq 0.02$ |
| H5 | The proportional internal-to-external decline is substantially larger for the Utility Score than for AUROC | none |
| H6 | The primary model is not inferior to the GBT by more than 0.02 AUROC | $|\Delta| \leq 0.02$ |

A spread is the maximum minus the minimum of a metric across a factor's
levels, so it is never negative. H3, H4 and the metric row are signed
differences, and H6 is an absolute difference. Every number in this notebook is
a point estimate. The bootstrap intervals and the Holm-corrected confirmatory
results of Protocol Amendment 1 are produced in `12_inference.Rmd`, and they
take precedence over everything here.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
})

source("config.R")
source("utils.R")

step("=== Phase 6a: Variance Decomposition ===")
```

## Load Metric Results

The metric results of `07_metrics_suite.Rmd`, the external-validation results
of `08_external_validation.Rmd` and the coefficient table of
`05_primary_model.Rmd` are loaded. The table below lists every temporal-split
result for Variants A--C as stage 07 wrote it.

Some entries in it need explanation. The Utility Score, alert burden and lead
time are read at the default threshold of 0.5. The primary model raises no
alert at that threshold, so its Utility Score is exactly 0, its alert burden is
infinite (Inf, because there is no true alert to divide by) and its lead time
is missing (NA). The treatment-anchored rows contain no onsets under Variants
A--C, for the reason given in the proposition below, so their AUROC and AUPRC
are also NA. The column `utility_max` is the maximum Utility Score over the
threshold grid of stage 07. It is chosen on the test set and is therefore an
optimistic ceiling. A small negative value means that no threshold raising at
least one alert did better than issuing no alerts.

```{r load-metrics}
metric_path <- file.path(DATA_DIR, "07_metric_results.parquet")
ext_path    <- file.path(DATA_DIR, "08_external_validation_results.parquet")
coef_path   <- file.path(DATA_DIR, "05_coef_all_variants.parquet")

if (!file.exists(metric_path))
  stop("07_metric_results.parquet not found. Run 07_metrics_suite.Rmd first.")

metrics    <- as.data.table(arrow::read_parquet(metric_path))
ext_res    <- if (file.exists(ext_path))  as.data.table(arrow::read_parquet(ext_path))  else NULL
coef_table <- if (file.exists(coef_path)) as.data.table(arrow::read_parquet(coef_path)) else NULL

cat(sprintf("Metric results loaded: %d rows\n", nrow(metrics)))
knitr::kable(metrics[, .(variant, model, eval_window, auroc, auprc,
                           utility_normalised, utility_max, alert_burden,
                           lead_time_h)],
             caption = "All evaluation results (from 07\\_metrics\\_suite.Rmd)")
```

## Variance by Factor

Each row of the decomposition is computed as follows, with the factors not
named held at their reference levels.

- **Model class:** the AUROC spread across the six models under Variant B. The
  Utility Score spread uses each model's maximum over the threshold grid, not
  its score at 0.5. At this event rate the 0.5 threshold separates models by
  how far their probabilities are inflated, so a spread taken there would
  measure calibration rather than model class.
- **Label operationalisation (H1):** the spread of the primary model's AUROC
  across Variants A--C. The Utility Score spread again uses the maxima.
- **Prediction-time anchoring (H3):** unrestricted minus treatment-anchored
  AUROC, for the primary model under Variant B.
- **Validation split (H4):** random-split minus temporal-split AUROC, for the
  primary model under Variant B. A negative value means the random split gave
  the lower AUROC.
- **Evaluation metric (H5):** internal minus external AUROC, for the primary
  model under Variant B, with the external value taken from the eICU Sepsis-3
  arm. A positive value means AUROC fell on external data. The Utility Score
  part of H5 is not computed here.
- **Model-class null (H6):** the absolute difference between the AUROCs of the
  primary model and the GBT under Variant B.

```{r variance-by-factor}
# Factor 1: MODEL CLASS - label=B, temporal, unrestricted
model_class_rows   <- metrics[variant == "B" & eval_window == "unrestricted"]
model_spread_auroc   <- if (nrow(model_class_rows) > 0) diff(range(model_class_rows$auroc, na.rm = TRUE)) else NA
# Utility spread is taken over the swept maximum, not over the 0.5 cut-off. At
# this event rate the cut-off separates models by how far their probabilities
# happen to be inflated rather than by anything clinical, so a spread computed
# there measures calibration, not the factor under study.
model_spread_utility <- if (nrow(model_class_rows) > 0) diff(range(model_class_rows$utility_max, na.rm = TRUE)) else NA

# Factor 2: LABEL VARIANT - model=primary, temporal, unrestricted
label_rows         <- metrics[model == "primary" & eval_window == "unrestricted"]
label_spread_auroc   <- if (nrow(label_rows) > 0) diff(range(label_rows$auroc, na.rm = TRUE)) else NA
label_spread_utility <- if (nrow(label_rows) > 0) diff(range(label_rows$utility_max, na.rm = TRUE)) else NA

# Factor 3: PREDICTION-TIME ANCHORING - model=primary, variant=B, temporal
anchor_rows_B  <- metrics[model == "primary" & variant == "B"]
anchor_unrestr <- anchor_rows_B[eval_window == "unrestricted",         auroc]
anchor_treatm  <- anchor_rows_B[eval_window == "treatment_anchored",   auroc]
h3_delta_auroc <- if (length(anchor_unrestr) > 0 && length(anchor_treatm) > 0)
                    anchor_unrestr - anchor_treatm else NA

# Factor 4: VALIDATION SPLIT (temporal vs random)
random_path <- file.path(DATA_DIR, "07_metric_results_random.parquet")
if (file.exists(random_path)) {
  metrics_random <- as.data.table(arrow::read_parquet(random_path))
  temporal_auroc <- metrics[model == "primary" & variant == "B" &
                               eval_window == "unrestricted", auroc]
  random_auroc   <- metrics_random[model == "primary" & variant == "B" &
                                    eval_window == "unrestricted", auroc]
  h4_delta_auroc <- if (length(random_auroc) > 0 && length(temporal_auroc) > 0)
                      random_auroc - temporal_auroc else NA
} else {
  h4_delta_auroc <- NA
  cat("Random split results not found; H4 cannot be tested.\n")
}

# Factor 5: METRIC
internal_primary_B <- metrics[model == "primary" & variant == "B" & eval_window == "unrestricted"]
h5_internal_auroc  <- if (nrow(internal_primary_B) > 0) internal_primary_B$auroc else NA
h5_internal_util   <- if (nrow(internal_primary_B) > 0) internal_primary_B$utility_normalised else NA
if (!is.null(ext_res)) {
  ext_primary_B <- ext_res[variant == "B" & model == "pred_sepsis"]
  h5_ext_auroc  <- if (nrow(ext_primary_B) > 0) ext_primary_B$auroc else NA
} else { h5_ext_auroc <- NA }

# Factor 6: MODEL CLASS NULL (H6)
if (nrow(model_class_rows) > 0) {
  prim_auroc <- model_class_rows[model == "primary", auroc]
  gbt_auroc  <- model_class_rows[model == "gbt",     auroc]
  h6_delta   <- if (length(prim_auroc) > 0 && length(gbt_auroc) > 0)
                  abs(prim_auroc - gbt_auroc) else NA
} else { h6_delta <- NA }
```

## The Variance Decomposition Table

The table below is the variance decomposition. Each row varies one factor and
names, in its factor column, the levels at which the other factors are held.
The column `spread_auroc` holds the quantity defined for that row in the
previous section. Despite the column name, not every row is a range: the split
and metric rows are signed differences, and the H6 row is an absolute
difference. The `hypothesis` and `h_threshold` columns give the pre-registered
hypothesis each row bears on. The model-class row is marked "-" because it
tests no hypothesis on its own; it enters H1 as the comparator for the label
spread. The anchoring row is NA because the treatment-anchored AUROC is
undefined under Variant B.

The rows are intended to be compared with one another, to show which decision
moves reported AUROC the most. On these point estimates, model class moves it
furthest, about twelve times as far as the label. The model-class spread runs
from the rule-based scores to the primary model, however. The gap between the
two fitted flexible models, the primary model and the GBT, is far smaller and
is given by the H6 row.

```{r decomp-table}
decomp_table <- data.frame(
  factor = c(
    "Model class (label=B, temporal, unrestricted)",
    "Label operationalisation (model=primary, temporal, unrestricted)",
    "Prediction-time anchoring (model=primary, variant=B, temporal)",
    "Validation split (temporal vs random, model=primary, variant=B)",
    "Metric (AUROC vs Utility, internal→external, model=primary, variant=B)",
    "Model-class null |primary − GBT| (variant=B, temporal, unrestricted)"
  ),
  spread_auroc = round(c(model_spread_auroc, label_spread_auroc,
                          h3_delta_auroc, h4_delta_auroc,
                          if (!is.na(h5_internal_auroc) && !is.na(h5_ext_auroc))
                            h5_internal_auroc - h5_ext_auroc else NA,
                          h6_delta), 4),
  spread_utility = round(c(model_spread_utility, label_spread_utility,
                             NA, NA, NA, NA), 4),
  hypothesis  = c("-", "H1", "H3", "H4", "H5", "H6"),
  h_threshold = c("-", "label ≥ model", "Δ ≥ 0.03", "Δ ≥ 0.02",
                   "utility drop > AUROC drop", "|Δ| ≤ 0.02")
)

knitr::kable(decomp_table, caption = "THE VARIANCE DECOMPOSITION TABLE")
save_parquet(decomp_table, "09_variance_decomposition_table", DATA_DIR)
```

## Point-Estimate Checks Against the Pre-Registered Thresholds

Each hypothesis that has a point estimate here is compared with its threshold.
These are descriptive checks, not statistical tests. In the printed output,
"CONFIRMED" means that the threshold is met and "NOT confirmed" that it is not.
For H6, "NOT rejected" means that the absolute difference lies within the 0.02
margin, which on a point estimate does not establish equivalence. The verdicts
of record come from `12_inference.Rmd`, which attaches bootstrap intervals and
applies the Holm correction. No later stage reads the file written here.

Four hypotheses are checked. H3 is omitted because its AUROC difference is
undefined under Variant B. H5 is omitted because the pre-registration gives it
no numerical threshold; its contrast is evaluated in
`08_external_validation.Rmd` and `12_inference.Rmd`.

- **H1:** the label spread is compared with the model-class spread.
- **H2:** the check covers every coefficient of the sepsis outcome present
  under all three of Variants A--C, except the intercept and the spline terms
  in time since admission, and applies the sign criterion only. The
  pre-registered H2 also counts a change in magnitude by a factor of 2 or more,
  and the full count is reported in `12_inference.Rmd`. In the output,
  `bilirubin_total` is total bilirubin and `vaso_anyTRUE` is any vasopressor
  exposure. A coefficient close to zero can change sign without meaningful
  instability. Protocol Amendment 7 therefore compares each change with its
  estimation noise, and both sign reversals found here are smaller than the
  pooled standard error of the estimates they compare.
- **H4:** the difference is negative, so the random split gave the lower
  AUROC, the opposite of the direction H4 predicted. "Split optimism" in the
  output is the hypothesis's pre-registered name.
- **H6:** the check uses the absolute difference. `12_inference.Rmd` uses the
  signed difference, GBT minus primary, and the conclusion drawn there from the
  interval is one-sided non-inferiority, not equivalence.

```{r hypothesis-verdicts}
step("Hypothesis test verdicts...")
verdicts <- list()

# H1
if (!is.na(label_spread_auroc) && !is.na(model_spread_auroc)) {
  h1_result <- label_spread_auroc >= model_spread_auroc
  cat(sprintf("H1 (Label dominance): label_spread=%.3f, model_spread=%.3f → %s\n",
              label_spread_auroc, model_spread_auroc,
              if (h1_result) "CONFIRMED" else "NOT confirmed"))
  verdicts[["H1"]] <- data.frame(hypothesis = "H1", finding = "Label dominance",
                                  threshold = "label >= model",
                                  observed  = sprintf("label=%.3f, model=%.3f",
                                                      label_spread_auroc, model_spread_auroc),
                                  confirmed = h1_result)
}

# H2
if (!is.null(coef_table) && "term" %in% names(coef_table)) {
  coef_sep <- coef_table[outcome == "sepsis" &
                           !grepl("^ns\\(|^\\(Intercept\\)", as.character(term))]
  if (nrow(coef_sep) > 0) {
    common_terms <- Reduce(intersect, lapply(c("A","B","C"), function(v)
      unique(as.character(coef_sep[variant == v, term]))))
    sign_changes <- character(0)
    for (tm in common_terms) {
      vals  <- sapply(c("A","B","C"), function(v) {
        r <- coef_sep[variant == v & as.character(term) == tm, coef]
        if (length(r) == 0) NA_real_ else as.numeric(r[1])
      })
      signs <- sign(vals[!is.na(vals)])
      if (length(signs) >= 2 && diff(range(signs)) != 0)
        sign_changes <- c(sign_changes, tm)
    }
    h2_result <- length(sign_changes) >= 1
    cat(sprintf("H2 (Coefficient instability): %d sign-changes: %s → %s\n",
                length(sign_changes), paste(head(sign_changes, 3), collapse = ", "),
                if (h2_result) "CONFIRMED" else "NOT confirmed"))
    verdicts[["H2"]] <- data.frame(hypothesis = "H2", finding = "Coefficient instability",
                                    threshold = ">=1 covariate changes sign",
                                    observed  = paste(sign_changes, collapse = "; "),
                                    confirmed = h2_result)
  }
}

# H3
if (!is.na(h3_delta_auroc)) {
  h3_result <- h3_delta_auroc >= 0.03
  cat(sprintf("H3 (Leakage): unrestricted − anchored AUROC Δ=%.3f → %s\n",
              h3_delta_auroc, if (h3_result) "CONFIRMED" else "NOT confirmed (< 0.03)"))
  verdicts[["H3"]] <- data.frame(hypothesis = "H3", finding = "Leakage via treatment anchoring",
                                  threshold = "Δ ≥ 0.03",
                                  observed  = sprintf("Δ=%.3f", h3_delta_auroc),
                                  confirmed = h3_result)
}

# H4
if (!is.na(h4_delta_auroc)) {
  h4_result <- h4_delta_auroc >= 0.02
  cat(sprintf("H4 (Split optimism): random − temporal AUROC Δ=%.3f → %s\n",
              h4_delta_auroc, if (h4_result) "CONFIRMED" else "NOT confirmed (< 0.02)"))
  verdicts[["H4"]] <- data.frame(hypothesis = "H4", finding = "Split optimism",
                                  threshold = "Δ ≥ 0.02",
                                  observed  = sprintf("Δ=%.3f", h4_delta_auroc),
                                  confirmed = h4_result)
}

# H6
if (!is.na(h6_delta)) {
  h6_result <- h6_delta <= 0.02
  cat(sprintf("H6 (Model-class null): |primary − GBT| AUROC=%.3f → %s\n",
              h6_delta, if (h6_result) "NOT rejected (interpretable ≈ GBT)"
                        else "REJECTED (GBT wins by > 0.02)"))
  verdicts[["H6"]] <- data.frame(hypothesis = "H6", finding = "Model-class null",
                                  threshold = "|Δ| ≤ 0.02",
                                  observed  = sprintf("Δ=%.3f", h6_delta),
                                  confirmed = h6_result)
}

if (length(verdicts) > 0) {
  verdicts_df <- do.call(rbind, verdicts)
  # These are point-estimate pre-checks computed before any uncertainty exists.
  # Protocol Amendment 1 made 12_inference.Rmd the adjudicating layer, and the
  # two disagree by construction: this table marks H2 and H6 "confirmed = TRUE"
  # where 12_confirmatory_tests records "Descriptive" and "H6 not rejected",
  # and it holds four hypotheses where the family has six. No stage and no
  # macro reads it. Labelled rather than left to be found by an examiner
  # reproducing the run.
  verdicts_df$status <- "point estimate only; superseded by 12_confirmatory_tests (Amendment 1)"
  knitr::kable(verdicts_df, caption = paste("Hypothesis pre-checks on point estimates.",
    "The adjudicated verdicts, with intervals and Holm correction, are in",
    "12_confirmatory_tests (Protocol Amendment 1)."))
  cat(paste0("\nNOTE: 09_hypothesis_verdicts holds point-estimate pre-checks only.\n",
             "      The confirmatory verdicts are 12_confirmatory_tests.\n"))
  save_parquet(verdicts_df, "09_hypothesis_verdicts", DATA_DIR)
}
```

## Variance Decomposition Plots

The two figures show the AUROC values behind the first two rows of the
decomposition. The first gives the primary model's AUROC under each label
variant, with a dashed line at its value under Variant B. The second gives each
model's AUROC under Variant B. H1 compares the range across the bars of the
first figure with the range across the bars of the second.

```{r variance-plots, fig.cap="Variance decomposition: label-variant and model-class effects on AUROC.", fig.width=6, fig.height=4}
if (nrow(metrics[!is.na(auroc)]) > 0) {
  # Label-variant effect
  tryCatch({
    p_label <- ggplot(metrics[model == "primary" & eval_window == "unrestricted"],
                       aes(x = variant, y = auroc, fill = variant)) +
      geom_col() +
      geom_hline(yintercept = if (!is.na(h5_internal_auroc)) h5_internal_auroc else 0.75,
                 linetype = "dashed") +
      labs(title    = "Label-variant effect on AUROC (primary model, temporal split)",
           subtitle = "H1: label spread should be ≥ model-class spread",
           x = "Label Variant", y = "AUROC") +
      theme_bw() + theme(legend.position = "none")
    print(p_label)
    ggsave(file.path(FIGURES_DIR, "09_label_variance_auroc.png"), p_label,
           width = 6, height = 4, dpi = 150)
  }, error = function(e) cat("Label plot failed:", e$message, "\n"))

  # Model-class effect
  tryCatch({
    p_model <- ggplot(metrics[variant == "B" & eval_window == "unrestricted" & !is.na(auroc)],
                       aes(x = model, y = auroc, fill = model)) +
      geom_col() +
      labs(title    = "Model-class effect on AUROC (Variant B, temporal split)",
           subtitle = "H6: |primary − GBT| ≤ 0.02",
           x = "Model", y = "AUROC") +
      theme_bw() + theme(axis.text.x = element_text(angle = 30, hjust = 1),
                         legend.position = "none")
    print(p_model)
    ggsave(file.path(FIGURES_DIR, "09_model_class_auroc.png"), p_model,
           width = 6, height = 4, dpi = 150)
  }, error = function(e) cat("Model plot failed:", e$message, "\n"))
}

cat("\nVariance decomposition complete.\n")
```

## Label-Anchor Attribution (Protocol Amendment 2)

The label variants above all lie within Sepsis-3. Variants A, B and C are
three operationalisations of one definition, and all three require the same
culture-plus-antibiotic trigger. The label spread in H1 therefore measures
sensitivity to how one definition is operationalised, not to the choice of
construct. Within that family it is not possible to separate the part of the
discrimination that comes from the patient's physiology from the part that
comes from the treatment decisions the trigger records.

Sepsis-3 requires suspected infection and acute organ dysfunction together.
Protocol Amendment 2, which is post-hoc, separates the two components. The
same model specification, with the same covariates and temporal split, is
refitted under a label built from each component alone:

| Variant | Suspected infection | Organ dysfunction | Uses a treatment timestamp? |
|---|---|---|---|
| B (Seymour-Standard) | required | required | yes |
| E (Anchor-Only) | required | not required | yes |
| D (Deterioration) | not required | required | **no** |

Variant D supplies the anchor-independent label that Variants A--C lack. Its
onset is the first hour at which a SOFA score computed from physiology alone
rises by at least 2 points above the admission baseline. The vasopressor terms
are withheld, so no antibiotic, culture or drug-administration timestamp enters
the label. The label is not independent of clinical care, however, because its
inputs are measurements that exist because they were ordered.

The comparison yields two descriptive quantities. The first is the
**anchor-attributable fraction**, the ratio of Variant E's above-chance AUROC
to Variant B's:

$$\phi_{\text{anchor}} = \frac{\mathrm{AUROC}_E - 0.5}{\mathrm{AUROC}_B - 0.5}.$$

It compares discrimination on two different outcomes and is not a causal share.
A value near 1 means that a model predicting the treatment anchor alone
discriminates about as well as one predicting the full Sepsis-3 label. A value
above 1 means it discriminates slightly better, not that treatment explains
more than all of Variant B's performance. The ratio would become unstable if
Variant B's AUROC approached 0.5, which it does not here, and its bootstrap
interval is reported in `12_inference.Rmd`. The second quantity is the set of
**label agreement** statistics in the next section, which measure how far the
alternative labels move the onsets.

```{r anchor-attribution}
step("Label-anchor attribution (Protocol Amendment 2)...")

alt_path <- file.path(DATA_DIR, "07_metric_results_altlabel.parquet")
alt_res  <- if (file.exists(alt_path)) as.data.table(arrow::read_parquet(alt_path)) else NULL

attrib <- NULL
if (is.null(alt_res)) {
  cat("07_metric_results_altlabel.parquet not found; run 07 with the alternative-label variants.\n")
} else {
  ref <- metrics[eval_window == "unrestricted" & variant == "B",
                 .(model, auroc_B = auroc, auprc_B = auprc,
                   utility_B = utility_normalised, utility_max_B = utility_max,
                   n_events_B = n_sepsis_events)]

  alt <- alt_res[eval_window == "unrestricted",
                 .(variant, variant_name, model, auroc, auprc,
                   utility_normalised, utility_max, n_sepsis_events)]

  attrib <- merge(alt, ref, by = "model")
  attrib[, `:=`(
    delta_auroc_vs_B  = round(auroc - auroc_B, 4),
    anchor_fraction   = ifelse(variant == "E" & auroc_B > 0.5,
                               round((auroc - 0.5) / (auroc_B - 0.5), 3), NA_real_),
    event_rate_ratio  = round(n_sepsis_events / n_events_B, 2)
  )]
  setorder(attrib, variant, -auroc)

  knitr::kable(
    attrib[, .(variant, label = variant_name, model,
               auroc = round(auroc, 4), auroc_B = round(auroc_B, 4),
               delta_auroc_vs_B, n_events = n_sepsis_events,
               event_rate_ratio, anchor_fraction)],
    caption = "Discrimination under each limb of the Sepsis-3 conjunction, against Variant B (temporal test, unrestricted window)"
  )
  save_parquet(attrib, "09_label_anchor_attribution", DATA_DIR)
  save_csv(attrib,     "09_label_anchor_attribution", DATA_DIR)

  phi <- attrib[variant == "E" & model == "primary", anchor_fraction]
  if (length(phi) == 1 && !is.na(phi))
    cat(sprintf("\nAnchor-attributable fraction (primary model): %.1f%% of Variant B's above-chance AUROC is reproduced by the treatment anchor alone.\n",
                100 * phi))
}
```

## Label Agreement Between Definitions

Agreement with Variant B is measured on every stay in the cohort. A stay is
positive under a label if its onset falls within the 72-hour panel horizon,
because onsets beyond it never become events. For each variant the table gives
the percentage of stays labelled the same as under Variant B, Cohen's $\kappa$,
and the share of stays positive under Variant B that are also positive under
the variant (`sensitivity_to_B`). Among the stays positive under both labels
(`n_positive_both`), it also gives the median signed onset shift (variant minus
Variant B), the percentage with the same onset hour, the median absolute shift
and the interquartile range of the shift.

Cohen's $\kappa$ corrects agreement for chance. When two labels differ greatly
in prevalence, as Variants D and B do, the largest attainable $\kappa$ is
limited, so $\kappa$ is read together with the counts of positive stays.
Substantial disagreement means that the definitions identify different stays
or different onset hours, and so define different prediction targets.

```{r label-agreement}
# Onsets come from the person-hour tables where those exist, and otherwise from
# the cohort tables, which carry the same stay-level onset_hour for every
# variant except D (whose onsets are only defined on the hourly panel).
load_onsets <- function(vid) {
  p <- file.path(DATA_DIR, sprintf("03_onsets_%s.parquet", vid))
  if (!file.exists(p)) p <- file.path(DATA_DIR, sprintf("02_cohort_%s.parquet", vid))
  if (!file.exists(p)) return(NULL)
  d <- unique(as.data.table(
    arrow::read_parquet(p, col_select = c("stay_id", "onset_hour"))))
  # Onsets beyond the modelled horizon never become events, so they count as
  # negative here; this keeps D, whose onsets exist only inside the panel,
  # comparable with the variants labelled on the full stay.
  d[!is.na(onset_hour) & onset_hour > MAX_PANEL_HOURS, onset_hour := NA_integer_]
  setnames(d, "onset_hour", paste0("onset_", vid))
  d
}

cohen_kappa <- function(a, b) {
  po <- mean(a == b)
  pe <- mean(a) * mean(b) + mean(!a) * mean(!b)
  if (pe >= 1) return(NA_real_)
  (po - pe) / (1 - pe)
}

agreement <- NULL
ons <- Filter(Negate(is.null), lapply(ALL_VARIANTS, load_onsets))
if (length(ons) < 2 || is.null(load_onsets("B"))) {
  cat("Per-variant onset tables not available; re-run 03_person_hours.Rmd.\n")
} else {
  wide <- Reduce(function(x, y) merge(x, y, by = "stay_id"), ons)
  others <- setdiff(ALL_VARIANTS, "B")
  others <- others[paste0("onset_", others) %in% names(wide)]

  agreement <- rbindlist(lapply(others, function(vid) {
    oc <- paste0("onset_", vid)
    pos_B <- !is.na(wide$onset_B); pos_v <- !is.na(wide[[oc]])
    both  <- pos_B & pos_v
    data.table(
      variant          = vid,
      label            = variant_name(vid),
      label_family     = variant_type(vid),
      n_stays          = nrow(wide),
      n_positive       = sum(pos_v),
      n_positive_B     = sum(pos_B),
      n_positive_both  = sum(both),
      pct_agree        = round(100 * mean(pos_B == pos_v), 1),
      kappa_vs_B       = round(cohen_kappa(pos_B, pos_v), 3),
      # of the stays Sepsis-3 (B) calls septic, how many does this label find
      sensitivity_to_B = round(sum(both) / max(sum(pos_B), 1), 3),
      median_shift_h   = if (sum(both) > 0)
                           round(median(wide[[oc]][both] - wide$onset_B[both]), 1)
                         else NA_real_,
      # Among stays both labels call septic, how often they agree on the hour,
      # and how far apart they are when they do not. The pre-registered variants
      # all write t_susp, so exact agreement there is near-total by construction;
      # it is the variants that disagree which carry information.
      pct_exact_match  = if (sum(both) > 0)
                           round(100 * mean(wide[[oc]][both] == wide$onset_B[both]), 1)
                         else NA_real_,
      median_abs_shift_h = if (sum(both) > 0)
                             round(median(abs(wide[[oc]][both] - wide$onset_B[both])), 1)
                           else NA_real_,
      iqr_shift_lo_h   = if (sum(both) > 0)
                           round(quantile(wide[[oc]][both] - wide$onset_B[both], 0.25), 1)
                         else NA_real_,
      iqr_shift_hi_h   = if (sum(both) > 0)
                           round(quantile(wide[[oc]][both] - wide$onset_B[both], 0.75), 1)
                         else NA_real_
    )
  }))

  knitr::kable(agreement, caption = "Onset agreement with Variant B across all label variants")
  save_parquet(agreement, "09_label_agreement", DATA_DIR)
  save_csv(agreement,     "09_label_agreement", DATA_DIR)
}
```

## The Pre-Treatment Window (H3), Analytically and Empirically

H3 concerns evaluation restricted to the hours before the first antibiotic or
culture. Under Variants A--C the number of onsets in that window follows from
the onset definition alone, so it is established analytically.

> **Proposition.** Let $t_{\text{anchor}}$ be the earlier of the admission's
> first antibiotic and first culture, and let $t_{\text{onset}}$ be the onset
> assigned by Variant A, B, C or E. Then $t_{\text{onset}} \geq t_{\text{anchor}}$
> for every labelled stay, so the pre-anchor window contains no onsets.
>
> *Proof.* Under Variant E, $t_{\text{onset}} = t_{\text{susp}}$, the later of
> the two timestamps in a qualifying antibiotic-culture pair. Neither member of
> the pair can be earlier than the first event of its kind, so
> $t_{\text{susp}} \geq t_{\text{anchor}}$. Under Variants A, B and C the SOFA
> criterion decides whether a stay is labelled, but the assigned onset is still
> $t_{\text{susp}}$, so the inequality holds there too. $\square$
>
> The proposition concerns this implementation of the labels, and in
> particular its timing of onset at the later member of the pair. The absence
> of pre-anchor onsets is a structural consequence of that onset rule, not a
> property of the MIMIC-IV population.

H3 therefore cannot be tested as pre-registered under Variants A--C. The
pre-treatment window is instead examined under definitions that allow an onset
before the anchor. Variant D drops the infection anchor. Variant F (Protocol
Amendment 4, post-hoc) labels the same stays as Variant B cohort-wide and
changes only the onset time, to $\min(t_{\text{susp}}, t_{\text{SOFA}})$, the
rule used by Seymour et al. (2016) and the PhysioNet/CinC 2019 Challenge. Under
that rule an onset precedes the anchor whenever the rise in organ dysfunction
is detected first. Variant F is an exploratory counterpart to H3, not a
redefinition of it.

The table gives, for the primary model and the GBT, the person-hours and onsets
in the pre-treatment window and the AUROC there. The column `structural_zero`
marks the variants whose zero follows from the proposition. The pre-treatment
person-hours differ slightly between Variants B and F because the panel holds
each stay's at-risk hours only, up to and including onset. When Variant F moves
an onset earlier, that stay's at-risk hours end earlier. Variant F's AUROC in
this window rests on few onsets and carries no interval here, so it is
descriptive. The lines printed below the table summarise the pre-treatment
onsets and Variant F's onset shift from `03_person_hours.Rmd`.

```{r pretreatment-onsets}
step("Pre-treatment (treatment-anchored) window across all label variants...")

pre_cols <- c("variant", "model", "n_rows", "n_sepsis_events", "auroc")
pre_src  <- list(metrics)
if (!is.null(alt_res)) pre_src <- c(pre_src, list(alt_res))

pretreat <- rbindlist(lapply(pre_src, function(d) {
  if (is.null(d) || !nrow(d)) return(NULL)
  d[eval_window == "treatment_anchored", ..pre_cols]
}), use.names = TRUE, fill = TRUE)

if (nrow(pretreat) == 0) {
  cat("No treatment-anchored rows found; re-run 07_metrics_suite.Rmd.\n")
} else {
  pretreat[, label        := vapply(variant, variant_name, character(1))]
  pretreat[, label_family := vapply(variant, variant_type, character(1))]
  # Zero by construction wherever the onset is the anchor (or gated by it)
  pretreat[, structural_zero := label_family %in% c("sepsis3", "anchor_only")]
  setcolorder(pretreat, c("variant", "label", "label_family", "model"))
  setorder(pretreat, variant, model)

  shift <- if (file.exists(file.path(DATA_DIR, "03_early_onset_shift.parquet")))
             as.data.table(arrow::read_parquet(
               file.path(DATA_DIR, "03_early_onset_shift.parquet"))) else NULL

  knitr::kable(
    pretreat[model %in% c("primary", "gbt"),
             .(variant, label, model, person_hours = n_rows,
               onsets = n_sepsis_events, auroc = round(auroc, 4),
               structural_zero)],
    caption = "Onsets and discrimination in the pre-treatment window, by label variant"
  )
  save_parquet(pretreat, "09_pretreatment_onsets", DATA_DIR)
  save_csv(pretreat,     "09_pretreatment_onsets", DATA_DIR)

  f_row <- pretreat[variant == "F" & model == "primary"]
  b_row <- pretreat[variant == "B" & model == "primary"]
  if (nrow(f_row) == 1 && nrow(b_row) == 1)
    cat(sprintf("\nPre-treatment onsets: B (conjunction timing): %d in %s person-h; F (standard min(t_susp, t_SOFA) timing, same stays): %d in %s person-h, AUROC %s\n",
                b_row$n_sepsis_events, format(b_row$n_rows, big.mark = ","),
                f_row$n_sepsis_events, format(f_row$n_rows, big.mark = ","),
                if (is.na(f_row$auroc)) "undefined" else sprintf("%.4f", f_row$auroc)))
  if (!is.null(shift) && nrow(shift) == 1)
    cat(sprintf("Variant F moved %s of %s onsets earlier (%.1f%%), median shift %s h.\n",
                format(shift$n_moved_earlier, big.mark = ","),
                format(shift$n_onsets, big.mark = ","),
                shift$pct_moved_earlier, format(shift$median_shift_h)))
}
```

```{r anchor-attribution-plot, fig.cap="Discrimination under each limb of the Sepsis-3 conjunction. Variant B requires suspected infection and organ dysfunction; E keeps only the treatment anchor; D keeps only the physiological deterioration and contains no treatment timestamp; F keeps the full conjunction but times the onset as min(t\\_susp, t\\_SOFA).", fig.width=7, fig.height=4}
if (!is.null(attrib) && nrow(attrib) > 0) {
  tryCatch({
    plot_dt <- rbind(
      metrics[eval_window == "unrestricted" & variant == "B",
              .(variant, model, auroc)],
      attrib[, .(variant, model, auroc)]
    )
    plot_dt[, label := factor(
      variant,
      levels = c("E", "B", "F", "D"),
      labels = c("E: anchor only\n(treatment)", "B: Sepsis-3\n(anchor AND dysfunction)",
                 "F: Sepsis-3\n(standard onset timing)",
                 "D: deterioration only\n(treatment-independent)"))]
    plot_dt <- plot_dt[!is.na(label) & !is.na(auroc)]

    p_anchor <- ggplot(plot_dt, aes(x = label, y = auroc, fill = model)) +
      geom_col(position = position_dodge(width = 0.8), width = 0.75) +
      geom_hline(yintercept = 0.5, linetype = "dashed") +
      labs(title    = "Splitting the Sepsis-3 conjunction into its two limbs",
           subtitle = "Temporal test set, unrestricted window; dashed line = chance",
           x = NULL, y = "AUROC", fill = "Model") +
      theme_bw() + theme(axis.text.x = element_text(size = 8))
    print(p_anchor)
    ggsave(file.path(FIGURES_DIR, "09_label_anchor_attribution.png"), p_anchor,
           width = 7, height = 4, dpi = 150)
  }, error = function(e) cat("Anchor attribution plot failed:", e$message, "\n"))
}
```

## Reading the Attribution

The comparison supports three observations. The first two follow from how the
labels are defined, and only the third depends on the data.

1. **Variant E's label is not a sepsis label.** Its onset is the
   suspected-infection time derived from the culture and antibiotic records.
   Discrimination under Variant E is discrimination of that treatment-related
   anchor, which reflects both the patient's condition and clinical practice.
   $\phi_{\text{anchor}}$ therefore compares how well the anchor and the full
   Sepsis-3 label can be predicted. It is not the share of Variant B's
   discrimination that treatment decisions explain.
2. **Variant D's AUROC does not estimate performance on Variant B's target.**
   Its label is a threshold on recorded physiology, and five of the six SOFA
   components (all but PaO$_2$/FiO$_2$) are also model inputs, so its AUROC is
   partly circular. The informative comparison is whether the label changes
   which stays are flagged and when, which the agreement table shows.
3. **The importance of label choice depends on how far Variants D and B
   disagree.** If a label without a treatment timestamp identifies
   substantially different stays at substantially different hours, then the
   choice of label changes the prediction target itself, not only the way one
   definition is operationalised.

The output below prints the primary model's AUROC under Variants B, E and D,
and the agreement between Variants D and B. In that line, "B's septic stays"
means the stays positive under Variant B. Cohen's $\kappa$ is computed over all
stays, whereas the onset shift is computed only over the stays positive under
both labels. A median signed shift of 0 hours is therefore compatible with a
low $\kappa$. The agreement table shows that the two labels assign the same
onset hour to few of those stays, and that the shifts spread in both
directions.

```{r attribution-summary}
if (!is.null(attrib)) {
  e_prim <- attrib[variant == "E" & model == "primary"]
  d_prim <- attrib[variant == "D" & model == "primary"]
  b_prim <- metrics[variant == "B" & model == "primary" & eval_window == "unrestricted"]
  if (nrow(e_prim) == 1 && nrow(b_prim) == 1)
    cat(sprintf("Primary model AUROC: B (Sepsis-3): %.3f | E (anchor only): %.3f | D (deterioration): %s\n",
                b_prim$auroc, e_prim$auroc,
                if (nrow(d_prim) == 1) sprintf("%.3f", d_prim$auroc) else "n/a"))
}
if (!is.null(agreement)) {
  d_row <- agreement[variant == "D"]
  if (nrow(d_row) == 1)
    cat(sprintf("Variant D vs B: kappa = %.3f, %.1f%% of B's septic stays also deteriorate, median onset shift %s h\n",
                d_row$kappa_vs_B, 100 * d_row$sensitivity_to_B,
                format(d_row$median_shift_h)))
}
```
