---
title: "10 Equity Analysis - Subgroup Performance and Label Sensitivity"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

Wang et al. (2022), working in MIMIC-III, reported two findings. Competing
sepsis identification criteria selected sub-populations that differed by race,
insurance and language. Separately, mortality prediction among patients already
identified as septic was worse for Asian, Hispanic and Spanish-speaking
patients. Their performance finding concerns mortality rather than sepsis
onset, so it gives no prior estimate for an onset model. Together, however, the
two findings motivate examining both quantities on the same subgroups of one
onset-prediction cohort, which this notebook does.

The first quantity is subgroup performance: AUROC, AUPRC and alert burden for
each recorded level of the race, language and insurance fields. The second is
subgroup label sensitivity: the range, in percentage points, of the proportion
of stays labelled septic across Variants A, B and C, computed within each
subgroup level. If that range differs between subgroups, the choice of label
changes subgroup event rates by different amounts, and any rate computed later
depends on which label was chosen.

The subgroup fields are taken from the hospital admission record of each stay.
`config.R` declares four equity variables, but MIMIC-IV records race and
ethnicity in a single field, so three carry data: race (which includes
ethnicity), language and insurance. These are administrative registration
fields. Insurance is included as a recorded administrative characteristic, not
as a measure of socioeconomic position, and none of the three fields defines a
homogeneous population.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(PRROC)
  library(ggplot2)
  library(tidyr)
  library(dplyr)
})

source("config.R")
source("utils.R")

step("=== Phase 6b: Equity Analysis ===")
```

## Subgroup Performance

This section estimates model performance within each subgroup level. Every
metric is computed from the same Variant B test-set predictions as the
aggregate results, stratified by subgroup, with no model refitted within a
subgroup. A difference between strata therefore cannot come from separate
fitting or from a different metric implementation. It can still reflect
differences in sample size, event prevalence, case mix or data completeness.

Two features of the data limit interpretation. Strata with few onsets have
wide intervals, often wide enough to include an AUROC of 0.5. The subgroup
fields are administrative categories, and some of their levels record missing
or unobtainable information rather than a demographic answer.

### Subgroup Metric Calculation

The function below computes AUROC, AUPRC and alert burden for one subgroup
level and one model. It returns nothing for a level with fewer than 50 scored
person-hours or with no onset. All three metrics are computed over
person-hours. Alert burden is the number of false alerts per true alert at the
default threshold of 0.5.

```{r subgroup-helper}
compute_subgroup_metrics <- function(dt, pred_col, subgroup_col, subgroup_val) {
  sub <- dt[get(subgroup_col) == subgroup_val & !is.na(get(pred_col))]
  if (nrow(sub) < 50) return(NULL)

  y     <- as.integer(sub$event == 1)
  pred  <- as.numeric(sub[[pred_col]])
  alert <- as.integer(pred >= 0.5)
  n_pos <- sum(y)
  if (n_pos == 0) return(NULL)

  roc_pr <- tryCatch({
    list(
      auroc = PRROC::roc.curve(pred, weights.class0 = y)$auc,
      auprc = PRROC::pr.curve(pred,  weights.class0 = y)$auc.integral
    )
  }, error = function(e) list(auroc = NA, auprc = NA))

  tp     <- sum(alert == 1 & y == 1, na.rm = TRUE)
  fp     <- sum(alert == 1 & y == 0, na.rm = TRUE)
  burden <- if (tp == 0) Inf else fp / tp

  data.frame(
    subgroup_col = subgroup_col,
    subgroup_val = subgroup_val,
    n_rows       = nrow(sub),
    n_stays      = length(unique(sub$stay_id)),
    n_sepsis     = n_pos,
    auroc        = round(roc_pr$auroc, 4),
    auprc        = round(roc_pr$auprc, 4),
    alert_burden = round(burden, 2)
  )
}
```

### Performance by Subgroup (Variant B)

For each subgroup field, the eight levels with the most test person-hours are
selected. Each is scored for three models: the primary model, gradient-boosted
trees fitted with XGBoost (GBT) and NEWS2, shown in the table as
`pred_sepsis`, `pred_gbt` and `pred_news2`. The table therefore covers all three
models, not only the primary model. The selection uses counts alone, not
performance or outcomes. The requirement of at least one onset, applied by the
function above, does use outcomes, and it removes levels that have none.

Two event floors apply to these strata, and they serve different purposes. The
floor here only decides whether a metric can be computed. `12_inference.Rmd`
computes a bootstrap interval for a stratum with at least 5 onsets. It
interprets a stratum, and includes it in the exploratory contrast family E1,
only if it has at least 100 onsets (Protocol Amendment 6). Strata below 100
onsets are still estimated and shown, but they support no comparison. Several
rows in the table rest on fewer than 20 onsets, and their AUROCs should not be
read or ranked. The race strata that reach 100 onsets are White and Unknown,
and Unknown records missing information.

Some entries need explanation. `n_rows` is the number of test person-hours and
`n_sepsis` the number of onsets. A blank `subgroup_val` means the field is
empty in the admission record. The race levels are reproduced exactly as
MIMIC-IV records them. That field mixes broad and specific levels from the same
population, such as WHITE and WHITE - OTHER EUROPEAN, so the levels overlap in
meaning and are not disjoint populations. The primary model's alert burden is
infinite in every row because it raises no alert at the 0.5 threshold, so
there is no true alert to divide by. The other models alert on many
person-hours at 0.5, which gives burdens in the hundreds. AUPRC is low
throughout because onsets are rare per person-hour, and it depends on each
stratum's event rate, so AUPRCs are not comparable between strata with
different rates.

```{r subgroup-performance}
step("Subgroup performance analysis...")

prim_path <- file.path(DATA_DIR, "05_predictions_primary_B.parquet")
comp_path <- file.path(DATA_DIR, "06_predictions_comparators_B.parquet")

perf_df <- NULL

if (!file.exists(prim_path)) {
  cat("Primary model predictions not found; skipping subgroup performance.\n")
} else {
  ph <- as.data.table(arrow::read_parquet(prim_path))
  if (file.exists(comp_path)) {
    comp       <- as.data.table(arrow::read_parquet(comp_path))
    merge_cols <- c("stay_id", "hour", setdiff(names(comp), names(ph)))
    ph <- merge(ph, comp[, ..merge_cols], by = c("stay_id", "hour"), all.x = TRUE)
  }

  ph_demo_path <- file.path(DATA_DIR, "03_person_hours_B.parquet")
  if (file.exists(ph_demo_path)) {
    ph_demo <- as.data.table(arrow::read_parquet(ph_demo_path,
                                                   col_select = c("stay_id","race",
                                                                   "language","insurance")))
    ph_demo <- unique(ph_demo)
    ph <- merge(ph, ph_demo, by = "stay_id", all.x = TRUE)
  }

  perf_results <- list()
  models_eval  <- c("pred_sepsis", "pred_gbt", "pred_news2")

  for (subg_col in EQUITY_VARS) {
    if (!subg_col %in% names(ph)) next
    val_counts <- ph[, .N, by = subg_col][order(-N)]
    top_vals   <- head(val_counts[[subg_col]], 8)
    for (val in top_vals) {
      for (mod_col in models_eval) {
        if (!mod_col %in% names(ph)) next
        res <- compute_subgroup_metrics(ph, mod_col, subg_col, val)
        if (!is.null(res)) {
          res$model   <- mod_col
          res$variant <- "B"
          perf_results[[paste(subg_col, val, mod_col, sep = "-")]] <- res
        }
      }
    }
  }

  if (length(perf_results) > 0) {
    perf_df <- do.call(rbind, perf_results)
    save_parquet(perf_df, "10_subgroup_performance", DATA_DIR)

    # 63 rows does not fit in a single float; longtable lets it break across
    # pages instead of overflowing (LaTeX "Float too large for page").
    knitr::kable(
      perf_df[order(perf_df$subgroup_col, perf_df$subgroup_val), ],
      longtable = TRUE,
      caption = "Subgroup performance (primary model, Variant B)"
    )
  }
}
```

### AUROC by Race

The figure shows the primary model's AUROC for each race level in the table
above, ordered from highest to lowest AUROC. The dashed line is the unweighted
mean of these stratum AUROCs, not the AUROC of the whole test set, so a stratum
with 13 onsets counts as much as one with 597. The subtitle refers to the
mortality finding of Wang et al. (2022) described in the introduction. Only
the White and Unknown strata reach the interpretation floor, so differences in
bar height support no comparison. The contrasts, their bootstrap intervals and
the false-discovery-rate correction are computed in `12_inference.Rmd`.

```{r auroc-by-race, fig.cap="AUROC stratified by race/ethnicity for the primary model (Variant B). Dashed line indicates the overall mean.", fig.width=8, fig.height=5}
if (!is.null(perf_df) && "race" %in% perf_df$subgroup_col) {
  race_df <- perf_df[perf_df$subgroup_col == "race" & perf_df$model == "pred_sepsis", ]
  if (nrow(race_df) > 0) {
    tryCatch({
      p <- ggplot(race_df, aes(x = reorder(subgroup_val, -auroc), y = auroc,
                                fill = subgroup_val)) +
        geom_col() +
        geom_hline(yintercept = mean(race_df$auroc, na.rm = TRUE),
                   linetype = "dashed", color = "gray40") +
        labs(title    = "AUROC by race/ethnicity (primary model, Variant B)",
             subtitle = "Wang et al. (2022): performance degraded for Asian, Hispanic, Spanish-speaking",
             x = "Race/Ethnicity", y = "AUROC") +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 40, hjust = 1), legend.position = "none")
      print(p)
      ggsave(file.path(FIGURES_DIR, "10_auroc_by_race.png"), p, width = 8, height = 5, dpi = 150)
    }, error = function(e) cat("Race plot failed:", e$message, "\n"))
  }
}
```

## Subgroup Label-Sensitivity

For each subgroup level, the percentage of stays labelled septic is computed
under each of Variants A, B and C over the whole cohort, not only the test set.
The label sensitivity of the level is the largest of the three percentages
minus the smallest, in percentage points (pp). A larger range means that the
level's event rate changes more with the choice of label. The measure compares
proportions, so it does not show which individual stays change status between
labels.

The levels are the eight with the most stays under Variant B. A field with
fewer than eight non-missing levels contributes all of them, however small.
Insurance is such a field: its "No charge" level holds only seven stays, so its
range of 28.6 pp is a difference of two stays and carries no information. As
above, a blank level means an empty field. The table is sorted by range. The
second table repeats the levels with a range above 5 pp, a display filter that
was not pre-specified. The stratum sizes are not printed here, so small levels
should be read with care. The stratum sizes and the bootstrap contrasts
against each field's reference level, corrected within exploratory family E2,
are produced in `12_inference.Rmd`.

```{r label-sensitivity}
step("Subgroup label-sensitivity analysis...")

label_sens_results <- list()

for (subg_col in EQUITY_VARS) {
  cohort_data <- lapply(names(LABEL_VARIANTS), function(vid) {
    coh_path <- file.path(DATA_DIR, paste0("02_cohort_", vid, ".parquet"))
    if (!file.exists(coh_path)) return(NULL)
    coh_full <- tryCatch(as.data.table(arrow::read_parquet(coh_path)),
                         error = function(e) NULL)
    if (is.null(coh_full)) return(NULL)
    want_cols <- intersect(c("stay_id","subject_id", subg_col, "onset_hour",
                              "label_variant"), names(coh_full))
    coh <- coh_full[, ..want_cols]
    coh[, variant   := vid]
    coh[, has_sepsis := !is.na(onset_hour)]
    coh
  })
  cohort_data <- Filter(Negate(is.null), cohort_data)
  if (length(cohort_data) == 0) next

  all_cohort <- rbindlist(cohort_data, fill = TRUE)
  if (!subg_col %in% names(all_cohort)) next

  val_counts <- all_cohort[variant == "B" & !is.na(get(subg_col)), .N, by = subg_col]
  top_vals   <- head(val_counts[order(-N)][[subg_col]], 8)

  for (val in top_vals) {
    subg_dt <- all_cohort[get(subg_col) == val]
    if (nrow(subg_dt) == 0) next

    by_variant <- subg_dt[, .(
      n_stays    = .N,
      n_sepsis   = sum(has_sepsis),
      pct_sepsis = round(100 * sum(has_sepsis) / .N, 1)
    ), by = variant]

    if (nrow(by_variant) >= 2) {
      pct_range <- diff(range(by_variant$pct_sepsis))
      get_pct   <- function(v) {
        r <- by_variant[variant == v, pct_sepsis]
        if (length(r) > 0) r[1] else NA_real_
      }
      label_sens_results[[paste(subg_col, val, sep = "-")]] <- data.frame(
        subgroup_col    = subg_col,
        subgroup_val    = as.character(val),
        pct_A           = get_pct("A"),
        pct_B           = get_pct("B"),
        pct_C           = get_pct("C"),
        label_range_pct = round(pct_range, 1)
      )
    }
  }
}

if (length(label_sens_results) > 0) {
  ls_df <- do.call(rbind, label_sens_results)
  save_parquet(ls_df, "10_subgroup_label_sensitivity", DATA_DIR)

  knitr::kable(
    ls_df[order(-abs(ls_df$label_range_pct)), ],
    caption = "Subgroup label-sensitivity (% sepsis by variant A/B/C per demographic group)"
  )

  high_sens <- ls_df[!is.na(ls_df$label_range_pct) & ls_df$label_range_pct > 5, ]
  if (nrow(high_sens) > 0) {
    cat("\nSubgroups with highest label sensitivity (>5 pp range):\n")
    knitr::kable(high_sens)
  }
}
```

### Label Sensitivity and AUROC

The figure plots each subgroup level's label sensitivity against the primary
model's AUROC for that level under Variant B. It is exploratory and tests
nothing. The fitted line is an unweighted least-squares line, so a level with
a handful of onsets counts as much as one with hundreds. It also pools three
different fields whose levels are not exchangeable and, for race, overlap in
meaning, and it treats both axes as exact when both are estimated. A downward
slope would mean only that levels whose event rate depends more on the label
also tend to have lower estimated AUROC. It would not show that one causes the
other.

```{r label-sens-vs-auroc, fig.cap="Subgroup label sensitivity (x) vs. AUROC (y). Each point is a demographic subgroup; a downward trend suggests label ambiguity compounds prediction difficulty.", fig.width=8, fig.height=4.8}
if (length(label_sens_results) > 0 && !is.null(perf_df)) {
  merged_eq <- merge(ls_df,
                     perf_df[perf_df$model == "pred_sepsis",
                              c("subgroup_col","subgroup_val","auroc")],
                     by = c("subgroup_col","subgroup_val"), all.x = TRUE)

  if (nrow(merged_eq[!is.na(merged_eq$auroc), ]) > 0) {
    tryCatch({
      p_eq <- ggplot(merged_eq[!is.na(merged_eq$auroc), ],
                     aes(x = label_range_pct, y = auroc,
                         label = paste0(subgroup_col, ":", subgroup_val))) +
        geom_point(aes(color = subgroup_col), size = 3) +
        geom_smooth(method = "lm", se = TRUE, color = "gray40") +
        geom_text(hjust = 1.1, size = 2.5) +
        labs(title    = "Label sensitivity vs. AUROC by subgroup",
             subtitle = "Novel conjunction: Wang et al. (2022) established each separately",
             x = "Label sensitivity (pp range across variants A–C)",
             y = "AUROC (primary model, Variant B)") +
        theme_bw()
      print(p_eq)
      ggsave(file.path(FIGURES_DIR, "10_label_sens_vs_auroc.png"), p_eq,
             width = 8, height = 6, dpi = 150)
    }, error = function(e) cat("Equity plot failed:", e$message, "\n"))
  }
}

cat("\nEquity analysis and all V2 analyses complete.\n")
```

## Reading the Equity Results

The two analyses support conclusions of different strength.

The subgroup performance estimates are limited mainly by their event counts.
Onsets are rare, and dividing them across subgroup levels leaves most levels
with too few onsets for a precise AUROC. A contrast that is not significant
means that the study did not detect a difference. It does not show that the
model performs equally across strata. `12_inference.Rmd` applies the 100-onset
interpretation floor, and only contrasts between strata that reach it are
corrected and interpreted. The analysis covers discrimination only. Subgroup
calibration is not estimated here, and the alert burden at a common threshold
depends on calibration and event rate as well as on discrimination.

The label-sensitivity analysis rests on stay-level proportions rather than on
a discrimination estimate, so it is less limited by onset counts. Its
proportions are still subject to sampling variability, as the seven-stay
"No charge" level shows. If the range across Variants A, B and C differs
between subgroups, then the proportion of stays labelled septic depends on the
definition to different degrees in different subgroups. Subgroup event rates,
and the performance estimates computed on them, then depend on the label
chosen. This variation arises when the outcome is constructed, so it exists
independently of any model.

The quality of the administrative fields limits both analyses. Levels such as
UNKNOWN, UNABLE TO OBTAIN and a blank entry record the absence of information,
so they are data-quality categories rather than demographic groups. A result
for such a level, even one that reaches the event floor, supports no
conclusion about any particular population.
