---
title: "11 Clinical Utility Metrics - Calibration Curves, Net Benefit, Operating Points, Event-Level Detection"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "85%", fig.width = 7, fig.height = 4.2
)
```

## Introduction

`07_metrics_suite.Rmd` reports discrimination (AUROC and AUPRC), the PhysioNet
Utility Score, the calibration intercept and slope, and a ratio-based alert
burden. This notebook extends that evaluation to six further aspects of
clinical usefulness:

1. **Calibration curves.** The intercept and slope summarise calibration in two
   numbers and do not show where along the range of predicted risk a model is
   miscalibrated. The flexible calibration curve is therefore plotted. It is
   summarised by three measures of the distance between predicted and observed
   risk, the Integrated Calibration Index (ICI), E50 and E90, and by the scaled
   Brier score (Austin & Steyerberg, 2019; Van Calster et al., 2019).
2. **Decision-curve analysis.** Stage 07 evaluated net benefit on a threshold
   grid of 0.01--0.20. The hourly event prevalence is about 0.2--0.4 per cent,
   so even the lowest threshold on that grid is several times the prevalence.
   On that grid the treat-all strategy has negative net benefit throughout, and
   the primary model's net benefit stays close to zero. Net benefit is
   therefore recomputed here on a grid scaled to the prevalence (Vickers &
   Elkin, 2006; Vickers et al., 2019).
3. **Performance at fixed sensitivity.** An alerting system is evaluated at a
   chosen operating threshold rather than at a default threshold of 0.5.
   Thresholds are selected to give per-hour sensitivities of 0.50, 0.80 and
   0.90. At each threshold, the PPV, specificity and number needed to evaluate
   are reported.
4. **False alerts per patient-day.** The ratio-based burden (false alerts per
   true alert) is not expressed per unit of time, so it is hard to compare with
   alert-frequency figures from clinical practice. Both false-alert hours and
   distinct alert episodes per patient-day are computed here.
5. **Per-patient detection.** All of stage 07's discrimination measures are
   per person-hour. Here the analysis also reports three per-patient measures:
   the proportion of label-positive stays alerted at or before onset, a
   per-patient PPV, and the proportion of label-negative stays that receive at
   least one alert. They appear alongside the per-hour figures. A
   label-positive stay is one with an onset under the variant's label
   definition.
6. **Lead time.** Lead time is reported as a distribution: its median and
   interquartile range, and the proportions of label-positive stays alerted at
   least 3 and 6 hours before onset.

All measures are computed on the temporal test set (2017--2019), from the
predictions written by `05_primary_model.Rmd`, `06_comparators.Rmd` and
`08_external_validation.Rmd`. No model is refitted.

The prediction scales differ between models, which matters for every
probability-based measure below. The primary model returns a probability of
onset in the hour. The GBT returns a probability from a model trained with the
onset rows up-weighted, so its values are far above the observed event rate.
The Cox score is the logistic transform of a relative hazard. NEWS2, qSOFA and
SIRS are integer scores divided by their maxima (20, 3 and 4). Only the primary
model's output is intended as a calibrated probability. For the other models,
the calibration measures, the Brier score and the probability thresholds of
the decision curves describe a score treated as if it were a probability.
Their ranking-based results (sensitivity, specificity and PPV at a chosen
threshold) are unaffected.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(arrow)
  library(splines)
  library(ggplot2)
  library(dplyr)
})

source("config.R")
source("utils.R")

step("=== Clinical Utility Metric Suite (11) ===")

# Sensitivity targets at which operating points are defined
SENS_TARGETS <- c(0.50, 0.80, 0.90)

# Headline operating point reported in the thesis
HEADLINE_SENS <- 0.80

# Models carried through this notebook
MODEL_NAMES <- c(primary = "pred_sepsis", gbt = "pred_gbt",
                  news2 = "pred_news2", cox = "pred_cox",
                  qsofa = "pred_qsofa", sirs = "pred_sirs")

# Models shown in the plots (the primary model and its two headline comparators)
PLOT_MODELS <- c("primary", "gbt", "news2")

MODEL_LABELS <- c(primary = "Primary (pooled logistic)", gbt = "GBT",
                   news2 = "NEWS2", cox = "Cox", qsofa = "qSOFA", sirs = "SIRS")
```

## Metric Functions

This section defines the functions used for the measures below. The
threshold-counting function (`counts_at_thresholds()`) gives the number of
alerts, true positives and false positives at each threshold. It sorts the
predictions once and derives all the counts from cumulative sums, so the cost
of a long threshold grid stays low.

- **Calibration.** A logistic regression of the outcome on a natural cubic
  spline (3 degrees of freedom) of the logit of the predicted risk gives the
  flexible calibration curve. ICI, E50 and E90 are the mean, median and 90th
  percentile of the absolute difference between that curve and the predicted
  risk. The intercept is estimated with the logit of the prediction as an
  offset, and the slope is the coefficient of that logit.
- **Net benefit.** Net benefit follows Vickers and Elkin (2006):
  $\text{NB}(t) = \text{TP}/N - (\text{FP}/N)\,t/(1-t)$.
- **Operating points.** The operating threshold for a target sensitivity is
  the corresponding quantile of the predictions in event hours. Every
  person-hour with a prediction at or above it raises an alert. For an integer
  score, all hours tied at the threshold alert together, so the achieved
  sensitivity can exceed the target. Patient-days are the number of evaluated
  person-hours divided by 24.
- **Per-patient measures.** These work at the level of the stay, as described
  under "Per-Patient (Event-Level) Detection".

```{r metric-fns}
# --------------------------------------------------------------------------- #
# Small helpers
# --------------------------------------------------------------------------- #
# clamp_p() lives in utils.R: stage 12 clamps the same predictions before
# taking their logit, and the two must use the same epsilon.

#' Count, for each threshold, how many predictions are >= that threshold,
#' and how many of those are events. O(n log n) once, then O(k log n).
#' Returns a data.table with threshold, n_alert, tp, fp.
counts_at_thresholds <- function(y, p, thresholds) {
  ord    <- order(p, decreasing = TRUE)
  p_desc <- p[ord]
  y_desc <- y[ord]
  cum_tp <- cumsum(y_desc)
  n      <- length(p_desc)
  p_asc  <- rev(p_desc)

  # number of predictions strictly below t  ->  n_alert = n - that
  n_below <- findInterval(thresholds, p_asc, left.open = TRUE)
  k       <- n - n_below
  tp      <- ifelse(k > 0, cum_tp[pmax(k, 1)], 0)

  data.table(threshold = thresholds,
             n_alert   = k,
             tp        = tp,
             fp        = k - tp)
}

# --------------------------------------------------------------------------- #
# 1. Calibration: flexible curve + ICI / E50 / E90 / Brier
#    Austin & Steyerberg (2019). The calibration curve is a restricted cubic
#    spline of the outcome on the logit of the predicted risk.
# --------------------------------------------------------------------------- #
calibration_curve <- function(y, p, n_bins = 20L, df_spline = 3L) {
  keep <- !is.na(y) & !is.na(p)
  y <- as.integer(y[keep]); p <- as.numeric(p[keep])
  if (length(y) < 100 || sum(y) < 10 || length(unique(p)) < 20)
    return(list(summary = NULL, curve = NULL, bins = NULL))

  pc <- clamp_p(p)
  lp <- qlogis(pc)

  fit <- tryCatch(
    glm(y ~ ns(lp, df = df_spline), family = binomial),
    error = function(e) NULL
  )
  if (is.null(fit)) return(list(summary = NULL, curve = NULL, bins = NULL))

  obs <- as.numeric(fitted(fit))
  d   <- abs(obs - pc)

  # Intercept / slope (recomputed here so the curve and the summary agree)
  int_fit   <- tryCatch(coef(glm(y ~ offset(lp), family = binomial))[1],
                        error = function(e) NA_real_)
  slope_fit <- tryCatch(coef(glm(y ~ lp, family = binomial))[2],
                        error = function(e) NA_real_)

  brier      <- mean((pc - y)^2)
  prev       <- mean(y)
  brier_null <- prev * (1 - prev)

  summ <- data.table(
    n            = length(y),
    n_events     = sum(y),
    prevalence   = prev,
    calib_int    = as.numeric(int_fit),
    calib_slope  = as.numeric(slope_fit),
    ici          = mean(d),
    e50          = as.numeric(stats::quantile(d, 0.50)),
    e90          = as.numeric(stats::quantile(d, 0.90)),
    emax         = max(d),
    brier        = brier,
    brier_scaled = 1 - brier / brier_null
  )

  # Smooth curve on a grid over the observed prediction range
  grid_lp <- seq(quantile(lp, 0.001), quantile(lp, 0.999), length.out = 200)
  grid_p  <- plogis(grid_lp)
  pr      <- predict(fit, newdata = data.frame(lp = grid_lp),
                     type = "link", se.fit = TRUE)
  curve <- data.table(
    pred     = grid_p,
    obs      = plogis(pr$fit),
    obs_low  = plogis(pr$fit - 1.96 * pr$se.fit),
    obs_high = plogis(pr$fit + 1.96 * pr$se.fit)
  )

  # Decile (default: vigintile) bins for the overlaid points
  brk <- unique(stats::quantile(pc, probs = seq(0, 1, length.out = n_bins + 1)))
  if (length(brk) > 2) {
    bin <- cut(pc, breaks = brk, include.lowest = TRUE, labels = FALSE)
    bins <- data.table(bin = bin, y = y, p = pc)[
      !is.na(bin), .(pred = mean(p), obs = mean(y), n = .N), by = bin]
    bins[, `:=`(obs_low  = pmax(0, obs - 1.96 * sqrt(obs * (1 - obs) / n)),
                obs_high = pmin(1, obs + 1.96 * sqrt(obs * (1 - obs) / n)))]
  } else bins <- NULL

  list(summary = summ, curve = curve, bins = bins)
}

# --------------------------------------------------------------------------- #
# 2. Net benefit (Vickers & Elkin, 2006)
#    NB(t) = TP/N - (FP/N) * t/(1-t);  NB_all(t) = prev - (1-prev)*t/(1-t)
# --------------------------------------------------------------------------- #
net_benefit_curve <- function(y, p, thresholds) {
  keep <- !is.na(y) & !is.na(p)
  y <- as.integer(y[keep]); p <- as.numeric(p[keep])
  n <- length(y)
  if (n < 100 || sum(y) == 0) return(NULL)

  cnt  <- counts_at_thresholds(y, p, thresholds)
  w    <- thresholds / (1 - thresholds)
  prev <- mean(y)

  data.table(
    threshold      = thresholds,
    net_benefit    = cnt$tp / n - (cnt$fp / n) * w,
    nb_treat_all   = prev - (1 - prev) * w,
    nb_treat_none  = 0,
    alert_rate     = cnt$n_alert / n,
    prevalence     = prev
  )
}

# --------------------------------------------------------------------------- #
# 3. Operating point at a fixed sensitivity
#    Threshold = the largest cut-off achieving at least the target sensitivity.
# --------------------------------------------------------------------------- #
operating_point <- function(y, p, target_sens, hours_per_row = 1) {
  keep <- !is.na(y) & !is.na(p)
  y <- as.integer(y[keep]); p <- as.numeric(p[keep])
  n  <- length(y); n_pos <- sum(y); n_neg <- n - n_pos
  if (n_pos < 5 || n_neg < 5) return(NULL)

  # P(p >= thr | y = 1) >= target  ->  thr = quantile of positives
  thr <- as.numeric(stats::quantile(p[y == 1], probs = 1 - target_sens,
                                    type = 1, names = FALSE))

  alert <- p >= thr
  tp <- sum(alert & y == 1); fp <- sum(alert & y == 0)
  fn <- n_pos - tp;          tn <- n_neg - fp

  sens <- tp / n_pos
  spec <- tn / n_neg
  ppv  <- if ((tp + fp) > 0) tp / (tp + fp) else NA_real_
  npv  <- if ((tn + fn) > 0) tn / (tn + fn) else NA_real_

  patient_days <- n * hours_per_row / 24

  data.table(
    target_sens        = target_sens,
    threshold          = thr,
    n_hours            = n,
    n_event_hours      = n_pos,
    patient_days       = patient_days,
    sensitivity        = sens,
    specificity        = spec,
    ppv                = ppv,
    npv                = npv,
    nne                = if (!is.na(ppv) && ppv > 0) 1 / ppv else NA_real_,
    alert_hours        = tp + fp,
    alert_rate         = (tp + fp) / n,
    false_alert_hours  = fp,
    fa_per_patient_day = fp / patient_days,
    alerts_per_pt_day  = (tp + fp) / patient_days,
    alert_burden_ratio = if (tp > 0) fp / tp else Inf
  )
}

# --------------------------------------------------------------------------- #
# 4. Per-patient (event-level) detection at a given threshold
#
#    A stay is a sepsis stay if it contributes at least one event hour.
#    Detection windows follow the Utility Score convention (Reyna et al., 2020):
#      - actionable window : [onset - 6h, onset]
#      - any pre-onset     : first alert at or before onset
#    A contiguous run of alert hours counts as ONE alarm episode; an episode
#    is a true alarm if it overlaps [onset - 6h, onset + 3h].
# --------------------------------------------------------------------------- #
patient_level_metrics <- function(dt, pred_col, thr) {
  d <- dt[!is.na(get(pred_col)), .(stay_id, hour,
                                    ev    = as.integer(event == 1),
                                    onset = as.numeric(onset_hour),
                                    alert = as.integer(get(pred_col) >= thr))]
  if (nrow(d) == 0) return(NULL)
  setorder(d, stay_id, hour)

  # alarm episodes: contiguous runs of alert hours within a stay
  d[, run := rleid(stay_id, alert)]
  ep <- d[alert == 1L, .(stay_id = stay_id[1],
                          h_start = min(hour), h_end = max(hour),
                          onset   = onset[1]), by = run]
  ep[, true_alarm := !is.na(onset) &
       h_end >= (onset + UTILITY_TP_MIN) & h_start <= (onset + UTILITY_TP_MAX)]

  stay <- d[, .(
    is_sepsis = as.integer(any(ev == 1L)),
    onset     = as.numeric(onset[1]),
    n_hours   = .N,
    any_alert = as.integer(any(alert == 1L))
  ), by = stay_id]

  # first alert inside the actionable window [onset - 6 h, onset], and
  # first alert at any point at or before onset
  win <- d[alert == 1L & !is.na(onset) & hour >= (onset + UTILITY_TP_MIN) &
             hour <= onset,
           .(first_win_alert = as.numeric(min(hour))), by = stay_id]
  pre <- d[alert == 1L & !is.na(onset) & hour <= onset,
           .(first_pre_alert = as.numeric(min(hour))), by = stay_id]

  stay <- merge(stay, win, by = "stay_id", all.x = TRUE)
  stay <- merge(stay, pre, by = "stay_id", all.x = TRUE)
  stay[, detected_window := as.integer(is_sepsis == 1L & !is.na(first_win_alert))]
  stay[, detected_pre    := as.integer(is_sepsis == 1L & !is.na(first_pre_alert))]
  stay[, lead_time_h     := ifelse(detected_pre == 1L, onset - first_pre_alert,
                                   NA_real_)]

  n_sepsis    <- stay[, sum(is_sepsis == 1L)]
  n_nonsepsis <- stay[, sum(is_sepsis == 0L)]
  n_alerted   <- stay[, sum(any_alert == 1L)]
  patient_days <- stay[, sum(n_hours)] / 24

  fa_episodes <- ep[true_alarm == FALSE, .N]
  tp_episodes <- ep[true_alarm == TRUE,  .N]

  lt <- stay[detected_pre == 1L & is.finite(lead_time_h), lead_time_h]

  list(
    summary = data.table(
      threshold             = thr,
      n_stays               = nrow(stay),
      n_sepsis_stays        = n_sepsis,
      n_nonsepsis_stays     = n_nonsepsis,
      patient_days          = patient_days,
      # event-level sensitivity
      episode_sens_window   = if (n_sepsis > 0)
        stay[, sum(detected_window)] / n_sepsis else NA_real_,
      episode_sens_pre      = if (n_sepsis > 0)
        stay[, sum(detected_pre)] / n_sepsis else NA_real_,
      # event-level PPV and false-alarm exposure
      stays_alerted         = n_alerted,
      frac_stays_alerted    = n_alerted / nrow(stay),
      frac_nonsepsis_alerted = if (n_nonsepsis > 0)
        stay[is_sepsis == 0L, sum(any_alert)] / n_nonsepsis else NA_real_,
      patient_ppv           = if (n_alerted > 0)
        stay[, sum(detected_pre)] / n_alerted else NA_real_,
      # alarm episodes
      n_alarm_episodes      = nrow(ep),
      n_false_alarm_episodes = fa_episodes,
      n_true_alarm_episodes  = tp_episodes,
      fa_episodes_per_pt_day = fa_episodes / patient_days,
      # lead time
      lead_median_h         = if (length(lt) > 0) median(lt) else NA_real_,
      lead_q25_h            = if (length(lt) > 0) as.numeric(quantile(lt, .25)) else NA_real_,
      lead_q75_h            = if (length(lt) > 0) as.numeric(quantile(lt, .75)) else NA_real_,
      frac_lead_ge_3h       = if (n_sepsis > 0) sum(lt >= 3) / n_sepsis else NA_real_,
      frac_lead_ge_6h       = if (n_sepsis > 0) sum(lt >= 6) / n_sepsis else NA_real_
    ),
    lead_times = lt
  )
}
```

## Data Assembly

As in stage 07, the primary-model predictions are joined to the comparator
predictions on `stay_id` and `hour`, for Variants A--C. Only the unrestricted
evaluation window is used. Under the onset rule of Variants A--C, onset is
placed at the suspicion-of-infection time, so the treatment-anchored window,
which ends before the first antibiotic or culture, contains no event hours.
No sensitivity-based threshold can be set in it.

```{r assemble}
load_variant <- function(vid) {
  prim_path <- file.path(DATA_DIR, paste0("05_predictions_primary_",     vid, ".parquet"))
  comp_path <- file.path(DATA_DIR, paste0("06_predictions_comparators_", vid, ".parquet"))
  if (!file.exists(prim_path)) return(NULL)

  ph <- as.data.table(arrow::read_parquet(prim_path))
  if (file.exists(comp_path)) {
    cp <- as.data.table(arrow::read_parquet(comp_path))
    add <- c("stay_id", "hour", setdiff(names(cp), names(ph)))
    ph <- merge(ph, cp[, ..add], by = c("stay_id", "hour"), all.x = TRUE)
  }
  ph
}

variants <- names(LABEL_VARIANTS)
ph_list  <- setNames(lapply(variants, load_variant), variants)
ph_list  <- ph_list[!vapply(ph_list, is.null, logical(1))]

for (vid in names(ph_list)) {
  v2_log(sprintf("Variant %s: %s test hours, %s event hours, %s stays",
                 vid,
                 format(nrow(ph_list[[vid]]), big.mark = ","),
                 format(sum(ph_list[[vid]]$event == 1), big.mark = ","),
                 format(uniqueN(ph_list[[vid]]$stay_id), big.mark = ",")))
}
```

## Calibration Curves

Each panel plots the observed event rate against the predicted value, and the
diagonal marks perfect calibration. The curve is the flexible calibration curve
from the spline model above. Its shaded band is a pointwise 95% confidence
interval for the curve. The points are 20 groups formed at quantiles of the
predicted value, so each holds about the same number of person-hours. Each
point is drawn with a normal-approximation (Wald) binomial 95% interval. In
every panel the outcome is onset in the hour, the event each model predicts.

The axes follow each panel's own range of predicted values, because those
ranges differ by orders of magnitude between models. Deviations should
therefore be compared between panels with care. The same distance from the
diagonal can look large in a narrow panel and small in a wide one.

Table 1 reports the calibration measures:

- The primary model's ICI, E50 and E90 are about an order of magnitude below
  the event rate.
- The GBT and NEWS2 rows show values about three orders of magnitude larger,
  and strongly negative scaled Brier scores. This reflects their scales: their
  outputs are not calibrated probabilities (see the Introduction).
- The scaled Brier score is $1 - \text{Brier}/\text{Brier}_{\text{null}}$,
  where the null model predicts the prevalence for every hour. A value near 0
  means no improvement over that model. A large negative value means the
  squared error is many times that of the null model.
- For NEWS2, the intercept and slope describe a logistic recalibration of the
  rescaled score, not the calibration of a probability.

```{r calibration, fig.height=4.6, fig.cap="Flexible calibration curves on the temporal test set. The dashed line is perfect calibration, points are twenty equal-sized bins of predicted risk, and the shaded band is the 95 per cent confidence interval. Each panel carries its own scale because the predicted-risk ranges differ by orders of magnitude."}
calib_summ  <- list()
calib_curve <- list()
calib_bins  <- list()

for (vid in names(ph_list)) {
  ph <- ph_list[[vid]]
  y  <- as.integer(ph$event == 1)
  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ph)) next
    res <- calibration_curve(y, as.numeric(ph[[pcol]]))
    if (is.null(res$summary)) next
    key <- paste(vid, mod, sep = "-")
    calib_summ[[key]]  <- cbind(variant = vid, model = mod, res$summary)
    if (mod %in% PLOT_MODELS) {
      calib_curve[[key]] <- cbind(variant = vid, model = mod, res$curve)
      if (!is.null(res$bins))
        calib_bins[[key]] <- cbind(variant = vid, model = mod, res$bins)
    }
  }
}

calib_summ_dt <- rbindlist(calib_summ, fill = TRUE)
save_parquet(calib_summ_dt, "11_calibration_summary", DATA_DIR)
save_csv(calib_summ_dt,     "11_calibration_summary", DATA_DIR)

curve_dt <- rbindlist(calib_curve, fill = TRUE)
bins_dt  <- rbindlist(calib_bins,  fill = TRUE)

if (nrow(curve_dt) > 0) {
  curve_dt[, model_lab := factor(MODEL_LABELS[model],
                                  levels = MODEL_LABELS[PLOT_MODELS])]
  bins_dt[,  model_lab := factor(MODEL_LABELS[model],
                                  levels = MODEL_LABELS[PLOT_MODELS])]
  curve_dt[, variant_lab := paste0("Variant ", variant)]
  bins_dt[,  variant_lab := paste0("Variant ", variant)]

  # Predicted-risk scales differ by orders of magnitude between the pooled
  # logistic model (calibrated to a 0.2% hourly event rate) and GBT, so every
  # panel needs its own scale. facet_grid(scales = "free") only frees axes per
  # row/column, so facet_wrap over the model-variant interaction is used
  # instead. The dashed diagonal is perfect calibration in every panel.
  mk_panel <- function(d) {
    d[, panel_lab := factor(
      paste0(model_lab, ", Variant ", variant),
      levels = unlist(lapply(MODEL_LABELS[PLOT_MODELS], function(m)
        paste0(m, ", Variant ", sort(unique(curve_dt$variant))))))]
    d[!is.na(panel_lab)]
  }
  curve_dt <- mk_panel(curve_dt)
  bins_dt  <- mk_panel(bins_dt)

  p_cal <- ggplot(curve_dt, aes(pred, obs, colour = model_lab, fill = model_lab)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey40", linewidth = 0.4) +
    geom_ribbon(aes(ymin = obs_low, ymax = obs_high), alpha = 0.18,
                colour = NA) +
    geom_line(linewidth = 0.7) +
    geom_point(data = bins_dt, aes(pred, obs), size = 1.0, alpha = 0.7,
               inherit.aes = TRUE) +
    facet_wrap(~ panel_lab, scales = "free", ncol = 3) +
    expand_limits(x = 0, y = 0) +
    labs(x = "Predicted risk of sepsis onset within 6 h",
         y = "Observed event rate",
         colour = NULL, fill = NULL) +
    theme_bw() +
    theme(legend.position = "none",
          panel.grid.minor = element_blank(),
          strip.text = element_text(size = 7.5),
          axis.text = element_text(size = 6.5))

  n_panels <- length(unique(curve_dt$panel_lab))
  ggsave(file.path(FIGURES_DIR, "11_calibration_curves.png"), p_cal,
         width = 9, height = 2.4 * ceiling(n_panels / 3) + 0.4, dpi = 150)
  print(p_cal)
}
```

```{r calibration-table}
if (nrow(calib_summ_dt) > 0) {
  tab <- calib_summ_dt[model %in% PLOT_MODELS,
    .(variant, model = MODEL_LABELS[model],
      Intercept = round(calib_int, 3), Slope = round(calib_slope, 3),
      ICI = signif(ici, 3), E50 = signif(e50, 3), E90 = signif(e90, 3),
      Brier = signif(brier, 4), `Brier scaled` = round(brier_scaled, 4))]
  knitr::kable(tab[order(variant, model)],
    caption = paste("Calibration summary. ICI = Integrated Calibration Index",
                    "(mean absolute difference between predicted risk and the",
                    "flexible calibration curve). E50/E90 = median and 90th",
                    "percentile of that difference. Scaled Brier = 1 - Brier /",
                    "Brier(null); values at or below 0 mean the model is no",
                    "better than predicting the prevalence for everyone."))
}
```

## Decision-Curve Analysis

For each variant, net benefit is evaluated on 80 thresholds from one tenth to
ten times the event prevalence, with the prevalence itself added. Choosing a
threshold probability $t$ implies that up to $(1-t)/t$ false alerts are
acceptable for each true detection. A threshold of 0.01, for example, accepts
at most 99. At an hourly prevalence of 0.2 per cent, a model alerting at
random would raise about 500 alerts per true detection. Thresholds near the
prevalence are therefore the range in which an hourly alert could add
benefit. Which threshold is clinically appropriate depends on the relative
harms of missed and false alerts, and this grid does not settle that.

Figure 2 shows the primary model, the GBT and NEWS2, with the treat-all and
treat-none strategies as references. To keep the region around zero readable,
values below $-1.75$ times the prevalence are not drawn. A curve that leaves
the bottom of a panel has lower net benefit than treat-none from that
threshold onwards.

Table 2 gives each model's net benefit at the threshold equal to the
prevalence. At that threshold, treat-all and treat-none both have a net
benefit of exactly zero, and a perfect model would reach the prevalence. The
standardised net benefit (net benefit divided by the prevalence) is therefore
the share of that maximum which the model achieves. The alert-rate span is the
change in the alert rate across the whole grid. A span of zero means that the
model's alerts are the same at every threshold on the grid. This happens for
two reasons:

- The GBT and Cox scores all lie above the highest threshold, so these models
  alert in every hour and their curves coincide with treat-all.
- NEWS2, qSOFA and SIRS take values in steps (multiples of 1/20, 1/3 and 1/4)
  that are coarser than the whole grid, so the alert depends only on whether
  the score is above zero.

Only the primary model can be tuned within this range.

```{r dca, fig.height=4.6, fig.cap="Decision curves on the temporal test set. Net benefit against threshold probability, with both default strategies shown. Curves are clipped below 1.75 times prevalence so the near-zero region stays legible; a line leaving the bottom of a panel is worse than treating nobody from that threshold onward."}
nb_list <- list()

for (vid in names(ph_list)) {
  ph   <- ph_list[[vid]]
  y    <- as.integer(ph$event == 1)
  prev <- mean(y)
  thr_grid <- unique(sort(c(
    seq(prev / 10, prev * 10, length.out = 80),
    prev
  )))
  thr_grid <- thr_grid[thr_grid > 0 & thr_grid < 0.5]

  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ph)) next
    nb <- net_benefit_curve(y, as.numeric(ph[[pcol]]), thr_grid)
    if (is.null(nb)) next
    nb_list[[paste(vid, mod, sep = "-")]] <- cbind(variant = vid, model = mod, nb)
  }
}

nb_dt <- rbindlist(nb_list, fill = TRUE)
save_parquet(nb_dt, "11_net_benefit", DATA_DIR)

if (nrow(nb_dt) > 0) {
  # Treat-all and treat-none are plotted as named series so that the reader can
  # see directly which model curves lie on top of a default strategy.
  ref <- unique(nb_dt[, .(variant, threshold, nb_treat_all)])
  series <- rbind(
    nb_dt[model %in% PLOT_MODELS,
          .(variant, threshold, nb = net_benefit, series = MODEL_LABELS[model])],
    ref[,  .(variant, threshold, nb = nb_treat_all, series = "Treat all")],
    ref[,  .(variant, threshold, nb = 0,           series = "Treat none")]
  )
  ser_levels <- c(unname(MODEL_LABELS[PLOT_MODELS]), "Treat all", "Treat none")
  series[, series := factor(series, levels = ser_levels)]
  series[, variant_lab := paste0("Variant ", variant)]

  # The maximum attainable net benefit is the prevalence, so the informative
  # band is a small multiple of it. Curves are clipped below -1.75 x prevalence
  # so that the near-zero region stays legible; a line that leaves the bottom
  # of a panel is worse than treating nobody from that threshold onwards.
  series <- merge(series, nb_dt[, .(prev = prevalence[1]), by = variant],
                  by = "variant", all.x = TRUE)
  series[, nb_clip := fifelse(nb < -1.75 * prev, NA_real_, nb)]

  p_dca <- ggplot(series, aes(threshold, nb_clip, colour = series,
                              linetype = series, linewidth = series)) +
    geom_line(na.rm = TRUE) +
    facet_wrap(~ variant_lab, nrow = 1, scales = "free") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_colour_manual(values = setNames(
      c(scales::hue_pal()(length(PLOT_MODELS)), "grey25", "grey55"), ser_levels)) +
    scale_linetype_manual(values = setNames(
      c(rep("solid", length(PLOT_MODELS)), "dotdash", "dashed"), ser_levels)) +
    scale_linewidth_manual(values = setNames(
      c(rep(0.75, length(PLOT_MODELS)), 0.5, 0.5), ser_levels)) +
    labs(x = "Threshold probability", y = "Net benefit",
         colour = NULL, linetype = NULL, linewidth = NULL) +
    theme_bw() +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  ggsave(file.path(FIGURES_DIR, "11_decision_curves.png"), p_dca,
         width = 9, height = 4, dpi = 150)
  print(p_dca)
}
```

```{r dca-table}
# Net benefit at the threshold equal to the event prevalence. At t = prevalence
# both defaults have net benefit exactly 0, so any positive value is a real
# gain over both, and the maximum attainable net benefit is the prevalence
# itself. Standardised net benefit (Vickers et al., 2019) = NB / prevalence,
# i.e. the fraction of the attainable benefit that the model captures.
if (nrow(nb_dt) > 0) {
  nb_prev <- nb_dt[, {
    i <- which.min(abs(threshold - prevalence[1]))
    .(threshold   = threshold[i],
      prevalence  = prevalence[1],
      net_benefit = net_benefit[i],
      nb_treat_all = nb_treat_all[i],
      snb         = net_benefit[i] / prevalence[1],
      alert_rate  = alert_rate[i],
      # does the alert rate respond to the threshold at all?
      alert_rate_range = max(alert_rate) - min(alert_rate))
  }, by = .(variant, model)]

  save_parquet(nb_prev, "11_net_benefit_at_prevalence", DATA_DIR)
  save_csv(nb_prev,     "11_net_benefit_at_prevalence", DATA_DIR)

  knitr::kable(
    nb_prev[order(variant, model),
      .(variant, model = MODEL_LABELS[model],
        Threshold = signif(threshold, 3),
        `Net benefit` = signif(net_benefit, 3),
        `Treat all` = signif(nb_treat_all, 2),
        `Std NB` = round(snb, 3),
        `Alert rate` = round(alert_rate, 3),
        `Alert rate span` = round(alert_rate_range, 3))],
    caption = paste("Net benefit at a threshold probability equal to the event",
                    "prevalence. Both default strategies have net benefit 0 at",
                    "this threshold, so Std NB (net benefit divided by the",
                    "prevalence) is the fraction of the attainable benefit",
                    "captured. 'Alert rate span' is the change in alert rate",
                    "across the whole threshold grid: a span of 0 means the",
                    "score cannot be tuned anywhere in the plausible range and",
                    "the decision curve collapses onto a default strategy."))
}
```

## Operating Points at Fixed Sensitivity

For each model and each target sensitivity
(`r paste(SENS_TARGETS, collapse = ", ")`), the threshold is the largest value
at which the per-hour sensitivity reaches the target. Because NEWS2 takes
discrete values, its achieved sensitivity can exceed the target, as the table
shows.

The table reports the `r HEADLINE_SENS * 100`% target, and Figure 3 adds the
other two. Its columns are calculated at the selected threshold:

- sensitivity and specificity (`Sens`, `Spec`);
- PPV;
- the number needed to evaluate (NNE $= 1/\text{PPV}$), the number of alerted
  person-hours per true event hour;
- alerted person-hours and false-alert person-hours per patient-day
  (`Alerts/pt-day`, `FA/pt-day`), with patient-days equal to the evaluated
  person-hours divided by 24;
- the ratio of false-positive to true-positive hours (`FP:TP`).

NNE and FP:TP differ by one by construction, since NNE counts the true alert
as well. An alert that stays on for several hours counts once for each hour in
these per-hour measures. Distinct alert episodes are counted in the
per-patient section.

```{r operating-points}
op_list <- list()

for (vid in names(ph_list)) {
  ph <- ph_list[[vid]]
  y  <- as.integer(ph$event == 1)
  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ph)) next
    p <- as.numeric(ph[[pcol]])
    for (ts in SENS_TARGETS) {
      op <- operating_point(y, p, ts)
      if (is.null(op)) next
      op_list[[paste(vid, mod, ts, sep = "-")]] <-
        cbind(variant = vid, model = mod, op)
    }
  }
}

op_dt <- rbindlist(op_list, fill = TRUE)
save_parquet(op_dt, "11_operating_points", DATA_DIR)
save_csv(op_dt,     "11_operating_points", DATA_DIR)

if (nrow(op_dt) > 0) {
  show <- op_dt[target_sens == HEADLINE_SENS & model %in% PLOT_MODELS,
    .(variant, model = MODEL_LABELS[model],
      Threshold   = signif(threshold, 3),
      Sens        = round(sensitivity, 3),
      Spec        = round(specificity, 3),
      PPV         = signif(ppv, 3),
      NNE         = round(nne, 0),
      `Alerts/pt-day` = round(alerts_per_pt_day, 2),
      `FA/pt-day` = round(fa_per_patient_day, 2),
      `FP:TP`     = round(alert_burden_ratio, 0))]
  knitr::kable(show[order(variant, model)],
    caption = sprintf(paste("Operating characteristics at a fixed per-hour",
                            "sensitivity of %.0f%%. NNE = number of alerted",
                            "person-hours per true event hour. FA/pt-day =",
                            "false-alert hours per patient-day."),
                      HEADLINE_SENS * 100))
}
```

```{r sens-tradeoff, fig.height=3.6, fig.cap="False alerts per patient-day against per-hour sensitivity, at sensitivity targets of 50, 80 and 90 per cent."}
if (nrow(op_dt) > 0) {
  tr <- op_dt[model %in% PLOT_MODELS]
  tr[, model_lab := factor(MODEL_LABELS[model], levels = MODEL_LABELS[PLOT_MODELS])]
  tr[, variant_lab := paste0("Variant ", variant)]

  p_tr <- ggplot(tr, aes(sensitivity, fa_per_patient_day,
                         colour = model_lab, group = model_lab)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.6) +
    facet_wrap(~ variant_lab, nrow = 1) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(x = "Per-hour sensitivity", y = "False alerts per patient-day",
         colour = NULL) +
    theme_bw() +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  ggsave(file.path(FIGURES_DIR, "11_false_alarms_vs_sensitivity.png"), p_tr,
         width = 9, height = 3.4, dpi = 150)
  print(p_tr)
}
```

## Per-Patient (Event-Level) Detection

The measures above are per person-hour. In the discrete-time competing-risks
coding, `event == 1` marks the onset hour itself, so each label-positive stay
contributes exactly one event hour. States 2 and 3 are discharge alive and
death without onset. Per-hour sensitivity is therefore strict: an alert counts
as a true positive only if it occurs in the onset hour. The per-patient
measures ask a different and complementary question, namely whether a stay was
alerted at all before its onset. They are computed at the thresholds of the
operating-point table:

- **Sensitivity in the 6-hour window.** The proportion of label-positive stays
  with an alert from 6 hours before onset to onset (`Episode sens (6h)`).
- **Sensitivity at any time before onset.** The proportion with an alert at
  any time from the start of the stay up to and including the onset hour
  (`Episode sens (any pre)`). This measure has no limit on how far ahead the
  alert may come, so an alert early in the stay counts even if it was
  unrelated to the later onset. The 6-hour measure is the stricter of the two.
- **Patient PPV.** The proportion of alerted stays that are label-positive and
  alerted at or before onset.
- **Stays alerted.** The percentages of all stays and of label-negative stays
  (`% non-sepsis alerted`) with at least one alert.
- **False-alert episodes.** An episode is a run of consecutive alert hours
  within a stay. It is false if it does not overlap the window from 6 hours
  before to 3 hours after onset, the true-positive window of the Utility Score
  (Reyna et al., 2020). The panel ends at onset, so in practice the window
  reaches only to the onset hour. Every episode in a label-negative stay is
  false. Episodes are counted per patient-day.

The column headed "Sepsis stays" counts label-positive stays, whose onset is
defined by the variant's label rather than by a clinical diagnosis. At the
`r HEADLINE_SENS * 100`% per-hour operating point, almost all stays,
label-negative ones included, receive at least one alert. That is the main
limitation on how usable these operating points would be in practice.

```{r patient-level}
pat_list  <- list()
lead_list <- list()

for (vid in names(ph_list)) {
  ph <- ph_list[[vid]]
  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ph)) next
    for (ts in SENS_TARGETS) {
      thr <- op_dt[variant == vid & model == mod & target_sens == ts, threshold]
      if (length(thr) == 0) next
      res <- tryCatch(patient_level_metrics(ph, pcol, thr[1]),
                      error = function(e) { v2_log(sprintf(
                        "patient_level_metrics failed (%s/%s/%.2f): %s",
                        vid, mod, ts, e$message), level = "WARN"); NULL })
      if (is.null(res)) next
      key <- paste(vid, mod, ts, sep = "-")
      pat_list[[key]] <- cbind(variant = vid, model = mod,
                                target_sens = ts, res$summary)
      if (ts == HEADLINE_SENS && mod %in% PLOT_MODELS && length(res$lead_times) > 0)
        lead_list[[key]] <- data.table(variant = vid, model = mod,
                                        lead_time_h = res$lead_times)
    }
  }
}

pat_dt <- rbindlist(pat_list, fill = TRUE)
save_parquet(pat_dt, "11_patient_level_metrics", DATA_DIR)
save_csv(pat_dt,     "11_patient_level_metrics", DATA_DIR)

if (nrow(pat_dt) > 0) {
  show_p <- pat_dt[target_sens == HEADLINE_SENS & model %in% PLOT_MODELS,
    .(variant, model = MODEL_LABELS[model],
      `Sepsis stays`      = n_sepsis_stays,
      `Episode sens (6h)` = round(episode_sens_window, 3),
      `Episode sens (any pre)` = round(episode_sens_pre, 3),
      `Patient PPV`       = signif(patient_ppv, 3),
      `% stays alerted`   = round(100 * frac_stays_alerted, 1),
      `% non-sepsis alerted` = round(100 * frac_nonsepsis_alerted, 1),
      `FA episodes/pt-day` = round(fa_episodes_per_pt_day, 2))]
  knitr::kable(show_p[order(variant, model)],
    caption = paste("Per-patient detection at the same thresholds as the",
                    "operating-point table. Episode sens (6h) = fraction of",
                    "septic stays alerted inside the actionable window",
                    "[onset - 6 h, onset]. Episode sens (any pre) = alerted at",
                    "any point before onset. FA episodes = contiguous runs of",
                    "alert hours that do not overlap [onset - 6 h, onset + 3 h]."))
}
```

### Per-Hour versus Per-Patient, Side by Side

This table sets the per-hour and per-patient measures of the same
`r HEADLINE_SENS * 100`% operating point side by side. `Sens (patient)` is the
same measure as `Episode sens (any pre)` above. In these results the
per-patient sensitivity and PPV are well above their per-hour counterparts:

- A single alert at any time before onset is enough to detect a stay.
- An alert lasting several hours counts as one alerted stay, not as many
  alerted hours.

A favourable per-hour figure can therefore coexist with a high proportion of
stays receiving false alerts. The table reports both false-alert hours and
false-alert episodes per patient-day for this reason.

```{r hour-vs-patient}
if (nrow(op_dt) > 0 && nrow(pat_dt) > 0) {
  side <- merge(
    op_dt[target_sens == HEADLINE_SENS,
          .(variant, model, target_sens,
            hour_sens = sensitivity, hour_ppv = ppv,
            hour_burden = alert_burden_ratio, fa_per_patient_day)],
    pat_dt[target_sens == HEADLINE_SENS,
           .(variant, model, target_sens,
             pt_sens = episode_sens_pre, pt_ppv = patient_ppv,
             fa_episodes_per_pt_day)],
    by = c("variant", "model", "target_sens"))

  save_parquet(side, "11_hour_vs_patient", DATA_DIR)
  save_csv(side,     "11_hour_vs_patient", DATA_DIR)

  knitr::kable(
    side[model %in% PLOT_MODELS][order(variant, model),
      .(variant, model = MODEL_LABELS[model],
        `Sens (hour)` = round(hour_sens, 3),
        `Sens (patient)` = round(pt_sens, 3),
        `PPV (hour)` = signif(hour_ppv, 3),
        `PPV (patient)` = signif(pt_ppv, 3),
        `FA hours/pt-day` = round(fa_per_patient_day, 2),
        `FA episodes/pt-day` = round(fa_episodes_per_pt_day, 2))],
    caption = paste("Per-hour and per-patient views of the same operating",
                    "point. Patient-level sensitivity is much higher than",
                    "per-hour sensitivity because one alert anywhere before",
                    "onset counts the whole episode as detected; patient-level",
                    "PPV is correspondingly higher than per-hour PPV because a",
                    "multi-hour alert run counts once."))
}
```

## Lead Time to Onset

Lead time is the interval from a stay's first alert at or before onset to the
onset hour. It is computed at the `r HEADLINE_SENS * 100`% operating threshold
for the label-positive stays that receive such an alert. The first alert can
come at any time in the stay, so a long lead time may reflect an alert that
was not related to the later onset. A lead time shows that an alert came early
enough to allow a response, not that a response would have helped.

The measures in the table have different denominators:

- The median and interquartile range (`IQR`, lower and upper quartiles)
  describe only the stays that were detected.
- `detected_pre` is the proportion of label-positive stays that were detected.
- `>= 3 h` and `>= 6 h` are proportions of all label-positive stays alerted at
  least that many hours ahead.

Figure 4 shows lead times from 0 to 48 hours as counts of stays. Longer lead
times are omitted from the figure but included in the table. The y-axes differ
between panels, and Variant C has about twice as many label-positive stays as
A and B, so the shapes of the distributions are easier to compare than their
heights. Similar medians can hide different distributions, which is why the
quartiles and proportions are also given. Lead time also depends on how the
onset is timed. Under Variants A--C, onset is placed at the
suspicion-of-infection time, so these lead times are not comparable with
published lead times computed under another onset rule.

```{r lead-time, fig.height=3.6, fig.cap="Distribution of lead time before onset at the 80 per cent sensitivity operating point, truncated at 48 hours."}
lead_dt <- rbindlist(lead_list, fill = TRUE)

lead_summary <- pat_dt[target_sens == HEADLINE_SENS,
  .(variant, model, n_sepsis_stays,
    detected_pre = round(episode_sens_pre, 3),
    lead_median_h, lead_q25_h, lead_q75_h,
    frac_lead_ge_3h = round(frac_lead_ge_3h, 3),
    frac_lead_ge_6h = round(frac_lead_ge_6h, 3))]
save_parquet(lead_summary, "11_lead_time_summary", DATA_DIR)
save_csv(lead_summary,     "11_lead_time_summary", DATA_DIR)

if (nrow(lead_dt) > 0) {
  lead_dt[, model_lab := factor(MODEL_LABELS[model], levels = MODEL_LABELS[PLOT_MODELS])]
  lead_dt[, variant_lab := paste0("Variant ", variant)]

  p_lead <- ggplot(lead_dt[lead_time_h >= 0 & lead_time_h <= 48],
                   aes(lead_time_h, fill = model_lab)) +
    geom_histogram(binwidth = 2, position = "identity", alpha = 0.55,
                   colour = NA) +
    facet_wrap(~ variant_lab, nrow = 1, scales = "free_y") +
    labs(x = "Lead time before sepsis onset (h)",
         y = "Septic stays", fill = NULL) +
    theme_bw() +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  ggsave(file.path(FIGURES_DIR, "11_lead_time_distribution.png"), p_lead,
         width = 9, height = 3.4, dpi = 150)
  print(p_lead)
}

knitr::kable(
  lead_summary[model %in% PLOT_MODELS][order(variant, model),
    .(variant, model = MODEL_LABELS[model], n_sepsis_stays, detected_pre,
      `Median lead (h)` = lead_median_h,
      `IQR` = paste0(lead_q25_h, "-", lead_q75_h),
      `>= 3 h` = frac_lead_ge_3h, `>= 6 h` = frac_lead_ge_6h)],
  caption = sprintf(paste("Lead time at the %.0f%% sensitivity operating",
                          "point. '>= 3 h' and '>= 6 h' are fractions of ALL",
                          "septic stays (not just detected ones) alerted at",
                          "least that far ahead of onset."),
                    HEADLINE_SENS * 100))
```

## External Validation: Clinical Utility on eICU-CRD

The operating-point and per-patient calculations are also applied to the
eICU-CRD predictions of `08_external_validation.Rmd`. These come from the
Sepsis-3 arm on the microbiology-covered stays, for the primary model, the GBT
and NEWS2. The onsets follow the same rule for each variant as internally, as
derived in stage 08.

The thresholds here are selected on the external cohort itself, to reach
`r HEADLINE_SENS * 100`% per-hour sensitivity among eICU-CRD event hours.
They are not the internal thresholds applied unchanged. These results
therefore show what the models would achieve if the threshold were reset at
the new site. They are not a fixed-threshold test of transportability, and
the selection uses the external labels, which may make them optimistic.

No variant reaches the `r MIN_EXTERNAL_EVENTS`-event minimum used in stage
08. The label-positive stays number only in the tens, as the table shows, and
no intervals are reported here, so the results are descriptive. A
per-patient sensitivity of 1.000 on a few dozen stays is compatible with a
much lower true value. The external PPVs are far lower than the internal ones,
corresponding to thousands of alerted person-hours per true event hour (NNE).
This low external event rate is examined in stage 08.

```{r external}
ext_op  <- list()
ext_pat <- list()

for (vid in variants) {
  ext_path <- file.path(DATA_DIR, paste0("08_eicu_predictions_", vid, ".parquet"))
  if (!file.exists(ext_path)) next
  ext <- as.data.table(arrow::read_parquet(ext_path))
  y   <- as.integer(ext$event == 1)
  if (sum(y) < 10) next

  for (mod in names(MODEL_NAMES)) {
    pcol <- MODEL_NAMES[[mod]]
    if (!pcol %in% names(ext)) next
    op <- operating_point(y, as.numeric(ext[[pcol]]), HEADLINE_SENS)
    if (is.null(op)) next
    ext_op[[paste(vid, mod, sep = "-")]] <- cbind(variant = vid, model = mod,
                                                   cohort = "eICU", op)
    res <- tryCatch(patient_level_metrics(ext, pcol, op$threshold),
                    error = function(e) NULL)
    if (!is.null(res))
      ext_pat[[paste(vid, mod, sep = "-")]] <-
        cbind(variant = vid, model = mod, cohort = "eICU",
              target_sens = HEADLINE_SENS, res$summary)
  }
  rm(ext); gc(verbose = FALSE)
}

ext_op_dt  <- rbindlist(ext_op,  fill = TRUE)
ext_pat_dt <- rbindlist(ext_pat, fill = TRUE)

if (nrow(ext_op_dt) > 0) {
  save_parquet(ext_op_dt,  "11_external_operating_points", DATA_DIR)
  save_csv(ext_op_dt,      "11_external_operating_points", DATA_DIR)
  knitr::kable(
    ext_op_dt[model %in% PLOT_MODELS][order(variant, model),
      .(variant, model = MODEL_LABELS[model],
        Threshold = signif(threshold, 3), Sens = round(sensitivity, 3),
        PPV = signif(ppv, 4), NNE = round(nne, 0),
        `FA/pt-day` = round(fa_per_patient_day, 2))],
    caption = sprintf(paste("eICU-CRD operating characteristics at the same",
                            "%.0f%% target sensitivity (thresholds re-derived",
                            "on the external cohort)."), HEADLINE_SENS * 100))
}
```

```{r external-patient}
if (nrow(ext_pat_dt) > 0) {
  save_parquet(ext_pat_dt, "11_external_patient_level", DATA_DIR)
  save_csv(ext_pat_dt,     "11_external_patient_level", DATA_DIR)
  knitr::kable(
    ext_pat_dt[model %in% PLOT_MODELS][order(variant, model),
      .(variant, model = MODEL_LABELS[model], n_sepsis_stays,
        `Episode sens (6h)` = round(episode_sens_window, 3),
        `Episode sens (any pre)` = round(episode_sens_pre, 3),
        `Patient PPV` = signif(patient_ppv, 3),
        `FA episodes/pt-day` = round(fa_episodes_per_pt_day, 2))],
    caption = "eICU-CRD per-patient detection at the external operating point.")
}
```

## Cross-Check Against Notebook 07

Net benefit is calculated here directly from the Vickers--Elkin formula,
whereas stage 07 uses the `dcurves` package. As a check, the stage-07 values
for the primary model and the GBT are compared with the values here. The
comparison uses the stage-07 thresholds that fall inside this notebook's
grid, each matched to the nearest grid point. Table 6 gives the number of
thresholds compared and the largest absolute difference. This notebook's
grid ends at ten times the prevalence. It overlaps only the lowest thresholds
of stage 07's grid, so one threshold is compared for Variants A and B and four
for Variant C. This makes it a check at a few points, not across the
whole curve. The small differences are consistent with matching each
stage-07 threshold to a nearby, not identical, grid point. A comparison on an
identical threshold vector would be needed to rule out any other source.

```{r crosscheck}
chk_rows <- list()
for (vid in variants) {
  f <- file.path(DATA_DIR, paste0("07_dca_", vid, ".parquet"))
  if (!file.exists(f)) next
  old <- as.data.table(arrow::read_parquet(f))
  old <- old[variable %in% c("pred_sepsis", "pred_gbt")]
  if (nrow(old) == 0) next
  old[, model := ifelse(variable == "pred_sepsis", "primary", "gbt")]

  new <- nb_dt[variant == vid & model %in% c("primary", "gbt")]
  if (nrow(new) == 0) next

  for (m in unique(old$model)) {
    # NB: the 07 DCA table carries a column literally called `n`, so local
    # objects here are named to avoid colliding with it inside `[.data.table`.
    o_dt   <- old[model == m]
    new_dt <- new[model == m]
    thr_new <- new_dt$threshold
    thr_lo  <- min(thr_new); thr_hi <- max(thr_new)
    # nearest new threshold to each old threshold within the shared range
    o_dt <- o_dt[threshold <= thr_hi & threshold >= thr_lo]
    if (nrow(o_dt) == 0) next
    idx <- vapply(o_dt$threshold, function(t) which.min(abs(thr_new - t)),
                  integer(1))
    chk_rows[[paste(vid, m)]] <- data.table(
      variant = vid, model = m, n_compared = nrow(o_dt),
      max_abs_diff = signif(max(abs(o_dt$net_benefit -
                                      new_dt$net_benefit[idx])), 3))
  }
}

if (length(chk_rows) > 0) {
  knitr::kable(rbindlist(chk_rows),
    caption = paste("Agreement between the dcurves output stored by notebook",
                    "07 and the net benefit recomputed here, over the shared",
                    "threshold range. Differences reflect only the threshold",
                    "grid offset, not a difference in definition."))
} else {
  cat("No overlapping threshold range between the 07 grid (0.01-0.20) and the",
      "prevalence-scaled grid used here; nothing to cross-check.\n")
}
```

## What These Metrics Add to Discrimination

AUROC measures discrimination. It does not assess calibration, the benefit of
acting at a particular threshold, alert burden or lead time.

Calibration assesses whether predicted probabilities agree with observed event
rates. A model can rank person-hours well and still give probabilities that
are systematically too high or too low. This matters as soon as a probability
threshold is used to guide decisions. Here the primary model is close to
calibrated. The GBT and NEWS2 outputs are not probabilities on the scale of
the event rate, so their calibration and net-benefit results describe that
scale rather than a probability forecast.

Net benefit compares acting on a model with the treat-all and treat-none
strategies. A model whose curve does not rise above both gives no additional
benefit at those thresholds, whatever its discrimination. Within the
prevalence-scaled grid, only the primary model's alerts vary with the
threshold.

The operating-point and per-patient tables show the cost of high sensitivity.
At `r HEADLINE_SENS * 100`% per-hour sensitivity, every model has a per-hour
PPV of well under 1 per cent. Almost every stay, label-negative stays
included, receives at least one alert. Externally, with thresholds reset on
the eICU-CRD cohort, the PPV is lower still, and the event counts are too
small for firm conclusions.

Lead time indicates whether an alert comes early enough to allow a clinical
response. It depends strongly on how onset is timed, so it should be read
against the onset rule used to build the label.

The ranking of models can differ between discrimination, calibration and
these clinical measures. Such differences are reported rather than resolved,
because the measures assess different aspects of performance.

## Summary

This stage produces the calibration, net-benefit, operating-point,
per-patient and lead-time tables listed below. They are written to
`output/processed_data/`, with the figures in `output/figures/`, for
`12_inference.Rmd` and for the thesis tables.

```{r summary-out}
cat("Artefacts written to", DATA_DIR, ":\n")
for (f in c("11_calibration_summary", "11_net_benefit", "11_operating_points",
            "11_patient_level_metrics", "11_hour_vs_patient",
            "11_lead_time_summary", "11_external_operating_points",
            "11_external_patient_level")) {
  p <- file.path(DATA_DIR, paste0(f, ".parquet"))
  if (file.exists(p)) cat("  -", basename(p), "\n")
}
cat("\nFigures written to", FIGURES_DIR, ":\n")
for (f in c("11_calibration_curves.png", "11_decision_curves.png",
            "11_false_alarms_vs_sensitivity.png",
            "11_lead_time_distribution.png")) {
  p <- file.path(FIGURES_DIR, f)
  if (file.exists(p)) cat("  -", f, "\n")
}
```
