---
title: "12 Inference - Bootstrap Uncertainty and Multiplicity Control"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

The preceding stages report point estimates. This notebook adds intervals to
them and controls the error rates that arise from testing many contrasts. It
has three components.

1. **Cluster-bootstrap confidence intervals** for the primary and subgroup
   estimates. Person-hours are nested within ICU stays, so the stay is the
   resampling unit. Resampling individual person-hours would ignore the
   dependence between hours of the same stay and would understate the
   uncertainty.

2. **Multiplicity control.** The six pre-registered hypotheses form the
   confirmatory family, and the Holm-Bonferroni procedure controls its
   family-wise error rate at $\alpha = 0.05$. The subgroup analyses and the
   alternative-label comparisons form three exploratory families: E1 (subgroup
   discrimination), E2 (subgroup label sensitivity) and E3 (alternative-label
   contrasts). Within each, the Benjamini-Hochberg (BH) procedure controls the
   expected false-discovery rate at $q = 0.05$. The two procedures differ
   because the families serve different purposes: a confirmatory claim must
   meet strict family-wise control, whereas an exploratory family is intended
   to generate hypotheses for later confirmation. BH is valid under
   independence and under positive dependence between the tests. The contrasts
   within a family share stays and, within a field, a reference level, so they
   are dependent. That dependence is expected to be positive, and a
   Bonferroni-adjusted p-value, which needs no assumption about dependence, is
   computed beside each BH value as a check.

3. **An analysis register** that tags every reported analysis as
   confirmatory, exploratory or descriptive and names the section of the
   pre-registration or the protocol amendment under which it was specified. It
   is saved as a table so that the thesis tables can state each analysis's
   status directly rather than through a general methodological statement.

**Status of this notebook.** The uncertainty and multiplicity plan is
Protocol Amendment 1 (`00_preregistration.md` §13), written after the analyses
in notebooks 02--11 had been run. The amendment is therefore itself post-hoc.
It does not alter the definitions of H1--H6, their thresholds or their
estimands, except where §13.6 records that H5's estimand cannot be formed
(see below). It adds intervals and a multiplicity correction that the original
protocol did not specify, and these can change a conclusion. Where the
correction changes a verdict, the corrected verdict is the one reported.

Family E3 belongs to Protocol Amendment 2 (`00_preregistration.md` §§16--17),
which adds two non-Sepsis-3 label definitions, and to Protocol Amendment 4,
which adds Variant F. Both amendments are post-hoc. They do not modify the
definitions of H1--H6: Variants D, E and F are kept out of H1 and out of the
variance decomposition, which are computed from Variants A--C alone.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(arrow)
  library(ggplot2)
  library(dplyr)
})

source("config.R")
source("utils.R")

step("=== Phase 9: Inference, uncertainty, and multiplicity ===")

# Smoke-test override: V2_BOOT_B=50 runs the whole notebook end to end in
# seconds. Reported results always use the values fixed in config.R.
.b_override <- suppressWarnings(as.integer(Sys.getenv("V2_BOOT_B", "")))
if (!is.na(.b_override) && .b_override > 0) {
  BOOT_B_MAIN <- BOOT_B_SUBGROUP <- BOOT_B_EXTERNAL <- .b_override
  v2_log(sprintf("V2_BOOT_B override active: B = %d (smoke test, not for reporting)",
                 .b_override), level = "WARN")
}

v2_log(sprintf("Bootstrap replicates: main=%d, subgroup=%d, external=%d (seed %d)",
               BOOT_B_MAIN, BOOT_B_SUBGROUP, BOOT_B_EXTERNAL, BOOT_SEED))

MODELS <- c(primary = "pred_sepsis", gbt = "pred_gbt", news2 = "pred_news2",
            cox = "pred_cox", qsofa = "pred_qsofa", sirs = "pred_sirs")

fmt_ci <- function(lo, hi, digits = 3)
  sprintf("[%.*f, %.*f]", digits, lo, digits, hi)
```

## Data Assembly

The temporal test predictions for each label variant are joined to the
comparator predictions and to stay-level demographics. Rows are ordered by
`stay_id`, so the hours of each stay form one block that can be resampled as a
cluster, and each stay is given an index into a **common resampling frame**
shared by Variants A, B and C. All three variants are built on the same 12,132
test stays, so the same bootstrap sample of stays is applied to all three
within each replicate, which makes the label-variant contrast in H1 a paired
comparison.

The pairing is by stay, not by person-hour. The output below shows that the
three variants contain different numbers of person-hours. The panel keeps each
stay's at-risk hours only, up to and including its onset, and the variants
place onsets at different hours or not at all, so the same stay contributes a
different number of hours under each variant. Variant C labels the most stays
and so has the fewest hours. Each onset contributes exactly one positive
person-hour, the hour in which it occurs, so the number of onsets equals the
number of positive person-hours (`n_events` in later tables).

```{r assemble}
load_test_table <- function(vid) {
  prim <- file.path(DATA_DIR, sprintf("05_predictions_primary_%s.parquet", vid))
  comp <- file.path(DATA_DIR, sprintf("06_predictions_comparators_%s.parquet", vid))
  if (!file.exists(prim)) return(NULL)

  dt <- as.data.table(read_parquet(prim))
  if (file.exists(comp)) {
    cp <- as.data.table(read_parquet(comp))
    keep <- intersect(c("pred_gbt", "pred_news2", "pred_cox", "pred_qsofa", "pred_sirs"),
                      names(cp))
    dt <- merge(dt, cp[, c("stay_id", "hour", keep), with = FALSE],
                by = c("stay_id", "hour"), all.x = TRUE)
  }

  demo_path <- file.path(DATA_DIR, sprintf("03_person_hours_%s.parquet", vid))
  if (file.exists(demo_path)) {
    demo <- unique(as.data.table(read_parquet(
      demo_path, col_select = c("stay_id", EQUITY_VARS[EQUITY_VARS != "ethnicity"]))))
    dt <- merge(dt, demo, by = "stay_id", all.x = TRUE)
  }

  dt[, y := as.integer(event == 1)]
  setorder(dt, stay_id, hour)
  dt[]
}

test_tab <- lapply(c(A = "A", B = "B", C = "C"), load_test_table)
test_tab <- Filter(Negate(is.null), test_tab)

# Common stay frame across variants (all variants share the same test stays)
stay_frame <- sort(Reduce(intersect, lapply(test_tab, function(d) unique(d$stay_id))))
N_STAYS    <- length(stay_frame)
for (v in names(test_tab)) test_tab[[v]][, sidx := match(stay_id, stay_frame)]

cat(sprintf("Common test-stay resampling frame: %d stays\n", N_STAYS))
for (v in names(test_tab))
  cat(sprintf("  Variant %s: %s person-hours, %d sepsis onsets\n",
              v, format(nrow(test_tab[[v]]), big.mark = ","), sum(test_tab[[v]]$y)))
```

## Internal Performance with Confidence Intervals

AUROC and AUPRC for every model on the Variant B temporal test set, with
percentile intervals from `r format(BOOT_B_MAIN, big.mark=",")` stay-level
bootstrap replicates. The same replicate samples are used for all six models,
so a difference between two models can be computed within each replicate and
its interval keeps the pairing.

Percentile intervals are used throughout. They respect the [0, 1] range of
AUROC and AUPRC, and the bootstrap p-values below are obtained by inverting the
same percentile distribution, so a test and its interval cannot disagree. A
replicate in which a metric is undefined, because the resampled stays contain
no onset or no non-onset hour, is dropped. An interval is reported only if at
least 10 replicates remain. In this table `n_events` is the number of onsets.

```{r main-ci}
tb <- test_tab[["B"]]
pre_B <- lapply(MODELS, function(m)
  if (m %in% names(tb)) auc_precompute(tb[[m]], tb$y, tb$sidx, N_STAYS) else NULL)
names(pre_B) <- names(MODELS)
pre_B <- Filter(Negate(is.null), pre_B)

stat_main <- function(w) {
  c(vapply(pre_B, boot_auroc_w, numeric(1), w_stay = w),
    vapply(pre_B, boot_auprc_w, numeric(1), w_stay = w)) |>
    setNames(c(paste0("auroc_", names(pre_B)), paste0("auprc_", names(pre_B))))
}

t_start   <- proc.time()
boot_main <- boot_run(N_STAYS, BOOT_B_MAIN, stat_main, seed = BOOT_SEED,
                      progress_every = 500L)
v2_log(sprintf("Main bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))

ones <- rep(1L, N_STAYS)
main_ci <- rbindlist(lapply(names(pre_B), function(m) {
  a <- perc_ci(boot_main[, paste0("auroc_", m)], CI_LEVEL)
  p <- perc_ci(boot_main[, paste0("auprc_", m)], CI_LEVEL)
  data.table(model = m,
             n_rows = pre_B[[m]]$n,
             n_events = sum(tb$y),
             auroc = boot_auroc_w(pre_B[[m]], ones), auroc_lo = a["lo"], auroc_hi = a["hi"],
             auprc = boot_auprc_w(pre_B[[m]], ones), auprc_lo = p["lo"], auprc_hi = p["hi"])
}))
main_ci[, `:=`(auroc_ci = fmt_ci(auroc_lo, auroc_hi),
               auprc_ci = fmt_ci(auprc_lo, auprc_hi, 4))]

save_parquet(main_ci, "12_internal_ci", DATA_DIR)
save_csv(main_ci, "12_internal_ci", DATA_DIR)

knitr::kable(main_ci[, .(model, n_events, auroc = round(auroc, 4), auroc_ci,
                          auprc = round(auprc, 5), auprc_ci)],
             caption = "Internal temporal test performance with 95\\% stay-level cluster bootstrap intervals (Variant B)")
```

## Confirmatory Family: the Six Pre-Registered Hypotheses

Each hypothesis is expressed as a single scalar contrast $\theta$ tested
against its pre-registered threshold $\theta_0$. The protocol states each
hypothesis directionally, so the tests are one-sided. The bootstrap replicates
$\theta^*_b$ are centred on the observed estimate $\hat\theta$, and
$\theta^*_b - \hat\theta$ is taken as the distribution of
$\hat\theta - \theta$ under the null. For the null $\theta \leq \theta_0$, the
p-value is

$$p = \frac{1 + \#\{b : \theta^*_b - \hat\theta \geq \hat\theta - \theta_0\}}{B + 1},$$

with the inequality reversed for a null of the form $\theta \geq \theta_0$.
The $(1 + \#)/(B + 1)$ form is the standard finite-simulation estimate of a
bootstrap p-value. It also sets the Monte Carlo resolution: with $B = 2{,}000$
the smallest attainable p-value is about 0.0005, so values near 0.001 should be
read as "at the resolution limit" rather than as precise.

### Contrast Estimation

H1 and H6 are computed from the same replicates. Both compare quantities
measured on one set of test stays, so a single resampling of stays serves both,
and the intervals keep the correlation between the quantities being
differenced rather than treating them as independent.

H4 and H5 cannot be paired in this way. H5 compares two different databases,
so its two bootstrap streams are drawn separately. H4 compares the temporal
test set with the random-split test set, and these overlap, because the random
split's test set is a 25 per cent sample from all admission years and so
includes stays from the temporal test years. H4's two streams are nevertheless
drawn separately. If the two AUROC estimates are positively correlated, which
the shared stays make likely, ignoring that correlation overstates the
variance of the difference, and H4's interval is then conservative. Linked
resampling of the shared stays would remove the approximation but is not
implemented.

```{r confirmatory-boot}
# --- H1 / H6: label spread, model spread, and the primary-vs-GBT gap --------
pre_label <- lapply(names(test_tab), function(v) {
  d <- test_tab[[v]]
  auc_precompute(d$pred_sepsis, d$y, d$sidx, N_STAYS)
})
names(pre_label) <- names(test_tab)

stat_h1h6 <- function(w) {
  lab <- vapply(pre_label, boot_auroc_w, numeric(1), w_stay = w)
  mod <- vapply(pre_B,     boot_auroc_w, numeric(1), w_stay = w)
  c(label_spread = diff(range(lab, na.rm = TRUE)),
    model_spread = diff(range(mod, na.rm = TRUE)),
    h1_delta     = diff(range(lab, na.rm = TRUE)) - diff(range(mod, na.rm = TRUE)),
    h6_delta     = unname(mod["gbt"] - mod["primary"]))
}

t_start  <- proc.time()
boot_h1  <- boot_run(N_STAYS, BOOT_B_MAIN, stat_h1h6, seed = BOOT_SEED + 1L,
                     progress_every = 500L)
v2_log(sprintf("H1/H6 bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
obs_h1h6 <- stat_h1h6(ones)
```

```{r confirmatory-h4}
# --- H4: random-split optimism (independent test set, so an independent draw)
rnd_prim <- file.path(DATA_DIR, "05_predictions_primary_random_B.parquet")
boot_h4 <- NULL; obs_h4 <- NA_real_
if (file.exists(rnd_prim)) {
  rd <- as.data.table(read_parquet(rnd_prim))
  rd[, y := as.integer(event == 1)]
  setorder(rd, stay_id, hour)
  rstays <- sort(unique(rd$stay_id)); rd[, sidx := match(stay_id, rstays)]
  pre_rnd <- auc_precompute(rd$pred_sepsis, rd$y, rd$sidx, length(rstays))

  b_rnd <- boot_run(length(rstays), BOOT_B_MAIN,
                    function(w) c(auroc = boot_auroc_w(pre_rnd, w)),
                    seed = BOOT_SEED + 2L, progress_every = 1000L)[, "auroc"]
  b_tmp <- boot_run(N_STAYS, BOOT_B_MAIN,
                    function(w) c(auroc = boot_auroc_w(pre_B$primary, w)),
                    seed = BOOT_SEED + 3L, progress_every = 1000L)[, "auroc"]
  # independent samples: pair the two bootstrap streams to propagate both variances
  boot_h4 <- b_rnd - b_tmp
  obs_h4  <- boot_auroc_w(pre_rnd, rep(1L, length(rstays))) -
             boot_auroc_w(pre_B$primary, ones)
  cat(sprintf("H4: random AUROC = %.4f, temporal AUROC = %.4f, delta = %.4f\n",
              boot_auroc_w(pre_rnd, rep(1L, length(rstays))),
              boot_auroc_w(pre_B$primary, ones), obs_h4))
}
```

```{r confirmatory-h5}
# --- H5: internal -> external degradation on eICU-CRD -----------------------
eicu_path <- file.path(DATA_DIR, "08_eicu_predictions_B.parquet")
boot_h5 <- NULL; obs_h5 <- NA_real_; ext_auroc <- NA_real_
ext_n_events <- NA_integer_; ext_interpretable <- NA
if (file.exists(eicu_path)) {
  ed <- as.data.table(read_parquet(eicu_path))
  ed[, y := as.integer(event == 1)]
  # The external event count governs whether the H5 estimate can be read at
  # all. Stage 08 restricts the external cohort to the stays in which the
  # Sepsis-3 anchor is recoverable; if that cohort still yields fewer than
  # MIN_EXTERNAL_EVENTS events the interval is wider than the effect H5
  # concerns, and the verdict is carried with that flag attached.
  ext_n_events      <- sum(ed$y)
  ext_interpretable <- ext_n_events >= MIN_EXTERNAL_EVENTS
  if (!ext_interpretable)
    v2_log(sprintf(paste0("H5: external cohort has %d events, below the %d-event ",
                          "floor; the AUROC estimate is not interpretable"),
                   ext_n_events, MIN_EXTERNAL_EVENTS), level = "WARN")
  setorder(ed, stay_id, hour)
  estays <- sort(unique(ed$stay_id)); ed[, sidx := match(stay_id, estays)]
  pre_ext <- auc_precompute(ed$pred_sepsis, ed$y, ed$sidx, length(estays))
  ext_auroc <- boot_auroc_w(pre_ext, rep(1L, length(estays)))

  t_start <- proc.time()
  b_ext <- boot_run(length(estays), BOOT_B_EXTERNAL,
                    function(w) c(auroc = boot_auroc_w(pre_ext, w)),
                    seed = BOOT_SEED + 4L, progress_every = 250L)[, "auroc"]
  v2_log(sprintf("External bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
  b_int <- boot_run(N_STAYS, BOOT_B_EXTERNAL,
                    function(w) c(auroc = boot_auroc_w(pre_B$primary, w)),
                    seed = BOOT_SEED + 5L, progress_every = 500L)[, "auroc"]

  boot_h5 <- b_int - b_ext
  obs_h5  <- boot_auroc_w(pre_B$primary, ones) - ext_auroc
  ext_ci  <- perc_ci(b_ext, CI_LEVEL)
  cat(sprintf("H5: internal AUROC = %.4f, external AUROC = %.4f %s, degradation = %.4f\n",
              boot_auroc_w(pre_B$primary, ones), ext_auroc,
              fmt_ci(ext_ci["lo"], ext_ci["hi"]), obs_h5))
}
```

```{r confirmatory-h2h3}
# --- H2: coefficient stability (descriptive; no test statistic) -------------
# H2's estimand is a threshold rule on point estimates -- sign reversal, or a
# doubling in magnitude, across the pre-registered label variants -- and it
# stays exactly that. It is not a contrast and carries no test, and it consumes
# a slot in the family size regardless, which makes the correction on the other
# hypotheses stricter.
#
# Protocol Amendment 7 adds a diagnostic beside it, not in place of it. Once
# stage 05 could compute a cluster-robust covariance the obvious question about
# the count became answerable: does a flagged coefficient move by more than it
# is estimated to within? A threshold rule cannot distinguish a coefficient
# that genuinely shifts from one that was never estimated precisely enough for
# a doubling to mean anything. The pre-registered count is reported first and
# unmodified; the diagnostic is reported second and separately.
coef_path <- file.path(DATA_DIR, "05_coef_all_variants.parquet")
h2_sign <- NA_integer_; h2_mag <- NA_integer_; h2_any <- NA_integer_; h2_n_terms <- NA_integer_
h2_n_robust <- NA_integer_; h2_n_noise <- NA_integer_
if (file.exists(coef_path)) {
  ct    <- as.data.table(read_parquet(coef_path))
  # Shared with the Amendment 5 ablation arm in 05_primary_model.Rmd, so the
  # two instability counts are produced by one rule and remain comparable.
  flags <- coef_stability_flags(ct, variants = c("A", "B", "C"))

  # Instability split by whether the covariate records physiology or clinician
  # behaviour. This is descriptive and sits outside the confirmatory family.
  kind_split <- stability_by_feature_kind(flags)
  save_parquet(kind_split, "12_h2_stability_by_kind", DATA_DIR)
  save_csv(kind_split,     "12_h2_stability_by_kind", DATA_DIR)
  cat("H2 instability by feature kind:\n"); print(kind_split)

  h2_sign    <- sum(flags$sign_flip)
  h2_mag     <- sum(flags$mag_flip)
  h2_any     <- sum(flags$sign_flip | flags$mag_flip)
  h2_n_terms <- nrow(flags)
  save_parquet(flags, "12_h2_coefficient_stability", DATA_DIR)
  save_csv(flags,     "12_h2_coefficient_stability", DATA_DIR)
  cat(sprintf("H2: of %d shared covariates, %d reverse sign, %d change magnitude >=2x, %d do either\n",
              h2_n_terms, h2_sign, h2_mag, h2_any))

  # Amendment 7: the uncertainty diagnostic on the flagged covariates.
  h2_unc <- h2_stability_uncertainty(ct, flags, variants = c("A", "B", "C"))
  if (nrow(h2_unc) > 0) {
    h2_unc[, is_action := vapply(as.character(term), function(tm)
      any(startsWith(tm, ACTION_DERIVED_FEATURES)), logical(1))]
    setorder(h2_unc, -z_ratio)
    save_parquet(h2_unc, "12_h2_stability_uncertainty", DATA_DIR)
    save_csv(h2_unc,     "12_h2_stability_uncertainty", DATA_DIR)
    h2_n_robust <- sum(h2_unc$exceeds_noise, na.rm = TRUE)
    h2_n_noise  <- sum(!h2_unc$exceeds_noise, na.rm = TRUE)
    cat(sprintf(paste0("H2 uncertainty diagnostic: of the %d flagged covariates, ",
                       "%d exceed their pooled cluster-robust SE and %d do not; ",
                       "%d of the %d that do are action-derived\n"),
                nrow(h2_unc), h2_n_robust, h2_n_noise,
                sum(h2_unc$exceeds_noise & h2_unc$is_action, na.rm = TRUE),
                h2_n_robust))
    print(h2_unc[, .(term, max_diff = round(max_diff, 4),
                     pooled_se = round(pooled_se, 4),
                     z = round(z_ratio, 2), exceeds_noise, is_action)])
  } else {
    v2_log(paste0("H2 uncertainty diagnostic empty: no se_cluster column in ",
                  "05_coef_all_variants, or no covariate flagged. Re-run ",
                  "stage 05 if the covariance was expected."), level = "WARN")
  }
}

# --- H3: treatment-anchored window (degenerate; exact bound instead) --------
mt <- as.data.table(read_parquet(file.path(DATA_DIR, "07_metric_results.parquet")))
anch <- mt[eval_window == "treatment_anchored" & variant == "B" & model == "primary"]
h3_n_hours  <- if (nrow(anch)) anch$n_rows[1] else NA_integer_
h3_n_events <- if (nrow(anch)) anch$n_sepsis_events[1] else NA_integer_
h3_upper    <- if (!is.na(h3_n_hours) && h3_n_events == 0) zero_event_upper(h3_n_hours, CI_LEVEL) else NA_real_
if (!is.na(h3_upper))
  cat(sprintf("H3: %d sepsis onsets in %s pre-treatment person-hours; exact 95%% upper bound on the per-hour onset rate = %.3g (1 in %s)\n",
              h3_n_events, format(h3_n_hours, big.mark = ","), h3_upper,
              format(round(1 / h3_upper), big.mark = ",")))

# The bound above is exact but it is a bound on a quantity the label definition
# fixes at zero (see the proposition in 09_variance_decomposition.Rmd), so it
# carries no information about MIMIC-IV. The informative counterpart is the
# same window under standard min(t_susp, t_SOFA) onset timing, which is Variant
# F: identical membership, timing branch restored. Reported here so the two
# numbers appear side by side.
alt_mt <- if (file.exists(file.path(DATA_DIR, "07_metric_results_altlabel.parquet")))
            as.data.table(read_parquet(file.path(DATA_DIR, "07_metric_results_altlabel.parquet"))) else NULL
h3_f <- if (!is.null(alt_mt))
          alt_mt[eval_window == "treatment_anchored" & variant == "F" & model == "primary"] else NULL
if (!is.null(h3_f) && nrow(h3_f) == 1)
  cat(sprintf("H3 (empirical counterpart, Variant F): %d sepsis onsets in %s pre-treatment person-hours, AUROC %s\n",
              h3_f$n_sepsis_events[1], format(h3_f$n_rows[1], big.mark = ","),
              if (is.na(h3_f$auroc[1])) "undefined" else sprintf("%.4f", h3_f$auroc[1])))
```

### Multiplicity-Adjusted Confirmatory Results

Each hypothesis is given its contrast, interval, null hypothesis, raw p-value
and Holm-adjusted p-value. The direction of subtraction is stated in each
contrast. H1's contrast is the label spread minus the model-class spread, so
the alternative H1 predicts is $\theta > 0$. H4's is random-split minus
temporal-split AUROC, and H6's is GBT AUROC minus primary AUROC. H1's two
spreads are also saved as separate fields, so the thesis table takes them from
the data rather than by transcription.

The verdict column reports the pre-registered test only. "NOT confirmed" means
that the pre-registered threshold was not met. It does not establish the
absence of an effect, and an estimate on the opposite side of its null is
shown by the interval rather than by the verdict. Four rows need more than
the verdict:

- **H2** has no p-value. It reports the pre-registered count, the number of
  covariates that change sign or differ at least twofold in magnitude across
  Variants A--C. A covariate meeting both criteria is counted once, so the
  count is the size of the union. Beside it is the Protocol Amendment 7
  diagnostic, which compares each flagged covariate's largest cross-label
  difference with the pooled cluster-robust standard error of the two
  estimates involved. The diagnostic is a heuristic benchmark, not a test. The
  three fits share stays, so their estimates are correlated, which the pooled
  standard error ignores, and the sandwich estimator is ridge-regularised. A
  difference below the benchmark is not shown to be noise; it is only not
  distinguishable from it by this measure.
- **H3** has no p-value. Under Variants A--C the pre-treatment window contains
  no onsets by construction (`09_variance_decomposition.Rmd`), so the AUROC
  difference is undefined. The row instead reports a one-sided 95 per cent
  upper bound on the hourly onset rate in that window, the exact
  Clopper-Pearson bound for zero events in $n$ hours, $1 - 0.05^{1/n}$. The
  bound treats hours as independent and the zero as a sampling outcome, whereas
  the zero is imposed by the onset rule, so it describes the window's size
  rather than the absence of leakage. Variant F's pre-treatment result is a
  post-hoc analysis of a different onset-timing rule, not a test of H3.
- **H5** is tested on a substitute estimand. The pre-registered H5 compares
  the proportional decline of the Utility Score with that of AUROC, and that
  contrast cannot be formed: the internal Utility Score under the primary label
  is 0.000 at the default threshold and only slightly above zero at its best
  threshold, so a proportional decline is undefined or has no usable sampling
  distribution. Section 13.6 of the pre-registration records the substitute,
  internal minus external AUROC for the primary model under Variant B against
  $\theta_0 = 0$, and places it in H5's slot in the Holm family. Its adjusted
  p-value refers to that AUROC contrast. It neither tests nor establishes the
  pre-registered metric-divergence claim. The external estimate also rests on
  29 onsets in the microbiology-covered eICU sub-cohort, below the 100-onset
  floor, so the contrast reflects differences in label ascertainment as well
  as transport.
- **H6** is discussed under item 3 of "Reading the Corrected Family" below.

```{r confirmatory-table}
mk <- function(id, label, contrast, theta, ci, null_txt, p, note = "") {
  data.table(hypothesis = id, claim = label, contrast = contrast,
             estimate = theta, ci_lo = ci[1], ci_hi = ci[2],
             null_hypothesis = null_txt, p_raw = p, note = note)
}

rows <- list()

# The two spreads and their intervals are the body of the decomposition table
# in the thesis. They used to exist only inside H1's free-text note, which meant
# the table transcribed them by hand; they are persisted as data instead.
spread_ci <- rbindlist(lapply(c("label", "model"), function(q) {
  col <- paste0(q, "_spread")
  ci  <- perc_ci(boot_h1[, col], CI_LEVEL)
  data.table(quantity = q,
             estimate = round(unname(obs_h1h6[col]), 4),
             ci_lo    = round(unname(ci["lo"]), 4),
             ci_hi    = round(unname(ci["hi"]), 4))
}))
save_parquet(spread_ci, "12_spread_ci", DATA_DIR)
save_csv(spread_ci,     "12_spread_ci", DATA_DIR)
print(spread_ci)

# H1: theta = label spread - model spread; H0: theta <= 0
ci_h1 <- perc_ci(boot_h1[, "h1_delta"], CI_LEVEL)
p_h1  <- boot_pvalue(boot_h1[, "h1_delta"], obs_h1h6["h1_delta"], 0, "greater")
rows[["H1"]] <- mk("H1", "Label dominance",
                   "AUROC spread across label variants - spread across model classes",
                   unname(obs_h1h6["h1_delta"]), ci_h1, "theta <= 0", p_h1,
                   sprintf("label spread %.4f %s; model spread %.4f %s",
                           obs_h1h6["label_spread"],
                           do.call(fmt_ci, as.list(perc_ci(boot_h1[, "label_spread"], CI_LEVEL))),
                           obs_h1h6["model_spread"],
                           do.call(fmt_ci, as.list(perc_ci(boot_h1[, "model_spread"], CI_LEVEL)))))

# H2: descriptive only
rows[["H2"]] <- mk("H2", "Coefficient instability",
                   "Covariates with sign reversal or >=2x magnitude change across variants",
                   as.numeric(h2_any), c(NA_real_, NA_real_),
                   "not testable (threshold rule on point estimates, not a contrast)", NA_real_,
                   sprintf(paste0("%d of %d covariates unstable (%d sign reversals, ",
                                  "%d magnitude >=2x). Amendment 7 diagnostic: %s of the %d ",
                                  "flagged exceed their pooled cluster-robust SE; the ",
                                  "remainder move by less than their own estimation noise. ",
                                  "The count above is the pre-registered estimand and is ",
                                  "unmodified by the diagnostic"),
                           h2_any, h2_n_terms, h2_sign, h2_mag,
                           ifelse(is.na(h2_n_robust), "an unrecorded number", h2_n_robust),
                           h2_any))

# H3: degenerate; exact bound rather than a test
rows[["H3"]] <- mk("H3", "Treatment leakage",
                   "Sepsis onsets in the pre-treatment (treatment-anchored) window",
                   as.numeric(h3_n_events), c(NA_real_, NA_real_),
                   "not testable (outcome absent from the window)", NA_real_,
                   sprintf("%d onsets in %s person-hours; exact 95%% upper bound on the per-hour onset rate = %.2e (1 in %s)",
                           h3_n_events, format(h3_n_hours, big.mark = ","), h3_upper,
                           format(round(1 / h3_upper), big.mark = ",")))

# H4: theta = random - temporal AUROC; H0: theta <= 0.02
if (!is.null(boot_h4)) {
  ci_h4 <- perc_ci(boot_h4, CI_LEVEL)
  p_h4  <- boot_pvalue(boot_h4, obs_h4, 0.02, "greater")
  rows[["H4"]] <- mk("H4", "Split optimism",
                     "Random-split AUROC - temporal-split AUROC (primary, Variant B)",
                     obs_h4, ci_h4, "theta <= 0.02", p_h4,
                     paste("the two test sets are resampled separately and both",
                           "bootstrap streams are propagated into the difference.",
                           "They are not disjoint: the random-split test set is",
                           "a 25% sample across all admission years and so shares",
                           "stays with the 2017-2019 temporal test set. Ignoring",
                           "that positive dependence over-states the variance of",
                           "the difference, so the interval is conservative"))
}

# H5: theta = internal - external AUROC; H0: theta <= 0
if (!is.null(boot_h5)) {
  ci_h5 <- perc_ci(boot_h5, CI_LEVEL)
  p_h5  <- boot_pvalue(boot_h5, obs_h5, 0, "greater")
  rows[["H5"]] <- mk("H5", "Metric divergence",
                     "Internal - external AUROC (primary, Variant B, eICU-CRD)",
                     obs_h5, ci_h5, "theta <= 0", p_h5,
                     sprintf("external AUROC %.4f on %d external events (%s the %d-event interpretability floor); Utility Score arm not testable: the internal score is %.3f at the default cut-off and %.3f at its best threshold, so a proportional decline has no baseline to be proportional to",
                             ext_auroc, ext_n_events,
                             if (isTRUE(ext_interpretable)) "at or above" else "BELOW",
                             MIN_EXTERNAL_EVENTS,
                             mt[variant == "B" & model == "primary" & eval_window == "unrestricted",
                                utility_normalised][1],
                             mt[variant == "B" & model == "primary" & eval_window == "unrestricted",
                                utility_max][1]))
}

# H6: equivalence bound 0.02; rejecting H6 requires evidence that |delta| > 0.02.
#
# Protocol Amendment 10. The null is an INTERVAL, -0.02 <= theta <= 0.02, so it
# has two boundaries and rejecting it means clearing one of them. Earlier runs
# tested the upper boundary only. That limb asks whether the boosted trees beat
# the hazard model by more than the margin, which is the pre-registered concern
# and the right thing to report -- but with theta-hat on the NEGATIVE side of
# zero it is also the limb the data cannot speak to, and it returns p ~ 1 for
# arithmetic reasons rather than evidential ones. Both limbs are now computed.
#
# The reported p_raw is min(p_upper, p_lower), which is a valid (conservative)
# p-value for an interval null: under any theta inside [-0.02, 0.02] at most one
# limb can be small, because a theta near one boundary is two margins away from
# the other. The verdict rule is unchanged, and so is the direction of the
# conclusion: rejection would mean non-equivalence, and neither limb rejects.
ci_h6 <- perc_ci(boot_h1[, "h6_delta"], CI_LEVEL)
p_h6_upper <- boot_pvalue(boot_h1[, "h6_delta"], obs_h1h6["h6_delta"],  0.02, "greater")
p_h6_lower <- boot_pvalue(boot_h1[, "h6_delta"], obs_h1h6["h6_delta"], -0.02, "less")
p_h6 <- min(p_h6_upper, p_h6_lower, na.rm = TRUE)
rows[["H6"]] <- mk("H6", "Model-class null",
                   "GBT AUROC - primary AUROC (Variant B, temporal)",
                   unname(obs_h1h6["h6_delta"]), ci_h6,
                   "-0.02 <= theta <= 0.02 (equivalence; interval null, two limbs)",
                   p_h6,
                   sprintf(paste0("paired: both models scored on the same ",
                                  "person-hours. Interval null tested at both ",
                                  "boundaries (Amendment 10): upper limb ",
                                  "(theta > +0.02, GBT better by more than the ",
                                  "margin) p = %.3f; lower limb (theta < -0.02, ",
                                  "primary better by more than the margin) ",
                                  "p = %.3f. The reported value is the smaller ",
                                  "of the two. With theta-hat = %.4f the lower ",
                                  "limb is the informative one; the upper limb ",
                                  "is large by construction and establishes ",
                                  "nothing on its own"),
                           p_h6_upper, p_h6_lower, unname(obs_h1h6["h6_delta"])))

conf <- rbindlist(rows)
setorder(conf, hypothesis)

# Holm-Bonferroni over the pre-registered family size, not over the number of
# tests that happened to be computable.
p_vec <- conf$p_raw
m_fam <- CONFIRMATORY_FAMILY_SIZE
ord   <- order(p_vec, na.last = NA)
p_adj <- rep(NA_real_, length(p_vec))
if (length(ord) > 0) {
  running <- 0
  for (k in seq_along(ord)) {
    running <- max(running, (m_fam - k + 1) * p_vec[ord[k]])
    p_adj[ord[k]] <- min(1, running)
  }
}
conf[, p_holm := p_adj]
conf[, significant_holm := !is.na(p_holm) & p_holm < FWER_ALPHA]
conf[, verdict := fifelse(is.na(p_raw), "Descriptive (not testable)",
                   fifelse(significant_holm,
                           fifelse(hypothesis == "H6", "H6 REJECTED (non-equivalence)", "CONFIRMED"),
                           fifelse(hypothesis == "H6", "H6 not rejected", "NOT confirmed")))]
conf[hypothesis == "H2", verdict := "Descriptive: instability present"]
conf[hypothesis == "H3", verdict := "Structural: outcome absent from window"]
# H5's slot is occupied by a substituted estimand (pre-registration 13.6): the
# proportional utility-versus-AUROC divergence cannot be formed against an
# internal Utility Score indistinguishable from zero, so the AUROC limb alone
# is tested. A bare "NOT confirmed" would read as a hypothesis that was tested
# and failed. The verdict names the substitution so that every table printing
# it carries the qualification with it.
conf[hypothesis == "H5",
     verdict := "Not evaluable as pre-registered; substituted AUROC contrast inconclusive"]
conf[, family := "confirmatory (pre-registered)"]
conf[, correction := sprintf("Holm-Bonferroni, m=%d, FWER=%.2f", m_fam, FWER_ALPHA)]

# H6's two limbs are persisted beside its reported p-value so the thesis can
# print both rather than describing one of them (Amendment 10). Every other
# hypothesis has a one-sided null with a single boundary, so the columns are NA.
conf[, `:=`(p_limb_upper = NA_real_, p_limb_lower = NA_real_)]
conf[hypothesis == "H6", `:=`(p_limb_upper = p_h6_upper, p_limb_lower = p_h6_lower)]

save_parquet(conf, "12_confirmatory_tests", DATA_DIR)
save_csv(conf, "12_confirmatory_tests", DATA_DIR)

knitr::kable(
  conf[, .(H = hypothesis, claim, estimate = round(estimate, 4),
           CI = fifelse(is.na(ci_lo), "--", fmt_ci(ci_lo, ci_hi)),
           H0 = null_hypothesis,
           p = fifelse(is.na(p_raw), "--", formatC(p_raw, format = "g", digits = 2)),
           p_holm = fifelse(is.na(p_holm), "--", formatC(p_holm, format = "g", digits = 2)),
           verdict)],
  caption = "Confirmatory family: six pre-registered hypotheses, one-sided cluster-bootstrap tests against their pre-registered thresholds, Holm-Bonferroni corrected at family size 6")
```

```{r confirmatory-notes}
knitr::kable(conf[nzchar(note), .(H = hypothesis, note)],
             caption = "Supporting detail for the confirmatory family")
```

## Exploratory Family E1: Subgroup Discrimination

Subgroup AUROC is estimated with a bootstrap stratified by subgroup level:
stays are resampled within each level, so the level sizes are held at their
observed values. The intervals are therefore conditional on the observed
subgroup composition and do not include variation in how many stays fall in
each level. That is the appropriate estimand when the subgroups themselves are
the object of interest.

Each level is contrasted with the largest level of its own field, the
reference, which is chosen by sample size rather than by outcome and is
recorded in the output. The levels of a field are disjoint, so each contrast
compares two independently resampled groups and is not paired. Protocol
Amendment 6, which is post-hoc, restricts the corrected family to contrasts
in which both levels have at least 100 onsets. Levels with at least 5 onsets
are still estimated and shown, but they enter no contrast. The family is
therefore smaller than the set of all level-by-model comparisons, and its size
is printed in the output. BH is applied across the whole family. The two
tables below show the race field only; the full family, including the language
and insurance contrasts, is saved in `12_subgroup_auroc_contrasts.csv`.
Whether a small level's interval includes 0.5 is shown row by row in the
first table.

```{r subgroup-auroc}
subgroup_models <- c(primary = "pred_sepsis", gbt = "pred_gbt")

sub_rows <- list(); sub_diff <- list()

for (svar in EQUITY_VARS) {
  if (!svar %in% names(tb)) next
  # blank strings are missing data, not a demographic category
  dsub <- tb[!is.na(get(svar)) & nzchar(trimws(as.character(get(svar))))]
  if (nrow(dsub) == 0) next

  lvl_n  <- dsub[, .N, by = svar][order(-N)]
  levels_keep <- head(as.character(lvl_n[[svar]]), MAX_SUBGROUP_LEVELS)
  dsub <- dsub[as.character(get(svar)) %chin% levels_keep]
  if (nrow(dsub) == 0) next

  # stay-level stratification frame for this variable
  stay_lvl <- unique(dsub[, .(stay_id, lvl = as.character(get(svar)))])
  stay_lvl <- stay_lvl[!duplicated(stay_id)]
  setorder(stay_lvl, lvl, stay_id)
  stay_lvl[, k := .I]
  dsub <- merge(dsub, stay_lvl[, .(stay_id, k)], by = "stay_id")
  n_k  <- nrow(stay_lvl)

  # eligible levels: enough rows and enough events to support an AUROC
  elig <- dsub[, .(n_rows = .N, n_events = sum(y), n_stays = uniqueN(stay_id)),
               by = .(lvl = as.character(get(svar)))]
  elig <- elig[n_rows >= MIN_SUBGROUP_ROWS & n_events >= MIN_SUBGROUP_EVENTS]
  if (nrow(elig) == 0) next
  ref_lvl <- elig[order(-n_rows)][1, lvl]

  # Two distinct thresholds, and the distinction is the point. The filter above
  # decides whether an AUROC can be COMPUTED at all; the flag below decides
  # whether it is INTERPRETED. Levels under the interpretability floor are
  # still estimated and reported -- suppressing them would hide which groups
  # the cohort contains and how few events they carry -- but they are kept out
  # of the contrast family, so the BH correction is applied over the analyses
  # the study is willing to read rather than over every level that exists.
  elig[, interpretable := n_events >= MIN_SUBGROUP_EVENTS_INTERPRET]
  v2_log(sprintf(
    "  %s: %d levels estimated, %d above the %d-event interpretability floor",
    svar, nrow(elig), sum(elig$interpretable), MIN_SUBGROUP_EVENTS_INTERPRET))

  # Estimand keys are opaque ids paired with a lookup table: subgroup labels
  # contain slashes, spaces and hyphens, so they are never parsed out of a key.
  pre_sub <- list(); key_tab <- list()
  for (mn in names(subgroup_models)) {
    mc <- subgroup_models[[mn]]
    if (!mc %in% names(dsub)) next
    for (lv in elig$lvl) {
      kk  <- sprintf("e%03d", length(pre_sub) + 1L)
      dd  <- dsub[as.character(get(svar)) == lv]
      pre_sub[[kk]] <- auc_precompute(dd[[mc]], dd$y, dd$k, n_k)
      key_tab[[kk]] <- data.table(estimand = kk, model = mn, lvl = lv)
    }
  }
  if (length(pre_sub) == 0) next
  key_tab <- rbindlist(key_tab)

  stat_sub <- function(w) vapply(pre_sub, boot_auroc_w, numeric(1), w_stay = w)
  v2_log(sprintf("Subgroup bootstrap: %s (%d levels, %d estimands)",
                 svar, nrow(elig), length(pre_sub)))
  b_sub <- boot_run(n_k, BOOT_B_SUBGROUP, stat_sub,
                    strata = stay_lvl$lvl, seed = BOOT_SEED + 10L,
                    progress_every = 1000L)
  obs_sub <- stat_sub(rep(1L, n_k))

  for (i in seq_len(nrow(key_tab))) {
    nm <- key_tab$estimand[i]; mn <- key_tab$model[i]; lv <- key_tab$lvl[i]
    ci <- perc_ci(b_sub[, nm], CI_LEVEL)
    e  <- elig[lvl == lv]
    sub_rows[[paste(svar, nm)]] <- data.table(
      subgroup_col = svar, subgroup_val = lv, model = mn,
      is_reference = lv == ref_lvl,
      n_rows = e$n_rows, n_stays = e$n_stays, n_events = e$n_events,
      auroc = unname(obs_sub[nm]), ci_lo = ci["lo"], ci_hi = ci["hi"],
      ci_width = unname(ci["hi"] - ci["lo"]),
      interpretable = e$interpretable,
      min_events_interpret = MIN_SUBGROUP_EVENTS_INTERPRET)

    ref_interpretable <- elig[lvl == ref_lvl, interpretable]
    if (lv != ref_lvl && isTRUE(e$interpretable) && isTRUE(ref_interpretable)) {
      ref_nm <- key_tab[model == mn & lvl == ref_lvl, estimand]
      if (length(ref_nm) == 1L) {
        dvec <- b_sub[, nm] - b_sub[, ref_nm]
        dobs <- unname(obs_sub[nm] - obs_sub[ref_nm])
        dci  <- perc_ci(dvec, CI_LEVEL)
        sub_diff[[paste(svar, nm)]] <- data.table(
          subgroup_col = svar, subgroup_val = lv, model = mn,
          reference = ref_lvl, n_events = e$n_events,
          delta_auroc = dobs, ci_lo = dci["lo"], ci_hi = dci["hi"],
          p_raw = boot_pvalue(dvec, dobs, 0, "two.sided"))
      }
    }
  }
}

sub_auroc <- rbindlist(sub_rows)
sub_delta <- rbindlist(sub_diff)
```

```{r subgroup-fdr}
if (nrow(sub_delta) > 0) {
  sub_delta[, p_bh := p.adjust(p_raw, method = "BH")]
  sub_delta[, p_bonf := p.adjust(p_raw, method = "bonferroni")]
  sub_delta[, significant_bh := p_bh < FDR_Q]
  sub_delta[, family := sprintf("exploratory E1 (m=%d subgroup AUROC contrasts)", nrow(sub_delta))]
  setorder(sub_delta, p_raw)
}
sub_auroc[, family := "exploratory E1"]

save_parquet(sub_auroc, "12_subgroup_auroc_ci", DATA_DIR)
save_csv(sub_auroc,     "12_subgroup_auroc_ci", DATA_DIR)
if (nrow(sub_delta) > 0) {
  save_parquet(sub_delta, "12_subgroup_auroc_contrasts", DATA_DIR)
  save_csv(sub_delta,     "12_subgroup_auroc_contrasts", DATA_DIR)
}

knitr::kable(
  sub_auroc[subgroup_col == "race"][order(model, -n_rows),
    .(model, subgroup = subgroup_val, n_rows, n_events,
      auroc = round(auroc, 3), CI = fmt_ci(ci_lo, ci_hi), width = round(ci_width, 3))],
  caption = "Subgroup AUROC by race with 95\\% stratified cluster-bootstrap intervals (Variant B, temporal test)")
```

```{r subgroup-fdr-contrasts}
if (nrow(sub_delta) > 0)
  knitr::kable(
    sub_delta[subgroup_col == "race"][order(model, p_raw),
      .(model, subgroup = subgroup_val, reference, n_events,
        delta = round(delta_auroc, 3), CI = fmt_ci(ci_lo, ci_hi),
        p = formatC(p_raw, format = "f", digits = 3),
        q_BH = formatC(p_bh, format = "f", digits = 3), significant_bh)],
    caption = "Subgroup AUROC differences vs.\\ the reference group, Benjamini-Hochberg corrected across the whole exploratory family E1")
```

## Exploratory Family E2: Subgroup Label Sensitivity

The proportion of stays labelled septic under each variant is a stay-level
proportion, so the bootstrap resamples whole stays within subgroup. The same
65,078 stays are labelled under all three variants, so the comparison of
Variants A, B and C is paired within each replicate. Within a replicate, a
level's range is the largest of its three proportions minus the smallest, and
its interval keeps the pairing. A range is not a smooth function of the
proportions, because the variant giving the maximum or minimum can change
between replicates, but percentile intervals remain usable for it.

Each level's range is contrasted with that of the reference level of its field
(the largest), giving a difference in percentage points, and BH is applied
across the whole family. As in E1, the tables below show the race field only;
the full family is saved in `12_subgroup_labelsens_contrasts.csv`. Levels such
as UNKNOWN and UNABLE TO OBTAIN record missing information, so a difference
for them concerns documentation, not a demographic group.

```{r label-sens-ci}
coh <- lapply(c(A = "A", B = "B", C = "C"), function(v) {
  p <- file.path(DATA_DIR, sprintf("02_cohort_%s.parquet", v))
  if (!file.exists(p)) return(NULL)
  d <- as.data.table(read_parquet(p))
  d[, .(stay_id, septic = as.integer(!is.na(onset_hour)),
        race, language, insurance)]
})
coh <- Filter(Negate(is.null), coh)

ls_rows <- list(); ls_diff <- list()

if (length(coh) >= 2) {
  base <- coh[[1]]
  common_stays <- sort(Reduce(intersect, lapply(coh, function(d) d$stay_id)))
  for (v in names(coh)) {
    coh[[v]] <- coh[[v]][stay_id %in% common_stays]
    setorder(coh[[v]], stay_id)
  }
  n_cs <- length(common_stays)

  for (svar in EQUITY_VARS) {
    if (!svar %in% names(coh[["B"]])) next
    lv_dt <- coh[["B"]][!is.na(get(svar)) & nzchar(trimws(as.character(get(svar)))),
                        .N, by = svar][order(-N)]
    levels_keep <- head(as.character(lv_dt[[svar]]), MAX_SUBGROUP_LEVELS)
    if (length(levels_keep) < 2) next

    lvl_vec <- as.character(coh[["B"]][[svar]])
    keep    <- !is.na(lvl_vec) & lvl_vec %chin% levels_keep
    idx     <- which(keep)
    lvl_k   <- lvl_vec[idx]
    ord     <- order(lvl_k, common_stays[idx])
    idx     <- idx[ord]; lvl_k <- lvl_k[ord]
    n_k     <- length(idx)

    sep_mat <- vapply(names(coh), function(v) coh[[v]]$septic[idx], numeric(n_k))
    lvl_f   <- factor(lvl_k, levels = levels_keep[levels_keep %in% lvl_k])
    ref_lvl <- names(sort(table(lvl_k), decreasing = TRUE))[1]

    stat_ls <- function(w) {
      num <- rowsum(w * sep_mat, lvl_f, reorder = FALSE)
      den <- as.vector(rowsum(w, lvl_f, reorder = FALSE))
      pct <- 100 * num / den
      rng <- apply(pct, 1, function(r) diff(range(r)))
      out <- c(as.vector(pct), rng)
      names(out) <- c(outer(rownames(pct), colnames(sep_mat),
                            function(a, b) paste0("pct_", b, "||", a)),
                      paste0("range||", rownames(pct)))
      out
    }

    v2_log(sprintf("Label-sensitivity bootstrap: %s (%d levels)", svar, nlevels(lvl_f)))
    b_ls  <- boot_run(n_k, BOOT_B_SUBGROUP, stat_ls, strata = lvl_k,
                      seed = BOOT_SEED + 20L, progress_every = 1000L)
    obs_ls <- stat_ls(rep(1L, n_k))

    for (lv in levels(lvl_f)) {
      r <- list(subgroup_col = svar, subgroup_val = lv,
                n_stays = sum(lvl_k == lv), is_reference = lv == ref_lvl)
      for (v in names(coh)) {
        nm <- paste0("pct_", v, "||", lv)
        ci <- perc_ci(b_ls[, nm], CI_LEVEL)
        r[[paste0("pct_", v)]]    <- unname(obs_ls[nm])
        r[[paste0("pct_", v, "_lo")]] <- unname(ci["lo"])
        r[[paste0("pct_", v, "_hi")]] <- unname(ci["hi"])
      }
      nm <- paste0("range||", lv)
      ci <- perc_ci(b_ls[, nm], CI_LEVEL)
      r$range_pp <- unname(obs_ls[nm]); r$range_lo <- unname(ci["lo"]); r$range_hi <- unname(ci["hi"])
      ls_rows[[paste(svar, lv)]] <- as.data.table(r)

      if (lv != ref_lvl) {
        dvec <- b_ls[, nm] - b_ls[, paste0("range||", ref_lvl)]
        dobs <- unname(obs_ls[nm] - obs_ls[paste0("range||", ref_lvl)])
        dci  <- perc_ci(dvec, CI_LEVEL)
        ls_diff[[paste(svar, lv)]] <- data.table(
          subgroup_col = svar, subgroup_val = lv, reference = ref_lvl,
          n_stays = sum(lvl_k == lv), delta_range_pp = dobs,
          ci_lo = dci["lo"], ci_hi = dci["hi"],
          p_raw = boot_pvalue(dvec, dobs, 0, "two.sided"))
      }
    }
  }
}

ls_ci   <- rbindlist(ls_rows, fill = TRUE)
ls_delta <- rbindlist(ls_diff, fill = TRUE)
if (nrow(ls_delta) > 0) {
  ls_delta[, p_bh := p.adjust(p_raw, method = "BH")]
  ls_delta[, p_bonf := p.adjust(p_raw, method = "bonferroni")]
  ls_delta[, significant_bh := p_bh < FDR_Q]
  ls_delta[, family := sprintf("exploratory E2 (m=%d label-sensitivity contrasts)", nrow(ls_delta))]
  setorder(ls_delta, p_raw)
}
if (nrow(ls_ci) > 0) {
  ls_ci[, family := "exploratory E2"]
  save_parquet(ls_ci, "12_subgroup_labelsens_ci", DATA_DIR)
  save_csv(ls_ci,     "12_subgroup_labelsens_ci", DATA_DIR)
}
if (nrow(ls_delta) > 0) {
  save_parquet(ls_delta, "12_subgroup_labelsens_contrasts", DATA_DIR)
  save_csv(ls_delta,     "12_subgroup_labelsens_contrasts", DATA_DIR)
}

if (nrow(ls_ci) > 0)
  knitr::kable(
    ls_ci[subgroup_col == "race"][order(-n_stays),
      .(subgroup = subgroup_val, n_stays,
        A = sprintf("%.1f %s", pct_A, fmt_ci(pct_A_lo, pct_A_hi, 1)),
        B = sprintf("%.1f %s", pct_B, fmt_ci(pct_B_lo, pct_B_hi, 1)),
        C = sprintf("%.1f %s", pct_C, fmt_ci(pct_C_lo, pct_C_hi, 1)),
        range = sprintf("%.1f %s", range_pp, fmt_ci(range_lo, range_hi, 1)))],
    caption = "Percentage of stays labelled septic by race and label variant, with 95\\% stratified cluster-bootstrap intervals")
```

```{r label-sens-contrasts}
if (nrow(ls_delta) > 0)
  knitr::kable(
    ls_delta[subgroup_col == "race"][order(p_raw),
      .(subgroup = subgroup_val, reference, n_stays,
        delta_pp = round(delta_range_pp, 2), CI = fmt_ci(ci_lo, ci_hi, 2),
        p = formatC(p_raw, format = "f", digits = 3),
        q_BH = formatC(p_bh, format = "f", digits = 3), significant_bh)],
    caption = "Difference in label sensitivity (A--C range) vs.\\ the reference group, Benjamini-Hochberg corrected across exploratory family E2")
```

## Exploratory Family E3: Alternative-Label Contrasts

Families E1 and E2 vary the label only within Sepsis-3, so they cannot
separate the effect of how one definition is operationalised from the effect
of using a label that depends on a treatment timestamp. This family attaches
intervals to the comparisons of Protocol Amendment 2. Variant E is Sepsis-3's
suspected-infection anchor with the organ-dysfunction requirement removed, and
Variant D is the organ-dysfunction component with no treatment timestamp.
Protocol Amendment 4 adds Variant F, which labels the same stays as Variant B
cohort-wide and moves the onset to $\min(t_{\text{susp}}, t_{\text{SOFA}})$.
Its contrast with Variant B is designed to isolate the effect of that timing
rule. It is not a pure timing effect, because moving an onset earlier also
shortens the stay's at-risk hours and brings some stays inside the 72-hour
horizon. Adding Variant F enlarged family E3, and every BH-adjusted p-value in
the family is computed over the enlarged set.

All variants are labelled on the same test stays, so the same bootstrap sample
of stays applies to all of them and each contrast is paired within replicate.
Reported beside the differences is the **anchor-attributable fraction**
$\phi = (\mathrm{AUROC}_E - 0.5)/(\mathrm{AUROC}_B - 0.5)$, the ratio of
Variant E's above-chance AUROC to Variant B's. It compares discrimination on
two different outcomes and is descriptive, not a causal share. A value above 1
means only that the model discriminates Variant E's label slightly better.
The natural reference value is 1, at which the two above-chance AUROCs are
equal. $\phi$ is reported as an interval and read against 1, and it is not
part of the corrected family.

```{r alt-label-ci}
alt_ids <- setNames(names(ALT_LABEL_VARIANTS), names(ALT_LABEL_VARIANTS))
alt_tab <- lapply(alt_ids, load_test_table)
alt_tab <- Filter(Negate(is.null), alt_tab)

alt_ci <- data.table(); alt_delta <- data.table()

if (length(alt_tab) == 0) {
  cat("No alternative-label predictions found; run 05-07 with the alternative-label variants.\n")
} else {
  ALT_MODELS <- c(primary = "pred_sepsis", gbt = "pred_gbt")

  # Reuse the A/B/C resampling frame so the contrasts against B are paired.
  pre_alt <- list()
  for (v in names(alt_tab)) {
    d <- alt_tab[[v]]
    d[, sidx := match(stay_id, stay_frame)]
    n_off <- d[is.na(sidx), uniqueN(stay_id)]
    if (n_off > 0)
      v2_log(sprintf("Variant %s: %d test stays outside the common frame, dropped",
                     v, n_off), level = "WARN")
    d <- d[!is.na(sidx)]
    alt_tab[[v]] <- d
    for (m in names(ALT_MODELS)) {
      col <- ALT_MODELS[[m]]
      if (col %in% names(d))
        pre_alt[[paste(v, m, sep = "_")]] <- auc_precompute(d[[col]], d$y, d$sidx, N_STAYS)
    }
  }
  for (m in names(ALT_MODELS))
    if (!is.null(pre_B[[m]])) pre_alt[[paste("B", m, sep = "_")]] <- pre_B[[m]]
  pre_alt <- Filter(Negate(is.null), pre_alt)

  stat_alt <- function(w) vapply(pre_alt, boot_auroc_w, numeric(1), w_stay = w)

  t_start  <- proc.time()
  boot_alt <- boot_run(N_STAYS, BOOT_B_MAIN, stat_alt, seed = BOOT_SEED + 30L,
                       progress_every = 500L)
  v2_log(sprintf("Alternative-label bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
  obs_alt <- stat_alt(ones)

  n_events_of <- function(v, m) {
    d <- if (v == "B") tb else alt_tab[[v]]
    if (is.null(d)) NA_integer_ else sum(d$y)
  }

  alt_ci <- rbindlist(lapply(names(pre_alt), function(k) {
    parts <- strsplit(k, "_", fixed = TRUE)[[1]]
    v <- parts[1]; m <- parts[2]
    ci <- perc_ci(boot_alt[, k], CI_LEVEL)
    data.table(variant = v, label = variant_name(v), label_family = variant_type(v),
               model = m, n_rows = pre_alt[[k]]$n, n_events = n_events_of(v, m),
               auroc = unname(obs_alt[k]), ci_lo = ci["lo"], ci_hi = ci["hi"])
  }))
  alt_ci[, `:=`(auroc_ci = fmt_ci(ci_lo, ci_hi), family = "exploratory E3")]
  setorder(alt_ci, model, variant)

  # Contrasts against Sepsis-3 (Variant B), paired within replicate
  rows <- list()
  for (m in names(ALT_MODELS)) {
    kb <- paste("B", m, sep = "_")
    if (!kb %in% colnames(boot_alt)) next
    for (v in names(alt_tab)) {
      kv <- paste(v, m, sep = "_")
      if (!kv %in% colnames(boot_alt)) next
      dvec <- boot_alt[, kv] - boot_alt[, kb]
      dobs <- unname(obs_alt[kv] - obs_alt[kb])
      dci  <- perc_ci(dvec, CI_LEVEL)
      rows[[kv]] <- data.table(
        variant = v, label = variant_name(v), model = m,
        delta_auroc = dobs, ci_lo = dci["lo"], ci_hi = dci["hi"],
        p_raw = boot_pvalue(dvec, dobs, 0, "two.sided"))
    }
  }
  alt_delta <- rbindlist(rows)
  if (nrow(alt_delta) > 0) {
    alt_delta[, p_bh := p.adjust(p_raw, method = "BH")]
    alt_delta[, p_bonf := p.adjust(p_raw, method = "bonferroni")]
    alt_delta[, significant_bh := p_bh < FDR_Q]
    alt_delta[, family := sprintf("exploratory E3 (m=%d alternative-label contrasts)",
                                  nrow(alt_delta))]
    setorder(alt_delta, p_raw)
  }

  # Anchor-attributable fraction, with the ratio bootstrapped end to end
  phi_rows <- list()
  for (m in names(ALT_MODELS)) {
    kb <- paste("B", m, sep = "_"); ke <- paste("E", m, sep = "_")
    if (!all(c(kb, ke) %in% colnames(boot_alt))) next
    denom <- boot_alt[, kb] - 0.5
    pvec  <- ifelse(denom > 0, (boot_alt[, ke] - 0.5) / denom, NA_real_)
    pci   <- perc_ci(pvec, CI_LEVEL)
    phi_rows[[m]] <- data.table(
      model = m,
      auroc_B = unname(obs_alt[kb]), auroc_E = unname(obs_alt[ke]),
      phi_anchor = unname((obs_alt[ke] - 0.5) / (obs_alt[kb] - 0.5)),
      ci_lo = pci["lo"], ci_hi = pci["hi"])
  }
  phi_dt <- rbindlist(phi_rows)

  save_parquet(alt_ci, "12_altlabel_auroc_ci", DATA_DIR)
  save_csv(alt_ci,     "12_altlabel_auroc_ci", DATA_DIR)
  if (nrow(alt_delta) > 0) {
    save_parquet(alt_delta, "12_altlabel_contrasts", DATA_DIR)
    save_csv(alt_delta,     "12_altlabel_contrasts", DATA_DIR)
  }
  if (nrow(phi_dt) > 0) {
    save_parquet(phi_dt, "12_anchor_attribution_ci", DATA_DIR)
    save_csv(phi_dt,     "12_anchor_attribution_ci", DATA_DIR)
  }

  knitr::kable(
    alt_ci[, .(variant, label, model, n_events,
               auroc = round(auroc, 4), CI = auroc_ci)],
    caption = "AUROC under each limb of the Sepsis-3 conjunction, with 95\\% stay-level cluster-bootstrap intervals (temporal test)")

  if (nrow(phi_dt) > 0)
    knitr::kable(
      phi_dt[, .(model, auroc_B = round(auroc_B, 4), auroc_E = round(auroc_E, 4),
                 phi_anchor = round(phi_anchor, 3), CI = fmt_ci(ci_lo, ci_hi))],
      caption = "Anchor-attributable fraction: share of Variant B's above-chance AUROC reproduced by the treatment anchor alone")
}
```

```{r alt-label-contrast-table}
if (nrow(alt_delta) > 0)
  knitr::kable(
    alt_delta[, .(variant, label, model, delta = round(delta_auroc, 4),
                  CI = fmt_ci(ci_lo, ci_hi),
                  p = formatC(p_raw, format = "f", digits = 4),
                  q_BH = formatC(p_bh, format = "f", digits = 4), significant_bh)],
    caption = "AUROC difference vs.\\ Sepsis-3 (Variant B), Benjamini-Hochberg corrected across exploratory family E3")
```

## Intervals Added Under Protocol Amendment 10

Three quantities belong to post-hoc analyses, and Amendment 1 attaches
intervals only to contrasts inside a corrected family. They are the effect on
discrimination of dropping the action-derived features, the effect of the
label definition on calibration, and the loss of discrimination when the
models are transported to the full external cohort under Variant D. Each
supports a statement in the thesis, and a point estimate alone cannot show how
precise it is. Protocol Amendment 10 therefore attaches an interval to each.
None of them enters a corrected family. Amendments 5 and 9 exclude the
ablation and the Variant D analysis from every family, and the calibration
contrasts are estimates, not tests. For all three, an interval that excludes
zero shows that the difference is detectable. Whether it is large enough to
matter is a separate question, answered by its magnitude.

### Action-Derived Feature Ablation (Amendment 5 arm)

The full and ablated specifications are fitted on the same training stays and
scored on the same test stays. The difference (ablated minus full) is
therefore paired within replicate, and its interval describes the difference
itself. Two separate intervals for the full and ablated AUROCs would not
measure the uncertainty of that difference. Each of the six contrasts (two
models under three variants) is reported with its own size and direction.

```{r ablation-ci}
abl_ci <- data.table()
abl_pre <- list(); abl_obs <- c()

for (vid in names(test_tab)) {
  full <- test_tab[[vid]]
  ap <- file.path(DATA_DIR, sprintf("05_predictions_primary_ablated_%s.parquet", vid))
  ag <- file.path(DATA_DIR, sprintf("06_predictions_gbt_ablated_%s.parquet", vid))
  for (m in c("primary", "gbt")) {
    path <- if (m == "primary") ap else ag
    col  <- if (m == "primary") "pred_sepsis" else "pred_gbt"
    if (!file.exists(path) || !col %in% names(full)) next
    ad <- as.data.table(read_parquet(path))
    if (!col %in% names(ad)) next
    ad[, y := as.integer(event == 1)]
    ad[, sidx := match(stay_id, stay_frame)]
    ad <- ad[!is.na(sidx)]
    abl_pre[[paste(vid, m, "full", sep = "_")]] <-
      auc_precompute(full[[col]], full$y, full$sidx, N_STAYS)
    abl_pre[[paste(vid, m, "abl", sep = "_")]] <-
      auc_precompute(ad[[col]], ad$y, ad$sidx, N_STAYS)
    rm(ad); gc()
  }
}
abl_pre <- Filter(Negate(is.null), abl_pre)

if (length(abl_pre) > 0) {
  stat_abl <- function(w) vapply(abl_pre, boot_auroc_w, numeric(1), w_stay = w)
  t_start  <- proc.time()
  boot_abl <- boot_run(N_STAYS, BOOT_B_MAIN, stat_abl, seed = BOOT_SEED + 40L,
                       progress_every = 500L)
  v2_log(sprintf("Ablation bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
  abl_obs <- stat_abl(ones)

  keys <- grep("_full$", names(abl_pre), value = TRUE)
  abl_ci <- rbindlist(lapply(keys, function(kf) {
    ka <- sub("_full$", "_abl", kf)
    if (!ka %in% colnames(boot_abl)) return(NULL)
    parts <- strsplit(kf, "_", fixed = TRUE)[[1]]
    dvec <- boot_abl[, ka] - boot_abl[, kf]
    dobs <- unname(abl_obs[ka] - abl_obs[kf])
    dci  <- perc_ci(dvec, CI_LEVEL)
    data.table(variant = parts[1], model = parts[2],
               auroc_full = unname(abl_obs[kf]), auroc_ablated = unname(abl_obs[ka]),
               delta_auroc = dobs, ci_lo = dci["lo"], ci_hi = dci["hi"],
               excludes_zero = dci["lo"] > 0 | dci["hi"] < 0)
  }))
  if (nrow(abl_ci) > 0) {
    abl_ci[, delta_ci := fmt_ci(ci_lo, ci_hi, 4)]
    abl_ci[, family := "none (Amendment 5 arm; enters no corrected family)"]
    setorder(abl_ci, model, variant)
    save_parquet(abl_ci, "12_ablation_ci", DATA_DIR)
    save_csv(abl_ci,     "12_ablation_ci", DATA_DIR)
    knitr::kable(abl_ci[, .(variant, model,
                            full = round(auroc_full, 4),
                            ablated = round(auroc_ablated, 4),
                            delta = round(delta_auroc, 4), delta_ci, excludes_zero)],
                 caption = "Paired ablation contrast (ablated minus full) with 95\\% stay-level cluster-bootstrap intervals")
  }
}
```

### Calibration Across Label Definitions

The calibration intercept and slope are refitted inside each replicate on the
same stays as every other quantity here, so the cross-label differences are
paired. The estimand is the change in the calibration intercept or slope
between each alternative label and the primary label, Variant B. Both are on
the logit scale. An intercept difference $d$ multiplies the predicted odds by
$e^{d}$, so a difference below 0.1 corresponds to a change of less than about
10 per cent in the predicted odds. The column `excludes_zero` is a descriptive
indicator of whether the interval excludes zero, not a test.

```{r calibration-ci}
calib_ci <- data.table(); calib_delta <- data.table()

cal_pre <- list()
for (vid in names(test_tab)) {
  d <- test_tab[[vid]]
  cal_pre[[vid]] <- calib_precompute(d$pred_sepsis, d$y, d$sidx, N_STAYS)
}
cal_pre <- Filter(Negate(is.null), cal_pre)

if (length(cal_pre) > 0) {
  # Warm-start values: the full-data fit. Every replicate starts from here, so
  # the IRLS converges in two or three steps instead of eight.
  cal_start <- lapply(cal_pre, function(pr)
    list(slope = boot_calib_slope_w(pr, ones),
         int   = boot_calib_int_w(pr, ones)))

  stat_cal <- function(w) {
    out <- numeric(0)
    for (v in names(cal_pre)) {
      sl <- boot_calib_slope_w(cal_pre[[v]], w, start = cal_start[[v]]$slope)
      it <- boot_calib_int_w(cal_pre[[v]], w, start = cal_start[[v]]$int)
      out <- c(out, setNames(c(sl[2], it), paste0(c("slope_", "int_"), v)))
    }
    out
  }

  t_start  <- proc.time()
  boot_cal <- boot_run(N_STAYS, BOOT_B_MAIN, stat_cal, seed = BOOT_SEED + 50L,
                       progress_every = 250L)
  v2_log(sprintf("Calibration bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
  obs_cal <- stat_cal(ones)

  calib_ci <- rbindlist(lapply(names(cal_pre), function(v) {
    cs <- perc_ci(boot_cal[, paste0("slope_", v)], CI_LEVEL)
    ci <- perc_ci(boot_cal[, paste0("int_",   v)], CI_LEVEL)
    data.table(variant = v, model = "primary", n_rows = cal_pre[[v]]$n,
               calib_slope = unname(obs_cal[paste0("slope_", v)]),
               slope_lo = cs["lo"], slope_hi = cs["hi"],
               calib_int = unname(obs_cal[paste0("int_", v)]),
               int_lo = ci["lo"], int_hi = ci["hi"])
  }))
  calib_ci[, `:=`(slope_ci = fmt_ci(slope_lo, slope_hi),
                  int_ci   = fmt_ci(int_lo, int_hi))]

  # Contrasts against the primary label, paired within replicate.
  if ("B" %in% names(cal_pre)) {
    calib_delta <- rbindlist(lapply(setdiff(names(cal_pre), "B"), function(v) {
      ds <- boot_cal[, paste0("slope_", v)] - boot_cal[, "slope_B"]
      di <- boot_cal[, paste0("int_",   v)] - boot_cal[, "int_B"]
      dso <- unname(obs_cal[paste0("slope_", v)] - obs_cal["slope_B"])
      dio <- unname(obs_cal[paste0("int_",   v)] - obs_cal["int_B"])
      cs <- perc_ci(ds, CI_LEVEL); ci <- perc_ci(di, CI_LEVEL)
      data.table(variant = v, reference = "B", model = "primary",
                 delta_slope = dso, slope_lo = cs["lo"], slope_hi = cs["hi"],
                 slope_excludes_zero = cs["lo"] > 0 | cs["hi"] < 0,
                 delta_int = dio, int_lo = ci["lo"], int_hi = ci["hi"],
                 int_excludes_zero = ci["lo"] > 0 | ci["hi"] < 0)
    }))
    calib_delta[, `:=`(slope_ci = fmt_ci(slope_lo, slope_hi),
                       int_ci   = fmt_ci(int_lo, int_hi))]
  }

  save_parquet(calib_ci, "12_calibration_ci", DATA_DIR)
  save_csv(calib_ci,     "12_calibration_ci", DATA_DIR)
  if (nrow(calib_delta) > 0) {
    save_parquet(calib_delta, "12_calibration_contrasts", DATA_DIR)
    save_csv(calib_delta,     "12_calibration_contrasts", DATA_DIR)
  }

  knitr::kable(calib_ci[, .(variant, slope = round(calib_slope, 4), slope_ci,
                            intercept = round(calib_int, 4), int_ci)],
               caption = "Primary-model calibration slope and intercept by label variant, with 95\\% stay-level cluster-bootstrap intervals")
  if (nrow(calib_delta) > 0)
    knitr::kable(calib_delta[, .(variant, reference,
                                 delta_slope = round(delta_slope, 4), slope_ci,
                                 slope_excludes_zero,
                                 delta_int = round(delta_int, 4), int_ci,
                                 int_excludes_zero)],
                 caption = "Paired cross-label calibration contrasts against Variant B")
}
```

### Variant D Transport to the Full External Cohort (Amendment 9 arm)

The internal and external cohorts are disjoint databases, so the two bootstrap
streams are drawn separately and differenced, as for H5. The external stream
uses fewer replicates than the internal one because it resamples several
million person-hours. Unlike the Sepsis-3 external arm, which has 29 onsets
under Variant B, Variant D on the full eICU cohort has far more than the 100
onsets needed for interpretation. Its label is a threshold on
physiology that is itself a model input, so the transport loss concerns a
deterioration target, not sepsis.

```{r altlabel-transport-ci}
alt_transport <- data.table()
alt_ext_path  <- file.path(DATA_DIR, "08_eicu_predictions_D.parquet")

if (file.exists(alt_ext_path) && "D" %in% names(alt_tab)) {
  ed <- as.data.table(read_parquet(alt_ext_path))
  ed[, y := as.integer(event == 1)]
  setorder(ed, stay_id, hour)
  estays_d <- sort(unique(ed$stay_id)); ed[, sidx := match(stay_id, estays_d)]
  di <- alt_tab[["D"]]

  ALT_T_MODELS <- c(primary = "pred_sepsis", gbt = "pred_gbt")
  rows_t <- list()
  for (m in names(ALT_T_MODELS)) {
    col <- ALT_T_MODELS[[m]]
    if (!col %in% names(ed) || !col %in% names(di)) next
    pre_e <- auc_precompute(ed[[col]], ed$y, ed$sidx, length(estays_d))
    pre_i <- auc_precompute(di[[col]], di$y, di$sidx, N_STAYS)
    if (is.null(pre_e) || is.null(pre_i)) next

    b_e <- boot_run(length(estays_d), BOOT_B_EXTERNAL,
                    function(w) c(auroc = boot_auroc_w(pre_e, w)),
                    seed = BOOT_SEED + 60L, progress_every = 250L)[, "auroc"]
    b_i <- boot_run(N_STAYS, BOOT_B_EXTERNAL,
                    function(w) c(auroc = boot_auroc_w(pre_i, w)),
                    seed = BOOT_SEED + 61L, progress_every = 250L)[, "auroc"]

    a_e <- boot_auroc_w(pre_e, rep(1L, length(estays_d)))
    a_i <- boot_auroc_w(pre_i, ones)
    dvec <- b_i - b_e
    dci  <- perc_ci(dvec, CI_LEVEL)
    rows_t[[m]] <- data.table(
      variant = "D", model = m,
      n_ext_stays = length(estays_d), n_ext_rows = pre_e$n, n_ext_events = sum(ed$y),
      auroc_internal = a_i, auroc_external = a_e,
      delta_auroc = a_i - a_e, ci_lo = dci["lo"], ci_hi = dci["hi"],
      excludes_zero = dci["lo"] > 0 | dci["hi"] < 0)
    rm(pre_e, pre_i); gc()
  }
  alt_transport <- rbindlist(rows_t)
  if (nrow(alt_transport) > 0) {
    alt_transport[, delta_ci := fmt_ci(ci_lo, ci_hi)]
    alt_transport[, family := "none (Amendment 9 arm; excluded from H5)"]
    save_parquet(alt_transport, "12_altlabel_transport_ci", DATA_DIR)
    save_csv(alt_transport,     "12_altlabel_transport_ci", DATA_DIR)
    knitr::kable(alt_transport[, .(model, n_ext_events,
                                   internal = round(auroc_internal, 4),
                                   external = round(auroc_external, 4),
                                   delta = round(delta_auroc, 4), delta_ci,
                                   excludes_zero)],
                 caption = "Variant D internal-to-external AUROC transport on the full eICU-CRD cohort, with 95\\% cluster-bootstrap intervals")
  }
  rm(ed); gc()
} else {
  cat("Variant D external predictions not found; run stage 08 to write 08_eicu_predictions_D.\n")
}
```

## Subgroup Interval Widths

The intervals show how precise each subgroup AUROC is, and that precision
falls as the number of onsets falls. A subgroup with few onsets gives too
little precision for a conclusion about a difference in either direction. The
figure plots each subgroup AUROC with its interval against the number of
onsets in the subgroup. Interval width generally increases as the number of
onsets decreases, and several of the smallest subgroups have intervals that
include an AUROC of 0.5. The race levels are reproduced as MIMIC-IV records
them, and UNKNOWN is a missing-information category, not a demographic group.

```{r subgroup-ci-plot, fig.cap="Subgroup AUROC with 95\\% cluster-bootstrap intervals (Variant B, temporal test). Interval width is driven almost entirely by the subgroup event count; the smallest groups cannot distinguish their AUROC from chance.", fig.width=8, fig.height=5.5}
if (nrow(sub_auroc) > 0) {
  pd <- sub_auroc[subgroup_col == "race"]
  if (nrow(pd) > 0) {
    pd[, lbl := paste0(subgroup_val, " (", n_events, ")")]
    p_ci <- ggplot(pd, aes(x = reorder(lbl, auroc), y = auroc, colour = model)) +
      geom_hline(yintercept = 0.5, linetype = "dotted", colour = "gray40") +
      geom_pointrange(aes(ymin = ci_lo, ymax = ci_hi),
                      position = position_dodge(width = 0.5), size = 0.35) +
      coord_flip() +
      labs(title    = "Subgroup AUROC with 95% cluster-bootstrap intervals",
           subtitle = "Race, Variant B, temporal test. Sepsis onset count in parentheses.",
           x = "Racial subgroup (n events)", y = "AUROC", colour = "Model") +
      theme_bw() + theme(legend.position = "bottom")
    print(p_ci)
    ggsave(file.path(FIGURES_DIR, "12_subgroup_auroc_ci.png"), p_ci,
           width = 8, height = 5.5, dpi = 150)
  }
}
```

## Analysis Register

Every analysis reported in the thesis is tagged here. The tag records the
analysis's inferential role, and a separate column records the section of the
pre-registration or the protocol amendment that specifies it. `confirmatory`
means that the estimand, and where one exists the threshold and direction,
were fixed in `00_preregistration.md` before any modelling. Some confirmatory
analyses are estimates with an interval and no test, and their correction
column says so. `exploratory` means that the analysis was pre-registered in
scope but not in its specific contrasts, so it generates hypotheses rather
than confirming them. `descriptive` means that no inferential claim is
attached. The rows for post-hoc amendments, including Amendment 6's
100-onset floor, which decides which contrasts enter family E1, state their
post-hoc status explicitly.

```{r register}
register <- rbindlist(list(
  data.table(analysis = "Cohort description by label variant",           status = "descriptive",
             prereg = "§2",        correction = "none",  reported_in = "Table: cohort"),
  data.table(analysis = "Sample size adequacy (Riley et al.)",           status = "confirmatory",
             prereg = "§10",       correction = "none (single pre-specified criterion)",
             reported_in = "Table: sample_size"),
  data.table(analysis = "Internal temporal performance (all models)",    status = "confirmatory",
             prereg = "§5, §6",    correction = "none (estimation, not testing)",
             reported_in = "Table: internal_results"),
  data.table(analysis = "Calibration intercept and slope",               status = "confirmatory",
             prereg = "§5",        correction = "none (estimation)",
             reported_in = "Table: calibration"),
  data.table(analysis = "Decision-curve net benefit",                    status = "confirmatory",
             prereg = "§6",        correction = "none (estimation)",
             reported_in = "Table: net_benefit"),
  data.table(analysis = "Operating points at fixed sensitivity",         status = "exploratory",
             prereg = "not specified (threshold chosen post hoc)",
             correction = "none; reported as descriptive operating characteristics",
             reported_in = "Table: operating_points"),
  data.table(analysis = "Per-patient detection and lead time",           status = "confirmatory",
             prereg = "§6",        correction = "none (estimation)",
             reported_in = "Table: patient_level, lead_time"),
  data.table(analysis = "H1 label dominance",                            status = "confirmatory",
             prereg = "§7",        correction = "Holm (m=6)", reported_in = "H1 box"),
  data.table(analysis = "H2 instability uncertainty diagnostic",         status = "exploratory",
             prereg = "Amendment 7 (post-hoc)",
             correction = "none; ratio to pooled SE, not a test",
             reported_in = "Table: h2_uncertainty"),
  data.table(analysis = "H2 coefficient instability",                    status = "confirmatory",
             prereg = "§7",        correction = "family slot reserved; no test statistic",
             reported_in = "H2 box, Table: coef_instability"),
  data.table(analysis = "H3 treatment leakage",                          status = "confirmatory",
             prereg = "§7",        correction = "family slot reserved; exact bound reported",
             reported_in = "H3 box"),
  data.table(analysis = "H4 split optimism",                             status = "confirmatory",
             prereg = "§7",        correction = "Holm (m=6)", reported_in = "H4 box"),
  data.table(analysis = "H5 metric divergence",                          status = "confirmatory",
             prereg = "§7",        correction = "Holm (m=6)", reported_in = "H5 box"),
  data.table(analysis = "H6 model-class null",                           status = "confirmatory",
             prereg = "§7",        correction = "Holm (m=6)", reported_in = "H6 box"),
  data.table(analysis = "External validation on eICU-CRD",               status = "confirmatory",
             prereg = "§4",        correction = "none (estimation); feeds H5",
             reported_in = "Table: external"),
  data.table(analysis = "Variance decomposition table",                  status = "confirmatory",
             prereg = "§1",        correction = "none (descriptive assembly of H1-H6 contrasts)",
             reported_in = "Table: decomposition"),
  data.table(analysis = "Proportional-hazards (Schoenfeld) testing",     status = "confirmatory",
             prereg = "§8",        correction = "none (single pre-specified check)",
             reported_in = "PH section"),
  data.table(analysis = "GBT feature importance",                        status = "exploratory",
             prereg = "not specified", correction = "none (ranking, no inference)",
             reported_in = "Table: feature_importance"),
  data.table(analysis = "Subgroup AUROC by race/language/insurance",     status = "exploratory",
             prereg = "§9 (scope only; contrasts not pre-specified)",
             correction = "Benjamini-Hochberg within family E1",
             reported_in = "Table: equity_performance"),
  data.table(analysis = "Subgroup label sensitivity across variants",    status = "exploratory",
             prereg = "§9 (scope only; contrasts not pre-specified)",
             correction = "Benjamini-Hochberg within family E2",
             reported_in = "Table: equity_label"),
  data.table(analysis = "Label sensitivity vs. AUROC association",       status = "exploratory",
             prereg = "not specified", correction = "none (visual, no test)",
             reported_in = "Figure: label_sens_vs_auroc"),
  data.table(analysis = "Alternative-label AUROC vs. Sepsis-3 (variants D, E)", status = "exploratory",
             prereg = "Protocol Amendment 2 (post hoc; §16-17)",
             correction = "Benjamini-Hochberg within family E3",
             reported_in = "Table: altlabel_contrasts"),
  data.table(analysis = "Standard onset timing min(t_susp, t_SOFA) (variant F)", status = "exploratory",
             prereg = "Protocol Amendment 4 (post hoc; §18)",
             correction = "Benjamini-Hochberg within family E3 (family enlarged)",
             reported_in = "Table: altlabel_contrasts, H3 section"),
  data.table(analysis = "Pre-treatment onsets under standard onset timing",     status = "descriptive",
             prereg = "Protocol Amendment 4 (post hoc; §18)",
             correction = "none (event count and window size, not a test)",
             reported_in = "Table: pretreatment_onsets"),
  data.table(analysis = "Anchor-attributable fraction of discrimination", status = "exploratory",
             prereg = "Protocol Amendment 2 (post hoc; §16-17)",
             correction = "none (ratio reported as an interval, no null at zero)",
             reported_in = "Table: anchor_attribution_ci"),
  data.table(analysis = "Onset agreement between label definitions",     status = "descriptive",
             prereg = "Protocol Amendment 2 (post hoc; §16-17)",
             correction = "none (descriptive agreement statistics)",
             reported_in = "Table: label_agreement"),
  # Amendments 5-7 were absent from the register, so the register and the
  # chapter's own register table disagreed about how many analyses had no
  # machine-generated counterpart. Amendment 6 is the load-bearing one: it is
  # the only place the reduction of family E1 from 30 computed contrasts to 6
  # is recorded in a file the pipeline writes.
  data.table(analysis = "Subgroup interpretability floor (family E1 reduction)",
             status = "exploratory",
             prereg = "Protocol Amendment 6 (post hoc; §20)",
             correction = sprintf(paste0("E1 restricted to contrasts in which both strata carry >= %d ",
                                         "events; %s contrasts retained. Below-floor strata are still ",
                                         "estimated and tabulated, and are excluded from the inferential ",
                                         "family only"),
                                  MIN_SUBGROUP_EVENTS_INTERPRET,
                                  if (exists("sub_delta") && !is.null(sub_delta))
                                    format(nrow(sub_delta)) else "?"),
             reported_in = "Table: equity_contrasts"),
  data.table(analysis = "Action-derived feature ablation (Amendment 5)", status = "exploratory",
             prereg = "Protocol Amendment 5 (post hoc; §19)",
             correction = "none (estimation); enters no corrected family",
             reported_in = "Table: ablation"),
  # Amendment 9's external arm had no register row at all, which is how it came
  # to be printed under a confirmatory tag in the thesis while its own table
  # said post-hoc. Registered here as descriptive, matching §23.2.
  data.table(analysis = "Variant D external validation on the full eICU-CRD cohort (Amendment 9)",
             status = "descriptive",
             prereg = "Protocol Amendment 9 (post hoc; §23)",
             correction = "none; joins no corrected family and is excluded from H5",
             reported_in = "Table: external_altlabel"),
  data.table(analysis = "Variant D internal-to-external transport interval (Amendment 10)",
             status = "descriptive",
             prereg = "Protocol Amendment 10 (post hoc; §24)",
             correction = "none (estimation); excluded from H5 on Amendment 9's terms",
             reported_in = "Table: altlabel_transport_ci"),
  data.table(analysis = "Ablation contrast intervals (Amendment 10)", status = "exploratory",
             prereg = "Protocol Amendment 10 (post hoc; §24)",
             correction = "none (estimation); enters no corrected family",
             reported_in = "Table: ablation_ci"),
  data.table(analysis = "Cross-label calibration contrast intervals (Amendment 10)",
             status = "exploratory",
             prereg = "Protocol Amendment 10 (post hoc; §24)",
             correction = "none (estimation); enters no corrected family",
             reported_in = "Table: calibration_contrasts")
  # Amendment 7 is NOT added here: the diagnostic block above already registers
  # it as "H2 instability uncertainty diagnostic". A second row was briefly
  # added in this round and removed -- a duplicated register entry is exactly
  # the kind of thing the register exists to make impossible.
))

save_parquet(register, "12_analysis_register", DATA_DIR)
save_csv(register,     "12_analysis_register", DATA_DIR)

knitr::kable(register, caption = "Analysis register: confirmatory vs.\\ exploratory status and the multiplicity correction applied to each")
```

## Reading the Corrected Family

Four principles guide the interpretation of the tables above.

1. **The family keeps its pre-registered size.** Holm is applied with
   $m = 6$, the number of pre-registered hypotheses, not the number that
   remained testable. H2 and H3 remain members of the family but contribute no
   p-value. In practice the four available p-values are ranked and multiplied
   by 6, 5, 4 and 3 in turn, with the usual running maximum. This is the same
   as Holm over six hypotheses with H2 and H3 entered as $p = 1$. Keeping the
   original family size prevents the correction from becoming less stringent
   after the data showed that two hypotheses could not be tested.
2. **Failure to meet a threshold does not establish the absence of an
   effect.** Failing to meet a pre-registered threshold and finding no effect
   are different outcomes. The estimate and its interval also show whether an
   effect is smaller than predicted or lies in the opposite direction, which
   the verdict column cannot show. The verdict should therefore be read
   together with the estimate and interval.
3. **H6's test and H6's conclusion answer different questions.** Here
   $\theta$ is GBT AUROC minus primary AUROC. The pre-registered test treats the
   margin as the null, $H_0: -0.02 \leq \theta \leq 0.02$, so rejecting it
   would show that the models differ by more than 0.02 in one direction or the
   other. Amendment 10 tests both boundaries and reports the smaller of the two
   one-sided p-values. The boundaries are 0.04 apart, about five standard
   errors of $\hat\theta$, so for any $\theta$ inside the null at most one of
   the two p-values can be small, and the test has essentially the size of a
   single one-sided test. With $\hat\theta$ negative, the upper-bound test
   returns a p-value near 1 for arithmetic reasons, and the lower-bound test is
   the informative one. This is not the two one-sided tests (TOST) procedure,
   in which equivalence is the alternative and the larger p-value governs.
   Failing to reject this null does not establish equivalence. What the data
   support is read from the interval. Its upper limit lies below $+0.02$, so
   the GBT is not better than the primary model by more than the margin, which
   is one-sided non-inferiority of the primary model. Its lower limit lies
   below $-0.02$, so two-sided equivalence within $\pm 0.02$ is not
   established; that side of the interval corresponds to the primary model
   being the better of the two.
4. **Exploratory families are for generating hypotheses, not for
   confirmatory claims.** E1, E2 and E3 are corrected by BH within themselves
   and are never pooled with the confirmatory family. A contrast that meets
   the BH criterion is an exploratory association that needs independent
   confirmation, not a confirmed effect.

## Summary

None of the four testable hypotheses meets its pre-registered threshold after
correction. H1's estimate lies on the opposite side of its null: AUROC varies
far more across model classes than across label variants, although the
model-class spread runs from the rule-based scores up to the primary model. H2
meets its point-estimate criterion, but only some of its flagged covariates
exceed the Amendment 7 benchmark. H3 cannot be tested, because the onset rule
leaves no onsets in the pre-treatment window. H4's estimate is negative, so the
random split gave the lower AUROC, the opposite of the predicted direction. H5
is tested only on its substitute AUROC contrast. That contrast has the
predicted sign but an interval that includes zero, and it says nothing about
the pre-registered metric-divergence claim. H6 is not rejected, and its
interval supports one-sided non-inferiority of the primary model but not
two-sided equivalence. The exploratory families are summarised by their
smallest adjusted p-values and the number of contrasts meeting the BH
criterion, and they require independent confirmation.

The lines below print these results in compact form. "NOT confirmed" means
that the threshold was not met, and "significant at BH q < 0.05" means that
the BH criterion was met. The exploratory lines cover each complete family,
not only the race rows shown in the tables above, which is why their smallest
p-values can differ from those tables. The thesis takes its values from the
saved tables, not from this output.

```{r summary}
cat("Confirmatory family (Holm, m = 6, FWER = 0.05):\n")
for (i in seq_len(nrow(conf)))
  cat(sprintf("  %s  %-26s est=%s  p_raw=%s  p_holm=%s  -> %s\n",
              conf$hypothesis[i], conf$claim[i],
              formatC(conf$estimate[i], format = "g", digits = 4),
              ifelse(is.na(conf$p_raw[i]),  "--", formatC(conf$p_raw[i],  format = "g", digits = 2)),
              ifelse(is.na(conf$p_holm[i]), "--", formatC(conf$p_holm[i], format = "g", digits = 2)),
              conf$verdict[i]))

if (nrow(sub_delta) > 0) {
  cat(sprintf("\nExploratory family E1: %d subgroup AUROC contrasts, %d significant at BH q < %.2f\n",
              nrow(sub_delta), sum(sub_delta$significant_bh), FDR_Q))
  cat(sprintf("  Smallest raw p = %.4f (%s, %s vs %s); smallest BH q = %.4f\n",
              min(sub_delta$p_raw), sub_delta$subgroup_val[1], sub_delta$model[1],
              sub_delta$reference[1], min(sub_delta$p_bh)))
  cat(sprintf("  Median 95%% CI width for subgroup AUROC: %.3f (max %.3f)\n",
              median(sub_auroc$ci_width, na.rm = TRUE), max(sub_auroc$ci_width, na.rm = TRUE)))
}
if (nrow(ls_delta) > 0)
  cat(sprintf("\nExploratory family E2: %d label-sensitivity contrasts, %d significant at BH q < %.2f\n",
              nrow(ls_delta), sum(ls_delta$significant_bh), FDR_Q))

if (nrow(alt_delta) > 0) {
  cat(sprintf("\nExploratory family E3: %d alternative-label contrasts, %d significant at BH q < %.2f\n",
              nrow(alt_delta), sum(alt_delta$significant_bh), FDR_Q))
  for (i in seq_len(nrow(alt_delta)))
    cat(sprintf("  %s (%s) vs B, %-7s delta=%+.4f %s  q_BH=%.4f\n",
                alt_delta$variant[i], alt_delta$label[i], alt_delta$model[i],
                alt_delta$delta_auroc[i],
                fmt_ci(alt_delta$ci_lo[i], alt_delta$ci_hi[i]), alt_delta$p_bh[i]))
}

cat("\nInference notebook complete.\n")
```
