---
title: "12 Inference - Bootstrap Uncertainty, Multiplicity Control, and Analysis Status"
author: "Abdul Ahad. Student ID: G00486649"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=2.3cm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \renewcommand{\contentsname}{Table of Contents}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
set.seed(42)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H",
  out.width = "80%", fig.width = 6, fig.height = 3.6
)
```

## Introduction

Every result reported so far is a point estimate. This notebook attaches
uncertainty to those estimates and controls the error rate incurred by testing
many of them at once. Three things are produced:

1. **Cluster-bootstrap confidence intervals** for every headline and subgroup
   estimate. Person-hours are nested within ICU stays, so the resampling unit
   is the stay: naive row-level intervals would be far too narrow.

2. **Multiplicity control.** The six pre-registered hypotheses form a
   *confirmatory* family and are corrected by Holm--Bonferroni at
   $\alpha = 0.05$ (FWER). The subgroup analyses and the alternative-label
   comparison form *exploratory* families (E1 subgroup discrimination, E2
   subgroup label sensitivity, E3 alternative-label contrasts) and are
   corrected by Benjamini--Hochberg at $q = 0.05$ (FDR). The two
   procedures are deliberately different: a confirmatory claim must survive
   strict family-wise control, whereas an exploratory scan is intended to
   generate hypotheses and only needs its false-discovery proportion bounded.

3. **An analysis register** tagging every reported analysis as confirmatory,
   exploratory, or descriptive, with the pre-registration clause that licenses
   it. The register is written to disk so the thesis tables can carry the tag
   in the table itself rather than relying on a blanket methodological
   statement.

**Status of this notebook.** The uncertainty and multiplicity plan is
*Protocol Amendment 1* (`00_preregistration.md` §13), written after the
analyses in notebooks 02--11 had been run. The amendment is therefore itself
post-hoc. It does not change any hypothesis, threshold, or estimand; it only
attaches uncertainty to contrasts that were pre-specified and imposes a
correction that the original protocol omitted. Where a correction changes a
verdict, that is reported as the corrected verdict.

Family E3 belongs to *Protocol Amendment 2* (`00_preregistration.md` §§16--17),
which adds two non-Sepsis-3 label definitions. It too is post-hoc, and it
likewise leaves every pre-registered hypothesis untouched: variants D and E are
never pooled with A--C, so H1 and the variance decomposition are computed from
exactly the same inputs as before.

```{r load-deps}
suppressPackageStartupMessages({
  library(data.table)
  library(arrow)
  library(ggplot2)
  library(dplyr)
})

source("config.R")
source("utils.R")

step("=== Phase 9: Inference, uncertainty, and multiplicity ===")

# Smoke-test override: V2_BOOT_B=50 runs the whole notebook end to end in
# seconds. Reported results always use the values fixed in config.R.
.b_override <- suppressWarnings(as.integer(Sys.getenv("V2_BOOT_B", "")))
if (!is.na(.b_override) && .b_override > 0) {
  BOOT_B_MAIN <- BOOT_B_SUBGROUP <- BOOT_B_EXTERNAL <- .b_override
  v2_log(sprintf("V2_BOOT_B override active: B = %d (smoke test, not for reporting)",
                 .b_override), level = "WARN")
}

v2_log(sprintf("Bootstrap replicates: main=%d, subgroup=%d, external=%d (seed %d)",
               BOOT_B_MAIN, BOOT_B_SUBGROUP, BOOT_B_EXTERNAL, BOOT_SEED))

MODELS <- c(primary = "pred_sepsis", gbt = "pred_gbt", news2 = "pred_news2",
            cox = "pred_cox", qsofa = "pred_qsofa", sirs = "pred_sirs")

fmt_ci <- function(lo, hi, digits = 3)
  sprintf("[%.*f, %.*f]", digits, lo, digits, hi)
```

## Data Assembly

The temporal test predictions for each label variant are joined to the
comparator predictions and to stay-level demographics. Rows are ordered by
`stay_id` so that a stay maps to a contiguous block, and each stay is given an
index into a **common resampling frame** shared by variants A, B and C. Because
all three variants are built on the same 12,132 test stays, one replicate draw
can be applied to all three simultaneously, which makes the label-variant
contrast in H1 a *paired* comparison.

```{r assemble}
load_test_table <- function(vid) {
  prim <- file.path(DATA_DIR, sprintf("05_predictions_primary_%s.parquet", vid))
  comp <- file.path(DATA_DIR, sprintf("06_predictions_comparators_%s.parquet", vid))
  if (!file.exists(prim)) return(NULL)

  dt <- as.data.table(read_parquet(prim))
  if (file.exists(comp)) {
    cp <- as.data.table(read_parquet(comp))
    keep <- intersect(c("pred_gbt", "pred_news2", "pred_cox", "pred_qsofa", "pred_sirs"),
                      names(cp))
    dt <- merge(dt, cp[, c("stay_id", "hour", keep), with = FALSE],
                by = c("stay_id", "hour"), all.x = TRUE)
  }

  demo_path <- file.path(DATA_DIR, sprintf("03_person_hours_%s.parquet", vid))
  if (file.exists(demo_path)) {
    demo <- unique(as.data.table(read_parquet(
      demo_path, col_select = c("stay_id", EQUITY_VARS[EQUITY_VARS != "ethnicity"]))))
    dt <- merge(dt, demo, by = "stay_id", all.x = TRUE)
  }

  dt[, y := as.integer(event == 1)]
  setorder(dt, stay_id, hour)
  dt[]
}

test_tab <- lapply(c(A = "A", B = "B", C = "C"), load_test_table)
test_tab <- Filter(Negate(is.null), test_tab)

# Common stay frame across variants (all variants share the same test stays)
stay_frame <- sort(Reduce(intersect, lapply(test_tab, function(d) unique(d$stay_id))))
N_STAYS    <- length(stay_frame)
for (v in names(test_tab)) test_tab[[v]][, sidx := match(stay_id, stay_frame)]

cat(sprintf("Common test-stay resampling frame: %d stays\n", N_STAYS))
for (v in names(test_tab))
  cat(sprintf("  Variant %s: %s person-hours, %d sepsis onsets\n",
              v, format(nrow(test_tab[[v]]), big.mark = ","), sum(test_tab[[v]]$y)))
```

## Internal Performance with Confidence Intervals

AUROC and AUPRC for every model on the Variant B temporal test set, with
percentile intervals from `r format(BOOT_B_MAIN, big.mark=",")` stay-level
bootstrap replicates. A single replicate is shared across all six models, so
the intervals are mutually consistent and model-to-model differences can be
read off the same replicates as paired contrasts.

```{r main-ci}
tb <- test_tab[["B"]]
pre_B <- lapply(MODELS, function(m)
  if (m %in% names(tb)) auc_precompute(tb[[m]], tb$y, tb$sidx, N_STAYS) else NULL)
names(pre_B) <- names(MODELS)
pre_B <- Filter(Negate(is.null), pre_B)

stat_main <- function(w) {
  c(vapply(pre_B, boot_auroc_w, numeric(1), w_stay = w),
    vapply(pre_B, boot_auprc_w, numeric(1), w_stay = w)) |>
    setNames(c(paste0("auroc_", names(pre_B)), paste0("auprc_", names(pre_B))))
}

t_start   <- proc.time()
boot_main <- boot_run(N_STAYS, BOOT_B_MAIN, stat_main, seed = BOOT_SEED,
                      progress_every = 500L)
v2_log(sprintf("Main bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))

ones <- rep(1L, N_STAYS)
main_ci <- rbindlist(lapply(names(pre_B), function(m) {
  a <- perc_ci(boot_main[, paste0("auroc_", m)], CI_LEVEL)
  p <- perc_ci(boot_main[, paste0("auprc_", m)], CI_LEVEL)
  data.table(model = m,
             n_rows = pre_B[[m]]$n,
             n_events = sum(tb$y),
             auroc = boot_auroc_w(pre_B[[m]], ones), auroc_lo = a["lo"], auroc_hi = a["hi"],
             auprc = boot_auprc_w(pre_B[[m]], ones), auprc_lo = p["lo"], auprc_hi = p["hi"])
}))
main_ci[, `:=`(auroc_ci = fmt_ci(auroc_lo, auroc_hi),
               auprc_ci = fmt_ci(auprc_lo, auprc_hi, 4))]

save_parquet(main_ci, "12_internal_ci", DATA_DIR)
save_csv(main_ci, "12_internal_ci", DATA_DIR)

knitr::kable(main_ci[, .(model, n_events, auroc = round(auroc, 4), auroc_ci,
                          auprc = round(auprc, 5), auprc_ci)],
             caption = "Internal temporal test performance with 95\\% stay-level cluster bootstrap intervals (Variant B)")
```

## Confirmatory Family: the Six Pre-Registered Hypotheses

Each hypothesis is expressed as a single scalar contrast $\theta$ tested
against its **pre-registered** threshold $\theta_0$. Because the protocol
states each hypothesis directionally, the tests are one-sided. P-values are
obtained by recentring the bootstrap distribution on the null boundary and
reading off the tail area, with the $(1+\#)/(B+1)$ convention so that no
p-value is reported as exactly zero.

### Contrast Estimation

H1 and H6 are drawn from the same replicates. Both compare quantities measured
on one set of test stays, so a single resampling of stays serves both, and the
resulting intervals keep the correlation between the spreads being differenced
rather than treating them as independent.

```{r confirmatory-boot}
# --- H1 / H6: label spread, model spread, and the primary-vs-GBT gap --------
pre_label <- lapply(names(test_tab), function(v) {
  d <- test_tab[[v]]
  auc_precompute(d$pred_sepsis, d$y, d$sidx, N_STAYS)
})
names(pre_label) <- names(test_tab)

stat_h1h6 <- function(w) {
  lab <- vapply(pre_label, boot_auroc_w, numeric(1), w_stay = w)
  mod <- vapply(pre_B,     boot_auroc_w, numeric(1), w_stay = w)
  c(label_spread = diff(range(lab, na.rm = TRUE)),
    model_spread = diff(range(mod, na.rm = TRUE)),
    h1_delta     = diff(range(lab, na.rm = TRUE)) - diff(range(mod, na.rm = TRUE)),
    h6_delta     = unname(mod["gbt"] - mod["primary"]))
}

t_start  <- proc.time()
boot_h1  <- boot_run(N_STAYS, BOOT_B_MAIN, stat_h1h6, seed = BOOT_SEED + 1L,
                     progress_every = 500L)
v2_log(sprintf("H1/H6 bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
obs_h1h6 <- stat_h1h6(ones)
```

```{r confirmatory-h4}
# --- H4: random-split optimism (independent test set, so an independent draw)
rnd_prim <- file.path(DATA_DIR, "05_predictions_primary_random_B.parquet")
boot_h4 <- NULL; obs_h4 <- NA_real_
if (file.exists(rnd_prim)) {
  rd <- as.data.table(read_parquet(rnd_prim))
  rd[, y := as.integer(event == 1)]
  setorder(rd, stay_id, hour)
  rstays <- sort(unique(rd$stay_id)); rd[, sidx := match(stay_id, rstays)]
  pre_rnd <- auc_precompute(rd$pred_sepsis, rd$y, rd$sidx, length(rstays))

  b_rnd <- boot_run(length(rstays), BOOT_B_MAIN,
                    function(w) c(auroc = boot_auroc_w(pre_rnd, w)),
                    seed = BOOT_SEED + 2L, progress_every = 1000L)[, "auroc"]
  b_tmp <- boot_run(N_STAYS, BOOT_B_MAIN,
                    function(w) c(auroc = boot_auroc_w(pre_B$primary, w)),
                    seed = BOOT_SEED + 3L, progress_every = 1000L)[, "auroc"]
  # independent samples: pair the two bootstrap streams to propagate both variances
  boot_h4 <- b_rnd - b_tmp
  obs_h4  <- boot_auroc_w(pre_rnd, rep(1L, length(rstays))) -
             boot_auroc_w(pre_B$primary, ones)
  cat(sprintf("H4: random AUROC = %.4f, temporal AUROC = %.4f, delta = %.4f\n",
              boot_auroc_w(pre_rnd, rep(1L, length(rstays))),
              boot_auroc_w(pre_B$primary, ones), obs_h4))
}
```

```{r confirmatory-h5}
# --- H5: internal -> external degradation on eICU-CRD -----------------------
eicu_path <- file.path(DATA_DIR, "08_eicu_predictions_B.parquet")
boot_h5 <- NULL; obs_h5 <- NA_real_; ext_auroc <- NA_real_
ext_n_events <- NA_integer_; ext_interpretable <- NA
if (file.exists(eicu_path)) {
  ed <- as.data.table(read_parquet(eicu_path))
  ed[, y := as.integer(event == 1)]
  # The external event count governs whether the H5 estimate can be read at
  # all. Stage 08 restricts the external cohort to the stays in which the
  # Sepsis-3 anchor is recoverable; if that cohort still yields fewer than
  # MIN_EXTERNAL_EVENTS events the interval is wider than the effect H5
  # concerns, and the verdict is carried with that flag attached.
  ext_n_events      <- sum(ed$y)
  ext_interpretable <- ext_n_events >= MIN_EXTERNAL_EVENTS
  if (!ext_interpretable)
    v2_log(sprintf(paste0("H5: external cohort has %d events, below the %d-event ",
                          "floor; the AUROC estimate is not interpretable"),
                   ext_n_events, MIN_EXTERNAL_EVENTS), level = "WARN")
  setorder(ed, stay_id, hour)
  estays <- sort(unique(ed$stay_id)); ed[, sidx := match(stay_id, estays)]
  pre_ext <- auc_precompute(ed$pred_sepsis, ed$y, ed$sidx, length(estays))
  ext_auroc <- boot_auroc_w(pre_ext, rep(1L, length(estays)))

  t_start <- proc.time()
  b_ext <- boot_run(length(estays), BOOT_B_EXTERNAL,
                    function(w) c(auroc = boot_auroc_w(pre_ext, w)),
                    seed = BOOT_SEED + 4L, progress_every = 250L)[, "auroc"]
  v2_log(sprintf("External bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
  b_int <- boot_run(N_STAYS, BOOT_B_EXTERNAL,
                    function(w) c(auroc = boot_auroc_w(pre_B$primary, w)),
                    seed = BOOT_SEED + 5L, progress_every = 500L)[, "auroc"]

  boot_h5 <- b_int - b_ext
  obs_h5  <- boot_auroc_w(pre_B$primary, ones) - ext_auroc
  ext_ci  <- perc_ci(b_ext, CI_LEVEL)
  cat(sprintf("H5: internal AUROC = %.4f, external AUROC = %.4f %s, degradation = %.4f\n",
              boot_auroc_w(pre_B$primary, ones), ext_auroc,
              fmt_ci(ext_ci["lo"], ext_ci["hi"]), obs_h5))
}
```

```{r confirmatory-h2h3}
# --- H2: coefficient stability (descriptive; no test statistic) -------------
# H2's estimand is a threshold rule on point estimates -- sign reversal, or a
# doubling in magnitude, across the pre-registered label variants -- and it
# stays exactly that. It is not a contrast and carries no test, and it consumes
# a slot in the family size regardless, which makes the correction on the other
# hypotheses stricter.
#
# Protocol Amendment 7 adds a diagnostic beside it, not in place of it. Once
# stage 05 could compute a cluster-robust covariance the obvious question about
# the count became answerable: does a flagged coefficient move by more than it
# is estimated to within? A threshold rule cannot distinguish a coefficient
# that genuinely shifts from one that was never estimated precisely enough for
# a doubling to mean anything. The pre-registered count is reported first and
# unmodified; the diagnostic is reported second and separately.
coef_path <- file.path(DATA_DIR, "05_coef_all_variants.parquet")
h2_sign <- NA_integer_; h2_mag <- NA_integer_; h2_any <- NA_integer_; h2_n_terms <- NA_integer_
h2_n_robust <- NA_integer_; h2_n_noise <- NA_integer_
if (file.exists(coef_path)) {
  ct    <- as.data.table(read_parquet(coef_path))
  # Shared with the Amendment 5 ablation arm in 05_primary_model.Rmd, so the
  # two instability counts are produced by one rule and remain comparable.
  flags <- coef_stability_flags(ct, variants = c("A", "B", "C"))

  # Instability split by whether the covariate records physiology or clinician
  # behaviour. This is descriptive and sits outside the confirmatory family.
  kind_split <- stability_by_feature_kind(flags)
  save_parquet(kind_split, "12_h2_stability_by_kind", DATA_DIR)
  save_csv(kind_split,     "12_h2_stability_by_kind", DATA_DIR)
  cat("H2 instability by feature kind:\n"); print(kind_split)

  h2_sign    <- sum(flags$sign_flip)
  h2_mag     <- sum(flags$mag_flip)
  h2_any     <- sum(flags$sign_flip | flags$mag_flip)
  h2_n_terms <- nrow(flags)
  save_parquet(flags, "12_h2_coefficient_stability", DATA_DIR)
  save_csv(flags,     "12_h2_coefficient_stability", DATA_DIR)
  cat(sprintf("H2: of %d shared covariates, %d reverse sign, %d change magnitude >=2x, %d do either\n",
              h2_n_terms, h2_sign, h2_mag, h2_any))

  # Amendment 7: the uncertainty diagnostic on the flagged covariates.
  h2_unc <- h2_stability_uncertainty(ct, flags, variants = c("A", "B", "C"))
  if (nrow(h2_unc) > 0) {
    h2_unc[, is_action := vapply(as.character(term), function(tm)
      any(startsWith(tm, ACTION_DERIVED_FEATURES)), logical(1))]
    setorder(h2_unc, -z_ratio)
    save_parquet(h2_unc, "12_h2_stability_uncertainty", DATA_DIR)
    save_csv(h2_unc,     "12_h2_stability_uncertainty", DATA_DIR)
    h2_n_robust <- sum(h2_unc$exceeds_noise, na.rm = TRUE)
    h2_n_noise  <- sum(!h2_unc$exceeds_noise, na.rm = TRUE)
    cat(sprintf(paste0("H2 uncertainty diagnostic: of the %d flagged covariates, ",
                       "%d exceed their pooled cluster-robust SE and %d do not; ",
                       "%d of the %d that do are action-derived\n"),
                nrow(h2_unc), h2_n_robust, h2_n_noise,
                sum(h2_unc$exceeds_noise & h2_unc$is_action, na.rm = TRUE),
                h2_n_robust))
    print(h2_unc[, .(term, max_diff = round(max_diff, 4),
                     pooled_se = round(pooled_se, 4),
                     z = round(z_ratio, 2), exceeds_noise, is_action)])
  } else {
    v2_log(paste0("H2 uncertainty diagnostic empty: no se_cluster column in ",
                  "05_coef_all_variants, or no covariate flagged. Re-run ",
                  "stage 05 if the covariance was expected."), level = "WARN")
  }
}

# --- H3: treatment-anchored window (degenerate; exact bound instead) --------
mt <- as.data.table(read_parquet(file.path(DATA_DIR, "07_metric_results.parquet")))
anch <- mt[eval_window == "treatment_anchored" & variant == "B" & model == "primary"]
h3_n_hours  <- if (nrow(anch)) anch$n_rows[1] else NA_integer_
h3_n_events <- if (nrow(anch)) anch$n_sepsis_events[1] else NA_integer_
h3_upper    <- if (!is.na(h3_n_hours) && h3_n_events == 0) zero_event_upper(h3_n_hours, CI_LEVEL) else NA_real_
if (!is.na(h3_upper))
  cat(sprintf("H3: %d sepsis onsets in %s pre-treatment person-hours; exact 95%% upper bound on the per-hour onset rate = %.3g (1 in %s)\n",
              h3_n_events, format(h3_n_hours, big.mark = ","), h3_upper,
              format(round(1 / h3_upper), big.mark = ",")))

# The bound above is exact but it is a bound on a quantity the label definition
# fixes at zero (see the proposition in 09_variance_decomposition.Rmd), so it
# carries no information about MIMIC-IV. The informative counterpart is the
# same window under standard min(t_susp, t_SOFA) onset timing, which is Variant
# F: identical membership, timing branch restored. Reported here so the two
# numbers appear side by side.
alt_mt <- if (file.exists(file.path(DATA_DIR, "07_metric_results_altlabel.parquet")))
            as.data.table(read_parquet(file.path(DATA_DIR, "07_metric_results_altlabel.parquet"))) else NULL
h3_f <- if (!is.null(alt_mt))
          alt_mt[eval_window == "treatment_anchored" & variant == "F" & model == "primary"] else NULL
if (!is.null(h3_f) && nrow(h3_f) == 1)
  cat(sprintf("H3 (empirical counterpart, Variant F): %d sepsis onsets in %s pre-treatment person-hours, AUROC %s\n",
              h3_f$n_sepsis_events[1], format(h3_f$n_rows[1], big.mark = ","),
              if (is.na(h3_f$auroc[1])) "undefined" else sprintf("%.4f", h3_f$auroc[1])))
```

### Corrected Confirmatory Verdicts

Each hypothesis becomes one row: the contrast, its interval, the null it is
referred to, and the raw p-value Holm will correct. The two spreads are
persisted as data here rather than left inside H1's free-text note, so the
thesis table reads them instead of transcribing them by hand.

```{r confirmatory-table}
mk <- function(id, label, contrast, theta, ci, null_txt, p, note = "") {
  data.table(hypothesis = id, claim = label, contrast = contrast,
             estimate = theta, ci_lo = ci[1], ci_hi = ci[2],
             null_hypothesis = null_txt, p_raw = p, note = note)
}

rows <- list()

# The two spreads and their intervals are the body of the decomposition table
# in the thesis. They used to exist only inside H1's free-text note, which meant
# the table transcribed them by hand; they are persisted as data instead.
spread_ci <- rbindlist(lapply(c("label", "model"), function(q) {
  col <- paste0(q, "_spread")
  ci  <- perc_ci(boot_h1[, col], CI_LEVEL)
  data.table(quantity = q,
             estimate = round(unname(obs_h1h6[col]), 4),
             ci_lo    = round(unname(ci["lo"]), 4),
             ci_hi    = round(unname(ci["hi"]), 4))
}))
save_parquet(spread_ci, "12_spread_ci", DATA_DIR)
save_csv(spread_ci,     "12_spread_ci", DATA_DIR)
print(spread_ci)

# H1: theta = label spread - model spread; H0: theta <= 0
ci_h1 <- perc_ci(boot_h1[, "h1_delta"], CI_LEVEL)
p_h1  <- boot_pvalue(boot_h1[, "h1_delta"], obs_h1h6["h1_delta"], 0, "greater")
rows[["H1"]] <- mk("H1", "Label dominance",
                   "AUROC spread across label variants - spread across model classes",
                   unname(obs_h1h6["h1_delta"]), ci_h1, "theta <= 0", p_h1,
                   sprintf("label spread %.4f %s; model spread %.4f %s",
                           obs_h1h6["label_spread"],
                           do.call(fmt_ci, as.list(perc_ci(boot_h1[, "label_spread"], CI_LEVEL))),
                           obs_h1h6["model_spread"],
                           do.call(fmt_ci, as.list(perc_ci(boot_h1[, "model_spread"], CI_LEVEL)))))

# H2: descriptive only
rows[["H2"]] <- mk("H2", "Coefficient instability",
                   "Covariates with sign reversal or >=2x magnitude change across variants",
                   as.numeric(h2_any), c(NA_real_, NA_real_),
                   "not testable (threshold rule on point estimates, not a contrast)", NA_real_,
                   sprintf(paste0("%d of %d covariates unstable (%d sign reversals, ",
                                  "%d magnitude >=2x). Amendment 7 diagnostic: %s of the %d ",
                                  "flagged exceed their pooled cluster-robust SE; the ",
                                  "remainder move by less than their own estimation noise. ",
                                  "The count above is the pre-registered estimand and is ",
                                  "unmodified by the diagnostic"),
                           h2_any, h2_n_terms, h2_sign, h2_mag,
                           ifelse(is.na(h2_n_robust), "an unrecorded number", h2_n_robust),
                           h2_any))

# H3: degenerate; exact bound rather than a test
rows[["H3"]] <- mk("H3", "Treatment leakage",
                   "Sepsis onsets in the pre-treatment (treatment-anchored) window",
                   as.numeric(h3_n_events), c(NA_real_, NA_real_),
                   "not testable (outcome absent from the window)", NA_real_,
                   sprintf("%d onsets in %s person-hours; exact 95%% upper bound on the per-hour onset rate = %.2e (1 in %s)",
                           h3_n_events, format(h3_n_hours, big.mark = ","), h3_upper,
                           format(round(1 / h3_upper), big.mark = ",")))

# H4: theta = random - temporal AUROC; H0: theta <= 0.02
if (!is.null(boot_h4)) {
  ci_h4 <- perc_ci(boot_h4, CI_LEVEL)
  p_h4  <- boot_pvalue(boot_h4, obs_h4, 0.02, "greater")
  rows[["H4"]] <- mk("H4", "Split optimism",
                     "Random-split AUROC - temporal-split AUROC (primary, Variant B)",
                     obs_h4, ci_h4, "theta <= 0.02", p_h4,
                     paste("the two test sets are resampled separately and both",
                           "bootstrap streams are propagated into the difference.",
                           "They are not disjoint: the random-split test set is",
                           "a 25% sample across all admission years and so shares",
                           "stays with the 2017-2019 temporal test set. Ignoring",
                           "that positive dependence over-states the variance of",
                           "the difference, so the interval is conservative"))
}

# H5: theta = internal - external AUROC; H0: theta <= 0
if (!is.null(boot_h5)) {
  ci_h5 <- perc_ci(boot_h5, CI_LEVEL)
  p_h5  <- boot_pvalue(boot_h5, obs_h5, 0, "greater")
  rows[["H5"]] <- mk("H5", "Metric divergence",
                     "Internal - external AUROC (primary, Variant B, eICU-CRD)",
                     obs_h5, ci_h5, "theta <= 0", p_h5,
                     sprintf("external AUROC %.4f on %d external events (%s the %d-event interpretability floor); Utility Score arm not testable: the internal score is %.3f at the default cut-off and %.3f at its best threshold, so a proportional decline has no baseline to be proportional to",
                             ext_auroc, ext_n_events,
                             if (isTRUE(ext_interpretable)) "at or above" else "BELOW",
                             MIN_EXTERNAL_EVENTS,
                             mt[variant == "B" & model == "primary" & eval_window == "unrestricted",
                                utility_normalised][1],
                             mt[variant == "B" & model == "primary" & eval_window == "unrestricted",
                                utility_max][1]))
}

# H6: equivalence bound 0.02; rejecting H6 requires evidence that |delta| > 0.02.
#
# Protocol Amendment 10. The null is an INTERVAL, -0.02 <= theta <= 0.02, so it
# has two boundaries and rejecting it means clearing one of them. Earlier runs
# tested the upper boundary only. That limb asks whether the boosted trees beat
# the hazard model by more than the margin, which is the pre-registered concern
# and the right thing to report -- but with theta-hat on the NEGATIVE side of
# zero it is also the limb the data cannot speak to, and it returns p ~ 1 for
# arithmetic reasons rather than evidential ones. Both limbs are now computed.
#
# The reported p_raw is min(p_upper, p_lower), which is a valid (conservative)
# p-value for an interval null: under any theta inside [-0.02, 0.02] at most one
# limb can be small, because a theta near one boundary is two margins away from
# the other. The verdict rule is unchanged, and so is the direction of the
# conclusion: rejection would mean non-equivalence, and neither limb rejects.
ci_h6 <- perc_ci(boot_h1[, "h6_delta"], CI_LEVEL)
p_h6_upper <- boot_pvalue(boot_h1[, "h6_delta"], obs_h1h6["h6_delta"],  0.02, "greater")
p_h6_lower <- boot_pvalue(boot_h1[, "h6_delta"], obs_h1h6["h6_delta"], -0.02, "less")
p_h6 <- min(p_h6_upper, p_h6_lower, na.rm = TRUE)
rows[["H6"]] <- mk("H6", "Model-class null",
                   "GBT AUROC - primary AUROC (Variant B, temporal)",
                   unname(obs_h1h6["h6_delta"]), ci_h6,
                   "-0.02 <= theta <= 0.02 (equivalence; interval null, two limbs)",
                   p_h6,
                   sprintf(paste0("paired: both models scored on the same ",
                                  "person-hours. Interval null tested at both ",
                                  "boundaries (Amendment 10): upper limb ",
                                  "(theta > +0.02, GBT better by more than the ",
                                  "margin) p = %.3f; lower limb (theta < -0.02, ",
                                  "primary better by more than the margin) ",
                                  "p = %.3f. The reported value is the smaller ",
                                  "of the two. With theta-hat = %.4f the lower ",
                                  "limb is the informative one; the upper limb ",
                                  "is large by construction and establishes ",
                                  "nothing on its own"),
                           p_h6_upper, p_h6_lower, unname(obs_h1h6["h6_delta"])))

conf <- rbindlist(rows)
setorder(conf, hypothesis)

# Holm-Bonferroni over the pre-registered family size, not over the number of
# tests that happened to be computable.
p_vec <- conf$p_raw
m_fam <- CONFIRMATORY_FAMILY_SIZE
ord   <- order(p_vec, na.last = NA)
p_adj <- rep(NA_real_, length(p_vec))
if (length(ord) > 0) {
  running <- 0
  for (k in seq_along(ord)) {
    running <- max(running, (m_fam - k + 1) * p_vec[ord[k]])
    p_adj[ord[k]] <- min(1, running)
  }
}
conf[, p_holm := p_adj]
conf[, significant_holm := !is.na(p_holm) & p_holm < FWER_ALPHA]
conf[, verdict := fifelse(is.na(p_raw), "Descriptive (not testable)",
                   fifelse(significant_holm,
                           fifelse(hypothesis == "H6", "H6 REJECTED (non-equivalence)", "CONFIRMED"),
                           fifelse(hypothesis == "H6", "H6 not rejected", "NOT confirmed")))]
conf[hypothesis == "H2", verdict := "Descriptive: instability present"]
conf[hypothesis == "H3", verdict := "Structural: outcome absent from window"]
# H5's slot is occupied by a substituted estimand (pre-registration 13.6): the
# proportional utility-versus-AUROC divergence cannot be formed against an
# internal Utility Score indistinguishable from zero, so the AUROC limb alone
# is tested. A bare "NOT confirmed" would read as a hypothesis that was tested
# and failed. The verdict names the substitution so that every table printing
# it carries the qualification with it.
conf[hypothesis == "H5",
     verdict := "Not evaluable as pre-registered; substituted AUROC contrast inconclusive"]
conf[, family := "confirmatory (pre-registered)"]
conf[, correction := sprintf("Holm-Bonferroni, m=%d, FWER=%.2f", m_fam, FWER_ALPHA)]

# H6's two limbs are persisted beside its reported p-value so the thesis can
# print both rather than describing one of them (Amendment 10). Every other
# hypothesis has a one-sided null with a single boundary, so the columns are NA.
conf[, `:=`(p_limb_upper = NA_real_, p_limb_lower = NA_real_)]
conf[hypothesis == "H6", `:=`(p_limb_upper = p_h6_upper, p_limb_lower = p_h6_lower)]

save_parquet(conf, "12_confirmatory_tests", DATA_DIR)
save_csv(conf, "12_confirmatory_tests", DATA_DIR)

knitr::kable(
  conf[, .(H = hypothesis, claim, estimate = round(estimate, 4),
           CI = fifelse(is.na(ci_lo), "--", fmt_ci(ci_lo, ci_hi)),
           H0 = null_hypothesis,
           p = fifelse(is.na(p_raw), "--", formatC(p_raw, format = "g", digits = 2)),
           p_holm = fifelse(is.na(p_holm), "--", formatC(p_holm, format = "g", digits = 2)),
           verdict)],
  caption = "Confirmatory family: six pre-registered hypotheses, one-sided cluster-bootstrap tests against their pre-registered thresholds, Holm-Bonferroni corrected at family size 6")
```

```{r confirmatory-notes}
knitr::kable(conf[nzchar(note), .(H = hypothesis, note)],
             caption = "Supporting detail for the confirmatory family")
```

## Exploratory Family E1: Subgroup Discrimination

Subgroup AUROC is estimated with a bootstrap **stratified by subgroup level**:
stays are resampled within each level, so the level sizes are held at their
observed values. This is the correct design when the subgroups themselves are
the estimand rather than a random draw.

Each subgroup is contrasted against the largest level of its own variable
(the reference), and the resulting family of differences is corrected by
Benjamini--Hochberg. The reference level is chosen by sample size, not by
outcome, and is recorded in the output.

```{r subgroup-auroc}
subgroup_models <- c(primary = "pred_sepsis", gbt = "pred_gbt")

sub_rows <- list(); sub_diff <- list()

for (svar in EQUITY_VARS) {
  if (!svar %in% names(tb)) next
  # blank strings are missing data, not a demographic category
  dsub <- tb[!is.na(get(svar)) & nzchar(trimws(as.character(get(svar))))]
  if (nrow(dsub) == 0) next

  lvl_n  <- dsub[, .N, by = svar][order(-N)]
  levels_keep <- head(as.character(lvl_n[[svar]]), MAX_SUBGROUP_LEVELS)
  dsub <- dsub[as.character(get(svar)) %chin% levels_keep]
  if (nrow(dsub) == 0) next

  # stay-level stratification frame for this variable
  stay_lvl <- unique(dsub[, .(stay_id, lvl = as.character(get(svar)))])
  stay_lvl <- stay_lvl[!duplicated(stay_id)]
  setorder(stay_lvl, lvl, stay_id)
  stay_lvl[, k := .I]
  dsub <- merge(dsub, stay_lvl[, .(stay_id, k)], by = "stay_id")
  n_k  <- nrow(stay_lvl)

  # eligible levels: enough rows and enough events to support an AUROC
  elig <- dsub[, .(n_rows = .N, n_events = sum(y), n_stays = uniqueN(stay_id)),
               by = .(lvl = as.character(get(svar)))]
  elig <- elig[n_rows >= MIN_SUBGROUP_ROWS & n_events >= MIN_SUBGROUP_EVENTS]
  if (nrow(elig) == 0) next
  ref_lvl <- elig[order(-n_rows)][1, lvl]

  # Two distinct thresholds, and the distinction is the point. The filter above
  # decides whether an AUROC can be COMPUTED at all; the flag below decides
  # whether it is INTERPRETED. Levels under the interpretability floor are
  # still estimated and reported -- suppressing them would hide which groups
  # the cohort contains and how few events they carry -- but they are kept out
  # of the contrast family, so the BH correction is applied over the analyses
  # the study is willing to read rather than over every level that exists.
  elig[, interpretable := n_events >= MIN_SUBGROUP_EVENTS_INTERPRET]
  v2_log(sprintf(
    "  %s: %d levels estimated, %d above the %d-event interpretability floor",
    svar, nrow(elig), sum(elig$interpretable), MIN_SUBGROUP_EVENTS_INTERPRET))

  # Estimand keys are opaque ids paired with a lookup table: subgroup labels
  # contain slashes, spaces and hyphens, so they are never parsed out of a key.
  pre_sub <- list(); key_tab <- list()
  for (mn in names(subgroup_models)) {
    mc <- subgroup_models[[mn]]
    if (!mc %in% names(dsub)) next
    for (lv in elig$lvl) {
      kk  <- sprintf("e%03d", length(pre_sub) + 1L)
      dd  <- dsub[as.character(get(svar)) == lv]
      pre_sub[[kk]] <- auc_precompute(dd[[mc]], dd$y, dd$k, n_k)
      key_tab[[kk]] <- data.table(estimand = kk, model = mn, lvl = lv)
    }
  }
  if (length(pre_sub) == 0) next
  key_tab <- rbindlist(key_tab)

  stat_sub <- function(w) vapply(pre_sub, boot_auroc_w, numeric(1), w_stay = w)
  v2_log(sprintf("Subgroup bootstrap: %s (%d levels, %d estimands)",
                 svar, nrow(elig), length(pre_sub)))
  b_sub <- boot_run(n_k, BOOT_B_SUBGROUP, stat_sub,
                    strata = stay_lvl$lvl, seed = BOOT_SEED + 10L,
                    progress_every = 1000L)
  obs_sub <- stat_sub(rep(1L, n_k))

  for (i in seq_len(nrow(key_tab))) {
    nm <- key_tab$estimand[i]; mn <- key_tab$model[i]; lv <- key_tab$lvl[i]
    ci <- perc_ci(b_sub[, nm], CI_LEVEL)
    e  <- elig[lvl == lv]
    sub_rows[[paste(svar, nm)]] <- data.table(
      subgroup_col = svar, subgroup_val = lv, model = mn,
      is_reference = lv == ref_lvl,
      n_rows = e$n_rows, n_stays = e$n_stays, n_events = e$n_events,
      auroc = unname(obs_sub[nm]), ci_lo = ci["lo"], ci_hi = ci["hi"],
      ci_width = unname(ci["hi"] - ci["lo"]),
      interpretable = e$interpretable,
      min_events_interpret = MIN_SUBGROUP_EVENTS_INTERPRET)

    ref_interpretable <- elig[lvl == ref_lvl, interpretable]
    if (lv != ref_lvl && isTRUE(e$interpretable) && isTRUE(ref_interpretable)) {
      ref_nm <- key_tab[model == mn & lvl == ref_lvl, estimand]
      if (length(ref_nm) == 1L) {
        dvec <- b_sub[, nm] - b_sub[, ref_nm]
        dobs <- unname(obs_sub[nm] - obs_sub[ref_nm])
        dci  <- perc_ci(dvec, CI_LEVEL)
        sub_diff[[paste(svar, nm)]] <- data.table(
          subgroup_col = svar, subgroup_val = lv, model = mn,
          reference = ref_lvl, n_events = e$n_events,
          delta_auroc = dobs, ci_lo = dci["lo"], ci_hi = dci["hi"],
          p_raw = boot_pvalue(dvec, dobs, 0, "two.sided"))
      }
    }
  }
}

sub_auroc <- rbindlist(sub_rows)
sub_delta <- rbindlist(sub_diff)
```

```{r subgroup-fdr}
if (nrow(sub_delta) > 0) {
  sub_delta[, p_bh := p.adjust(p_raw, method = "BH")]
  sub_delta[, p_bonf := p.adjust(p_raw, method = "bonferroni")]
  sub_delta[, significant_bh := p_bh < FDR_Q]
  sub_delta[, family := sprintf("exploratory E1 (m=%d subgroup AUROC contrasts)", nrow(sub_delta))]
  setorder(sub_delta, p_raw)
}
sub_auroc[, family := "exploratory E1"]

save_parquet(sub_auroc, "12_subgroup_auroc_ci", DATA_DIR)
save_csv(sub_auroc,     "12_subgroup_auroc_ci", DATA_DIR)
if (nrow(sub_delta) > 0) {
  save_parquet(sub_delta, "12_subgroup_auroc_contrasts", DATA_DIR)
  save_csv(sub_delta,     "12_subgroup_auroc_contrasts", DATA_DIR)
}

knitr::kable(
  sub_auroc[subgroup_col == "race"][order(model, -n_rows),
    .(model, subgroup = subgroup_val, n_rows, n_events,
      auroc = round(auroc, 3), CI = fmt_ci(ci_lo, ci_hi), width = round(ci_width, 3))],
  caption = "Subgroup AUROC by race with 95\\% stratified cluster-bootstrap intervals (Variant B, temporal test)")
```

```{r subgroup-fdr-contrasts}
if (nrow(sub_delta) > 0)
  knitr::kable(
    sub_delta[subgroup_col == "race"][order(model, p_raw),
      .(model, subgroup = subgroup_val, reference, n_events,
        delta = round(delta_auroc, 3), CI = fmt_ci(ci_lo, ci_hi),
        p = formatC(p_raw, format = "f", digits = 3),
        q_BH = formatC(p_bh, format = "f", digits = 3), significant_bh)],
    caption = "Subgroup AUROC differences vs.\\ the reference group, Benjamini-Hochberg corrected across the whole exploratory family E1")
```

## Exploratory Family E2: Subgroup Label Sensitivity

The fraction of stays labelled septic under each variant is a stay-level
proportion, so the bootstrap resamples whole stays within subgroup. Because the
same 65,078 stays are labelled under all three variants, the A/B/C comparison
is paired within replicate, and the interval on the A--C *range* correctly
reflects that pairing.

```{r label-sens-ci}
coh <- lapply(c(A = "A", B = "B", C = "C"), function(v) {
  p <- file.path(DATA_DIR, sprintf("02_cohort_%s.parquet", v))
  if (!file.exists(p)) return(NULL)
  d <- as.data.table(read_parquet(p))
  d[, .(stay_id, septic = as.integer(!is.na(onset_hour)),
        race, language, insurance)]
})
coh <- Filter(Negate(is.null), coh)

ls_rows <- list(); ls_diff <- list()

if (length(coh) >= 2) {
  base <- coh[[1]]
  common_stays <- sort(Reduce(intersect, lapply(coh, function(d) d$stay_id)))
  for (v in names(coh)) {
    coh[[v]] <- coh[[v]][stay_id %in% common_stays]
    setorder(coh[[v]], stay_id)
  }
  n_cs <- length(common_stays)

  for (svar in EQUITY_VARS) {
    if (!svar %in% names(coh[["B"]])) next
    lv_dt <- coh[["B"]][!is.na(get(svar)) & nzchar(trimws(as.character(get(svar)))),
                        .N, by = svar][order(-N)]
    levels_keep <- head(as.character(lv_dt[[svar]]), MAX_SUBGROUP_LEVELS)
    if (length(levels_keep) < 2) next

    lvl_vec <- as.character(coh[["B"]][[svar]])
    keep    <- !is.na(lvl_vec) & lvl_vec %chin% levels_keep
    idx     <- which(keep)
    lvl_k   <- lvl_vec[idx]
    ord     <- order(lvl_k, common_stays[idx])
    idx     <- idx[ord]; lvl_k <- lvl_k[ord]
    n_k     <- length(idx)

    sep_mat <- vapply(names(coh), function(v) coh[[v]]$septic[idx], numeric(n_k))
    lvl_f   <- factor(lvl_k, levels = levels_keep[levels_keep %in% lvl_k])
    ref_lvl <- names(sort(table(lvl_k), decreasing = TRUE))[1]

    stat_ls <- function(w) {
      num <- rowsum(w * sep_mat, lvl_f, reorder = FALSE)
      den <- as.vector(rowsum(w, lvl_f, reorder = FALSE))
      pct <- 100 * num / den
      rng <- apply(pct, 1, function(r) diff(range(r)))
      out <- c(as.vector(pct), rng)
      names(out) <- c(outer(rownames(pct), colnames(sep_mat),
                            function(a, b) paste0("pct_", b, "||", a)),
                      paste0("range||", rownames(pct)))
      out
    }

    v2_log(sprintf("Label-sensitivity bootstrap: %s (%d levels)", svar, nlevels(lvl_f)))
    b_ls  <- boot_run(n_k, BOOT_B_SUBGROUP, stat_ls, strata = lvl_k,
                      seed = BOOT_SEED + 20L, progress_every = 1000L)
    obs_ls <- stat_ls(rep(1L, n_k))

    for (lv in levels(lvl_f)) {
      r <- list(subgroup_col = svar, subgroup_val = lv,
                n_stays = sum(lvl_k == lv), is_reference = lv == ref_lvl)
      for (v in names(coh)) {
        nm <- paste0("pct_", v, "||", lv)
        ci <- perc_ci(b_ls[, nm], CI_LEVEL)
        r[[paste0("pct_", v)]]    <- unname(obs_ls[nm])
        r[[paste0("pct_", v, "_lo")]] <- unname(ci["lo"])
        r[[paste0("pct_", v, "_hi")]] <- unname(ci["hi"])
      }
      nm <- paste0("range||", lv)
      ci <- perc_ci(b_ls[, nm], CI_LEVEL)
      r$range_pp <- unname(obs_ls[nm]); r$range_lo <- unname(ci["lo"]); r$range_hi <- unname(ci["hi"])
      ls_rows[[paste(svar, lv)]] <- as.data.table(r)

      if (lv != ref_lvl) {
        dvec <- b_ls[, nm] - b_ls[, paste0("range||", ref_lvl)]
        dobs <- unname(obs_ls[nm] - obs_ls[paste0("range||", ref_lvl)])
        dci  <- perc_ci(dvec, CI_LEVEL)
        ls_diff[[paste(svar, lv)]] <- data.table(
          subgroup_col = svar, subgroup_val = lv, reference = ref_lvl,
          n_stays = sum(lvl_k == lv), delta_range_pp = dobs,
          ci_lo = dci["lo"], ci_hi = dci["hi"],
          p_raw = boot_pvalue(dvec, dobs, 0, "two.sided"))
      }
    }
  }
}

ls_ci   <- rbindlist(ls_rows, fill = TRUE)
ls_delta <- rbindlist(ls_diff, fill = TRUE)
if (nrow(ls_delta) > 0) {
  ls_delta[, p_bh := p.adjust(p_raw, method = "BH")]
  ls_delta[, p_bonf := p.adjust(p_raw, method = "bonferroni")]
  ls_delta[, significant_bh := p_bh < FDR_Q]
  ls_delta[, family := sprintf("exploratory E2 (m=%d label-sensitivity contrasts)", nrow(ls_delta))]
  setorder(ls_delta, p_raw)
}
if (nrow(ls_ci) > 0) {
  ls_ci[, family := "exploratory E2"]
  save_parquet(ls_ci, "12_subgroup_labelsens_ci", DATA_DIR)
  save_csv(ls_ci,     "12_subgroup_labelsens_ci", DATA_DIR)
}
if (nrow(ls_delta) > 0) {
  save_parquet(ls_delta, "12_subgroup_labelsens_contrasts", DATA_DIR)
  save_csv(ls_delta,     "12_subgroup_labelsens_contrasts", DATA_DIR)
}

if (nrow(ls_ci) > 0)
  knitr::kable(
    ls_ci[subgroup_col == "race"][order(-n_stays),
      .(subgroup = subgroup_val, n_stays,
        A = sprintf("%.1f %s", pct_A, fmt_ci(pct_A_lo, pct_A_hi, 1)),
        B = sprintf("%.1f %s", pct_B, fmt_ci(pct_B_lo, pct_B_hi, 1)),
        C = sprintf("%.1f %s", pct_C, fmt_ci(pct_C_lo, pct_C_hi, 1)),
        range = sprintf("%.1f %s", range_pp, fmt_ci(range_lo, range_hi, 1)))],
    caption = "Percentage of stays labelled septic by race and label variant, with 95\\% stratified cluster-bootstrap intervals")
```

```{r label-sens-contrasts}
if (nrow(ls_delta) > 0)
  knitr::kable(
    ls_delta[subgroup_col == "race"][order(p_raw),
      .(subgroup = subgroup_val, reference, n_stays,
        delta_pp = round(delta_range_pp, 2), CI = fmt_ci(ci_lo, ci_hi, 2),
        p = formatC(p_raw, format = "f", digits = 3),
        q_BH = formatC(p_bh, format = "f", digits = 3), significant_bh)],
    caption = "Difference in label sensitivity (A--C range) vs.\\ the reference group, Benjamini-Hochberg corrected across exploratory family E2")
```

## Exploratory Family E3: Alternative-Label Contrasts

Families E1 and E2 vary the label only *within* Sepsis-3, so they cannot say
whether the label sensitivity is empirical or definitional. This family attaches
intervals to the Protocol Amendment 2 comparison: Variant E is Sepsis-3's
treatment anchor with the organ-dysfunction requirement removed, Variant D is
the organ-dysfunction limb with no treatment timestamp at all. Protocol
Amendment 4 adds Variant F, which keeps the Sepsis-3 conjunction intact and
changes only the onset timestamp to the standard
$\min(t_{\text{susp}}, t_{\text{SOFA}})$; its contrast against B is therefore a
pure timing effect. Adding it enlarges family E3, so every BH-adjusted p-value
in this family is computed over the larger set and is correspondingly stricter.

All variants are labelled on the same test stays, so one replicate draw applies
to all of them and each contrast is paired within replicate. Reported alongside
the differences is the **anchor-attributable fraction**
$\phi = (\mathrm{AUROC}_E - 0.5)/(\mathrm{AUROC}_B - 0.5)$, the share of Variant
B's above-chance discrimination that the treatment anchor reproduces on its own.
$\phi$ is a ratio with no meaningful null at zero, so it is reported as an
interval and is not part of the corrected family.

```{r alt-label-ci}
alt_ids <- setNames(names(ALT_LABEL_VARIANTS), names(ALT_LABEL_VARIANTS))
alt_tab <- lapply(alt_ids, load_test_table)
alt_tab <- Filter(Negate(is.null), alt_tab)

alt_ci <- data.table(); alt_delta <- data.table()

if (length(alt_tab) == 0) {
  cat("No alternative-label predictions found; run 05-07 with the alternative-label variants.\n")
} else {
  ALT_MODELS <- c(primary = "pred_sepsis", gbt = "pred_gbt")

  # Reuse the A/B/C resampling frame so the contrasts against B are paired.
  pre_alt <- list()
  for (v in names(alt_tab)) {
    d <- alt_tab[[v]]
    d[, sidx := match(stay_id, stay_frame)]
    n_off <- d[is.na(sidx), uniqueN(stay_id)]
    if (n_off > 0)
      v2_log(sprintf("Variant %s: %d test stays outside the common frame, dropped",
                     v, n_off), level = "WARN")
    d <- d[!is.na(sidx)]
    alt_tab[[v]] <- d
    for (m in names(ALT_MODELS)) {
      col <- ALT_MODELS[[m]]
      if (col %in% names(d))
        pre_alt[[paste(v, m, sep = "_")]] <- auc_precompute(d[[col]], d$y, d$sidx, N_STAYS)
    }
  }
  for (m in names(ALT_MODELS))
    if (!is.null(pre_B[[m]])) pre_alt[[paste("B", m, sep = "_")]] <- pre_B[[m]]
  pre_alt <- Filter(Negate(is.null), pre_alt)

  stat_alt <- function(w) vapply(pre_alt, boot_auroc_w, numeric(1), w_stay = w)

  t_start  <- proc.time()
  boot_alt <- boot_run(N_STAYS, BOOT_B_MAIN, stat_alt, seed = BOOT_SEED + 30L,
                       progress_every = 500L)
  v2_log(sprintf("Alternative-label bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
  obs_alt <- stat_alt(ones)

  n_events_of <- function(v, m) {
    d <- if (v == "B") tb else alt_tab[[v]]
    if (is.null(d)) NA_integer_ else sum(d$y)
  }

  alt_ci <- rbindlist(lapply(names(pre_alt), function(k) {
    parts <- strsplit(k, "_", fixed = TRUE)[[1]]
    v <- parts[1]; m <- parts[2]
    ci <- perc_ci(boot_alt[, k], CI_LEVEL)
    data.table(variant = v, label = variant_name(v), label_family = variant_type(v),
               model = m, n_rows = pre_alt[[k]]$n, n_events = n_events_of(v, m),
               auroc = unname(obs_alt[k]), ci_lo = ci["lo"], ci_hi = ci["hi"])
  }))
  alt_ci[, `:=`(auroc_ci = fmt_ci(ci_lo, ci_hi), family = "exploratory E3")]
  setorder(alt_ci, model, variant)

  # Contrasts against Sepsis-3 (Variant B), paired within replicate
  rows <- list()
  for (m in names(ALT_MODELS)) {
    kb <- paste("B", m, sep = "_")
    if (!kb %in% colnames(boot_alt)) next
    for (v in names(alt_tab)) {
      kv <- paste(v, m, sep = "_")
      if (!kv %in% colnames(boot_alt)) next
      dvec <- boot_alt[, kv] - boot_alt[, kb]
      dobs <- unname(obs_alt[kv] - obs_alt[kb])
      dci  <- perc_ci(dvec, CI_LEVEL)
      rows[[kv]] <- data.table(
        variant = v, label = variant_name(v), model = m,
        delta_auroc = dobs, ci_lo = dci["lo"], ci_hi = dci["hi"],
        p_raw = boot_pvalue(dvec, dobs, 0, "two.sided"))
    }
  }
  alt_delta <- rbindlist(rows)
  if (nrow(alt_delta) > 0) {
    alt_delta[, p_bh := p.adjust(p_raw, method = "BH")]
    alt_delta[, p_bonf := p.adjust(p_raw, method = "bonferroni")]
    alt_delta[, significant_bh := p_bh < FDR_Q]
    alt_delta[, family := sprintf("exploratory E3 (m=%d alternative-label contrasts)",
                                  nrow(alt_delta))]
    setorder(alt_delta, p_raw)
  }

  # Anchor-attributable fraction, with the ratio bootstrapped end to end
  phi_rows <- list()
  for (m in names(ALT_MODELS)) {
    kb <- paste("B", m, sep = "_"); ke <- paste("E", m, sep = "_")
    if (!all(c(kb, ke) %in% colnames(boot_alt))) next
    denom <- boot_alt[, kb] - 0.5
    pvec  <- ifelse(denom > 0, (boot_alt[, ke] - 0.5) / denom, NA_real_)
    pci   <- perc_ci(pvec, CI_LEVEL)
    phi_rows[[m]] <- data.table(
      model = m,
      auroc_B = unname(obs_alt[kb]), auroc_E = unname(obs_alt[ke]),
      phi_anchor = unname((obs_alt[ke] - 0.5) / (obs_alt[kb] - 0.5)),
      ci_lo = pci["lo"], ci_hi = pci["hi"])
  }
  phi_dt <- rbindlist(phi_rows)

  save_parquet(alt_ci, "12_altlabel_auroc_ci", DATA_DIR)
  save_csv(alt_ci,     "12_altlabel_auroc_ci", DATA_DIR)
  if (nrow(alt_delta) > 0) {
    save_parquet(alt_delta, "12_altlabel_contrasts", DATA_DIR)
    save_csv(alt_delta,     "12_altlabel_contrasts", DATA_DIR)
  }
  if (nrow(phi_dt) > 0) {
    save_parquet(phi_dt, "12_anchor_attribution_ci", DATA_DIR)
    save_csv(phi_dt,     "12_anchor_attribution_ci", DATA_DIR)
  }

  knitr::kable(
    alt_ci[, .(variant, label, model, n_events,
               auroc = round(auroc, 4), CI = auroc_ci)],
    caption = "AUROC under each limb of the Sepsis-3 conjunction, with 95\\% stay-level cluster-bootstrap intervals (temporal test)")

  if (nrow(phi_dt) > 0)
    knitr::kable(
      phi_dt[, .(model, auroc_B = round(auroc_B, 4), auroc_E = round(auroc_E, 4),
                 phi_anchor = round(phi_anchor, 3), CI = fmt_ci(ci_lo, ci_hi))],
      caption = "Anchor-attributable fraction: share of Variant B's above-chance AUROC reproduced by the treatment anchor alone")
}
```

```{r alt-label-contrast-table}
if (nrow(alt_delta) > 0)
  knitr::kable(
    alt_delta[, .(variant, label, model, delta = round(delta_auroc, 4),
                  CI = fmt_ci(ci_lo, ci_hi),
                  p = formatC(p_raw, format = "f", digits = 4),
                  q_BH = formatC(p_bh, format = "f", digits = 4), significant_bh)],
    caption = "AUROC difference vs.\\ Sepsis-3 (Variant B), Benjamini-Hochberg corrected across exploratory family E3")
```

## Intervals Added Under Protocol Amendment 10

Three quantities were reported as point estimates in earlier rounds, and each
carried a claim in the thesis prose that a point estimate cannot support: that
dropping the action-derived features leaves discrimination "essentially
unaffected", that the three label definitions make no practical difference to
calibration, and that the frozen models degrade on transport to the full
external cohort under Variant D. Amendment 10 attaches an interval to each.
None of them enters a corrected family: the ablation arm and the Variant D arm
are excluded from every family by Amendments 5 and 9 respectively, and the
calibration contrasts are estimation, not testing.

### Action-Derived Feature Ablation (Amendment 5 arm)

Both arms are fitted on the same training stays and scored on the same test
stays, so the difference is paired within replicate and the interval is on the
difference rather than on two marginal estimates that happen to overlap.

```{r ablation-ci}
abl_ci <- data.table()
abl_pre <- list(); abl_obs <- c()

for (vid in names(test_tab)) {
  full <- test_tab[[vid]]
  ap <- file.path(DATA_DIR, sprintf("05_predictions_primary_ablated_%s.parquet", vid))
  ag <- file.path(DATA_DIR, sprintf("06_predictions_gbt_ablated_%s.parquet", vid))
  for (m in c("primary", "gbt")) {
    path <- if (m == "primary") ap else ag
    col  <- if (m == "primary") "pred_sepsis" else "pred_gbt"
    if (!file.exists(path) || !col %in% names(full)) next
    ad <- as.data.table(read_parquet(path))
    if (!col %in% names(ad)) next
    ad[, y := as.integer(event == 1)]
    ad[, sidx := match(stay_id, stay_frame)]
    ad <- ad[!is.na(sidx)]
    abl_pre[[paste(vid, m, "full", sep = "_")]] <-
      auc_precompute(full[[col]], full$y, full$sidx, N_STAYS)
    abl_pre[[paste(vid, m, "abl", sep = "_")]] <-
      auc_precompute(ad[[col]], ad$y, ad$sidx, N_STAYS)
    rm(ad); gc()
  }
}
abl_pre <- Filter(Negate(is.null), abl_pre)

if (length(abl_pre) > 0) {
  stat_abl <- function(w) vapply(abl_pre, boot_auroc_w, numeric(1), w_stay = w)
  t_start  <- proc.time()
  boot_abl <- boot_run(N_STAYS, BOOT_B_MAIN, stat_abl, seed = BOOT_SEED + 40L,
                       progress_every = 500L)
  v2_log(sprintf("Ablation bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
  abl_obs <- stat_abl(ones)

  keys <- grep("_full$", names(abl_pre), value = TRUE)
  abl_ci <- rbindlist(lapply(keys, function(kf) {
    ka <- sub("_full$", "_abl", kf)
    if (!ka %in% colnames(boot_abl)) return(NULL)
    parts <- strsplit(kf, "_", fixed = TRUE)[[1]]
    dvec <- boot_abl[, ka] - boot_abl[, kf]
    dobs <- unname(abl_obs[ka] - abl_obs[kf])
    dci  <- perc_ci(dvec, CI_LEVEL)
    data.table(variant = parts[1], model = parts[2],
               auroc_full = unname(abl_obs[kf]), auroc_ablated = unname(abl_obs[ka]),
               delta_auroc = dobs, ci_lo = dci["lo"], ci_hi = dci["hi"],
               excludes_zero = dci["lo"] > 0 | dci["hi"] < 0)
  }))
  if (nrow(abl_ci) > 0) {
    abl_ci[, delta_ci := fmt_ci(ci_lo, ci_hi, 4)]
    abl_ci[, family := "none (Amendment 5 arm; enters no corrected family)"]
    setorder(abl_ci, model, variant)
    save_parquet(abl_ci, "12_ablation_ci", DATA_DIR)
    save_csv(abl_ci,     "12_ablation_ci", DATA_DIR)
    knitr::kable(abl_ci[, .(variant, model,
                            full = round(auroc_full, 4),
                            ablated = round(auroc_ablated, 4),
                            delta = round(delta_auroc, 4), delta_ci, excludes_zero)],
                 caption = "Paired ablation contrast (ablated minus full) with 95\\% stay-level cluster-bootstrap intervals")
  }
}
```

### Calibration Across Label Definitions

The calibration intercept and slope are refit inside each replicate on the same
stays as every other quantity here, so the cross-label differences are paired.
The claim under test is the thesis's, that the label definition does not change
how faithfully the fitted hazard tracks the observed rate; the contrast is
therefore each variant against the primary label B.

```{r calibration-ci}
calib_ci <- data.table(); calib_delta <- data.table()

cal_pre <- list()
for (vid in names(test_tab)) {
  d <- test_tab[[vid]]
  cal_pre[[vid]] <- calib_precompute(d$pred_sepsis, d$y, d$sidx, N_STAYS)
}
cal_pre <- Filter(Negate(is.null), cal_pre)

if (length(cal_pre) > 0) {
  # Warm-start values: the full-data fit. Every replicate starts from here, so
  # the IRLS converges in two or three steps instead of eight.
  cal_start <- lapply(cal_pre, function(pr)
    list(slope = boot_calib_slope_w(pr, ones),
         int   = boot_calib_int_w(pr, ones)))

  stat_cal <- function(w) {
    out <- numeric(0)
    for (v in names(cal_pre)) {
      sl <- boot_calib_slope_w(cal_pre[[v]], w, start = cal_start[[v]]$slope)
      it <- boot_calib_int_w(cal_pre[[v]], w, start = cal_start[[v]]$int)
      out <- c(out, setNames(c(sl[2], it), paste0(c("slope_", "int_"), v)))
    }
    out
  }

  t_start  <- proc.time()
  boot_cal <- boot_run(N_STAYS, BOOT_B_MAIN, stat_cal, seed = BOOT_SEED + 50L,
                       progress_every = 250L)
  v2_log(sprintf("Calibration bootstrap: %.1f s", (proc.time() - t_start)["elapsed"]))
  obs_cal <- stat_cal(ones)

  calib_ci <- rbindlist(lapply(names(cal_pre), function(v) {
    cs <- perc_ci(boot_cal[, paste0("slope_", v)], CI_LEVEL)
    ci <- perc_ci(boot_cal[, paste0("int_",   v)], CI_LEVEL)
    data.table(variant = v, model = "primary", n_rows = cal_pre[[v]]$n,
               calib_slope = unname(obs_cal[paste0("slope_", v)]),
               slope_lo = cs["lo"], slope_hi = cs["hi"],
               calib_int = unname(obs_cal[paste0("int_", v)]),
               int_lo = ci["lo"], int_hi = ci["hi"])
  }))
  calib_ci[, `:=`(slope_ci = fmt_ci(slope_lo, slope_hi),
                  int_ci   = fmt_ci(int_lo, int_hi))]

  # Contrasts against the primary label, paired within replicate.
  if ("B" %in% names(cal_pre)) {
    calib_delta <- rbindlist(lapply(setdiff(names(cal_pre), "B"), function(v) {
      ds <- boot_cal[, paste0("slope_", v)] - boot_cal[, "slope_B"]
      di <- boot_cal[, paste0("int_",   v)] - boot_cal[, "int_B"]
      dso <- unname(obs_cal[paste0("slope_", v)] - obs_cal["slope_B"])
      dio <- unname(obs_cal[paste0("int_",   v)] - obs_cal["int_B"])
      cs <- perc_ci(ds, CI_LEVEL); ci <- perc_ci(di, CI_LEVEL)
      data.table(variant = v, reference = "B", model = "primary",
                 delta_slope = dso, slope_lo = cs["lo"], slope_hi = cs["hi"],
                 slope_excludes_zero = cs["lo"] > 0 | cs["hi"] < 0,
                 delta_int = dio, int_lo = ci["lo"], int_hi = ci["hi"],
                 int_excludes_zero = ci["lo"] > 0 | ci["hi"] < 0)
    }))
    calib_delta[, `:=`(slope_ci = fmt_ci(slope_lo, slope_hi),
                       int_ci   = fmt_ci(int_lo, int_hi))]
  }

  save_parquet(calib_ci, "12_calibration_ci", DATA_DIR)
  save_csv(calib_ci,     "12_calibration_ci", DATA_DIR)
  if (nrow(calib_delta) > 0) {
    save_parquet(calib_delta, "12_calibration_contrasts", DATA_DIR)
    save_csv(calib_delta,     "12_calibration_contrasts", DATA_DIR)
  }

  knitr::kable(calib_ci[, .(variant, slope = round(calib_slope, 4), slope_ci,
                            intercept = round(calib_int, 4), int_ci)],
               caption = "Primary-model calibration slope and intercept by label variant, with 95\\% stay-level cluster-bootstrap intervals")
  if (nrow(calib_delta) > 0)
    knitr::kable(calib_delta[, .(variant, reference,
                                 delta_slope = round(delta_slope, 4), slope_ci,
                                 slope_excludes_zero,
                                 delta_int = round(delta_int, 4), int_ci,
                                 int_excludes_zero)],
                 caption = "Paired cross-label calibration contrasts against Variant B")
}
```

### Variant D Transport to the Full External Cohort (Amendment 9 arm)

The internal and external cohorts are disjoint databases, so the two bootstrap
streams are drawn separately and differenced, exactly as for H5. This arm is
the only external estimate in the study that clears the interpretability floor,
and it was the only one carrying no interval.

```{r altlabel-transport-ci}
alt_transport <- data.table()
alt_ext_path  <- file.path(DATA_DIR, "08_eicu_predictions_D.parquet")

if (file.exists(alt_ext_path) && "D" %in% names(alt_tab)) {
  ed <- as.data.table(read_parquet(alt_ext_path))
  ed[, y := as.integer(event == 1)]
  setorder(ed, stay_id, hour)
  estays_d <- sort(unique(ed$stay_id)); ed[, sidx := match(stay_id, estays_d)]
  di <- alt_tab[["D"]]

  ALT_T_MODELS <- c(primary = "pred_sepsis", gbt = "pred_gbt")
  rows_t <- list()
  for (m in names(ALT_T_MODELS)) {
    col <- ALT_T_MODELS[[m]]
    if (!col %in% names(ed) || !col %in% names(di)) next
    pre_e <- auc_precompute(ed[[col]], ed$y, ed$sidx, length(estays_d))
    pre_i <- auc_precompute(di[[col]], di$y, di$sidx, N_STAYS)
    if (is.null(pre_e) || is.null(pre_i)) next

    b_e <- boot_run(length(estays_d), BOOT_B_EXTERNAL,
                    function(w) c(auroc = boot_auroc_w(pre_e, w)),
                    seed = BOOT_SEED + 60L, progress_every = 250L)[, "auroc"]
    b_i <- boot_run(N_STAYS, BOOT_B_EXTERNAL,
                    function(w) c(auroc = boot_auroc_w(pre_i, w)),
                    seed = BOOT_SEED + 61L, progress_every = 250L)[, "auroc"]

    a_e <- boot_auroc_w(pre_e, rep(1L, length(estays_d)))
    a_i <- boot_auroc_w(pre_i, ones)
    dvec <- b_i - b_e
    dci  <- perc_ci(dvec, CI_LEVEL)
    rows_t[[m]] <- data.table(
      variant = "D", model = m,
      n_ext_stays = length(estays_d), n_ext_rows = pre_e$n, n_ext_events = sum(ed$y),
      auroc_internal = a_i, auroc_external = a_e,
      delta_auroc = a_i - a_e, ci_lo = dci["lo"], ci_hi = dci["hi"],
      excludes_zero = dci["lo"] > 0 | dci["hi"] < 0)
    rm(pre_e, pre_i); gc()
  }
  alt_transport <- rbindlist(rows_t)
  if (nrow(alt_transport) > 0) {
    alt_transport[, delta_ci := fmt_ci(ci_lo, ci_hi)]
    alt_transport[, family := "none (Amendment 9 arm; excluded from H5)"]
    save_parquet(alt_transport, "12_altlabel_transport_ci", DATA_DIR)
    save_csv(alt_transport,     "12_altlabel_transport_ci", DATA_DIR)
    knitr::kable(alt_transport[, .(model, n_ext_events,
                                   internal = round(auroc_internal, 4),
                                   external = round(auroc_external, 4),
                                   delta = round(delta_auroc, 4), delta_ci,
                                   excludes_zero)],
                 caption = "Variant D internal-to-external AUROC transport on the full eICU-CRD cohort, with 95\\% cluster-bootstrap intervals")
  }
  rm(ed); gc()
} else {
  cat("Variant D external predictions not found; run stage 08 to write 08_eicu_predictions_D.\n")
}
```

## Subgroup Interval Widths

The point of attaching intervals to the equity tables is that a subgroup with
few events cannot support a claim in either direction. This figure plots each
subgroup AUROC with its interval against the number of sepsis onsets in that
subgroup.

```{r subgroup-ci-plot, fig.cap="Subgroup AUROC with 95\\% cluster-bootstrap intervals (Variant B, temporal test). Interval width is driven almost entirely by the subgroup event count; the smallest groups cannot distinguish their AUROC from chance.", fig.width=8, fig.height=5.5}
if (nrow(sub_auroc) > 0) {
  pd <- sub_auroc[subgroup_col == "race"]
  if (nrow(pd) > 0) {
    pd[, lbl := paste0(subgroup_val, " (", n_events, ")")]
    p_ci <- ggplot(pd, aes(x = reorder(lbl, auroc), y = auroc, colour = model)) +
      geom_hline(yintercept = 0.5, linetype = "dotted", colour = "gray40") +
      geom_pointrange(aes(ymin = ci_lo, ymax = ci_hi),
                      position = position_dodge(width = 0.5), size = 0.35) +
      coord_flip() +
      labs(title    = "Subgroup AUROC with 95% cluster-bootstrap intervals",
           subtitle = "Race, Variant B, temporal test. Sepsis onset count in parentheses.",
           x = "Racial subgroup (n events)", y = "AUROC", colour = "Model") +
      theme_bw() + theme(legend.position = "bottom")
    print(p_ci)
    ggsave(file.path(FIGURES_DIR, "12_subgroup_auroc_ci.png"), p_ci,
           width = 8, height = 5.5, dpi = 150)
  }
}
```

## Analysis Register

Every analysis reported in the thesis is tagged here. `confirmatory` means the
estimand, threshold, and direction were fixed in `00_preregistration.md` before
any modelling. `exploratory` means the analysis was pre-registered in *scope*
but not in specific contrasts, so it generates hypotheses rather than testing
them. `descriptive` means no inferential claim is attached.

```{r register}
register <- rbindlist(list(
  data.table(analysis = "Cohort description by label variant",           status = "descriptive",
             prereg = "§2",        correction = "none",  reported_in = "Table: cohort"),
  data.table(analysis = "Sample size adequacy (Riley et al.)",           status = "confirmatory",
             prereg = "§10",       correction = "none (single pre-specified criterion)",
             reported_in = "Table: sample_size"),
  data.table(analysis = "Internal temporal performance (all models)",    status = "confirmatory",
             prereg = "§5, §6",    correction = "none (estimation, not testing)",
             reported_in = "Table: internal_results"),
  data.table(analysis = "Calibration intercept and slope",               status = "confirmatory",
             prereg = "§5",        correction = "none (estimation)",
             reported_in = "Table: calibration"),
  data.table(analysis = "Decision-curve net benefit",                    status = "confirmatory",
             prereg = "§6",        correction = "none (estimation)",
             reported_in = "Table: net_benefit"),
  data.table(analysis = "Operating points at fixed sensitivity",         status = "exploratory",
             prereg = "not specified (threshold chosen post hoc)",
             correction = "none; reported as descriptive operating characteristics",
             reported_in = "Table: operating_points"),
  data.table(analysis = "Per-patient detection and lead time",           status = "confirmatory",
             prereg = "§6",        correction = "none (estimation)",
             reported_in = "Table: patient_level, lead_time"),
  data.table(analysis = "H1 label dominance",                            status = "confirmatory",
             prereg = "§7",        correction = "Holm (m=6)", reported_in = "H1 box"),
  data.table(analysis = "H2 instability uncertainty diagnostic",         status = "exploratory",
             prereg = "Amendment 7 (post-hoc)",
             correction = "none; ratio to pooled SE, not a test",
             reported_in = "Table: h2_uncertainty"),
  data.table(analysis = "H2 coefficient instability",                    status = "confirmatory",
             prereg = "§7",        correction = "family slot reserved; no test statistic",
             reported_in = "H2 box, Table: coef_instability"),
  data.table(analysis = "H3 treatment leakage",                          status = "confirmatory",
             prereg = "§7",        correction = "family slot reserved; exact bound reported",
             reported_in = "H3 box"),
  data.table(analysis = "H4 split optimism",                             status = "confirmatory",
             prereg = "§7",        correction = "Holm (m=6)", reported_in = "H4 box"),
  data.table(analysis = "H5 metric divergence",                          status = "confirmatory",
             prereg = "§7",        correction = "Holm (m=6)", reported_in = "H5 box"),
  data.table(analysis = "H6 model-class null",                           status = "confirmatory",
             prereg = "§7",        correction = "Holm (m=6)", reported_in = "H6 box"),
  data.table(analysis = "External validation on eICU-CRD",               status = "confirmatory",
             prereg = "§4",        correction = "none (estimation); feeds H5",
             reported_in = "Table: external"),
  data.table(analysis = "Variance decomposition table",                  status = "confirmatory",
             prereg = "§1",        correction = "none (descriptive assembly of H1-H6 contrasts)",
             reported_in = "Table: decomposition"),
  data.table(analysis = "Proportional-hazards (Schoenfeld) testing",     status = "confirmatory",
             prereg = "§8",        correction = "none (single pre-specified check)",
             reported_in = "PH section"),
  data.table(analysis = "GBT feature importance",                        status = "exploratory",
             prereg = "not specified", correction = "none (ranking, no inference)",
             reported_in = "Table: feature_importance"),
  data.table(analysis = "Subgroup AUROC by race/language/insurance",     status = "exploratory",
             prereg = "§9 (scope only; contrasts not pre-specified)",
             correction = "Benjamini-Hochberg within family E1",
             reported_in = "Table: equity_performance"),
  data.table(analysis = "Subgroup label sensitivity across variants",    status = "exploratory",
             prereg = "§9 (scope only; contrasts not pre-specified)",
             correction = "Benjamini-Hochberg within family E2",
             reported_in = "Table: equity_label"),
  data.table(analysis = "Label sensitivity vs. AUROC association",       status = "exploratory",
             prereg = "not specified", correction = "none (visual, no test)",
             reported_in = "Figure: label_sens_vs_auroc"),
  data.table(analysis = "Alternative-label AUROC vs. Sepsis-3 (variants D, E)", status = "exploratory",
             prereg = "Protocol Amendment 2 (post hoc; §16-17)",
             correction = "Benjamini-Hochberg within family E3",
             reported_in = "Table: altlabel_contrasts"),
  data.table(analysis = "Standard onset timing min(t_susp, t_SOFA) (variant F)", status = "exploratory",
             prereg = "Protocol Amendment 4 (post hoc; §18)",
             correction = "Benjamini-Hochberg within family E3 (family enlarged)",
             reported_in = "Table: altlabel_contrasts, H3 section"),
  data.table(analysis = "Pre-treatment onsets under standard onset timing",     status = "descriptive",
             prereg = "Protocol Amendment 4 (post hoc; §18)",
             correction = "none (event count and window size, not a test)",
             reported_in = "Table: pretreatment_onsets"),
  data.table(analysis = "Anchor-attributable fraction of discrimination", status = "exploratory",
             prereg = "Protocol Amendment 2 (post hoc; §16-17)",
             correction = "none (ratio reported as an interval, no null at zero)",
             reported_in = "Table: anchor_attribution_ci"),
  data.table(analysis = "Onset agreement between label definitions",     status = "descriptive",
             prereg = "Protocol Amendment 2 (post hoc; §16-17)",
             correction = "none (descriptive agreement statistics)",
             reported_in = "Table: label_agreement"),
  # Amendments 5-7 were absent from the register, so the register and the
  # chapter's own register table disagreed about how many analyses had no
  # machine-generated counterpart. Amendment 6 is the load-bearing one: it is
  # the only place the reduction of family E1 from 30 computed contrasts to 6
  # is recorded in a file the pipeline writes.
  data.table(analysis = "Subgroup interpretability floor (family E1 reduction)",
             status = "exploratory",
             prereg = "Protocol Amendment 6 (post hoc; §20)",
             correction = sprintf(paste0("E1 restricted to contrasts in which both strata carry >= %d ",
                                         "events; %s contrasts retained. Below-floor strata are still ",
                                         "estimated and tabulated, and are excluded from the inferential ",
                                         "family only"),
                                  MIN_SUBGROUP_EVENTS_INTERPRET,
                                  if (exists("sub_delta") && !is.null(sub_delta))
                                    format(nrow(sub_delta)) else "?"),
             reported_in = "Table: equity_contrasts"),
  data.table(analysis = "Action-derived feature ablation (Amendment 5)", status = "exploratory",
             prereg = "Protocol Amendment 5 (post hoc; §19)",
             correction = "none (estimation); enters no corrected family",
             reported_in = "Table: ablation"),
  # Amendment 9's external arm had no register row at all, which is how it came
  # to be printed under a confirmatory tag in the thesis while its own table
  # said post-hoc. Registered here as descriptive, matching §23.2.
  data.table(analysis = "Variant D external validation on the full eICU-CRD cohort (Amendment 9)",
             status = "descriptive",
             prereg = "Protocol Amendment 9 (post hoc; §23)",
             correction = "none; joins no corrected family and is excluded from H5",
             reported_in = "Table: external_altlabel"),
  data.table(analysis = "Variant D internal-to-external transport interval (Amendment 10)",
             status = "descriptive",
             prereg = "Protocol Amendment 10 (post hoc; §24)",
             correction = "none (estimation); excluded from H5 on Amendment 9's terms",
             reported_in = "Table: altlabel_transport_ci"),
  data.table(analysis = "Ablation contrast intervals (Amendment 10)", status = "exploratory",
             prereg = "Protocol Amendment 10 (post hoc; §24)",
             correction = "none (estimation); enters no corrected family",
             reported_in = "Table: ablation_ci"),
  data.table(analysis = "Cross-label calibration contrast intervals (Amendment 10)",
             status = "exploratory",
             prereg = "Protocol Amendment 10 (post hoc; §24)",
             correction = "none (estimation); enters no corrected family",
             reported_in = "Table: calibration_contrasts")
  # Amendment 7 is NOT added here: the diagnostic block above already registers
  # it as "H2 instability uncertainty diagnostic". A second row was briefly
  # added in this round and removed -- a duplicated register entry is exactly
  # the kind of thing the register exists to make impossible.
))

save_parquet(register, "12_analysis_register", DATA_DIR)
save_csv(register,     "12_analysis_register", DATA_DIR)

knitr::kable(register, caption = "Analysis register: confirmatory vs.\\ exploratory status and the multiplicity correction applied to each")
```

## Reading the Corrected Family

Four things govern how the tables above should be read.

1. **The family is deliberately larger than the number of testable
   hypotheses.** Holm runs at $m = 6$, the count pre-registered, not the count
   that survived contact with the data. H2 and H3 hold slots without supplying
   a p-value. A hypothesis that becomes untestable after unblinding must not
   loosen the correction on the ones that remain, so the cost of that choice is
   carried here rather than passed to the surviving hypotheses.
2. **A non-confirmation is not a null result.** Failing to meet a
   pre-registered threshold and finding no effect are different outcomes, and
   the direction of a miss carries information the p-value does not. Where an
   estimate lands on the opposite side of its null from the one predicted, the
   interval is what shows it; the verdict column alone will not.
3. **Non-inferiority is read off the interval, not the test.** For H6 the
   question is whether the confidence limit falls inside the pre-registered
   margin. A large p-value on its own establishes nothing, because failing to
   reject is not evidence of equivalence. H6's null is an interval with two
   boundaries, and Amendment 10 tests both: the upper limb is large whenever
   the estimate is negative, so it is the lower limb that carries whatever
   information the test has.
4. **Exploratory families generate hypotheses and do not test them.** E1, E2
   and E3 are corrected by Benjamini--Hochberg within themselves and never
   pooled with the confirmatory family. A contrast surviving BH is a candidate
   worth a future study, not a finding this design can claim.

## Summary

The confirmatory family as it stands after correction, printed for the run log.
The thesis reads these values from the saved tables, not from this output.

```{r summary}
cat("Confirmatory family (Holm, m = 6, FWER = 0.05):\n")
for (i in seq_len(nrow(conf)))
  cat(sprintf("  %s  %-26s est=%s  p_raw=%s  p_holm=%s  -> %s\n",
              conf$hypothesis[i], conf$claim[i],
              formatC(conf$estimate[i], format = "g", digits = 4),
              ifelse(is.na(conf$p_raw[i]),  "--", formatC(conf$p_raw[i],  format = "g", digits = 2)),
              ifelse(is.na(conf$p_holm[i]), "--", formatC(conf$p_holm[i], format = "g", digits = 2)),
              conf$verdict[i]))

if (nrow(sub_delta) > 0) {
  cat(sprintf("\nExploratory family E1: %d subgroup AUROC contrasts, %d significant at BH q < %.2f\n",
              nrow(sub_delta), sum(sub_delta$significant_bh), FDR_Q))
  cat(sprintf("  Smallest raw p = %.4f (%s, %s vs %s); smallest BH q = %.4f\n",
              min(sub_delta$p_raw), sub_delta$subgroup_val[1], sub_delta$model[1],
              sub_delta$reference[1], min(sub_delta$p_bh)))
  cat(sprintf("  Median 95%% CI width for subgroup AUROC: %.3f (max %.3f)\n",
              median(sub_auroc$ci_width, na.rm = TRUE), max(sub_auroc$ci_width, na.rm = TRUE)))
}
if (nrow(ls_delta) > 0)
  cat(sprintf("\nExploratory family E2: %d label-sensitivity contrasts, %d significant at BH q < %.2f\n",
              nrow(ls_delta), sum(ls_delta$significant_bh), FDR_Q))

if (nrow(alt_delta) > 0) {
  cat(sprintf("\nExploratory family E3: %d alternative-label contrasts, %d significant at BH q < %.2f\n",
              nrow(alt_delta), sum(alt_delta$significant_bh), FDR_Q))
  for (i in seq_len(nrow(alt_delta)))
    cat(sprintf("  %s (%s) vs B, %-7s delta=%+.4f %s  q_BH=%.4f\n",
                alt_delta$variant[i], alt_delta$label[i], alt_delta$model[i],
                alt_delta$delta_auroc[i],
                fmt_ci(alt_delta$ci_lo[i], alt_delta$ci_hi[i]), alt_delta$p_bh[i]))
}

cat("\nInference notebook complete.\n")
```
